### 2024-10-24 template/edumer_template.qmd
Create template first version

---
pagetitle: "EDUMER Template"
Author: Equipo EDUMER
bibliography: "../input/bib/refs.bib"
csl: "../input/bib/apa6.csl"
lang: es

knitr:
  opts_chunk:
    collapse: true
    comment: "#>"
    R.options:
      knitr.graphics.auto_pdf: true

format:
  revealjs:
    #logo: images/edumer.png
    slide-number: true
    theme: "../libs/edumer.scss"
    auto-stretch: false
    title-slide-attributes:
      visibility: false
    transition: fade
    transition-speed: slow
# data-background-image: images/cover.jpg
# data-background-size: cover
    auto-play-media: true
  pdf:
    format: beamer 
    keep-tex: true  # Debug LaTeX output if needed
---

```{r}
#| label: setup
#| include: false
library(knitr)
knitr::opts_chunk$set(echo = F,
                      warning = F,
                      error = F, 
                      message = F) 
```

```{r}
#| label: packages
#| include: false

if (! require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               sjlabelled, 
               here,
               kableExtra,
               ggdist,
               summarytools)

options(scipen=999)
rm(list = ls())
```

```{r}
#| label: data
#| include: false

load(url("https://github.com/educacion-meritocracia/market-justice-preferences-school/raw/refs/heads/main/input/data/mlm_models.RData"))
load(url("https://github.com/educacion-meritocracia/market-justice-preferences-school/raw/refs/heads/main/input/data/dependientes.RData"))

```

::: columns
::: {.column width="15%"}

![](images/coes.png)

![](images/edumer.png)

:::

::: {.column .column-right width="85%"}
<br>

## **Mi increíble presentación**

------------------------------------------------------------------------

**Autores**

##### Afiliaciones

::: {.red2 .medium}

**FONDECYT No. 1210847 "Meritocracia en la escuela: fundamentos morales del mercado educativo y sus implicancias para la formación ciudadana en Chile"**

:::

Conferencia X

:::
:::

::: {.notes}
Aquí mis notas
:::

# Contexto y motivación {.xlarge}

## Antecedentes

::: {.incremental .highlight-last style="font-size: 100%;"}

- The justification of social inequality based on market-type criteria has been conceptualized as the individuals' adherence to the deservingness of social goods and services (such as health, education, and pensions) based on prices and individuals' ability to pay [@boltanski_new_2005; @lane_market_1986; @streeck_citizens_2012].

- Research on social stratification beliefs [@kluegel_beliefs_1987], highlights that people's explanations and justifications of social inequality are closely tied to their judgments of deservingness.

-  The expansion of marketization has been related to a larger justification of market mechanisms, whereby societies with larger private spending on services show larger market justice preferences [@lindh_public_2015].

:::

## Antecedentes

::::: columns
::: {.column width="50%" .content-box-red}

#### A

- The original definition of merit is a combination of effort and talent [@young_rise_1958]
- 2
- 3

:::

::: {.column width="50%" .content-box-yellow}

#### B

- Adolescents from low-socioeconomic-status families are likelier to have an egalitarian fairness view and consider an equal distribution as fair in a situation with unequal merits [@almas_fairness_2017]
- 2
- 3

:::
:::::

## Hipótesis

::: {.incremental .highlight-last style="font-size: 100%;"}

$H_{1}$

$H_{2}$

$H_{3}$

$H_{n}$

:::

# Datos, variables y métodos {.xlarge}

## Datos

::: {.incremental .highlight-last style="font-size: 100%;"}
-   Datos
:::

## Variables

```{r}
#| label: tbl-desc-dependientes
#| tbl-cap: "Variables dependientes"
#| results: asis
#| echo: false

df<-dfSummary(dependientes,
               plain.ascii = FALSE,
               style = "multiline",
               tmp.img.dir = "/tmp",
               headings = F,  # encabezado
               varnumbers = F, # num variable
               labels.col = T, # etiquetas
               na.col = F,    # missing
               graph.col = F, # plot
               valid.col = T) # n valido

df$Variable <- NULL # delete variable column

print(df)
```

:::

## Métodos

- Métodos

# Resultados {.xlarge}

# Modelos
::: {style="font-size: 100%;"}

```{r}
#| label: fig-lineal-reg
#| fig-cap: "Linear mixed-effects models for meritocracy perception and market justice preferences"
#| fig-cap-location: top
#| fig-align: center
#| out-width: '80%'
#| echo: false
#| warning: false
#| message: false
#| results: asis

theme_set(theme_ggdist())

sjPlot::plot_model(model = reg1_bienestar, show.values = T, show.p = T)

```

:::

# Discusión y conclusiones {data-background-color="#F2A0A7"}

# Discusión y conclusiones {data-background-color="#F2A0A7"}

::: { .incremental .highlight-last style="font-size: 100%;"}

- Discusión

  - Conclusión
:::

# Gracias por su atención!

-   **Github del proyecto:** <https://github.com/educacion-meritocracia>

## Referencias

### 2025-01-29 processing/01-proc-data.qmd
Up primeros ajustes

---
title: "Data preparation"
subtitle: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
author: "Andreas Laffert, Research asistant"
date: today
lang: en
fontsize: 12pt
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    toc-expand: 2
    toc-title: Contents
    number-sections: true
    number-depth: 3
    theme:
      - cosmo
      - edumer_html.scss
    code-link: true
    title-block-banner: true
  pdf:
    number-sections: true
    number-depth: 3
editor_options: 
  chunk_output_type: console
---

# Presentation

This is the data preparation code for the paper "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)". The prepared dataset is `ELSOC_Long_2016_2023_1.00.RData`.

```{r}
#| label: set
#| echo: false
#| message: false
#| warning: false

library(knitr)
knitr::opts_chunk$set(echo = TRUE, include = TRUE, warning = FALSE, message = FALSE)

table_format <- if(is_html_output()) {
  "html"
} else if(is_latex_output()) {
  "latex"
}
table_format2 <- if(is_html_output()) {
  T
} else if(is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")
```

# Libraries

```{r}
#| label: packages
#| include: false

if (! require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse,
               sjmisc, 
               here,
               sjlabelled)

options(scipen=999)
rm(list = ls())
```

# Data

```{r}
#| label: data
#| echo: true
#| output: false

load(file = here("input", "data", "original", "ELSOC_Long_2016_2023_1.00.RData"))

glimpse(elsoc_long_2016_2023)

```

# Processing

## Select

## Filter

## Recode and transform

## Missing values

# Save and export

### 2025-03-04 processing/01-proc-data.qmd
Up data preparation and analysis V1

@@
 #| label: packages
-#| include: false
+#| include: true
 
@@
                here,
-               sjlabelled)
+               sjlabelled,
+               SciViews,
+               naniar)
 
@@
 
-load(file = here("input", "data", "original", "ELSOC_Long_2016_2023_1.00.RData"))
+load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))
 
@@
 
+```{r}
+db_proc <- elsoc_long_2016_2023 %>% 
+  select(idencuesta, ola, 
+         ponderador_long_total,segmento, estrato,
+         just_educ = d02_02, just_pension = d02_01, 
+         just_health = d02_03, merit_effort = c18_09,
+         merit_talent = c18_10, perc_sal_gerente = d03_01,
+         perc_sal_obrero = d03_02, just_sal_gerente = d04_01, 
+         just_sal_obrero = d04_02, age = m0_edad, m01, 
+         sex = m0_sexo, ess = d01_01, ideo = c15
+         ) %>% 
+  as_tibble()
+```
+
 ## Filter
+
+```{r}
+#db_proc %>% 
+#  group_by(ola) %>% 
+#  count(just_educ,just_pension,just_health) %>% 
+#  na.omit() %>% 
+#  print(n = nrow(.))
+
+db_proc <- db_proc %>% dplyr::filter(ola %in% c(1,2,3,4,6,7))
+
+```
 
@@
 
+```{r}
+# general na's
+db_proc[db_proc ==-999] <- NA
+db_proc[db_proc ==-888] <- NA
+db_proc[db_proc ==-777] <- NA
+db_proc[db_proc ==-666] <- NA
+
+# mjp
+
+frq(db_proc$just_health)
+frq(db_proc$just_pension)
+frq(db_proc$just_educ)
+
+labels1 <- c("Strongly desagree" = 1, 
+             "Desagree" = 2, 
+             "Neither agree nor desagree" = 3, 
+             "Agree" = 4, 
+             "Strongly agree" = 5)
+
+db_proc <- db_proc %>% 
+  mutate(
+    across(
+      .cols = c(just_health, just_pension, just_educ),
+      .fns = ~ sjlabelled::set_labels(., labels = labels1)
+      )
+    )
+
+db_proc <- cbind(db_proc, "mjp" = rowMeans(db_proc %>% select(just_health, just_pension, just_educ), na.rm=TRUE))
+
+summary(db_proc$mjp)
+
+db_proc <- db_proc %>% 
+  mutate(
+    across(
+      .cols = c(just_health, just_pension, just_educ),
+      .fns = ~ car::recode(., recodes = c("1='Strongly desagree'; 2='Desagree';
+                                          3='Neither agree nor desagree'; 4='Agree';
+                                          5='Strongly agree'"), 
+                           levels = c("Strongly desagree", "Desagree", "Neither agree nor desagree", "Agree", "Strongly agree"),
+                           as.factor = T)
+    )
+  )
+
+db_proc$just_health <- sjlabelled::set_label(db_proc$just_health, 
+                        label = "Health distributive justice")
+
+db_proc$just_pension <- sjlabelled::set_label(db_proc$just_pension, 
+                        label = "Pension distributive justice")
+
+db_proc$just_educ <- sjlabelled::set_label(db_proc$just_educ, 
+                        label = "Education distributive justice")
+
+db_proc$mjp <- sjlabelled::set_label(db_proc$mjp, 
+                        label = "Market justice preferences")
+
+# merit
+
+frq(db_proc$merit_effort)
+frq(db_proc$merit_talent)
+
+db_proc <- db_proc %>% 
+  mutate(
+    across(
+      .cols = c(merit_effort, merit_talent),
+      .fns = ~ sjlabelled::set_labels(., labels = labels1)
+      )
+    )
+
+db_proc <- db_proc %>% 
+  mutate(
+    across(
+      .cols = c(merit_effort, merit_talent),
+      .fns = ~ car::recode(., recodes = c("1='Strongly desagree'; 2='Desagree';
+                                          3='Neither agree nor desagree'; 4='Agree';
+                                          5='Strongly agree'"), 
+                           levels = c("Strongly desagree", "Desagree", "Neither agree nor desagree", "Agree", "Strongly agree"),
+                           as.factor = T)
+    )
+  )
+
+db_proc$merit_effort <- sjlabelled::set_label(db_proc$merit_effort, 
+                        label = "People are rewarded for their efforts")
+
+db_proc$merit_talent <- sjlabelled::set_label(db_proc$merit_talent, 
+                        label = "People are rewarded for their intelligence")
+
+# perc inequality gap
+
+frq(db_proc$perc_sal_gerente)
+frq(db_proc$perc_sal_obrero)
+
+db_proc <- db_proc %>% 
+  mutate(
+    across(
+      .cols = c(perc_sal_gerente, perc_sal_obrero),
+           .fns = ~ sjlabelled::set_na(., na = 0)
+      )
+    )
+
+db_proc$perc_inequality <- SciViews::ln(db_proc$perc_sal_gerente/db_proc$perc_sal_obrero)
+
+summary(db_proc$perc_inequality)
+
+db_proc$perc_inequality <- sjlabelled::set_label(db_proc$perc_inequality, 
+                        label = "Inequality gap perception")
+
+# just inequality gap
+
+frq(db_proc$just_sal_gerente)
+frq(db_proc$just_sal_obrero)
+
+db_proc <- db_proc %>% 
+  mutate(
+    across(
+      .cols = c(just_sal_gerente, just_sal_obrero),
+           .fns = ~ sjlabelled::set_na(., na = 0)
+      )
+    )
+
+db_proc$just_inequality <- SciViews::ln(db_proc$just_sal_gerente/db_proc$just_sal_obrero)
+
+summary(db_proc$just_inequality)
+
+db_proc$perc_inequality <- sjlabelled::set_label(db_proc$perc_inequality, 
+                        label = "Inequality gap justification")
+
+# sex
+db_proc$sex <- car::recode(db_proc$sex, 
+                           recodes = c("1='Male'; 2='Female'"), 
+                           levels = c("Male", "Female"),
+                           as.factor = T)
+
+db_proc$sex <- sjlabelled::set_label(db_proc$sex, 
+                        label = "Gender")
+
+# wave 
+db_proc <- db_proc %>% 
+  rename(wave = ola) %>% 
+  mutate(wave = case_when(wave == 1 ~ 1,
+                          wave == 2 ~ 2,
+                          wave == 3 ~ 3,
+                          wave == 4 ~ 4,
+                          wave == 6 ~ 5,
+                          wave == 7 ~ 6,
+                          TRUE ~ NA_real_))
+
+db_proc$wave <- sjlabelled::set_label(db_proc$wave, 
+                        label = "Wave")
+
+# age
+frq(db_proc$age)
+
+db_proc$age <- 
+  factor(car::recode(db_proc$age, 
+                     "18:29=1;30:49=2;50:64=3;65:150=4"),
+         labels = c('18-29', '30-49', '50-64', '65 or more'))
+db_proc$age <-
+  sjlabelled::set_label(db_proc$age, 
+                        label = c("Age groups"))
+
+# political indentification
+
+frq(db_proc$ideo)
+
+db_proc$ideo<-
+factor(
+  car::recode(
+    db_proc$ideo,
+    "c(11,12,-888,-999)='Does not identify';c(0,1,2,3,4)='Left';
+     c(5)='Center';c(6,7,8,9,10)='Right'"
+  ),
+  levels = c('Left', 'Center', 'Right', 'Does not identify')
+)
+
+db_proc$ideo<- factor(db_proc$ideo,levels = levels(db_proc$ideo))
+
+db_proc$ideo <- 
+sjlabelled::set_label(x = db_proc$ideo, 
+                      label = "Political identification")
+
+frq(db_proc$ideo)
+
+# ess
+sjmisc::frq(db_proc$ess)
+
+db_proc$ess <- as.numeric(db_proc$ess)
+
+db_proc$ess <-  sjlabelled::set_label(x = db_proc$ess,
+                      label = "Subjective Social Status")
+
+# educ
+frq(db_proc$m01)
+
+db_proc$educ <- 
+  car::recode(db_proc$m01,
+              "c(1,2,3,4,5,6,7)=1;c(8,9,10)=2; c(-888,-999)=NA")
+db_proc$educ <-
+  factor(db_proc$educ,
+         labels = c("Less than Universitary","Universitary"))
+
+db_proc$educ <- 
+sjlabelled::set_label(x = db_proc$educ,
+                      label = "Education")
+
+#Recoding of education to years based on casen 2017.
+db_proc$educyear<- as.numeric(
+  car::recode(db_proc$m01, 
+              "1=0;2=4.3;3=7.5;4=9.8;5=12.02;6=13.9;
+               7=14.8;8=14.9;9=16.9;10=19.07;c(-888,-999)=NA", 
+              as.numeric = T))
+
+db_proc$educyear <- 
+sjlabelled::set_label(x = db_proc$educyear,
+                      label = "Education in years")
+
+class(db_proc$educyear)
+frq(db_proc$educyear)
+```
+
 ## Missing values
 
+```{r}
+colSums(is.na(db_proc))
+
+n_miss(db_proc)
+
+prop_miss(db_proc)*100
+
+miss_var_summary(db_proc)
+
+miss_var_table(db_proc)
+
+vis_miss(db_proc) + theme(axis.text.x = element_text(angle=80))
+
+```
+
 # Save and export
+
+```{r}
+db_proc <- db_proc %>% 
+  select(-c(perc_sal_gerente, perc_sal_obrero, just_sal_gerente, just_sal_obrero, m01))
+
+df1 <- db_proc %>% na.omit()
+
+df1 <- sjlabelled::copy_labels(df_new = df1, df_origin = db_proc)
+
+save(df1, file = here("input/data/proc/df1.RData"))
+```

### 2025-03-04 processing/02-analysis.qmd
Up data preparation and analysis V1

---
title: "Data analysis"
subtitle: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
author: "Andreas Laffert, Research asistant"
date: today
lang: en
fontsize: 12pt
format:
  html:
    code-fold: true
    code-summary: "Show the code"
    toc: true
    toc-location: right
    toc-depth: 2
    toc-expand: 2
    toc-title: Contents
    number-sections: true
    number-depth: 3
    theme:
      - cosmo
      - edumer_html.scss
    code-link: true
    title-block-banner: true
  pdf:
    number-sections: true
    number-depth: 3
editor_options: 
  chunk_output_type: console
---

# Presentation

This is the analysis code for the paper "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)". The dataset used is `df1.RData`.

```{r}
#| label: set
#| echo: false
#| message: false
#| warning: false

library(knitr)
knitr::opts_chunk$set(echo = TRUE, include = TRUE, warning = FALSE, message = FALSE)

table_format <- if(is_html_output()) {
  "html"
} else if(is_latex_output()) {
  "latex"
}
table_format2 <- if(is_html_output()) {
  T
} else if(is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")
```

# Libraries

```{r}
#| label: packages
#| include: true

if (! require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               influence.ME, 
               marginaleffects,
               MLMusingR,
               texreg, 
               ggdist,
               misty,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled)

options(scipen=999)
rm(list = ls())
```

# Data

```{r}
#| label: data
#| echo: true

load(file = here("input/data/proc/df1.RData"))

glimpse(df1)

```

# Analysis

## Descriptives

```{r}
#| label: tbl-summary
#| tbl-cap: "Estadísticos descriptivos"
#| tbl-cap-location: top
t1 <- df1 %>% 
  select(just_health, just_pension, just_educ, mjp, perc_inequality, just_inequality, merit_effort, merit_talent, ess, ideo, educ, sex, age)

print(summarytools::dfSummary(t1), method="render")
```

```{r}
#| label: fig-alluvial
#| out-width: '100%'
#| fig-asp: 1
#| fig-cap: "Change in the justification of inequality in health, pensions and education over time (2016-2023)"
#| fig-cap-location: top

datos.health <- df1 %>% 
  group_by(idencuesta, wave) %>% 
  count(just_health) %>% 
  group_by(wave) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(wave == 1 ~ "2016",
                          wave == 2 ~ "2017",
                          wave == 3 ~ "2018",
                          wave == 4 ~ "2019",
                          wave == 5 ~ "2022",
                          wave == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                      "2017",
                                      "2018",
                                      "2019",
                                      "2022",
                                      "2023")))

etiquetas.health <- df1 %>%
  group_by(wave, just_health) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(wave) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(wave == 1 ~ "2016",
                          wave == 2 ~ "2017",
                          wave == 3 ~ "2018",
                          wave == 4 ~ "2019",
                          wave == 5 ~ "2022",
                          wave == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

p1 <- datos.health %>% 
  ggplot(aes(x = wave, fill = just_health, stratum = just_health,
             alluvium = idencuesta, y = porcentaje)) +
  ggalluvial::geom_flow(alpha = .4) + 
  ggalluvial::geom_stratum(linetype = 0) +
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(values =  c("#CA0020","#F4A582","#b3b3b3ff","#92C5DE","#0571B0")) +
  geom_shadowtext(data = etiquetas.health,
                  aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
                  position = position_stack(vjust = .5),
                  show.legend = FALSE,
                  size = 3,
                  color = rep('white'),
                  bg.colour='grey30')+
  labs(y = "%",
       x = NULL,
       fill = NULL,
       title = "a. Health") +
  theme_ggdist() +
  theme(legend.position = "none")

datos.pension <- df1 %>% 
  group_by(idencuesta, wave) %>% 
  count(just_pension) %>% 
  group_by(wave) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(wave == 1 ~ "2016",
                          wave == 2 ~ "2017",
                          wave == 3 ~ "2018",
                          wave == 4 ~ "2019",
                          wave == 5 ~ "2022",
                          wave == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

etiquetas.pension <- df1 %>%
  group_by(wave, just_pension) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(wave) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(wave == 1 ~ "2016",
                          wave == 2 ~ "2017",
                          wave == 3 ~ "2018",
                          wave == 4 ~ "2019",
                          wave == 5 ~ "2022",
                          wave == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

p2 <- datos.pension %>% 
  ggplot(aes(x = wave, fill = just_pension, stratum = just_pension,
             alluvium = idencuesta, y = porcentaje)) +
  ggalluvial::geom_flow(alpha = .4) + 
  ggalluvial::geom_stratum(linetype = 0) +
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(values =  c("#CA0020","#F4A582","#b3b3b3ff","#92C5DE","#0571B0")) +
  geom_shadowtext(data = etiquetas.pension,
                  aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
                  position = position_stack(vjust = .5),
                  show.legend = FALSE,
                  size = 3,
                  color = rep('white'),
                  bg.colour='grey30')+
  labs(y = "%",
       x = NULL,
       fill = NULL,
       title = "b. Pension") +
  theme_ggdist() +
  theme(legend.position = "none")

datos.educ <- df1 %>% 
  group_by(idencuesta, wave) %>% 
  count(just_educ) %>% 
  group_by(wave) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(wave == 1 ~ "2016",
                          wave == 2 ~ "2017",
                          wave == 3 ~ "2018",
                          wave == 4 ~ "2019",
                          wave == 5 ~ "2022",
                          wave == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

etiquetas.educ <- df1 %>%
  group_by(wave, just_educ) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(wave) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(wave == 1 ~ "2016",
                          wave == 2 ~ "2017",
                          wave == 3 ~ "2018",
                          wave == 4 ~ "2019",
                          wave == 5 ~ "2022",
                          wave == 6 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

p3 <- datos.educ %>% 
  ggplot(aes(x = wave, fill = just_educ, stratum = just_educ,
             alluvium = idencuesta, y = porcentaje)) +
  ggalluvial::geom_flow(alpha = .4) + 
  ggalluvial::geom_stratum(linetype = 0) +
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(values =  c("#CA0020","#F4A582","#b3b3b3ff","#92C5DE","#0571B0")) +
  geom_shadowtext(data = etiquetas.educ,
                  aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
                  position = position_stack(vjust = .5),
                  show.legend = FALSE,
                  size = 3,
                  color = rep('white'),
                  bg.colour='grey30')+
  labs(y = "%",
       x = NULL,
       fill = NULL,
       title = "c. Education") +
  theme_ggdist() +
  theme(legend.position = "bottom")

(p1 / p2 / p3)

```

## Correlations

```{r}
df1 %>% 
  select(mjp, perc_inequality, just_inequality, merit_effort, merit_talent) %>% 
  mutate_all(.funs = ~ as.numeric(.)) %>% 
  sjPlot::tab_corr(., triangle = "lower")
```

## Longitudinal multilevel models

```{r}
m0 <- lmer(mjp ~ 1 + (1 | idencuesta), 
                data = df1)

performance::icc(m0, by_group = T)
```

```{r}
df1$merit_effort <- as_numeric(df1$merit_effort)
df1$merit_talent <- as_numeric(df1$merit_talent)
df1 <- df1 %>% 
  mutate(wave_f = case_when(wave == 1 ~ "2016",
                          wave == 2 ~ "2017",
                          wave == 3 ~ "2018",
                          wave == 4 ~ "2019",
                          wave == 5 ~ "2022",
                          wave == 6 ~ "2023"),
         wave_f = factor(wave_f, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

df1 <- df1 %>% 
  group_by(idencuesta) %>% 
  mutate(perc_inequality_mean = mean(perc_inequality, na.rm = T),
         perc_inequality_cwc = perc_inequality - perc_inequality_mean,
         just_inequality_mean = mean(just_inequality, na.rm = T),
         just_inequality_cwc = just_inequality - just_inequality_mean,
         merit_effort_mean = mean(merit_effort, na.rm = T),
         merit_effort_cwc = merit_effort - merit_effort_mean,
         merit_talent_mean = mean(merit_talent, na.rm = T),
         merit_talent_cwc = merit_talent - merit_talent_mean,
         ) %>% 
  ungroup()

m1 <- lmer(mjp ~ 1 + perc_inequality + (1 | idencuesta),
                data = df1)

m2 <- lmer(mjp ~ 1 + just_inequality + (1 | idencuesta),
                data = df1)

m3 <- lmer(mjp ~ 1 + perc_inequality + just_inequality + (1 | idencuesta),
                data = df1)

m4 <- lmer(mjp ~ 1 + perc_inequality + just_inequality + 
             merit_effort +  (1 | idencuesta),
                data = df1)

m5 <- lmer(mjp ~ 1 + perc_inequality + just_inequality + 
             merit_effort + merit_talent + (1 | idencuesta),
                data = df1)

m6 <- lmer(mjp ~ 1 + perc_inequality + just_inequality + 
             merit_effort + merit_talent + perc_inequality_mean +
             just_inequality_mean + merit_effort_mean + 
             merit_talent_mean + (1 | idencuesta),
                data = df1)

m7 <- lmer(mjp ~ 1 + perc_inequality + just_inequality + 
             merit_effort + merit_talent + perc_inequality_mean +
             just_inequality_mean + merit_effort_mean + 
             merit_talent_mean + age + sex + ess + ideo + educ + wave_f + (1 | idencuesta),
                data = df1)
```

```{r}
#| label: tbl-modelos
#| results: asis
#| tbl-cap: "Longitudinal multilevel models for market justice preferences"
#| tbl-cap-location: top

ccoef <- list(
  "(Intercept)" = "Intercept",
  perc_inequality = "Perception inequality (WE)",
  just_inequality = "Justification inequality (WE)",
  merit_effort = "Merit: Effort (WE)",
  merit_talent = "Merit: Talent (WE)",
  perc_inequality_mean = "Perception inequality (BE)",
  just_inequality_mean = "Justification inequality (BE)",
  merit_effort_mean = "Merit: Effort (BE)",
  merit_talent_mean = "Merit: Talent (BE)",
  "age30-49" = "30-49",
  "age50-64" = "50-64",
  "age65 or more" = "65 or more",
  sexFemale = "Female (Ref.= Male)",
  ess = "Subjective social status",
  ideoCenter = "Center",
  ideoRight = "Right",
  "ideoDoes not identify" = "Does not identify",
  educUniversitary = "Universitary (Ref.= Less than universitary)",
  wave_f2017 = "2017",
  wave_f2018 = "2018",
  wave_f2019 = "2019",
  wave_f2022 = "2022",
  wave_f2023 = "2023"
  )

texreg::htmlreg(list(m1, m2, m3, m4, m5, m6, m7),
               custom.model.names = c(paste0("Model ", seq(1:7))),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               groups = list("Age (Ref. = 18-29)" = 10:12,"Political identification (Ref.= Left)" = 15:17, "Wave (Ref.= 2016)" = 19:23),
               custom.note = "\\item Note: Cells contain regression coefficients with standard errors in parentheses. %stars. \\\\ \\item Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 16219).",
               threeparttable = T,
               leading.zero = T,
               float.pos = "h!",
               use.packages = F,
               booktabs = T,
               scalebox = 1)
```

### 2025-03-04 processing/01-proc-data.qmd
Up data prep and analysis V1

@@
 subtitle: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
-author: "Andreas Laffert, Research asistant"
+author: "Researcher"
 date: today

### 2025-03-04 processing/02-analysis.qmd
Up data prep and analysis V1

@@
 %stars.
-\\\\ \\item Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 16219).",
+\\\\ \\item Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 4364).",
                threeparttable = T,

### 2025-03-05 processing/01-proc-data.qmd
Update analysis with income and gaps

@@
 db_proc <- db_proc %>% 
-  mutate(
-    across(
-      .cols = c(perc_sal_gerente, perc_sal_obrero),
-           .fns = ~ sjlabelled::set_na(., na = 0)
-      )
-    )
+  mutate(perc_sal_obrero=replace(perc_sal_obrero, perc_sal_obrero <= 40000  | perc_sal_obrero>= 1000000, NA)) %>%
+  mutate(perc_sal_gerente=replace(perc_sal_gerente, perc_sal_gerente <= 250000 | perc_sal_gerente>= 100000001, NA))
 
@@
 db_proc <- db_proc %>% 
-  mutate(
-    across(
-      .cols = c(just_sal_gerente, just_sal_obrero),
-           .fns = ~ sjlabelled::set_na(., na = 0)
-      )
-    )
+  mutate(just_sal_obrero=replace(just_sal_obrero, just_sal_obrero <= 40000  | just_sal_obrero>=999999999999999, NA)) %>%
+  mutate(just_sal_gerente=replace(just_sal_gerente, just_sal_gerente <= 100000 | just_sal_gerente>=999999999999999, NA))
 
@@
 
-db_proc$perc_inequality <- sjlabelled::set_label(db_proc$perc_inequality, 
+db_proc$just_inequality <- sjlabelled::set_label(db_proc$just_inequality, 
                         label = "Inequality gap justification")
@@
 
+```{r}
+# Household income----------------------------------------------------------------
+
+# Ltop is the lower limit of the top category, Ltop-1 is the lower limit of the category before the
+# top one, ftop is the frequency in the top category, and ftop-1 is the frequency in the
+# category before the top one.
+
+sjmisc::frq( elsoc_long_2016_2023$m30)
+Ltop_1<- 1850001
+Ltop  <- 2700000
+
+ftop_1 <- 42
+ftop   <- 34
+
+V  = (log(ftop_1 + ftop) - log(ftop)) / (log(Ltop) -log(Ltop_1))
+M_top = 0.5* Ltop *(1+(V/(V-1)));M_top # = $3.897.232
+
+#Impute midpoint of income ranges
+elsoc_long_2016_2023$m30_rec <-
+  as.numeric(car::recode( elsoc_long_2016_2023$m30,
+                         "1=110000;2=251000;3=305000;4=355000;5=400000;
+            6=445000;7=490000;8=535000;9=585000;10=640000;11=700000;12=765000;
+            13=845000;14=935000;15=1040000;16=1180000;17=1375000;18=1670000;
+            19=2275000;20=3897232;NA=NA;c(-888,-999)=NA"))
+
+summary( elsoc_long_2016_2023$m30_rec)
+
+#Impute midpoint of income ranges (2021)
+
+frq(elsoc_long_2016_2023$m30b)
+Ltop_1<- 450000
+Ltop  <- 700000
+ftop_1 <- 18
+ftop   <- 27
+
+V  = (log(ftop_1 + ftop) - log(ftop)) / (log(Ltop) -log(Ltop_1))
+M_top = 0.5* Ltop *(1+(V/(V-1))) # = $2.941.412
+
+ elsoc_long_2016_2023$m30b_rec <-
+  as.numeric(car::recode( elsoc_long_2016_2023$m30b,
+                         "1=125000;2=300000;3=400000;4=575000;5=2941412;NA=NA;c(-888,-999)=NA"))
+summary( elsoc_long_2016_2023$m30b_rec)
+
+# Combine m30_rec with m30b_rec
+ elsoc_long_2016_2023$m30_rec <- 
+  ifelse(is.na( elsoc_long_2016_2023$m30_rec),
+         yes =  elsoc_long_2016_2023$m30b_rec,
+         no =  elsoc_long_2016_2023$m30_rec)
+summary( elsoc_long_2016_2023$m30_rec)
+
+#Recode DK/DA of Income to NA
+elsoc_long_2016_2023$m29_rec <-
+  as.numeric(car::recode( elsoc_long_2016_2023$m29,"c(-888,-999)=NA"))
+
+#replace NA of income with new imputed variable
+elsoc_long_2016_2023$m29_imp <- 
+  ifelse(test = !is.na( elsoc_long_2016_2023$m29_rec),
+         yes =   elsoc_long_2016_2023$m29_rec,
+         no =   elsoc_long_2016_2023$m30_rec)
+summary( elsoc_long_2016_2023$m29_imp)
+
+#  elsoc_long_2016_2023$m29_imp <- 
+#   ifelse(test = is.na( elsoc_long_2016_2023$m29_imp),
+#          yes =   elsoc_long_2016_2023$m30b_rec,
+#          no =   elsoc_long_2016_2023$m29_imp)
+
+summary( elsoc_long_2016_2023$m29_imp)
+
+# N Household:
+elsoc_long_2016_2023 <-
+   elsoc_long_2016_2023 %>%
+  mutate(n_hogar =
+           dplyr::case_when(ola == 1 ~ nhogar1,
+                            ola == 2 ~ m46_nhogar,
+                            ola == 3 ~ m54,
+                            ola == 4 ~ m54,
+                            ola == 5 ~ m54,
+                            ola == 6 ~ m54,
+                            ola == 7 ~ m54))
+
+summary( elsoc_long_2016_2023$n_hogar)
+#imputar tamanio del hogar de ola 5 a ola 6
+elsoc_long_2016_2023 <- 
+   elsoc_long_2016_2023 %>%  
+  group_by(idencuesta) %>%
+  mutate(n_hogar = if_else(ola == 6 & is.na(n_hogar), last(n_hogar[ola == 5]), n_hogar)) %>% 
+  ungroup()
+
+sjmisc::frq( elsoc_long_2016_2023$n_hogar)
+table( elsoc_long_2016_2023$n_hogar, elsoc_long_2016_2023$ola)
+
+#Recode DK/DA to NA
+ elsoc_long_2016_2023$n_hogar_r<-
+  car::recode( elsoc_long_2016_2023$n_hogar,"c(-888,-999)=NA")
+
+table( elsoc_long_2016_2023$m13, elsoc_long_2016_2023$n_hogar_r)
+
+#ingresos individuales para hogares de 1 persona
+ elsoc_long_2016_2023 %>% sjmisc::find_var('Ingreso')
+ elsoc_long_2016_2023$inc_ind <- ifelse( elsoc_long_2016_2023$n_hogar_r==1, elsoc_long_2016_2023$m13,NA)
+summary( elsoc_long_2016_2023$inc_ind)
+# 
+# c(40000,
+# (40001 + 85000) / 2,
+# (85001 + 125000) / 2,
+# (125001 + 170000) / 2,
+# (170001 + 210000) / 2,
+# (210001 + 230000) / 2,
+# (230001 + 280000) / 2,
+# (280001 + 320000) / 2,
+# (320001 + 360000) / 2,
+# (360001 + 400000) / 2,
+# (400001 + 465000) / 2,
+# (465001 + 540000) / 2,
+# (540001 + 665000) / 2,
+# (665001 + 850000) / 2,
+# (850001 + 1300000) / 2,
+# 1300001)
+
+frq( elsoc_long_2016_2023$m14)
+Ltop_1<- 850001
+Ltop  <- 1300001 
+ftop_1 <- 65
+ftop   <- 40
+
+V  = (log(ftop_1 + ftop) - log(ftop)) / (log(Ltop) -log(Ltop_1))
+M_top = 0.5* Ltop *(1+(V/(V-1))) # = $1.811.247
+
+ elsoc_long_2016_2023$m14_rec<- 
+  car::recode( elsoc_long_2016_2023$m14,
+              "1=40000.0;2=62500.5;3=105000.5;4=147500.5;5=190000.5;6=220000.5;
+              7=255000.5;8=300000.5;9=340000.5;10=380000.5;11=432500.5;12=502500.5;
+              13=602500.5;14=757500.5;15=1075000.5;16=1811247"
+  )
+
+ elsoc_long_2016_2023$inc_ind <- ifelse( elsoc_long_2016_2023$n_hogar_r==1 & is.na( elsoc_long_2016_2023$inc_ind), elsoc_long_2016_2023$m14_rec, elsoc_long_2016_2023$inc_ind)
+summary( elsoc_long_2016_2023$inc_ind)
+
+# Ingreso del hogar pegarle ingreso del hogar unipersonal
+summary( elsoc_long_2016_2023$m29_imp)
+ elsoc_long_2016_2023$m29_imp <- 
+  ifelse(is.na( elsoc_long_2016_2023$m29_imp) &  elsoc_long_2016_2023$n_hogar_r==1,
+          elsoc_long_2016_2023$inc_ind,
+          elsoc_long_2016_2023$m29_imp) #recupero 14 casos
+summary( elsoc_long_2016_2023$m29_imp)
+
+# Household Equivalent income 
+# 1.0 to the first adult;
+# 0.5 to the second and each subsequent person aged 14 and over;
+# 0.3 to each child aged under 14.
+
+ elsoc_long_2016_2023$n_hogar_r_eq <- ( elsoc_long_2016_2023$n_hogar_r + 1) /2
+table( elsoc_long_2016_2023$n_hogar_r_eq)
+
+frq( elsoc_long_2016_2023$n_hogar_r_eq)
+
+# Per capita household income:
+ elsoc_long_2016_2023$ing_pc <- 
+  (elsoc_long_2016_2023$m29_imp/ elsoc_long_2016_2023$n_hogar_r)
+
+# Per capita household income equivalized (OECD):
+ elsoc_long_2016_2023$ing_pc_eq <- 
+  ( elsoc_long_2016_2023$m29_imp/ elsoc_long_2016_2023$n_hogar_r_eq)
+
+descr( elsoc_long_2016_2023$ing_pc_eq)
+descr( elsoc_long_2016_2023$ing_pc)
+ elsoc_long_2016_2023$ing_pc <-
+  sjlabelled::set_label(x =  elsoc_long_2016_2023$ing_pc,
+                        label = "Household income per capita")
+
+sjmisc::descr( elsoc_long_2016_2023$ing_pc)
+
+# Compute income quintiles
+ elsoc_long_2016_2023 <-  elsoc_long_2016_2023 %>% 
+  group_by(ola) %>% 
+  mutate(quintil = ntile(-desc(ing_pc), 5)) %>% 
+  ungroup()
+
+ elsoc_long_2016_2023$quintil <- 
+  factor( elsoc_long_2016_2023$quintil,
+         levels = c(1, 2, 3, 4, 5),
+         labels = c('Q1', 'Q2', 'Q3', 'Q4', 'Q5')) # Quintiles as factors
+
+elsoc_long_2016_2023$quintil <- 
+  sjlabelled::set_label(x =  elsoc_long_2016_2023$quintil,
+                        label = "Household income quintile per capita")
+
+sjmisc::frq( elsoc_long_2016_2023$quintil)
+
+#include new quintile category with missing cases
+ elsoc_long_2016_2023$quintil1<-
+  car::recode( elsoc_long_2016_2023$quintil, 
+              "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")
+
+# elsoc_long_2016_2023$quintil1 <- factor( elsoc_long_2016_2023$quintil1, c("Q1","Q2","Q3","Q4","Q5","QNA"))
+
+ elsoc_long_2016_2023$quintil1 <- 
+  sjlabelled::set_label(x =  elsoc_long_2016_2023$quintil1,
+                        label = "Household income quintile per capita (NA)") 
+sjmisc::frq( elsoc_long_2016_2023$quintil1)
+
+```
+
+```{r}
+elsoc_income <- elsoc_long_2016_2023 %>% 
+  dplyr::select(idencuesta, ola, ing_pc, quintil, quintil1) %>% 
+  dplyr::filter(ola %in% c(1,2,3,4,6,7))
+
+db_proc <- left_join(db_proc, elsoc_income, by = c("idencuesta", "wave" = "ola"))
+
+```
+
 ## Missing values

### 2025-03-05 processing/02-analysis.qmd
Update analysis with income and gaps

@@
              just_inequality_mean + merit_effort_mean + 
-             merit_talent_mean + age + sex + ess + ideo + educ + wave_f + (1 | idencuesta),
+             merit_talent_mean + ideo + quintil + educ + ess + age + sex + wave_f + (1 | idencuesta),
                 data = df1)
@@
   merit_talent_mean = "Merit: Talent (BE)",
+  ideoCenter = "Center",
+  ideoRight = "Right",
+  "ideoDoes not identify" = "Does not identify",
+  quintilQ2 = "Quintile 2",
+  quintilQ3 = "Quintile 3",
+  quintilQ4 = "Quintile 4",
+  quintilQ5 = "Quintile 5",
+  educUniversitary = "Universitary (Ref.= Less than universitary)",
+  ess = "Subjective social status",
   "age30-49" = "30-49",
@@
   sexFemale = "Female (Ref.= Male)",
-  ess = "Subjective social status",
-  ideoCenter = "Center",
-  ideoRight = "Right",
-  "ideoDoes not identify" = "Does not identify",
-  educUniversitary = "Universitary (Ref.= Less than universitary)",
   wave_f2017 = "2017",
@@
                digits = 3,
-               groups = list("Age (Ref. = 18-29)" = 10:12,"Political identification (Ref.= Left)" = 15:17, "Wave (Ref.= 2016)" = 19:23),
+               groups = list("Political identification (Ref.= Left)" = 10:12, "Income (Ref.= Quintile 1)" = 13:16, "Age (Ref. = 18-29)" = 19:21, "Wave (Ref.= 2016)" = 23:26),
                custom.note = "\\item Note: Cells contain regression coefficients with standard errors in parentheses.
 %stars.
-\\\\ \\item Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 4364).",
+\\\\ \\item Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 4257).",
                threeparttable = T,

### 2025-04-04 processing/01-proc-data.qmd
Update analysis

--- 2025-03-04 processing/01-proc-data.qmd
@@
 pacman::p_load(tidyverse,
+               car,
                sjmisc, 
@@
 
-## Select
+```{r}
+
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-999] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-888] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-777] <- NA
+elsoc_long_2016_2023[elsoc_long_2016_2023 ==-666] <- NA
+
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(just_educ = d02_02, 
+         just_pension = d02_01, 
+         just_health = d02_03, 
+         merit_effort = c18_09,
+         merit_talent = c18_10, 
+         perc_sal_gerente = d03_01,
+         perc_sal_obrero = d03_02, 
+         just_sal_gerente = d04_01, 
+         just_sal_obrero = d04_02, 
+         age = m0_edad, m01, 
+         sex = m0_sexo, 
+         ess = d01_01, 
+         ideo = c15) %>% 
+  as_tibble() %>% 
+  sjlabelled::drop_labels(., drop.na = FALSE)
+```
 
 ```{r}
-db_proc <- elsoc_long_2016_2023 %>% 
-  select(idencuesta, ola, 
-         ponderador_long_total,segmento, estrato,
-         just_educ = d02_02, just_pension = d02_01, 
-         just_health = d02_03, merit_effort = c18_09,
-         merit_talent = c18_10, perc_sal_gerente = d03_01,
-         perc_sal_obrero = d03_02, just_sal_gerente = d04_01, 
-         just_sal_obrero = d04_02, age = m0_edad, m01, 
-         sex = m0_sexo, ess = d01_01, ideo = c15
-         ) %>% 
-  as_tibble()
-```
+# Market Justice Preferences
 
-## Filter
+frq(elsoc_long_2016_2023$just_health)
+frq(elsoc_long_2016_2023$just_pension)
+frq(elsoc_long_2016_2023$just_educ)
 
-```{r}
-#db_proc %>% 
-#  group_by(ola) %>% 
-#  count(just_educ,just_pension,just_health) %>% 
-#  na.omit() %>% 
-#  print(n = nrow(.))
+elsoc_long_2016_2023 <- cbind(elsoc_long_2016_2023, "mjp" = rowMeans(elsoc_long_2016_2023 %>% select(just_health, just_pension, just_educ), na.rm=TRUE))
 
-db_proc <- db_proc %>% dplyr::filter(ola %in% c(1,2,3,4,6,7))
+summary(elsoc_long_2016_2023$mjp)
 
-```
-
-## Recode and transform
-
-```{r}
-# general na's
-db_proc[db_proc ==-999] <- NA
-db_proc[db_proc ==-888] <- NA
-db_proc[db_proc ==-777] <- NA
-db_proc[db_proc ==-666] <- NA
-
-# mjp
-
-frq(db_proc$just_health)
-frq(db_proc$just_pension)
-frq(db_proc$just_educ)
-
-labels1 <- c("Strongly desagree" = 1, 
-             "Desagree" = 2, 
-             "Neither agree nor desagree" = 3, 
-             "Agree" = 4, 
-             "Strongly agree" = 5)
-
-db_proc <- db_proc %>% 
-  mutate(
-    across(
-      .cols = c(just_health, just_pension, just_educ),
-      .fns = ~ sjlabelled::set_labels(., labels = labels1)
-      )
-    )
-
-db_proc <- cbind(db_proc, "mjp" = rowMeans(db_proc %>% select(just_health, just_pension, just_educ), na.rm=TRUE))
-
-summary(db_proc$mjp)
-
-db_proc <- db_proc %>% 
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
   mutate(
@@
 
-db_proc$just_health <- sjlabelled::set_label(db_proc$just_health, 
+elsoc_long_2016_2023$just_health <- sjlabelled::set_label(elsoc_long_2016_2023$just_health, 
                         label = "Health distributive justice")
 
-db_proc$just_pension <- sjlabelled::set_label(db_proc$just_pension, 
+elsoc_long_2016_2023$just_pension <- sjlabelled::set_label(elsoc_long_2016_2023$just_pension, 
                         label = "Pension distributive justice")
 
-db_proc$just_educ <- sjlabelled::set_label(db_proc$just_educ, 
+elsoc_long_2016_2023$just_educ <- sjlabelled::set_label(elsoc_long_2016_2023$just_educ, 
                         label = "Education distributive justice")
 
-db_proc$mjp <- sjlabelled::set_label(db_proc$mjp, 
+elsoc_long_2016_2023$mjp <- sjlabelled::set_label(elsoc_long_2016_2023$mjp, 
                         label = "Market justice preferences")
+```
 
-# merit
+```{r}
+# Meritocracy
 
-frq(db_proc$merit_effort)
-frq(db_proc$merit_talent)
+frq(elsoc_long_2016_2023$merit_effort)
+frq(elsoc_long_2016_2023$merit_talent)
 
-db_proc <- db_proc %>% 
-  mutate(
-    across(
-      .cols = c(merit_effort, merit_talent),
-      .fns = ~ sjlabelled::set_labels(., labels = labels1)
-      )
-    )
-
-db_proc <- db_proc %>% 
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
   mutate(
@@
 
-db_proc$merit_effort <- sjlabelled::set_label(db_proc$merit_effort, 
+elsoc_long_2016_2023$merit_effort <- sjlabelled::set_label(elsoc_long_2016_2023$merit_effort, 
                         label = "People are rewarded for their efforts")
 
-db_proc$merit_talent <- sjlabelled::set_label(db_proc$merit_talent, 
+elsoc_long_2016_2023$merit_talent <- sjlabelled::set_label(elsoc_long_2016_2023$merit_talent, 
                         label = "People are rewarded for their intelligence")
+
+```
+
+```{r}
+# Inequality
 
@@
 
-frq(db_proc$perc_sal_gerente)
-frq(db_proc$perc_sal_obrero)
+frq(elsoc_long_2016_2023$perc_sal_gerente)
+frq(elsoc_long_2016_2023$perc_sal_obrero)
 
-db_proc <- db_proc %>% 
-  mutate(
-    across(
-      .cols = c(perc_sal_gerente, perc_sal_obrero),
-           .fns = ~ sjlabelled::set_na(., na = 0)
-      )
-    )
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(perc_sal_obrero=replace(perc_sal_obrero, perc_sal_obrero <= 40000  | perc_sal_obrero>= 1000000, NA)) %>%
+  mutate(perc_sal_gerente=replace(perc_sal_gerente, perc_sal_gerente <= 250000 | perc_sal_gerente>= 100000001, NA))
 
-db_proc$perc_inequality <- SciViews::ln(db_proc$perc_sal_gerente/db_proc$perc_sal_obrero)
+elsoc_long_2016_2023$perc_inequality <- SciViews::ln(elsoc_long_2016_2023$perc_sal_gerente/elsoc_long_2016_2023$perc_sal_obrero)
 
-summary(db_proc$perc_inequality)
+summary(elsoc_long_2016_2023$perc_inequality)
 
-db_proc$perc_inequality <- sjlabelled::set_label(db_proc$perc_inequality, 
+elsoc_long_2016_2023$perc_inequality <- sjlabelled::set_label(elsoc_long_2016_2023$perc_inequality, 
                         label = "Inequality gap perception")
@@
 
-frq(db_proc$just_sal_gerente)
-frq(db_proc$just_sal_obrero)
+frq(elsoc_long_2016_2023$just_sal_gerente)
+frq(elsoc_long_2016_2023$just_sal_obrero)
 
-db_proc <- db_proc %>% 
-  mutate(
-    across(
-      .cols = c(just_sal_gerente, just_sal_obrero),
-           .fns = ~ sjlabelled::set_na(., na = 0)
-      )
-    )
+elsoc_long_2016_2023 <- elsoc_long_2016_2023 %>% 
+  mutate(just_sal_obrero=replace(just_sal_obrero, just_sal_obrero <= 40000  | just_sal_obrero>=999999999999999, NA)) %>%
+  mutate(just_sal_gerente=replace(just_sal_gerente, just_sal_gerente <= 100000 | just_sal_gerente>=999999999999999, NA))
 
-db_proc$just_inequality <- SciViews::ln(db_proc$just_sal_gerente/db_proc$just_sal_obrero)
+elsoc_long_2016_2023$just_inequality <- SciViews::ln(elsoc_long_2016_2023$just_sal_gerente/elsoc_long_2016_2023$just_sal_obrero)
 
-summary(db_proc$just_inequality)
+summary(elsoc_long_2016_2023$just_inequality)
 
-db_proc$perc_inequality <- sjlabelled::set_label(db_proc$perc_inequality, 
+elsoc_long_2016_2023$just_inequality <- sjlabelled::set_label(elsoc_long_2016_2023$just_inequality, 
                         label = "Inequality gap justification")
 
+```
+
+```{r}
+# Controls
+
 # sex
-db_proc$sex <- car::recode(db_proc$sex, 
+elsoc_long_2016_2023$sex <- car::recode(elsoc_long_2016_2023$sex, 
                            recodes = c("1='Male'; 2='Female'"), 
@@
 
-db_proc$sex <- sjlabelled::set_label(db_proc$sex, 
+elsoc_long_2016_2023$sex <- sjlabelled::set_label(elsoc_long_2016_2023$sex, 
                         label = "Gender")
 
-# wave 
-db_proc <- db_proc %>% 
-  rename(wave = ola) %>% 
-  mutate(wave = case_when(wave == 1 ~ 1,
-                          wave == 2 ~ 2,
-                          wave == 3 ~ 3,
-                          wave == 4 ~ 4,
-                          wave == 6 ~ 5,
-                          wave == 7 ~ 6,
-                          TRUE ~ NA_real_))
+# age
+frq(elsoc_long_2016_2023$age)
 
-db_proc$wave <- sjlabelled::set_label(db_proc$wave, 
-                        label = "Wave")
-
-# age
-frq(db_proc$age)
-
-db_proc$age <- 
-  factor(car::recode(db_proc$age, 
+elsoc_long_2016_2023$age <- 
+  factor(car::recode(elsoc_long_2016_2023$age, 
                      "18:29=1;30:49=2;50:64=3;65:150=4"),
          labels = c('18-29', '30-49', '50-64', '65 or more'))
-db_proc$age <-
-  sjlabelled::set_label(db_proc$age, 
+elsoc_long_2016_2023$age <-
+  sjlabelled::set_label(elsoc_long_2016_2023$age, 
                         label = c("Age groups"))
@@
 
-frq(db_proc$ideo)
+frq(elsoc_long_2016_2023$ideo)
 
-db_proc$ideo<-
+elsoc_long_2016_2023$ideo<-
 factor(
   car::recode(
-    db_proc$ideo,
+    elsoc_long_2016_2023$ideo,
     "c(11,12,-888,-999)='Does not identify';c(0,1,2,3,4)='Left';
@@
 
-db_proc$ideo<- factor(db_proc$ideo,levels = levels(db_proc$ideo))
+elsoc_long_2016_2023$ideo<- factor(elsoc_long_2016_2023$ideo,levels = levels(elsoc_long_2016_2023$ideo))
 
-db_proc$ideo <- 
-sjlabelled::set_label(x = db_proc$ideo, 
+elsoc_long_2016_2023$ideo <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$ideo, 
                       label = "Political identification")
 
-frq(db_proc$ideo)
+frq(elsoc_long_2016_2023$ideo)
 
 # ess
-sjmisc::frq(db_proc$ess)
+sjmisc::frq(elsoc_long_2016_2023$ess)
 
-db_proc$ess <- as.numeric(db_proc$ess)
+elsoc_long_2016_2023$ess <- as.numeric(elsoc_long_2016_2023$ess)
 
-db_proc$ess <-  sjlabelled::set_label(x = db_proc$ess,
+elsoc_long_2016_2023$ess <-  sjlabelled::set_label(x = elsoc_long_2016_2023$ess,
                       label = "Subjective Social Status")
 
-# educ
-frq(db_proc$m01)
+```
 
-db_proc$educ <- 
-  car::recode(db_proc$m01,
+```{r}
+# Socioeconomic_________________________________________________________________
+
+# Education_______________________________________
+elsoc_long_2016_2023$educ <- 
+  car::recode(elsoc_long_2016_2023$m01,
               "c(1,2,3,4,5,6,7)=1;c(8,9,10)=2; c(-888,-999)=NA")
-db_proc$educ <-
-  factor(db_proc$educ,
+elsoc_long_2016_2023$educ <-
+  factor(elsoc_long_2016_2023$educ,
          labels = c("Less than Universitary","Universitary"))
 
-db_proc$educ <- 
-sjlabelled::set_label(x = db_proc$educ,
+#reverse education, reference level is the highest level
+#elsoc_long_2016_2023$educ <- forcats::fct_rev(elsoc_long_2016_2023$educ)
+
+elsoc_long_2016_2023$educ <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$educ,
                       label = "Education")
+sjmisc::frq(elsoc_long_2016_2023$educ)
 
 #Recoding of education to years based on casen 2017.
-db_proc$educyear<- as.numeric(
-  car::recode(db_proc$m01, 
+elsoc_long_2016_2023$educyear<- as.numeric(
+  car::recode(elsoc_long_2016_2023$m01, 
               "1=0;2=4.3;3=7.5;4=9.8;5=12.02;6=13.9;
@@
 
-db_proc$educyear <- 
-sjlabelled::set_label(x = db_proc$educyear,
+elsoc_long_2016_2023$educyear <- 
+sjlabelled::set_label(x = elsoc_long_2016_2023$educyear,
                       label = "Education in years")
 
-class(db_proc$educyear)
-frq(db_proc$educyear)
+class(elsoc_long_2016_2023$educyear)
+sjmisc::frq(elsoc_long_2016_2023$educyear)
 ```
 
-## Missing values
+```{r}
+# N Household:
+# Select variables______________________________________________________________
+
+df_study1 <- 
+elsoc_long_2016_2023 %>% 
+  filter(muestra == 1) %>% #keep original sample (vs refresh)
+  select(idencuesta,ola,region,region_cod,comuna,comunacod=comuna_cod,
+         muestra,
+         ponderador_long_total, segmento, estrato,
+         just_educ,
+         just_pension,
+         just_health,
+         mjp,
+         merit_effort,
+         merit_talent,
+         perc_sal_gerente,
+         perc_sal_obrero,
+         just_sal_gerente,
+         just_sal_obrero,
+         perc_inequality,
+         just_inequality,
+         educ,
+         educyear,
+         sex,
+         age,
+         ess,
+         ideo,
+         nhogar1,
+         m46_nhogar,
+         m54, m30, m30b, m29)
+
+# Household income_________________________________________
+
+#Impute midpoint of income ranges
+df_study1$m30_rec <-
+as.numeric(car::recode(df_study1$m30,
+           "1=110000;2=251000;3=305000;4=355000;5=400000;
+            6=445000;7=490000;8=535000;9=585000;10=640000;11=700000;12=765000;
+            13=845000;14=935000;15=1040000;16=1180000;17=1375000;18=1670000;
+            19=2275000;20=2700000;NA=NA;c(-888,-999)=NA"))
+
+#Impute midpoint of income ranges (2021)
+df_study1$m30b_rec <-
+as.numeric(car::recode(df_study1$m30b,
+           "1=125000;2=300000;3=400000;4=575000;5=70000;NA=NA;c(-888,-999)=NA"))
+
+sjmisc::frq(df_study1$m30b_rec)
+
+#Recode DK/DA of Income to NA
+df_study1$m29_rec <-
+  as.numeric(car::recode(df_study1$m29,"c(-888,-999)=NA"))
+
+#replace NA of income with new imputed variable
+df_study1$m29_imp <- 
+  ifelse(test = !is.na(df_study1$m29_rec),
+         yes =  df_study1$m29_rec,
+         no =  df_study1$m30_rec)
+summary(df_study1$m29_imp)
+
+df_study1$m29_imp <- 
+  ifelse(test = is.na(df_study1$m29_imp),
+         yes =  df_study1$m30b_rec,
+         no =  df_study1$m29_imp)
+summary(df_study1$m29_imp)
+
+# deflate at each year's prices
+df_study1$deflactor <-
+  with(df_study1, case_when(
+    ola == 2016 ~ 113.88 / 123.82,
+    ola == 2017 ~ 116.46 / 123.82,
+    ola == 2018 ~ 119.45 / 123.82,
+    ola == 2019 ~ 123.82 / 123.82
+  ))
+
+# Reshape long to wide
+df_study1_wide <- df_study1 %>% 
+  tidyr::pivot_wider(id_cols = c("idencuesta","muestra"),
+                     names_from = "ola",
+                     # names_prefix = c("educ","educyear","ing_pc","quintil","quintil1"),
+                     values_from = names(select(df_study1,comuna,comunacod,ponderador_long_total,segmento, estrato, just_educ:ideo, nhogar1, m46_nhogar, m54, m30, m30b, m29_imp))
+                     )
+
+df_study1_wide$m54_6 <- df_study1_wide$m54_7
+
+# reshape from long to wide
+pacman::p_load(datasets,data.table)
+df_study1_long <- data.table::melt.data.table(data.table::setDT(df_study1_wide),
+              id.vars = c("idencuesta","muestra"),
+              variable.name = c("ola"),
+              measure = patterns("^comuna_","^comunacod_","^ponderador_long_total_", "^segmento_", "^estrato_", "^just_educ_", "^just_pension_", "^just_health_", "^mjp_","^merit_effort_","^merit_talent_", "^perc_sal_gerente_", "^perc_sal_obrero_", "^just_sal_gerente_", "^just_sal_obrero_", "^perc_inequality_", "^just_inequality_", "^educ_","^educyear_","^sex_", "^age_","^ess_", "^ideo_","^nhogar1_", "^m46_nhogar_", "^m54_", "^m30_", "^m30b_", "^m29_imp_"),
+              value.name = c("comuna","comunacod","ponderador_long_total","segmento","estrato", "just_educ", "just_pension","just_health","mjp","merit_effort","merit_talent", "perc_sal_gerente","perc_sal_obrero","just_sal_gerente","just_sal_obrero", "perc_inequality", "just_inequality", "educ","educyear","sex", "age", "ess", "ideo", "nhogar1", "m46_nhogar", "m54", "m30", "m30b", "m29_imp"),
+              na.rm = F,value.factor = TRUE
+              )
+
+df_study1_long <-
+  df_study1_long %>%
+  mutate(n_hogar =
+           dplyr::case_when(ola == 1 ~ nhogar1,
+                            ola == 2 ~ m46_nhogar,
+                            ola == 3 ~ m54,
+                            ola == 4 ~ m54,
+                            ola == 5 ~ m54,
+                            ola == 6 ~ m54,
+                            ola == 7 ~ m54))
+sjmisc::frq(df_study1_long$n_hogar)
+
+#Recode DK/DA to NA
+df_study1_long$n_hogar_r<-
+  car::recode(df_study1_long$n_hogar,"c(-888,-999)=NA")
+
+# Per capita household income:
+df_study1_long$ing_pc <- 
+  (df_study1_long$m29_imp/df_study1_long$n_hogar_r)
+
+df_study1_long$ing_pc <-
+sjlabelled::set_label(x = df_study1_long$ing_pc,
+                      label = "Household income per capita")
+
+sjmisc::descr(df_study1_long$ing_pc)
+
+# Compute income quintiles
+df_study1_long <- df_study1_long %>% 
+  group_by(ola) %>% 
+  mutate(quintil = ntile(-desc(ing_pc), 5)) %>% 
+  ungroup()
+
+df_study1_long$quintil <- 
+  factor(df_study1_long$quintil,
+         levels = c(1, 2, 3, 4, 5),
+         labels = c('Q1', 'Q2', 'Q3', 'Q4', 'Q5')) # Quintiles as factors
+
+#reverse quintile, reference level is the highest quintile
+#elsoc_long_2016_2023$quintil <- forcats::fct_rev(elsoc_long_2016_2023$quintil)
+
+df_study1_long$quintil <- 
+sjlabelled::set_label(x = df_study1_long$quintil,
+                      label = "Household income quintile per capita")
+
+sjmisc::frq(df_study1_long$quintil)
+
+#include new quintile category with missing cases
+df_study1_long$quintil1<-
+  car::recode(df_study1_long$quintil, 
+              "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")
+
+#elsoc_long_2016_2023$quintil1 <- factor(elsoc_long_2016_2023$quintil1, c("Q1","Q2","Q3","Q4","Q5","QNA"))
+
+df_study1_long$quintil1 <- 
+sjlabelled::set_label(x = df_study1_long$quintil1,
+                      label = "Household income quintile per capita (NA)") 
+sjmisc::frq(df_study1_long$quintil1)
+```
 
 ```{r}
-colSums(is.na(db_proc))
+# Reshape long to wide
+df_study1_wide <- df_study1_long %>% 
+  tidyr::pivot_wider(id_cols = c("idencuesta","muestra"),
+                     names_from = "ola",
+                     # names_prefix = c("educ","educyear","ing_pc","quintil","quintil1"),
+                     values_from = names(select(df_study1_long,comuna,comunacod,ponderador_long_total,segmento, estrato, just_educ:ideo, quintil, quintil1, ing_pc)))
 
-n_miss(db_proc)
+# fix data to w01 values
+df_study1_wide$age_2 <-df_study1_wide$age_1 #age
+df_study1_wide$age_3 <-df_study1_wide$age_1
+df_study1_wide$age_4 <-df_study1_wide$age_1
+df_study1_wide$age_5 <-df_study1_wide$age_1
+df_study1_wide$age_6 <-df_study1_wide$age_1
+df_study1_wide$age_7 <-df_study1_wide$age_1
 
-prop_miss(db_proc)*100
+df_study1_wide$sex_2 <-df_study1_wide$sex_1 #sex
+df_study1_wide$sex_3 <-df_study1_wide$sex_1
+df_study1_wide$sex_4 <-df_study1_wide$sex_1
+df_study1_wide$sex_5 <-df_study1_wide$sex_1
+df_study1_wide$sex_6 <-df_study1_wide$sex_1
+df_study1_wide$sex_7 <-df_study1_wide$sex_1
 
-miss_var_summary(db_proc)
+df_study1_wide$educ_2 <-df_study1_wide$educ_1 #education
+df_study1_wide$educ_3 <-df_study1_wide$educ_1
+df_study1_wide$educ_4 <-df_study1_wide$educ_1
+df_study1_wide$educ_5 <-df_study1_wide$educ_1
+df_study1_wide$educ_6 <-df_study1_wide$educ_1
+df_study1_wide$educ_7 <-df_study1_wide$educ_1
 
-miss_var_table(db_proc)
+df_study1_wide$educyear_2 <-df_study1_wide$educyear_1 #education years
+df_study1_wide$educyear_3 <-df_study1_wide$educyear_1
+df_study1_wide$educyear_4 <-df_study1_wide$educyear_1
+df_study1_wide$educyear_5 <-df_study1_wide$educyear_1
+df_study1_wide$educyear_6 <-df_study1_wide$educyear_1
+df_study1_wide$educyear_7 <-df_study1_wide$educyear_1
 
-vis_miss(db_proc) + theme(axis.text.x = element_text(angle=80))
+df_study1_wide$quintil1_2 <-df_study1_wide$quintil1_1 #quintiles
+df_study1_wide$quintil1_3 <-df_study1_wide$quintil1_1
+df_study1_wide$quintil1_4 <-df_study1_wide$quintil1_1
+df_study1_wide$quintil1_5 <-df_study1_wide$quintil1_1
+df_study1_wide$quintil1_6 <-df_study1_wide$quintil1_1
+df_study1_wide$quintil1_7 <-df_study1_wide$quintil1_1
 
+df_study1_wide$ing_pc_2 <-df_study1_wide$ing_pc_1 #income per capita
+df_study1_wide$ing_pc_3 <-df_study1_wide$ing_pc_1
+df_study1_wide$ing_pc_4 <-df_study1_wide$ing_pc_1
+df_study1_wide$ing_pc_5 <-df_study1_wide$ing_pc_1
+df_study1_wide$ing_pc_6 <-df_study1_wide$ing_pc_1
+df_study1_wide$ing_pc_7 <-df_study1_wide$ing_pc_1
+
+df_study1_wide$ess_2 <-df_study1_wide$ess_1 # subjective status
+df_study1_wide$ess_3 <-df_study1_wide$ess_1
+df_study1_wide$ess_4 <-df_study1_wide$ess_1
+df_study1_wide$ess_5 <-df_study1_wide$ess_1
+df_study1_wide$ess_6 <-df_study1_wide$ess_1
+df_study1_wide$ess_7 <-df_study1_wide$ess_1
+
+df_study1_wide$ideo_2 <-df_study1_wide$ideo_1 # political position
+df_study1_wide$ideo_3 <-df_study1_wide$ideo_1
+df_study1_wide$ideo_4 <-df_study1_wide$ideo_1
+df_study1_wide$ideo_5 <-df_study1_wide$ideo_1
+df_study1_wide$ideo_6 <-df_study1_wide$ideo_1
+df_study1_wide$ideo_7 <-df_study1_wide$ideo_1
+
+df_study1_wide$comuna_2 <-df_study1_wide$comuna_1 #comuna
+df_study1_wide$comuna_3 <-df_study1_wide$comuna_1
+df_study1_wide$comuna_4 <-df_study1_wide$comuna_1
+df_study1_wide$comuna_5 <-df_study1_wide$comuna_1
+df_study1_wide$comuna_6 <-df_study1_wide$comuna_1
+df_study1_wide$comuna_7 <-df_study1_wide$comuna_1
+
+df_study1_wide$comunacod_2 <-df_study1_wide$comunacod_1 #comuna
+df_study1_wide$comunacod_3 <-df_study1_wide$comunacod_1
+df_study1_wide$comunacod_4 <-df_study1_wide$comunacod_1
+df_study1_wide$comunacod_5 <-df_study1_wide$comunacod_1
+df_study1_wide$comunacod_6 <-df_study1_wide$comunacod_1
+df_study1_wide$comunacod_7 <-df_study1_wide$comunacod_1
+
+dim(df_study1_wide)
+
+sjPlot::view_df(df_study1_wide,
+                show.frq = T,show.values = T,show.na = T,show.prc = T, show.type = T)
 ```
 
-# Save and export
+```{r}
+# reshape from long to wide
+pacman::p_load(datasets,data.table)
+df_study1_long <- data.table::melt.data.table(data.table::setDT(df_study1_wide),
+              id.vars = c("idencuesta","muestra"),
+              variable.name = c("ola"),
+              measure = 
+patterns("^comuna_","^comunacod_","^ponderador_long_total_", "^segmento_", "^estrato_", "^just_educ_", "^just_pension_", "^just_health_", "^mjp_","^merit_effort_","^merit_talent_", "^perc_sal_gerente_", "^perc_sal_obrero_", "^just_sal_gerente_", "^just_sal_obrero_", "^perc_inequality_", "^just_inequality_", "^educ_","^educyear_","^sex_", "^age_","^ess_", "^ideo_","^quintil_", "^quintil1_", "^ing_pc_"),
+              value.name = c("comuna","comunacod","ponderador_long_total","segmento","estrato", "just_educ", "just_pension","just_health","mjp","merit_effort","merit_talent", "perc_sal_gerente","perc_sal_obrero","just_sal_gerente","just_sal_obrero", "perc_inequality", "just_inequality", "educ","educyear","sex", "age", "ess", "ideo", "quintil", "quintil1", "ing_pc"),
+              na.rm = F,value.factor = TRUE
+              )
+```
 
 ```{r}
-db_proc <- db_proc %>% 
-  select(-c(perc_sal_gerente, perc_sal_obrero, just_sal_gerente, just_sal_obrero, m01))
+names(df_study1_long) #check names of long dataset
+dim(df_study1_long) #check dimensions of the dataframe
+# Original dataset with 7 waves
+df_study2_long <- df_study1_long
 
-df1 <- db_proc %>% na.omit()
+# filter the dataset for the waves 1 to 4 and 6 to 7
+df_study1_long <-
+df_study1_long %>% 
+  filter(ola %in% c(1,2,3,4,6,7)) %>% 
+  mutate(ola=factor(ola,levels = 1:7,labels = 1:7))
+dim(df_study1_long) #check, now is OK
 
-df1 <- sjlabelled::copy_labels(df_new = df1, df_origin = db_proc)
+# df_study1_long <- 
+# set_label(x = df_study1_long,
+#           label = get_label(select(df_study1,names(df_study1_long))))
 
-save(df1, file = here("input/data/proc/df1.RData"))
+sjPlot::view_df(df_study1_long,
+                show.frq = T,show.values = T,show.na = T,show.prc = T,
+                show.type = T)
+#______________________________________________________________________________
+# obtain the idencuesta for wave 7
+ids <- 
+  df_study1 %>% 
+  select(idencuesta,ola) %>% 
+  filter(ola==7) %>% 
+  sjmisc::frq(idencuesta,show.na = F) %>% as.data.frame()
+
+# filter data by the idencuesta of t7
+df_study1_long_t7 <- 
+  df_study1_long %>%
+  filter(idencuesta %in% ids$val)
+
+names(df_study1_long_t7)
+dim(df_study1_long_t7)
+sjmisc::frq(df_study1_long_t7$ola)
 ```
+
+```{r}
+save(df_study1_long,file = here::here("input/data/proc/df_study1_long.RData"))
+save(df_study1_long_t7,file = here::here("input/data/proc/df_study1_long_t7.RData"))
+save(df_study2_long,file = here::here("input/data/proc/df_study2_long.RData"))
+```

### 2025-04-04 processing/02-analysis.qmd
Update analysis

--- 2025-03-04 processing/02-analysis.qmd
@@
 
-This is the analysis code for the paper "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)". The dataset used is `df1.RData`.
+This is the analysis code for the paper "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)".
+The dataset used is `df_study1_long_t7.RData`.
 
@@
 
-load(file = here("input/data/proc/df1.RData"))
+load(file = here("input/data/proc/df_study1_long_t7.RData"))
 
-glimpse(df1)
+glimpse(df_study1_long_t7)
 
@@
 #| tbl-cap-location: top
-t1 <- df1 %>% 
+t1 <- df_study1_long_t7 %>% 
   select(just_health, just_pension, just_educ, mjp, perc_inequality, just_inequality, merit_effort, merit_talent, ess, ideo, educ, sex, age)
@@
 
-datos.health <- df1 %>% 
-  group_by(idencuesta, wave) %>% 
+datos.health <- df_study1_long_t7 %>% 
+  group_by(idencuesta, ola) %>% 
   count(just_health) %>% 
-  group_by(wave) %>% 
+  group_by(ola) %>% 
   mutate(porcentaje=n/sum(n)) %>% 
@@
   na.omit() %>% 
-  mutate(wave = case_when(wave == 1 ~ "2016",
-                          wave == 2 ~ "2017",
-                          wave == 3 ~ "2018",
-                          wave == 4 ~ "2019",
-                          wave == 5 ~ "2022",
-                          wave == 6 ~ "2023"),
+  mutate(wave = case_when(ola == 1 ~ "2016",
+                          ola == 2 ~ "2017",
+                          ola == 3 ~ "2018",
+                          ola == 4 ~ "2019",
+                          ola == 6 ~ "2022",
+                          ola == 7 ~ "2023"),
          wave = factor(wave, levels = c("2016",
@@
 
-etiquetas.health <- df1 %>%
-  group_by(wave, just_health) %>%
+etiquetas.health <- df_study1_long_t7 %>%
+  group_by(ola, just_health) %>%
   summarise(count = n(), .groups = "drop") %>%
-  group_by(wave) %>%
+  group_by(ola) %>%
   mutate(porcentaje = count / sum(count)) %>% 
@@
   mutate(idencuesta = 1,
-         wave = case_when(wave == 1 ~ "2016",
-                          wave == 2 ~ "2017",
-                          wave == 3 ~ "2018",
-                          wave == 4 ~ "2019",
-                          wave == 5 ~ "2022",
-                          wave == 6 ~ "2023"),
+         wave = case_when(ola == 1 ~ "2016",
+                          ola == 2 ~ "2017",
+                          ola == 3 ~ "2018",
+                          ola == 4 ~ "2019",
+                          ola == 6 ~ "2022",
+                          ola == 7 ~ "2023"),
          wave = factor(wave, levels = c("2016",
@@
 
-datos.pension <- df1 %>% 
-  group_by(idencuesta, wave) %>% 
+datos.pension <- df_study1_long_t7 %>% 
+  group_by(idencuesta, ola) %>% 
   count(just_pension) %>% 
-  group_by(wave) %>% 
+  group_by(ola) %>% 
   mutate(porcentaje=n/sum(n)) %>% 
@@
 
-etiquetas.pension <- df1 %>%
-  group_by(wave, just_pension) %>%
+etiquetas.pension <- df_study1_long_t7 %>%
+  group_by(ola, just_pension) %>%
   summarise(count = n(), .groups = "drop") %>%
-  group_by(wave) %>%
+  group_by(ola) %>%
   mutate(porcentaje = count / sum(count)) %>% 
@@
 
-datos.educ <- df1 %>% 
-  group_by(idencuesta, wave) %>% 
+datos.educ <- df_study1_long_t7 %>% 
+  group_by(idencuesta, ola) %>% 
   count(just_educ) %>% 
-  group_by(wave) %>% 
+  group_by(ola) %>% 
   mutate(porcentaje=n/sum(n)) %>% 
@@
 
-etiquetas.educ <- df1 %>%
-  group_by(wave, just_educ) %>%
+etiquetas.educ <- df_study1_long_t7 %>%
+  group_by(ola, just_educ) %>%
   summarise(count = n(), .groups = "drop") %>%
-  group_by(wave) %>%
+  group_by(ola) %>%
   mutate(porcentaje = count / sum(count)) %>% 
@@
 ```{r}
-df1 %>% 
+df_study1_long_t7 %>%
+  filter(ola == 7) %>% 
   select(mjp, perc_inequality, just_inequality, merit_effort, merit_talent) %>% 
@@
 ```{r}
-m0 <- lmer(mjp ~ 1 + (1 | idencuesta), 
-                data = df1)
+# Generate analytical sample
 
-performance::icc(m0, by_group = T)
-```
+df_study1 <- df_study1_long_t7 %>%
+  select(idencuesta,
+         ola,
+         ponderador_long_total,
+         mjp, 
+         perc_inequality, 
+         just_inequality,
+         merit_effort, 
+         merit_talent, 
+         educ,
+         quintil1,
+         sex,
+         age,
+         ess, 
+         ideo) %>% 
+  na.omit() %>% 
+  mutate(ola = as.factor(ola),
+         ola_num = as.numeric(ola),
+         ola_2=as.numeric(ola)^2)
 
-```{r}
-df1$merit_effort <- as_numeric(df1$merit_effort)
-df1$merit_talent <- as_numeric(df1$merit_talent)
-df1 <- df1 %>% 
-  mutate(wave_f = case_when(wave == 1 ~ "2016",
-                          wave == 2 ~ "2017",
-                          wave == 3 ~ "2018",
-                          wave == 4 ~ "2019",
-                          wave == 5 ~ "2022",
-                          wave == 6 ~ "2023"),
-         wave_f = factor(wave_f, levels = c("2016",
+df_study1 <- df_study1 %>%
+  group_by(idencuesta) %>%             # Agrupar por el identificador del participante
+  mutate(n_participaciones = n()) %>%  # Contar el número de filas (participaciones) por participante
+  ungroup()
+
+df_study1 <- df_study1 %>% filter(n_participaciones>1)
+
+df_study1$merit_effort <- as_numeric(df_study1$merit_effort)
+df_study1$merit_talent <- as_numeric(df_study1$merit_talent)
+
+df_study1 <- df_study1 %>% 
+  mutate(ola = case_when(ola == 1 ~ "2016",
+                         ola == 2 ~ "2017",
+                         ola == 3 ~ "2018",
+                         ola == 4 ~ "2019",
+                         ola == 6 ~ "2022",
+                         ola == 7 ~ "2023"),
+         ola = factor(ola, levels = c("2016",
                                         "2017",
@@
 
-df1 <- df1 %>% 
+df_study1 <- df_study1 %>% 
   group_by(idencuesta) %>% 
@@
   ungroup()
+```
 
-m1 <- lmer(mjp ~ 1 + perc_inequality + (1 | idencuesta),
-                data = df1)
+```{r}
+m0 <- lmer(mjp ~ 1 + (1 | idencuesta), 
+                data = df_study1)
 
-m2 <- lmer(mjp ~ 1 + just_inequality + (1 | idencuesta),
-                data = df1)
+performance::icc(m0, by_group = T)
 
-m3 <- lmer(mjp ~ 1 + perc_inequality + just_inequality + (1 | idencuesta),
-                data = df1)
+m1 <- lmer(mjp ~ 1 + ola + (1 | idencuesta),
+                data = df_study1)
 
-m4 <- lmer(mjp ~ 1 + perc_inequality + just_inequality + 
-             merit_effort +  (1 | idencuesta),
-                data = df1)
+m2 <- lmer(mjp ~ 1 + ola_num + ola_2 + (1 | idencuesta),
+                data = df_study1)
 
-m5 <- lmer(mjp ~ 1 + perc_inequality + just_inequality + 
-             merit_effort + merit_talent + (1 | idencuesta),
-                data = df1)
+m3 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + (1 | idencuesta), data = df_study1)
 
-m6 <- lmer(mjp ~ 1 + perc_inequality + just_inequality + 
-             merit_effort + merit_talent + perc_inequality_mean +
-             just_inequality_mean + merit_effort_mean + 
-             merit_talent_mean + (1 | idencuesta),
-                data = df1)
+m4 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + (1 | idencuesta),
+                data = df_study1)
 
-m7 <- lmer(mjp ~ 1 + perc_inequality + just_inequality + 
-             merit_effort + merit_talent + perc_inequality_mean +
-             just_inequality_mean + merit_effort_mean + 
-             merit_talent_mean + age + sex + ess + ideo + educ + wave_f + (1 | idencuesta),
-                data = df1)
+m5 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + (1 | idencuesta),
+                data = df_study1)
+
+m6 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + (1 | idencuesta),
+                data = df_study1)
+
+m7 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + (1 | idencuesta),
+                data = df_study1)
+
+m8 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 | idencuesta),
+                data = df_study1)
+
 ```
@@
   "(Intercept)" = "Intercept",
+  "ola2017" = "Wave 2017",
+  "ola2018" = "Wave 2018",
+  "ola2019" = "Wave 2019",
+  "ola2022" = "Wave 2022",
+  "ola2023" = "Wave 2023",
+  ola_num = "Wave",
+  ola_2 = "Wave^2",
   perc_inequality = "Perception inequality (WE)",
@@
   merit_talent_mean = "Merit: Talent (BE)",
-  "age30-49" = "30-49",
-  "age50-64" = "50-64",
-  "age65 or more" = "65 or more",
-  sexFemale = "Female (Ref.= Male)",
+  educUniversitary = "Universitary (Ref.= Less than universitary)",
+  quintil1Q2 = "Quintile 2",
+  quintil1Q3 = "Quintile 3",
+  quintil1Q4 = "Quintile 4",
+  quintil1Q5 = "Quintile 5",
+  quintil1QNA = "Quintile NA",
   ess = "Subjective social status",
@@
   "ideoDoes not identify" = "Does not identify",
-  educUniversitary = "Universitary (Ref.= Less than universitary)",
-  wave_f2017 = "2017",
-  wave_f2018 = "2018",
-  wave_f2019 = "2019",
-  wave_f2022 = "2022",
-  wave_f2023 = "2023"
+  sexFemale = "Female (Ref.= Male)",
+  "age30-49" = "30-49",
+  "age50-64" = "50-64",
+  "age65 or more" = "65 or more"
   )
 
-texreg::htmlreg(list(m1, m2, m3, m4, m5, m6, m7),
-               custom.model.names = c(paste0("Model ", seq(1:7))),
+texreg::htmlreg(list(m1, m2, m3, m4, m5, m6, m7, m8),
+               custom.model.names = c(paste0("Model ", seq(1:8))),
                caption = NULL,
@@
                digits = 3,
-               groups = list("Age (Ref. = 18-29)" = 10:12,"Political identification (Ref.= Left)" = 15:17, "Wave (Ref.= 2016)" = 19:23),
-               custom.note = "\\item Note: Cells contain regression coefficients with standard errors in parentheses.
-%stars.
-\\\\ \\item Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 4364).",
+               groups = list("Wave (Ref.= 2016)" = 2:6, "Income (Ref.= Quintile 1)" = 18:22, "Political identification (Ref.= Left)" = 24:26, "Age (Ref. = 18-29)" = 28:30),
+               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
                threeparttable = T,

### 2025-04-04 processing/borrador_prep.qmd
Update analysis

---
title: "Data preparation"
subtitle: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
author: "Researcher"
date: today
lang: en
fontsize: 12pt
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 2
    toc-expand: 2
    toc-title: Contents
    number-sections: true
    number-depth: 3
    theme:
      - cosmo
      - edumer_html.scss
    code-link: true
    title-block-banner: true
  pdf:
    number-sections: true
    number-depth: 3
editor_options: 
  chunk_output_type: console
---

# Presentation

This is the data preparation code for the paper "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)". The prepared dataset is `ELSOC_Long_2016_2023_1.00.RData`.

```{r}
#| label: set
#| echo: false
#| message: false
#| warning: false

library(knitr)
knitr::opts_chunk$set(echo = TRUE, include = TRUE, warning = FALSE, message = FALSE)

table_format <- if(is_html_output()) {
  "html"
} else if(is_latex_output()) {
  "latex"
}
table_format2 <- if(is_html_output()) {
  T
} else if(is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")
```

# Libraries

```{r}
#| label: packages
#| include: true

if (! require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse,
               sjmisc, 
               here,
               sjlabelled,
               SciViews,
               naniar)

options(scipen=999)
rm(list = ls())
```

# Data

```{r}
#| label: data
#| echo: true
#| output: false

load(url("https://dataverse.harvard.edu/api/access/datafile/10797987"))

glimpse(elsoc_long_2016_2023)

```

# Processing

## Select

```{r}
db_proc <- elsoc_long_2016_2023 %>% 
  select(idencuesta, ola, 
         ponderador_long_total,segmento, estrato,
         just_educ = d02_02, just_pension = d02_01, 
         just_health = d02_03, merit_effort = c18_09,
         merit_talent = c18_10, perc_sal_gerente = d03_01,
         perc_sal_obrero = d03_02, just_sal_gerente = d04_01, 
         just_sal_obrero = d04_02, age = m0_edad, m01, 
         sex = m0_sexo, ess = d01_01, ideo = c15
         ) %>% 
  as_tibble()
```

## Filter

```{r}
#db_proc %>% 
#  group_by(ola) %>% 
#  count(just_educ,just_pension,just_health) %>% 
#  na.omit() %>% 
#  print(n = nrow(.))

db_proc <- db_proc %>% dplyr::filter(ola %in% c(1,2,3,4,6,7))

```

## Recode and transform

```{r}
# general na's
db_proc[db_proc ==-999] <- NA
db_proc[db_proc ==-888] <- NA
db_proc[db_proc ==-777] <- NA
db_proc[db_proc ==-666] <- NA

```

```{r}
# Market Justice Preferences

frq(db_proc$just_health)
frq(db_proc$just_pension)
frq(db_proc$just_educ)

db_proc <- cbind(db_proc, "mjp" = rowMeans(db_proc %>% select(just_health, just_pension, just_educ), na.rm=TRUE))

summary(db_proc$mjp)

db_proc <- db_proc %>% 
  mutate(
    across(
      .cols = c(just_health, just_pension, just_educ),
      .fns = ~ car::recode(., recodes = c("1='Strongly desagree'; 2='Desagree';
                                          3='Neither agree nor desagree'; 4='Agree';
                                          5='Strongly agree'"), 
                           levels = c("Strongly desagree", "Desagree", "Neither agree nor desagree", "Agree", "Strongly agree"),
                           as.factor = T)
    )
  )

db_proc$just_health <- sjlabelled::set_label(db_proc$just_health, 
                        label = "Health distributive justice")

db_proc$just_pension <- sjlabelled::set_label(db_proc$just_pension, 
                        label = "Pension distributive justice")

db_proc$just_educ <- sjlabelled::set_label(db_proc$just_educ, 
                        label = "Education distributive justice")

db_proc$mjp <- sjlabelled::set_label(db_proc$mjp, 
                        label = "Market justice preferences")
```

```{r}
# Meritocracy

frq(db_proc$merit_effort)
frq(db_proc$merit_talent)

db_proc <- db_proc %>% 
  mutate(
    across(
      .cols = c(merit_effort, merit_talent),
      .fns = ~ car::recode(., recodes = c("1='Strongly desagree'; 2='Desagree';
                                          3='Neither agree nor desagree'; 4='Agree';
                                          5='Strongly agree'"), 
                           levels = c("Strongly desagree", "Desagree", "Neither agree nor desagree", "Agree", "Strongly agree"),
                           as.factor = T)
    )
  )

db_proc$merit_effort <- sjlabelled::set_label(db_proc$merit_effort, 
                        label = "People are rewarded for their efforts")

db_proc$merit_talent <- sjlabelled::set_label(db_proc$merit_talent, 
                        label = "People are rewarded for their intelligence")

```

```{r}
# Inequality

# perc inequality gap

frq(db_proc$perc_sal_gerente)
frq(db_proc$perc_sal_obrero)

db_proc <- db_proc %>% 
  mutate(perc_sal_obrero=replace(perc_sal_obrero, perc_sal_obrero <= 40000  | perc_sal_obrero>= 1000000, NA)) %>%
  mutate(perc_sal_gerente=replace(perc_sal_gerente, perc_sal_gerente <= 250000 | perc_sal_gerente>= 100000001, NA))

db_proc$perc_inequality <- SciViews::ln(db_proc$perc_sal_gerente/db_proc$perc_sal_obrero)

summary(db_proc$perc_inequality)

db_proc$perc_inequality <- sjlabelled::set_label(db_proc$perc_inequality, 
                        label = "Inequality gap perception")

# just inequality gap

frq(db_proc$just_sal_gerente)
frq(db_proc$just_sal_obrero)

db_proc <- db_proc %>% 
  mutate(just_sal_obrero=replace(just_sal_obrero, just_sal_obrero <= 40000  | just_sal_obrero>=999999999999999, NA)) %>%
  mutate(just_sal_gerente=replace(just_sal_gerente, just_sal_gerente <= 100000 | just_sal_gerente>=999999999999999, NA))

db_proc$just_inequality <- SciViews::ln(db_proc$just_sal_gerente/db_proc$just_sal_obrero)

summary(db_proc$just_inequality)

db_proc$just_inequality <- sjlabelled::set_label(db_proc$just_inequality, 
                        label = "Inequality gap justification")

```

```{r}
# Controls

# sex
db_proc$sex <- car::recode(db_proc$sex, 
                           recodes = c("1='Male'; 2='Female'"), 
                           levels = c("Male", "Female"),
                           as.factor = T)

db_proc$sex <- sjlabelled::set_label(db_proc$sex, 
                        label = "Gender")

# age
frq(db_proc$age)

db_proc$age <- 
  factor(car::recode(db_proc$age, 
                     "18:29=1;30:49=2;50:64=3;65:150=4"),
         labels = c('18-29', '30-49', '50-64', '65 or more'))
db_proc$age <-
  sjlabelled::set_label(db_proc$age, 
                        label = c("Age groups"))

# political indentification

frq(db_proc$ideo)

db_proc$ideo<-
factor(
  car::recode(
    db_proc$ideo,
    "c(11,12,-888,-999)='Does not identify';c(0,1,2,3,4)='Left';
     c(5)='Center';c(6,7,8,9,10)='Right'"
  ),
  levels = c('Left', 'Center', 'Right', 'Does not identify')
)

db_proc$ideo<- factor(db_proc$ideo,levels = levels(db_proc$ideo))

db_proc$ideo <- 
sjlabelled::set_label(x = db_proc$ideo, 
                      label = "Political identification")

frq(db_proc$ideo)

# ess
sjmisc::frq(db_proc$ess)

db_proc$ess <- as.numeric(db_proc$ess)

db_proc$ess <-  sjlabelled::set_label(x = db_proc$ess,
                      label = "Subjective Social Status")

```

```{r}
# Socioeconomic

# Education

frq(db_proc$m01)

db_proc$educ <- 
  car::recode(db_proc$m01,
              "c(1,2,3,4,5,6,7)=1;c(8,9,10)=2; c(-888,-999)=NA")
db_proc$educ <-
  factor(db_proc$educ,
         labels = c("Less than Universitary","Universitary"))

db_proc$educ <- 
sjlabelled::set_label(x = db_proc$educ,
                      label = "Education")

#Recoding of education to years based on casen 2017.
db_proc$educyear<- as.numeric(
  car::recode(db_proc$m01, 
              "1=0;2=4.3;3=7.5;4=9.8;5=12.02;6=13.9;
               7=14.8;8=14.9;9=16.9;10=19.07;c(-888,-999)=NA", 
              as.numeric = T))

db_proc$educyear <- 
sjlabelled::set_label(x = db_proc$educyear,
                      label = "Education in years")

class(db_proc$educyear)
frq(db_proc$educyear)

```

```{r}
# Household income----------------------------------------------------------------

```

```{r}
# Household income----------------------------------------------------------------

# Ltop is the lower limit of the top category, Ltop-1 is the lower limit of the category before the
# top one, ftop is the frequency in the top category, and ftop-1 is the frequency in the
# category before the top one.

sjmisc::frq( elsoc_long_2016_2023$m30)
Ltop_1<- 1850001
Ltop  <- 2700000

ftop_1 <- 42
ftop   <- 34

V  = (log(ftop_1 + ftop) - log(ftop)) / (log(Ltop) -log(Ltop_1))
M_top = 0.5* Ltop *(1+(V/(V-1)));M_top # = $3.897.232

#Impute midpoint of income ranges
elsoc_long_2016_2023$m30_rec <-
  as.numeric(car::recode( elsoc_long_2016_2023$m30,
                         "1=110000;2=251000;3=305000;4=355000;5=400000;
            6=445000;7=490000;8=535000;9=585000;10=640000;11=700000;12=765000;
            13=845000;14=935000;15=1040000;16=1180000;17=1375000;18=1670000;
            19=2275000;20=3897232;NA=NA;c(-888,-999)=NA"))

summary( elsoc_long_2016_2023$m30_rec)

#Impute midpoint of income ranges (2021)

frq(elsoc_long_2016_2023$m30b)
Ltop_1<- 450000
Ltop  <- 700000
ftop_1 <- 18
ftop   <- 27

V  = (log(ftop_1 + ftop) - log(ftop)) / (log(Ltop) -log(Ltop_1))
M_top = 0.5* Ltop *(1+(V/(V-1))) # = $2.941.412

 elsoc_long_2016_2023$m30b_rec <-
  as.numeric(car::recode( elsoc_long_2016_2023$m30b,
                         "1=125000;2=300000;3=400000;4=575000;5=2941412;NA=NA;c(-888,-999)=NA"))
summary( elsoc_long_2016_2023$m30b_rec)

# Combine m30_rec with m30b_rec
 elsoc_long_2016_2023$m30_rec <- 
  ifelse(is.na( elsoc_long_2016_2023$m30_rec),
         yes =  elsoc_long_2016_2023$m30b_rec,
         no =  elsoc_long_2016_2023$m30_rec)
summary( elsoc_long_2016_2023$m30_rec)

#Recode DK/DA of Income to NA
elsoc_long_2016_2023$m29_rec <-
  as.numeric(car::recode( elsoc_long_2016_2023$m29,"c(-888,-999)=NA"))

#replace NA of income with new imputed variable
elsoc_long_2016_2023$m29_imp <- 
  ifelse(test = !is.na( elsoc_long_2016_2023$m29_rec),
         yes =   elsoc_long_2016_2023$m29_rec,
         no =   elsoc_long_2016_2023$m30_rec)
summary( elsoc_long_2016_2023$m29_imp)

#  elsoc_long_2016_2023$m29_imp <- 
#   ifelse(test = is.na( elsoc_long_2016_2023$m29_imp),
#          yes =   elsoc_long_2016_2023$m30b_rec,
#          no =   elsoc_long_2016_2023$m29_imp)

summary( elsoc_long_2016_2023$m29_imp)

# N Household:
elsoc_long_2016_2023 <-
   elsoc_long_2016_2023 %>%
  mutate(n_hogar =
           dplyr::case_when(ola == 1 ~ nhogar1,
                            ola == 2 ~ m46_nhogar,
                            ola == 3 ~ m54,
                            ola == 4 ~ m54,
                            ola == 5 ~ m54,
                            ola == 6 ~ m54,
                            ola == 7 ~ m54))

summary( elsoc_long_2016_2023$n_hogar)
#imputar tamanio del hogar de ola 5 a ola 6
elsoc_long_2016_2023 <- 
   elsoc_long_2016_2023 %>%  
  group_by(idencuesta) %>%
  mutate(n_hogar = if_else(ola == 6 & is.na(n_hogar), last(n_hogar[ola == 5]), n_hogar)) %>% 
  ungroup()

sjmisc::frq( elsoc_long_2016_2023$n_hogar)
table( elsoc_long_2016_2023$n_hogar, elsoc_long_2016_2023$ola)

#Recode DK/DA to NA
 elsoc_long_2016_2023$n_hogar_r<-
  car::recode( elsoc_long_2016_2023$n_hogar,"c(-888,-999)=NA")

table( elsoc_long_2016_2023$m13, elsoc_long_2016_2023$n_hogar_r)

#ingresos individuales para hogares de 1 persona
 elsoc_long_2016_2023 %>% sjmisc::find_var('Ingreso')
 elsoc_long_2016_2023$inc_ind <- ifelse( elsoc_long_2016_2023$n_hogar_r==1, elsoc_long_2016_2023$m13,NA)
summary( elsoc_long_2016_2023$inc_ind)
# 
# c(40000,
# (40001 + 85000) / 2,
# (85001 + 125000) / 2,
# (125001 + 170000) / 2,
# (170001 + 210000) / 2,
# (210001 + 230000) / 2,
# (230001 + 280000) / 2,
# (280001 + 320000) / 2,
# (320001 + 360000) / 2,
# (360001 + 400000) / 2,
# (400001 + 465000) / 2,
# (465001 + 540000) / 2,
# (540001 + 665000) / 2,
# (665001 + 850000) / 2,
# (850001 + 1300000) / 2,
# 1300001)

frq( elsoc_long_2016_2023$m14)
Ltop_1<- 850001
Ltop  <- 1300001 
ftop_1 <- 65
ftop   <- 40

V  = (log(ftop_1 + ftop) - log(ftop)) / (log(Ltop) -log(Ltop_1))
M_top = 0.5* Ltop *(1+(V/(V-1))) # = $1.811.247

 elsoc_long_2016_2023$m14_rec<- 
  car::recode( elsoc_long_2016_2023$m14,
              "1=40000.0;2=62500.5;3=105000.5;4=147500.5;5=190000.5;6=220000.5;
              7=255000.5;8=300000.5;9=340000.5;10=380000.5;11=432500.5;12=502500.5;
              13=602500.5;14=757500.5;15=1075000.5;16=1811247"
  )

 elsoc_long_2016_2023$inc_ind <- ifelse( elsoc_long_2016_2023$n_hogar_r==1 & is.na( elsoc_long_2016_2023$inc_ind), elsoc_long_2016_2023$m14_rec, elsoc_long_2016_2023$inc_ind)
summary( elsoc_long_2016_2023$inc_ind)

# Ingreso del hogar pegarle ingreso del hogar unipersonal
summary( elsoc_long_2016_2023$m29_imp)
 elsoc_long_2016_2023$m29_imp <- 
  ifelse(is.na( elsoc_long_2016_2023$m29_imp) &  elsoc_long_2016_2023$n_hogar_r==1,
          elsoc_long_2016_2023$inc_ind,
          elsoc_long_2016_2023$m29_imp) #recupero 14 casos
summary( elsoc_long_2016_2023$m29_imp)

# Household Equivalent income 
# 1.0 to the first adult;
# 0.5 to the second and each subsequent person aged 14 and over;
# 0.3 to each child aged under 14.

 elsoc_long_2016_2023$n_hogar_r_eq <- ( elsoc_long_2016_2023$n_hogar_r + 1) /2
table( elsoc_long_2016_2023$n_hogar_r_eq)

frq( elsoc_long_2016_2023$n_hogar_r_eq)

# Per capita household income:
 elsoc_long_2016_2023$ing_pc <- 
  (elsoc_long_2016_2023$m29_imp/ elsoc_long_2016_2023$n_hogar_r)

# Per capita household income equivalized (OECD):
 elsoc_long_2016_2023$ing_pc_eq <- 
  ( elsoc_long_2016_2023$m29_imp/ elsoc_long_2016_2023$n_hogar_r_eq)

descr( elsoc_long_2016_2023$ing_pc_eq)
descr( elsoc_long_2016_2023$ing_pc)
 elsoc_long_2016_2023$ing_pc <-
  sjlabelled::set_label(x =  elsoc_long_2016_2023$ing_pc,
                        label = "Household income per capita")

sjmisc::descr( elsoc_long_2016_2023$ing_pc)

# Compute income quintiles
 elsoc_long_2016_2023 <-  elsoc_long_2016_2023 %>% 
  group_by(ola) %>% 
  mutate(quintil = ntile(-desc(ing_pc), 5)) %>% 
  ungroup()

 elsoc_long_2016_2023$quintil <- 
  factor( elsoc_long_2016_2023$quintil,
         levels = c(1, 2, 3, 4, 5),
         labels = c('Q1', 'Q2', 'Q3', 'Q4', 'Q5')) # Quintiles as factors

elsoc_long_2016_2023$quintil <- 
  sjlabelled::set_label(x =  elsoc_long_2016_2023$quintil,
                        label = "Household income quintile per capita")

sjmisc::frq( elsoc_long_2016_2023$quintil)

#include new quintile category with missing cases
 elsoc_long_2016_2023$quintil1<-
  car::recode( elsoc_long_2016_2023$quintil, 
              "'Q1'='Q1';'Q2'= 'Q2';'Q3'='Q3';'Q4'='Q4';'Q5'='Q5'; NA='QNA'")

# elsoc_long_2016_2023$quintil1 <- factor( elsoc_long_2016_2023$quintil1, c("Q1","Q2","Q3","Q4","Q5","QNA"))

 elsoc_long_2016_2023$quintil1 <- 
  sjlabelled::set_label(x =  elsoc_long_2016_2023$quintil1,
                        label = "Household income quintile per capita (NA)") 
sjmisc::frq( elsoc_long_2016_2023$quintil1)

```

```{r}
elsoc_income <- elsoc_long_2016_2023 %>% 
  dplyr::select(idencuesta, ola, ing_pc, quintil, quintil1) %>% 
  dplyr::filter(ola %in% c(1,2,3,4,6,7))

db_proc <- left_join(db_proc, elsoc_income, by = c("idencuesta", "wave" = "ola"))

```

```{r}
# wave 
db_proc <- db_proc %>% 
  rename(wave = ola) %>% 
  mutate(wave = case_when(wave == 1 ~ 1,
                          wave == 2 ~ 2,
                          wave == 3 ~ 3,
                          wave == 4 ~ 4,
                          wave == 6 ~ 5,
                          wave == 7 ~ 6,
                          TRUE ~ NA_real_))

db_proc$wave <- sjlabelled::set_label(db_proc$wave, 
                        label = "Wave")
```

## Missing values

```{r}
colSums(is.na(db_proc))

n_miss(db_proc)

prop_miss(db_proc)*100

miss_var_summary(db_proc)

miss_var_table(db_proc)

vis_miss(db_proc) + theme(axis.text.x = element_text(angle=80))

```

# Save and export

```{r}
db_proc <- db_proc %>% 
  select(-c(perc_sal_gerente, perc_sal_obrero, just_sal_gerente, just_sal_obrero, m01))

df1 <- db_proc %>% na.omit()

df1 <- sjlabelled::copy_labels(df_new = df1, df_origin = db_proc)

save(df1, file = here("input/data/proc/df1.RData"))
```

### 2025-04-04 processing/02-analysis.qmd
Up analysis

@@
 
-m2 <- lmer(mjp ~ 1 + ola_num + ola_2 + (1 | idencuesta),
+m1.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + (1 | idencuesta),
                 data = df_study1)
 
-m3 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + (1 | idencuesta), data = df_study1)
-
-m4 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + (1 | idencuesta),
+m2 <- lmer(mjp ~ 1 + ola_num + ola_2 + (1 + ola_num | idencuesta),
                 data = df_study1)
 
-m5 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + (1 | idencuesta),
+anova(m1.1, m2)
+
+m3 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + (1 + ola_num | idencuesta), 
            data = df_study1)
 
-m6 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + (1 | idencuesta),
+m4 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + (1 + ola_num | idencuesta),
                 data = df_study1)
 
-m7 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + (1 | idencuesta),
+m5 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + (1 + ola_num | idencuesta),
                 data = df_study1)
 
-m8 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 | idencuesta),
+m6 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + (1 + ola_num | idencuesta),
                 data = df_study1)
 
+m7 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + (1 + ola_num | idencuesta),
+                data = df_study1)
+
+m8 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
+                data = df_study1)
+
+m9 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
+                data = df_study1)
+
+m10 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + just_inequality*ola_num + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
+                data = df_study1)
+
+m11 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + just_inequality*ola_num + merit_effort*ola_num + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
+                data = df_study1)
+
+m12 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + just_inequality*ola_num + merit_effort*ola_num + merit_talent*ola_num + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
+                data = df_study1)
 ```
@@
 ```
+
+```{r}
+#| label: tbl-interactions
+#| results: asis
+#| tbl-cap: "Time interactions within effects for market justice preferences"
+#| tbl-cap-location: top
+
+ccoef <- list(
+  "(Intercept)" = "Intercept",
+  "ola_num:perc_inequality" = "Perception inequality (WE) x Wave",
+  "ola_num:just_inequality" = "Justification inequality (WE) x Wave",
+  "ola_num:merit_effort" = "Merit: Effort (WE) x Wave",
+  "ola_num:merit_talent" = "Merit: Talent (WE) x Wave"
+  )
+
+texreg::htmlreg(list(m9,m10,m11,m12),
+               custom.model.names = c(paste0("Model ", seq(9:12))),
+               caption = NULL,
+               stars = c(0.05, 0.01, 0.001),
+               custom.coef.map = ccoef,
+               digits = 3,
+               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
+               threeparttable = T,
+               leading.zero = T,
+               float.pos = "h!",
+               use.packages = F,
+               booktabs = T,
+               scalebox = 1)
+```

### 2025-04-11 03-methods.qmd
Update analysis and repo order

---
author: "Equipo EDUMER"
bibliography: "input/bib/refs.bib"
csl: "input/bib/apa6.csl"
---

# Data, Variables and Methods

## Data

## Variables

## Methods

### 2025-04-11 paper.qmd
Update analysis and repo order

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  docx:
    number-sections: true
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  pdf: 
    geometry: 
      - top=30mm
      - left=20mm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/refs.bib"
csl: "input/bib/apa6.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  My abstract. \newline
  **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
header-includes:
  - \usepackage{threeparttable}
---

```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 rm(list=ls())
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
```

```{r}
# This document was last modified at {{< modified >}}
# 
# and it was last rendered at {{< now >}}

```

{{< include 01-introduction.qmd >}}

{{< include 02-antecedents.qmd >}}

{{< include 03-methods.qmd >}}

{{< include 04-analisis.qmd >}}

{{< include 05-discussion.qmd >}}

{{< include 06-conclusion.qmd >}}

{{< include 07-supplementary-material.qmd >}}

### 2025-04-11 paper.tex
Update analysis and repo order

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Times New Roman}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[top=30mm,left=20mm]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\usepackage{threeparttable}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)},
  pdfauthor={Equipo EDUMER},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Perceptions of Inequality and Meritocracy: Their Interplay in
Shaping Preferences for Market Justice in Chile (2016-2023)}

  \author{Juan Carlos Castillo}
            \affil{%
                  Departamento de Sociología, Universidad de Chile
              }
          \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
          \affil{%
                  Núcleo milenio de desigualdades y oportunidades
                  digitales (NUDOS)
              }
        \author{Andreas Laffert}
            \affil{%
                  Instituto de Sociología, Pontificia Universidad
                  Católica de Chile
              }
        \author{Kevin Carrasco}
            \affil{%
                  Centro de estudios del conflicto y cohesión social
                  (COES)
              }
        \author{Julio Iturra}
            \affil{%
                  International Graduate School of Social Sciencies
                  (BIGSSS), University of Bremen, Germany
              }

\date{}
\begin{document}
\maketitle
\begin{abstract}
My abstract. \newline \textbf{Keywords}: meritocracy, social inequality,
inequality justification, COVID-19
\end{abstract}

\setstretch{1.15}
\section{Introduction}\label{introduction}

\section{Data, Variables and Methods}\label{data-variables-and-methods}

\subsection{Data}\label{data}

\subsection{Variables}\label{variables}

\subsection{Methods}\label{methods}

\section{Results}\label{results}

\section{Discussion}\label{discussion}

\section{Conclusion}\label{conclusion}

\section{Supplementary material}\label{supplementary-material}

\end{document}

### 2025-04-11 processing/02-analysis.qmd
Update analysis and repo order

@@
 
+# Generate analytical sample
+
+df_study1 <- df_study1_long_t7 %>%
+  select(idencuesta,
+         ola,
+         ponderador_long_total,
+         segmento,
+         estrato,
+         just_health,
+         just_pension,
+         just_educ,
+         mjp, 
+         perc_inequality, 
+         just_inequality,
+         merit_effort, 
+         merit_talent, 
+         educ,
+         quintil1,
+         sex,
+         age,
+         ess, 
+         ideo) %>% 
+  na.omit() %>% 
+  mutate(ola = as.factor(ola),
+         ola_num = as.numeric(ola),
+         ola_2=as.numeric(ola)^2)
+
+df_study1 <- df_study1 %>%
+  group_by(idencuesta) %>%             # Agrupar por el identificador del participante
+  mutate(n_participaciones = n()) %>%  # Contar el número de filas (participaciones) por participante
+  ungroup()
+
+df_study1 <- df_study1 %>% filter(n_participaciones>1)
+
+# Corregir etiquetas
+
+df_study1$just_health <- sjlabelled::set_label(df_study1$just_health, 
+                        label = "Health distributive justice")
+
+df_study1$just_pension <- sjlabelled::set_label(df_study1$just_pension, 
+                        label = "Pension distributive justice")
+
+df_study1$just_educ <- sjlabelled::set_label(df_study1$just_educ, 
+                        label = "Education distributive justice")
+
+df_study1$merit_effort <- sjlabelled::set_label(df_study1$merit_effort, 
+                        label = "People are rewarded for their efforts")
+
+df_study1$merit_talent <- sjlabelled::set_label(df_study1$merit_talent, 
+                        label = "People are rewarded for their intelligence")
+
 ```
@@
 ```{r}
-#| label: tbl-summary
-#| tbl-cap: "Estadísticos descriptivos"
+#| label: tbl-summary1
+#| tbl-cap: "Estadísticos descriptivos items justicia de mercado"
 #| tbl-cap-location: top
-t1 <- df_study1_long_t7 %>% 
-  select(just_health, just_pension, just_educ, mjp, perc_inequality, just_inequality, merit_effort, merit_talent, ess, ideo, educ, sex, age)
+t1 <- df_study1 %>% 
+  filter(ola == 1) %>% 
+  select(just_health, just_pension, just_educ, mjp)
 
 print(summarytools::dfSummary(t1), method="render")
+```
+
+```{r}
+#| label: tbl-summary2
+#| tbl-cap: "Estadísticos descriptivos variables independientes"
+#| tbl-cap-location: top
+t2 <- df_study1 %>% 
+  filter(ola == 1) %>% 
+  select(perc_inequality, just_inequality, merit_effort, merit_talent)
+
+print(summarytools::dfSummary(t2), method="render")
 ```
@@
 
-datos.health <- df_study1_long_t7 %>% 
+datos.health <- df_study1 %>% 
+  mutate(just_health = factor(just_health, 
+                              levels = c("Strongly agree",
+                                         "Agree",
+                                         "Neither agree nor desagree",
+                                         "Desagree",
+                                         "Strongly desagree"))) %>% 
   group_by(idencuesta, ola) %>% 
@@
 
-etiquetas.health <- df_study1_long_t7 %>%
+etiquetas.health <- df_study1 %>%
+  mutate(just_health = factor(just_health, 
+                              levels = c("Strongly agree",
+                                         "Agree",
+                                         "Neither agree nor desagree",
+                                         "Desagree",
+                                         "Strongly desagree"))) %>% 
   group_by(ola, just_health) %>%
@@
   scale_y_continuous(labels = scales::percent) + 
-  scale_fill_manual(values =  c("#CA0020","#F4A582","#b3b3b3ff","#92C5DE","#0571B0")) +
+  scale_fill_manual(values =  c("#0571B0","#92C5DE","#b3b3b3ff","#F4A582","#CA0020")) +
   geom_shadowtext(data = etiquetas.health,
@@
 
-datos.pension <- df_study1_long_t7 %>% 
+datos.pension <- df_study1 %>% 
+   mutate(just_pension = factor(just_pension, 
+                              levels = c("Strongly agree",
+                                         "Agree",
+                                         "Neither agree nor desagree",
+                                         "Desagree",
+                                         "Strongly desagree"))) %>% 
   group_by(idencuesta, ola) %>% 
@@
 
-etiquetas.pension <- df_study1_long_t7 %>%
+etiquetas.pension <- df_study1 %>%
+  mutate(just_pension = factor(just_pension, 
+                              levels = c("Strongly agree",
+                                         "Agree",
+                                         "Neither agree nor desagree",
+                                         "Desagree",
+                                         "Strongly desagree"))) %>% 
   group_by(ola, just_pension) %>%
@@
   scale_y_continuous(labels = scales::percent) + 
-  scale_fill_manual(values =  c("#CA0020","#F4A582","#b3b3b3ff","#92C5DE","#0571B0")) +
+  scale_fill_manual(values =  c("#0571B0","#92C5DE","#b3b3b3ff","#F4A582","#CA0020")) +
   geom_shadowtext(data = etiquetas.pension,
@@
 
-datos.educ <- df_study1_long_t7 %>% 
+datos.educ <- df_study1 %>% 
+  mutate(just_educ = factor(just_educ, 
+                              levels = c("Strongly agree",
+                                         "Agree",
+                                         "Neither agree nor desagree",
+                                         "Desagree",
+                                         "Strongly desagree"))) %>%
   group_by(idencuesta, ola) %>% 
@@
 
-etiquetas.educ <- df_study1_long_t7 %>%
+etiquetas.educ <- df_study1 %>%
+    mutate(just_educ = factor(just_educ, 
+                              levels = c("Strongly agree",
+                                         "Agree",
+                                         "Neither agree nor desagree",
+                                         "Desagree",
+                                         "Strongly desagree"))) %>%
   group_by(ola, just_educ) %>%
@@
   scale_y_continuous(labels = scales::percent) + 
-  scale_fill_manual(values =  c("#CA0020","#F4A582","#b3b3b3ff","#92C5DE","#0571B0")) +
+  scale_fill_manual(values =  c("#0571B0","#92C5DE","#b3b3b3ff","#F4A582","#CA0020")) +
   geom_shadowtext(data = etiquetas.educ,
@@
        fill = NULL,
-       title = "c. Education") +
+       title = "c. Education",
+       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)") +
   theme_ggdist() +
@@
 
+```{r}
+library(srvyr)
+
+df_pond <- df_study1 %>% 
+  mutate(ola = case_when(ola == 1 ~ "2016",
+                         ola == 2 ~ "2017",
+                         ola == 3 ~ "2018",
+                         ola == 4 ~ "2019",
+                         ola == 6 ~ "2022",
+                         ola == 7 ~ "2023"),
+         ola = factor(ola, levels = c("2016",
+                                        "2017",
+                                        "2018",
+                                        "2019",
+                                        "2022",
+                                        "2023"))) %>% 
+  as_survey_design(.data = .,
+                   ids = segmento, 
+                   strata = estrato, 
+                   weights = ponderador_long_total)
+
+df_pond %>% 
+  select(ola, mjp, perc_inequality, just_inequality, merit_effort, merit_talent) %>% 
+  mutate_at(.vars = 2:6, .funs = ~ as.numeric(.)) %>% 
+  group_by(ola) %>% 
+  summarise_all(~survey_mean(., vartype = "ci")) %>% 
+  pivot_longer(cols = -ola,
+               names_to = "temp",
+               values_to = "valor") %>%
+  mutate(
+    ci = case_when(
+      str_ends(temp, "_low") ~ "ic_low",
+      str_ends(temp, "_upp") ~ "ic_upp",
+      TRUE                   ~ "mean"
+    ),
+    variable = str_remove(temp, "_low|_upp")
+  ) %>%
+  select(ola, variable, ci, valor) %>%
+  pivot_wider(
+    names_from  = ci,
+    values_from = valor
+  ) %>% 
+  mutate(variable = case_when(variable == "mjp" ~ "Market justice preferences",
+                              variable == "perc_inequality"  ~ "Inequality perception",
+                              variable == "just_inequality"  ~ "Inequality justification",
+                              variable == "merit_effort"     ~ "Merit: Effort",
+                              variable == "merit_talent"     ~ "Merit: Talent"),
+         variable = factor(variable, levels = c("Market justice preferences",
+                                      "Inequality perception",
+                                      "Inequality justification",
+                                      "Merit: Effort",
+                                      "Merit: Talent")
+                              )) %>% 
+  ggplot(aes(x = ola, y = mean, group = variable)) +
+  geom_point(aes(shape=variable, color=variable), size = 3.5) +
+  geom_line(aes(color = variable), linewidth = 0.8) +
+  geom_errorbar(aes(ymin = ic_low, ymax = ic_upp, color = variable),
+                width = 0.1) +
+  scale_color_manual(values = c("#023858", # azul oscuro
+                                "#0571B0", # azul intermedio
+                                "#b3b3b3", # gris
+                                "#F4A582", # salmón
+                                "#B2182B"  # rojo profundo
+                                         )) +
+  scale_shape_manual(values=c(15, 16, 17, 18, 4)) +
+  labs(y = "Mean",
+       x = NULL,
+       color = NULL,
+       shape = NULL,
+       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)") +
+  theme_ggdist() +
+  theme(legend.position = "top",
+        text = element_text(size = 12))
+
+```
+
 ## Correlations
@@
 ```{r}
-df_study1_long_t7 %>%
+M <- df_study1 %>%
   filter(ola == 7) %>% 
@@
   mutate_all(.funs = ~ as.numeric(.)) %>% 
-  sjPlot::tab_corr(., triangle = "lower")
+  cor(use = "pairwise.complete.obs", method = "pearson")
+
+diag(M) <- NA
+
+rownames(M) <- c("A. Marjet justice preferences",
+                 "B. Inequality gap perception",
+                 "C. Inequality gap justification",
+                 "D. People are rewarded for their efforts ",
+                 "E. People are rewarded for their intelligence")
+
+#set Column names of the matrix
+colnames(M) <-c("(A)", "(B)","(C)","(D)","(E)")
+
+#Plot the matrix using corrplot
+corrplot::corrplot(M,
+                   method = "color",
+                   addCoef.col = "black",
+                   type = "upper",
+                   tl.col = "black",
+                   col = colorRampPalette(c("#E16462", "white", "#0D0887"))(12),
+                   bg = "white",
+                   na.label = "-")
+
 ```
@@
 ```{r}
-# Generate analytical sample
-
-df_study1 <- df_study1_long_t7 %>%
-  select(idencuesta,
-         ola,
-         ponderador_long_total,
-         mjp, 
-         perc_inequality, 
-         just_inequality,
-         merit_effort, 
-         merit_talent, 
-         educ,
-         quintil1,
-         sex,
-         age,
-         ess, 
-         ideo) %>% 
-  na.omit() %>% 
-  mutate(ola = as.factor(ola),
-         ola_num = as.numeric(ola),
-         ola_2=as.numeric(ola)^2)
-
-df_study1 <- df_study1 %>%
-  group_by(idencuesta) %>%             # Agrupar por el identificador del participante
-  mutate(n_participaciones = n()) %>%  # Contar el número de filas (participaciones) por participante
-  ungroup()
-
-df_study1 <- df_study1 %>% filter(n_participaciones>1)
 
@@
 
-m9 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
+m9 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality | idencuesta),
                 data = df_study1)
 
-m10 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + just_inequality*ola_num + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
+m10 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality*ola_num + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + just_inequality| idencuesta),
                 data = df_study1)
 
-m11 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + just_inequality*ola_num + merit_effort*ola_num + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
+m11 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort*ola_num + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort | idencuesta),
                 data = df_study1)
 
-m12 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + just_inequality*ola_num + merit_effort*ola_num + merit_talent*ola_num + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
+m12 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent*ola_num + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_talent | idencuesta),
                 data = df_study1)

### 2025-04-11 processing/02-analysis.tex
Update analysis and repo order

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{3}
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{241,243,245}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
\usepackage[bidi=basic]{babel}
\else
\usepackage[bidi=default]{babel}
\fi
\babelprovide[main,import]{english}
% get rid of language-specific shorthands (see #6817):
\let\LanguageShortHands\languageshorthands
\def\languageshorthands#1{}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Data analysis},
  pdfauthor={Researcher},
  pdflang={en},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Data analysis}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{Perceptions of Inequality and Meritocracy: Their Interplay in
Shaping Preferences for Market Justice in Chile (2016-2023)}
\author{Researcher}
\date{2025-04-07}

\begin{document}
\maketitle

\section{Presentation}\label{presentation}

This is the analysis code for the paper ``Perceptions of Inequality and
Meritocracy: Their Interplay in Shaping Preferences for Market Justice
in Chile (2016-2023)''. The dataset used is
\texttt{df\_study1\_long\_t7.RData}.

\section{Libraries}\label{libraries}

\begin{Shaded}
\begin{Highlighting}[]
\ControlFlowTok{if}\NormalTok{ (}\SpecialCharTok{!} \FunctionTok{require}\NormalTok{(}\StringTok{"pacman"}\NormalTok{)) }\FunctionTok{install.packages}\NormalTok{(}\StringTok{"pacman"}\NormalTok{)}

\NormalTok{pacman}\SpecialCharTok{::}\FunctionTok{p\_load}\NormalTok{(tidyverse, }
\NormalTok{               sjmisc, }
\NormalTok{               sjPlot, }
\NormalTok{               lme4, }
\NormalTok{               here, }
\NormalTok{               performance,}
\NormalTok{               influence.ME, }
\NormalTok{               marginaleffects,}
\NormalTok{               MLMusingR,}
\NormalTok{               texreg, }
\NormalTok{               ggdist,}
\NormalTok{               misty,}
\NormalTok{               kableExtra,}
\NormalTok{               ggalluvial, }
\NormalTok{               shadowtext,}
\NormalTok{               MetBrewer,}
\NormalTok{               patchwork,}
\NormalTok{               sjlabelled)}

\FunctionTok{options}\NormalTok{(}\AttributeTok{scipen=}\DecValTok{999}\NormalTok{)}
\FunctionTok{rm}\NormalTok{(}\AttributeTok{list =} \FunctionTok{ls}\NormalTok{())}
\end{Highlighting}
\end{Shaded}

\section{Data}\label{data}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{load}\NormalTok{(}\AttributeTok{file =} \FunctionTok{here}\NormalTok{(}\StringTok{"input/data/proc/df\_study1\_long\_t7.RData"}\NormalTok{))}

\FunctionTok{glimpse}\NormalTok{(df\_study1\_long\_t7)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
Rows: 10,422
Columns: 29
$ idencuesta            <dbl> 1101011, 1101012, 1101021, 1101023, 1101032, 110~
$ muestra               <dbl> 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, ~
$ ola                   <fct> 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, ~
$ comuna                <chr> "Iquique", "Iquique", "Iquique", "Iquique", "Iqu~
$ comunacod             <dbl> 1101, 1101, 1101, 1101, 1101, 1101, 1101, 1101, ~
$ ponderador_long_total <dbl> 0.11821742, 0.11821742, 0.05633656, 0.07703080, ~
$ segmento              <dbl> 110101, 110101, 110102, 110102, 110103, 110103, ~
$ estrato               <dbl> 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, 4, ~
$ just_educ             <fct> Strongly desagree, Desagree, Agree, Strongly des~
$ just_pension          <fct> Strongly desagree, Desagree, Neither agree nor d~
$ just_health           <fct> Desagree, Desagree, Agree, Strongly desagree, St~
$ mjp                   <dbl> 1.333333, 2.000000, 3.666667, 1.333333, 1.000000~
$ merit_effort          <fct> Agree, Agree, Agree, Neither agree nor desagree,~
$ merit_talent          <fct> Agree, Agree, Agree, Strongly agree, Strongly de~
$ perc_sal_gerente      <dbl> 2000000, 500000, 20000000, 6000000, 4000000, 500~
$ perc_sal_obrero       <dbl> 300000, 300000, 400000, 280000, 300000, 400000, ~
$ just_sal_gerente      <dbl> 2000000, 3000000, 200000000, 1800000, 300000, 50~
$ just_sal_obrero       <dbl> 500000, 500000, 700000, 600000, 600000, 500000, ~
$ perc_inequality       <dbl> 1.8971200, 0.5108256, 3.9120230, 3.0647251, 2.59~
$ just_inequality       <dbl> 1.3862944, 1.7917595, 5.6549923, 1.0986123, -0.6~
$ educ                  <fct> Less than Universitary, Less than Universitary, ~
$ educyear              <dbl> 4.30, 9.80, 14.90, 9.80, 12.02, 13.90, 12.02, 7.~
$ sex                   <fct> Female, Female, Male, Male, Female, Female, Male~
$ age                   <fct> 50-64, 50-64, 50-64, 50-64, 50-64, 30-49, 30-49,~
$ ess                   <dbl> 5, 5, 6, 5, 6, 7, 4, 0, 4, 5, 4, 3, 6, 3, 5, 8, ~
$ ideo                  <fct> Does not identify, Does not identify, Does not i~
$ quintil               <fct> Q1, Q4, Q5, Q5, Q4, Q4, Q2, Q2, Q2, Q5, Q4, Q2, ~
$ quintil1              <fct> Q1, Q4, Q5, Q5, Q4, Q4, Q2, Q2, Q2, Q5, Q4, Q2, ~
$ ing_pc                <dbl> 75000.0, 250000.0, 765000.0, 490000.0, 250000.0,~
\end{verbatim}

\section{Analysis}\label{analysis}

\subsection{Descriptives}\label{descriptives}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{t1 }\OtherTok{\textless{}{-}}\NormalTok{ df\_study1\_long\_t7 }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{select}\NormalTok{(just\_health, just\_pension, just\_educ, mjp, perc\_inequality, just\_inequality, merit\_effort, merit\_talent, ess, ideo, educ, sex, age) }

\FunctionTok{print}\NormalTok{(summarytools}\SpecialCharTok{::}\FunctionTok{dfSummary}\NormalTok{(t1), }\AttributeTok{method=}\StringTok{"render"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[]{@{}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}
  >{\centering\arraybackslash}p{(\columnwidth - 14\tabcolsep) * \real{0.1250}}@{}}

\caption{\label{tbl-summary}Estadísticos descriptivos}

\tabularnewline

\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\centering
\textbf{No}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Variable}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Label}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Stats / Values}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Freqs (\% of Valid)}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Graph}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Valid}
\end{minipage} & \begin{minipage}[b]{\linewidth}\centering
\textbf{Missing}
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1 & just\_health {[}factor{]} & &
\begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable*}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1. Strongly desagree \\
2. Desagree \\
3. Neither agree nor desagre \\
4. Agree \\
5. Strongly agree \\
\end{longtable}
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}rlrl@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
3620 & ( & 36.9\% & ) \\
4746 & ( & 48.4\% & ) \\
513 & ( & 5.2\% & ) \\
802 & ( & 8.2\% & ) \\
131 & ( & 1.3\% & ) \\
\end{longtable}
\end{minipage} & & 9812 (94.1\%) & 610 (5.9\%) \\
2 & just\_pension {[}factor{]} & &
\begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1. Strongly desagree \\
2. Desagree \\
3. Neither agree nor desagre \\
4. Agree \\
5. Strongly agree \\
\end{longtable}
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}rlrl@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
2588 & ( & 26.4\% & ) \\
4289 & ( & 43.8\% & ) \\
918 & ( & 9.4\% & ) \\
1759 & ( & 18.0\% & ) \\
244 & ( & 2.5\% & ) \\
\end{longtable}
\end{minipage} & & 9798 (94.0\%) & 624 (6.0\%) \\
3 & just\_educ {[}factor{]} & &
\begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1. Strongly desagree \\
2. Desagree \\
3. Neither agree nor desagre \\
4. Agree \\
5. Strongly agree \\
\end{longtable}
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}rlrl@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
3341 & ( & 34.1\% & ) \\
4901 & ( & 50.0\% & ) \\
574 & ( & 5.9\% & ) \\
863 & ( & 8.8\% & ) \\
128 & ( & 1.3\% & ) \\
\end{longtable}
\end{minipage} & & 9807 (94.1\%) & 615 (5.9\%) \\
4 & mjp {[}numeric{]} & Market justice preferences &
\begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mean (sd) : 2 (0.9) \\
min ≤ med ≤ max: \\
1 ≤ 2 ≤ 5 \\
IQR (CV) : 1 (0.4) \\
\end{longtable}
\end{minipage} & 16 distinct values & & 9817 (94.2\%) & 605 (5.8\%) \\
5 & perc\_inequality {[}numeric{]} & Inequality gap perception &
\begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mean (sd) : 3.4 (1.1) \\
min ≤ med ≤ max: \\
-0.5 ≤ 3.4 ≤ 6.9 \\
IQR (CV) : 1.5 (0.3) \\
\end{longtable}
\end{minipage} & 798 distinct values & & 8898 (85.4\%) & 1524
(14.6\%) \\
6 & just\_inequality {[}numeric{]} & Inequality gap justification &
\begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mean (sd) : 1.9 (1.2) \\
min ≤ med ≤ max: \\
-6.6 ≤ 1.8 ≤ 10.2 \\
IQR (CV) : 1.5 (0.6) \\
\end{longtable}
\end{minipage} & 535 distinct values & & 9190 (88.2\%) & 1232
(11.8\%) \\
7 & merit\_effort {[}factor{]} & &
\begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1. Strongly desagree \\
2. Desagree \\
3. Neither agree nor desagre \\
4. Agree \\
5. Strongly agree \\
\end{longtable}
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}rlrl@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1064 & ( & 10.9\% & ) \\
4448 & ( & 45.5\% & ) \\
1928 & ( & 19.7\% & ) \\
2077 & ( & 21.2\% & ) \\
262 & ( & 2.7\% & ) \\
\end{longtable}
\end{minipage} & & 9779 (93.8\%) & 643 (6.2\%) \\
8 & merit\_talent {[}factor{]} & &
\begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1. Strongly desagree \\
2. Desagree \\
3. Neither agree nor desagre \\
4. Agree \\
5. Strongly agree \\
\end{longtable}
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}rlrl@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
852 & ( & 8.7\% & ) \\
3838 & ( & 39.2\% & ) \\
2180 & ( & 22.3\% & ) \\
2618 & ( & 26.8\% & ) \\
292 & ( & 3.0\% & ) \\
\end{longtable}
\end{minipage} & & 9780 (93.8\%) & 642 (6.2\%) \\
9 & ess {[}numeric{]} & Subjective Social Status &
\begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mean (sd) : 4.3 (1.5) \\
min ≤ med ≤ max: \\
0 ≤ 4 ≤ 10 \\
IQR (CV) : 2 (0.4) \\
\end{longtable}
\end{minipage} & 11 distinct values & & 10398 (99.8\%) & 24 (0.2\%) \\
10 & ideo {[}factor{]} & & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1. Left \\
2. Center \\
3. Right \\
4. Does not identify \\
\end{longtable}
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}rlrl@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
2136 & ( & 20.9\% & ) \\
2130 & ( & 20.8\% & ) \\
1350 & ( & 13.2\% & ) \\
4620 & ( & 45.1\% & ) \\
\end{longtable}
\end{minipage} & & 10236 (98.2\%) & 186 (1.8\%) \\
11 & educ {[}factor{]} & & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1. Less than Universitary \\
2. Universitary \\
\end{longtable}
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}rlrl@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
8652 & ( & 83.1\% & ) \\
1764 & ( & 16.9\% & ) \\
\end{longtable}
\end{minipage} & & 10416 (99.9\%) & 6 (0.1\%) \\
12 & sex {[}factor{]} & & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1. Male \\
2. Female \\
\end{longtable}
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}rlrl@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
3708 & ( & 35.6\% & ) \\
6714 & ( & 64.4\% & ) \\
\end{longtable}
\end{minipage} & & 10422 (100.0\%) & 0 (0.0\%) \\
13 & age {[}factor{]} & & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}l@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1. 18-29 \\
2. 30-49 \\
3. 50-64 \\
4. 65 or more \\
\end{longtable}
\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
\begin{longtable}[]{@{}rlrl@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
1524 & ( & 14.6\% & ) \\
4056 & ( & 38.9\% & ) \\
3348 & ( & 32.1\% & ) \\
1494 & ( & 14.3\% & ) \\
\end{longtable*}
\end{minipage} & & 10422 (100.0\%) & 0 (0.0\%) \\

\end{longtable}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{datos.health }\OtherTok{\textless{}{-}}\NormalTok{ df\_study1\_long\_t7 }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{group\_by}\NormalTok{(idencuesta, ola) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{count}\NormalTok{(just\_health) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{group\_by}\NormalTok{(ola) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{porcentaje=}\NormalTok{n}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(n)) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{ungroup}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{na.omit}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{wave =} \FunctionTok{case\_when}\NormalTok{(ola }\SpecialCharTok{==} \DecValTok{1} \SpecialCharTok{\textasciitilde{}} \StringTok{"2016"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{2} \SpecialCharTok{\textasciitilde{}} \StringTok{"2017"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{3} \SpecialCharTok{\textasciitilde{}} \StringTok{"2018"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{4} \SpecialCharTok{\textasciitilde{}} \StringTok{"2019"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{6} \SpecialCharTok{\textasciitilde{}} \StringTok{"2022"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{7} \SpecialCharTok{\textasciitilde{}} \StringTok{"2023"}\NormalTok{),}
         \AttributeTok{wave =} \FunctionTok{factor}\NormalTok{(wave, }\AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"2016"}\NormalTok{,}
                                      \StringTok{"2017"}\NormalTok{,}
                                      \StringTok{"2018"}\NormalTok{,}
                                      \StringTok{"2019"}\NormalTok{,}
                                      \StringTok{"2022"}\NormalTok{,}
                                      \StringTok{"2023"}\NormalTok{)))}

\NormalTok{etiquetas.health }\OtherTok{\textless{}{-}}\NormalTok{ df\_study1\_long\_t7 }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(ola, just\_health) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{summarise}\NormalTok{(}\AttributeTok{count =} \FunctionTok{n}\NormalTok{(), }\AttributeTok{.groups =} \StringTok{"drop"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(ola) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{porcentaje =}\NormalTok{ count }\SpecialCharTok{/} \FunctionTok{sum}\NormalTok{(count)) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{na.omit}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{idencuesta =} \DecValTok{1}\NormalTok{,}
         \AttributeTok{wave =} \FunctionTok{case\_when}\NormalTok{(ola }\SpecialCharTok{==} \DecValTok{1} \SpecialCharTok{\textasciitilde{}} \StringTok{"2016"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{2} \SpecialCharTok{\textasciitilde{}} \StringTok{"2017"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{3} \SpecialCharTok{\textasciitilde{}} \StringTok{"2018"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{4} \SpecialCharTok{\textasciitilde{}} \StringTok{"2019"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{6} \SpecialCharTok{\textasciitilde{}} \StringTok{"2022"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{7} \SpecialCharTok{\textasciitilde{}} \StringTok{"2023"}\NormalTok{),}
         \AttributeTok{wave =} \FunctionTok{factor}\NormalTok{(wave, }\AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"2016"}\NormalTok{,}
                                        \StringTok{"2017"}\NormalTok{,}
                                        \StringTok{"2018"}\NormalTok{,}
                                        \StringTok{"2019"}\NormalTok{,}
                                        \StringTok{"2022"}\NormalTok{,}
                                        \StringTok{"2023"}\NormalTok{)))}

\NormalTok{p1 }\OtherTok{\textless{}{-}}\NormalTok{ datos.health }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ wave, }\AttributeTok{fill =}\NormalTok{ just\_health, }\AttributeTok{stratum =}\NormalTok{ just\_health,}
             \AttributeTok{alluvium =}\NormalTok{ idencuesta, }\AttributeTok{y =}\NormalTok{ porcentaje)) }\SpecialCharTok{+}
\NormalTok{  ggalluvial}\SpecialCharTok{::}\FunctionTok{geom\_flow}\NormalTok{(}\AttributeTok{alpha =}\NormalTok{ .}\DecValTok{4}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{  ggalluvial}\SpecialCharTok{::}\FunctionTok{geom\_stratum}\NormalTok{(}\AttributeTok{linetype =} \DecValTok{0}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{scale\_y\_continuous}\NormalTok{(}\AttributeTok{labels =}\NormalTok{ scales}\SpecialCharTok{::}\NormalTok{percent) }\SpecialCharTok{+} 
  \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =}  \FunctionTok{c}\NormalTok{(}\StringTok{"\#CA0020"}\NormalTok{,}\StringTok{"\#F4A582"}\NormalTok{,}\StringTok{"\#b3b3b3ff"}\NormalTok{,}\StringTok{"\#92C5DE"}\NormalTok{,}\StringTok{"\#0571B0"}\NormalTok{)) }\SpecialCharTok{+}
  \FunctionTok{geom\_shadowtext}\NormalTok{(}\AttributeTok{data =}\NormalTok{ etiquetas.health,}
                  \FunctionTok{aes}\NormalTok{(}\AttributeTok{label =} \FunctionTok{ifelse}\NormalTok{(porcentaje }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{ , scales}\SpecialCharTok{::}\FunctionTok{percent}\NormalTok{(porcentaje, }\AttributeTok{accuracy =}\NormalTok{ .}\DecValTok{1}\NormalTok{),}\StringTok{""}\NormalTok{)),}
                  \AttributeTok{position =} \FunctionTok{position\_stack}\NormalTok{(}\AttributeTok{vjust =}\NormalTok{ .}\DecValTok{5}\NormalTok{),}
                  \AttributeTok{show.legend =} \ConstantTok{FALSE}\NormalTok{,}
                  \AttributeTok{size =} \DecValTok{3}\NormalTok{,}
                  \AttributeTok{color =} \FunctionTok{rep}\NormalTok{(}\StringTok{\textquotesingle{}white\textquotesingle{}}\NormalTok{),}
                  \AttributeTok{bg.colour=}\StringTok{\textquotesingle{}grey30\textquotesingle{}}\NormalTok{)}\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}\AttributeTok{y =} \StringTok{"\%"}\NormalTok{,}
       \AttributeTok{x =} \ConstantTok{NULL}\NormalTok{,}
       \AttributeTok{fill =} \ConstantTok{NULL}\NormalTok{,}
       \AttributeTok{title =} \StringTok{"a. Health"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{theme\_ggdist}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"none"}\NormalTok{) }

\NormalTok{datos.pension }\OtherTok{\textless{}{-}}\NormalTok{ df\_study1\_long\_t7 }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{group\_by}\NormalTok{(idencuesta, ola) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{count}\NormalTok{(just\_pension) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{group\_by}\NormalTok{(ola) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{porcentaje=}\NormalTok{n}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(n)) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{ungroup}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{na.omit}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{wave =} \FunctionTok{case\_when}\NormalTok{(ola }\SpecialCharTok{==} \DecValTok{1} \SpecialCharTok{\textasciitilde{}} \StringTok{"2016"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{2} \SpecialCharTok{\textasciitilde{}} \StringTok{"2017"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{3} \SpecialCharTok{\textasciitilde{}} \StringTok{"2018"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{4} \SpecialCharTok{\textasciitilde{}} \StringTok{"2019"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{6} \SpecialCharTok{\textasciitilde{}} \StringTok{"2022"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{7} \SpecialCharTok{\textasciitilde{}} \StringTok{"2023"}\NormalTok{),}
         \AttributeTok{wave =} \FunctionTok{factor}\NormalTok{(wave, }\AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"2016"}\NormalTok{,}
                                        \StringTok{"2017"}\NormalTok{,}
                                        \StringTok{"2018"}\NormalTok{,}
                                        \StringTok{"2019"}\NormalTok{,}
                                        \StringTok{"2022"}\NormalTok{,}
                                        \StringTok{"2023"}\NormalTok{)))}

\NormalTok{etiquetas.pension }\OtherTok{\textless{}{-}}\NormalTok{ df\_study1\_long\_t7 }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(ola, just\_pension) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{summarise}\NormalTok{(}\AttributeTok{count =} \FunctionTok{n}\NormalTok{(), }\AttributeTok{.groups =} \StringTok{"drop"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(ola) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{porcentaje =}\NormalTok{ count }\SpecialCharTok{/} \FunctionTok{sum}\NormalTok{(count)) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{na.omit}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{idencuesta =} \DecValTok{1}\NormalTok{,}
         \AttributeTok{wave =} \FunctionTok{case\_when}\NormalTok{(ola }\SpecialCharTok{==} \DecValTok{1} \SpecialCharTok{\textasciitilde{}} \StringTok{"2016"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{2} \SpecialCharTok{\textasciitilde{}} \StringTok{"2017"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{3} \SpecialCharTok{\textasciitilde{}} \StringTok{"2018"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{4} \SpecialCharTok{\textasciitilde{}} \StringTok{"2019"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{6} \SpecialCharTok{\textasciitilde{}} \StringTok{"2022"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{7} \SpecialCharTok{\textasciitilde{}} \StringTok{"2023"}\NormalTok{),}
         \AttributeTok{wave =} \FunctionTok{factor}\NormalTok{(wave, }\AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"2016"}\NormalTok{,}
                                        \StringTok{"2017"}\NormalTok{,}
                                        \StringTok{"2018"}\NormalTok{,}
                                        \StringTok{"2019"}\NormalTok{,}
                                        \StringTok{"2022"}\NormalTok{,}
                                        \StringTok{"2023"}\NormalTok{)))}

\NormalTok{p2 }\OtherTok{\textless{}{-}}\NormalTok{ datos.pension }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ wave, }\AttributeTok{fill =}\NormalTok{ just\_pension, }\AttributeTok{stratum =}\NormalTok{ just\_pension,}
             \AttributeTok{alluvium =}\NormalTok{ idencuesta, }\AttributeTok{y =}\NormalTok{ porcentaje)) }\SpecialCharTok{+}
\NormalTok{  ggalluvial}\SpecialCharTok{::}\FunctionTok{geom\_flow}\NormalTok{(}\AttributeTok{alpha =}\NormalTok{ .}\DecValTok{4}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{  ggalluvial}\SpecialCharTok{::}\FunctionTok{geom\_stratum}\NormalTok{(}\AttributeTok{linetype =} \DecValTok{0}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{scale\_y\_continuous}\NormalTok{(}\AttributeTok{labels =}\NormalTok{ scales}\SpecialCharTok{::}\NormalTok{percent) }\SpecialCharTok{+} 
  \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =}  \FunctionTok{c}\NormalTok{(}\StringTok{"\#CA0020"}\NormalTok{,}\StringTok{"\#F4A582"}\NormalTok{,}\StringTok{"\#b3b3b3ff"}\NormalTok{,}\StringTok{"\#92C5DE"}\NormalTok{,}\StringTok{"\#0571B0"}\NormalTok{)) }\SpecialCharTok{+}
  \FunctionTok{geom\_shadowtext}\NormalTok{(}\AttributeTok{data =}\NormalTok{ etiquetas.pension,}
                  \FunctionTok{aes}\NormalTok{(}\AttributeTok{label =} \FunctionTok{ifelse}\NormalTok{(porcentaje }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{ , scales}\SpecialCharTok{::}\FunctionTok{percent}\NormalTok{(porcentaje, }\AttributeTok{accuracy =}\NormalTok{ .}\DecValTok{1}\NormalTok{),}\StringTok{""}\NormalTok{)),}
                  \AttributeTok{position =} \FunctionTok{position\_stack}\NormalTok{(}\AttributeTok{vjust =}\NormalTok{ .}\DecValTok{5}\NormalTok{),}
                  \AttributeTok{show.legend =} \ConstantTok{FALSE}\NormalTok{,}
                  \AttributeTok{size =} \DecValTok{3}\NormalTok{,}
                  \AttributeTok{color =} \FunctionTok{rep}\NormalTok{(}\StringTok{\textquotesingle{}white\textquotesingle{}}\NormalTok{),}
                  \AttributeTok{bg.colour=}\StringTok{\textquotesingle{}grey30\textquotesingle{}}\NormalTok{)}\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}\AttributeTok{y =} \StringTok{"\%"}\NormalTok{,}
       \AttributeTok{x =} \ConstantTok{NULL}\NormalTok{,}
       \AttributeTok{fill =} \ConstantTok{NULL}\NormalTok{,}
       \AttributeTok{title =} \StringTok{"b. Pension"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{theme\_ggdist}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"none"}\NormalTok{) }

\NormalTok{datos.educ }\OtherTok{\textless{}{-}}\NormalTok{ df\_study1\_long\_t7 }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{group\_by}\NormalTok{(idencuesta, ola) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{count}\NormalTok{(just\_educ) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{group\_by}\NormalTok{(ola) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{porcentaje=}\NormalTok{n}\SpecialCharTok{/}\FunctionTok{sum}\NormalTok{(n)) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{ungroup}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{na.omit}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{wave =} \FunctionTok{case\_when}\NormalTok{(ola }\SpecialCharTok{==} \DecValTok{1} \SpecialCharTok{\textasciitilde{}} \StringTok{"2016"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{2} \SpecialCharTok{\textasciitilde{}} \StringTok{"2017"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{3} \SpecialCharTok{\textasciitilde{}} \StringTok{"2018"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{4} \SpecialCharTok{\textasciitilde{}} \StringTok{"2019"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{6} \SpecialCharTok{\textasciitilde{}} \StringTok{"2022"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{7} \SpecialCharTok{\textasciitilde{}} \StringTok{"2023"}\NormalTok{),}
         \AttributeTok{wave =} \FunctionTok{factor}\NormalTok{(wave, }\AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"2016"}\NormalTok{,}
                                        \StringTok{"2017"}\NormalTok{,}
                                        \StringTok{"2018"}\NormalTok{,}
                                        \StringTok{"2019"}\NormalTok{,}
                                        \StringTok{"2022"}\NormalTok{,}
                                        \StringTok{"2023"}\NormalTok{)))}

\NormalTok{etiquetas.educ }\OtherTok{\textless{}{-}}\NormalTok{ df\_study1\_long\_t7 }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(ola, just\_educ) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{summarise}\NormalTok{(}\AttributeTok{count =} \FunctionTok{n}\NormalTok{(), }\AttributeTok{.groups =} \StringTok{"drop"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(ola) }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{porcentaje =}\NormalTok{ count }\SpecialCharTok{/} \FunctionTok{sum}\NormalTok{(count)) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{na.omit}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{idencuesta =} \DecValTok{1}\NormalTok{,}
         \AttributeTok{wave =} \FunctionTok{case\_when}\NormalTok{(ola }\SpecialCharTok{==} \DecValTok{1} \SpecialCharTok{\textasciitilde{}} \StringTok{"2016"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{2} \SpecialCharTok{\textasciitilde{}} \StringTok{"2017"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{3} \SpecialCharTok{\textasciitilde{}} \StringTok{"2018"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{4} \SpecialCharTok{\textasciitilde{}} \StringTok{"2019"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{6} \SpecialCharTok{\textasciitilde{}} \StringTok{"2022"}\NormalTok{,}
\NormalTok{                          ola }\SpecialCharTok{==} \DecValTok{7} \SpecialCharTok{\textasciitilde{}} \StringTok{"2023"}\NormalTok{),}
         \AttributeTok{wave =} \FunctionTok{factor}\NormalTok{(wave, }\AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"2016"}\NormalTok{,}
                                        \StringTok{"2017"}\NormalTok{,}
                                        \StringTok{"2018"}\NormalTok{,}
                                        \StringTok{"2019"}\NormalTok{,}
                                        \StringTok{"2022"}\NormalTok{,}
                                        \StringTok{"2023"}\NormalTok{)))}

\NormalTok{p3 }\OtherTok{\textless{}{-}}\NormalTok{ datos.educ }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ wave, }\AttributeTok{fill =}\NormalTok{ just\_educ, }\AttributeTok{stratum =}\NormalTok{ just\_educ,}
             \AttributeTok{alluvium =}\NormalTok{ idencuesta, }\AttributeTok{y =}\NormalTok{ porcentaje)) }\SpecialCharTok{+}
\NormalTok{  ggalluvial}\SpecialCharTok{::}\FunctionTok{geom\_flow}\NormalTok{(}\AttributeTok{alpha =}\NormalTok{ .}\DecValTok{4}\NormalTok{) }\SpecialCharTok{+} 
\NormalTok{  ggalluvial}\SpecialCharTok{::}\FunctionTok{geom\_stratum}\NormalTok{(}\AttributeTok{linetype =} \DecValTok{0}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{scale\_y\_continuous}\NormalTok{(}\AttributeTok{labels =}\NormalTok{ scales}\SpecialCharTok{::}\NormalTok{percent) }\SpecialCharTok{+} 
  \FunctionTok{scale\_fill\_manual}\NormalTok{(}\AttributeTok{values =}  \FunctionTok{c}\NormalTok{(}\StringTok{"\#CA0020"}\NormalTok{,}\StringTok{"\#F4A582"}\NormalTok{,}\StringTok{"\#b3b3b3ff"}\NormalTok{,}\StringTok{"\#92C5DE"}\NormalTok{,}\StringTok{"\#0571B0"}\NormalTok{)) }\SpecialCharTok{+}
  \FunctionTok{geom\_shadowtext}\NormalTok{(}\AttributeTok{data =}\NormalTok{ etiquetas.educ,}
                  \FunctionTok{aes}\NormalTok{(}\AttributeTok{label =} \FunctionTok{ifelse}\NormalTok{(porcentaje }\SpecialCharTok{\textgreater{}} \DecValTok{0}\NormalTok{ , scales}\SpecialCharTok{::}\FunctionTok{percent}\NormalTok{(porcentaje, }\AttributeTok{accuracy =}\NormalTok{ .}\DecValTok{1}\NormalTok{),}\StringTok{""}\NormalTok{)),}
                  \AttributeTok{position =} \FunctionTok{position\_stack}\NormalTok{(}\AttributeTok{vjust =}\NormalTok{ .}\DecValTok{5}\NormalTok{),}
                  \AttributeTok{show.legend =} \ConstantTok{FALSE}\NormalTok{,}
                  \AttributeTok{size =} \DecValTok{3}\NormalTok{,}
                  \AttributeTok{color =} \FunctionTok{rep}\NormalTok{(}\StringTok{\textquotesingle{}white\textquotesingle{}}\NormalTok{),}
                  \AttributeTok{bg.colour=}\StringTok{\textquotesingle{}grey30\textquotesingle{}}\NormalTok{)}\SpecialCharTok{+}
  \FunctionTok{labs}\NormalTok{(}\AttributeTok{y =} \StringTok{"\%"}\NormalTok{,}
       \AttributeTok{x =} \ConstantTok{NULL}\NormalTok{,}
       \AttributeTok{fill =} \ConstantTok{NULL}\NormalTok{,}
       \AttributeTok{title =} \StringTok{"c. Education"}\NormalTok{) }\SpecialCharTok{+}
  \FunctionTok{theme\_ggdist}\NormalTok{() }\SpecialCharTok{+}
  \FunctionTok{theme}\NormalTok{(}\AttributeTok{legend.position =} \StringTok{"bottom"}\NormalTok{) }

\NormalTok{(p1 }\SpecialCharTok{/}\NormalTok{ p2 }\SpecialCharTok{/}\NormalTok{ p3)}
\end{Highlighting}
\end{Shaded}

\begin{figure}[H]

\caption{\label{fig-alluvial}Change in the justification of inequality
in health, pensions and education over time (2016-2023)}

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{02-analysis_files/figure-pdf/fig-alluvial-1.pdf}

}

\end{figure}%

\subsection{Correlations}\label{correlations}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{df\_study1\_long\_t7 }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{filter}\NormalTok{(ola }\SpecialCharTok{==} \DecValTok{7}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{select}\NormalTok{(mjp, perc\_inequality, just\_inequality, merit\_effort, merit\_talent) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate\_all}\NormalTok{(}\AttributeTok{.funs =} \SpecialCharTok{\textasciitilde{}} \FunctionTok{as.numeric}\NormalTok{(.)) }\SpecialCharTok{\%\textgreater{}\%} 
\NormalTok{  sjPlot}\SpecialCharTok{::}\FunctionTok{tab\_corr}\NormalTok{(., }\AttributeTok{triangle =} \StringTok{"lower"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{longtable}[]{@{}llllll@{}}
\toprule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
~ & mjp & perc\_inequality & just\_inequality & merit\_effort &
merit\_talent \\
mjp & ~ & ~ & ~ & ~ & ~ \\
perc\_inequality & -0.049{} & ~ & ~ & ~ & ~ \\
just\_inequality & 0.162{***} & 0.524{***} & ~ & ~ & ~ \\
merit\_effort & 0.140{***} & -0.041{} & 0.046{} & ~ & ~ \\
merit\_talent & 0.117{***} & -0.039{} & 0.062{*} & 0.688{***} & ~ \\
\multicolumn{6}{@{}r@{}}{%
Computed correlation used pearson-method with listwise-deletion.} \\
\end{longtable}

\subsection{Longitudinal multilevel
models}\label{longitudinal-multilevel-models}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Generate analytical sample}

\NormalTok{df\_study1 }\OtherTok{\textless{}{-}}\NormalTok{ df\_study1\_long\_t7 }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{select}\NormalTok{(idencuesta,}
\NormalTok{         ola,}
\NormalTok{         ponderador\_long\_total,}
\NormalTok{         mjp, }
\NormalTok{         perc\_inequality, }
\NormalTok{         just\_inequality,}
\NormalTok{         merit\_effort, }
\NormalTok{         merit\_talent, }
\NormalTok{         educ,}
\NormalTok{         quintil1,}
\NormalTok{         sex,}
\NormalTok{         age,}
\NormalTok{         ess, }
\NormalTok{         ideo) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{na.omit}\NormalTok{() }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{ola =} \FunctionTok{as.factor}\NormalTok{(ola),}
         \AttributeTok{ola\_num =} \FunctionTok{as.numeric}\NormalTok{(ola),}
         \AttributeTok{ola\_2=}\FunctionTok{as.numeric}\NormalTok{(ola)}\SpecialCharTok{\^{}}\DecValTok{2}\NormalTok{)}

\NormalTok{df\_study1 }\OtherTok{\textless{}{-}}\NormalTok{ df\_study1 }\SpecialCharTok{\%\textgreater{}\%}
  \FunctionTok{group\_by}\NormalTok{(idencuesta) }\SpecialCharTok{\%\textgreater{}\%}             \CommentTok{\# Agrupar por el identificador del participante}
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{n\_participaciones =} \FunctionTok{n}\NormalTok{()) }\SpecialCharTok{\%\textgreater{}\%}  \CommentTok{\# Contar el número de filas (participaciones) por participante}
  \FunctionTok{ungroup}\NormalTok{()}

\NormalTok{df\_study1 }\OtherTok{\textless{}{-}}\NormalTok{ df\_study1 }\SpecialCharTok{\%\textgreater{}\%} \FunctionTok{filter}\NormalTok{(n\_participaciones}\SpecialCharTok{\textgreater{}}\DecValTok{1}\NormalTok{)}

\NormalTok{df\_study1}\SpecialCharTok{$}\NormalTok{merit\_effort }\OtherTok{\textless{}{-}} \FunctionTok{as\_numeric}\NormalTok{(df\_study1}\SpecialCharTok{$}\NormalTok{merit\_effort)}
\NormalTok{df\_study1}\SpecialCharTok{$}\NormalTok{merit\_talent }\OtherTok{\textless{}{-}} \FunctionTok{as\_numeric}\NormalTok{(df\_study1}\SpecialCharTok{$}\NormalTok{merit\_talent)}

\NormalTok{df\_study1 }\OtherTok{\textless{}{-}}\NormalTok{ df\_study1 }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{ola =} \FunctionTok{case\_when}\NormalTok{(ola }\SpecialCharTok{==} \DecValTok{1} \SpecialCharTok{\textasciitilde{}} \StringTok{"2016"}\NormalTok{,}
\NormalTok{                         ola }\SpecialCharTok{==} \DecValTok{2} \SpecialCharTok{\textasciitilde{}} \StringTok{"2017"}\NormalTok{,}
\NormalTok{                         ola }\SpecialCharTok{==} \DecValTok{3} \SpecialCharTok{\textasciitilde{}} \StringTok{"2018"}\NormalTok{,}
\NormalTok{                         ola }\SpecialCharTok{==} \DecValTok{4} \SpecialCharTok{\textasciitilde{}} \StringTok{"2019"}\NormalTok{,}
\NormalTok{                         ola }\SpecialCharTok{==} \DecValTok{6} \SpecialCharTok{\textasciitilde{}} \StringTok{"2022"}\NormalTok{,}
\NormalTok{                         ola }\SpecialCharTok{==} \DecValTok{7} \SpecialCharTok{\textasciitilde{}} \StringTok{"2023"}\NormalTok{),}
         \AttributeTok{ola =} \FunctionTok{factor}\NormalTok{(ola, }\AttributeTok{levels =} \FunctionTok{c}\NormalTok{(}\StringTok{"2016"}\NormalTok{,}
                                        \StringTok{"2017"}\NormalTok{,}
                                        \StringTok{"2018"}\NormalTok{,}
                                        \StringTok{"2019"}\NormalTok{,}
                                        \StringTok{"2022"}\NormalTok{,}
                                        \StringTok{"2023"}\NormalTok{)))}

\NormalTok{df\_study1 }\OtherTok{\textless{}{-}}\NormalTok{ df\_study1 }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{group\_by}\NormalTok{(idencuesta) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{perc\_inequality\_mean =} \FunctionTok{mean}\NormalTok{(perc\_inequality, }\AttributeTok{na.rm =}\NormalTok{ T),}
         \AttributeTok{perc\_inequality\_cwc =}\NormalTok{ perc\_inequality }\SpecialCharTok{{-}}\NormalTok{ perc\_inequality\_mean,}
         \AttributeTok{just\_inequality\_mean =} \FunctionTok{mean}\NormalTok{(just\_inequality, }\AttributeTok{na.rm =}\NormalTok{ T),}
         \AttributeTok{just\_inequality\_cwc =}\NormalTok{ just\_inequality }\SpecialCharTok{{-}}\NormalTok{ just\_inequality\_mean,}
         \AttributeTok{merit\_effort\_mean =} \FunctionTok{mean}\NormalTok{(merit\_effort, }\AttributeTok{na.rm =}\NormalTok{ T),}
         \AttributeTok{merit\_effort\_cwc =}\NormalTok{ merit\_effort }\SpecialCharTok{{-}}\NormalTok{ merit\_effort\_mean,}
         \AttributeTok{merit\_talent\_mean =} \FunctionTok{mean}\NormalTok{(merit\_talent, }\AttributeTok{na.rm =}\NormalTok{ T),}
         \AttributeTok{merit\_talent\_cwc =}\NormalTok{ merit\_talent }\SpecialCharTok{{-}}\NormalTok{ merit\_talent\_mean,}
\NormalTok{         ) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{ungroup}\NormalTok{()}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{m0 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ idencuesta), }
                \AttributeTok{data =}\NormalTok{ df\_study1)}

\NormalTok{performance}\SpecialCharTok{::}\FunctionTok{icc}\NormalTok{(m0, }\AttributeTok{by\_group =}\NormalTok{ T)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
# ICC by Group

Group      |   ICC
------------------
idencuesta | 0.232
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{m1 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ idencuesta),}
                \AttributeTok{data =}\NormalTok{ df\_study1)}

\NormalTok{m1}\FloatTok{.1} \OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{+}\NormalTok{ ola\_2 }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ idencuesta),}
                \AttributeTok{data =}\NormalTok{ df\_study1)}

\NormalTok{m2 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{+}\NormalTok{ ola\_2 }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{|}\NormalTok{ idencuesta),}
                \AttributeTok{data =}\NormalTok{ df\_study1)}

\FunctionTok{anova}\NormalTok{(m1}\FloatTok{.1}\NormalTok{, m2)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
Data: df_study1
Models:
m1.1: mjp ~ 1 + ola_num + ola_2 + (1 | idencuesta)
m2: mjp ~ 1 + ola_num + ola_2 + (1 + ola_num | idencuesta)
     npar   AIC   BIC logLik deviance  Chisq Df           Pr(>Chisq)    
m1.1    5 20352 20387 -10171    20342                                   
m2      7 20288 20338 -10137    20274 67.966  2 0.000000000000001743 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{m3 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{+}\NormalTok{ ola\_2 }\SpecialCharTok{+}\NormalTok{ perc\_inequality }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{|}\NormalTok{ idencuesta), }
           \AttributeTok{data =}\NormalTok{ df\_study1)}

\NormalTok{m4 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{+}\NormalTok{ ola\_2 }\SpecialCharTok{+}\NormalTok{ perc\_inequality }\SpecialCharTok{+}\NormalTok{ just\_inequality }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{|}\NormalTok{ idencuesta),}
                \AttributeTok{data =}\NormalTok{ df\_study1)}

\NormalTok{m5 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{+}\NormalTok{ ola\_2 }\SpecialCharTok{+}\NormalTok{ perc\_inequality }\SpecialCharTok{+}\NormalTok{ just\_inequality }\SpecialCharTok{+}\NormalTok{ merit\_effort }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{|}\NormalTok{ idencuesta),}
                \AttributeTok{data =}\NormalTok{ df\_study1)}

\NormalTok{m6 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{+}\NormalTok{ ola\_2 }\SpecialCharTok{+}\NormalTok{ perc\_inequality }\SpecialCharTok{+}\NormalTok{ just\_inequality }\SpecialCharTok{+}\NormalTok{ merit\_effort }\SpecialCharTok{+}\NormalTok{ merit\_talent }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{|}\NormalTok{ idencuesta),}
                \AttributeTok{data =}\NormalTok{ df\_study1)}

\NormalTok{m7 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{+}\NormalTok{ ola\_2 }\SpecialCharTok{+}\NormalTok{ perc\_inequality }\SpecialCharTok{+}\NormalTok{ just\_inequality }\SpecialCharTok{+}\NormalTok{ merit\_effort }\SpecialCharTok{+}\NormalTok{ merit\_talent }\SpecialCharTok{+}\NormalTok{ perc\_inequality\_mean }\SpecialCharTok{+}\NormalTok{ just\_inequality\_mean }\SpecialCharTok{+}\NormalTok{ merit\_effort\_mean }\SpecialCharTok{+}\NormalTok{ merit\_talent\_mean }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{|}\NormalTok{ idencuesta),}
                \AttributeTok{data =}\NormalTok{ df\_study1)}

\NormalTok{m8 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{+}\NormalTok{ ola\_2 }\SpecialCharTok{+}\NormalTok{ perc\_inequality }\SpecialCharTok{+}\NormalTok{ just\_inequality }\SpecialCharTok{+}\NormalTok{ merit\_effort }\SpecialCharTok{+}\NormalTok{ merit\_talent }\SpecialCharTok{+}\NormalTok{ perc\_inequality\_mean }\SpecialCharTok{+}\NormalTok{ just\_inequality\_mean }\SpecialCharTok{+}\NormalTok{ merit\_effort\_mean }\SpecialCharTok{+}\NormalTok{ merit\_talent\_mean }\SpecialCharTok{+}\NormalTok{ educ }\SpecialCharTok{+}\NormalTok{ quintil1 }\SpecialCharTok{+}\NormalTok{ ess }\SpecialCharTok{+}\NormalTok{ ideo }\SpecialCharTok{+}\NormalTok{ sex }\SpecialCharTok{+}\NormalTok{ age }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{|}\NormalTok{ idencuesta),}
                \AttributeTok{data =}\NormalTok{ df\_study1)}

\NormalTok{m9 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{+}\NormalTok{ ola\_2 }\SpecialCharTok{+}\NormalTok{ perc\_inequality}\SpecialCharTok{*}\NormalTok{ola\_num }\SpecialCharTok{+}\NormalTok{ just\_inequality }\SpecialCharTok{+}\NormalTok{ merit\_effort }\SpecialCharTok{+}\NormalTok{ merit\_talent }\SpecialCharTok{+}\NormalTok{ perc\_inequality\_mean }\SpecialCharTok{+}\NormalTok{ just\_inequality\_mean }\SpecialCharTok{+}\NormalTok{ merit\_effort\_mean }\SpecialCharTok{+}\NormalTok{ merit\_talent\_mean }\SpecialCharTok{+}\NormalTok{ educ }\SpecialCharTok{+}\NormalTok{ quintil1 }\SpecialCharTok{+}\NormalTok{ ess }\SpecialCharTok{+}\NormalTok{ ideo }\SpecialCharTok{+}\NormalTok{ sex }\SpecialCharTok{+}\NormalTok{ age }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{|}\NormalTok{ idencuesta),}
                \AttributeTok{data =}\NormalTok{ df\_study1)}

\NormalTok{m10 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{+}\NormalTok{ ola\_2 }\SpecialCharTok{+}\NormalTok{ perc\_inequality}\SpecialCharTok{*}\NormalTok{ola\_num }\SpecialCharTok{+}\NormalTok{ just\_inequality}\SpecialCharTok{*}\NormalTok{ola\_num }\SpecialCharTok{+}\NormalTok{ merit\_effort }\SpecialCharTok{+}\NormalTok{ merit\_talent }\SpecialCharTok{+}\NormalTok{ perc\_inequality\_mean }\SpecialCharTok{+}\NormalTok{ just\_inequality\_mean }\SpecialCharTok{+}\NormalTok{ merit\_effort\_mean }\SpecialCharTok{+}\NormalTok{ merit\_talent\_mean }\SpecialCharTok{+}\NormalTok{ educ }\SpecialCharTok{+}\NormalTok{ quintil1 }\SpecialCharTok{+}\NormalTok{ ess }\SpecialCharTok{+}\NormalTok{ ideo }\SpecialCharTok{+}\NormalTok{ sex }\SpecialCharTok{+}\NormalTok{ age }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{|}\NormalTok{ idencuesta),}
                \AttributeTok{data =}\NormalTok{ df\_study1)}

\NormalTok{m11 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{+}\NormalTok{ ola\_2 }\SpecialCharTok{+}\NormalTok{ perc\_inequality}\SpecialCharTok{*}\NormalTok{ola\_num }\SpecialCharTok{+}\NormalTok{ just\_inequality}\SpecialCharTok{*}\NormalTok{ola\_num }\SpecialCharTok{+}\NormalTok{ merit\_effort}\SpecialCharTok{*}\NormalTok{ola\_num }\SpecialCharTok{+}\NormalTok{ merit\_talent }\SpecialCharTok{+}\NormalTok{ perc\_inequality\_mean }\SpecialCharTok{+}\NormalTok{ just\_inequality\_mean }\SpecialCharTok{+}\NormalTok{ merit\_effort\_mean }\SpecialCharTok{+}\NormalTok{ merit\_talent\_mean }\SpecialCharTok{+}\NormalTok{ educ }\SpecialCharTok{+}\NormalTok{ quintil1 }\SpecialCharTok{+}\NormalTok{ ess }\SpecialCharTok{+}\NormalTok{ ideo }\SpecialCharTok{+}\NormalTok{ sex }\SpecialCharTok{+}\NormalTok{ age }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{|}\NormalTok{ idencuesta),}
                \AttributeTok{data =}\NormalTok{ df\_study1)}

\NormalTok{m12 }\OtherTok{\textless{}{-}} \FunctionTok{lmer}\NormalTok{(mjp }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{+}\NormalTok{ ola\_2 }\SpecialCharTok{+}\NormalTok{ perc\_inequality}\SpecialCharTok{*}\NormalTok{ola\_num }\SpecialCharTok{+}\NormalTok{ just\_inequality}\SpecialCharTok{*}\NormalTok{ola\_num }\SpecialCharTok{+}\NormalTok{ merit\_effort}\SpecialCharTok{*}\NormalTok{ola\_num }\SpecialCharTok{+}\NormalTok{ merit\_talent}\SpecialCharTok{*}\NormalTok{ola\_num }\SpecialCharTok{+}\NormalTok{ perc\_inequality\_mean }\SpecialCharTok{+}\NormalTok{ just\_inequality\_mean }\SpecialCharTok{+}\NormalTok{ merit\_effort\_mean }\SpecialCharTok{+}\NormalTok{ merit\_talent\_mean }\SpecialCharTok{+}\NormalTok{ educ }\SpecialCharTok{+}\NormalTok{ quintil1 }\SpecialCharTok{+}\NormalTok{ ess }\SpecialCharTok{+}\NormalTok{ ideo }\SpecialCharTok{+}\NormalTok{ sex }\SpecialCharTok{+}\NormalTok{ age }\SpecialCharTok{+}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ ola\_num }\SpecialCharTok{|}\NormalTok{ idencuesta),}
                \AttributeTok{data =}\NormalTok{ df\_study1)}
\end{Highlighting}
\end{Shaded}

\begin{table}

\caption{\label{tbl-modelos}Longitudinal multilevel models for market
justice preferences}

\centering{

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ccoef }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(}
  \StringTok{"(Intercept)"} \OtherTok{=} \StringTok{"Intercept"}\NormalTok{,}
  \StringTok{"ola2017"} \OtherTok{=} \StringTok{"Wave 2017"}\NormalTok{,}
  \StringTok{"ola2018"} \OtherTok{=} \StringTok{"Wave 2018"}\NormalTok{,}
  \StringTok{"ola2019"} \OtherTok{=} \StringTok{"Wave 2019"}\NormalTok{,}
  \StringTok{"ola2022"} \OtherTok{=} \StringTok{"Wave 2022"}\NormalTok{,}
  \StringTok{"ola2023"} \OtherTok{=} \StringTok{"Wave 2023"}\NormalTok{,}
  \AttributeTok{ola\_num =} \StringTok{"Wave"}\NormalTok{,}
  \AttributeTok{ola\_2 =} \StringTok{"Wave\^{}2"}\NormalTok{,}
  \AttributeTok{perc\_inequality =} \StringTok{"Perception inequality (WE)"}\NormalTok{,}
  \AttributeTok{just\_inequality =} \StringTok{"Justification inequality (WE)"}\NormalTok{,}
  \AttributeTok{merit\_effort =} \StringTok{"Merit: Effort (WE)"}\NormalTok{,}
  \AttributeTok{merit\_talent =} \StringTok{"Merit: Talent (WE)"}\NormalTok{,}
  \AttributeTok{perc\_inequality\_mean =} \StringTok{"Perception inequality (BE)"}\NormalTok{,}
  \AttributeTok{just\_inequality\_mean =} \StringTok{"Justification inequality (BE)"}\NormalTok{,}
  \AttributeTok{merit\_effort\_mean =} \StringTok{"Merit: Effort (BE)"}\NormalTok{,}
  \AttributeTok{merit\_talent\_mean =} \StringTok{"Merit: Talent (BE)"}\NormalTok{,}
  \AttributeTok{educUniversitary =} \StringTok{"Universitary (Ref.= Less than universitary)"}\NormalTok{,}
  \AttributeTok{quintil1Q2 =} \StringTok{"Quintile 2"}\NormalTok{,}
  \AttributeTok{quintil1Q3 =} \StringTok{"Quintile 3"}\NormalTok{,}
  \AttributeTok{quintil1Q4 =} \StringTok{"Quintile 4"}\NormalTok{,}
  \AttributeTok{quintil1Q5 =} \StringTok{"Quintile 5"}\NormalTok{,}
  \AttributeTok{quintil1QNA =} \StringTok{"Quintile NA"}\NormalTok{,}
  \AttributeTok{ess =} \StringTok{"Subjective social status"}\NormalTok{,}
  \AttributeTok{ideoCenter =} \StringTok{"Center"}\NormalTok{,}
  \AttributeTok{ideoRight =} \StringTok{"Right"}\NormalTok{,}
  \StringTok{"ideoDoes not identify"} \OtherTok{=} \StringTok{"Does not identify"}\NormalTok{,}
  \AttributeTok{sexFemale =} \StringTok{"Female (Ref.= Male)"}\NormalTok{,}
  \StringTok{"age30{-}49"} \OtherTok{=} \StringTok{"30{-}49"}\NormalTok{,}
  \StringTok{"age50{-}64"} \OtherTok{=} \StringTok{"50{-}64"}\NormalTok{,}
  \StringTok{"age65 or more"} \OtherTok{=} \StringTok{"65 or more"}
\NormalTok{  )}

\NormalTok{texreg}\SpecialCharTok{::}\FunctionTok{htmlreg}\NormalTok{(}\FunctionTok{list}\NormalTok{(m1, m2, m3, m4, m5, m6, m7, m8),}
               \AttributeTok{custom.model.names =} \FunctionTok{c}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(}\StringTok{"Model "}\NormalTok{, }\FunctionTok{seq}\NormalTok{(}\DecValTok{1}\SpecialCharTok{:}\DecValTok{8}\NormalTok{))),}
               \AttributeTok{caption =} \ConstantTok{NULL}\NormalTok{,}
               \AttributeTok{stars =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.05}\NormalTok{, }\FloatTok{0.01}\NormalTok{, }\FloatTok{0.001}\NormalTok{),}
               \AttributeTok{custom.coef.map =}\NormalTok{ ccoef,}
               \AttributeTok{digits =} \DecValTok{3}\NormalTok{,}
               \AttributeTok{groups =} \FunctionTok{list}\NormalTok{(}\StringTok{"Wave (Ref.= 2016)"} \OtherTok{=} \DecValTok{2}\SpecialCharTok{:}\DecValTok{6}\NormalTok{, }\StringTok{"Income (Ref.= Quintile 1)"} \OtherTok{=} \DecValTok{18}\SpecialCharTok{:}\DecValTok{22}\NormalTok{, }\StringTok{"Political identification (Ref.= Left)"} \OtherTok{=} \DecValTok{24}\SpecialCharTok{:}\DecValTok{26}\NormalTok{, }\StringTok{"Age (Ref. = 18{-}29)"} \OtherTok{=} \DecValTok{28}\SpecialCharTok{:}\DecValTok{30}\NormalTok{),}
               \AttributeTok{custom.note =} \StringTok{"Note: Cells contain regression coefficients with standard errors in parentheses. \%stars."}\NormalTok{,}
               \AttributeTok{threeparttable =}\NormalTok{ T,}
               \AttributeTok{leading.zero =}\NormalTok{ T,}
               \AttributeTok{float.pos =} \StringTok{"h!"}\NormalTok{,}
               \AttributeTok{use.packages =}\NormalTok{ F,}
               \AttributeTok{booktabs =}\NormalTok{ T,}
               \AttributeTok{scalebox =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

~

Model 1

Model 2

Model 3

Model 4

Model 5

Model 6

Model 7

Model 8

Intercept

1.966***

1.968***

2.108***

2.063***

1.755***

1.725***

1.537***

1.565***

~

(0.022)

(0.033)

(0.046)

(0.045)

(0.052)

(0.053)

(0.095)

(0.112)

Wave (Ref.= 2016)

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~~~~~Wave 2017

-0.163***

~

~

~

~

~

~

~

~

(0.028)

~

~

~

~

~

~

~

~~~~~Wave 2018

-0.013

~

~

~

~

~

~

~

~

(0.027)

~

~

~

~

~

~

~

~~~~~Wave 2019

-0.043

~

~

~

~

~

~

~

~

(0.027)

~

~

~

~

~

~

~

~~~~~Wave 2022

0.289***

~

~

~

~

~

~

~

~

(0.028)

~

~

~

~

~

~

~

~~~~~Wave 2023

0.284***

~

~

~

~

~

~

~

~

(0.027)

~

~

~

~

~

~

~

Wave

~

-0.069***

-0.072***

-0.057**

-0.057**

-0.058**

-0.064***

-0.064***

~

~

(0.018)

(0.018)

(0.018)

(0.018)

(0.018)

(0.018)

(0.018)

Wave\^{}2

~

0.017***

0.017***

0.015***

0.015***

0.015***

0.016***

0.016***

~

~

(0.002)

(0.002)

(0.002)

(0.002)

(0.002)

(0.002)

(0.002)

Perception inequality (WE)

~

~

-0.038***

-0.087***

-0.076***

-0.076***

-0.058***

-0.058***

~

~

~

(0.009)

(0.010)

(0.010)

(0.010)

(0.011)

(0.011)

Justification inequality (WE)

~

~

~

0.101***

0.096***

0.095***

0.056***

0.056***

~

~

~

~

(0.009)

(0.009)

(0.009)

(0.010)

(0.010)

Merit: Effort (WE)

~

~

~

~

0.107***

0.088***

0.071***

0.071***

~

~

~

~

~

(0.009)

(0.011)

(0.012)

(0.012)

Merit: Talent (WE)

~

~

~

~

~

0.028*

0.025*

0.024*

~

~

~

~

~

~

(0.011)

(0.012)

(0.012)

Perception inequality (BE)

~

~

~

~

~

~

-0.077***

-0.078***

~

~

~

~

~

~

~

(0.023)

(0.023)

Justification inequality (BE)

~

~

~

~

~

~

0.159***

0.134***

~

~

~

~

~

~

~

(0.021)

(0.021)

Merit: Effort (BE)

~

~

~

~

~

~

0.095**

0.091**

~

~

~

~

~

~

~

(0.033)

(0.033)

Merit: Talent (BE)

~

~

~

~

~

~

-0.008

-0.017

~

~

~

~

~

~

~

(0.033)

(0.033)

Universitary (Ref.= Less than universitary)

~

~

~

~

~

~

~

-0.006

~

~

~

~

~

~

~

~

(0.035)

Income (Ref.= Quintile 1)

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~~~~~Quintile 2

~

~

~

~

~

~

~

0.002

~

~

~

~

~

~

~

~

(0.038)

~~~~~Quintile 3

~

~

~

~

~

~

~

0.035

~

~

~

~

~

~

~

~

(0.038)

~~~~~Quintile 4

~

~

~

~

~

~

~

0.097*

~

~

~

~

~

~

~

~

(0.039)

~~~~~Quintile 5

~

~

~

~

~

~

~

0.147***

~

~

~

~

~

~

~

~

(0.042)

~~~~~Quintile NA

~

~

~

~

~

~

~

0.173**

~

~

~

~

~

~

~

~

(0.066)

Subjective social status

~

~

~

~

~

~

~

-0.007

~

~

~

~

~

~

~

~

(0.008)

Political identification (Ref.= Left)

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~~~~~Center

~

~

~

~

~

~

~

0.057

~

~

~

~

~

~

~

~

(0.037)

~~~~~Right

~

~

~

~

~

~

~

0.245***

~

~

~

~

~

~

~

~

(0.042)

~~~~~Does not identify

~

~

~

~

~

~

~

0.095**

~

~

~

~

~

~

~

~

(0.032)

Female (Ref.= Male)

~

~

~

~

~

~

~

-0.081**

~

~

~

~

~

~

~

~

(0.026)

Age (Ref. = 18-29)

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~

~~~~~30-49

~

~

~

~

~

~

~

-0.009

~

~

~

~

~

~

~

~

(0.037)

~~~~~50-64

~

~

~

~

~

~

~

0.003

~

~

~

~

~

~

~

~

(0.039)

~~~~~65 or more

~

~

~

~

~

~

~

0.002

~

~

~

~

~

~

~

~

(0.046)

AIC

20324.667

20314.582

20304.765

20190.877

20051.570

20054.362

20003.452

20033.225

BIC

20381.029

20363.898

20361.127

20254.284

20122.022

20131.859

20109.130

20237.537

Log Likelihood

-10154.334

-10150.291

-10144.383

-10086.439

-10015.785

-10016.181

-9986.726

-9987.613

Num. obs.

8478

8478

8478

8478

8478

8478

8478

8478

Num. groups: idencuesta

1682

1682

1682

1682

1682

1682

1682

1682

Var: idencuesta (Intercept)

0.176

0.218

0.213

0.206

0.181

0.180

0.182

0.173

Var: Residual

0.528

0.495

0.494

0.493

0.491

0.490

0.488

0.488

Var: idencuesta ola\_num

~

0.007

0.007

0.007

0.006

0.006

0.006

0.006

Cov: idencuesta (Intercept) ola\_num

~

-0.018

-0.017

-0.018

-0.015

-0.015

-0.016

-0.016

Note: Cells contain regression coefficients with standard errors in
parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
\textless{} 0.05.

}

\end{table}%

\begin{table}

\caption{\label{tbl-interactions}Time interactions within effects for
market justice preferences}

\centering{

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{ccoef }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(}
  \StringTok{"(Intercept)"} \OtherTok{=} \StringTok{"Intercept"}\NormalTok{,}
  \StringTok{"ola\_num:perc\_inequality"} \OtherTok{=} \StringTok{"Perception inequality (WE) x Wave"}\NormalTok{,}
  \StringTok{"ola\_num:just\_inequality"} \OtherTok{=} \StringTok{"Justification inequality (WE) x Wave"}\NormalTok{,}
  \StringTok{"ola\_num:merit\_effort"} \OtherTok{=} \StringTok{"Merit: Effort (WE) x Wave"}\NormalTok{,}
  \StringTok{"ola\_num:merit\_talent"} \OtherTok{=} \StringTok{"Merit: Talent (WE) x Wave"}
\NormalTok{  )}

\NormalTok{texreg}\SpecialCharTok{::}\FunctionTok{htmlreg}\NormalTok{(}\FunctionTok{list}\NormalTok{(m9,m10,m11,m12),}
               \AttributeTok{custom.model.names =} \FunctionTok{c}\NormalTok{(}\FunctionTok{paste0}\NormalTok{(}\StringTok{"Model "}\NormalTok{, }\FunctionTok{seq}\NormalTok{(}\DecValTok{9}\SpecialCharTok{:}\DecValTok{12}\NormalTok{))),}
               \AttributeTok{caption =} \ConstantTok{NULL}\NormalTok{,}
               \AttributeTok{stars =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.05}\NormalTok{, }\FloatTok{0.01}\NormalTok{, }\FloatTok{0.001}\NormalTok{),}
               \AttributeTok{custom.coef.map =}\NormalTok{ ccoef,}
               \AttributeTok{digits =} \DecValTok{3}\NormalTok{,}
               \AttributeTok{custom.note =} \StringTok{"Note: Cells contain regression coefficients with standard errors in parentheses. \%stars."}\NormalTok{,}
               \AttributeTok{threeparttable =}\NormalTok{ T,}
               \AttributeTok{leading.zero =}\NormalTok{ T,}
               \AttributeTok{float.pos =} \StringTok{"h!"}\NormalTok{,}
               \AttributeTok{use.packages =}\NormalTok{ F,}
               \AttributeTok{booktabs =}\NormalTok{ T,}
               \AttributeTok{scalebox =} \DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

~

Model 1

Model 2

Model 3

Model 4

Intercept

1.589***

1.601***

1.472***

1.455***

~

(0.123)

(0.123)

(0.129)

(0.131)

Perception inequality (WE) x Wave

0.002

-0.007

-0.009*

-0.009*

~

(0.004)

(0.004)

(0.004)

(0.004)

Justification inequality (WE) x Wave

~

0.018***

0.018***

0.019***

~

~

(0.004)

(0.004)

(0.004)

Merit: Effort (WE) x Wave

~

~

-0.013**

-0.010

~

~

~

(0.004)

(0.005)

Merit: Talent (WE) x Wave

~

~

~

-0.005

~

~

~

~

(0.005)

AIC

20044.276

20035.653

20036.766

20046.613

BIC

20255.633

20254.055

20262.214

20279.105

Log Likelihood

-9992.138

-9986.826

-9986.383

-9990.306

Num. obs.

8478

8478

8478

8478

Num. groups: idencuesta

1682

1682

1682

1682

Var: idencuesta (Intercept)

0.173

0.168

0.165

0.164

Var: idencuesta ola\_num

0.006

0.006

0.006

0.006

Cov: idencuesta (Intercept) ola\_num

-0.016

-0.015

-0.014

-0.014

Var: Residual

0.488

0.489

0.489

0.489

Note: Cells contain regression coefficients with standard errors in
parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
\textless{} 0.05.

}

\end{table}%

\end{document}

### 2025-04-23 01-introduction.qmd
Update paper

---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
---

# Introduction

Since 1980, economic inequality and wealth concentration have dramatically increased worldwide, becoming one of the main challenges for the social sciences. Globally, in 2021, less than 50% of the world population owned only 2% of the wealth, while the richest 10% concentrated 76%, and the wealthiest 1% captured nearly 38% of total assets [@chancel_world_2022]. This context of economic disparity has sparked renewed interest in studying not only the objective aspects of inequality, such as income and access to resources, but also its subjective dimensions, including perceptions, beliefs, and associated attitudes [@janmaat_subjective_2013]. This has led, among other reasons, to widespread concern in the social sciences about the political delegitimization that economic inequality can produce [@castillo_perception_2022]. Understanding how people perceive inequality is crucial, as these perceptions can influence how societies comprehend and justify the distribution of goods and services.

The perception of economic inequality can be understood as an individual's subjective assessment of how resources are allocated among members of a given society [@akyelken_urban_2020]. Regardless of their measurement, various studies have shown that this perception often underestimates the gap between the rich and the poor, which could have implications for attitudes toward the distribution of goods and services [@castillo_perception_2022; @schroder_income_2017]. Moreover, in recent years, there has been a discussion about the subjective and objective aspects of inequality, demonstrating inconsistencies between the two [@trump_income_2018]. This tension is relevant because advancing the study of perceptions could help understand how perceived inequality affects the distribution of resources within a society (Becker, 2020).

Perception of economic inequality has been associated with important outcomes such as redistributive preferences, justification of inequality, and legitimacy of the economic system (Castillo et al., 2022 \[acá citar gente de Granada\]). Research shows that greater perceived inequality fosters stronger preferences for redistribution, regardless of actual inequality levels [@garcia-sanchez_vicious_2019; @gimpelson_misperceiving_2018]. However, other findings suggest that heightened perceptions of income gaps between high- and low-paying occupations may lead to greater justification of inequality [@Castillo2011]. A less explored area in this regard concerns market justice, which denotes the degree to which individuals consider just that social goods and services (e.g., health, pensions, education) are allocated according to individual contribution, competition, and ability to pay [@lane_market_1986; @lindh_public_2015; @kluegel_legitimation_1999]. Market justice attitudes reflect the belief that the market promotes procedural fairness---equality of opportunity---so that outcomes result from individual achievement [@lane_market_1986]. Indeed, Lindh [-@lindh_public_2015] shows that market justice preferences are stronger in countries with higher private spending on services than in those with more comprehensive welfare systems. From this perspective, perceptions of economic inequality may be key, as lower perceived gaps can reinforce acceptance of free-market systems and justify existing inequalities.

Meritocracy posits that inequality is an inherent feature of societies but can be legitimized through fairer principles, such as effort and talent [@young_rise_1962; @davis_principles_2001]. Previous studies have demonstrated that people with stronger meritocratic beliefs tend to perceive less inequality as they attribute economic differences to individual achievements [@mijs_paradox_2021; @wilson_role_2003] and also to justify more inequality since these beliefs are associated with attitudes that maintain the legitimization of social differences [@batruch_belief_2023]. In a context like Chile, where the distribution of goods and services is predominantly governed by market logics strongly introduced during the military dictatorship (1973-1989) [@boccardo_30_2020], these beliefs can play a crucial role in the acceptance of social inequalities. Incorporating the perception of meritocracy in this literature allows for an understanding of how attitudes toward inequality are shaped not only by the perception of economic gaps, but also by beliefs about how these gaps are justified. Furthermore, these beliefs are often consolidated from an early age, reinforced by institutions promoting values such as effort and individual skills to climb socially [@castillo_socialization_2024; @reynolds_perceptions_2014].

The perception of meritocracy and the perception of economic inequality would interact intricately in shaping preferences for market justice. On the one hand, a stronger perception of merit as a criterion for distribution could minimize the perception of economic gaps, thus justifying an unequal system [@castillo_percepcion_2019; @mijs_paradox_2021]. On the other hand, the perception of inequality could moderate the impact of meritocracy, as a higher perception of economic gaps might challenge the idea that these are solely based on merit.

The primary objective of this study is to analyze the interplay between perceptions of inequality and meritocracy and their joint influence on preferences for market justice in Chile from 2016 to 2023, using longitudinal data from the ELSOC survey. This interaction is expected to provide a more comprehensive explanation of attitudes toward market justice in a country characterized by high inequality and strong free-market influence [@boccardo_30_2020; @flores_top_2020]. Additionally, this analysis seeks to elucidate how political and social contingencies---such as the 2019 and 2022 social movements---might have moderated these relationships by prompting more critical reflection on the commodification of social services. Recognizing the temporal dimension in shaping market justice preferences is essential, given that such preferences are not static but are influenced by historical and contextual factors that challenge or reaffirm prevailing social norms. In this regard, variations in perceived economic inequality and meritocracy over time can affect how individuals endorse market-based approaches. Consequently, public opinion may shift from supporting market justice to embracing redistributive policies aimed at mitigating inequalities.

### 2025-04-23 02-antecedents.qmd
Update paper

---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
---

# Theoretical views on market justice, inequality perception and meritocracy

## Justification of market inequality

Market justice has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit and effort. The most detailed theoretical account of this concept is provided by Lane [-@lane_market_1986], who defines market justice as a system of "earned deserts," whereby individuals are seen as entitled to their outcomes insofar as these are a product of personal productivity and competition. This view directly contrasts with political justice, which emphasizes redistributive mechanisms based on the principles of equality and need, and is often materialized through welfare policies. For Lane [-@lane_market_1986], market justice relies on the conception that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement. What matters is not whether everyone ends up with the same, but that everyone has the chance to compete, and that success reflects effort and ability. Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open. These beliefs align with a broader ideological framework that values individual autonomy, private property, and limited government intervention. In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of personal responsibility and societal efficiency.

Empirical studies have developed different strategies to capture and operationalize market justice preferences. A common approach in the literature is to gauge attitudes toward the legitimacy of inequality in specific domains, especially when linked to income differences. Kluegel and Smith [-@kluegel_beliefs_1981] and later Kluegel et al. [-@kluegel_legitimation_1999] laid the groundwork for measuring beliefs in market justice through survey items that assess whether people support economic inequality when it reflects personal effort and skill. Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, or pensions. For example, Knesebeck et al. (2016) and Immergut and Schneider [-@immergut_it_2020] examine whether citizens consider it fair that individuals with higher incomes can access better healthcare, while Lee and Stacey [-@lee_fairness_2023] apply a similar method in the context of education in Australia. These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts. Some research, like that of Lindh [-@lindh_public_2015] and Svallfors [-@svallfors_political_2007], uses these indicators in combination to measure broader preferences for market-based distribution in welfare systems. More recently, Castillo et al. [-@castillo_socialization_2024] introduced a single-item composite measure of market justice to assess student attitudes toward income-based access to education, healthcare, and pensions in Chile. These different operationalizations all aim to capture the extent to which individuals accept inequality when it is framed as a reflection of market outcomes.

The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded. Across the literature, there is strong empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher income, education, and occupational status---are more likely to endorse market-based distribution as fair and legitimate [@svallfors_political_2007; @koos_moral_2019]. This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo. At the same time, normative and ideological orientations---such as conservative, liberal, or neoliberal values---are consistently associated with greater support for meritocratic principles and skepticism toward redistribution. This is particularly visible in contexts such as Chile and Australia, where studies by Castillo et al. [-@castillo_socialization_2024] and Lee and Stacey [-@lee_fairness_2023] show that those with right-leaning political preferences tend to view income-based access to welfare services as legitimate. Beyond individual-level characteristics, country-level institutions also play a central role. In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015]. These findings suggest that market justice is not a fixed belief but a contextually shaped orientation, mediated by structural conditions, cultural values, and the historical development of welfare institutions.

## Perception of inequality

Perceptions of inequality refer to individuals' subjective evaluations of the extent, causes, and consequences of income and wealth disparities. Unlike objective measures such as the Gini index, perceived inequality captures how individuals make sense of distributive hierarchies in their everyday lives, shaped by reference groups, social comparisons, and information environments [@mijs_stratified_2016; @gimpelson_misperceiving_2018; @garcia-castro_perceiving_2020]. Scholars have proposed multiple dimensions of perceived inequality, including its magnitude (how significant are the gaps), vertical structure (between which groups), trend over time (increasing or decreasing), and legitimacy (whether it is just or not) [@engelhardt_what_2018; @garcia-sanchez_vicious_2019]. These dimensions encompass both cognitive and normative aspects of perceptions of inequality and can vary across societies and social groups, depending on exposure, ideology, and personal experience [@castillo_perception_2022; @garcia-sanchez_perceptions_2018].

Among the different approaches to measuring perceived inequality, one of the most widely used is the estimation of wage gaps between occupational extremes, such as between a CEO and a manual worker. This type of item provides a concrete frame that respondents can relate to more easily than abstract questions about national income distribution [@castillo_percepcion_2019; @easterbrook_social_2021; Willis et al., 2015]. While it enables the estimation of inequality using simple heuristics, this method is not without challenges. For instance, people often lack reliable knowledge about the earnings of those at the top of the income ladder, which leads to high variability in responses and the use of biased mental shortcuts [@knell_perceptions_2020]. Despite this, perceived wage gaps are strong predictors of political attitudes[@pedersen_attitudes_2019; @garcia-sanchez_perceptions_2018], making them a valuable tool for understanding public responses to economic disparities.

Another relevant critique of this approach lies in its conflation of different psychological constructs. Many surveys assess perceived inequality through Likert-type items that ask respondents to agree or disagree with statements, such as "income differences are too large," which captures general concern or discomfort rather than a specific perception [@Castillo2011; @garcia-sanchez_vicious_2019].These items mix cognitive estimations with affective evaluations, complicating the interpretation of what respondents perceive versus what they morally reject. As a result, the conceptual clarity between perceived inequality and inequality aversion remains blurred in many empirical studies. To address this limitation, recent work has emphasized the need to distinguish between absolute and comparative measures, as well as between ideal and actual estimates of economic gaps [@garcia-sanchez_creencias_2022; Auspurg et al., 2017].

Importantly, perceptions of inequality have been linked to attitudes about market justice---the belief that goods and services should be allocated according to individual contribution, competition, or willingness to pay [@lindh_public_2015; @kluegel_social_1995a]. Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011]. In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].

## Perception of meritocracy

Meritocracy constitutes a central ideological framework for legitimizing social inequality. Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions. As initially conceptualized by Michael Young [-@young_rise_1962], the term was meant to critique a system in which merit-based stratification becomes a new form of inequality. However, over time, meritocracy has been widely adopted in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2021; @sandel_tyranny_2020]. From a sociological perspective, the belief in meritocracy is more than a cognitive assessment; it reflects a moral lens through which individuals interpret inequality. People who believe that success results from hard work and talent are more likely to view social and economic disparities as legitimate [@castillo_percepcion_2019; @batruch_belief_2023]. Conversely, if they see outcomes as driven by luck, social origin, or systemic barriers, inequality is more likely to be perceived as unjust. This distinction becomes crucial in societies with persistent structural inequality, where public narratives often emphasize personal responsibility and merit while overlooking entrenched disadvantages.

We adopt a multidimensional perspective on meritocracy, distinguishing between two key dimensions: effort-based and talent-based perceptions. This distinction is essential, as it captures different pathways through which individuals justify inequality. Effort-based meritocracy emphasizes hard work and perseverance as the basis for social rewards, aligning closely with cultural narratives of personal responsibility. A talent-based meritocracy, by contrast, emphasizes intelligence and innate abilities, which are often perceived as less malleable and more unequally distributed. Both dimensions have been shown to correlate with acceptance of inequality, but they may carry distinct implications for how inequality is justified in specific domains [@castillo_multidimensional_2023]. The relevance of this distinction is supported by recent studies, which show that individuals respond differently to these dimensions. For instance, perceptions that effort is rewarded in society are more strongly associated with positive evaluations of fairness and acceptance of unequal outcomes [@batruch_belief_2023]. This may be because effort is seen as a controllable and morally virtuous trait, whereas talent is often perceived as a natural advantage. Consequently, effort-based meritocracy is likely more potent in legitimizing inequality, particularly in neoliberal contexts.

These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they personally endorse meritocratic principles. This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice (Tejero-Peregrina et al., 2025). Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity. Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_percepcion_2019; @castillo_socialization_2024].

In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups. Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor (Hoyt & Burnette, 2023). This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds. Moreover, Busemeyer, Abrassart, and Nezi (2019) argue that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures. This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.

## This study (hypotheses)

### 2025-04-23 03-methods.qmd
Update paper

@@
 author: "Equipo EDUMER"
-bibliography: "input/bib/refs.bib"
+bibliography: "input/bib/merit-edjust.bib"
 csl: "input/bib/apa6.csl"
 ---
+
+```{r}
+#| echo: false
+#| message: false
+#| warning: false
+
+library(knitr)
+knitr::opts_chunk$set(echo = TRUE, include = TRUE, warning = FALSE, message = FALSE)
+
+table_format <- if(is_html_output()) {
+  "html"
+} else if(is_latex_output()) {
+  "latex"
+}
+table_format2 <- if(is_html_output()) {
+  T
+} else if(is_latex_output()) {
+  F
+}
+
+options(kableExtra.html.bsTable = T)
+options(knitr.kable.NA = "")
+```
+
+```{r}
+
+#| include: false
+#| echo: false
+
+if (! require("pacman")) install.packages("pacman")
+
+pacman::p_load(tidyverse, 
+               sjmisc, 
+               sjPlot, 
+               lme4, 
+               here, 
+               performance,
+               influence.ME, 
+               marginaleffects,
+               MLMusingR,
+               texreg, 
+               ggdist,
+               misty,
+               kableExtra,
+               ggalluvial, 
+               shadowtext,
+               MetBrewer,
+               patchwork,
+               sjlabelled,
+               summarytools)
+
+options(scipen=999)
+rm(list = ls())
+```
+
+```{r}
+#| echo: false
+#| include: false
+
+load(file = here("input/data/proc/df_study1_long_t7.RData"))
+
+glimpse(df_study1_long_t7)
+
+# Generate analytical sample
+
+df_study1 <- df_study1_long_t7 %>%
+  select(idencuesta,
+         ola,
+         ponderador_long_total,
+         segmento,
+         estrato,
+         just_health,
+         just_pension,
+         just_educ,
+         mjp, 
+         perc_inequality, 
+         just_inequality,
+         merit_effort, 
+         merit_talent, 
+         educ,
+         quintil1,
+         sex,
+         age,
+         ess, 
+         ideo) %>% 
+  na.omit() %>% 
+  mutate(ola = as.factor(ola),
+         ola_num = as.numeric(ola),
+         ola_2=as.numeric(ola)^2)
+
+df_study1 <- df_study1 %>%
+  group_by(idencuesta) %>%             # Agrupar por el identificador del participante
+  mutate(n_participaciones = n()) %>%  # Contar el número de filas (participaciones) por participante
+  ungroup()
+
+df_study1 <- df_study1 %>% filter(n_participaciones>1)
+
+# Corregir etiquetas
+
+df_study1$just_health <- sjlabelled::set_label(df_study1$just_health, 
+                        label = "Health distributive justice")
+
+df_study1$just_pension <- sjlabelled::set_label(df_study1$just_pension, 
+                        label = "Pension distributive justice")
+
+df_study1$just_educ <- sjlabelled::set_label(df_study1$just_educ, 
+                        label = "Education distributive justice")
+
+df_study1$merit_effort <- sjlabelled::set_label(df_study1$merit_effort, 
+                        label = "People are rewarded for their efforts")
+
+df_study1$merit_talent <- sjlabelled::set_label(df_study1$merit_talent, 
+                        label = "People are rewarded for their intelligence")
+
+```
 
@@
 
+This study draws on data from the Chilean Longitudinal Social Survey (ELSOC), a panel study collected annually from 2016 to 2023.
+The survey evaluates how individuals think, feel, and behave regarding conflict and social cohesion in Chile.
+ELSOC employs a probabilistic, stratified, clustered, multistage sampling design encompassing major urban centers (Santiago, Valparaíso, and Concepción) and smaller cities.
+The target population includes women and men aged 18--75 who are habitual residents of private dwellings.
+The first wave included 2,927 participants, representing northern and southern regions, covering 77% of the country's total population and 93% of its urban population, with a response rate of 62.4% [@elsoc_estudio_2022].
+
+The survey has been conducted yearly since 2016, except in 2020, when it was suspended due to the pandemic.
+Waves 2016, 2017, 2018, 2019, 2022, and 2023 used computer-assisted personal interviewing (CAPI), while a reduced wave in 2021 employed computer-assisted telephone interviews (CATI).
+Wave 3 included a refreshment sample (N = 1,519) to counter attrition, but these cases are not used here to capture longer response trends.
+Between wave 1 and wave 7, attrition amounted to 40%, achieving a final sample of N = 1,741.
+Longitudinal weights adjust for both the sampling design and potential biases arising from systematic non-participation.
+Further details on sampling, attrition, and weighting can be found at <https://coes.cl/encuesta-panel/>, and the dataset is publicly available at <https://dataverse.harvard.edu/dataverse/elsoc>.
+
 ## Variables
 
+*Dependent variable*
+
+**Market justice preferences**: The dependent variable in this study is preferences for market justice.
+This construct is operationalized through three items that capture how strongly individuals justify conditioning access to social services---healthcare, pensions, and education---on income.
+Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is posed for pensions and education.
+In all cases, respondents indicate their level of agreement on a five-point Likert scale ranging from 1 ("strongly disagree") to 5 ("strongly agree").
+Additionally, we include a composite measure of "market justice preferences", calculated as the average of these three items ($\alpha$ = 0.84).
+This index ranges from 1 to 5, with higher values indicating stronger preferences for market justice (see @tbl-summary1).
+
+```{r}
+#| label: tbl-summary1
+#| tbl-cap: "Dependent variables for the first wave (2016)"
+#| tbl-cap-location: top
+#| results: asis
+#| echo: false
+
+t1 <- df_study1 %>% 
+  filter(ola == 1) %>% 
+  select(just_health, just_pension, just_educ, mjp)
+
+df<-dfSummary(t1,
+               plain.ascii = FALSE,
+               style = "multiline",
+               tmp.img.dir = "/tmp",
+               graph.magnif = 0.75,
+               headings = F,  # encabezado
+               varnumbers = F, # num variable
+               labels.col = T, # etiquetas
+               na.col = F,    # missing
+               graph.col = F, # plot
+               valid.col = T, # n valido
+               col.widths = c(30,10,10,10))
+
+df$Variable <- NULL # delete variable column
+
+print(df)
+```
+
+*Independent variables*
+
+**Perception of economic inequality**: The main independent variable refers to the perception of economic inequality, measured through the perceived wage gap, which serves as an absolute indicator of inequality perceptions [@castillo_perception_2022].
+This measure is derived from the gap between the perceived salaries of jobs at opposite ends of the occupational hierarchy.
+Specifically, it relies on the logarithm of the division between the perceived salary of a large-company president and that of an unskilled worker [@Castillo2011].
+Higher values of this term indicate a greater perception of economic inequality between occupations located at the extremes of the status continuum.
+
+**Justification of economic inequality**: This construct is measured as the just earnings gap, an approach to measuring the justification of economic inequality based on earning differences considered just between high and low status occupations [@Castillo2011].
+Similarly to economic inequality perception, this indicator relies on the logarithmic ratio between the just salary of a large-company president and that of an unskilled worker.
+Higher values of this term indicate a greater justification of economic inequality.
+
+**Meritocracy**: Meritocratic perception is operationalized through two components: one addressing effort and another focusing on talent [@young_rise_1962].
+The item used to gauge effort is: "In Chile, people are rewarded for their efforts," while the item for talent is: "In Chile, people are rewarded for their intelligence and skills." In both cases, respondents indicate their level of agreement on a five-point Likert scale, ranging from 1 ("strongly disagree") to 5 ("strongly agree").
+
+Table 2 shows the independent variables used, their response categories and their frequencies.
+
+```{r}
+#| label: tbl-summary2
+#| tbl-cap: "Independent variables for the first wave (2016)"
+#| tbl-cap-location: top
+#| results: asis
+#| echo: false
+
+t2 <- df_study1 %>% 
+  filter(ola == 1) %>% 
+  select(perc_inequality, just_inequality, merit_effort, merit_talent)
+
+df<-dfSummary(t2,
+               plain.ascii = FALSE,
+               style = "multiline",
+               tmp.img.dir = "/tmp",
+               graph.magnif = 0.75,
+               headings = F,  # encabezado
+               varnumbers = F, # num variable
+               labels.col = T, # etiquetas
+               na.col = F,    # missing
+               graph.col = F, # plot
+               valid.col = T, # n valido
+               col.widths = c(30,10,10,10))
+
+df$Variable <- NULL # delete variable column
+
+print(df)
+```
+
+*Controls*
+
+Sociodemographic and attitudinal variables are included to control for potential composition effects in the population.
+In terms of sociodemographic characteristics, we incorporate per capita household income quintile, educational level (1=Less than Universitary, 2=Universitary), age (in years), and sex (1=Male, 2=Female), which have been shown to influence market justice preferences significantly [@lindh_public_2015; @castillo_socialization_2024].
+Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) because they may confound the relationship between market justice preferences and perceptions of inequality and meritocracy [@schneider_poverty_2015].
+
 ## Methods
+
+Given the data's hierarchical structure, in which observations are nested in survey waves, we employ longitudinal multilevel linear models [@singer_applied_2009].
+In a panel-data framework, within-person effects capture how shifts in individual-level variables across waves are associated with variations in market justice preferences.
+By contrast, between-person effects focus on differences among individuals, explaining how long-term (or average) values relate to overall levels of market justice preferences.
+
+To estimate within-person effects, we use group-mean centering, where each respondent functions as the "group" (i.e., observations nested within persons).
+Meanwhile, the between-person effects are derived from each individual's average on these variables, calculated across the waves of panel data.
+
+All the analyses were conducted using R software and the *lme4* package [@bates_fitting_2015].

### 2025-04-23 04-analisis.qmd
Update paper

---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
---

```{r}
#| echo: false
#| message: false
#| warning: false

library(knitr)
knitr::opts_chunk$set(echo = TRUE, include = TRUE, warning = FALSE, message = FALSE)

table_format <- if(is_html_output()) {
  "html"
} else if(is_latex_output()) {
  "latex"
}
table_format2 <- if(is_html_output()) {
  T
} else if(is_latex_output()) {
  F
}

options(kableExtra.html.bsTable = T)
options(knitr.kable.NA = "")
```

```{r}
#| label: packages
#| include: true
#| echo: false

if (! require("pacman")) install.packages("pacman")

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               influence.ME, 
               marginaleffects,
               MLMusingR,
               texreg, 
               ggdist,
               misty,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools)

options(scipen=999)
rm(list = ls())
```

```{r}
#| label: data
#| echo: false
#| include: false

load(file = here("input/data/proc/df_study1_long_t7.RData"))

glimpse(df_study1_long_t7)

# Generate analytical sample

df_study1 <- df_study1_long_t7 %>%
  select(idencuesta,
         ola,
         ponderador_long_total,
         segmento,
         estrato,
         just_health,
         just_pension,
         just_educ,
         mjp, 
         perc_inequality, 
         just_inequality,
         merit_effort, 
         merit_talent, 
         educ,
         quintil1,
         sex,
         age,
         ess, 
         ideo) %>% 
  na.omit() %>% 
  mutate(ola = as.factor(ola),
         ola_num = as.numeric(ola),
         ola_2=as.numeric(ola)^2)

df_study1 <- df_study1 %>%
  group_by(idencuesta) %>%             # Agrupar por el identificador del participante
  mutate(n_participaciones = n()) %>%  # Contar el número de filas (participaciones) por participante
  ungroup()

df_study1 <- df_study1 %>% filter(n_participaciones>1)

# Corregir etiquetas

df_study1$just_health <- sjlabelled::set_label(df_study1$just_health, 
                        label = "Health distributive justice")

df_study1$just_pension <- sjlabelled::set_label(df_study1$just_pension, 
                        label = "Pension distributive justice")

df_study1$just_educ <- sjlabelled::set_label(df_study1$just_educ, 
                        label = "Education distributive justice")

df_study1$merit_effort <- sjlabelled::set_label(df_study1$merit_effort, 
                        label = "People are rewarded for their efforts")

df_study1$merit_talent <- sjlabelled::set_label(df_study1$merit_talent, 
                        label = "People are rewarded for their intelligence")

```
# Results

## Descriptive

@fig-alluvial shows the annual frequencies of market justice preferences for healthcare, pensions, and education from 2016 to 2023. Each year presents stacked percentage frequencies, and the flows between them reflect opinion changes among the same individuals from one year to the next, given that we are using panel data. For instance, of the 40.9% who strongly disagreed with justifying inequality in healthcare in 2019, around 24.3% maintained that position in 2022, while the remaining 16.6% shifted toward other response categories—primarily moving into disagreement rather than strong disagreement. Overall, more than half of the respondents exhibit a high level of disagreement (disagree + strongly disagree) with inequality in these three social service areas over time. Despite this general pattern, recent waves show a slight decrease in disagreement and a corresponding rise in support for inequality. Specifically, in healthcare and education, although disagreement remains substantial, agreement increased from 8.9% and 7.6% in 2019 to 12% and 12.5% in 2023, respectively. This shift is most evident in pensions, where the combined agree/strongly agree category grew by about 10 percentage points, from 16.9% in 2016 to 27.8% in 2023.

```{r}
#| label: fig-alluvial
#| out-width: '100%'
#| fig-asp: 1
#| fig-cap: "Change in the justification of inequality in health, pensions and education over time (2016-2023)"
#| fig-cap-location: top
#| echo: false
#| results: asis

datos.health <- df_study1 %>% 
  mutate(just_health = factor(just_health, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"))) %>% 
  group_by(idencuesta, ola) %>% 
  count(just_health) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 6 ~ "2022",
                          ola == 7 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                      "2017",
                                      "2018",
                                      "2019",
                                      "2022",
                                      "2023")))

etiquetas.health <- df_study1 %>%
  mutate(just_health = factor(just_health, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"))) %>% 
  group_by(ola, just_health) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 6 ~ "2022",
                          ola == 7 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

p1 <- datos.health %>% 
  ggplot(aes(x = wave, fill = just_health, stratum = just_health,
             alluvium = idencuesta, y = porcentaje)) +
  ggalluvial::geom_flow(alpha = .4) + 
  ggalluvial::geom_stratum(linetype = 0) +
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(values =  c("#0571B0","#92C5DE","#b3b3b3ff","#F4A582","#CA0020")) +
  geom_shadowtext(data = etiquetas.health,
                  aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
                  position = position_stack(vjust = .5),
                  show.legend = FALSE,
                  size = 3,
                  color = rep('white'),
                  bg.colour='grey30')+
  labs(y = "%",
       x = "Wave",
       fill = NULL,
       title = "a. Health") +
  theme_ggdist() +
  theme(legend.position = "none")

datos.pension <- df_study1 %>% 
   mutate(just_pension = factor(just_pension, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"))) %>% 
  group_by(idencuesta, ola) %>% 
  count(just_pension) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 6 ~ "2022",
                          ola == 7 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

etiquetas.pension <- df_study1 %>%
  mutate(just_pension = factor(just_pension, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"))) %>% 
  group_by(ola, just_pension) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 6 ~ "2022",
                          ola == 7 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

p2 <- datos.pension %>% 
  ggplot(aes(x = wave, fill = just_pension, stratum = just_pension,
             alluvium = idencuesta, y = porcentaje)) +
  ggalluvial::geom_flow(alpha = .4) + 
  ggalluvial::geom_stratum(linetype = 0) +
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(values =  c("#0571B0","#92C5DE","#b3b3b3ff","#F4A582","#CA0020")) +
  geom_shadowtext(data = etiquetas.pension,
                  aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
                  position = position_stack(vjust = .5),
                  show.legend = FALSE,
                  size = 3,
                  color = rep('white'),
                  bg.colour='grey30')+
  labs(y = "%",
       x = "Wave",
       fill = NULL,
       title = "b. Pension") +
  theme_ggdist() +
  theme(legend.position = "none")

datos.educ <- df_study1 %>% 
  mutate(just_educ = factor(just_educ, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"))) %>%
  group_by(idencuesta, ola) %>% 
  count(just_educ) %>% 
  group_by(ola) %>% 
  mutate(porcentaje=n/sum(n)) %>% 
  ungroup() %>% 
  na.omit() %>% 
  mutate(wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 6 ~ "2022",
                          ola == 7 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

etiquetas.educ <- df_study1 %>%
    mutate(just_educ = factor(just_educ, 
                              levels = c("Strongly agree",
                                         "Agree",
                                         "Neither agree nor desagree",
                                         "Desagree",
                                         "Strongly desagree"))) %>%
  group_by(ola, just_educ) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(ola) %>%
  mutate(porcentaje = count / sum(count)) %>% 
  na.omit() %>% 
  mutate(idencuesta = 1,
         wave = case_when(ola == 1 ~ "2016",
                          ola == 2 ~ "2017",
                          ola == 3 ~ "2018",
                          ola == 4 ~ "2019",
                          ola == 6 ~ "2022",
                          ola == 7 ~ "2023"),
         wave = factor(wave, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

p3 <- datos.educ %>% 
  ggplot(aes(x = wave, fill = just_educ, stratum = just_educ,
             alluvium = idencuesta, y = porcentaje)) +
  ggalluvial::geom_flow(alpha = .4) + 
  ggalluvial::geom_stratum(linetype = 0) +
  scale_y_continuous(labels = scales::percent) + 
  scale_fill_manual(values =  c("#0571B0","#92C5DE","#b3b3b3ff","#F4A582","#CA0020")) +
  geom_shadowtext(data = etiquetas.educ,
                  aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
                  position = position_stack(vjust = .5),
                  show.legend = FALSE,
                  size = 3,
                  color = rep('white'),
                  bg.colour='grey30')+
  labs(y = "%",
       x = "Wave",
       fill = NULL,
       title = "c. Education",
       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)") +
  theme_ggdist() +
  theme(legend.position = "bottom")

(p1 / p2 / p3)

```

Regarding the main variables of this study, @fig-meanchange depicts their average changes over the years. We observe an increase in the average level of market justice preferences in the most recent waves, which begins in 2019. The highest average consistently appears for perceived economic inequality, although this variable shows a downward trend of roughly one point over time. Interestingly, while perceptions of economic inequality declined in the latest measurements (2022–2023), market justice preferences increased. With respect to economic inequality justification, there is a steady decrease in average levels over the years, registering the lowest mean compared to the other variables. The meritocracy measures remain stable, though the perception that individuals are rewarded for intelligence is slightly higher than the perception that they are rewarded for their effort.

```{r}
#| label: fig-meanchange
#| out-width: '100%'
#| fig-asp: 1
#| fig-cap: "Change in the mean of market justice preferences, economic inequality perception and justification, and meritocracy (2016-2023)"
#| fig-cap-location: top
#| echo: false
#| results: asis
library(srvyr)

df_pond <- df_study1 %>% 
  mutate(ola = case_when(ola == 1 ~ "2016",
                         ola == 2 ~ "2017",
                         ola == 3 ~ "2018",
                         ola == 4 ~ "2019",
                         ola == 6 ~ "2022",
                         ola == 7 ~ "2023"),
         ola = factor(ola, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023"))) %>% 
  as_survey_design(.data = .,
                   ids = segmento, 
                   strata = estrato, 
                   weights = ponderador_long_total)

df_pond %>% 
  select(ola, mjp, perc_inequality, just_inequality, merit_effort, merit_talent) %>% 
  mutate_at(.vars = 2:6, .funs = ~ as.numeric(.)) %>% 
  group_by(ola) %>% 
  summarise_all(~survey_mean(., vartype = "ci")) %>% 
  pivot_longer(cols = -ola,
               names_to = "temp",
               values_to = "valor") %>%
  mutate(
    ci = case_when(
      str_ends(temp, "_low") ~ "ic_low",
      str_ends(temp, "_upp") ~ "ic_upp",
      TRUE                   ~ "mean"
    ),
    variable = str_remove(temp, "_low|_upp")
  ) %>%
  select(ola, variable, ci, valor) %>%
  pivot_wider(
    names_from  = ci,
    values_from = valor
  ) %>% 
  mutate(variable = case_when(variable == "mjp" ~ "Market justice preferences",
                              variable == "perc_inequality"  ~ "Inequality perception",
                              variable == "just_inequality"  ~ "Inequality justification",
                              variable == "merit_effort"     ~ "Merit: Effort",
                              variable == "merit_talent"     ~ "Merit: Talent"),
         variable = factor(variable, levels = c("Market justice preferences",
                                      "Inequality perception",
                                      "Inequality justification",
                                      "Merit: Effort",
                                      "Merit: Talent")
                              )) %>% 
  ggplot(aes(x = ola, y = mean, group = variable)) +
  geom_point(aes(shape=variable, color=variable), size = 3.5) +
  geom_line(aes(color = variable), linewidth = 0.8) +
  geom_errorbar(aes(ymin = ic_low, ymax = ic_upp, color = variable),
                width = 0.1) +
  scale_color_manual(values = c("#023858", # azul oscuro
                                "#0571B0", # azul intermedio
                                "#b3b3b3", # gris
                                "#F4A582", # salmón
                                "#B2182B"  # rojo profundo
                                         )) +
  scale_shape_manual(values=c(15, 16, 17, 18, 4)) +
  labs(y = "Mean",
       x = "Wave",
       color = NULL,
       shape = NULL,
       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)") +
  theme_ggdist() +
  theme(legend.position = "top",
        text = element_text(size = 12))

```

@fig-matrix presents the correlation matrix for the main variables in the latest wave (2023). Overall, the coefficients range from low to moderate. The association between market justice preferences and economic inequality perception is negative but small and not statistically significant ($r$ = -0.05, $p$ > .05). By contrast, market justice preferences positively and significantly correlate with the justification of economic inequality ($r$ = 0.16, $p$ < .01), as well as show low, positive associations with the two meritocracy variables ($r$ = 0.14, $p$ < .01; $r$ = 0.12, $p$ < .01). Perceived economic inequality has a strong, positive correlation with justifying economic inequality ($r$ = 0.53, $p$ < .01), but negative and nonsignificant correlations with both meritocracy perceptions ($r$ = -0.04, $p$ > .05; $r$ = -0.04, $p$ > .05). The justification of economic inequality indicator also has small yet statistically significant positive correlations with the meritocracy variables ($r$ = 0.05, $p$ < .05; $r$ = 0.07, $p$ < .01). Finally, the two meritocracy variables exhibit a strong positive association with each other ($r$ = 0.69, $p$ < .01).

```{r}
#| label: fig-matrix
#| out-width: '100%'
#| fig-asp: 1
#| fig-cap: "Correlation matrix of the main variables for the last wave (2023)"
#| fig-cap-location: top
#| echo: false
#| results: asis

M <- df_study1 %>%
  filter(ola == 7) %>% 
  select(mjp, perc_inequality, just_inequality, merit_effort, merit_talent) %>% 
  mutate_all(.funs = ~ as.numeric(.)) %>% 
  cor(use = "pairwise.complete.obs", method = "pearson")

diag(M) <- NA

rownames(M) <- c("A. Marjet justice preferences",
                 "B. Inequality gap perception",
                 "C. Inequality gap justification",
                 "D. People are rewarded for their efforts ",
                 "E. People are rewarded for their intelligence")

#set Column names of the matrix
colnames(M) <-c("(A)", "(B)","(C)","(D)","(E)")

#Plot the matrix using corrplot
corrplot::corrplot(M,
                   method = "color",
                   addCoef.col = "black",
                   type = "upper",
                   tl.col = "black",
                   col = colorRampPalette(c("#E16462", "white", "#0D0887"))(12),
                   bg = "white",
                   na.label = "-")

```

## Multilevel models

@tbl-modelos presents the results of the multilevel models estimated for market justice preferences, examining both individuals (within) and group-level (between) effects. The intraclass correlation [ICC, @hoxMultilevelAnalysisTechniques2017] from the empty model (see Supplementary Material), which decomposes the variance of market justice preferences, is 0.23, indicating that approximately 23% of the variation is attributable to differences between individuals. Complementary, 77% of the variation corresponds to within-individual differences over time.

According to Model 1, which includes the survey waves to capture intertemporal variations in the dependent variable, there is a decrease in 2017 ($\beta$ = -0.162, $p$ < .001) relative to 2016, and similarly in 2018 ($\beta$ = -0.015, $p$ > .05) and 2019 ($\beta$ = -0.041, $p$ > .05), although the latter effects are not statistically significant. In contrast, in the more recent waves of 2022 and 2023, there is a statistically significant increase in market justice preferences ($\beta$ = 0.289, $p$ < .001; $\beta$ = 0.286, $p$ < .001), suggesting a non-linear effect. To model this trajectory over time, Model 2 incorporates time (survey waves) as a continuous variable, along with its quadratic term, representing the non-linear association initially observed in Model 1. While the linear term (survey wave) shows a negative association, reflecting an overall decline in market preferences over time, the positive quadratic term indicates a reversal of this pattern in the final measurement points.

```{r, echo=FALSE, include=FALSE}

df_study1$merit_effort <- as_numeric(df_study1$merit_effort)
df_study1$merit_talent <- as_numeric(df_study1$merit_talent)

df_study1 <- df_study1 %>% 
  mutate(ola = case_when(ola == 1 ~ "2016",
                         ola == 2 ~ "2017",
                         ola == 3 ~ "2018",
                         ola == 4 ~ "2019",
                         ola == 6 ~ "2022",
                         ola == 7 ~ "2023"),
         ola = factor(ola, levels = c("2016",
                                        "2017",
                                        "2018",
                                        "2019",
                                        "2022",
                                        "2023")))

df_study1 <- df_study1 %>% 
  group_by(idencuesta) %>% 
  mutate(perc_inequality_mean = mean(perc_inequality, na.rm = T),
         perc_inequality_cwc = perc_inequality - perc_inequality_mean,
         just_inequality_mean = mean(just_inequality, na.rm = T),
         just_inequality_cwc = just_inequality - just_inequality_mean,
         merit_effort_mean = mean(merit_effort, na.rm = T),
         merit_effort_cwc = merit_effort - merit_effort_mean,
         merit_talent_mean = mean(merit_talent, na.rm = T),
         merit_talent_cwc = merit_talent - merit_talent_mean,
         ) %>% 
  ungroup()
```

```{r}
#| echo: false
#| include: false

m0 <- lmer(mjp ~ 1 + (1 | idencuesta), 
                data = df_study1)

performance::icc(m0, by_group = T)

m1 <- lmer(mjp ~ 1 + ola + (1 | idencuesta),
                data = df_study1)

m1.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + (1 | idencuesta),
                data = df_study1)

m2 <- lmer(mjp ~ 1 + ola_num + ola_2 + (1 + ola_num | idencuesta),
                data = df_study1)

anova(m1.1, m2)

m3 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + (1 + ola_num | idencuesta), 
           data = df_study1)

m4 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + (1 + ola_num | idencuesta),
                data = df_study1)

m5 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + (1 + ola_num | idencuesta),
                data = df_study1)

m6 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + (1 + ola_num | idencuesta),
                data = df_study1)

m7 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + (1 + ola_num | idencuesta),
                data = df_study1)

m8 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
                data = df_study1)

m9 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality +ola_num | idencuesta),
                data = df_study1)

m10 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality*ola_num + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + just_inequality+ola_num| idencuesta),
                data = df_study1)

m11 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort*ola_num + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort+ola_num | idencuesta),
                data = df_study1)

m12 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent*ola_num + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_talent +ola_num| idencuesta),
                data = df_study1)
```

```{r}
#| label: tbl-modelos
#| tbl-cap: "Longitudinal multilevel models for market justice preferences"
#| tbl-cap-location: top
#| results: asis
#| echo: false

ccoef <- list(
  "(Intercept)" = "Intercept",
  "ola2017" = "Wave 2017",
  "ola2018" = "Wave 2018",
  "ola2019" = "Wave 2019",
  "ola2022" = "Wave 2022",
  "ola2023" = "Wave 2023",
  ola_num = "Wave",
  ola_2 = "Wave^2",
  perc_inequality = "Perception inequality (WE)",
  just_inequality = "Justification inequality (WE)",
  merit_effort = "Merit: Effort (WE)",
  merit_talent = "Merit: Talent (WE)",
  perc_inequality_mean = "Perception inequality (BE)",
  just_inequality_mean = "Justification inequality (BE)",
  merit_effort_mean = "Merit: Effort (BE)",
  merit_talent_mean = "Merit: Talent (BE)"
  )

texreg::texreg(list(m1, m2, m3, m4, m5, m6, m7, m8),
               custom.model.names = c(paste0("Model ", seq(1:8))),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               groups = list("Wave (Ref.= 2016)" = 2:6),
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               threeparttable = T,
               leading.zero = T,
               float.pos = "H",
               use.packages = F,
               booktabs = T,
               scalebox = 1,
              custom.gof.rows = list("Controls"=c(rep("No",7), "Yes")))

```

Models 3 to 6 incorporate the within-group effects (WE) of the primary independent variables, capturing how individual changes in these variables over time shape the dependent variable. The results in Model 3 suggest that the within effect of perceived economic inequality is negative and statistically significant ($p$ < .001). Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.037 point decrease in market justice preferences. By contrast, in Model 4, the within effect of justifying economic inequality is positive and significant ($\beta$ = 0.100, $p$ < .001), implying that a rise in an individual’s justification of economic inequality over time corresponds to higher market justice preferences. Regarding meritocratic variables, Model 5 indicates that the perception of being rewarded for effort exerts a positive within effect ($\beta$ = 0.107, $p$ < .001). Similarly, Model 6 shows that viewing intelligence and ability (talent) as rewarded is also positively related to market justice preferences ($\beta$ = 0.029, $p$ < .05). Taken together, these results suggest that individuals who increasingly perceive meritocracy—whether through effort or talent—tend to hold stronger market justice preferences.

When examining the between-group effects (BE) in Model 7, which capture differences between individuals in the average of the main variables, a similar pattern emerges. Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.077, $p$ < .001), whereas those who justify economic inequality more strongly exhibit higher market justice preferences ($\beta$ = 0.156, $p$ < .001). Furthermore, the meritocratic perception that effort is rewarded has a positive impact on market justice preferences ($\beta$ = 0.092, $p$ < .05). By contrast, the perception that talent is rewarded shows a negative coefficient ($\beta$ = -0.006), though this relationship is not statistically significant at the 95% confidence level.

Model 8 introduces the control variables. The within- and between-effects of the main predictors retain both their direction and statistical significance, indicating that the associations are robust to adjustment. Regarding the controls, belonging to higher household-income quintiles is associated with stronger market-justice preferences: individuals in the fourth and fifth quintiles score higher than those in the bottom quintile ($\beta$ = 0.093, $p$ < .05; $\beta$ = 0.147, $p$ < .001, respectively). Respondents who did not report their income also express greater support for market justice ($\beta$ = 0.172, $p$ < .01). Political orientation matters as well: compared with those on the left, individuals who place themselves on the right ($\beta$ = 0.246, $p$ < .001) or who declare no political position ($\beta$ = 0.092, $p$ < .01) display higher market-justice preferences. Finally, women exhibit lower market-justice preferences than men ($\beta$ = −0.080, $p$ < .01).

```{r}
#| label: tbl-interactions
#| tbl-cap: "Intereactions for within effects and time"
#| tbl-cap-location: top

ccoef <- list(
  "(Intercept)" = "Intercept",
  "ola_num:perc_inequality" = "Perception inequality (WE) x Wave",
  "ola_num:just_inequality" = "Justification inequality (WE) x Wave",
  "ola_num:merit_effort" = "Merit: Effort (WE) x Wave",
  "ola_num:merit_talent" = "Merit: Talent (WE) x Wave"
  )

texreg::texreg(list(m9,m10,m11,m12),
               custom.model.names = c("Model 9", "Model 10", "Model 11", "Model 12"),
               caption = NULL,
               stars = c(0.05, 0.01, 0.001),
               custom.coef.map = ccoef,
               digits = 3,
               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
               threeparttable = T,
               leading.zero = T,
               float.pos = "h!",
               use.packages = F,
               booktabs = T,
               scalebox = 1,
               custom.gof.rows = list("Controls"=c(rep("Yes",4))))
```

### 2025-04-23 paper.qmd
Update paper

@@
   pdf: 
-    geometry: 
-      - top=30mm
-      - left=20mm
+    geometry: margin=2cm
     template-partials: 
@@
 editor: source
-bibliography: "input/bib/refs.bib"
+bibliography: "input/bib/merit-edjust.bib"
 csl: "input/bib/apa6.csl"
@@
 
-{{< include 07-supplementary-material.qmd >}}
+# References
+
+::: {#refs}
+:::

### 2025-04-23 paper.tex
Update paper

@@
 \usepackage{xcolor}
-\usepackage[top=30mm,left=20mm]{geometry}
+\usepackage[margin=2cm]{geometry}
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
+  \fi
+  % set entry spacing
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
+Since 1980, economic inequality and wealth concentration have
+dramatically increased worldwide, becoming one of the main challenges
+for the social sciences. Globally, in 2021, less than 50\% of the world
+population owned only 2\% of the wealth, while the richest 10\%
+concentrated 76\%, and the wealthiest 1\% captured nearly 38\% of total
+assets (\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}). This
+context of economic disparity has sparked renewed interest in studying
+not only the objective aspects of inequality, such as income and access
+to resources, but also its subjective dimensions, including perceptions,
+beliefs, and associated attitudes
+(\citeproc{ref-janmaat_subjective_2013}{Janmaat, 2013}). This has led,
+among other reasons, to widespread concern in the social sciences about
+the political delegitimization that economic inequality can produce
+(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}).
+Understanding how people perceive inequality is crucial, as these
+perceptions can influence how societies comprehend and justify the
+distribution of goods and services.
+
+The perception of economic inequality can be understood as an
+individual's subjective assessment of how resources are allocated among
+members of a given society (\citeproc{ref-akyelken_urban_2020}{Akyelken,
+2020}). Regardless of their measurement, various studies have shown that
+this perception often underestimates the gap between the rich and the
+poor, which could have implications for attitudes toward the
+distribution of goods and services
+(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
+\citeproc{ref-schroder_income_2017}{Schröder, 2017}). Moreover, in
+recent years, there has been a discussion about the subjective and
+objective aspects of inequality, demonstrating inconsistencies between
+the two (\citeproc{ref-trump_income_2018}{Trump, 2018}). This tension is
+relevant because advancing the study of perceptions could help
+understand how perceived inequality affects the distribution of
+resources within a society (Becker, 2020).
+
+Perception of economic inequality has been associated with important
+outcomes such as redistributive preferences, justification of
+inequality, and legitimacy of the economic system (Castillo et al., 2022
+{[}acá citar gente de Granada{]}). Research shows that greater perceived
+inequality fosters stronger preferences for redistribution, regardless
+of actual inequality levels
+(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
+2019}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
+Treisman, 2018}). However, other findings suggest that heightened
+perceptions of income gaps between high- and low-paying occupations may
+lead to greater justification of inequality
+(\citeproc{ref-Castillo2011}{Castillo, 2011}). A less explored area in
+this regard concerns market justice, which denotes the degree to which
+individuals consider just that social goods and services (e.g., health,
+pensions, education) are allocated according to individual contribution,
+competition, and ability to pay
+(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
+\citeproc{ref-lane_market_1986}{Lane, 1986};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Market justice attitudes
+reflect the belief that the market promotes procedural
+fairness---equality of opportunity---so that outcomes result from
+individual achievement (\citeproc{ref-lane_market_1986}{Lane, 1986}).
+Indeed, Lindh (\citeproc{ref-lindh_public_2015}{2015}) shows that market
+justice preferences are stronger in countries with higher private
+spending on services than in those with more comprehensive welfare
+systems. From this perspective, perceptions of economic inequality may
+be key, as lower perceived gaps can reinforce acceptance of free-market
+systems and justify existing inequalities.
+
+Meritocracy posits that inequality is an inherent feature of societies
+but can be legitimized through fairer principles, such as effort and
+talent (\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
+\citeproc{ref-young_rise_1962}{Young, 1962}). Previous studies have
+demonstrated that people with stronger meritocratic beliefs tend to
+perceive less inequality as they attribute economic differences to
+individual achievements (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
+\citeproc{ref-wilson_role_2003}{Wilson, 2003}) and also to justify more
+inequality since these beliefs are associated with attitudes that
+maintain the legitimization of social differences
+(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In a context
+like Chile, where the distribution of goods and services is
+predominantly governed by market logics strongly introduced during the
+military dictatorship (1973-1989)
+(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}), these beliefs can
+play a crucial role in the acceptance of social inequalities.
+Incorporating the perception of meritocracy in this literature allows
+for an understanding of how attitudes toward inequality are shaped not
+only by the perception of economic gaps, but also by beliefs about how
+these gaps are justified. Furthermore, these beliefs are often
+consolidated from an early age, reinforced by institutions promoting
+values such as effort and individual skills to climb socially
+(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
+\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).
+
+The perception of meritocracy and the perception of economic inequality
+would interact intricately in shaping preferences for market justice. On
+the one hand, a stronger perception of merit as a criterion for
+distribution could minimize the perception of economic gaps, thus
+justifying an unequal system
+(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
+\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). On the other hand, the
+perception of inequality could moderate the impact of meritocracy, as a
+higher perception of economic gaps might challenge the idea that these
+are solely based on merit.
+
+The primary objective of this study is to analyze the interplay between
+perceptions of inequality and meritocracy and their joint influence on
+preferences for market justice in Chile from 2016 to 2023, using
+longitudinal data from the ELSOC survey. This interaction is expected to
+provide a more comprehensive explanation of attitudes toward market
+justice in a country characterized by high inequality and strong
+free-market influence (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
+\citeproc{ref-flores_top_2020}{Flores et al., 2020}). Additionally, this
+analysis seeks to elucidate how political and social
+contingencies---such as the 2019 and 2022 social movements---might have
+moderated these relationships by prompting more critical reflection on
+the commodification of social services. Recognizing the temporal
+dimension in shaping market justice preferences is essential, given that
+such preferences are not static but are influenced by historical and
+contextual factors that challenge or reaffirm prevailing social norms.
+In this regard, variations in perceived economic inequality and
+meritocracy over time can affect how individuals endorse market-based
+approaches. Consequently, public opinion may shift from supporting
+market justice to embracing redistributive policies aimed at mitigating
+inequalities.
+
+\section{Theoretical views on market justice, inequality perception and
+meritocracy}\label{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}
+
+\subsection{Justification of market
+inequality}\label{justification-of-market-inequality}
+
+Market justice has been discussed in the literature as a normative
+principle that legitimates the distribution of economic rewards based on
+individual merit and effort. The most detailed theoretical account of
+this concept is provided by Lane
+(\citeproc{ref-lane_market_1986}{1986}), who defines market justice as a
+system of ``earned deserts,'' whereby individuals are seen as entitled
+to their outcomes insofar as these are a product of personal
+productivity and competition. This view directly contrasts with
+political justice, which emphasizes redistributive mechanisms based on
+the principles of equality and need, and is often materialized through
+welfare policies. For Lane (\citeproc{ref-lane_market_1986}{1986}),
+market justice relies on the conception that markets are neutral and
+self-regulating arenas, where individuals are treated fairly because
+they face the same formal rules of engagement. What matters is not
+whether everyone ends up with the same, but that everyone has the chance
+to compete, and that success reflects effort and ability. Consequently,
+the legitimacy of market justice stems from the assumption that
+inequalities are not only inevitable but fair---so long as the rules are
+transparent and opportunities are open. These beliefs align with a
+broader ideological framework that values individual autonomy, private
+property, and limited government intervention. In this way, market
+justice provides a moral justification for inequality by framing it as a
+necessary outcome of personal responsibility and societal efficiency.
+
+Empirical studies have developed different strategies to capture and
+operationalize market justice preferences. A common approach in the
+literature is to gauge attitudes toward the legitimacy of inequality in
+specific domains, especially when linked to income differences. Kluegel
+and Smith (\citeproc{ref-kluegel_beliefs_1981}{1981}) and later Kluegel
+et al. (\citeproc{ref-kluegel_legitimation_1999}{1999}) laid the
+groundwork for measuring beliefs in market justice through survey items
+that assess whether people support economic inequality when it reflects
+personal effort and skill. Over time, this approach has been extended
+beyond income to include other market-mediated outcomes, such as
+education, healthcare, or pensions. For example, Knesebeck et al.~(2016)
+and Immergut and Schneider (\citeproc{ref-immergut_it_2020}{2020})
+examine whether citizens consider it fair that individuals with higher
+incomes can access better healthcare, while Lee and Stacey
+(\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
+context of education in Australia. These studies usually rely on a
+survey item asking respondents to evaluate the fairness of income-based
+access to welfare services, allowing for comparing justice perceptions
+across different contexts. Some research, like that of Lindh
+(\citeproc{ref-lindh_public_2015}{2015}) and Svallfors
+(\citeproc{ref-svallfors_political_2007}{2007}), uses these indicators
+in combination to measure broader preferences for market-based
+distribution in welfare systems. More recently, Castillo et al.
+(\citeproc{ref-castillo_socialization_2024}{2024}) introduced a
+single-item composite measure of market justice to assess student
+attitudes toward income-based access to education, healthcare, and
+pensions in Chile. These different operationalizations all aim to
+capture the extent to which individuals accept inequality when it is
+framed as a reflection of market outcomes.
+
+The study of market justice preferences has increasingly focused on how
+they are shaped by individuals' socioeconomic position, normative
+beliefs, and the institutional context in which they are embedded.
+Across the literature, there is strong empirical evidence suggesting
+that individuals in more advantaged socioeconomic positions---those with
+higher income, education, and occupational status---are more likely to
+endorse market-based distribution as fair and legitimate
+(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
+\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). This tendency
+reflects not only material self-interest but also a broader moral
+economy, in which winners of the market system internalize
+justifications for the status quo. At the same time, normative and
+ideological orientations---such as conservative, liberal, or neoliberal
+values---are consistently associated with greater support for
+meritocratic principles and skepticism toward redistribution. This is
+particularly visible in contexts such as Chile and Australia, where
+studies by Castillo et al.
+(\citeproc{ref-castillo_socialization_2024}{2024}) and Lee and Stacey
+(\citeproc{ref-lee_fairness_2023}{2023}) show that those with
+right-leaning political preferences tend to view income-based access to
+welfare services as legitimate. Beyond individual-level characteristics,
+country-level institutions also play a central role. In liberal welfare
+regimes like those of the United States or the United Kingdom, market
+justice preferences are more widespread, while in coordinated or
+social-democratic regimes---such as Sweden or Germany---citizens are
+generally more critical of market-based inequalities
+(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
+that market justice is not a fixed belief but a contextually shaped
+orientation, mediated by structural conditions, cultural values, and the
+historical development of welfare institutions.
+
+\subsection{Perception of inequality}\label{perception-of-inequality}
+
+Perceptions of inequality refer to individuals' subjective evaluations
+of the extent, causes, and consequences of income and wealth
+disparities. Unlike objective measures such as the Gini index, perceived
+inequality captures how individuals make sense of distributive
+hierarchies in their everyday lives, shaped by reference groups, social
+comparisons, and information environments
+(\citeproc{ref-garcia-castro_perceiving_2020}{García-Castro et al.,
+2020}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
+Treisman, 2018}; \citeproc{ref-mijs_stratified_2016}{Mijs, 2016}).
+Scholars have proposed multiple dimensions of perceived inequality,
+including its magnitude (how significant are the gaps), vertical
+structure (between which groups), trend over time (increasing or
+decreasing), and legitimacy (whether it is just or not)
+(\citeproc{ref-engelhardt_what_2018}{Engelhardt \& Wagener, 2018};
+\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
+2019}). These dimensions encompass both cognitive and normative aspects
+of perceptions of inequality and can vary across societies and social
+groups, depending on exposure, ideology, and personal experience
+(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
+\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
+2018}).
+
+Among the different approaches to measuring perceived inequality, one of
+the most widely used is the estimation of wage gaps between occupational
+extremes, such as between a CEO and a manual worker. This type of item
+provides a concrete frame that respondents can relate to more easily
+than abstract questions about national income distribution {[}Castillo
+et al. (\citeproc{ref-castillo_percepcion_2019}{2012}); Easterbrook
+(\citeproc{ref-easterbrook_social_2021}{2021}); Willis et al., 2015{]}.
+While it enables the estimation of inequality using simple heuristics,
+this method is not without challenges. For instance, people often lack
+reliable knowledge about the earnings of those at the top of the income
+ladder, which leads to high variability in responses and the use of
+biased mental shortcuts (\citeproc{ref-knell_perceptions_2020}{Knell \&
+Stix, 2020}). Despite this, perceived wage gaps are strong predictors of
+political
+attitudes(\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez
+et al., 2018}; \citeproc{ref-pedersen_attitudes_2019}{Pedersen \& Mutz,
+2019}), making them a valuable tool for understanding public responses
+to economic disparities.
+
+Another relevant critique of this approach lies in its conflation of
+different psychological constructs. Many surveys assess perceived
+inequality through Likert-type items that ask respondents to agree or
+disagree with statements, such as ``income differences are too large,''
+which captures general concern or discomfort rather than a specific
+perception (\citeproc{ref-Castillo2011}{Castillo, 2011};
+\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
+2019}).These items mix cognitive estimations with affective evaluations,
+complicating the interpretation of what respondents perceive versus what
+they morally reject. As a result, the conceptual clarity between
+perceived inequality and inequality aversion remains blurred in many
+empirical studies. To address this limitation, recent work has
+emphasized the need to distinguish between absolute and comparative
+measures, as well as between ideal and actual estimates of economic gaps
+{[}García-Sánchez \& De Carvalho
+(\citeproc{ref-garcia-sanchez_creencias_2022}{2022}); Auspurg et al.,
+2017{]}.
+
+Importantly, perceptions of inequality have been linked to attitudes
+about market justice---the belief that goods and services should be
+allocated according to individual contribution, competition, or
+willingness to pay (\citeproc{ref-kluegel_social_1995a}{Kluegel et al.,
+1995}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research
+indicates that lower perceived inequality can reinforce support for
+market-based distributive arrangements by suggesting that the system is
+fair and that outcomes reflect effort and ability
+(\citeproc{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when inequality
+is perceived as excessive or structurally determined, individuals could
+question the legitimacy of market justice and become more supportive of
+redistributive policies
+(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
+\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
+2019}).
+
+\subsection{Perception of meritocracy}\label{perception-of-meritocracy}
+
+Meritocracy constitutes a central ideological framework for legitimizing
+social inequality. Rooted in the belief that rewards and positions
+should be allocated based on individual effort and talent, meritocracy
+operates as a normative ideal and a descriptive belief about how society
+functions. As initially conceptualized by Michael Young
+(\citeproc{ref-young_rise_1962}{1962}), the term was meant to critique a
+system in which merit-based stratification becomes a new form of
+inequality. However, over time, meritocracy has been widely adopted in
+many societies as a fair and desirable principle of distribution,
+particularly within liberal democracies and market-oriented economies
+(\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
+\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
+perspective, the belief in meritocracy is more than a cognitive
+assessment; it reflects a moral lens through which individuals interpret
+inequality. People who believe that success results from hard work and
+talent are more likely to view social and economic disparities as
+legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
+\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}).
+Conversely, if they see outcomes as driven by luck, social origin, or
+systemic barriers, inequality is more likely to be perceived as unjust.
+This distinction becomes crucial in societies with persistent structural
+inequality, where public narratives often emphasize personal
+responsibility and merit while overlooking entrenched disadvantages.
+
+We adopt a multidimensional perspective on meritocracy, distinguishing
+between two key dimensions: effort-based and talent-based perceptions.
+This distinction is essential, as it captures different pathways through
+which individuals justify inequality. Effort-based meritocracy
+emphasizes hard work and perseverance as the basis for social rewards,
+aligning closely with cultural narratives of personal responsibility. A
+talent-based meritocracy, by contrast, emphasizes intelligence and
+innate abilities, which are often perceived as less malleable and more
+unequally distributed. Both dimensions have been shown to correlate with
+acceptance of inequality, but they may carry distinct implications for
+how inequality is justified in specific domains
+(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}).
+The relevance of this distinction is supported by recent studies, which
+show that individuals respond differently to these dimensions. For
+instance, perceptions that effort is rewarded in society are more
+strongly associated with positive evaluations of fairness and acceptance
+of unequal outcomes (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
+2023}). This may be because effort is seen as a controllable and morally
+virtuous trait, whereas talent is often perceived as a natural
+advantage. Consequently, effort-based meritocracy is likely more potent
+in legitimizing inequality, particularly in neoliberal contexts.
+
+These dimensions of meritocracy reflect how respondents perceive
+society's distributive logic, regardless of whether they personally
+endorse meritocratic principles. This distinction aligns with recent
+findings indicating that individuals distinguish between how merit is
+perceived in society and how it should ideally operate, which in turn
+shapes their preferences for redistribution and justice
+(Tejero-Peregrina et al., 2025). Meritocratic beliefs serve as symbolic
+justifications for unequal outcomes, particularly when access is
+stratified by income or social opportunity. Prior studies have shown
+that individuals who perceive higher levels of meritocracy tend to
+express stronger support for unequal distributions that reflect market
+outcomes (\citeproc{ref-castillo_percepcion_2019}{Castillo et al.,
+2012}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
+2024}).
+
+In addition to influencing individual attitudes toward inequality,
+meritocratic beliefs can contribute to social division and the
+stigmatization of disadvantaged groups. Recent research has demonstrated
+that exposure to meritocratic narratives can reinforce the belief that
+poverty results from individual failings rather than systemic
+conditions, reducing support for redistributive measures and increasing
+the stigmatization of the poor (Hoyt \& Burnette, 2023). This reinforces
+negative stereotypes and reduces empathy toward individuals from lower
+socioeconomic backgrounds. Moreover, Busemeyer, Abrassart, and Nezi
+(2019) argue that meritocratic narratives can serve as feedback
+mechanisms that shape public opinion and well-being by framing
+individuals' understanding of welfare outcomes as deserved or undeserved
+within existing institutional structures. This psychological mechanism
+highlights the normative power of meritocracy in stabilizing unequal
+systems by shaping political attitudes and personal perceptions of
+success and failure.
+
+\subsection{This study (hypotheses)}\label{this-study-hypotheses}
+
 \section{Data, Variables and Methods}\label{data-variables-and-methods}
@@
 
+This study draws on data from the Chilean Longitudinal Social Survey
+(ELSOC), a panel study collected annually from 2016 to 2023. The survey
+evaluates how individuals think, feel, and behave regarding conflict and
+social cohesion in Chile. ELSOC employs a probabilistic, stratified,
+clustered, multistage sampling design encompassing major urban centers
+(Santiago, Valparaíso, and Concepción) and smaller cities. The target
+population includes women and men aged 18--75 who are habitual residents
+of private dwellings. The first wave included 2,927 participants,
+representing northern and southern regions, covering 77\% of the
+country's total population and 93\% of its urban population, with a
+response rate of 62.4\% (\citeproc{ref-elsoc_estudio_2022}{ELSOC,
+2022}).
+
+The survey has been conducted yearly since 2016, except in 2020, when it
+was suspended due to the pandemic. Waves 2016, 2017, 2018, 2019, 2022,
+and 2023 used computer-assisted personal interviewing (CAPI), while a
+reduced wave in 2021 employed computer-assisted telephone interviews
+(CATI). Wave 3 included a refreshment sample (N = 1,519) to counter
+attrition, but these cases are not used here to capture longer response
+trends. Between wave 1 and wave 7, attrition amounted to 40\%, achieving
+a final sample of N = 1,741. Longitudinal weights adjust for both the
+sampling design and potential biases arising from systematic
+non-participation. Further details on sampling, attrition, and weighting
+can be found at \url{https://coes.cl/encuesta-panel/}, and the dataset
+is publicly available at
+\url{https://dataverse.harvard.edu/dataverse/elsoc}.
+
 \subsection{Variables}\label{variables}
+
+\emph{Dependent variable}
+
+\textbf{Market justice preferences}: The dependent variable in this
+study is preferences for market justice. This construct is
+operationalized through three items that capture how strongly
+individuals justify conditioning access to social services---healthcare,
+pensions, and education---on income. Specifically, the justification of
+inequality in healthcare is assessed by the question: ``Is it fair in
+Chile that people with higher incomes can access better healthcare than
+people with lower incomes?'' The same question is posed for pensions and
+education. In all cases, respondents indicate their level of agreement
+on a five-point Likert scale ranging from 1 (``strongly disagree'') to 5
+(``strongly agree''). Additionally, we include a composite measure of
+``market justice preferences'', calculated as the average of these three
+items (\(\alpha\) = 0.84). This index ranges from 1 to 5, with higher
+values indicating stronger preferences for market justice (see
+Table~\ref{tbl-summary1}).
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3367}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3367}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2143}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1122}}@{}}
+\caption{Dependent variables for the first wave
+(2016)}\label{tbl-summary1}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Label
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Stats / Values
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Freqs (\% of Valid)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Valid
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Label
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Stats / Values
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Freqs (\% of Valid)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Valid
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Health distributive justice &
+\begin{minipage}[t]{\linewidth}\raggedright
+1. Strongly desagree\\
+2. Desagree\\
+3. Neither agree nor desagre\\
+4. Agree\\
+5. Strongly agree\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+554 (37.5\%)\\
+713 (48.3\%)\\
+61 ( 4.1\%)\\
+131 ( 8.9\%)\\
+18 ( 1.2\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+1477\\
+(100.0\%)\strut
+\end{minipage} \\
+Pension distributive justice &
+\begin{minipage}[t]{\linewidth}\raggedright
+1. Strongly desagree\\
+2. Desagree\\
+3. Neither agree nor desagre\\
+4. Agree\\
+5. Strongly agree\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+425 (28.8\%)\\
+704 (47.7\%)\\
+106 ( 7.2\%)\\
+219 (14.8\%)\\
+23 ( 1.6\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+1477\\
+(100.0\%)\strut
+\end{minipage} \\
+Education distributive justice &
+\begin{minipage}[t]{\linewidth}\raggedright
+1. Strongly desagree\\
+2. Desagree\\
+3. Neither agree nor desagre\\
+4. Agree\\
+5. Strongly agree\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+519 (35.1\%)\\
+764 (51.7\%)\\
+71 ( 4.8\%)\\
+112 ( 7.6\%)\\
+11 ( 0.7\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+1477\\
+(100.0\%)\strut
+\end{minipage} \\
+Market justice preferences & \begin{minipage}[t]{\linewidth}\raggedright
+Mean (sd) : 2 (0.8)\\
+min \textless{} med \textless{} max:\\
+1 \textless{} 2 \textless{} 5\\
+IQR (CV) : 1 (0.4)\strut
+\end{minipage} & 12 distinct values &
+\begin{minipage}[t]{\linewidth}\raggedright
+1477\\
+(100.0\%)\strut
+\end{minipage} \\
+\end{longtable}
+
+\emph{Independent variables}
+
+\textbf{Perception of economic inequality}: The main independent
+variable refers to the perception of economic inequality, measured
+through the perceived wage gap, which serves as an absolute indicator of
+inequality perceptions (\citeproc{ref-castillo_perception_2022}{Castillo
+et al., 2022}). This measure is derived from the gap between the
+perceived salaries of jobs at opposite ends of the occupational
+hierarchy. Specifically, it relies on the logarithm of the division
+between the perceived salary of a large-company president and that of an
+unskilled worker (\citeproc{ref-Castillo2011}{Castillo, 2011}). Higher
+values of this term indicate a greater perception of economic inequality
+between occupations located at the extremes of the status continuum.
+
+\textbf{Justification of economic inequality}: This construct is
+measured as the just earnings gap, an approach to measuring the
+justification of economic inequality based on earning differences
+considered just between high and low status occupations
+(\citeproc{ref-Castillo2011}{Castillo, 2011}). Similarly to economic
+inequality perception, this indicator relies on the logarithmic ratio
+between the just salary of a large-company president and that of an
+unskilled worker. Higher values of this term indicate a greater
+justification of economic inequality.
+
+\textbf{Meritocracy}: Meritocratic perception is operationalized through
+two components: one addressing effort and another focusing on talent
+(\citeproc{ref-young_rise_1962}{Young, 1962}). The item used to gauge
+effort is: ``In Chile, people are rewarded for their efforts,'' while
+the item for talent is: ``In Chile, people are rewarded for their
+intelligence and skills.'' In both cases, respondents indicate their
+level of agreement on a five-point Likert scale, ranging from 1
+(``strongly disagree'') to 5 (``strongly agree'').
+
+Table 2 shows the independent variables used, their response categories
+and their frequencies.
+
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3774}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3113}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2075}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1038}}@{}}
+\caption{Independent variables for the first wave
+(2016)}\label{tbl-summary2}\tabularnewline
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Label
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Stats / Values
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Freqs (\% of Valid)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Valid
+\end{minipage} \\
+\midrule\noalign{}
+\endfirsthead
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Label
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Stats / Values
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Freqs (\% of Valid)
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Valid
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+Inequality gap perception & \begin{minipage}[t]{\linewidth}\raggedright
+Mean (sd) : 3.7 (1.1)\\
+min \textless{} med \textless{} max:\\
+0.4 \textless{} 3.7 \textless{} 6.9\\
+IQR (CV) : 1.6 (0.3)\strut
+\end{minipage} & 294 distinct values &
+\begin{minipage}[t]{\linewidth}\raggedright
+1477\\
+(100.0\%)\strut
+\end{minipage} \\
+Inequality gap justification &
+\begin{minipage}[t]{\linewidth}\raggedright
+Mean (sd) : 2.1 (1.4)\\
+min \textless{} med \textless{} max:\\
+-5.5 \textless{} 2.1 \textless{} 9.9\\
+IQR (CV) : 1.7 (0.6)\strut
+\end{minipage} & 215 distinct values &
+\begin{minipage}[t]{\linewidth}\raggedright
+1477\\
+(100.0\%)\strut
+\end{minipage} \\
+People are rewarded for their efforts &
+\begin{minipage}[t]{\linewidth}\raggedright
+1. Strongly desagree\\
+2. Desagree\\
+3. Neither agree nor desagre\\
+4. Agree\\
+5. Strongly agree\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+168 (11.4\%)\\
+679 (46.0\%)\\
+258 (17.5\%)\\
+324 (21.9\%)\\
+48 ( 3.2\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+1477\\
+(100.0\%)\strut
+\end{minipage} \\
+People are rewarded for their intelligence &
+\begin{minipage}[t]{\linewidth}\raggedright
+1. Strongly desagree\\
+2. Desagree\\
+3. Neither agree nor desagre\\
+4. Agree\\
+5. Strongly agree\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+134 ( 9.1\%)\\
+604 (40.9\%)\\
+289 (19.6\%)\\
+395 (26.7\%)\\
+55 ( 3.7\%)\strut
+\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
+1477\\
+(100.0\%)\strut
+\end{minipage} \\
+\end{longtable}
+
+\emph{Controls}
+
+Sociodemographic and attitudinal variables are included to control for
+potential composition effects in the population. In terms of
+sociodemographic characteristics, we incorporate per capita household
+income quintile, educational level (1=Less than Universitary,
+2=Universitary), age (in years), and sex (1=Male, 2=Female), which have
+been shown to influence market justice preferences significantly
+(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding attitudinal
+variables, we include political identification (1=Left, 2=Center,
+3=Right, 4=No identification) and subjective social status (ranging from
+1 to 10) because they may confound the relationship between market
+justice preferences and perceptions of inequality and meritocracy
+(\citeproc{ref-schneider_poverty_2015}{Schneider \& Castillo, 2015}).
 
@@
 
+Given the data's hierarchical structure, in which observations are
+nested in survey waves, we employ longitudinal multilevel linear models
+(\citeproc{ref-singer_applied_2009}{Singer \& Willett, 2009}). In a
+panel-data framework, within-person effects capture how shifts in
+individual-level variables across waves are associated with variations
+in market justice preferences. By contrast, between-person effects focus
+on differences among individuals, explaining how long-term (or average)
+values relate to overall levels of market justice preferences.
+
+To estimate within-person effects, we use group-mean centering, where
+each respondent functions as the ``group'' (i.e., observations nested
+within persons). Meanwhile, the between-person effects are derived from
+each individual's average on these variables, calculated across the
+waves of panel data.
+
+All the analyses were conducted using R software and the \emph{lme4}
+package (\citeproc{ref-bates_fitting_2015}{Bates et al., 2015}).
+
 \section{Results}\label{results}
+
+\subsection{Descriptive}\label{descriptive}
+
+Figure~\ref{fig-alluvial} shows the annual frequencies of market justice
+preferences for healthcare, pensions, and education from 2016 to 2023.
+Each year presents stacked percentage frequencies, and the flows between
+them reflect opinion changes among the same individuals from one year to
+the next, given that we are using panel data. For instance, of the
+40.9\% who strongly disagreed with justifying inequality in healthcare
+in 2019, around 24.3\% maintained that position in 2022, while the
+remaining 16.6\% shifted toward other response categories---primarily
+moving into disagreement rather than strong disagreement. Overall, more
+than half of the respondents exhibit a high level of disagreement
+(disagree + strongly disagree) with inequality in these three social
+service areas over time. Despite this general pattern, recent waves show
+a slight decrease in disagreement and a corresponding rise in support
+for inequality. Specifically, in healthcare and education, although
+disagreement remains substantial, agreement increased from 8.9\% and
+7.6\% in 2019 to 12\% and 12.5\% in 2023, respectively. This shift is
+most evident in pensions, where the combined agree/strongly agree
+category grew by about 10 percentage points, from 16.9\% in 2016 to
+27.8\% in 2023.
+
+\begin{figure}[H]
+
+\caption{\label{fig-alluvial}Change in the justification of inequality
+in health, pensions and education over time (2016-2023)}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-alluvial-1.pdf}
+
+}
+
+\end{figure}%
+
+Regarding the main variables of this study, Figure~\ref{fig-meanchange}
+depicts their average changes over the years. We observe an increase in
+the average level of market justice preferences in the most recent
+waves, which begins in 2019. The highest average consistently appears
+for perceived economic inequality, although this variable shows a
+downward trend of roughly one point over time. Interestingly, while
+perceptions of economic inequality declined in the latest measurements
+(2022--2023), market justice preferences increased. With respect to
+economic inequality justification, there is a steady decrease in average
+levels over the years, registering the lowest mean compared to the other
+variables. The meritocracy measures remain stable, though the perception
+that individuals are rewarded for intelligence is slightly higher than
+the perception that they are rewarded for their effort.
+
+\begin{figure}[H]
+
+\caption{\label{fig-meanchange}Change in the mean of market justice
+preferences, economic inequality perception and justification, and
+meritocracy (2016-2023)}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-meanchange-1.pdf}
+
+}
+
+\end{figure}%
+
+Figure~\ref{fig-matrix} presents the correlation matrix for the main
+variables in the latest wave (2023). Overall, the coefficients range
+from low to moderate. The association between market justice preferences
+and economic inequality perception is negative but small and not
+statistically significant (\(r\) = -0.05, \(p\) \textgreater{} .05). By
+contrast, market justice preferences positively and significantly
+correlate with the justification of economic inequality (\(r\) = 0.16,
+\(p\) \textless{} .01), as well as show low, positive associations with
+the two meritocracy variables (\(r\) = 0.14, \(p\) \textless{} .01;
+\(r\) = 0.12, \(p\) \textless{} .01). Perceived economic inequality has
+a strong, positive correlation with justifying economic inequality
+(\(r\) = 0.53, \(p\) \textless{} .01), but negative and nonsignificant
+correlations with both meritocracy perceptions (\(r\) = -0.04, \(p\)
+\textgreater{} .05; \(r\) = -0.04, \(p\) \textgreater{} .05). The
+justification of economic inequality indicator also has small yet
+statistically significant positive correlations with the meritocracy
+variables (\(r\) = 0.05, \(p\) \textless{} .05; \(r\) = 0.07, \(p\)
+\textless{} .01). Finally, the two meritocracy variables exhibit a
+strong positive association with each other (\(r\) = 0.69, \(p\)
+\textless{} .01).
+
+\begin{figure}[H]
+
+\caption{\label{fig-matrix}Correlation matrix of the main variables for
+the last wave (2023)}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-matrix-1.pdf}
+
+}
+
+\end{figure}%
+
+\subsection{Multilevel models}\label{multilevel-models}
+
+Table~\ref{tbl-modelos} presents the results of the multilevel models
+estimated for market justice preferences, examining both individuals
+(within) and group-level (between) effects. The intraclass correlation
+(ICC,
+\citeproc{ref-hoxMultilevelAnalysisTechniques2017}{\textbf{hoxMultilevelAnalysisTechniques2017?}})
+from the empty model (see Supplementary Material), which decomposes the
+variance of market justice preferences, is 0.23, indicating that
+approximately 23\% of the variation is attributable to differences
+between individuals. Complementary, 77\% of the variation corresponds to
+within-individual differences over time.
+
+According to Model 1, which includes the survey waves to capture
+intertemporal variations in the dependent variable, there is a decrease
+in 2017 (\(\beta\) = -0.162, \(p\) \textless{} .001) relative to 2016,
+and similarly in 2018 (\(\beta\) = -0.015, \(p\) \textgreater{} .05) and
+2019 (\(\beta\) = -0.041, \(p\) \textgreater{} .05), although the latter
+effects are not statistically significant. In contrast, in the more
+recent waves of 2022 and 2023, there is a statistically significant
+increase in market justice preferences (\(\beta\) = 0.289, \(p\)
+\textless{} .001; \(\beta\) = 0.286, \(p\) \textless{} .001), suggesting
+a non-linear effect. To model this trajectory over time, Model 2
+incorporates time (survey waves) as a continuous variable, along with
+its quadratic term, representing the non-linear association initially
+observed in Model 1. While the linear term (survey wave) shows a
+negative association, reflecting an overall decline in market
+preferences over time, the positive quadratic term indicates a reversal
+of this pattern in the final measurement points.
+
+\begin{table}
+
+\caption{\label{tbl-modelos}Longitudinal multilevel models for market
+justice preferences}
+
+\centering{
+
+[H]
+\begin{center}
+\scalebox{1}{
+\begin{threeparttable}
+\begin{tabular}{l c c c c c c c c}
+\toprule
+ & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 & Model 8 \\
+\midrule
+Intercept                            & $1.964^{***}$  & $1.966^{***}$  & $2.103^{***}$  & $2.059^{***}$  & $1.752^{***}$  & $1.721^{***}$  & $1.541^{***}$  & $1.574^{***}$  \\
+                                     & $(0.022)$      & $(0.033)$      & $(0.046)$      & $(0.045)$      & $(0.052)$      & $(0.053)$      & $(0.095)$      & $(0.112)$      \\
+Wave (Ref.= 2016)                    &                &                &                &                &                &                &                &                \\
+                                     &                &                &                &                &                &                &                &                \\
+\quad Wave 2017                      & $-0.162^{***}$ &                &                &                &                &                &                &                \\
+                                     & $(0.028)$      &                &                &                &                &                &                &                \\
+\quad Wave 2018                      & $-0.015$       &                &                &                &                &                &                &                \\
+                                     & $(0.027)$      &                &                &                &                &                &                &                \\
+\quad Wave 2019                      & $-0.041$       &                &                &                &                &                &                &                \\
+                                     & $(0.027)$      &                &                &                &                &                &                &                \\
+\quad Wave 2022                      & $0.289^{***}$  &                &                &                &                &                &                &                \\
+                                     & $(0.028)$      &                &                &                &                &                &                &                \\
+\quad Wave 2023                      & $0.286^{***}$  &                &                &                &                &                &                &                \\
+                                     & $(0.027)$      &                &                &                &                &                &                &                \\
+Wave                                 &                & $-0.069^{***}$ & $-0.072^{***}$ & $-0.057^{**}$  & $-0.057^{**}$  & $-0.058^{**}$  & $-0.064^{***}$ & $-0.064^{***}$ \\
+                                     &                & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
+Wave$^2$                             &                & $0.017^{***}$  & $0.017^{***}$  & $0.015^{***}$  & $0.015^{***}$  & $0.015^{***}$  & $0.016^{***}$  & $0.016^{***}$  \\
+                                     &                & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      \\
+Perception inequality (WE)           &                &                & $-0.037^{***}$ & $-0.086^{***}$ & $-0.075^{***}$ & $-0.074^{***}$ & $-0.056^{***}$ & $-0.056^{***}$ \\
+                                     &                &                & $(0.009)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.011)$      & $(0.011)$      \\
+Justification inequality (WE)        &                &                &                & $0.100^{***}$  & $0.095^{***}$  & $0.094^{***}$  & $0.055^{***}$  & $0.055^{***}$  \\
+                                     &                &                &                & $(0.009)$      & $(0.009)$      & $(0.009)$      & $(0.010)$      & $(0.010)$      \\
+Merit: Effort (WE)                   &                &                &                &                & $0.107^{***}$  & $0.088^{***}$  & $0.071^{***}$  & $0.071^{***}$  \\
+                                     &                &                &                &                & $(0.009)$      & $(0.011)$      & $(0.012)$      & $(0.012)$      \\
+Merit: Talent (WE)                   &                &                &                &                &                & $0.029^{*}$    & $0.025^{*}$    & $0.025^{*}$    \\
+                                     &                &                &                &                &                & $(0.011)$      & $(0.012)$      & $(0.012)$      \\
+Perception inequality (BE)           &                &                &                &                &                &                & $-0.077^{***}$ & $-0.078^{***}$ \\
+                                     &                &                &                &                &                &                & $(0.023)$      & $(0.023)$      \\
+Justification inequality (BE)        &                &                &                &                &                &                & $0.156^{***}$  & $0.131^{***}$  \\
+                                     &                &                &                &                &                &                & $(0.021)$      & $(0.021)$      \\
+Merit: Effort (BE)                   &                &                &                &                &                &                & $0.092^{**}$   & $0.088^{**}$   \\
+                                     &                &                &                &                &                &                & $(0.033)$      & $(0.033)$      \\
+Merit: Talent (BE)                   &                &                &                &                &                &                & $-0.006$       & $-0.016$       \\
+                                     &                &                &                &                &                &                & $(0.033)$      & $(0.033)$      \\
+\midrule
+Controls                             & No             & No             & No             & No             & No             & No             & No             & Yes            \\
+AIC                                  & $20273.132$    & $20260.158$    & $20251.351$    & $20140.808$    & $20002.562$    & $20005.214$    & $19957.283$    & $19986.833$    \\
+BIC                                  & $20329.477$    & $20309.460$    & $20307.696$    & $20204.196$    & $20072.993$    & $20082.688$    & $20062.929$    & $20191.083$    \\
+Log Likelihood                       & $-10128.566$   & $-10123.079$   & $-10117.676$   & $-10061.404$   & $-9991.281$    & $-9991.607$    & $-9963.641$    & $-9964.417$    \\
+Num. obs.                            & $8460$         & $8460$         & $8460$         & $8460$         & $8460$         & $8460$         & $8460$         & $8460$         \\
+Num. groups: idencuesta              & $1681$         & $1681$         & $1681$         & $1681$         & $1681$         & $1681$         & $1681$         & $1681$         \\
+Var: idencuesta (Intercept)          & $0.174$        & $0.214$        & $0.210$        & $0.204$        & $0.179$        & $0.178$        & $0.180$        & $0.172$        \\
+Var: Residual                        & $0.528$        & $0.494$        & $0.494$        & $0.493$        & $0.490$        & $0.490$        & $0.488$        & $0.488$        \\
+Var: idencuesta ola\_num             & $$             & $0.007$        & $0.007$        & $0.007$        & $0.006$        & $0.006$        & $0.006$        & $0.006$        \\
+Cov: idencuesta (Intercept) ola\_num & $$             & $-0.018$       & $-0.017$       & $-0.018$       & $-0.015$       & $-0.015$       & $-0.016$       & $-0.016$       \\
+\bottomrule
+\end{tabular}
+\begin{tablenotes}[flushleft]
+\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}
+\end{tablenotes}
+\end{threeparttable}
+}
+\caption{}
+\label{table:coefficients}
+\end{center}
+
+}
+
+\end{table}%
+
+Models 3 to 6 incorporate the within-group effects (WE) of the primary
+independent variables, capturing how individual changes in these
+variables over time shape the dependent variable. The results in Model 3
+suggest that the within effect of perceived economic inequality is
+negative and statistically significant (\(p\) \textless{} .001).
+Specifically, each one-point increase in an individual's perception of
+economic inequality between waves is associated with a 0.037 point
+decrease in market justice preferences. By contrast, in Model 4, the
+within effect of justifying economic inequality is positive and
+significant (\(\beta\) = 0.100, \(p\) \textless{} .001), implying that a
+rise in an individual's justification of economic inequality over time
+corresponds to higher market justice preferences. Regarding meritocratic
+variables, Model 5 indicates that the perception of being rewarded for
+effort exerts a positive within effect (\(\beta\) = 0.107, \(p\)
+\textless{} .001). Similarly, Model 6 shows that viewing intelligence
+and ability (talent) as rewarded is also positively related to market
+justice preferences (\(\beta\) = 0.029, \(p\) \textless{} .05). Taken
+together, these results suggest that individuals who increasingly
+perceive meritocracy---whether through effort or talent---tend to hold
+stronger market justice preferences.
+
+When examining the between-group effects (BE) in Model 7, which capture
+differences between individuals in the average of the main variables, a
+similar pattern emerges. Individuals who perceive higher levels of
+economic inequality tend to prefer less market justice (\(\beta\) =
+-0.077, \(p\) \textless{} .001), whereas those who justify economic
+inequality more strongly exhibit higher market justice preferences
+(\(\beta\) = 0.156, \(p\) \textless{} .001). Furthermore, the
+meritocratic perception that effort is rewarded has a positive impact on
+market justice preferences (\(\beta\) = 0.092, \(p\) \textless{} .05).
+By contrast, the perception that talent is rewarded shows a negative
+coefficient (\(\beta\) = -0.006), though this relationship is not
+statistically significant at the 95\% confidence level.
+
+Model 8 introduces the control variables. The within- and
+between-effects of the main predictors retain both their direction and
+statistical significance, indicating that the associations are robust to
+adjustment. Regarding the controls, belonging to higher household-income
+quintiles is associated with stronger market-justice preferences:
+individuals in the fourth and fifth quintiles score higher than those in
+the bottom quintile (\(\beta\) = 0.093, \(p\) \textless{} .05; \(\beta\)
+= 0.147, \(p\) \textless{} .001, respectively). Respondents who did not
+report their income also express greater support for market justice
+(\(\beta\) = 0.172, \(p\) \textless{} .01). Political orientation
+matters as well: compared with those on the left, individuals who place
+themselves on the right (\(\beta\) = 0.246, \(p\) \textless{} .001) or
+who declare no political position (\(\beta\) = 0.092, \(p\) \textless{}
+.01) display higher market-justice preferences. Finally, women exhibit
+lower market-justice preferences than men (\(\beta\) = −0.080, \(p\)
+\textless{} .01).
+
+\begin{table}
+
+\caption{\label{tbl-interactions}Intereactions for within effects and
+time}
+
+\centering{
+
+\begin{verbatim}
+
+\begin{table}[h!]
+\begin{center}
+\scalebox{1}{
+\begin{threeparttable}
+\begin{tabular}{l c c c c}
+\toprule
+ & Model 9 & Model 10 & Model 11 & Model 12 \\
+\midrule
+Intercept                                    & $1.605^{***}$ & $1.679^{***}$ & $1.425^{***}$ & $1.444^{***}$ \\
+                                             & $(0.123)$     & $(0.112)$     & $(0.117)$     & $(0.117)$     \\
+Perception inequality (WE) x Wave            & $0.002$       &               &               &               \\
+                                             & $(0.004)$     &               &               &               \\
+Justification inequality (WE) x Wave         &               & $0.013^{***}$ &               &               \\
+                                             &               & $(0.003)$     &               &               \\
+Merit: Effort (WE) x Wave                    &               &               & $-0.012^{**}$ &               \\
+                                             &               &               & $(0.004)$     &               \\
+Merit: Talent (WE) x Wave                    &               &               &               & $-0.010^{*}$  \\
+                                             &               &               &               & $(0.004)$     \\
+\midrule
+Controls                                     & Yes           & Yes           & Yes           & Yes           \\
+AIC                                          & $19995.908$   & $19923.216$   & $19940.577$   & $19932.452$   \\
+BIC                                          & $20228.331$   & $20155.638$   & $20172.999$   & $20164.875$   \\
+Log Likelihood                               & $-9964.954$   & $-9928.608$   & $-9937.289$   & $-9933.226$   \\
+Num. obs.                                    & $8460$        & $8460$        & $8460$        & $8460$        \\
+Num. groups: idencuesta                      & $1681$        & $1681$        & $1681$        & $1681$        \\
+Var: idencuesta (Intercept)                  & $0.305$       & $0.198$       & $0.137$       & $0.123$       \\
+Var: idencuesta perc\_inequality             & $0.004$       & $$            & $$            & $$            \\
+Var: idencuesta ola\_num                     & $0.007$       & $0.006$       & $0.006$       & $0.006$       \\
+Cov: idencuesta (Intercept) perc\_inequality & $-0.026$      & $$            & $$            & $$            \\
+Cov: idencuesta (Intercept) ola\_num         & $-0.032$      & $-0.027$      & $-0.002$      & $-0.000$      \\
+Cov: idencuesta perc\_inequality ola\_num    & $0.004$       & $$            & $$            & $$            \\
+Var: Residual                                & $0.485$       & $0.484$       & $0.472$       & $0.471$       \\
+Var: idencuesta just\_inequality             & $$            & $0.010$       & $$            & $$            \\
+Cov: idencuesta (Intercept) just\_inequality & $$            & $-0.022$      & $$            & $$            \\
+Cov: idencuesta just\_inequality ola\_num    & $$            & $0.007$       & $$            & $$            \\
+Var: idencuesta merit\_effort                & $$            & $$            & $0.022$       & $$            \\
+Cov: idencuesta (Intercept) merit\_effort    & $$            & $$            & $-0.024$      & $$            \\
+Cov: idencuesta merit\_effort ola\_num       & $$            & $$            & $-0.005$      & $$            \\
+Var: idencuesta merit\_talent                & $$            & $$            & $$            & $0.021$       \\
+Cov: idencuesta (Intercept) merit\_talent    & $$            & $$            & $$            & $-0.023$      \\
+Cov: idencuesta merit\_talent ola\_num       & $$            & $$            & $$            & $-0.005$      \\
+\bottomrule
+\end{tabular}
+\begin{tablenotes}[flushleft]
+\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}
+\end{tablenotes}
+\end{threeparttable}
+}
+\caption{}
+\label{table:coefficients}
+\end{center}
+\end{table}
+\end{verbatim}
+
+}
+
+\end{table}%
 
@@
 
-\section{Supplementary material}\label{supplementary-material}
+\section{References}\label{references}
+
+\phantomsection\label{refs}
+\begin{CSLReferences}{1}{0}
+\bibitem[\citeproctext]{ref-akyelken_urban_2020}
+Akyelken, N. (2020). Urban conceptions of economic inequalities.
+\emph{Regional Studies}, \emph{54}(6), 863--872.
+\url{https://doi.org/10.1080/00343404.2020.1732902}
+
+\bibitem[\citeproctext]{ref-bates_fitting_2015}
+Bates, D., Mächler, M., Bolker, B., \& Walker, S. (2015). Fitting linear
+mixed-effects models using {lme4}. \emph{Journal of Statistical
+Software}, \emph{67}(1), 1--48.
+\url{https://doi.org/10.18637/jss.v067.i01}
+
+\bibitem[\citeproctext]{ref-batruch_belief_2023}
+Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F.
+(2023). Belief in {School Meritocracy} and the {Legitimization} of
+{Social} and {Income Inequality}. \emph{Social Psychological and
+Personality Science}, \emph{14}(5), 621--635.
+\url{https://doi.org/10.1177/19485506221111017}
+
+\bibitem[\citeproctext]{ref-boccardo_30_2020}
+Boccardo, G. (2020). \emph{30 a{ñ}os de privatizaciones en {Chile}: Lo
+que la pandemia revel{ó}} (Nodo XXI). Santiago.
+
+\bibitem[\citeproctext]{ref-Castillo2011}
+Castillo, J. C. (2011). Legitimacy of {Inequality} in a {Highly Unequal
+Context}: {Evidence} from the {Chilean Case}. \emph{Social Justice
+Research}, \emph{24}(4), 314--340.
+\url{https://doi.org/10.1007/s11211-011-0144-5}
+
+\bibitem[\citeproctext]{ref-castillo_perception_2022}
+Castillo, J. C., García-Castro, J.-D., \& Venegas, M. (2022). Perception
+of economic inequality: Concepts, associated factors and prospects of a
+burgeoning research agenda. \emph{International Journal of Social
+Psychology}, \emph{37}(1), 180--207.
+\url{https://doi.org/10.1080/02134748.2021.2009275}
+
+\bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
+Castillo, J. C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F.
+(2023). A {Multidimensional Approach} for {Measuring Meritocratic
+Beliefs}: {Advantages}, {Limitations} and {Alternatives} to the {ISSP
+Social Inequality Survey}. \emph{International Journal of Sociology},
+1--25. \url{https://doi.org/10.1080/00207659.2023.2274712}
+
+\bibitem[\citeproctext]{ref-castillo_percepcion_2019}
+Castillo, J. C., Miranda, D., \& Carrasco, D. (2012). Percepci{ó}n de
+{Desigualdad Econ{ó}mica} en {Chile}: {Medici{ó}n}, {Diferencias} y
+{Determinantes}. \emph{Psykhe (Santiago)}, \emph{21}(1), 99--114.
+\url{https://doi.org/10.4067/S0718-22282012000100007}
+
+\bibitem[\citeproctext]{ref-castillo_socialization_2024}
+Castillo, J. C., Salgado, M., Carrasco, K., \& Laffert, A. (2024). The
+{Socialization} of {Meritocracy} and {Market Justice Preferences} at
+{School}. \emph{Societies}, \emph{14}(11), 214.
+\url{https://doi.org/10.3390/soc14110214}
+
+\bibitem[\citeproctext]{ref-chancel_world_2022}
+Chancel, L., Piketty, T., Saez, E., \& Zucman, G. (2022). World
+inequality report 2022.
+https://bibliotecadigital.ccb.org.co/handle/11520/27510.
+
+\bibitem[\citeproctext]{ref-davis_principles_2001}
+Davis, K., \& Moore, W. E. (2001). Some {Principles} of
+{Stratification}. In \emph{Social {Stratification}, {Class}, {Race}, and
+{Gender} in {Sociological Perspective}, {Second Edition}} (2nd ed.).
+Routledge.
+
+\bibitem[\citeproctext]{ref-easterbrook_social_2021}
+Easterbrook, M. J. (2021). \emph{The social psychology of economic
+inequality} (43rd ed., Vol. 2021). UNU-WIDER.
+\url{https://doi.org/10.35188/UNU-WIDER/2021/981-5}
+
+\bibitem[\citeproctext]{ref-elsoc_estudio_2022}
+ELSOC, S. T. (2022). Estudio {Longitudinal Social} de {Chile}. Harvard
+Dataverse. \url{https://doi.org/10.7910/dvn/0kirbj}
+
+\bibitem[\citeproctext]{ref-engelhardt_what_2018}
+Engelhardt, C., \& Wagener, A. (2018). What do {Germans} think and know
+about income inequality? {A} survey experiment. \emph{Socio-Economic
+Review}, \emph{16}(4), 743--767.
+\url{https://doi.org/10.1093/ser/mwx036}
+
+\bibitem[\citeproctext]{ref-flores_top_2020}
+Flores, I., Sanhueza, C., Atria, J., \& Mayer, R. (2020). Top {Incomes}
+in {Chile}: {A Historical Perspective} on {Income Inequality},
+1964--2017. \emph{Review of Income and Wealth}, \emph{66}(4), 850--874.
+\url{https://doi.org/10.1111/roiw.12441}
+
+\bibitem[\citeproctext]{ref-garcia-castro_perceiving_2020}
+García-Castro, J. D., Rodríguez-Bailón, R., \& Willis, G. B. (2020).
+Perceiving economic inequality in everyday life decreases tolerance to
+inequality. \emph{Journal of Experimental Social Psychology}, \emph{90},
+104019. \url{https://doi.org/10.1016/j.jesp.2020.104019}
+
+\bibitem[\citeproctext]{ref-garcia-sanchez_creencias_2022}
+García-Sánchez, E., \& De Carvalho, S. (2022). Las creencias que
+justifican la desigualdad moderan la relaci{ó}n entre el estatus
+socioecon{ó}mico y el apoyo a la redistribuci{ó}n. \emph{Revista
+Internacional de Sociolog{í}a}, \emph{80}(3), e210.
+\url{https://doi.org/10.3989/ris.2022.80.3.21.29}
+
+\bibitem[\citeproctext]{ref-garcia-sanchez_vicious_2019}
+García-Sánchez, E., Van Der Toorn, J., Rodríguez-Bailón, R., \& Willis,
+G. B. (2019). The {Vicious Cycle} of {Economic Inequality}: {The Role}
+of {Ideology} in {Shaping} the {Relationship Between} {``{What Is}''}
+and {``{What Ought} to {Be}''} in 41 {Countries}. \emph{Social
+Psychological and Personality Science}, \emph{10}(8), 991--1001.
+\url{https://doi.org/10.1177/1948550618811500}
+
+\bibitem[\citeproctext]{ref-garcia-sanchez_perceptions_2018}
+García-Sánchez, E., Willis, G. B., Rodríguez-Bailón, R., Palacio Sañudo,
+J., David Polo, J., \& Rentería Pérez, E. (2018). Perceptions of
+{Economic Inequality} and {Support} for {Redistribution}: {The} role of
+{Existential} and {Utopian Standards}. \emph{Social Justice Research},
+\emph{31}(4), 335--354. \url{https://doi.org/10.1007/s11211-018-0317-6}
+
+\bibitem[\citeproctext]{ref-gimpelson_misperceiving_2018}
+Gimpelson, V., \& Treisman, D. (2018). Misperceiving inequality.
+\emph{Economics \& Politics}, \emph{30}(1), 27--54.
+\url{https://doi.org/10.1111/ecpo.12103}
+
+\bibitem[\citeproctext]{ref-immergut_it_2020}
+Immergut, E. M., \& Schneider, S. M. (2020). Is it unfair for the
+affluent to be able to purchase {``better''} healthcare? {Existential}
+standards and institutional norms in healthcare attitudes across 28
+countries. \emph{Social Science \& Medicine}, \emph{267}, 113146.
+\url{https://doi.org/10.1016/j.socscimed.2020.113146}
+
+\bibitem[\citeproctext]{ref-janmaat_subjective_2013}
+Janmaat, J. G. (2013). Subjective inequality: {A} review of
+international comparative studies on people's views about inequality.
+\emph{Archives Europeennes de Sociologie}, \emph{54}(3), 357--389.
+\url{https://doi.org/10.1017/S0003975613000209}
+
+\bibitem[\citeproctext]{ref-kluegel_social_1995a}
+Kluegel, J. R., Mason, D. S., \& Wegener, B. (Eds.). (1995).
+\emph{Social {Justice} and {Political Change}: {Public Opinion} in
+{Capitalist} and {Post-Communist States}} (1st ed.). Routledge.
+
+\bibitem[\citeproctext]{ref-kluegel_legitimation_1999}
+Kluegel, J. R., Mason, D. S., \& Wegener, B. (1999). The {Legitimation}
+of {Capitalism} in the {Postcommunist Transition}: {Public Opinion}
+about {Market Justice}, 1991-1996. \emph{European Sociological Review},
+\emph{15}(3), 251--283. Retrieved from
+\url{https://www.jstor.org/stable/522731}
+
+\bibitem[\citeproctext]{ref-kluegel_beliefs_1981}
+Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}.
+\emph{Annual Review of Sociology}, 29--56.
+
+\bibitem[\citeproctext]{ref-knell_perceptions_2020}
+Knell, M., \& Stix, H. (2020). Perceptions of inequality. \emph{European
+Journal of Political Economy}, \emph{65}, 101927.
+\url{https://doi.org/10.1016/j.ejpoleco.2020.101927}
+
+\bibitem[\citeproctext]{ref-koos_moral_2019}
+Koos, S., \& Sachweh, P. (2019). The moral economies of market
+societies: Popular attitudes towards market competition, redistribution
+and reciprocity in comparative perspective. \emph{Socio-Economic
+Review}, \emph{17}(4), 793--821.
+\url{https://doi.org/10.1093/ser/mwx045}
+
+\bibitem[\citeproctext]{ref-kuhn_eye_2011}
+Kuhn, A. (2011). In the eye of the beholder: {Subjective} inequality
+measures and individuals' assessment of market justice. \emph{European
+Journal of Political Economy}, \emph{27}(4), 625--641.
+\url{https://doi.org/10.1016/j.ejpoleco.2011.06.002}
+
+\bibitem[\citeproctext]{ref-lane_market_1986}
+Lane, R. E. (1986). Market {Justice}, {Political Justice}.
+\emph{American Political Science Review}, \emph{80}(2), 383--402.
+\url{https://doi.org/10.2307/1958264}
+
+\bibitem[\citeproctext]{ref-lee_fairness_2023}
+Lee, J.-S., \& Stacey, M. (2023). Fairness perceptions of educational
+inequality: The effects of self-interest and neoliberal orientations.
+\emph{The Australian Educational Researcher}.
+\url{https://doi.org/10.1007/s13384-023-00636-6}
+
+\bibitem[\citeproctext]{ref-lindh_public_2015}
+Lindh, A. (2015). Public {Opinion} against {Markets}? {Attitudes}
+towards {Market Distribution} of {Social Services} -- {A Comparison} of
+17 {Countries}. \emph{Social Policy \& Administration}, \emph{49}(7),
+887--910. \url{https://doi.org/10.1111/spol.12105}
+
+\bibitem[\citeproctext]{ref-mijs_stratified_2016}
+Mijs, J. (2016). Stratified {Failure}: {Educational Stratification} and
+{Students}' {Attributions} of {Their Mathematics Performance} in 24
+{Countries}. \emph{Sociology of Education}, \emph{89}(2), 137--153.
+\url{https://doi.org/10.1177/0038040716636434}
+
+\bibitem[\citeproctext]{ref-mijs_paradox_2021}
+Mijs, J. (2021). The paradox of inequality: Income inequality and belief
+in meritocracy go hand in hand. \emph{Socio-Economic Review},
+\emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}
+
+\bibitem[\citeproctext]{ref-pedersen_attitudes_2019}
+Pedersen, R. T., \& Mutz, D. C. (2019). Attitudes {Toward Economic
+Inequality}: {The Illusory Agreement}. \emph{Political Science Research
+and Methods}, \emph{7}(04), 835--851.
+\url{https://doi.org/10.1017/psrm.2018.18}
+
+\bibitem[\citeproctext]{ref-reynolds_perceptions_2014}
+Reynolds, J., \& Xian, H. (2014). Perceptions of meritocracy in the land
+of opportunity. \emph{Research in Social Stratification and Mobility},
+\emph{36}, 121--137. \url{https://doi.org/10.1016/j.rssm.2014.03.001}
+
+\bibitem[\citeproctext]{ref-sandel_tyranny_2020}
+Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the
+common good?} (First edition). New York: {Farrar, Straus and Giroux}.
+
+\bibitem[\citeproctext]{ref-schneider_poverty_2015}
+Schneider, S. M., \& Castillo, J. C. (2015). Poverty {Attributions} and
+the {Perceived Justice} of {Income Inequality} : {A Comparison} of
+{East} and {West Germany}.
+\url{https://doi.org/10.1177/0190272515589298}
+
+\bibitem[\citeproctext]{ref-schroder_income_2017}
+Schröder, M. (2017). Is {Income Inequality Related} to {Tolerance} for
+{Inequality}? \emph{Social Justice Research}, \emph{30}(1), 23--47.
+\url{https://doi.org/10.1007/s11211-016-0276-8}
+
+\bibitem[\citeproctext]{ref-singer_applied_2009}
+Singer, J. D., \& Willett, J. B. (2009). \emph{Applied longitudinal data
+analysis: Modeling change and event occurence}. New York: Oxford
+University Press, Incorporated.
+
+\bibitem[\citeproctext]{ref-svallfors_political_2007}
+Svallfors, S. (Ed.). (2007). \emph{The {Political Sociology} of the
+{Welfare State}: {Institutions}, {Social Cleavages}, and {Orientations}}
+(1st ed.). Stanford University Press.
+\url{https://doi.org/10.2307/j.ctvr0qv0q}
+
+\bibitem[\citeproctext]{ref-trump_income_2018}
+Trump, K.-S. (2018). Income {Inequality Influences Perceptions} of
+{Legitimate Income Differences}. \emph{British Journal of Political
+Science}, \emph{48}(4), 929--952.
+\url{https://doi.org/10.1017/S0007123416000326}
+
+\bibitem[\citeproctext]{ref-wilson_role_2003}
+Wilson, C. (2003). The {Role} of a {Merit Principle} in {Distributive
+Justice}. \emph{The Journal of Ethics}, \emph{7}(3), 277--314.
+\url{https://doi.org/10.1023/A:1024667228488}
+
+\bibitem[\citeproctext]{ref-young_rise_1962}
+Young, M. (1962). \emph{The rise of the meritocracy}. Baltimore: Penguin
+Books.
+
+\end{CSLReferences}
 

### 2025-04-25 04-analisis.qmd
Up paper and link

@@
 #| out-width: '100%'
-#| fig-asp: 1
+#| fig-asp: 1.2
 #| fig-cap: "Change in the justification of inequality in health, pensions and education over time (2016-2023)"
@@
   labs(y = "%",
-       x = "Wave",
+       x = NULL,
        fill = NULL,
@@
 #| out-width: '100%'
-#| fig-asp: 1
+#| fig-asp: 0.8
 #| fig-cap: "Change in the mean of market justice preferences, economic inequality perception and justification, and meritocracy (2016-2023)"
@@
 #| results: asis
+
 library(srvyr)
@@
   ) %>% 
-  mutate(variable = case_when(variable == "mjp" ~ "Market justice preferences",
+  mutate(variable = case_when(variable == "mjp" ~ "Market justice",
                               variable == "perc_inequality"  ~ "Inequality perception",
@@
                               variable == "merit_talent"     ~ "Merit: Talent"),
-         variable = factor(variable, levels = c("Market justice preferences",
+         variable = factor(variable, levels = c("Market justice",
                                       "Inequality perception",
@@
 #| label: fig-matrix
-#| out-width: '100%'
+#| out-width: '90%'
 #| fig-asp: 1
@@
 @tbl-modelos presents the results of the multilevel models estimated for market justice preferences, examining both individuals (within) and group-level (between) effects.
-The intraclass correlation [ICC, @hoxMultilevelAnalysisTechniques2017] from the empty model (see Supplementary Material), which decomposes the variance of market justice preferences, is 0.23, indicating that approximately 23% of the variation is attributable to differences between individuals.
+The intraclass correlation [@hoxMultilevelAnalysisTechniques2017a] from the empty model (see Supplementary Material), which decomposes the variance of market justice preferences, is 0.23, indicating that approximately 23% of the variation is attributable to differences between individuals.
 Complementary, 77% of the variation corresponds to within-individual differences over time.
@@
 
-```{r}
+```{r echo=FALSE, results='asis'}
 #| label: tbl-modelos
@@
 #| tbl-cap-location: top
-#| results: asis
-#| echo: false
+
+omit <- "(educ)|(quintil1)|(ess)|(ideo)|(sex)|(age)"
 
@@
                stars = c(0.05, 0.01, 0.001),
+               omit.coef = omit,
                custom.coef.map = ccoef,
@@
                custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
-               threeparttable = T,
                leading.zero = T,
-               float.pos = "H",
                use.packages = F,
-               booktabs = T,
-               scalebox = 1,
+               booktabs = F,
+               scalebox = 0.75,
+               include.loglik = FALSE,
+               include.aic = FALSE,
+               center = T,
                custom.gof.rows = list("Controls"=c(rep("No",7), "Yes")))
@@
 
-```{r}
+@tbl-interactions displays
+
+```{r echo=FALSE, results='asis'}
 #| label: tbl-interactions
@@
                custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
-               threeparttable = T,
                leading.zero = T,
-               float.pos = "h!",
                use.packages = F,
-               booktabs = T,
-               scalebox = 1,
+               booktabs = F,
+               scalebox = 0.95,
+               include.loglik = FALSE,
+               include.aic = FALSE,
+               center = T,
                custom.gof.rows = list("Controls"=c(rep("Yes",4))))
+
 ```

### 2025-04-25 paper.qmd
Up paper and link

@@
 abstract: | 
-  My abstract. \newline
-  **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
+  This study examines the interplay between perceptions of economic inequality and meritocracy in shaping preferences for market justice in Chile from 2016 to 2023.
+Economic inequality has intensified globally, raising concerns about its social impacts and the legitimacy of market-based distributions of resources.
+By utilizing longitudinal data from the Chilean Longitudinal Social Survey (ELSOC), the research investigates how subjective assessments of inequality influence attitudes towards social justice, focusing on the extent to which individuals believe access to essential services like healthcare, education, and pensions should depend on personal merit.
+The findings suggest that as perceptions of economic inequality rise, individuals tend to favor redistributive policies, although those with strong meritocratic beliefs are more likely to justify existing inequalities.
+The analysis also highlights the effect of major social movements in Chile during this period, which have shifted public discourse and potentially altered perceptions of inequality and meritocracy.
+This paper contributes to the understanding of how individual beliefs about fairness and equity evolve within a socio-political context characterized by high inequality and extensive free-market principles.
+  \newline
+  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences
 link-citations: true
@@
 fig-width: 7.5
-header-includes:
-  - \usepackage{threeparttable}
 ---

### 2025-04-25 paper.tex
Up paper and link

@@
 \fi
+
+\usepackage{color}
+\usepackage{fancyvrb}
+\newcommand{\VerbBar}{|}
+\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
+\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
+% Add ',fontsize=\small' for more characters per line
+\usepackage{framed}
+\definecolor{shadecolor}{RGB}{241,243,245}
+\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
+\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.40,0.45,0.13}{#1}}
+\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\BuiltInTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\CharTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
+\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
+\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\ExtensionTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.28,0.35,0.67}{#1}}
+\newcommand{\ImportTok}[1]{\textcolor[rgb]{0.00,0.46,0.62}{#1}}
+\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\NormalTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.68,0.00,0.00}{#1}}
+\newcommand{\RegionMarkerTok}[1]{\textcolor[rgb]{0.00,0.23,0.31}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{#1}}
+\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\StringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.07,0.07,0.07}{#1}}
+\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.13,0.47,0.30}{#1}}
+\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.37,0.37,0.37}{\textit{#1}}}
 
@@
 
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 \usepackage[noblocks]{authblk}
@@
 \renewcommand\Affilfont{\small}
-\usepackage{threeparttable}
 \makeatletter
@@
 \begin{abstract}
-My abstract. \newline \textbf{Keywords}: meritocracy, social inequality,
-inequality justification, COVID-19
+This study examines the interplay between perceptions of economic
+inequality and meritocracy in shaping preferences for market justice in
+Chile from 2016 to 2023. Economic inequality has intensified globally,
+raising concerns about its social impacts and the legitimacy of
+market-based distributions of resources. By utilizing longitudinal data
+from the Chilean Longitudinal Social Survey (ELSOC), the research
+investigates how subjective assessments of inequality influence
+attitudes towards social justice, focusing on the extent to which
+individuals believe access to essential services like healthcare,
+education, and pensions should depend on personal merit. The findings
+suggest that as perceptions of economic inequality rise, individuals
+tend to favor redistributive policies, although those with strong
+meritocratic beliefs are more likely to justify existing inequalities.
+The analysis also highlights the effect of major social movements in
+Chile during this period, which have shifted public discourse and
+potentially altered perceptions of inequality and meritocracy. This
+paper contributes to the understanding of how individual beliefs about
+fairness and equity evolve within a socio-political context
+characterized by high inequality and extensive free-market principles.
+\newline \textbf{Keywords}: Economic inequality, meritocracy, market
+justice, Chile, public preferences
 \end{abstract}
@@
 \subsection{This study (hypotheses)}\label{this-study-hypotheses}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\#| include: false}
+\CommentTok{\#| echo: false}
+
+\ControlFlowTok{if}\NormalTok{ (}\SpecialCharTok{!} \FunctionTok{require}\NormalTok{(}\StringTok{"pacman"}\NormalTok{)) }\FunctionTok{install.packages}\NormalTok{(}\StringTok{"pacman"}\NormalTok{)}
+
+\NormalTok{pacman}\SpecialCharTok{::}\FunctionTok{p\_load}\NormalTok{(tidyverse, }
+\NormalTok{               sjmisc, }
+\NormalTok{               sjPlot, }
+\NormalTok{               lme4, }
+\NormalTok{               here, }
+\NormalTok{               performance,}
+\NormalTok{               influence.ME, }
+\NormalTok{               marginaleffects,}
+\NormalTok{               MLMusingR,}
+\NormalTok{               texreg, }
+\NormalTok{               ggdist,}
+\NormalTok{               misty,}
+\NormalTok{               kableExtra,}
+\NormalTok{               ggalluvial, }
+\NormalTok{               shadowtext,}
+\NormalTok{               MetBrewer,}
+\NormalTok{               patchwork,}
+\NormalTok{               sjlabelled,}
+\NormalTok{               summarytools)}
+
+\FunctionTok{options}\NormalTok{(}\AttributeTok{scipen=}\DecValTok{999}\NormalTok{)}
+\FunctionTok{rm}\NormalTok{(}\AttributeTok{list =} \FunctionTok{ls}\NormalTok{())}
+\end{Highlighting}
+\end{Shaded}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-matrix-1.pdf}
+\includegraphics[width=0.9\textwidth,height=\textheight]{paper_files/figure-pdf/fig-matrix-1.pdf}
 
@@
 (within) and group-level (between) effects. The intraclass correlation
-(ICC,
-\citeproc{ref-hoxMultilevelAnalysisTechniques2017}{\textbf{hoxMultilevelAnalysisTechniques2017?}})
+(\citeproc{ref-hoxMultilevelAnalysisTechniques2017a}{Hox et al., 2017})
 from the empty model (see Supplementary Material), which decomposes the
@@
 
-[H]
 \begin{center}
-\scalebox{1}{
-\begin{threeparttable}
+\scalebox{0.75}{
 \begin{tabular}{l c c c c c c c c}
-\toprule
+\hline
  & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 & Model 8 \\
-\midrule
+\hline
 Intercept                            & $1.964^{***}$  & $1.966^{***}$  & $2.103^{***}$  & $2.059^{***}$  & $1.752^{***}$  & $1.721^{***}$  & $1.541^{***}$  & $1.574^{***}$  \\
@@
                                      &                &                &                &                &                &                & $(0.033)$      & $(0.033)$      \\
-\midrule
+\hline
 Controls                             & No             & No             & No             & No             & No             & No             & No             & Yes            \\
-AIC                                  & $20273.132$    & $20260.158$    & $20251.351$    & $20140.808$    & $20002.562$    & $20005.214$    & $19957.283$    & $19986.833$    \\
 BIC                                  & $20329.477$    & $20309.460$    & $20307.696$    & $20204.196$    & $20072.993$    & $20082.688$    & $20062.929$    & $20191.083$    \\
-Log Likelihood                       & $-10128.566$   & $-10123.079$   & $-10117.676$   & $-10061.404$   & $-9991.281$    & $-9991.607$    & $-9963.641$    & $-9964.417$    \\
 Num. obs.                            & $8460$         & $8460$         & $8460$         & $8460$         & $8460$         & $8460$         & $8460$         & $8460$         \\
@@
 Cov: idencuesta (Intercept) ola\_num & $$             & $-0.018$       & $-0.017$       & $-0.018$       & $-0.015$       & $-0.015$       & $-0.016$       & $-0.016$       \\
-\bottomrule
+\hline
+\multicolumn{9}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
 \end{tabular}
-\begin{tablenotes}[flushleft]
-\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}
-\end{tablenotes}
-\end{threeparttable}
 }
@@
 
+Table~\ref{tbl-interactions} displays
+
 \begin{table}
@@
 
-\begin{verbatim}
-
-\begin{table}[h!]
 \begin{center}
-\scalebox{1}{
-\begin{threeparttable}
+\scalebox{0.95}{
 \begin{tabular}{l c c c c}
-\toprule
+\hline
  & Model 9 & Model 10 & Model 11 & Model 12 \\
-\midrule
+\hline
 Intercept                                    & $1.605^{***}$ & $1.679^{***}$ & $1.425^{***}$ & $1.444^{***}$ \\
@@
                                              &               &               &               & $(0.004)$     \\
-\midrule
+\hline
 Controls                                     & Yes           & Yes           & Yes           & Yes           \\
-AIC                                          & $19995.908$   & $19923.216$   & $19940.577$   & $19932.452$   \\
 BIC                                          & $20228.331$   & $20155.638$   & $20172.999$   & $20164.875$   \\
-Log Likelihood                               & $-9964.954$   & $-9928.608$   & $-9937.289$   & $-9933.226$   \\
 Num. obs.                                    & $8460$        & $8460$        & $8460$        & $8460$        \\
@@
 Cov: idencuesta merit\_talent ola\_num       & $$            & $$            & $$            & $-0.005$      \\
-\bottomrule
+\hline
+\multicolumn{5}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
 \end{tabular}
-\begin{tablenotes}[flushleft]
-\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}
-\end{tablenotes}
-\end{threeparttable}
 }
@@
 \end{center}
-\end{table}
-\end{verbatim}
 
@@
 
+\bibitem[\citeproctext]{ref-hoxMultilevelAnalysisTechniques2017a}
+Hox, J. J., Moerbeek, M., \& Van de Schoot, R. (2017). \emph{Multilevel
+{Analysis}: {Techniques} and {Applications}}.
+
 \bibitem[\citeproctext]{ref-immergut_it_2020}

### 2025-04-28 01-introduction.qmd
review introduction and antecedents

@@
 
-Meritocracy posits that inequality is an inherent feature of societies but can be legitimized through fairer principles, such as effort and talent [@young_rise_1962; @davis_principles_2001].
-Previous studies have demonstrated that people with stronger meritocratic beliefs tend to perceive less inequality as they attribute economic differences to individual achievements [@mijs_paradox_2021; @wilson_role_2003] and also to justify more inequality since these beliefs are associated with attitudes that maintain the legitimization of social differences [@batruch_belief_2023].
+Meritocracy posits that inequality can be legitimized through distributive criteria such as effort and talent [@young_rise_1962; @davis_principles_2001].
+Previous studies have demonstrated that people with stronger meritocratic beliefs tend to perceive less inequality as they attribute economic differences to individual achievements [@mijs_paradox_2021; @wilson_role_2003] and also to justify more inequality as these beliefs are associated with attitudes that  legitimize status differences [@batruch_belief_2023].
 In a context like Chile, where the distribution of goods and services is predominantly governed by market logics strongly introduced during the military dictatorship (1973-1989) [@boccardo_30_2020], these beliefs can play a crucial role in the acceptance of social inequalities.

### 2025-04-28 02-antecedents.qmd
review introduction and antecedents

@@
 
-## Justification of market inequality
+## The justification of market inequality
 
@@
 Kluegel and Smith [-@kluegel_beliefs_1981] and later Kluegel et al.
-[-@kluegel_legitimation_1999] laid the groundwork for measuring beliefs in market justice through survey items that assess whether people support economic inequality when it reflects personal effort and skill.
-Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, or pensions.
+[-@kluegel_legitimation_1999] laid the groundwork for measuring beliefs in market justice through survey items that assess whether people support economic inequality when it reflects personal effort and skills.
+Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, and/or pensions.
 For example, Knesebeck et al.
@@
 
+Perceptions of inequality have been associated to attitudes about market justice[@lindh_public_2015; @kluegel_social_1995a].
+Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011].
+In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].
+
 Perceptions of inequality refer to individuals' subjective evaluations of the extent, causes, and consequences of income and wealth disparities.
@@
 
-Importantly, perceptions of inequality have been linked to attitudes about market justice---the belief that goods and services should be allocated according to individual contribution, competition, or willingness to pay [@lindh_public_2015; @kluegel_social_1995a].
-Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011].
-In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].
-
 ## Perception of meritocracy
 
-Meritocracy constitutes a central ideological framework for legitimizing social inequality.
+Meritocracy constitutes a central ideological framework for legitimizing different types od social inequality, for instance through market justice beliefs.
 Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions.
 As initially conceptualized by Michael Young [-@young_rise_1962], the term was meant to critique a system in which merit-based stratification becomes a new form of inequality.
-However, over time, meritocracy has been widely adopted in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2021; @sandel_tyranny_2020].
+However, over time, meritocracy has been widely supported in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2021; @sandel_tyranny_2020].
 From a sociological perspective, the belief in meritocracy is more than a cognitive assessment; it reflects a moral lens through which individuals interpret inequality.
@@
 
-## This study (hypotheses)
+## This study
+
+Building upon the previous literature, this study proposes that attitudes toward market justice are shaped by a dynamic interplay between individuals' perceptions of economic inequality and their beliefs in meritocracy.
+Specifically, we argue that both perceptions independently and interactively influence the extent to which individuals endorse market-based distributions of social goods and services in Chile.
+
+First, consistent with previous findings, we expect that a lower perception of economic inequality will be associated with stronger market justice preferences.
+When individuals perceive smaller economic gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing acceptance of outcomes based on competition and ability to pay.
+Conversely, a heightened perception of inequality may erode confidence in market fairness, weakening support for market-based distribution.
+This relationship is particularly relevant in the context of Chile, where the neoliberal economic model has been a dominant force shaping public attitudes toward inequality and justice.
+
+Second, stronger meritocratic beliefs are expected to be positively associated with market justice preferences.
+Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit.
+This aligns with the idea that meritocratic beliefs serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit.
+This is particularly relevant in contexts where neoliberal ideologies dominate, as they often emphasize individual responsibility and competition as the basis for social order.
+
+Third, we propose that perceptions of meritocracy and perceptions of economic inequality interact in shaping market justice attitudes.
+Specifically, we argue that the legitimizing effect of meritocratic beliefs on market justice preferences is moderated by perceptions of economic inequality: when perceived inequality is low, the positive association between meritocratic beliefs and market justice preferences will be stronger.
+However, when perceived inequality is high, this association will weaken, as greater awareness of large economic gaps may challenge the view that outcomes are purely merit-based.
+
+Additionally, this study examines whether these relationships vary over time in response to significant social and political events.
+In particular, we explore whether the political outburst of 2019 and the subsequent constitutional processes in Chile have altered how perceptions of inequality and meritocracy shape market justice attitudes.
+We expect that after these critical events, perceptions of inequality may have a stronger negative effect on market justice preferences, reflecting increased societal questioning of market-based allocation mechanisms.
+
+Based on these arguments, we propose the following hypotheses:
+
+H1: Lower perceived economic inequality is associated with stronger market justice preferences.
+
+H2: Stronger meritocratic beliefs are associated with stronger market justice preferences.
+
+H3: The positive association between meritocratic beliefs and market justice preferences is moderated by perceived economic inequality; specifically, this association is weaker when perceived economic inequality is high.
+
+H4: The effects described in H1–H3 are attenuated after major social mobilizations (2019–2022), reflecting increased critical views of the market's role in allocating social goods.

### 2025-04-28 03-methods.qmd
methods-jc review

@@
 
-*Dependent variable*
-
 **Market justice preferences**: The dependent variable in this study is preferences for market justice.
@@
 
-*Independent variables*
+**Perception of economic inequality**: The main independent variable refers to the perception of economic inequality, measured through the perceived wage gap.
+[@castillo_perception_2022] [acá citar a Hadler, Gijsberts, Castillo 2009 (CEP)].
+This measure is derived from the salary gap between the perceived salaries of jobs at opposite ends of the occupational hierarchy.
+Specifically, it relies on the division between the perceived salary of a large-company president and that of an unskilled worker [@Castillo2011].
+Higher values of this term indicate a greater perception of economic inequality between occupations located at the extremes of the status continuum.
+This measure includes a logarithmic term in order to adjust income magnitudes (usually fewer cases with high income):
 
-**Perception of economic inequality**: The main independent variable refers to the perception of economic inequality, measured through the perceived wage gap, which serves as an absolute indicator of inequality perceptions [@castillo_perception_2022].
-This measure is derived from the gap between the perceived salaries of jobs at opposite ends of the occupational hierarchy.
-Specifically, it relies on the logarithm of the division between the perceived salary of a large-company president and that of an unskilled worker [@Castillo2011].
-Higher values of this term indicate a greater perception of economic inequality between occupations located at the extremes of the status continuum.
+$$
+\text{perceived wage gap} = \log_{10}\left(\frac{\text{perceived salary of a large-company president}}{\text{perceived salary of an unskilled worker}}\right)
+$$
 
-**Justification of economic inequality**: This construct is measured as the just earnings gap, an approach to measuring the justification of economic inequality based on earning differences considered just between high and low status occupations [@Castillo2011].
-Similarly to economic inequality perception, this indicator relies on the logarithmic ratio between the just salary of a large-company president and that of an unskilled worker.
-Higher values of this term indicate a greater justification of economic inequality.
-
-**Meritocracy**: Meritocratic perception is operationalized through two components: one addressing effort and another focusing on talent [@young_rise_1962].
-The item used to gauge effort is: "In Chile, people are rewarded for their efforts," while the item for talent is: "In Chile, people are rewarded for their intelligence and skills." In both cases, respondents indicate their level of agreement on a five-point Likert scale, ranging from 1 ("strongly disagree") to 5 ("strongly agree").
+**Perception of Meritocracy**: this variable is operationalized through two components, namely effort and talent [@young_rise_1962].
+The item used to gauge effort is: "In Chile, people are rewarded for their efforts," while the item for talent is: "In Chile, people are rewarded for their intelligence and skills".
+In both cases, respondents indicate their level of agreement on a five-point Likert scale, ranging from 1 ("strongly disagree") to 5 ("strongly agree").
 
@@
 Sociodemographic and attitudinal variables are included to control for potential composition effects in the population.
-In terms of sociodemographic characteristics, we incorporate per capita household income quintile, educational level (1=Less than Universitary, 2=Universitary), age (in years), and sex (1=Male, 2=Female), which have been shown to influence market justice preferences significantly [@lindh_public_2015; @castillo_socialization_2024].
-Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) because they may confound the relationship between market justice preferences and perceptions of inequality and meritocracy [@schneider_poverty_2015].
+In terms of sociodemographic characteristics, we incorporate per capita household income quintile, educational level (1=Less than Universitary, 2=Universitary), age (in years), and sex (1=Male, 2=Female), which have been previously shown to influence market justice preferences significantly [@lindh_public_2015; @castillo_socialization_2024].
+Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) as they may confound the relationship between market justice preferences and perceptions of inequality and meritocracy [@schneider_poverty_2015].
 

### 2025-04-28 01-introduction.qmd
referencias intro y antecedentes

@@
 
-Perception of economic inequality has been associated with important outcomes such as redistributive preferences, justification of inequality, and legitimacy of the economic system (Castillo et al., 2022 \[acá citar gente de Granada\]).
+Perception of economic inequality has been associated with important outcomes such as redistributive preferences, justification of inequality, and legitimacy of the economic system [@castillo_perception_2022; @garcia-sanchez_perceptions_2018; @garcia-sanchez_attitudes_2020].
 Research shows that greater perceived inequality fosters stronger preferences for redistribution, regardless of actual inequality levels [@garcia-sanchez_vicious_2019; @gimpelson_misperceiving_2018].

### 2025-04-28 02-antecedents.qmd
referencias intro y antecedentes

@@
 Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, and/or pensions.
-For example, Knesebeck et al.
-(2016) and Immergut and Schneider [-@immergut_it_2020] examine whether citizens consider it fair that individuals with higher incomes can access better healthcare, while Lee and Stacey [-@lee_fairness_2023] apply a similar method in the context of education in Australia.
+For example, @vondemknesebeck_are_2016 and Immergut and Schneider [-@immergut_it_2020] examine whether citizens consider it fair that individuals with higher incomes can access better healthcare, while Lee and Stacey [-@lee_fairness_2023] apply a similar method in the context of education in Australia.
 These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts.
@@
 
-Perceptions of inequality have been associated to attitudes about market justice[@lindh_public_2015; @kluegel_social_1995a].
+Perceptions of inequality have been associated to attitudes about market justice [@lindh_public_2015; @kluegel_social_1995a].
 Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011].
@@
 Among the different approaches to measuring perceived inequality, one of the most widely used is the estimation of wage gaps between occupational extremes, such as between a CEO and a manual worker.
-This type of item provides a concrete frame that respondents can relate to more easily than abstract questions about national income distribution [@castillo_percepcion_2019; @easterbrook_social_2021; Willis et al., 2015].
+This type of item provides a concrete frame that respondents can relate to more easily than abstract questions about national income distribution [@castillo_percepcion_2019; @easterbrook_social_2021; @willis_legitimacy_2015].
 While it enables the estimation of inequality using simple heuristics, this method is not without challenges.
 For instance, people often lack reliable knowledge about the earnings of those at the top of the income ladder, which leads to high variability in responses and the use of biased mental shortcuts [@knell_perceptions_2020].
-Despite this, perceived wage gaps are strong predictors of political attitudes[@pedersen_attitudes_2019; @garcia-sanchez_perceptions_2018], making them a valuable tool for understanding public responses to economic disparities.
+Despite this, perceived wage gaps are strong predictors of political attitudes [@pedersen_attitudes_2019; @garcia-sanchez_perceptions_2018], making them a valuable tool for understanding public responses to economic disparities.
 
 Another relevant critique of this approach lies in its conflation of different psychological constructs.
-Many surveys assess perceived inequality through Likert-type items that ask respondents to agree or disagree with statements, such as "income differences are too large," which captures general concern or discomfort rather than a specific perception [@Castillo2011; @garcia-sanchez_vicious_2019].These items mix cognitive estimations with affective evaluations, complicating the interpretation of what respondents perceive versus what they morally reject.
+Many surveys assess perceived inequality through Likert-type items that ask respondents to agree or disagree with statements, such as "income differences are too large," which captures general concern or discomfort rather than a specific perception [@Castillo2011; @garcia-sanchez_vicious_2019].
+These items mix cognitive estimations with affective evaluations, complicating the interpretation of what respondents perceive versus what they morally reject.
 As a result, the conceptual clarity between perceived inequality and inequality aversion remains blurred in many empirical studies.
-To address this limitation, recent work has emphasized the need to distinguish between absolute and comparative measures, as well as between ideal and actual estimates of economic gaps [@garcia-sanchez_creencias_2022; Auspurg et al., 2017].
+To address this limitation, recent work has emphasized the need to distinguish between absolute and comparative measures, as well as between ideal and actual estimates of economic gaps [@garcia-sanchez_creencias_2022; @auspurg_why_2017].
 
@@
 These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they personally endorse meritocratic principles.
-This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice (Tejero-Peregrina et al., 2025).
-Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity.
+This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025] Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity.
 Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_percepcion_2019; @castillo_socialization_2024].
@@
 In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups.
-Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor (Hoyt & Burnette, 2023).
+Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023].
 This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds.
-Moreover, Busemeyer, Abrassart, and Nezi (2019) argue that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures.
+Moreover, @busemeyer_positive_2021 argue that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures.
 This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.

### 2025-04-28 02-antecedents.qmd
Update results, method and render AL

@@
 
-H1: Lower perceived economic inequality is associated with stronger market justice preferences.
+$H1$: Greater perceived economic inequality is associated with stronger market justice preferences.
 
-H2: Stronger meritocratic beliefs are associated with stronger market justice preferences.
+$H2$: Stronger meritocratic beliefs are associated with stronger market justice preferences.
 
-H3: The positive association between meritocratic beliefs and market justice preferences is moderated by perceived economic inequality; specifically, this association is weaker when perceived economic inequality is high.
+$H3$: The positive association between meritocratic beliefs and market justice preferences is moderated by perceived economic inequality; specifically, this association is weaker when perceived economic inequality is high.
 
-H4: The effects described in H1–H3 are attenuated after major social mobilizations (2019–2022), reflecting increased critical views of the market's role in allocating social goods.
+$H4$: The effects described in H1–H3 are attenuated after major social mobilizations (2019–2022), reflecting increased critical views of the market's role in allocating social goods.

### 2025-04-28 03-methods.qmd
Update results, method and render AL

@@
          perc_inequality, 
-         just_inequality,
          merit_effort, 
@@
 **Perception of economic inequality**: The main independent variable refers to the perception of economic inequality, measured through the perceived wage gap.
-[@castillo_perception_2022] [acá citar a Hadler, Gijsberts, Castillo 2009 (CEP)].
+[@castillo_cual_2009; @hadler_why_2005; @gijsberts_thelegitimation_1999].
 This measure is derived from the salary gap between the perceived salaries of jobs at opposite ends of the occupational hierarchy.
@@
   filter(ola == 1) %>% 
-  select(perc_inequality, just_inequality, merit_effort, merit_talent)
+  select(perc_inequality, merit_effort, merit_talent)
 

### 2025-04-28 04-analisis.qmd
Up links
Update results, method and render AL

@@
 #| label: packages
-#| include: true
 #| echo: false
+#| include: false
 
@@
 Each year presents stacked percentage frequencies, and the flows between them reflect opinion changes among the same individuals from one year to the next, given that we are using panel data.
-For instance, of the 40.9% who strongly disagreed with justifying inequality in healthcare in 2019, around 24.3% maintained that position in 2022, while the remaining 16.6% shifted toward other response categories—primarily moving into disagreement rather than strong disagreement.
+For instance, of the 40.8% who strongly disagreed with justifying inequality in healthcare in 2019, around 24.3% maintained that position in 2022, while the remaining 16.5% shifted toward other response categories—primarily moving into disagreement rather than strong disagreement.
 Overall, more than half of the respondents exhibit a high level of disagreement (disagree + strongly disagree) with inequality in these three social service areas over time.
 Despite this general pattern, recent waves show a slight decrease in disagreement and a corresponding rise in support for inequality.
-Specifically, in healthcare and education, although disagreement remains substantial, agreement increased from 8.9% and 7.6% in 2019 to 12% and 12.5% in 2023, respectively.
-This shift is most evident in pensions, where the combined agree/strongly agree category grew by about 10 percentage points, from 16.9% in 2016 to 27.8% in 2023.
+Specifically, in healthcare and education, although disagreement remains substantial, agreement (agree + strongly agree) increased from 7.4% and 7.2% in 2019 to 13.1% and 14.2% in 2023, respectively.
+This shift is most evident in pensions, where the combined agree/strongly agree category grew by about 10 percentage points, from 16.9% in 2016 to 28% in 2023.
 
@@
 #| fig-asp: 1.2
-#| fig-cap: "Change in the justification of inequality in health, pensions and education over time (2016-2023)"
+#| fig-cap: "Change in the justification of inequality in healthcare, pensions and education over time (2016-2023)"
 #| fig-cap-location: top
@@
        fill = NULL,
-       title = "a. Health") +
+       title = "a. Healthcare") +
   theme_ggdist() +
@@
        fill = NULL,
-       title = "b. Pension") +
+       title = "b. Pensions") +
   theme_ggdist() +
@@
 Interestingly, while perceptions of economic inequality declined in the latest measurements (2022–2023), market justice preferences increased.
-With respect to economic inequality justification, there is a steady decrease in average levels over the years, registering the lowest mean compared to the other variables.
 The meritocracy measures remain stable, though the perception that individuals are rewarded for intelligence is slightly higher than the perception that they are rewarded for their effort.
@@
 #| fig-asp: 0.8
-#| fig-cap: "Change in the mean of market justice preferences, economic inequality perception and justification, and meritocracy (2016-2023)"
+#| fig-cap: "Change in the mean of market justice preferences, economic inequality perception, and meritocracy (2016-2023)"
 #| fig-cap-location: top
@@
 df_pond %>% 
-  select(ola, mjp, perc_inequality, just_inequality, merit_effort, merit_talent) %>% 
-  mutate_at(.vars = 2:6, .funs = ~ as.numeric(.)) %>% 
+  select(ola, mjp, perc_inequality, merit_effort, merit_talent) %>% 
+  mutate_at(.vars = 2:5, .funs = ~ as.numeric(.)) %>% 
   group_by(ola) %>% 
@@
   ) %>% 
-  mutate(variable = case_when(variable == "mjp" ~ "Market justice",
+  mutate(variable = case_when(variable == "mjp" ~ "Market justice preferences",
                               variable == "perc_inequality"  ~ "Inequality perception",
-                              variable == "just_inequality"  ~ "Inequality justification",
                               variable == "merit_effort"     ~ "Merit: Effort",
                               variable == "merit_talent"     ~ "Merit: Talent"),
-         variable = factor(variable, levels = c("Market justice",
+         variable = factor(variable, levels = c("Market justice preferences",
                                       "Inequality perception",
-                                      "Inequality justification",
                                       "Merit: Effort",
@@
                                 "#0571B0", # azul intermedio
-                                "#b3b3b3", # gris
                                 "#F4A582", # salmón
@@
                                          )) +
-  scale_shape_manual(values=c(15, 16, 17, 18, 4)) +
+  scale_shape_manual(values=c(15, 16, 17, 18)) +
   labs(y = "Mean",
@@
 Overall, the coefficients range from low to moderate.
-The association between market justice preferences and economic inequality perception is negative but small and not statistically significant ($r$ = -0.05, $p$ > .05).
-By contrast, market justice preferences positively and significantly correlate with the justification of economic inequality ($r$ = 0.16, $p$ < .01), as well as show low, positive associations with the two meritocracy variables ($r$ = 0.14, $p$ < .01; $r$ = 0.12, $p$ < .01).
-Perceived economic inequality has a strong, positive correlation with justifying economic inequality ($r$ = 0.53, $p$ < .01), but negative and nonsignificant correlations with both meritocracy perceptions ($r$ = -0.04, $p$ > .05; $r$ = -0.04, $p$ > .05).
-The justification of economic inequality indicator also has small yet statistically significant positive correlations with the meritocracy variables ($r$ = 0.05, $p$ < .05; $r$ = 0.07, $p$ < .01).
+The association between market justice preferences and economic inequality perception is negative but small and statistically significant ($r$ = -0.05, $p$ < .05).
+By contrast, market justice preferences positively and significantly correlate with the two meritocracy variables ($r$ = 0.15, $p$ < .01; $r$ = 0.13, $p$ < .01).
+Perceived economic inequality has negative and nonsignificant correlations with both meritocracy perceptions ($r$ = -0.04, $p$ > .05; $r$ = -0.04, $p$ > .05).
 Finally, the two meritocracy variables exhibit a strong positive association with each other ($r$ = 0.69, $p$ < .01).
@@
   filter(ola == 7) %>% 
-  select(mjp, perc_inequality, just_inequality, merit_effort, merit_talent) %>% 
+  select(mjp, perc_inequality, merit_effort, merit_talent) %>% 
   mutate_all(.funs = ~ as.numeric(.)) %>% 
@@
                  "B. Inequality gap perception",
-                 "C. Inequality gap justification",
-                 "D. People are rewarded for their efforts ",
-                 "E. People are rewarded for their intelligence")
+                 "C. People are rewarded for their efforts ",
+                 "D. People are rewarded for their intelligence")
 
 #set Column names of the matrix
-colnames(M) <-c("(A)", "(B)","(C)","(D)","(E)")
+colnames(M) <-c("(A)", "(B)","(C)","(D)")
 
@@
 Complementary, 77% of the variation corresponds to within-individual differences over time.
-
-According to Model 1, which includes the survey waves to capture intertemporal variations in the dependent variable, there is a decrease in 2017 ($\beta$ = -0.162, $p$ < .001) relative to 2016, and similarly in 2018 ($\beta$ = -0.015, $p$ > .05) and 2019 ($\beta$ = -0.041, $p$ > .05), although the latter effects are not statistically significant.
-In contrast, in the more recent waves of 2022 and 2023, there is a statistically significant increase in market justice preferences ($\beta$ = 0.289, $p$ < .001; $\beta$ = 0.286, $p$ < .001), suggesting a non-linear effect.
-To model this trajectory over time, Model 2 incorporates time (survey waves) as a continuous variable, along with its quadratic term, representing the non-linear association initially observed in Model 1.
-While the linear term (survey wave) shows a negative association, reflecting an overall decline in market preferences over time, the positive quadratic term indicates a reversal of this pattern in the final measurement points.
 
@@
          perc_inequality_cwc = perc_inequality - perc_inequality_mean,
-         just_inequality_mean = mean(just_inequality, na.rm = T),
-         just_inequality_cwc = just_inequality - just_inequality_mean,
          merit_effort_mean = mean(merit_effort, na.rm = T),
@@
 
+## WE and BE
+
 m1 <- lmer(mjp ~ 1 + ola + (1 | idencuesta),
@@
 
-m3 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + (1 + ola_num | idencuesta), 
+m3 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + (1 + ola_num | idencuesta), 
            data = df_study1)
 
-m4 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + (1 + ola_num | idencuesta),
+m4 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + (1 + ola_num | idencuesta),
                 data = df_study1)
 
-m5 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + (1 + ola_num | idencuesta),
+m5 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  (1 + ola_num | idencuesta),
                 data = df_study1)
 
-m6 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + (1 + ola_num | idencuesta),
+m6 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + (1 + ola_num | idencuesta),
                 data = df_study1)
 
-m7 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + (1 + ola_num | idencuesta),
+m7 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
                 data = df_study1)
 
-m8 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
+# interactions
+
+m8 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+             merit_effort_cwc*perc_inequality_cwc + merit_talent_cwc*perc_inequality_cwc +
+              (1 + merit_effort_cwc + merit_talent_cwc + perc_inequality_cwc | idencuesta),
                 data = df_study1)
 
-m9 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + just_inequality + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality +ola_num | idencuesta),
+m9 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+             merit_effort_mean*perc_inequality_mean + merit_talent_mean*perc_inequality_mean +
+              (1 | idencuesta),
                 data = df_study1)
 
-m10 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality*ola_num + merit_effort + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + just_inequality+ola_num| idencuesta),
+# growth curves
+
+m10 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + merit_effort + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality + ola_num| idencuesta),
                 data = df_study1)
 
-m11 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort*ola_num + merit_talent + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort+ola_num | idencuesta),
+m11 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + merit_effort*ola_num  + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort + ola_num| idencuesta),
                 data = df_study1)
 
-m12 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + just_inequality + merit_effort + merit_talent*ola_num + perc_inequality_mean + just_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_talent +ola_num| idencuesta),
+m12 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality +  merit_effort + merit_talent*ola_num + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+              (1 + merit_talent + ola_num| idencuesta),
                 data = df_study1)
+
+m13 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality  + merit_effort + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
+             merit_effort*perc_inequality*ola_num +
+              (1 + perc_inequality + merit_effort  + ola_num| idencuesta),
+                data = df_study1)
+
+m14 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality +  merit_effort + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
+              merit_talent*perc_inequality*ola_num +
+              (1 + perc_inequality + merit_talent + ola_num| idencuesta),
+                data = df_study1)
+
 ```
@@
   ola_2 = "Wave^2",
-  perc_inequality = "Perception inequality (WE)",
-  just_inequality = "Justification inequality (WE)",
-  merit_effort = "Merit: Effort (WE)",
-  merit_talent = "Merit: Talent (WE)",
+  perc_inequality_cwc = "Perception inequality (WE)",
+  merit_effort_cwc = "Merit: Effort (WE)",
+  merit_talent_cwc = "Merit: Talent (WE)",
   perc_inequality_mean = "Perception inequality (BE)",
-  just_inequality_mean = "Justification inequality (BE)",
   merit_effort_mean = "Merit: Effort (BE)",
@@
 
-texreg::texreg(list(m1, m2, m3, m4, m5, m6, m7, m8),
-               custom.model.names = c(paste0("Model ", seq(1:8))),
+texreg::texreg(list(m1, m2, m3, m4, m5, m6, m7),
+               custom.model.names = c(paste0("Model ", seq(1:7))),
                caption = NULL,
@@
                booktabs = F,
-               scalebox = 0.75,
+               scalebox = 0.79,
                include.loglik = FALSE,
@@
                center = T,
-               custom.gof.rows = list("Controls"=c(rep("No",7), "Yes")))
+               custom.gof.rows = list("Controls"=c(rep("No",6), "Yes")))
 
@@
 
-Models 3 to 6 incorporate the within-group effects (WE) of the primary independent variables, capturing how individual changes in these variables over time shape the dependent variable.
+According to Model 1, which includes the survey waves to capture intertemporal variations in the dependent variable, there is a decrease in 2017 ($\beta$ = -0.167, $p$ < .001) relative to 2016, and similarly in 2018 ($\beta$ = -0.025, $p$ > .05) and 2019 ($\beta$ = -0.046, $p$ > .05), although the latter effects are not statistically significant.
+In contrast, in the more recent waves of 2022 and 2023, there is a statistically significant increase in market justice preferences ($\beta$ = 0.285, $p$ < .001; $\beta$ = 0.284, $p$ < .001), suggesting a non-linear effect.
+To model this trajectory over time, Model 2 incorporates time (survey waves) as a continuous variable, along with its quadratic term, representing the non-linear association initially observed in Model 1.
+While the linear term (survey wave) shows a negative association, reflecting an overall decline in market preferences over time, the positive quadratic term indicates a reversal of this pattern in the final measurement points.
+
+Models 3 and 4 incorporate the within-group effects (WE) of the primary independent variables, capturing how individual changes in these variables over time shape the dependent variable.
 The results in Model 3 suggest that the within effect of perceived economic inequality is negative and statistically significant ($p$ < .001).
-Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.037 point decrease in market justice preferences.
-By contrast, in Model 4, the within effect of justifying economic inequality is positive and significant ($\beta$ = 0.100, $p$ < .001), implying that a rise in an individual’s justification of economic inequality over time corresponds to higher market justice preferences.
-Regarding meritocratic variables, Model 5 indicates that the perception of being rewarded for effort exerts a positive within effect ($\beta$ = 0.107, $p$ < .001).
-Similarly, Model 6 shows that viewing intelligence and ability (talent) as rewarded is also positively related to market justice preferences ($\beta$ = 0.029, $p$ < .05).
+Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.034 point decrease in market justice preferences.
+Model 4 shows that meritocratic beliefs operate in the opposite direction.
+An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.071, $p$ < .001), and a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with higher market-justice preferences ($\beta$ = 0.028, $p$ < .05).
 Taken together, these results suggest that individuals who increasingly perceive meritocracy—whether through effort or talent—tend to hold stronger market justice preferences.
 
-When examining the between-group effects (BE) in Model 7, which capture differences between individuals in the average of the main variables, a similar pattern emerges.
-Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.077, $p$ < .001), whereas those who justify economic inequality more strongly exhibit higher market justice preferences ($\beta$ = 0.156, $p$ < .001).
-Furthermore, the meritocratic perception that effort is rewarded has a positive impact on market justice preferences ($\beta$ = 0.092, $p$ < .05).
-By contrast, the perception that talent is rewarded shows a negative coefficient ($\beta$ = -0.006), though this relationship is not statistically significant at the 95% confidence level.
+When examining the between-group effects (BE) in Model 5 and 6, which capture differences between individuals in the average of the main variables, a similar pattern emerges.
+Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.042, $p$ < .05).
+In Model 6, the meritocratic belief that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.186, $p$ < .001).
+, whereas the belief that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.040).
+Notably, once the meritocratic variables are included, the negative coefficient for perceived economic inequality remains but is no longer statistically significant.
 
-Model 8 introduces the control variables.
-The within- and between-effects of the main predictors retain both their direction and statistical significance, indicating that the associations are robust to adjustment.
-Regarding the controls, belonging to higher household-income quintiles is associated with stronger market-justice preferences: individuals in the fourth and fifth quintiles score higher than those in the bottom quintile ($\beta$ = 0.093, $p$ < .05; $\beta$ = 0.147, $p$ < .001, respectively).
-Respondents who did not report their income also express greater support for market justice ($\beta$ = 0.172, $p$ < .01).
-Political orientation matters as well: compared with those on the left, individuals who place themselves on the right ($\beta$ = 0.246, $p$ < .001) or who declare no political position ($\beta$ = 0.092, $p$ < .01) display higher market-justice preferences.
-Finally, women exhibit lower market-justice preferences than men ($\beta$ = −0.080, $p$ < .01).
-
-@tbl-interactions displays
+Model 7 adds the control variables.
+With the exception of the between-effect for perceived economic inequality—which becomes nonsignificant—the within- and between-effects of the principal predictors retain both their direction and statistical significance, confirming the robustness of the associations (see Supplementary Material for effects of control variables).
 
 ```{r echo=FALSE, results='asis'}
-#| label: tbl-interactions
-#| tbl-cap: "Intereactions for within effects and time"
+#| label: tbl-interactions1
+#| tbl-cap: "Interactions for meritocracy, perceived economic inequality and market justice preferences"
 #| tbl-cap-location: top
@@
   "(Intercept)" = "Intercept",
-  "ola_num:perc_inequality" = "Perception inequality (WE) x Wave",
-  "ola_num:just_inequality" = "Justification inequality (WE) x Wave",
-  "ola_num:merit_effort" = "Merit: Effort (WE) x Wave",
-  "ola_num:merit_talent" = "Merit: Talent (WE) x Wave"
+  "perc_inequality_cwc:merit_effort_cwc" = "Merit: Effort (WE) x Perception inequality (WE)",
+  "perc_inequality_cwc:merit_talent_cwc" = "Merit: Talent (WE) x Perception inequality (WE)",
+  "perc_inequality_mean:merit_effort_mean" = "Merit: Effort (BE) x Perception inequality (BE)",
+  "perc_inequality_mean:merit_talent_mean" = "Merit: Talent (BE) x Perception inequality (BE)"
   )
 
-texreg::texreg(list(m9,m10,m11,m12),
-               custom.model.names = c("Model 9", "Model 10", "Model 11", "Model 12"),
+texreg::texreg(list(m8,m9),
+               custom.model.names = c("Model 8", "Model 9"),
                caption = NULL,
@@
                center = T,
-               custom.gof.rows = list("Controls"=c(rep("Yes",4))))
+               custom.gof.rows = list("Controls"=c(rep("Yes",2))))
 
 ```
+@tbl-interactions1 reports whether perceived economic inequality moderates the impact of meritocratic beliefs on market-justice preferences.
+In neither the within-person specification (Model 8) nor the between-person specification (Model 9) do the interaction terms reach statistical significance, indicating that higher perceptions of economic inequality do not meaningfully alter the effect of meritocracy on support for market allocation of social services.
+
+With regard to how the effects of perceived economic inequality, meritocratic beliefs, and their interaction evolve over time...
+
+Valores predichos o marginales??

### 2025-04-28 paper.tex
Up links
Update results, method and render AL

--- 2025-04-23 paper.tex
@@
 outcomes such as redistributive preferences, justification of
-inequality, and legitimacy of the economic system (Castillo et al., 2022
-{[}acá citar gente de Granada{]}). Research shows that greater perceived
-inequality fosters stronger preferences for redistribution, regardless
-of actual inequality levels
+inequality, and legitimacy of the economic system
+(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
+\citeproc{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et al.,
+2020}, \citeproc{ref-garcia-sanchez_perceptions_2018}{2018}). Research
+shows that greater perceived inequality fosters stronger preferences for
+redistribution, regardless of actual inequality levels
 (\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
@@
 
-Meritocracy posits that inequality is an inherent feature of societies
-but can be legitimized through fairer principles, such as effort and
-talent (\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
+Meritocracy posits that inequality can be legitimized through
+distributive criteria such as effort and talent
+(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
 \citeproc{ref-young_rise_1962}{Young, 1962}). Previous studies have
@@
 \citeproc{ref-wilson_role_2003}{Wilson, 2003}) and also to justify more
-inequality since these beliefs are associated with attitudes that
-maintain the legitimization of social differences
+inequality as these beliefs are associated with attitudes that
+legitimize status differences
 (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In a context
@@
 
-\subsection{Justification of market
-inequality}\label{justification-of-market-inequality}
+\subsection{The justification of market
+inequality}\label{the-justification-of-market-inequality}
 
@@
 that assess whether people support economic inequality when it reflects
-personal effort and skill. Over time, this approach has been extended
+personal effort and skills. Over time, this approach has been extended
 beyond income to include other market-mediated outcomes, such as
-education, healthcare, or pensions. For example, Knesebeck et al.~(2016)
-and Immergut and Schneider (\citeproc{ref-immergut_it_2020}{2020})
-examine whether citizens consider it fair that individuals with higher
-incomes can access better healthcare, while Lee and Stacey
+education, healthcare, and/or pensions. For example, Von Dem Knesebeck
+et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) and Immergut and
+Schneider (\citeproc{ref-immergut_it_2020}{2020}) examine whether
+citizens consider it fair that individuals with higher incomes can
+access better healthcare, while Lee and Stacey
 (\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
@@
 
+Perceptions of inequality have been associated to attitudes about market
+justice (\citeproc{ref-kluegel_social_1995a}{Kluegel et al., 1995};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research indicates that
+lower perceived inequality can reinforce support for market-based
+distributive arrangements by suggesting that the system is fair and that
+outcomes reflect effort and ability (\citeproc{ref-kuhn_eye_2011}{Kuhn,
+2011}). In contrast, when inequality is perceived as excessive or
+structurally determined, individuals could question the legitimacy of
+market justice and become more supportive of redistributive policies
+(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
+\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
+2019}).
+
 Perceptions of inequality refer to individuals' subjective evaluations
@@
 provides a concrete frame that respondents can relate to more easily
-than abstract questions about national income distribution {[}Castillo
-et al. (\citeproc{ref-castillo_percepcion_2019}{2012}); Easterbrook
-(\citeproc{ref-easterbrook_social_2021}{2021}); Willis et al., 2015{]}.
-While it enables the estimation of inequality using simple heuristics,
-this method is not without challenges. For instance, people often lack
+than abstract questions about national income distribution
+(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
+\citeproc{ref-easterbrook_social_2021}{Easterbrook, 2021};
+\citeproc{ref-willis_legitimacy_2015}{Willis et al., 2015}). While it
+enables the estimation of inequality using simple heuristics, this
+method is not without challenges. For instance, people often lack
 reliable knowledge about the earnings of those at the top of the income
@@
 Stix, 2020}). Despite this, perceived wage gaps are strong predictors of
-political
-attitudes(\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez
-et al., 2018}; \citeproc{ref-pedersen_attitudes_2019}{Pedersen \& Mutz,
-2019}), making them a valuable tool for understanding public responses
-to economic disparities.
+political attitudes
+(\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
+2018}; \citeproc{ref-pedersen_attitudes_2019}{Pedersen \& Mutz, 2019}),
+making them a valuable tool for understanding public responses to
+economic disparities.
 
@@
 \citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}).These items mix cognitive estimations with affective evaluations,
-complicating the interpretation of what respondents perceive versus what
-they morally reject. As a result, the conceptual clarity between
-perceived inequality and inequality aversion remains blurred in many
-empirical studies. To address this limitation, recent work has
-emphasized the need to distinguish between absolute and comparative
-measures, as well as between ideal and actual estimates of economic gaps
-{[}García-Sánchez \& De Carvalho
-(\citeproc{ref-garcia-sanchez_creencias_2022}{2022}); Auspurg et al.,
-2017{]}.
-
-Importantly, perceptions of inequality have been linked to attitudes
-about market justice---the belief that goods and services should be
-allocated according to individual contribution, competition, or
-willingness to pay (\citeproc{ref-kluegel_social_1995a}{Kluegel et al.,
-1995}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research
-indicates that lower perceived inequality can reinforce support for
-market-based distributive arrangements by suggesting that the system is
-fair and that outcomes reflect effort and ability
-(\citeproc{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when inequality
-is perceived as excessive or structurally determined, individuals could
-question the legitimacy of market justice and become more supportive of
-redistributive policies
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
-\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}).
+2019}). These items mix cognitive estimations with affective
+evaluations, complicating the interpretation of what respondents
+perceive versus what they morally reject. As a result, the conceptual
+clarity between perceived inequality and inequality aversion remains
+blurred in many empirical studies. To address this limitation, recent
+work has emphasized the need to distinguish between absolute and
+comparative measures, as well as between ideal and actual estimates of
+economic gaps (\citeproc{ref-auspurg_why_2017}{Auspurg et al., 2017};
+\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De
+Carvalho, 2022}).
 
@@
 Meritocracy constitutes a central ideological framework for legitimizing
-social inequality. Rooted in the belief that rewards and positions
-should be allocated based on individual effort and talent, meritocracy
-operates as a normative ideal and a descriptive belief about how society
+different types od social inequality, for instance through market
+justice beliefs. Rooted in the belief that rewards and positions should
+be allocated based on individual effort and talent, meritocracy operates
+as a normative ideal and a descriptive belief about how society
 functions. As initially conceptualized by Michael Young
@@
 system in which merit-based stratification becomes a new form of
-inequality. However, over time, meritocracy has been widely adopted in
+inequality. However, over time, meritocracy has been widely supported in
 many societies as a fair and desirable principle of distribution,
@@
 shapes their preferences for redistribution and justice
-(Tejero-Peregrina et al., 2025). Meritocratic beliefs serve as symbolic
-justifications for unequal outcomes, particularly when access is
-stratified by income or social opportunity. Prior studies have shown
-that individuals who perceive higher levels of meritocracy tend to
-express stronger support for unequal distributions that reflect market
-outcomes (\citeproc{ref-castillo_percepcion_2019}{Castillo et al.,
-2012}; \citeproc{ref-castillo_socialization_2024}{Castillo et al.,
-2024}).
+(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
+2025}) Meritocratic beliefs serve as symbolic justifications for unequal
+outcomes, particularly when access is stratified by income or social
+opportunity. Prior studies have shown that individuals who perceive
+higher levels of meritocracy tend to express stronger support for
+unequal distributions that reflect market outcomes
+(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
+\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
 
@@
 conditions, reducing support for redistributive measures and increasing
-the stigmatization of the poor (Hoyt \& Burnette, 2023). This reinforces
-negative stereotypes and reduces empathy toward individuals from lower
-socioeconomic backgrounds. Moreover, Busemeyer, Abrassart, and Nezi
-(2019) argue that meritocratic narratives can serve as feedback
-mechanisms that shape public opinion and well-being by framing
-individuals' understanding of welfare outcomes as deserved or undeserved
-within existing institutional structures. This psychological mechanism
-highlights the normative power of meritocracy in stabilizing unequal
-systems by shaping political attitudes and personal perceptions of
-success and failure.
-
-\subsection{This study (hypotheses)}\label{this-study-hypotheses}
+the stigmatization of the poor (\citeproc{ref-hoyt_mindsets_2023}{Hoyt
+et al., 2023}). This reinforces negative stereotypes and reduces empathy
+toward individuals from lower socioeconomic backgrounds. Moreover,
+Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argue
+that meritocratic narratives can serve as feedback mechanisms that shape
+public opinion and well-being by framing individuals' understanding of
+welfare outcomes as deserved or undeserved within existing institutional
+structures. This psychological mechanism highlights the normative power
+of meritocracy in stabilizing unequal systems by shaping political
+attitudes and personal perceptions of success and failure.
+
+\subsection{This study}\label{this-study}
+
+Building upon the previous literature, this study proposes that
+attitudes toward market justice are shaped by a dynamic interplay
+between individuals' perceptions of economic inequality and their
+beliefs in meritocracy. Specifically, we argue that both perceptions
+independently and interactively influence the extent to which
+individuals endorse market-based distributions of social goods and
+services in Chile.
+
+First, consistent with previous findings, we expect that a lower
+perception of economic inequality will be associated with stronger
+market justice preferences. When individuals perceive smaller economic
+gaps, they are more likely to view market mechanisms as fair and
+legitimate, reinforcing acceptance of outcomes based on competition and
+ability to pay. Conversely, a heightened perception of inequality may
+erode confidence in market fairness, weakening support for market-based
+distribution. This relationship is particularly relevant in the context
+of Chile, where the neoliberal economic model has been a dominant force
+shaping public attitudes toward inequality and justice.
+
+Second, stronger meritocratic beliefs are expected to be positively
+associated with market justice preferences. Individuals who believe that
+effort and talent primarily determine success are more likely to justify
+unequal outcomes and endorse the notion that markets allocate resources
+fairly according to individual merit. This aligns with the idea that
+meritocratic beliefs serve as a moral framework that legitimizes
+market-based inequalities, as individuals perceive the system as just
+when they believe that rewards are based on individual merit. This is
+particularly relevant in contexts where neoliberal ideologies dominate,
+as they often emphasize individual responsibility and competition as the
+basis for social order.
+
+Third, we propose that perceptions of meritocracy and perceptions of
+economic inequality interact in shaping market justice attitudes.
+Specifically, we argue that the legitimizing effect of meritocratic
+beliefs on market justice preferences is moderated by perceptions of
+economic inequality: when perceived inequality is low, the positive
+association between meritocratic beliefs and market justice preferences
+will be stronger. However, when perceived inequality is high, this
+association will weaken, as greater awareness of large economic gaps may
+challenge the view that outcomes are purely merit-based.
+
+Additionally, this study examines whether these relationships vary over
+time in response to significant social and political events. In
+particular, we explore whether the political outburst of 2019 and the
+subsequent constitutional processes in Chile have altered how
+perceptions of inequality and meritocracy shape market justice
+attitudes. We expect that after these critical events, perceptions of
+inequality may have a stronger negative effect on market justice
+preferences, reflecting increased societal questioning of market-based
+allocation mechanisms.
+
+Based on these arguments, we propose the following hypotheses:
+
+\(H1\): Greater perceived economic inequality is associated with
+stronger market justice preferences.
+
+\(H2\): Stronger meritocratic beliefs are associated with stronger
+market justice preferences.
+
+\(H3\): The positive association between meritocratic beliefs and market
+justice preferences is moderated by perceived economic inequality;
+specifically, this association is weaker when perceived economic
+inequality is high.
+
+\(H4\): The effects described in H1--H3 are attenuated after major
+social mobilizations (2019--2022), reflecting increased critical views
+of the market's role in allocating social goods.
 
@@
 \subsection{Variables}\label{variables}
-
-\emph{Dependent variable}
 
@@
 
-\emph{Independent variables}
-
 \textbf{Perception of economic inequality}: The main independent
 variable refers to the perception of economic inequality, measured
-through the perceived wage gap, which serves as an absolute indicator of
-inequality perceptions (\citeproc{ref-castillo_perception_2022}{Castillo
-et al., 2022}). This measure is derived from the gap between the
-perceived salaries of jobs at opposite ends of the occupational
-hierarchy. Specifically, it relies on the logarithm of the division
-between the perceived salary of a large-company president and that of an
-unskilled worker (\citeproc{ref-Castillo2011}{Castillo, 2011}). Higher
-values of this term indicate a greater perception of economic inequality
-between occupations located at the extremes of the status continuum.
-
-\textbf{Justification of economic inequality}: This construct is
-measured as the just earnings gap, an approach to measuring the
-justification of economic inequality based on earning differences
-considered just between high and low status occupations
-(\citeproc{ref-Castillo2011}{Castillo, 2011}). Similarly to economic
-inequality perception, this indicator relies on the logarithmic ratio
-between the just salary of a large-company president and that of an
-unskilled worker. Higher values of this term indicate a greater
-justification of economic inequality.
-
-\textbf{Meritocracy}: Meritocratic perception is operationalized through
-two components: one addressing effort and another focusing on talent
+through the perceived wage gap.
+(\citeproc{ref-castillo_cual_2009}{Castillo, 2009};
+\citeproc{ref-gijsberts_thelegitimation_1999}{Gijsberts, 1999};
+\citeproc{ref-hadler_why_2005}{Hadler, 2005}). This measure is derived
+from the salary gap between the perceived salaries of jobs at opposite
+ends of the occupational hierarchy. Specifically, it relies on the
+division between the perceived salary of a large-company president and
+that of an unskilled worker (\citeproc{ref-Castillo2011}{Castillo,
+2011}). Higher values of this term indicate a greater perception of
+economic inequality between occupations located at the extremes of the
+status continuum. This measure includes a logarithmic term in order to
+adjust income magnitudes (usually fewer cases with high income):
+
+\[
+\text{perceived wage gap} = \log_{10}\left(\frac{\text{perceived salary of a large-company president}}{\text{perceived salary of an unskilled worker}}\right)
+\]
+
+\textbf{Perception of Meritocracy}: this variable is operationalized
+through two components, namely effort and talent
 (\citeproc{ref-young_rise_1962}{Young, 1962}). The item used to gauge
@@
 the item for talent is: ``In Chile, people are rewarded for their
-intelligence and skills.'' In both cases, respondents indicate their
+intelligence and skills''. In both cases, respondents indicate their
 level of agreement on a five-point Likert scale, ranging from 1
@@
 IQR (CV) : 1.6 (0.3)\strut
-\end{minipage} & 294 distinct values &
+\end{minipage} & 296 distinct values &
 \begin{minipage}[t]{\linewidth}\raggedright
-1477\\
-(100.0\%)\strut
-\end{minipage} \\
-Inequality gap justification &
-\begin{minipage}[t]{\linewidth}\raggedright
-Mean (sd) : 2.1 (1.4)\\
-min \textless{} med \textless{} max:\\
--5.5 \textless{} 2.1 \textless{} 9.9\\
-IQR (CV) : 1.7 (0.6)\strut
-\end{minipage} & 215 distinct values &
-\begin{minipage}[t]{\linewidth}\raggedright
-1477\\
+1501\\
 (100.0\%)\strut
@@
 \end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-168 (11.4\%)\\
-679 (46.0\%)\\
-258 (17.5\%)\\
-324 (21.9\%)\\
+169 (11.3\%)\\
+693 (46.2\%)\\
+263 (17.5\%)\\
+328 (21.9\%)\\
 48 ( 3.2\%)\strut
 \end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-1477\\
+1501\\
 (100.0\%)\strut
@@
 \end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-134 ( 9.1\%)\\
-604 (40.9\%)\\
-289 (19.6\%)\\
-395 (26.7\%)\\
+134 ( 8.9\%)\\
+617 (41.1\%)\\
+294 (19.6\%)\\
+401 (26.7\%)\\
 55 ( 3.7\%)\strut
 \end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-1477\\
+1501\\
 (100.0\%)\strut
@@
 2=Universitary), age (in years), and sex (1=Male, 2=Female), which have
-been shown to influence market justice preferences significantly
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding attitudinal
-variables, we include political identification (1=Left, 2=Center,
-3=Right, 4=No identification) and subjective social status (ranging from
-1 to 10) because they may confound the relationship between market
-justice preferences and perceptions of inequality and meritocracy
+been previously shown to influence market justice preferences
+significantly (\citeproc{ref-castillo_socialization_2024}{Castillo et
+al., 2024}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding
+attitudinal variables, we include political identification (1=Left,
+2=Center, 3=Right, 4=No identification) and subjective social status
+(ranging from 1 to 10) as they may confound the relationship between
+market justice preferences and perceptions of inequality and meritocracy
 (\citeproc{ref-schneider_poverty_2015}{Schneider \& Castillo, 2015}).
@@
 the next, given that we are using panel data. For instance, of the
-40.9\% who strongly disagreed with justifying inequality in healthcare
+40.8\% who strongly disagreed with justifying inequality in healthcare
 in 2019, around 24.3\% maintained that position in 2022, while the
-remaining 16.6\% shifted toward other response categories---primarily
+remaining 16.5\% shifted toward other response categories---primarily
 moving into disagreement rather than strong disagreement. Overall, more
@@
 for inequality. Specifically, in healthcare and education, although
-disagreement remains substantial, agreement increased from 8.9\% and
-7.6\% in 2019 to 12\% and 12.5\% in 2023, respectively. This shift is
-most evident in pensions, where the combined agree/strongly agree
-category grew by about 10 percentage points, from 16.9\% in 2016 to
-27.8\% in 2023.
+disagreement remains substantial, agreement (agree + strongly agree)
+increased from 7.4\% and 7.2\% in 2019 to 13.1\% and 14.2\% in 2023,
+respectively. This shift is most evident in pensions, where the combined
+agree/strongly agree category grew by about 10 percentage points, from
+16.9\% in 2016 to 28\% in 2023.
 
@@
 \caption{\label{fig-alluvial}Change in the justification of inequality
-in health, pensions and education over time (2016-2023)}
+in healthcare, pensions and education over time (2016-2023)}
 
@@
 perceptions of economic inequality declined in the latest measurements
-(2022--2023), market justice preferences increased. With respect to
-economic inequality justification, there is a steady decrease in average
-levels over the years, registering the lowest mean compared to the other
-variables. The meritocracy measures remain stable, though the perception
-that individuals are rewarded for intelligence is slightly higher than
-the perception that they are rewarded for their effort.
+(2022--2023), market justice preferences increased. The meritocracy
+measures remain stable, though the perception that individuals are
+rewarded for intelligence is slightly higher than the perception that
+they are rewarded for their effort.
 
@@
 \caption{\label{fig-meanchange}Change in the mean of market justice
-preferences, economic inequality perception and justification, and
-meritocracy (2016-2023)}
+preferences, economic inequality perception, and meritocracy
+(2016-2023)}
 
@@
 from low to moderate. The association between market justice preferences
-and economic inequality perception is negative but small and not
-statistically significant (\(r\) = -0.05, \(p\) \textgreater{} .05). By
+and economic inequality perception is negative but small and
+statistically significant (\(r\) = -0.05, \(p\) \textless{} .05). By
 contrast, market justice preferences positively and significantly
-correlate with the justification of economic inequality (\(r\) = 0.16,
-\(p\) \textless{} .01), as well as show low, positive associations with
-the two meritocracy variables (\(r\) = 0.14, \(p\) \textless{} .01;
-\(r\) = 0.12, \(p\) \textless{} .01). Perceived economic inequality has
-a strong, positive correlation with justifying economic inequality
-(\(r\) = 0.53, \(p\) \textless{} .01), but negative and nonsignificant
-correlations with both meritocracy perceptions (\(r\) = -0.04, \(p\)
-\textgreater{} .05; \(r\) = -0.04, \(p\) \textgreater{} .05). The
-justification of economic inequality indicator also has small yet
-statistically significant positive correlations with the meritocracy
-variables (\(r\) = 0.05, \(p\) \textless{} .05; \(r\) = 0.07, \(p\)
-\textless{} .01). Finally, the two meritocracy variables exhibit a
-strong positive association with each other (\(r\) = 0.69, \(p\)
-\textless{} .01).
+correlate with the two meritocracy variables (\(r\) = 0.15, \(p\)
+\textless{} .01; \(r\) = 0.13, \(p\) \textless{} .01). Perceived
+economic inequality has negative and nonsignificant correlations with
+both meritocracy perceptions (\(r\) = -0.04, \(p\) \textgreater{} .05;
+\(r\) = -0.04, \(p\) \textgreater{} .05). Finally, the two meritocracy
+variables exhibit a strong positive association with each other (\(r\) =
+0.69, \(p\) \textless{} .01).
 
@@
 
+\begin{table}
+
+\caption{\label{tbl-modelos}Longitudinal multilevel models for market
+justice preferences}
+
+\centering{
+
+\begin{center}
+\scalebox{0.79}{
+\begin{tabular}{l c c c c c c c}
+\hline
+ & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 \\
+\hline
+Intercept                            & $1.968^{***}$  & $1.973^{***}$  & $1.985^{***}$  & $1.982^{***}$  & $2.122^{***}$  & $1.408^{***}$  & $1.499^{***}$  \\
+                                     & $(0.021)$      & $(0.032)$      & $(0.032)$      & $(0.032)$      & $(0.069)$      & $(0.096)$      & $(0.114)$      \\
+Wave (Ref.= 2016)                    &                &                &                &                &                &                &                \\
+                                     &                &                &                &                &                &                &                \\
+\quad Wave 2017                      & $-0.167^{***}$ &                &                &                &                &                &                \\
+                                     & $(0.027)$      &                &                &                &                &                &                \\
+\quad Wave 2018                      & $-0.025$       &                &                &                &                &                &                \\
+                                     & $(0.027)$      &                &                &                &                &                &                \\
+\quad Wave 2019                      & $-0.046$       &                &                &                &                &                &                \\
+                                     & $(0.027)$      &                &                &                &                &                &                \\
+\quad Wave 2022                      & $0.285^{***}$  &                &                &                &                &                &                \\
+                                     & $(0.028)$      &                &                &                &                &                &                \\
+\quad Wave 2023                      & $0.284^{***}$  &                &                &                &                &                &                \\
+                                     & $(0.027)$      &                &                &                &                &                &                \\
+Wave                                 &                & $-0.075^{***}$ & $-0.078^{***}$ & $-0.079^{***}$ & $-0.079^{***}$ & $-0.079^{***}$ & $-0.079^{***}$ \\
+                                     &                & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
+Wave$^2$                             &                & $0.017^{***}$  & $0.018^{***}$  & $0.018^{***}$  & $0.018^{***}$  & $0.018^{***}$  & $0.018^{***}$  \\
+                                     &                & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      \\
+Perception inequality (WE)           &                &                & $-0.034^{***}$ & $-0.029^{**}$  & $-0.029^{**}$  & $-0.029^{**}$  & $-0.029^{**}$  \\
+                                     &                &                & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
+Merit: Effort (WE)                   &                &                &                & $0.071^{***}$  & $0.071^{***}$  & $0.070^{***}$  & $0.070^{***}$  \\
+                                     &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
+Merit: Talent (WE)                   &                &                &                & $0.028^{*}$    & $0.028^{*}$    & $0.028^{*}$    & $0.027^{*}$    \\
+                                     &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
+Perception inequality (BE)           &                &                &                &                & $-0.042^{*}$   & $-0.005$       & $-0.032$       \\
+                                     &                &                &                &                & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
+Merit: Effort (BE)                   &                &                &                &                &                & $0.186^{***}$  & $0.178^{***}$  \\
+                                     &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
+Merit: Talent (BE)                   &                &                &                &                &                & $0.040$        & $0.019$        \\
+                                     &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
+\hline
+Controls                             & No             & No             & No             & No             & No             & No             & Yes            \\
+BIC                                  & $20780.594$    & $20757.596$    & $20761.847$    & $20704.882$    & $20714.843$    & $20633.143$    & $20727.029$    \\
+Num. obs.                            & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         \\
+Num. groups: idencuesta              & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         \\
+Var: idencuesta (Intercept)          & $0.172$        & $0.207$        & $0.204$        & $0.198$        & $0.196$        & $0.172$        & $0.160$        \\
+Var: Residual                        & $0.530$        & $0.497$        & $0.497$        & $0.493$        & $0.493$        & $0.492$        & $0.493$        \\
+Var: idencuesta ola\_num             & $$             & $0.007$        & $0.007$        & $0.006$        & $0.006$        & $0.006$        & $0.006$        \\
+Cov: idencuesta (Intercept) ola\_num & $$             & $-0.017$       & $-0.017$       & $-0.014$       & $-0.014$       & $-0.013$       & $-0.013$       \\
+\hline
+\multicolumn{8}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
+\end{tabular}
+}
+\caption{}
+\label{table:coefficients}
+\end{center}
+
+}
+
+\end{table}%
+
 According to Model 1, which includes the survey waves to capture
 intertemporal variations in the dependent variable, there is a decrease
-in 2017 (\(\beta\) = -0.162, \(p\) \textless{} .001) relative to 2016,
-and similarly in 2018 (\(\beta\) = -0.015, \(p\) \textgreater{} .05) and
-2019 (\(\beta\) = -0.041, \(p\) \textgreater{} .05), although the latter
+in 2017 (\(\beta\) = -0.167, \(p\) \textless{} .001) relative to 2016,
+and similarly in 2018 (\(\beta\) = -0.025, \(p\) \textgreater{} .05) and
+2019 (\(\beta\) = -0.046, \(p\) \textgreater{} .05), although the latter
 effects are not statistically significant. In contrast, in the more
 recent waves of 2022 and 2023, there is a statistically significant
-increase in market justice preferences (\(\beta\) = 0.289, \(p\)
-\textless{} .001; \(\beta\) = 0.286, \(p\) \textless{} .001), suggesting
+increase in market justice preferences (\(\beta\) = 0.285, \(p\)
+\textless{} .001; \(\beta\) = 0.284, \(p\) \textless{} .001), suggesting
 a non-linear effect. To model this trajectory over time, Model 2
@@
 
-\begin{table}
-
-\caption{\label{tbl-modelos}Longitudinal multilevel models for market
-justice preferences}
-
-\centering{
-
-[H]
-\begin{center}
-\scalebox{1}{
-\begin{threeparttable}
-\begin{tabular}{l c c c c c c c c}
-\toprule
- & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 & Model 8 \\
-\midrule
-Intercept                            & $1.964^{***}$  & $1.966^{***}$  & $2.103^{***}$  & $2.059^{***}$  & $1.752^{***}$  & $1.721^{***}$  & $1.541^{***}$  & $1.574^{***}$  \\
-                                     & $(0.022)$      & $(0.033)$      & $(0.046)$      & $(0.045)$      & $(0.052)$      & $(0.053)$      & $(0.095)$      & $(0.112)$      \\
-Wave (Ref.= 2016)                    &                &                &                &                &                &                &                &                \\
-                                     &                &                &                &                &                &                &                &                \\
-\quad Wave 2017                      & $-0.162^{***}$ &                &                &                &                &                &                &                \\
-                                     & $(0.028)$      &                &                &                &                &                &                &                \\
-\quad Wave 2018                      & $-0.015$       &                &                &                &                &                &                &                \\
-                                     & $(0.027)$      &                &                &                &                &                &                &                \\
-\quad Wave 2019                      & $-0.041$       &                &                &                &                &                &                &                \\
-                                     & $(0.027)$      &                &                &                &                &                &                &                \\
-\quad Wave 2022                      & $0.289^{***}$  &                &                &                &                &                &                &                \\
-                                     & $(0.028)$      &                &                &                &                &                &                &                \\
-\quad Wave 2023                      & $0.286^{***}$  &                &                &                &                &                &                &                \\
-                                     & $(0.027)$      &                &                &                &                &                &                &                \\
-Wave                                 &                & $-0.069^{***}$ & $-0.072^{***}$ & $-0.057^{**}$  & $-0.057^{**}$  & $-0.058^{**}$  & $-0.064^{***}$ & $-0.064^{***}$ \\
-                                     &                & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
-Wave$^2$                             &                & $0.017^{***}$  & $0.017^{***}$  & $0.015^{***}$  & $0.015^{***}$  & $0.015^{***}$  & $0.016^{***}$  & $0.016^{***}$  \\
-                                     &                & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      \\
-Perception inequality (WE)           &                &                & $-0.037^{***}$ & $-0.086^{***}$ & $-0.075^{***}$ & $-0.074^{***}$ & $-0.056^{***}$ & $-0.056^{***}$ \\
-                                     &                &                & $(0.009)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.011)$      & $(0.011)$      \\
-Justification inequality (WE)        &                &                &                & $0.100^{***}$  & $0.095^{***}$  & $0.094^{***}$  & $0.055^{***}$  & $0.055^{***}$  \\
-                                     &                &                &                & $(0.009)$      & $(0.009)$      & $(0.009)$      & $(0.010)$      & $(0.010)$      \\
-Merit: Effort (WE)                   &                &                &                &                & $0.107^{***}$  & $0.088^{***}$  & $0.071^{***}$  & $0.071^{***}$  \\
-                                     &                &                &                &                & $(0.009)$      & $(0.011)$      & $(0.012)$      & $(0.012)$      \\
-Merit: Talent (WE)                   &                &                &                &                &                & $0.029^{*}$    & $0.025^{*}$    & $0.025^{*}$    \\
-                                     &                &                &                &                &                & $(0.011)$      & $(0.012)$      & $(0.012)$      \\
-Perception inequality (BE)           &                &                &                &                &                &                & $-0.077^{***}$ & $-0.078^{***}$ \\
-                                     &                &                &                &                &                &                & $(0.023)$      & $(0.023)$      \\
-Justification inequality (BE)        &                &                &                &                &                &                & $0.156^{***}$  & $0.131^{***}$  \\
-                                     &                &                &                &                &                &                & $(0.021)$      & $(0.021)$      \\
-Merit: Effort (BE)                   &                &                &                &                &                &                & $0.092^{**}$   & $0.088^{**}$   \\
-                                     &                &                &                &                &                &                & $(0.033)$      & $(0.033)$      \\
-Merit: Talent (BE)                   &                &                &                &                &                &                & $-0.006$       & $-0.016$       \\
-                                     &                &                &                &                &                &                & $(0.033)$      & $(0.033)$      \\
-\midrule
-Controls                             & No             & No             & No             & No             & No             & No             & No             & Yes            \\
-AIC                                  & $20273.132$    & $20260.158$    & $20251.351$    & $20140.808$    & $20002.562$    & $20005.214$    & $19957.283$    & $19986.833$    \\
-BIC                                  & $20329.477$    & $20309.460$    & $20307.696$    & $20204.196$    & $20072.993$    & $20082.688$    & $20062.929$    & $20191.083$    \\
-Log Likelihood                       & $-10128.566$   & $-10123.079$   & $-10117.676$   & $-10061.404$   & $-9991.281$    & $-9991.607$    & $-9963.641$    & $-9964.417$    \\
-Num. obs.                            & $8460$         & $8460$         & $8460$         & $8460$         & $8460$         & $8460$         & $8460$         & $8460$         \\
-Num. groups: idencuesta              & $1681$         & $1681$         & $1681$         & $1681$         & $1681$         & $1681$         & $1681$         & $1681$         \\
-Var: idencuesta (Intercept)          & $0.174$        & $0.214$        & $0.210$        & $0.204$        & $0.179$        & $0.178$        & $0.180$        & $0.172$        \\
-Var: Residual                        & $0.528$        & $0.494$        & $0.494$        & $0.493$        & $0.490$        & $0.490$        & $0.488$        & $0.488$        \\
-Var: idencuesta ola\_num             & $$             & $0.007$        & $0.007$        & $0.007$        & $0.006$        & $0.006$        & $0.006$        & $0.006$        \\
-Cov: idencuesta (Intercept) ola\_num & $$             & $-0.018$       & $-0.017$       & $-0.018$       & $-0.015$       & $-0.015$       & $-0.016$       & $-0.016$       \\
-\bottomrule
-\end{tabular}
-\begin{tablenotes}[flushleft]
-\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}
-\end{tablenotes}
-\end{threeparttable}
-}
-\caption{}
-\label{table:coefficients}
-\end{center}
-
-}
-
-\end{table}%
-
-Models 3 to 6 incorporate the within-group effects (WE) of the primary
+Models 3 and 4 incorporate the within-group effects (WE) of the primary
 independent variables, capturing how individual changes in these
@@
 Specifically, each one-point increase in an individual's perception of
-economic inequality between waves is associated with a 0.037 point
-decrease in market justice preferences. By contrast, in Model 4, the
-within effect of justifying economic inequality is positive and
-significant (\(\beta\) = 0.100, \(p\) \textless{} .001), implying that a
-rise in an individual's justification of economic inequality over time
-corresponds to higher market justice preferences. Regarding meritocratic
-variables, Model 5 indicates that the perception of being rewarded for
-effort exerts a positive within effect (\(\beta\) = 0.107, \(p\)
-\textless{} .001). Similarly, Model 6 shows that viewing intelligence
-and ability (talent) as rewarded is also positively related to market
-justice preferences (\(\beta\) = 0.029, \(p\) \textless{} .05). Taken
-together, these results suggest that individuals who increasingly
-perceive meritocracy---whether through effort or talent---tend to hold
-stronger market justice preferences.
-
-When examining the between-group effects (BE) in Model 7, which capture
-differences between individuals in the average of the main variables, a
-similar pattern emerges. Individuals who perceive higher levels of
-economic inequality tend to prefer less market justice (\(\beta\) =
--0.077, \(p\) \textless{} .001), whereas those who justify economic
-inequality more strongly exhibit higher market justice preferences
-(\(\beta\) = 0.156, \(p\) \textless{} .001). Furthermore, the
-meritocratic perception that effort is rewarded has a positive impact on
-market justice preferences (\(\beta\) = 0.092, \(p\) \textless{} .05).
-By contrast, the perception that talent is rewarded shows a negative
-coefficient (\(\beta\) = -0.006), though this relationship is not
-statistically significant at the 95\% confidence level.
-
-Model 8 introduces the control variables. The within- and
-between-effects of the main predictors retain both their direction and
-statistical significance, indicating that the associations are robust to
-adjustment. Regarding the controls, belonging to higher household-income
-quintiles is associated with stronger market-justice preferences:
-individuals in the fourth and fifth quintiles score higher than those in
-the bottom quintile (\(\beta\) = 0.093, \(p\) \textless{} .05; \(\beta\)
-= 0.147, \(p\) \textless{} .001, respectively). Respondents who did not
-report their income also express greater support for market justice
-(\(\beta\) = 0.172, \(p\) \textless{} .01). Political orientation
-matters as well: compared with those on the left, individuals who place
-themselves on the right (\(\beta\) = 0.246, \(p\) \textless{} .001) or
-who declare no political position (\(\beta\) = 0.092, \(p\) \textless{}
-.01) display higher market-justice preferences. Finally, women exhibit
-lower market-justice preferences than men (\(\beta\) = −0.080, \(p\)
-\textless{} .01).
+economic inequality between waves is associated with a 0.034 point
+decrease in market justice preferences. Model 4 shows that meritocratic
+beliefs operate in the opposite direction. An upward shift in the
+perception that effort is rewarded exerts a positive within effect
+(\(\beta\) = 0.071, \(p\) \textless{} .001), and a parallel increase in
+the perception that intelligence and ability are rewarded is likewise
+associated with higher market-justice preferences (\(\beta\) = 0.028,
+\(p\) \textless{} .05). Taken together, these results suggest that
+individuals who increasingly perceive meritocracy---whether through
+effort or talent---tend to hold stronger market justice preferences.
+
+When examining the between-group effects (BE) in Model 5 and 6, which
+capture differences between individuals in the average of the main
+variables, a similar pattern emerges. Individuals who perceive higher
+levels of economic inequality tend to prefer less market justice
+(\(\beta\) = -0.042, \(p\) \textless{} .05). In Model 6, the
+meritocratic belief that effort is rewarded is positively associated
+with market justice preferences (\(\beta\) = 0.186, \(p\) \textless{}
+.001). , whereas the belief that talent is rewarded shows a positive but
+non-significant coefficient (\(\beta\) = 0.040). Notably, once the
+meritocratic variables are included, the negative coefficient for
+perceived economic inequality remains but is no longer statistically
+significant.
+
+Model 7 adds the control variables. With the exception of the
+between-effect for perceived economic inequality---which becomes
+nonsignificant---the within- and between-effects of the principal
+predictors retain both their direction and statistical significance,
+confirming the robustness of the associations (see Supplementary
+Material for effects of control variables).
 
@@
 
-\caption{\label{tbl-interactions}Intereactions for within effects and
-time}
+\caption{\label{tbl-interactions1}Interactions for meritocracy,
+perceived economic inequality and market justice preferences}
 
@@
 
-\begin{verbatim}
-
-\begin{table}[h!]
 \begin{center}
-\scalebox{1}{
-\begin{threeparttable}
-\begin{tabular}{l c c c c}
-\toprule
- & Model 9 & Model 10 & Model 11 & Model 12 \\
-\midrule
-Intercept                                    & $1.605^{***}$ & $1.679^{***}$ & $1.425^{***}$ & $1.444^{***}$ \\
-                                             & $(0.123)$     & $(0.112)$     & $(0.117)$     & $(0.117)$     \\
-Perception inequality (WE) x Wave            & $0.002$       &               &               &               \\
-                                             & $(0.004)$     &               &               &               \\
-Justification inequality (WE) x Wave         &               & $0.013^{***}$ &               &               \\
-                                             &               & $(0.003)$     &               &               \\
-Merit: Effort (WE) x Wave                    &               &               & $-0.012^{**}$ &               \\
-                                             &               &               & $(0.004)$     &               \\
-Merit: Talent (WE) x Wave                    &               &               &               & $-0.010^{*}$  \\
-                                             &               &               &               & $(0.004)$     \\
-\midrule
-Controls                                     & Yes           & Yes           & Yes           & Yes           \\
-AIC                                          & $19995.908$   & $19923.216$   & $19940.577$   & $19932.452$   \\
-BIC                                          & $20228.331$   & $20155.638$   & $20172.999$   & $20164.875$   \\
-Log Likelihood                               & $-9964.954$   & $-9928.608$   & $-9937.289$   & $-9933.226$   \\
-Num. obs.                                    & $8460$        & $8460$        & $8460$        & $8460$        \\
-Num. groups: idencuesta                      & $1681$        & $1681$        & $1681$        & $1681$        \\
-Var: idencuesta (Intercept)                  & $0.305$       & $0.198$       & $0.137$       & $0.123$       \\
-Var: idencuesta perc\_inequality             & $0.004$       & $$            & $$            & $$            \\
-Var: idencuesta ola\_num                     & $0.007$       & $0.006$       & $0.006$       & $0.006$       \\
-Cov: idencuesta (Intercept) perc\_inequality & $-0.026$      & $$            & $$            & $$            \\
-Cov: idencuesta (Intercept) ola\_num         & $-0.032$      & $-0.027$      & $-0.002$      & $-0.000$      \\
-Cov: idencuesta perc\_inequality ola\_num    & $0.004$       & $$            & $$            & $$            \\
-Var: Residual                                & $0.485$       & $0.484$       & $0.472$       & $0.471$       \\
-Var: idencuesta just\_inequality             & $$            & $0.010$       & $$            & $$            \\
-Cov: idencuesta (Intercept) just\_inequality & $$            & $-0.022$      & $$            & $$            \\
-Cov: idencuesta just\_inequality ola\_num    & $$            & $0.007$       & $$            & $$            \\
-Var: idencuesta merit\_effort                & $$            & $$            & $0.022$       & $$            \\
-Cov: idencuesta (Intercept) merit\_effort    & $$            & $$            & $-0.024$      & $$            \\
-Cov: idencuesta merit\_effort ola\_num       & $$            & $$            & $-0.005$      & $$            \\
-Var: idencuesta merit\_talent                & $$            & $$            & $$            & $0.021$       \\
-Cov: idencuesta (Intercept) merit\_talent    & $$            & $$            & $$            & $-0.023$      \\
-Cov: idencuesta merit\_talent ola\_num       & $$            & $$            & $$            & $-0.005$      \\
-\bottomrule
+\scalebox{0.95}{
+\begin{tabular}{l c c}
+\hline
+ & Model 8 & Model 9 \\
+\hline
+Intercept                                                & $1.511^{***}$ & $2.309^{***}$ \\
+                                                         & $(0.114)$     & $(0.287)$     \\
+Merit: Effort (WE) x Perception inequality (WE)          & $0.018$       &               \\
+                                                         & $(0.016)$     &               \\
+Merit: Talent (WE) x Perception inequality (WE)          & $-0.002$      &               \\
+                                                         & $(0.016)$     &               \\
+Merit: Effort (BE) x Perception inequality (BE)          &               & $0.016$       \\
+                                                         &               & $(0.045)$     \\
+Merit: Talent (BE) x Perception inequality (BE)          &               & $0.069$       \\
+                                                         &               & $(0.045)$     \\
+\hline
+Controls                                                 & Yes           & Yes           \\
+BIC                                                      & $20858.742$   & $20799.038$   \\
+Num. obs.                                                & $8643$        & $8643$        \\
+Num. groups: idencuesta                                  & $1687$        & $1687$        \\
+Var: idencuesta (Intercept)                              & $0.147$       & $0.141$       \\
+Var: idencuesta merit\_effort\_cwc                       & $0.017$       & $$            \\
+Var: idencuesta merit\_talent\_cwc                       & $0.018$       & $$            \\
+Var: idencuesta perc\_inequality\_cwc                    & $0.000$       & $$            \\
+Cov: idencuesta (Intercept) merit\_effort\_cwc           & $-0.005$      & $$            \\
+Cov: idencuesta (Intercept) merit\_talent\_cwc           & $0.015$       & $$            \\
+Cov: idencuesta (Intercept) perc\_inequality\_cwc        & $-0.006$      & $$            \\
+Cov: idencuesta merit\_effort\_cwc merit\_talent\_cwc    & $-0.003$      & $$            \\
+Cov: idencuesta merit\_effort\_cwc perc\_inequality\_cwc & $0.000$       & $$            \\
+Cov: idencuesta merit\_talent\_cwc perc\_inequality\_cwc & $0.001$       & $$            \\
+Var: Residual                                            & $0.501$       & $0.527$       \\
+\hline
+\multicolumn{3}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
 \end{tabular}
-\begin{tablenotes}[flushleft]
-\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}
-\end{tablenotes}
-\end{threeparttable}
 }
@@
 \end{table}%
+
+Table~\ref{tbl-interactions1} reports whether perceived economic
+inequality moderates the impact of meritocratic beliefs on
+market-justice preferences. In neither the within-person specification
+(Model 8) nor the between-person specification (Model 9) do the
+interaction terms reach statistical significance, indicating that higher
+perceptions of economic inequality do not meaningfully alter the effect
+of meritocracy on support for market allocation of social services.
+
+With regard to how the effects of perceived economic inequality,
+meritocratic beliefs, and their interaction evolve over time\ldots{}
+
+Valores predichos o marginales??
 
@@
 
+\bibitem[\citeproctext]{ref-auspurg_why_2017}
+Auspurg, K., Hinz, T., \& Sauer, C. (2017). Why {Should Women Get Less}?
+{Evidence} on the {Gender Pay Gap} from {Multifactorial Survey
+Experiments}. \emph{American Sociological Review}, \emph{82}(1),
+179--210. \url{https://doi.org/10.1177/0003122416683393}
+
 \bibitem[\citeproctext]{ref-bates_fitting_2015}
@@
 
+\bibitem[\citeproctext]{ref-busemeyer_positive_2021}
+Busemeyer, M. R., Abrassart, A., \& Nezi, R. (2021). Beyond {Positive}
+and {Negative}: {New Perspectives} on {Feedback Effects} in {Public
+Opinion} on the {Welfare State}. \emph{British Journal of Political
+Science}, \emph{51}(1), 137--162.
+\url{https://doi.org/10.1017/S0007123418000534}
+
+\bibitem[\citeproctext]{ref-castillo_cual_2009}
+Castillo, J. C. (2009). {{\textquestiondown}Cu{á}l es la brecha salarial
+justa? Opini{ó}n p{ú}blica y legitimaci{ó}n de la desigualdad en Chile}.
+\emph{Estudios P{ú}blicos}, (113).
+
 \bibitem[\citeproctext]{ref-Castillo2011}
@@
 \url{https://doi.org/10.3989/ris.2022.80.3.21.29}
+
+\bibitem[\citeproctext]{ref-garcia-sanchez_attitudes_2020}
+García-Sánchez, E., Osborne, D., Willis, G. B., \& Rodríguez-Bailón, R.
+(2020). Attitudes towards redistribution and the interplay between
+perceptions and beliefs about inequality. \emph{British Journal of
+Social Psychology}, \emph{59}(1), 111--136.
+\url{https://doi.org/10.1111/bjso.12326}
 
@@
 
+\bibitem[\citeproctext]{ref-gijsberts_thelegitimation_1999}
+Gijsberts, M. (1999). \emph{{Thelegitimation of inequality in state-
+socialist and market societies, 1987 - 1996}}. Amsterdam: Thela Thesis.
+
 \bibitem[\citeproctext]{ref-gimpelson_misperceiving_2018}
@@
 \url{https://doi.org/10.1111/ecpo.12103}
+
+\bibitem[\citeproctext]{ref-hadler_why_2005}
+Hadler, M. (2005). Why {Do People Accept Different Income Ratios}?: {A
+Multi-level Comparison} of {Thirty Countries}. \emph{Acta Sociologica},
+\emph{48}(2), 131--154. \url{https://doi.org/10.1177/0001699305053768}
+
+\bibitem[\citeproctext]{ref-hoxMultilevelAnalysisTechniques2017a}
+Hox, J. J., Moerbeek, M., \& Van de Schoot, R. (2017). \emph{Multilevel
+{Analysis}: {Techniques} and {Applications}}.
+
+\bibitem[\citeproctext]{ref-hoyt_mindsets_2023}
+Hoyt, C. L., Burnette, J. L., Billingsley, J., Becker, W., \& Babij, A.
+D. (2023). Mindsets of poverty: {Implications} for redistributive policy
+support. \emph{Analyses of Social Issues and Public Policy},
+\emph{23}(3), 668--693. \url{https://doi.org/10.1111/asap.12367}
 
@@
 
+\bibitem[\citeproctext]{ref-tejero-peregrina_perceived_2025}
+Tejero-Peregrina, L., Willis, G., Sánchez-Rodríguez, Á., \&
+Rodríguez-Bailón, R. (2025). From {Perceived Economic Inequality} to
+{Support} for {Redistribution}: {The Role} of {Meritocracy Perception}.
+\emph{International Review of Social Psychology}, \emph{38}(1), 4.
+\url{https://doi.org/10.5334/irsp.1013}
+
 \bibitem[\citeproctext]{ref-trump_income_2018}
@@
 
+\bibitem[\citeproctext]{ref-vondemknesebeck_are_2016}
+Von Dem Knesebeck, O., Vonneilich, N., \& Kim, T. J. (2016). Are health
+care inequalities unfair? {A} study on public attitudes in 23 countries.
+\emph{International Journal for Equity in Health}, \emph{15}(1), 61.
+\url{https://doi.org/10.1186/s12939-016-0350-8}
+
+\bibitem[\citeproctext]{ref-willis_legitimacy_2015}
+Willis, G. B., Rodríguez-Bailón, R., López-Rodríguez, L., \&
+García-Sánchez, E. (2015). Legitimacy {Moderates} the {Relation Between
+Perceived} and {Ideal Economic Inequalities}. \emph{Social Justice
+Research}, \emph{28}(4), 493--508.
+\url{https://doi.org/10.1007/s11211-015-0253-7}
+
 \bibitem[\citeproctext]{ref-wilson_role_2003}

### 2025-04-28 processing/otros.R

@@ -1,0 +1,30 @@
+elsoc_long_2016_2023 %>% 
+  filter(ola == 1) %>% 
+  select(just_educ, just_pension, just_health) %>% 
+  mutate_all(~ as.numeric(.)) %>% 
+  psych::alpha()
+
+texreg::screenreg(list(m10, m11, m12, m13, m14),
+                  digits = 3)
+
+library(ggeffects)
+
+pred(m11, terms = c("ola_num", 
+                         "merit_effort" )) %>% 
+  plot() 
+
+
+marginaleffects::plot_slopes(model = m11,
+                             variables = "merit_effort",
+                             condition = "ola_num",
+                             conf_level = .95) +
+  geom_hline(yintercept = 0,
+             color = "red",
+             linetype = "dashed") +
+  labs(x = "Meritocracy: Effort [WE]",
+       y = "Marginal effect of meritocracy: effort",
+       fill = NULL,
+       caption = "Fuente: Elaboración propia en base a ISSP (1999-2019) y WID
+       Intervalos de confianza al 95%") +
+  my_pretty_theme() +
+  theme(legend.position = "none")

### 2025-04-29 02-antecedents.qmd
Update 02-antecedents.qmd

@@
----
-author: "Equipo EDUMER"
-bibliography: "input/bib/merit-edjust.bib"
-csl: "input/bib/apa6.csl"
----
-
-# Theoretical views on market justice, inequality perception and meritocracy
+# Theoretical views on market justice, inequality perception, and meritocracy
 
@@
 
-Market justice has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit and effort.
-The most detailed theoretical account of this concept is provided by Lane [-@lane_market_1986], who defines market justice as a system of "earned deserts," whereby individuals are seen as entitled to their outcomes insofar as these are a product of personal productivity and competition.
-This view directly contrasts with political justice, which emphasizes redistributive mechanisms based on the principles of equality and need, and is often materialized through welfare policies.
-For Lane [-@lane_market_1986], market justice relies on the conception that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement.
-What matters is not whether everyone ends up with the same, but that everyone has the chance to compete, and that success reflects effort and ability.
+Conceptually, *market justice* has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit and effort.
+It is possible to trace the concept to the understanding of Lane [-@lane_market_1986], who defines market justice as a system of "earned deserts," whereby individuals are seen as deserving of a determined distributive outcome due to their effort and skills.
+In contrast, political justice,  emphasizes principles of equality and need, which are often represented by the welfare state action through social policies.
+An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement [@lane_market_1986].
 Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open.
-These beliefs align with a broader ideological framework that values individual autonomy, private property, and limited government intervention.
-In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of personal responsibility and societal efficiency.
+In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of individual responsibility.
 
-Empirical studies have developed different strategies to capture and operationalize market justice preferences.
+Empirical studies have shown different strategies for the study of market justice preferences.
 A common approach in the literature is to gauge attitudes toward the legitimacy of inequality in specific domains, especially when linked to income differences.
-Kluegel and Smith [-@kluegel_beliefs_1981] and later Kluegel et al.
-[-@kluegel_legitimation_1999] laid the groundwork for measuring beliefs in market justice through survey items that assess whether people support economic inequality when it reflects personal effort and skills.
+This can be traced to the seminal work of Kluegel and Smith [-@kluegel_beliefs_1981] who assessed the normative foundations that explain public support for economic inequality.
 Over time, this approach has been extended beyond income to include other market-mediated outcomes, such as education, healthcare, and/or pensions.
@@
 These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts.
-Some research, like that of Lindh [-@lindh_public_2015] and Svallfors [-@svallfors_political_2007], uses these indicators in combination to measure broader preferences for market-based distribution in welfare systems.
+Similarly, comparative studies have also relied on these indicators to study the support for market-based distribution in welfare systems [@svallfors_political_2007;@lindh_public_2015].
 More recently, Castillo et al.
 [-@castillo_socialization_2024] introduced a single-item composite measure of market justice to assess student attitudes toward income-based access to education, healthcare, and pensions in Chile.
-These different operationalizations all aim to capture the extent to which individuals accept inequality when it is framed as a reflection of market outcomes.
+These empirical strategies all aim to capture the extent to which individuals accept inequality when framed as a reflection of market outcomes.
 
 The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded.
-Across the literature, there is strong empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher income, education, and occupational status---are more likely to endorse market-based distribution as fair and legitimate [@svallfors_political_2007; @koos_moral_2019].
+Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019].
 This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo.
-At the same time, normative and ideological orientations---such as conservative, liberal, or neoliberal values---are consistently associated with greater support for meritocratic principles and skepticism toward redistribution.
-This is particularly visible in contexts such as Chile and Australia, where studies by Castillo et al.
-[-@castillo_socialization_2024] and Lee and Stacey [-@lee_fairness_2023] show that those with right-leaning political preferences tend to view income-based access to welfare services as legitimate.
-Beyond individual-level characteristics, country-level institutions also play a central role.
+At the same time, political ideology also plays a role ---such as economically conservative values ---  where more right-wing individuals show higher support for meritocracy and more skepticism towards redistribution.
+This is particularly salient in countries with more restricted public provision of social services.
+For example, Castillo et al.
+[-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that those with right-leaning individuals are more supportive of market-based distribution of welfare.
+Beyond individual characteristics, country-level institutions also play a central role.
 In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015].
-These findings suggest that market justice is not a fixed belief but a contextually shaped orientation, mediated by structural conditions, cultural values, and the historical development of welfare institutions.
+These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions.
 
@@
 
-Perceptions of inequality have been associated to attitudes about market justice [@lindh_public_2015; @kluegel_social_1995a].
+Perceptions of inequality have been associated with attitudes about market justice [@lindh_public_2015; @kluegel_social_1995a].
 Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011].
@@
 Unlike objective measures such as the Gini index, perceived inequality captures how individuals make sense of distributive hierarchies in their everyday lives, shaped by reference groups, social comparisons, and information environments [@mijs_stratified_2016; @gimpelson_misperceiving_2018; @garcia-castro_perceiving_2020].
-Scholars have proposed multiple dimensions of perceived inequality, including its magnitude (how significant are the gaps), vertical structure (between which groups), trend over time (increasing or decreasing), and legitimacy (whether it is just or not) [@engelhardt_what_2018; @garcia-sanchez_vicious_2019].
+Scholars have proposed multiple dimensions of perceived inequality, including its magnitude (how significant are the gaps), vertical structure (between which groups), the trend over time (increasing or decreasing), and legitimacy (whether it is just or not) [@engelhardt_what_2018; @garcia-sanchez_vicious_2019].
 These dimensions encompass both cognitive and normative aspects of perceptions of inequality and can vary across societies and social groups, depending on exposure, ideology, and personal experience [@castillo_perception_2022; @garcia-sanchez_perceptions_2018].
@@
 
-Meritocracy constitutes a central ideological framework for legitimizing different types od social inequality, for instance through market justice beliefs.
+Meritocracy constitutes a central ideological framework for legitimizing different types of social inequality, for instance through market justice beliefs.
 Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions.
@@
 
-These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they personally endorse meritocratic principles.
+These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they endorse meritocratic principles.
 This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025] Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity.
@@
 This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds.
-Moreover, @busemeyer_positive_2021 argue that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures.
+Moreover, @busemeyer_positive_2021 argues that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures.
 This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.
@@
 First, consistent with previous findings, we expect that a lower perception of economic inequality will be associated with stronger market justice preferences.
-When individuals perceive smaller economic gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing acceptance of outcomes based on competition and ability to pay.
+When individuals perceive smaller income gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing the acceptance of outcomes based on competition and ability to pay.
 Conversely, a heightened perception of inequality may erode confidence in market fairness, weakening support for market-based distribution.
-This relationship is particularly relevant in the context of Chile, where the neoliberal economic model has been a dominant force shaping public attitudes toward inequality and justice.
+This relationship is particularly relevant in the context of Chile, where the neoliberal economic model has been a dominant force in shaping public attitudes toward inequality and justice.
 
-Second, stronger meritocratic beliefs are expected to be positively associated with market justice preferences.
+Second, higher perceived meritocracy is expected to be positively associated with market justice preferences.
 Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit.
@@
 Third, we propose that perceptions of meritocracy and perceptions of economic inequality interact in shaping market justice attitudes.
-Specifically, we argue that the legitimizing effect of meritocratic beliefs on market justice preferences is moderated by perceptions of economic inequality: when perceived inequality is low, the positive association between meritocratic beliefs and market justice preferences will be stronger.
+Specifically, we argue that the legitimizing effect of perceived meritocracy on market justice preferences is moderated by perceived economic inequality: when perceived inequality is low, the positive association between meritocratic beliefs and market justice preferences will be stronger.
 However, when perceived inequality is high, this association will weaken, as greater awareness of large economic gaps may challenge the view that outcomes are purely merit-based.
@@
 
-$H1$: Greater perceived economic inequality is associated with stronger market justice preferences.
+$H1$: Higher perceived economic inequality is associated with higher market justice preferences.
 
-$H2$: Stronger meritocratic beliefs are associated with stronger market justice preferences.
+$H2$: Higher meritocratic beliefs are associated with higher market justice preferences.
 

### 2025-04-30 03-methods.qmd
Up interacciones WExTime AL

@@
 ```{r}
-
+#| echo: false
 #| include: false
-#| echo: false
 
@@
 rm(list = ls())
-```
-
-```{r}
-#| echo: false
-#| include: false
 
 load(file = here("input/data/proc/df_study1_long_t7.RData"))
-
-glimpse(df_study1_long_t7)
 
@@
   na.omit() %>% 
+  mutate(ola = case_when(ola == 1 ~ 1,
+                         ola == 2 ~ 2, 
+                         ola == 3 ~ 3,
+                         ola == 4 ~ 4,
+                         ola == 6 ~ 5,
+                         ola == 7 ~ 6)) %>% 
   mutate(ola = as.factor(ola),

### 2025-04-30 04-analisis.qmd
Up interacciones WExTime AL

@@
                           ola == 4 ~ "2019",
-                          ola == 6 ~ "2022",
-                          ola == 7 ~ "2023"),
+                          ola == 5 ~ "2022",
+                          ola == 6 ~ "2023"),
          wave = factor(wave, levels = c("2016",
@@
                          ola == 4 ~ "2019",
-                         ola == 6 ~ "2022",
-                         ola == 7 ~ "2023"),
+                         ola == 5 ~ "2022",
+                         ola == 6 ~ "2023"),
          ola = factor(ola, levels = c("2016",
@@
 M <- df_study1 %>%
-  filter(ola == 7) %>% 
+  filter(ola == 6) %>% 
   select(mjp, perc_inequality, merit_effort, merit_talent) %>% 
@@
 Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.042, $p$ < .05).
-In Model 6, the meritocratic belief that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.186, $p$ < .001).
-, whereas the belief that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.040).
+In Model 6, the meritocratic perception that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.186, $p$ < .001), whereas the perception that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.040).
 Notably, once the meritocratic variables are included, the negative coefficient for perceived economic inequality remains but is no longer statistically significant.
@@
                booktabs = F,
-               scalebox = 0.95,
+               scalebox = 0.8,
                include.loglik = FALSE,
@@
 ```
-@tbl-interactions1 reports whether perceived economic inequality moderates the impact of meritocratic beliefs on market-justice preferences.
+
+@tbl-interactions1 reports whether perceived economic inequality moderates the impact of meritocratic perceptions on market-justice preferences.
 In neither the within-person specification (Model 8) nor the between-person specification (Model 9) do the interaction terms reach statistical significance, indicating that higher perceptions of economic inequality do not meaningfully alter the effect of meritocracy on support for market allocation of social services.
 
-With regard to how the effects of perceived economic inequality, meritocratic beliefs, and their interaction evolve over time...
+```{r}
+#| label: fig-growth
+#| out-width: '100%'
+#| fig-asp: 1
+#| fig-cap: "Time varying within effects of meritocratic perceptions on market justice preferences"
+#| fig-cap-location: top
+#| echo: false
+#| results: asis
 
-Valores predichos o marginales??
+library(ggeffects)
+
+p4 <- ggpredict(m11, terms = c("ola_num [all]", "merit_effort")) %>% 
+  plot() + 
+  scale_x_continuous(n.breaks = 6) +
+  labs(x = "Wave",
+       title = "(A) People are rewarded for their efforts",
+       color = NULL) +
+  ggdist::theme_ggdist() +
+  theme(legend.position = "none")
+
+p5 <- ggpredict(m12, terms = c("ola_num [all]", "merit_talent")) %>% 
+  plot() + 
+  scale_x_continuous(n.breaks = 6) +
+  labs(x = "Wave",
+       title = "(B) People are rewarded for their intelligence",
+       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)",
+       color = NULL) +
+  ggdist::theme_ggdist() +
+  theme(legend.position = "bottom")
+
+p4 / p5
+
+```
+
+With respect to the temporal dynamics of the key predictors, @fig-growth illustrates the interaction between survey wave and the within-person effects of the meritocracy variables on market-justice preferences (complete models appear in the Supplementary Material).
+As time advances, the positive within-person effect of perceiving that effort is rewarded (panel A) or that intelligence is rewarded (panel B) progressively attenuates.
+Put differently, the uplift in market justice preferences associated with a strong meritocratic perception becomes smaller in each successive wave.

### 2025-04-30 paper.tex
Up interacciones WExTime AL

@@
 
-\section{Theoretical views on market justice, inequality perception and
+\section{Theoretical views on market justice, inequality perception, and
 meritocracy}\label{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}
@@
 
-Market justice has been discussed in the literature as a normative
-principle that legitimates the distribution of economic rewards based on
-individual merit and effort. The most detailed theoretical account of
-this concept is provided by Lane
+Conceptually, \emph{market justice} has been discussed in the literature
+as a normative principle that legitimates the distribution of economic
+rewards based on individual merit and effort. It is possible to trace
+the concept to the understanding of Lane
 (\citeproc{ref-lane_market_1986}{1986}), who defines market justice as a
-system of ``earned deserts,'' whereby individuals are seen as entitled
-to their outcomes insofar as these are a product of personal
-productivity and competition. This view directly contrasts with
-political justice, which emphasizes redistributive mechanisms based on
-the principles of equality and need, and is often materialized through
-welfare policies. For Lane (\citeproc{ref-lane_market_1986}{1986}),
-market justice relies on the conception that markets are neutral and
-self-regulating arenas, where individuals are treated fairly because
-they face the same formal rules of engagement. What matters is not
-whether everyone ends up with the same, but that everyone has the chance
-to compete, and that success reflects effort and ability. Consequently,
-the legitimacy of market justice stems from the assumption that
-inequalities are not only inevitable but fair---so long as the rules are
-transparent and opportunities are open. These beliefs align with a
-broader ideological framework that values individual autonomy, private
-property, and limited government intervention. In this way, market
-justice provides a moral justification for inequality by framing it as a
-necessary outcome of personal responsibility and societal efficiency.
-
-Empirical studies have developed different strategies to capture and
-operationalize market justice preferences. A common approach in the
-literature is to gauge attitudes toward the legitimacy of inequality in
-specific domains, especially when linked to income differences. Kluegel
-and Smith (\citeproc{ref-kluegel_beliefs_1981}{1981}) and later Kluegel
-et al. (\citeproc{ref-kluegel_legitimation_1999}{1999}) laid the
-groundwork for measuring beliefs in market justice through survey items
-that assess whether people support economic inequality when it reflects
-personal effort and skills. Over time, this approach has been extended
-beyond income to include other market-mediated outcomes, such as
-education, healthcare, and/or pensions. For example, Von Dem Knesebeck
-et al. (\citeproc{ref-vondemknesebeck_are_2016}{2016}) and Immergut and
+system of ``earned deserts,'' whereby individuals are seen as deserving
+of a determined distributive outcome due to their effort and skills. In
+contrast, political justice, emphasizes principles of equality and need,
+which are often represented by the welfare state action through social
+policies. An important remark is that the principles of market justice
+rely on the assumption that markets are neutral and self-regulating
+arenas, where individuals are treated fairly because they face the same
+formal rules of engagement (\citeproc{ref-lane_market_1986}{Lane,
+1986}). Consequently, the legitimacy of market justice stems from the
+assumption that inequalities are not only inevitable but fair---so long
+as the rules are transparent and opportunities are open. In this way,
+market justice provides a moral justification for inequality by framing
+it as a necessary outcome of individual responsibility.
+
+Empirical studies have shown different strategies for the study of
+market justice preferences. A common approach in the literature is to
+gauge attitudes toward the legitimacy of inequality in specific domains,
+especially when linked to income differences. This can be traced to the
+seminal work of Kluegel and Smith
+(\citeproc{ref-kluegel_beliefs_1981}{1981}) who assessed the normative
+foundations that explain public support for economic inequality. Over
+time, this approach has been extended beyond income to include other
+market-mediated outcomes, such as education, healthcare, and/or
+pensions. For example, Von Dem Knesebeck et al.
+(\citeproc{ref-vondemknesebeck_are_2016}{2016}) and Immergut and
 Schneider (\citeproc{ref-immergut_it_2020}{2020}) examine whether
@@
 access to welfare services, allowing for comparing justice perceptions
-across different contexts. Some research, like that of Lindh
-(\citeproc{ref-lindh_public_2015}{2015}) and Svallfors
-(\citeproc{ref-svallfors_political_2007}{2007}), uses these indicators
-in combination to measure broader preferences for market-based
-distribution in welfare systems. More recently, Castillo et al.
+across different contexts. Similarly, comparative studies have also
+relied on these indicators to study the support for market-based
+distribution in welfare systems (\citeproc{ref-lindh_public_2015}{Lindh,
+2015}; \citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). More
+recently, Castillo et al.
 (\citeproc{ref-castillo_socialization_2024}{2024}) introduced a
@@
 attitudes toward income-based access to education, healthcare, and
-pensions in Chile. These different operationalizations all aim to
-capture the extent to which individuals accept inequality when it is
-framed as a reflection of market outcomes.
+pensions in Chile. These empirical strategies all aim to capture the
+extent to which individuals accept inequality when framed as a
+reflection of market outcomes.
 
@@
 beliefs, and the institutional context in which they are embedded.
-Across the literature, there is strong empirical evidence suggesting
-that individuals in more advantaged socioeconomic positions---those with
-higher income, education, and occupational status---are more likely to
-endorse market-based distribution as fair and legitimate
-(\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
-\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). This tendency
-reflects not only material self-interest but also a broader moral
-economy, in which winners of the market system internalize
-justifications for the status quo. At the same time, normative and
-ideological orientations---such as conservative, liberal, or neoliberal
-values---are consistently associated with greater support for
-meritocratic principles and skepticism toward redistribution. This is
-particularly visible in contexts such as Chile and Australia, where
-studies by Castillo et al.
-(\citeproc{ref-castillo_socialization_2024}{2024}) and Lee and Stacey
-(\citeproc{ref-lee_fairness_2023}{2023}) show that those with
-right-leaning political preferences tend to view income-based access to
-welfare services as legitimate. Beyond individual-level characteristics,
+Across the literature, there is empirical evidence suggesting that
+individuals in more advantaged socioeconomic positions---those with
+higher occupational class, income, and education---are more likely to
+support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
+\& Sachweh, 2019}; \citeproc{ref-svallfors_political_2007}{Svallfors,
+2007}). This tendency reflects not only material self-interest but also
+a broader moral economy, in which winners of the market system
+internalize justifications for the status quo. At the same time,
+political ideology also plays a role ---such as economically
+conservative values --- where more right-wing individuals show higher
+support for meritocracy and more skepticism towards redistribution. This
+is particularly salient in countries with more restricted public
+provision of social services. For example, Castillo et al.
+(\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
+Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
+those with right-leaning individuals are more supportive of market-based
+distribution of welfare. Beyond individual characteristics,
 country-level institutions also play a central role. In liberal welfare
@@
 \citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
-that market justice is not a fixed belief but a contextually shaped
-orientation, mediated by structural conditions, cultural values, and the
-historical development of welfare institutions.
+that market justice is also shaped contextually, which is represented by
+the development of welfare institutions.
 
@@
 
-Perceptions of inequality have been associated to attitudes about market
-justice (\citeproc{ref-kluegel_social_1995a}{Kluegel et al., 1995};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research indicates that
-lower perceived inequality can reinforce support for market-based
-distributive arrangements by suggesting that the system is fair and that
-outcomes reflect effort and ability (\citeproc{ref-kuhn_eye_2011}{Kuhn,
-2011}). In contrast, when inequality is perceived as excessive or
-structurally determined, individuals could question the legitimacy of
-market justice and become more supportive of redistributive policies
+Perceptions of inequality have been associated with attitudes about
+market justice (\citeproc{ref-kluegel_social_1995a}{Kluegel et al.,
+1995}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research
+indicates that lower perceived inequality can reinforce support for
+market-based distributive arrangements by suggesting that the system is
+fair and that outcomes reflect effort and ability
+(\citeproc{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when inequality
+is perceived as excessive or structurally determined, individuals could
+question the legitimacy of market justice and become more supportive of
+redistributive policies
 (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
@@
 including its magnitude (how significant are the gaps), vertical
-structure (between which groups), trend over time (increasing or
+structure (between which groups), the trend over time (increasing or
 decreasing), and legitimacy (whether it is just or not)
@@
 Meritocracy constitutes a central ideological framework for legitimizing
-different types od social inequality, for instance through market
+different types of social inequality, for instance through market
 justice beliefs. Rooted in the belief that rewards and positions should
@@
 These dimensions of meritocracy reflect how respondents perceive
-society's distributive logic, regardless of whether they personally
-endorse meritocratic principles. This distinction aligns with recent
-findings indicating that individuals distinguish between how merit is
-perceived in society and how it should ideally operate, which in turn
-shapes their preferences for redistribution and justice
+society's distributive logic, regardless of whether they endorse
+meritocratic principles. This distinction aligns with recent findings
+indicating that individuals distinguish between how merit is perceived
+in society and how it should ideally operate, which in turn shapes their
+preferences for redistribution and justice
 (\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
@@
 toward individuals from lower socioeconomic backgrounds. Moreover,
-Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argue
+Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argues
 that meritocratic narratives can serve as feedback mechanisms that shape
@@
 perception of economic inequality will be associated with stronger
-market justice preferences. When individuals perceive smaller economic
+market justice preferences. When individuals perceive smaller income
 gaps, they are more likely to view market mechanisms as fair and
-legitimate, reinforcing acceptance of outcomes based on competition and
-ability to pay. Conversely, a heightened perception of inequality may
-erode confidence in market fairness, weakening support for market-based
-distribution. This relationship is particularly relevant in the context
-of Chile, where the neoliberal economic model has been a dominant force
-shaping public attitudes toward inequality and justice.
-
-Second, stronger meritocratic beliefs are expected to be positively
+legitimate, reinforcing the acceptance of outcomes based on competition
+and ability to pay. Conversely, a heightened perception of inequality
+may erode confidence in market fairness, weakening support for
+market-based distribution. This relationship is particularly relevant in
+the context of Chile, where the neoliberal economic model has been a
+dominant force in shaping public attitudes toward inequality and
+justice.
+
+Second, higher perceived meritocracy is expected to be positively
 associated with market justice preferences. Individuals who believe that
@@
 economic inequality interact in shaping market justice attitudes.
-Specifically, we argue that the legitimizing effect of meritocratic
-beliefs on market justice preferences is moderated by perceptions of
+Specifically, we argue that the legitimizing effect of perceived
+meritocracy on market justice preferences is moderated by perceived
 economic inequality: when perceived inequality is low, the positive
@@
 
-\(H1\): Greater perceived economic inequality is associated with
-stronger market justice preferences.
-
-\(H2\): Stronger meritocratic beliefs are associated with stronger
+\(H1\): Higher perceived economic inequality is associated with higher
 market justice preferences.
+
+\(H2\): Higher meritocratic beliefs are associated with higher market
+justice preferences.
 
@@
 \end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-554 (37.5\%)\\
-713 (48.3\%)\\
-61 ( 4.1\%)\\
-131 ( 8.9\%)\\
+558 (37.2\%)\\
+729 (48.6\%)\\
+63 ( 4.2\%)\\
+133 ( 8.9\%)\\
 18 ( 1.2\%)\strut
 \end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-1477\\
+1501\\
 (100.0\%)\strut
@@
 \end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-425 (28.8\%)\\
-704 (47.7\%)\\
-106 ( 7.2\%)\\
-219 (14.8\%)\\
-23 ( 1.6\%)\strut
+426 (28.4\%)\\
+718 (47.8\%)\\
+108 ( 7.2\%)\\
+226 (15.1\%)\\
+23 ( 1.5\%)\strut
 \end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-1477\\
+1501\\
 (100.0\%)\strut
@@
 \end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-519 (35.1\%)\\
-764 (51.7\%)\\
-71 ( 4.8\%)\\
-112 ( 7.6\%)\\
+521 (34.7\%)\\
+783 (52.2\%)\\
+73 ( 4.9\%)\\
+113 ( 7.5\%)\\
 11 ( 0.7\%)\strut
 \end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-1477\\
+1501\\
 (100.0\%)\strut
@@
 1 \textless{} 2 \textless{} 5\\
-IQR (CV) : 1 (0.4)\strut
+IQR (CV) : 0.7 (0.4)\strut
 \end{minipage} & 12 distinct values &
 \begin{minipage}[t]{\linewidth}\raggedright
-1477\\
+1501\\
 (100.0\%)\strut
@@
 \hline
-Intercept                            & $1.968^{***}$  & $1.973^{***}$  & $1.985^{***}$  & $1.982^{***}$  & $2.122^{***}$  & $1.408^{***}$  & $1.499^{***}$  \\
-                                     & $(0.021)$      & $(0.032)$      & $(0.032)$      & $(0.032)$      & $(0.069)$      & $(0.096)$      & $(0.114)$      \\
+Intercept                            & $1.968^{***}$  & $2.015^{***}$  & $2.030^{***}$  & $2.028^{***}$  & $2.169^{***}$  & $1.456^{***}$  & $1.547^{***}$  \\
+                                     & $(0.021)$      & $(0.035)$      & $(0.035)$      & $(0.035)$      & $(0.070)$      & $(0.097)$      & $(0.115)$      \\
 Wave (Ref.= 2016)                    &                &                &                &                &                &                &                \\
@@
                                      & $(0.027)$      &                &                &                &                &                &                \\
-Wave                                 &                & $-0.075^{***}$ & $-0.078^{***}$ & $-0.079^{***}$ & $-0.079^{***}$ & $-0.079^{***}$ & $-0.079^{***}$ \\
-                                     &                & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
-Wave$^2$                             &                & $0.017^{***}$  & $0.018^{***}$  & $0.018^{***}$  & $0.018^{***}$  & $0.018^{***}$  & $0.018^{***}$  \\
-                                     &                & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      & $(0.002)$      \\
-Perception inequality (WE)           &                &                & $-0.034^{***}$ & $-0.029^{**}$  & $-0.029^{**}$  & $-0.029^{**}$  & $-0.029^{**}$  \\
+Wave                                 &                & $-0.121^{***}$ & $-0.125^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.129^{***}$ \\
+                                     &                & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      \\
+Wave$^2$                             &                & $0.028^{***}$  & $0.028^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  \\
+                                     &                & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      \\
+Perception inequality (WE)           &                &                & $-0.039^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ \\
                                      &                &                & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
-Merit: Effort (WE)                   &                &                &                & $0.071^{***}$  & $0.071^{***}$  & $0.070^{***}$  & $0.070^{***}$  \\
+Merit: Effort (WE)                   &                &                &                & $0.070^{***}$  & $0.070^{***}$  & $0.069^{***}$  & $0.069^{***}$  \\
                                      &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
-Merit: Talent (WE)                   &                &                &                & $0.028^{*}$    & $0.028^{*}$    & $0.028^{*}$    & $0.027^{*}$    \\
+Merit: Talent (WE)                   &                &                &                & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    \\
                                      &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
-Perception inequality (BE)           &                &                &                &                & $-0.042^{*}$   & $-0.005$       & $-0.032$       \\
+Perception inequality (BE)           &                &                &                &                & $-0.042^{*}$   & $-0.006$       & $-0.033$       \\
                                      &                &                &                &                & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
@@
 Controls                             & No             & No             & No             & No             & No             & No             & Yes            \\
-BIC                                  & $20780.594$    & $20757.596$    & $20761.847$    & $20704.882$    & $20714.843$    & $20633.143$    & $20727.029$    \\
+BIC                                  & $20780.594$    & $20780.617$    & $20781.113$    & $20726.753$    & $20736.627$    & $20655.172$    & $20749.047$    \\
 Num. obs.                            & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         \\
 Num. groups: idencuesta              & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         \\
-Var: idencuesta (Intercept)          & $0.172$        & $0.207$        & $0.204$        & $0.198$        & $0.196$        & $0.172$        & $0.160$        \\
-Var: Residual                        & $0.530$        & $0.497$        & $0.497$        & $0.493$        & $0.493$        & $0.492$        & $0.493$        \\
-Var: idencuesta ola\_num             & $$             & $0.007$        & $0.007$        & $0.006$        & $0.006$        & $0.006$        & $0.006$        \\
-Cov: idencuesta (Intercept) ola\_num & $$             & $-0.017$       & $-0.017$       & $-0.014$       & $-0.014$       & $-0.013$       & $-0.013$       \\
+Var: idencuesta (Intercept)          & $0.172$        & $0.223$        & $0.219$        & $0.209$        & $0.207$        & $0.182$        & $0.171$        \\
+Var: Residual                        & $0.530$        & $0.499$        & $0.498$        & $0.495$        & $0.495$        & $0.495$        & $0.495$        \\
+Var: idencuesta ola\_num             & $$             & $0.011$        & $0.010$        & $0.009$        & $0.009$        & $0.009$        & $0.009$        \\
+Cov: idencuesta (Intercept) ola\_num & $$             & $-0.025$       & $-0.024$       & $-0.021$       & $-0.020$       & $-0.019$       & $-0.019$       \\
 \hline
@@
 (\(\beta\) = -0.042, \(p\) \textless{} .05). In Model 6, the
-meritocratic belief that effort is rewarded is positively associated
+meritocratic perception that effort is rewarded is positively associated
 with market justice preferences (\(\beta\) = 0.186, \(p\) \textless{}
-.001). , whereas the belief that talent is rewarded shows a positive but
-non-significant coefficient (\(\beta\) = 0.040). Notably, once the
+.001), whereas the perception that talent is rewarded shows a positive
+but non-significant coefficient (\(\beta\) = 0.040). Notably, once the
 meritocratic variables are included, the negative coefficient for
@@
 \begin{center}
-\scalebox{0.95}{
+\scalebox{0.8}{
 \begin{tabular}{l c c}
@@
 \hline
-Intercept                                                & $1.511^{***}$ & $2.309^{***}$ \\
-                                                         & $(0.114)$     & $(0.287)$     \\
-Merit: Effort (WE) x Perception inequality (WE)          & $0.018$       &               \\
+Intercept                                                & $1.558^{***}$ & $2.361^{***}$ \\
+                                                         & $(0.115)$     & $(0.287)$     \\
+Merit: Effort (WE) x Perception inequality (WE)          & $0.019$       &               \\
                                                          & $(0.016)$     &               \\
@@
                                                          &               & $(0.045)$     \\
-Merit: Talent (BE) x Perception inequality (BE)          &               & $0.069$       \\
+Merit: Talent (BE) x Perception inequality (BE)          &               & $0.070$       \\
                                                          &               & $(0.045)$     \\
@@
 Controls                                                 & Yes           & Yes           \\
-BIC                                                      & $20858.742$   & $20799.038$   \\
+BIC                                                      & $20879.829$   & $20819.684$   \\
 Num. obs.                                                & $8643$        & $8643$        \\
@@
 Var: idencuesta (Intercept)                              & $0.147$       & $0.141$       \\
-Var: idencuesta merit\_effort\_cwc                       & $0.017$       & $$            \\
+Var: idencuesta merit\_effort\_cwc                       & $0.016$       & $$            \\
 Var: idencuesta merit\_talent\_cwc                       & $0.018$       & $$            \\
@@
 Cov: idencuesta merit\_talent\_cwc perc\_inequality\_cwc & $0.001$       & $$            \\
-Var: Residual                                            & $0.501$       & $0.527$       \\
+Var: Residual                                            & $0.503$       & $0.528$       \\
 \hline
@@
 Table~\ref{tbl-interactions1} reports whether perceived economic
-inequality moderates the impact of meritocratic beliefs on
+inequality moderates the impact of meritocratic perceptions on
 market-justice preferences. In neither the within-person specification
@@
 
-With regard to how the effects of perceived economic inequality,
-meritocratic beliefs, and their interaction evolve over time\ldots{}
-
-Valores predichos o marginales??
+\begin{figure}[H]
+
+\caption{\label{fig-growth}Time varying within effects of meritocratic
+perceptions on market justice preferences}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-growth-1.pdf}
+
+}
+
+\end{figure}%
+
+With respect to the temporal dynamics of the key predictors,
+Figure~\ref{fig-growth} illustrates the interaction between survey wave
+and the within-person effects of the meritocracy variables on
+market-justice preferences (complete models appear in the Supplementary
+Material). As time advances, the positive within-person effect of
+perceiving that effort is rewarded (panel A) or that intelligence is
+rewarded (panel B) progressively attenuates. Put differently, the uplift
+in market justice preferences associated with a strong meritocratic
+perception becomes smaller in each successive wave.
 

### 2025-05-03 04-analisis.qmd
update interaction effects

@@
 
+df_study1 <- df_study1 %>%
+  mutate(
+    perc_inequality_cwc_dic = ifelse(perc_inequality_cwc > mean(perc_inequality_cwc, na.rm = TRUE), "mayor", "menor"),
+    merit_effort_cwc_dic = ifelse(merit_effort_cwc > mean(merit_effort_cwc, na.rm = TRUE), "mayor", "menor"),
+    merit_talent_cwc_dic = ifelse(merit_talent_cwc > mean(merit_talent_cwc, na.rm = TRUE), "mayor", "menor"),
+    perc_inequality_mean_dic = ifelse(perc_inequality_mean > mean(perc_inequality_mean, na.rm = TRUE), "mayor", "menor"),
+    merit_effort_mean_dic = ifelse(merit_effort_mean > mean(merit_effort_mean, na.rm = TRUE), "mayor", "menor"),
+    merit_talent_mean_dic = ifelse(merit_talent_mean > mean(merit_talent_mean, na.rm = TRUE), "mayor", "menor")
+  ) %>%
+  mutate(across(ends_with("_dic"), factor, levels = c("menor", "mayor")))
+
 ## WE and BE
@@
 # interactions 
+m8 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+             merit_effort_cwc_dic*perc_inequality_cwc_dic + merit_talent_cwc*perc_inequality_cwc +
+              (1 + merit_effort_cwc + merit_talent_cwc + perc_inequality_cwc | idencuesta),
+                data = df_study1)
 
-m8 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
-             merit_effort_cwc*perc_inequality_cwc + merit_talent_cwc*perc_inequality_cwc +
-              (1 + merit_effort_cwc + merit_talent_cwc + perc_inequality_cwc | idencuesta),
+m8.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+             merit_effort_cwc*perc_inequality_cwc +
+              (1 + merit_effort_cwc + perc_inequality_cwc | idencuesta),
+                data = df_study1)
+
+m8.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+             merit_talent_cwc*perc_inequality_cwc +
+              (1 + merit_talent_cwc + perc_inequality_cwc | idencuesta),
                 data = df_study1)
@@
              merit_effort_mean*perc_inequality_mean + merit_talent_mean*perc_inequality_mean +
+              (1 | idencuesta),
+                data = df_study1)
+
+m9.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+             merit_effort_mean*perc_inequality_mean +
+              (1 | idencuesta),
+                data = df_study1)
+
+m9.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+             merit_talent_mean*perc_inequality_mean +
               (1 | idencuesta),
@@
 
+m10.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc*ola_num + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality_cwc + ola_num| idencuesta),
+                data = df_study1)
+
+m10.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean*ola_num + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality_mean + ola_num| idencuesta),
+                data = df_study1)
+
 m11 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + merit_effort*ola_num  + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort + ola_num| idencuesta),
@@
 
-m12 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality +  merit_effort + merit_talent*ola_num + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+m11.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc*ola_num  + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort_cwc + ola_num| idencuesta),
+                data = df_study1)
+
+m11.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc  + merit_talent_cwc + perc_inequality_mean + merit_effort_mean*ola_num + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort_mean + ola_num| idencuesta),
+                data = df_study1)
+
+m12 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort + merit_talent*ola_num + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
               (1 + merit_talent + ola_num| idencuesta),
+                data = df_study1)
+
+m12.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc*ola_num + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+              (1 + merit_talent_cwc + ola_num| idencuesta),
+                data = df_study1)
+
+m12.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  merit_effort_mean + merit_talent_mean*ola_num + educ + quintil1 + ess + ideo + sex + age +
+              (1 + merit_talent_mean + ola_num| idencuesta),
                 data = df_study1)
@@
 
+m13.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc  + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+             merit_effort_cwc*perc_inequality_cwc*ola_num +
+              (1 + perc_inequality_cwc + merit_effort_cwc + ola_num| idencuesta),
+                data = df_study1)
+
+m13.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc  + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+             ola_num*merit_effort_mean*perc_inequality_mean +
+              (1 + perc_inequality_mean + merit_effort_mean  + ola_num| idencuesta),
+                data = df_study1)
+
 m14 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality +  merit_effort + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + 
@@
 
+m14.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+              merit_talent_cwc*perc_inequality_cwc*ola_num +
+              (1 + perc_inequality_cwc + merit_talent_cwc + ola_num| idencuesta),
+                data = df_study1)
+
+m14.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+              merit_talent_mean*perc_inequality_mean*ola_num +
+              (1 + perc_inequality_mean + merit_talent_mean + ola_num| idencuesta),
+                data = df_study1)
+
+#knitreg(list(m13.1, m13.2,m14.1, m14.2))
+```
+
+```{r}
+interactions::interact_plot(m13.2, pred = "merit_effort_mean", modx="perc_inequality_mean", mod2="ola_num")
 ```
@@
 
-texreg::texreg(list(m1, m2, m3, m4, m5, m6, m7),
+texreg::knitreg(list(m1, m2, m3, m4, m5, m6, m7),
                custom.model.names = c(paste0("Model ", seq(1:7))),
@@
 
-texreg::texreg(list(m8,m9),
-               custom.model.names = c("Model 8", "Model 9"),
+texreg::knitreg(list(m8.1,m8.2,m9.1,m9.2),
+               custom.model.names = c("Model 8", "Model 9", "Model 10", "Model 11"),
                caption = NULL,
@@
                center = T,
-               custom.gof.rows = list("Controls"=c(rep("Yes",2))))
+               custom.gof.rows = list("Controls"=c(rep("Yes",4))))
+
+```
+
+```{r}
+#| label: fig-interact
+#| out-width: '100%'
+#| fig-asp: 1
+#| fig-cap: "Within effects of perception of inequality and meritocratic perceptions on market justice preferences"
+#| fig-cap-location: top
+#| echo: false
+#| results: asis
+
+library(ggeffects)
+
+p6 <- ggpredict(m9.1, terms = c("merit_effort_mean", "perc_inequality_mean")) %>% 
+  plot() + 
+ # scale_x_continuous(n.breaks = 6) +
+  scale_color_manual(
+    name = "Perception\nInequality (BE)",
+    values = c("#D55E00", "#0072B2", "#009E73"),
+    labels = c("-1 SD", "Mean", "+1 SD")
+  ) +
+  scale_fill_manual(
+    name = "",
+    values = c("#D55E00", "#0072B2", "#009E73"),
+    labels = c("-1 SD", "Mean", "+1 SD")) +
+  labs(x = "Merit: Effort (BE)",
+       title = "(A) People are rewarded for their efforts",
+       color = NULL) +
+  ggdist::theme_ggdist() +
+  theme(legend.position = "bottom")
+
+p7 <- ggpredict(m9.2, terms = c("merit_talent_mean", "perc_inequality_mean")) %>% 
+  plot() +
+#  scale_x_continuous(n.breaks = 6) +
+  scale_color_manual(
+    name = "Perception\nInequality (BE)",
+    values = c("#D55E00", "#0072B2", "#009E73"),
+    labels = c("-1 SD", "Mean", "+1 SD")
+  ) +
+  scale_fill_manual(
+    name = "",
+    values = c("#D55E00", "#0072B2", "#009E73"),
+    labels = c("-1 SD", "Mean", "+1 SD")
+  ) +
+  labs(
+    x = "Merit: Talent (BE)",
+    title = "(B) People are rewarded for their intelligence",
+    caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)"
+  ) +
+  ggdist::theme_ggdist() +
+  theme(legend.position = "bottom")
+
+ggpubr::ggarrange(p6, p7, common.legend=TRUE, ncol=1)
 
@@
 @tbl-interactions1 reports whether perceived economic inequality moderates the impact of meritocratic perceptions on market-justice preferences.
-In neither the within-person specification (Model 8) nor the between-person specification (Model 9) do the interaction terms reach statistical significance, indicating that higher perceptions of economic inequality do not meaningfully alter the effect of meritocracy on support for market allocation of social services.
+The interaction terms in the within-person specification (Model 8 and Model 9) do not reach statistical significance.
+However, in the between-person specification (Model 10 and Model 11), interaction terms reach statistical significance, indicating that perceptions of economic inequality do meaningfully alter the effect of meritocracy on support for market allocation of social services.
+These effects are positive and mean that when the perception of inequality is greater, the positive effect of the meritocratic variables on market justice preferences is greater.
+These interaction effects are visualized in @fig-interact
+
+```{r echo=FALSE, results='asis'}
+#| label: tbl-interactions2
+#| tbl-cap: "Interactions for meritocracy, perceived economic inequality and market justice preferences"
+#| tbl-cap-location: top
+
+ccoef <- list(
+  "(Intercept)" = "Intercept",
+  "ola_num:perc_inequality_cwc" = "Wave x Perception inequality (WE)",
+  "ola_num:merit_effort_cwc" = "Wave x Merit: effort (WE)",
+  "ola_num:merit_talent_cwc" = "Wave x Merit: talent (WE)",
+  "ola_num:perc_inequality_mean" = "Wave x Perception inequality (BE)",
+  "ola_num:merit_effort_mean" = "Wave x Merit: effort (BE)",
+  "ola_num:merit_talent_mean" = "Wave x Merit: talent (BE)"
+    )
+
+texreg::knitreg(list(m10.1, m11.1, m12.1, m10.2, m11.2, m12.2),
+               custom.model.names = c("Model 12", "Model 13", "Model 14", "Model 15", "Model 16", "Model 17"),
+               caption = NULL,
+               stars = c(0.05, 0.01, 0.001),
+               custom.coef.map = ccoef,
+               digits = 3,
+               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
+               leading.zero = T,
+               use.packages = F,
+               booktabs = F,
+               scalebox = 0.8,
+               include.loglik = FALSE,
+               include.aic = FALSE,
+               center = T,
+               custom.gof.rows = list("Controls"=c(rep("Yes",6))))
+
+```
 
@@
 
-p4 <- ggpredict(m11, terms = c("ola_num [all]", "merit_effort")) %>% 
+p4 <- ggpredict(m11.2, terms = c("ola_num [all]", "merit_effort_mean")) %>% 
   plot() + 
   scale_x_continuous(n.breaks = 6) +
+  scale_color_manual(
+    name = "Merit:\nEffort",
+    values = c("#D55E00", "#0072B2", "#009E73"),
+    labels = c("-1 SD", "Mean", "+1 SD")
+  ) +
+  scale_fill_manual(
+    name = "Merit:\nEffort",
+    values = c("#D55E00", "#0072B2", "#009E73"),
+    labels = c("-1 SD", "Mean", "+1 SD")) +
   labs(x = "Wave",
@@
   ggdist::theme_ggdist() +
-  theme(legend.position = "none")
+  theme(legend.position = "top")
 
-p5 <- ggpredict(m12, terms = c("ola_num [all]", "merit_talent")) %>% 
+p5 <- ggpredict(m12.2, terms = c("ola_num [all]", "merit_talent_mean")) %>% 
   plot() +
   scale_x_continuous(n.breaks = 6) +
-  labs(x = "Wave",
+  scale_color_manual(
+    name = "Merit:\nTalent",
+    values = c("#D55E00", "#0072B2", "#009E73"),
+    labels = c("-1 SD", "Mean", "+1 SD")
+  ) +
+  scale_fill_manual(
+    name = "Merit:\nTalent",
+    values = c("#D55E00", "#0072B2", "#009E73"),
+    labels = c("-1 SD", "Mean", "+1 SD")
+  ) +
+  labs(
+    x = "Wave",
     title = "(B) People are rewarded for their intelligence",
-       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)",
-       color = NULL) +
+    caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)"
+  ) +
   ggdist::theme_ggdist() +
-  theme(legend.position = "bottom")
+  theme(legend.position = "top")
 
-p4 / p5
+ggpubr::ggarrange(p4, p5, common.legend=FALSE, ncol=1)
 
@@
 
-With respect to the temporal dynamics of the key predictors, @fig-growth illustrates the interaction between survey wave and the within-person effects of the meritocracy variables on market-justice preferences (complete models appear in the Supplementary Material).
+With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time only has significant effects on the between effects of the meritocratic variables (Model 16 and Model 17).
+These interaction effects are negative and mean that in the last waves, the impact of meritocratic variables on market justice preferences is attenuated.
+@fig-growth illustrates the interaction between survey wave and the within-person effects of the meritocracy variables on market-justice preferences.
 As time advances, the positive within-person effect of perceiving that effort is rewarded (panel A) or that intelligence is rewarded (panel B) progressively attenuates.

### 2025-05-21 02-antecedents.qmd
update interaction effects
review - not rendered

@@
 
-Conceptually, *market justice* has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit and effort.
-It is possible to trace the concept to the understanding of Lane [-@lane_market_1986], who defines market justice as a system of "earned deserts," whereby individuals are seen as deserving of a determined distributive outcome due to their effort and skills.
-In contrast, political justice,  emphasizes principles of equality and need, which are often represented by the welfare state action through social policies.
+Conceptually, *market justice* has been discussed in the literature as a normative principle that legitimates the distribution of economic rewards based on individual merit.
+It is possible to trace the concept to the understanding of Lane [-@lane_market_1986], who makes a contrast between market justice and political justice.
+The author defines market justice as a system of "earned deserts," whereby individuals are seen as deserving of a determined distributive outcome due to their effort and skills.
+In contrast, political justice emphasizes principles of equality and need, which are often represented by the welfare state action through social policies.
 An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement [@lane_market_1986].
@@
 These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts.
-Similarly, comparative studies have also relied on these indicators to study the support for market-based distribution in welfare systems [@svallfors_political_2007;@lindh_public_2015].
+Similarly, comparative studies have also considered these indicators to study the support for market-based distribution in welfare systems [@svallfors_political_2007;@lindh_public_2015].
 More recently, Castillo et al.
@@
 In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015].
-These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions.
+These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions, in line with the policy-feedbacks literature [campbell_policy_2012; pierson_when_1993].
 
@@
 Additionally, this study examines whether these relationships vary over time in response to significant social and political events.
-In particular, we explore whether the political outburst of 2019 and the subsequent constitutional processes in Chile have altered how perceptions of inequality and meritocracy shape market justice attitudes.
+In particular, we explore whether the political outburst of 2019 and the subsequent constitutional processes in Chile --- where socio-economic inequality was widely challenged --- have altered how perceptions of inequality and meritocracy shape market justice attitudes.
 We expect that after these critical events, perceptions of inequality may have a stronger negative effect on market justice preferences, reflecting increased societal questioning of market-based allocation mechanisms.
@@
 
-$H1$: Higher perceived economic inequality is associated with higher market justice preferences.
+$H1$: Higher perceived economic inequality is associated with less market justice preferences.
 

### 2025-05-21 03-methods.qmd
review - not rendered

@@
 **Market justice preferences**: The dependent variable in this study is preferences for market justice.
-This construct is operationalized through three items that capture how strongly individuals justify conditioning access to social services---healthcare, pensions, and education---on income.
+This construct is operationalized through three items that capture how strongly individuals justify conditioning access to social services---healthcare, pensions, and education--- basen on individual income.
 Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is posed for pensions and education.
@@
 
-**Perception of economic inequality**: The main independent variable refers to the perception of economic inequality, measured through the perceived wage gap.
-[@castillo_cual_2009; @hadler_why_2005; @gijsberts_thelegitimation_1999].
+**Perception of economic inequality**: The main independent variable refers to the perception of economic inequality, measured through the perceived wage gap [@castillo_cual_2009; @hadler_why_2005; @gijsberts_thelegitimation_1999].
 This measure is derived from the salary gap between the perceived salaries of jobs at opposite ends of the occupational hierarchy.
@@
 #| label: tbl-summary2
-#| tbl-cap: "Independent variables for the first wave (2016)"
+#| tbl-cap: "Independent variables ELSOC survey (descriptives for first wave 2016)"
 #| tbl-cap-location: top
@@
 In terms of sociodemographic characteristics, we incorporate per capita household income quintile, educational level (1=Less than Universitary, 2=Universitary), age (in years), and sex (1=Male, 2=Female), which have been previously shown to influence market justice preferences significantly [@lindh_public_2015; @castillo_socialization_2024].
-Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) as they may confound the relationship between market justice preferences and perceptions of inequality and meritocracy [@schneider_poverty_2015].
+Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) as they may affect the relationship between market justice preferences, perceptions of inequality, and meritocracy [@schneider_poverty_2015].
 

### 2025-05-21 04-analisis.qmd
review - not rendered

@@
 Overall, more than half of the respondents exhibit a high level of disagreement (disagree + strongly disagree) with inequality in these three social service areas over time.
-Despite this general pattern, recent waves show a slight decrease in disagreement and a corresponding rise in support for inequality.
+Despite this general pattern, recent waves show a slight decrease in disagreement and a corresponding rise in support for market-justice inequality.
 Specifically, in healthcare and education, although disagreement remains substantial, agreement (agree + strongly agree) increased from 7.4% and 7.2% in 2019 to 13.1% and 14.2% in 2023, respectively.
@@
                                          "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"))) %>% 
+                                         "Neither agree nor disagree",
+                                         "Disagree",
+                                         "Strongly disagree"))) %>% 
   group_by(idencuesta, ola) %>% 
@@
                                          "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"))) %>% 
+                                         "Neither agree nor disagree",
+                                         "Disagree",
+                                         "Strongly disagree"))) %>% 
   group_by(ola, just_health) %>%
@@
                                          "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"))) %>% 
+                                         "Neither agree nor disagree",
+                                         "Disagree",
+                                         "Strongly disagree"))) %>% 
   group_by(ola, just_pension) %>%
@@
                                          "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"))) %>%
+                                         "Neither agree nor disagree",
+                                         "Disagree",
+                                         "Strongly disagree"))) %>%
   group_by(idencuesta, ola) %>% 
@@
                                          "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"))) %>%
+                                         "Neither agree nor disagree",
+                                         "Disagree",
+                                         "Strongly disagree"))) %>%
   group_by(ola, just_educ) %>%
@@
 
-Regarding the main variables of this study, @fig-meanchange depicts their average changes over the years.
+Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their average changes over the years.
 We observe an increase in the average level of market justice preferences in the most recent waves, which begins in 2019.
@@
 @tbl-interactions1 reports whether perceived economic inequality moderates the impact of meritocratic perceptions on market-justice preferences.
-The interaction terms in the within-person specification (Model 8 and Model 9) do not reach statistical significance.
+Contrary to our hypothesis, the interaction terms in the within-person specification (Model 8 and Model 9) do not reach statistical significance.
 However, in the between-person specification (Model 10 and Model 11), interaction terms reach statistical significance, indicating that perceptions of economic inequality do meaningfully alter the effect of meritocracy on support for market allocation of social services.

### 2025-05-21 05-discussion.qmd
review - not rendered

---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
---

# Discussion

This study tests several key hypotheses concerning the relationships between perceptions of economic inequality, meritocratic beliefs, and attitudes toward market justice in Chile. It posits that higher perceived inequality and stronger meritocratic beliefs both independently contribute to greater support for market-based distributions of social goods. Furthermore, the study examines how perceived economic inequality moderates the influence of meritocratic beliefs on market justice preferences. Additionally, it considers how major social and political mobilizations, such as those occurring around 2019 and subsequent years, may have weakened these relationships, reflecting increased critical perspectives on systemic inequalities and market mechanisms. The subsequent discussion interprets the empirical findings in light of these hypotheses, considering their implications for understanding attitudes toward inequality and social justice in a context marked by profound social change.

The first hypothesis suggested that perceptions of economic inequality were strongly linked to attitudes toward market justice. The findings demonstrated that individuals perceiving higher income disparities showed stronger support for the distribution of social services based on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies (Castillo et al., 2012; Mijs, 2021). This relationship reflected a moral economy where perceptions of systemic unfairness threatened the legitimacy of existing distributions, prompting calls for greater equity. Interestingly, the negative within-effects observed over time implied that as perceptions of inequality became more salient, some individuals experienced a decline in support for market justice, possibly due to increasing distrust in market mechanisms or disillusionment with the fairness of the system. This indicated that while perception heightened awareness, prolonged exposure or social mobilization could also foster skepticism about whether redistribution effectively addressed inequality or simply perpetuated systemic flaws.

Regarding the second hypothesis, the results reinforced that meritocratic perceptions ---particularly those emphasizing individual effort and talent--- were associated with stronger support for market-based distribution systems. Individuals who viewed success as primarily earned through effort were more inclined to justify existing inequalities, seeing them as outcomes of individual merit rather than systemic injustice. This aligned with research indicating that meritocratic narratives served as moral justifications that reinforced societal stratification (Castillo et al., 2012; Hoyt et al., 2023). These beliefs functioned symbolically to legitimize structural disparities, reducing support for redistributive measures by framing inequality as fair and deserved. The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes.

The third hypothesis concerned the moderating role of perceptions of inequality on the relationship between meritocracy and market justice attitudes. The analysis indicated that the positive association between meritocratic beliefs and support for market-based distribution tended to weaken when perceptions of economic inequality were particularly high. This pattern suggested that high awareness of disparities could challenge the moral high ground of meritocracy, as individuals might question whether success was truly based on effort or whether systemic biases distorted outcomes. Although the moderation was not consistently statistically significant across all models (i.e.within subjects effects), the observed trend aligned with theoretical expectations that perceptions of inequality undermined the legitimacy of meritocratic narratives and fragmented support for existing economic arrangements. This interaction revealed the complex psychological landscape in which deeply held beliefs about effort and reward could be destabilized by heightened awareness of systemic disparities, especially during times of social upheaval or mobilization.

Finally, the fourth hypothesis addressed the impact of recent social movements, particularly from 2019 onwards, on the relationship between meritocratic beliefs, perceptions of inequality, and attitudes toward market justice. The results suggested that these movements contributed to a more critical stance toward the fairness of existing distributional mechanisms. Post-mobilization, the positive relationship between meritocratic beliefs and support for inequality appeared attenuated, indicating that collective action and public discourse during this period fostered questioning of the legitimacy of systemic inequalities justified solely through meritocratic narratives. These social mobilizations challenged dominant narratives, emphasizing structural barriers and systemic injustices, which in turn weakened the association between individual effort and deservingness. This is aligned with literature emphasizing that social protests act as catalysts for transforming societal attitudes by disrupting normative beliefs and encouraging more critical evaluations of institutional legitimacy [@tilly_contentious_2015]. The findings underscored the importance of socio-political context in shaping and reshaping attitudes toward inequality and justice, highlighting how collective action could temporarily or permanently alter the normative foundations that sustained support for market justice.

### 2025-05-21 06-conclusion.qmd
review - not rendered

---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
---

# Conclusions

This study examined the complex interplay between perceptions of economic inequality, meritocratic beliefs, and attitudes toward market justice in Chile from 2016 to 2023, drawing on longitudinal data from the ELSOC survey. By exploring how subjective assessments and social contexts influence support for redistribution and market-based resource allocation, the research offers different elements that contribute to the understanding of the normative foundations underpinning social justice attitudes in a highly unequal, neoliberal environment.

The findings confirm that higher perceptions of economic inequality are associated with increased support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health. Meritocratic beliefs strongly reinforce support for market justice, serving as moral justifications for structural inequalities. Additionally, perceptions of inequality moderate the influence of meritocratic beliefs, such that high awareness of disparities can weaken support for existing market-based allocations. Notably, the social mobilizations of 2019 and subsequent periods seem to have contributed to shifting attitudes, fostering greater skepticism of systemic fairness and reducing the legitimizing power of meritocratic narratives.

This research advances the extant literature by integrating subjective perceptions with social and political contexts to explain attitudes toward economic inequality and distributional justice. While previous studies primarily focused on objective measures or individual characteristics, this work emphasizes the dynamic and interactional nature of perceptions and beliefs over time. It highlights the importance of socio-political upheavals in reshaping normative attitudes, underscoring the role of collective action in challenging entrenched narratives of meritocracy and fairness. The longitudinal approach provides a deeper temporal perspective on how societal events could influence individual beliefs and preferences.

Despite its contributions, the study has several limitations. First, the reliance on self-reported perceptions and attitudes may be subject to social desirability bias. Second, the focus on Chile, while offering valuable insights into a specific context, limits the generalizability of results to other countries with different institutional and cultural backgrounds. Third, the observational nature of the data prevents causal inferences, and unobserved confounding factors may influence the observed relationships. Finally, the potential influence of media, political discourse, and peer effects on perceptions was not directly assessed.

Future research could explore the causal mechanisms linking perceptions and attitudes, possibly through experimental or mixed-method designs. Comparative studies across different national contexts would help determine the universality or specificity of these dynamics. Additionally, investigating the role of media, political communication, and education in shaping perceptions of inequality and meritocracy would deepen understanding of the normative foundations of social justice attitudes. Finally, examining how these perceptions influence behavioral outcomes, such as political participation or support for social movements, would provide valuable insights into the pathways from beliefs to collective action and policy change.

### 2025-05-21 paper.qmd
review - not rendered

--- 2025-04-23 paper.qmd
@@
 format:
-  docx:
-    number-sections: true
   html: 
@@
     html-math-method: katex
+    number-sections: true
+  docx:
     number-sections: true
@@
 abstract: | 
-  My abstract. \newline
-  **Keywords**: meritocracy, social inequality, inequality justification, COVID-19
+  This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and preferences for market justice in Chile between 2016 and 2023.
+Using six waves of panel data from the Chilean Longitudinal Social Survey (ELSOC; N = 1,519), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions.
+Results indicate that rising perceptions of inequality are associated with greater support for redistributive policies; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities.
+The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness.
+These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
+  \newline
+  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
 link-citations: true

### 2025-05-21 04-analisis.qmd
Up graficos interaccion, tablas, citas y material suplementario + render

@@
 csl: "input/bib/apa6.csl"
+editor_options: 
+  chunk_output_type: console
 ---
@@
                sjlabelled,
-               summarytools)
+               summarytools,
+               interactions)
 
@@
 load(file = here("input/data/proc/df_study1_long_t7.RData"))
-
-glimpse(df_study1_long_t7)
 
@@
                                          "Agree",
-                                         "Neither agree nor disagree",
-                                         "Disagree",
-                                         "Strongly disagree"))) %>% 
+                                         "Neither agree nor desagree",
+                                         "Desagree",
+                                         "Strongly desagree"),
+                              ordered = T)) %>% 
   group_by(idencuesta, ola) %>% 
@@
                                          "Agree",
-                                         "Neither agree nor disagree",
-                                         "Disagree",
-                                         "Strongly disagree"))) %>% 
+                                         "Neither agree nor desagree",
+                                         "Desagree",
+                                         "Strongly desagree"),
+                              ordered = T)) %>% 
   group_by(ola, just_health) %>%
@@
                                          "Agree",
-                                         "Neither agree nor disagree",
-                                         "Disagree",
-                                         "Strongly disagree"))) %>% 
+                                         "Neither agree nor desagree",
+                                         "Desagree",
+                                         "Strongly desagree"),
+                              ordered = T)) %>% 
   group_by(ola, just_pension) %>%
@@
                                          "Agree",
-                                         "Neither agree nor disagree",
-                                         "Disagree",
-                                         "Strongly disagree"))) %>%
+                                         "Neither agree nor desagree",
+                                         "Desagree",
+                                         "Strongly desagree"),
+                              ordered = T)) %>%
   group_by(idencuesta, ola) %>% 
@@
                                          "Agree",
-                                         "Neither agree nor disagree",
-                                         "Disagree",
-                                         "Strongly disagree"))) %>%
+                                         "Neither agree nor desagree",
+                                         "Desagree",
+                                         "Strongly desagree"),
+                              ordered = T)) %>%
   group_by(ola, just_educ) %>%
@@
 
-```{r}
+```{r, include=FALSE, eval=FALSE}
 interactions::interact_plot(m13.2, pred = "merit_effort_mean", modx="perc_inequality_mean", mod2="ola_num")
@@
                center = T,
+               custom.gof.names = c("BIC", "Numb. obs.", "Num. groups: individuals", "Var: individuals (Intercept)", "Var: Residual", "Var: individuals, wave", "Cov: individuals (Intercept), wave"),
                custom.gof.rows = list("Controls"=c(rep("No",6), "Yes")))
@@
                digits = 3,
-               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
+               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.",
                leading.zero = T,
@@
                center = T,
+               custom.gof.names = c("BIC", 
+                                    "Numb. obs.", 
+                                    "Num. groups: individuals", 
+                                    "Var: individuals (Intercept)", 
+                                    "Var: individuals, merit effort cwc", 
+                                    "Var: individuals, perception inequality cwc", 
+                                    "Cov: individuals (Intercept), merit effort cwc",
+                                    "Cov: individuals (Intercept), perception inequality cwc",
+                                    "Cov: individuals, merit effort cwc, perception inequality cwc",
+                                    "Var: Residuals",
+                                    "Var: individuals, merit talent cwc", 
+                                    "Cov: individuals (Intercept), merit talent cwc",
+                                    "Cov: individuals, merit talent cwc, perception inequality cwc"),
                custom.gof.rows = list("Controls"=c(rep("Yes",4))))
@@
 #| fig-asp: 1
-#| fig-cap: "Within effects of perception of inequality and meritocratic perceptions on market justice preferences"
+#| fig-cap: "Between effects of meritocratic perceptions on market justice preferences by economic inequality perception"
 #| fig-cap-location: top
@@
 
-p6 <- ggpredict(m9.1, terms = c("merit_effort_mean", "perc_inequality_mean")) %>% 
-  plot() + 
- # scale_x_continuous(n.breaks = 6) +
-  scale_color_manual(
-    name = "Perception\nInequality (BE)",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  scale_fill_manual(
-    name = "",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")) +
-  labs(x = "Merit: Effort (BE)",
-       title = "(A) People are rewarded for their efforts",
-       color = NULL) +
-  ggdist::theme_ggdist() +
-  theme(legend.position = "bottom")
+p6 <- marginaleffects::plot_slopes(model = m9.1,
+                             variables = "merit_effort_mean",
+                             condition = "perc_inequality_mean",
+                             conf_level = .95) +
+  geom_hline(yintercept = 0,
+             color = "red",
+             linetype = "dashed") +
+  labs(x = "Perception inequality (BE)",
+       y = "Marginal effect of Merit: Effort (BE)",
+       fill = NULL,
+       title = "(A) People are rewarded by their efforts") +
+  theme_ggdist()
 
-p7 <- ggpredict(m9.2, terms = c("merit_talent_mean", "perc_inequality_mean")) %>% 
-  plot() +
-#  scale_x_continuous(n.breaks = 6) +
-  scale_color_manual(
-    name = "Perception\nInequality (BE)",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  scale_fill_manual(
-    name = "",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  labs(
-    x = "Merit: Talent (BE)",
+p7 <- marginaleffects::plot_slopes(model = m9.2,
+                             variables = "merit_talent_mean",
+                             condition = "perc_inequality_mean",
+                             conf_level = .95) +
+  geom_hline(yintercept = 0,
+             color = "red",
+             linetype = "dashed") +
+  labs(x = "Perception inequality (BE)",
+       y = "Marginal effect of Merit: Talent (BE)",
+       fill = NULL,
        title = "(B) People are rewarded for their intelligence",
-    caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)"
-  ) +
-  ggdist::theme_ggdist() +
-  theme(legend.position = "bottom")
+       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)
+       Confidence intervals at 95%") +
+  theme_ggdist()
 
+library(ggpubr)
 ggpubr::ggarrange(p6, p7, common.legend=TRUE, ncol=1)
@@
 
-@tbl-interactions1 reports whether perceived economic inequality moderates the impact of meritocratic perceptions on market-justice preferences.
-Contrary to our hypothesis, the interaction terms in the within-person specification (Model 8 and Model 9) do not reach statistical significance.
-However, in the between-person specification (Model 10 and Model 11), interaction terms reach statistical significance, indicating that perceptions of economic inequality do meaningfully alter the effect of meritocracy on support for market allocation of social services.
-These effects are positive and mean that when the perception of inequality is greater, the positive effect of the meritocratic variables on market justice preferences is greater.
-These interaction effects are visualized in @fig-interact
+@tbl-interactions1 examines whether perceived economic inequality moderates the effect of meritocratic beliefs on market justice preferences.
+Contrary to our hypothesis, the interaction terms in the within-person specification (Models 8 and 9) do not reach statistical significance.
+However, in the between-person specification (Models 10 and 11), the interaction terms are statistically significant, suggesting that perceptions of economic inequality meaningfully alter the effect of meritocracy on support for market-based allocation of social services.
+These positive effects indicate that, as perceptions of inequality increase, the positive association between meritocratic perceptions and market justice preferences becomes stronger.
+The marginal effects of this interaction are illustrated in @fig-interact, which shows that the effect of meritocratic perceptions on market justice preferences intensifies as perceived economic inequality rises.
 
@@
                center = T,
+               custom.gof.names = c("BIC", 
+                                    "Numb. obs.", 
+                                    "Num. groups: individuals", 
+                                    "Var: individuals (Intercept)",
+
+                                    "Var: individuals, perception inequality cwc",
+                                    "Var: individuals, wave",
+                                    "Cov: individuals (Intercept), perception inequality cwc",
+                                    "Cov: individuals (Intercept), wave",
+                                    "Cov: individuals, perception inequality cwc, wave",
+                                    "Var: Residuals",
+                                    "Var: individuals, merit effort cwc", 
+                                    "Cov: individuals (Intercept), merit effort cwc",
+                                    "Cov: individuals, merit effort cwc, wave",
+
+                                    "Var: individuals, merit talent cwc", 
+                                    "Cov: individuals (Intercept), merit talent cwc",
+                                    "Cov: individuals, merit talent cwc, wave",
+
+                                    "Var: individuals, perception inequality mean", 
+                                    "Cov: individuals (Intercept), perception inequality mean",
+                                    "Cov: individuals, perception inequality mean, wave",
+
+                                     "Var: individuals, merit effort mean", 
+                                    "Cov: individuals (Intercept), merit effort mean",
+                                    "Cov: individuals, merit effort mean, wave",
+
+                                    "Var: individuals, merit talent mean", 
+                                    "Cov: individuals (Intercept), merit talent mean",
+                                    "Cov: individuals, merit talent mean, wave"
+                                    ),
                custom.gof.rows = list("Controls"=c(rep("Yes",6))))
@@
 #| results: asis
+#| include: false
+#| eval: false
 
@@
 
-With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time only has significant effects on the between effects of the meritocratic variables (Model 16 and Model 17).
+With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time only has significant effects on the between effects of the meritocratic variables.
 These interaction effects are negative and mean that in the last waves, the impact of meritocratic variables on market justice preferences is attenuated.
-@fig-growth illustrates the interaction between survey wave and the within-person effects of the meritocracy variables on market-justice preferences.
-As time advances, the positive within-person effect of perceiving that effort is rewarded (panel A) or that intelligence is rewarded (panel B) progressively attenuates.
+Specifically, as time advances, the positive within-person effect of perceiving that effort is rewarded (Model 16) or that intelligence is rewarded (Model 17) progressively attenuates.
 Put differently, the uplift in market justice preferences associated with a strong meritocratic perception becomes smaller in each successive wave.

### 2025-05-21 05-discussion.qmd
Up graficos interaccion, tablas, citas y material suplementario + render

@@
 The first hypothesis suggested that perceptions of economic inequality were strongly linked to attitudes toward market justice.
-The findings demonstrated that individuals perceiving higher income disparities showed stronger support for the distribution of social services based on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies (Castillo et al., 2012; Mijs, 2021).
+The findings demonstrated that individuals perceiving higher income disparities showed stronger support for the distribution of social services based on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021].
 This relationship reflected a moral economy where perceptions of systemic unfairness threatened the legitimacy of existing distributions, prompting calls for greater equity.
@@
 Individuals who viewed success as primarily earned through effort were more inclined to justify existing inequalities, seeing them as outcomes of individual merit rather than systemic injustice.
-This aligned with research indicating that meritocratic narratives served as moral justifications that reinforced societal stratification (Castillo et al., 2012; Hoyt et al., 2023).
+This aligned with research indicating that meritocratic narratives served as moral justifications that reinforced societal stratification [@Castillo2012a_justice; @hoyt_mindsets_2023].
 These beliefs functioned symbolically to legitimize structural disparities, reducing support for redistributive measures by framing inequality as fair and deserved.
-The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes.
+The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes [@canalesceron_sujeto_2021].
 

### 2025-05-21 07-supplementary-material.qmd
Up graficos interaccion, tablas, citas y material suplementario + render

--- 2025-05-03 04-analisis.qmd
@@
 ---
-
 ```{r}
@@
 ```{r}
-#| label: packages
 #| echo: false
@@
 ```{r}
-#| label: data
 #| echo: false
@@
 ```
-# Results
-
-## Descriptive
-
-@fig-alluvial shows the annual frequencies of market justice preferences for healthcare, pensions, and education from 2016 to 2023.
-Each year presents stacked percentage frequencies, and the flows between them reflect opinion changes among the same individuals from one year to the next, given that we are using panel data.
-For instance, of the 40.8% who strongly disagreed with justifying inequality in healthcare in 2019, around 24.3% maintained that position in 2022, while the remaining 16.5% shifted toward other response categories—primarily moving into disagreement rather than strong disagreement.
-Overall, more than half of the respondents exhibit a high level of disagreement (disagree + strongly disagree) with inequality in these three social service areas over time.
-Despite this general pattern, recent waves show a slight decrease in disagreement and a corresponding rise in support for inequality.
-Specifically, in healthcare and education, although disagreement remains substantial, agreement (agree + strongly agree) increased from 7.4% and 7.2% in 2019 to 13.1% and 14.2% in 2023, respectively.
-This shift is most evident in pensions, where the combined agree/strongly agree category grew by about 10 percentage points, from 16.9% in 2016 to 28% in 2023.
-
-```{r}
-#| label: fig-alluvial
-#| out-width: '100%'
-#| fig-asp: 1.2
-#| fig-cap: "Change in the justification of inequality in healthcare, pensions and education over time (2016-2023)"
-#| fig-cap-location: top
-#| echo: false
-#| results: asis
-
-datos.health <- df_study1 %>% 
-  mutate(just_health = factor(just_health, 
-                              levels = c("Strongly agree",
-                                         "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"))) %>% 
-  group_by(idencuesta, ola) %>% 
-  count(just_health) %>% 
-  group_by(ola) %>% 
-  mutate(porcentaje=n/sum(n)) %>% 
-  ungroup() %>% 
-  na.omit() %>% 
-  mutate(wave = case_when(ola == 1 ~ "2016",
-                          ola == 2 ~ "2017",
-                          ola == 3 ~ "2018",
-                          ola == 4 ~ "2019",
-                          ola == 5 ~ "2022",
-                          ola == 6 ~ "2023"),
-         wave = factor(wave, levels = c("2016",
-                                      "2017",
-                                      "2018",
-                                      "2019",
-                                      "2022",
-                                      "2023")))
-
-etiquetas.health <- df_study1 %>%
-  mutate(just_health = factor(just_health, 
-                              levels = c("Strongly agree",
-                                         "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"))) %>% 
-  group_by(ola, just_health) %>%
-  summarise(count = n(), .groups = "drop") %>%
-  group_by(ola) %>%
-  mutate(porcentaje = count / sum(count)) %>% 
-  na.omit() %>% 
-  mutate(idencuesta = 1,
-         wave = case_when(ola == 1 ~ "2016",
-                          ola == 2 ~ "2017",
-                          ola == 3 ~ "2018",
-                          ola == 4 ~ "2019",
-                          ola == 5 ~ "2022",
-                          ola == 6 ~ "2023"),
-         wave = factor(wave, levels = c("2016",
-                                        "2017",
-                                        "2018",
-                                        "2019",
-                                        "2022",
-                                        "2023")))
-
-p1 <- datos.health %>% 
-  ggplot(aes(x = wave, fill = just_health, stratum = just_health,
-             alluvium = idencuesta, y = porcentaje)) +
-  ggalluvial::geom_flow(alpha = .4) + 
-  ggalluvial::geom_stratum(linetype = 0) +
-  scale_y_continuous(labels = scales::percent) + 
-  scale_fill_manual(values =  c("#0571B0","#92C5DE","#b3b3b3ff","#F4A582","#CA0020")) +
-  geom_shadowtext(data = etiquetas.health,
-                  aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
-                  position = position_stack(vjust = .5),
-                  show.legend = FALSE,
-                  size = 3,
-                  color = rep('white'),
-                  bg.colour='grey30')+
-  labs(y = "%",
-       x = NULL,
-       fill = NULL,
-       title = "a. Healthcare") +
-  theme_ggdist() +
-  theme(legend.position = "none")
-
-datos.pension <- df_study1 %>% 
-   mutate(just_pension = factor(just_pension, 
-                              levels = c("Strongly agree",
-                                         "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"))) %>% 
-  group_by(idencuesta, ola) %>% 
-  count(just_pension) %>% 
-  group_by(ola) %>% 
-  mutate(porcentaje=n/sum(n)) %>% 
-  ungroup() %>% 
-  na.omit() %>% 
-  mutate(wave = case_when(ola == 1 ~ "2016",
-                          ola == 2 ~ "2017",
-                          ola == 3 ~ "2018",
-                          ola == 4 ~ "2019",
-                          ola == 5 ~ "2022",
-                          ola == 6 ~ "2023"),
-         wave = factor(wave, levels = c("2016",
-                                        "2017",
-                                        "2018",
-                                        "2019",
-                                        "2022",
-                                        "2023")))
-
-etiquetas.pension <- df_study1 %>%
-  mutate(just_pension = factor(just_pension, 
-                              levels = c("Strongly agree",
-                                         "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"))) %>% 
-  group_by(ola, just_pension) %>%
-  summarise(count = n(), .groups = "drop") %>%
-  group_by(ola) %>%
-  mutate(porcentaje = count / sum(count)) %>% 
-  na.omit() %>% 
-  mutate(idencuesta = 1,
-         wave = case_when(ola == 1 ~ "2016",
-                          ola == 2 ~ "2017",
-                          ola == 3 ~ "2018",
-                          ola == 4 ~ "2019",
-                          ola == 5 ~ "2022",
-                          ola == 6 ~ "2023"),
-         wave = factor(wave, levels = c("2016",
-                                        "2017",
-                                        "2018",
-                                        "2019",
-                                        "2022",
-                                        "2023")))
-
-p2 <- datos.pension %>% 
-  ggplot(aes(x = wave, fill = just_pension, stratum = just_pension,
-             alluvium = idencuesta, y = porcentaje)) +
-  ggalluvial::geom_flow(alpha = .4) + 
-  ggalluvial::geom_stratum(linetype = 0) +
-  scale_y_continuous(labels = scales::percent) + 
-  scale_fill_manual(values =  c("#0571B0","#92C5DE","#b3b3b3ff","#F4A582","#CA0020")) +
-  geom_shadowtext(data = etiquetas.pension,
-                  aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
-                  position = position_stack(vjust = .5),
-                  show.legend = FALSE,
-                  size = 3,
-                  color = rep('white'),
-                  bg.colour='grey30')+
-  labs(y = "%",
-       x = NULL,
-       fill = NULL,
-       title = "b. Pensions") +
-  theme_ggdist() +
-  theme(legend.position = "none")
-
-datos.educ <- df_study1 %>% 
-  mutate(just_educ = factor(just_educ, 
-                              levels = c("Strongly agree",
-                                         "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"))) %>%
-  group_by(idencuesta, ola) %>% 
-  count(just_educ) %>% 
-  group_by(ola) %>% 
-  mutate(porcentaje=n/sum(n)) %>% 
-  ungroup() %>% 
-  na.omit() %>% 
-  mutate(wave = case_when(ola == 1 ~ "2016",
-                          ola == 2 ~ "2017",
-                          ola == 3 ~ "2018",
-                          ola == 4 ~ "2019",
-                          ola == 5 ~ "2022",
-                          ola == 6 ~ "2023"),
-         wave = factor(wave, levels = c("2016",
-                                        "2017",
-                                        "2018",
-                                        "2019",
-                                        "2022",
-                                        "2023")))
-
-etiquetas.educ <- df_study1 %>%
-    mutate(just_educ = factor(just_educ, 
-                              levels = c("Strongly agree",
-                                         "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"))) %>%
-  group_by(ola, just_educ) %>%
-  summarise(count = n(), .groups = "drop") %>%
-  group_by(ola) %>%
-  mutate(porcentaje = count / sum(count)) %>% 
-  na.omit() %>% 
-  mutate(idencuesta = 1,
-         wave = case_when(ola == 1 ~ "2016",
-                          ola == 2 ~ "2017",
-                          ola == 3 ~ "2018",
-                          ola == 4 ~ "2019",
-                          ola == 5 ~ "2022",
-                          ola == 6 ~ "2023"),
-         wave = factor(wave, levels = c("2016",
-                                        "2017",
-                                        "2018",
-                                        "2019",
-                                        "2022",
-                                        "2023")))
-
-p3 <- datos.educ %>% 
-  ggplot(aes(x = wave, fill = just_educ, stratum = just_educ,
-             alluvium = idencuesta, y = porcentaje)) +
-  ggalluvial::geom_flow(alpha = .4) + 
-  ggalluvial::geom_stratum(linetype = 0) +
-  scale_y_continuous(labels = scales::percent) + 
-  scale_fill_manual(values =  c("#0571B0","#92C5DE","#b3b3b3ff","#F4A582","#CA0020")) +
-  geom_shadowtext(data = etiquetas.educ,
-                  aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
-                  position = position_stack(vjust = .5),
-                  show.legend = FALSE,
-                  size = 3,
-                  color = rep('white'),
-                  bg.colour='grey30')+
-  labs(y = "%",
-       x = "Wave",
-       fill = NULL,
-       title = "c. Education",
-       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)") +
-  theme_ggdist() +
-  theme(legend.position = "bottom")
-
-(p1 / p2 / p3)
-
-```
-
-Regarding the main variables of this study, @fig-meanchange depicts their average changes over the years.
-We observe an increase in the average level of market justice preferences in the most recent waves, which begins in 2019.
-The highest average consistently appears for perceived economic inequality, although this variable shows a downward trend of roughly one point over time.
-Interestingly, while perceptions of economic inequality declined in the latest measurements (2022–2023), market justice preferences increased.
-The meritocracy measures remain stable, though the perception that individuals are rewarded for intelligence is slightly higher than the perception that they are rewarded for their effort.
-
-```{r}
-#| label: fig-meanchange
-#| out-width: '100%'
-#| fig-asp: 0.8
-#| fig-cap: "Change in the mean of market justice preferences, economic inequality perception, and meritocracy (2016-2023)"
-#| fig-cap-location: top
-#| echo: false
-#| results: asis
-
-library(srvyr)
-
-df_pond <- df_study1 %>% 
-  mutate(ola = case_when(ola == 1 ~ "2016",
-                         ola == 2 ~ "2017",
-                         ola == 3 ~ "2018",
-                         ola == 4 ~ "2019",
-                         ola == 5 ~ "2022",
-                         ola == 6 ~ "2023"),
-         ola = factor(ola, levels = c("2016",
-                                        "2017",
-                                        "2018",
-                                        "2019",
-                                        "2022",
-                                        "2023"))) %>% 
-  as_survey_design(.data = .,
-                   ids = segmento, 
-                   strata = estrato, 
-                   weights = ponderador_long_total)
-
-df_pond %>% 
-  select(ola, mjp, perc_inequality, merit_effort, merit_talent) %>% 
-  mutate_at(.vars = 2:5, .funs = ~ as.numeric(.)) %>% 
-  group_by(ola) %>% 
-  summarise_all(~survey_mean(., vartype = "ci")) %>% 
-  pivot_longer(cols = -ola,
-               names_to = "temp",
-               values_to = "valor") %>%
-  mutate(
-    ci = case_when(
-      str_ends(temp, "_low") ~ "ic_low",
-      str_ends(temp, "_upp") ~ "ic_upp",
-      TRUE                   ~ "mean"
-    ),
-    variable = str_remove(temp, "_low|_upp")
-  ) %>%
-  select(ola, variable, ci, valor) %>%
-  pivot_wider(
-    names_from  = ci,
-    values_from = valor
-  ) %>% 
-  mutate(variable = case_when(variable == "mjp" ~ "Market justice preferences",
-                              variable == "perc_inequality"  ~ "Inequality perception",
-                              variable == "merit_effort"     ~ "Merit: Effort",
-                              variable == "merit_talent"     ~ "Merit: Talent"),
-         variable = factor(variable, levels = c("Market justice preferences",
-                                      "Inequality perception",
-                                      "Merit: Effort",
-                                      "Merit: Talent")
-                              )) %>% 
-  ggplot(aes(x = ola, y = mean, group = variable)) +
-  geom_point(aes(shape=variable, color=variable), size = 3.5) +
-  geom_line(aes(color = variable), linewidth = 0.8) +
-  geom_errorbar(aes(ymin = ic_low, ymax = ic_upp, color = variable),
-                width = 0.1) +
-  scale_color_manual(values = c("#023858", # azul oscuro
-                                "#0571B0", # azul intermedio
-                                "#F4A582", # salmón
-                                "#B2182B"  # rojo profundo
-                                         )) +
-  scale_shape_manual(values=c(15, 16, 17, 18)) +
-  labs(y = "Mean",
-       x = "Wave",
-       color = NULL,
-       shape = NULL,
-       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)") +
-  theme_ggdist() +
-  theme(legend.position = "top",
-        text = element_text(size = 12))
-
-```
-
-@fig-matrix presents the correlation matrix for the main variables in the latest wave (2023).
-Overall, the coefficients range from low to moderate.
-The association between market justice preferences and economic inequality perception is negative but small and statistically significant ($r$ = -0.05, $p$ < .05).
-By contrast, market justice preferences positively and significantly correlate with the two meritocracy variables ($r$ = 0.15, $p$ < .01; $r$ = 0.13, $p$ < .01).
-Perceived economic inequality has negative and nonsignificant correlations with both meritocracy perceptions ($r$ = -0.04, $p$ > .05; $r$ = -0.04, $p$ > .05).
-Finally, the two meritocracy variables exhibit a strong positive association with each other ($r$ = 0.69, $p$ < .01).
-
-```{r}
-#| label: fig-matrix
-#| out-width: '90%'
-#| fig-asp: 1
-#| fig-cap: "Correlation matrix of the main variables for the last wave (2023)"
-#| fig-cap-location: top
-#| echo: false
-#| results: asis
-
-M <- df_study1 %>%
-  filter(ola == 6) %>% 
-  select(mjp, perc_inequality, merit_effort, merit_talent) %>% 
-  mutate_all(.funs = ~ as.numeric(.)) %>% 
-  cor(use = "pairwise.complete.obs", method = "pearson")
-
-diag(M) <- NA
-
-rownames(M) <- c("A. Marjet justice preferences",
-                 "B. Inequality gap perception",
-                 "C. People are rewarded for their efforts ",
-                 "D. People are rewarded for their intelligence")
-
-#set Column names of the matrix
-colnames(M) <-c("(A)", "(B)","(C)","(D)")
-
-#Plot the matrix using corrplot
-corrplot::corrplot(M,
-                   method = "color",
-                   addCoef.col = "black",
-                   type = "upper",
-                   tl.col = "black",
-                   col = colorRampPalette(c("#E16462", "white", "#0D0887"))(12),
-                   bg = "white",
-                   na.label = "-")
-
-```
-
-## Multilevel models
-
-@tbl-modelos presents the results of the multilevel models estimated for market justice preferences, examining both individuals (within) and group-level (between) effects.
-The intraclass correlation [@hoxMultilevelAnalysisTechniques2017a] from the empty model (see Supplementary Material), which decomposes the variance of market justice preferences, is 0.23, indicating that approximately 23% of the variation is attributable to differences between individuals.
-Complementary, 77% of the variation corresponds to within-individual differences over time.
 
@@
 
-```{r}
-interactions::interact_plot(m13.2, pred = "merit_effort_mean", modx="perc_inequality_mean", mod2="ola_num")
-```
+# Supplementary material
 
 ```{r echo=FALSE, results='asis'}
-#| label: tbl-modelos
-#| tbl-cap: "Longitudinal multilevel models for market justice preferences"
+#| label: tbl-suple1
+#| tbl-cap: "Complete longitudinal multilevel models for market justice preferences"
 #| tbl-cap-location: top
-
-omit <- "(educ)|(quintil1)|(ess)|(ideo)|(sex)|(age)"
 
@@
   merit_effort_mean = "Merit: Effort (BE)",
-  merit_talent_mean = "Merit: Talent (BE)"
+  merit_talent_mean = "Merit: Talent (BE)",
+  "educUniversitary" = "Universitary education (Ref.= Less than Universitary)",
+  quintil1Q2 = "Quintile Q2",
+  quintil1Q3 = "Quintile Q3",
+  quintil1Q4 = "Quintile Q4",
+  quintil1Q5 = "Quintile Q5",
+  quintil1QNA = "Quintile no information",
+  ess = "Subjective social status",
+  ideoCenter = "Center",
+  ideoRight = "Right",
+  "ideoDoes not identify" = "Does not identify",
+  sexFemale = "Female (Ref.= Male)",
+  "age30-49" = "Age 30-49",
+  "age50-64" = "Age 50-64",
+  "age65 or more" = "Age 65 or more"
   )
 
-texreg::knitreg(list(m1, m2, m3, m4, m5, m6, m7),
-               custom.model.names = c(paste0("Model ", seq(1:7))),
+texreg::knitreg(list(m0, m1, m2, m3, m4, m5, m6, m7),
+               custom.model.names = c("Model 0",paste0("Model ", seq(1:7))),
                caption = NULL,
                stars = c(0.05, 0.01, 0.001),
-               omit.coef = omit,
                custom.coef.map = ccoef,
                digits = 3,
-               groups = list("Wave (Ref.= 2016)" = 2:6),
+               groups = list("Wave (Ref.= 2016)" = 2:6,
+                             "Income quintile (Ref.= Quintile 1)" = 16:20,
+                             "Political identification (Ref.= Left)" = 21:23,
+                             "Age (Ref.= 18-29)" = 25:27),
                custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
@@
                booktabs = F,
-               scalebox = 0.79,
+               scalebox = 0.7,
                include.loglik = FALSE,
@@
                center = T,
-               custom.gof.rows = list("Controls"=c(rep("No",6), "Yes")))
+               custom.gof.names = c("BIC", "Numb. obs.", "Num. groups: individuals", "Var: individuals (Intercept)", "Var: Residual", "Var: individuals, wave", "Cov: individuals (Intercept), wave"))
 
 ```
-
-According to Model 1, which includes the survey waves to capture intertemporal variations in the dependent variable, there is a decrease in 2017 ($\beta$ = -0.167, $p$ < .001) relative to 2016, and similarly in 2018 ($\beta$ = -0.025, $p$ > .05) and 2019 ($\beta$ = -0.046, $p$ > .05), although the latter effects are not statistically significant.
-In contrast, in the more recent waves of 2022 and 2023, there is a statistically significant increase in market justice preferences ($\beta$ = 0.285, $p$ < .001; $\beta$ = 0.284, $p$ < .001), suggesting a non-linear effect.
-To model this trajectory over time, Model 2 incorporates time (survey waves) as a continuous variable, along with its quadratic term, representing the non-linear association initially observed in Model 1.
-While the linear term (survey wave) shows a negative association, reflecting an overall decline in market preferences over time, the positive quadratic term indicates a reversal of this pattern in the final measurement points.
-
-Models 3 and 4 incorporate the within-group effects (WE) of the primary independent variables, capturing how individual changes in these variables over time shape the dependent variable.
-The results in Model 3 suggest that the within effect of perceived economic inequality is negative and statistically significant ($p$ < .001).
-Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.034 point decrease in market justice preferences.
-Model 4 shows that meritocratic beliefs operate in the opposite direction.
-An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.071, $p$ < .001), and a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with higher market-justice preferences ($\beta$ = 0.028, $p$ < .05).
-Taken together, these results suggest that individuals who increasingly perceive meritocracy—whether through effort or talent—tend to hold stronger market justice preferences.
-
-When examining the between-group effects (BE) in Model 5 and 6, which capture differences between individuals in the average of the main variables, a similar pattern emerges.
-Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.042, $p$ < .05).
-In Model 6, the meritocratic perception that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.186, $p$ < .001), whereas the perception that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.040).
-Notably, once the meritocratic variables are included, the negative coefficient for perceived economic inequality remains but is no longer statistically significant.
-
-Model 7 adds the control variables.
-With the exception of the between-effect for perceived economic inequality—which becomes nonsignificant—the within- and between-effects of the principal predictors retain both their direction and statistical significance, confirming the robustness of the associations (see Supplementary Material for effects of control variables).
-
-```{r echo=FALSE, results='asis'}
-#| label: tbl-interactions1
-#| tbl-cap: "Interactions for meritocracy, perceived economic inequality and market justice preferences"
-#| tbl-cap-location: top
-
-ccoef <- list(
-  "(Intercept)" = "Intercept",
-  "perc_inequality_cwc:merit_effort_cwc" = "Merit: Effort (WE) x Perception inequality (WE)",
-  "perc_inequality_cwc:merit_talent_cwc" = "Merit: Talent (WE) x Perception inequality (WE)",
-  "perc_inequality_mean:merit_effort_mean" = "Merit: Effort (BE) x Perception inequality (BE)",
-  "perc_inequality_mean:merit_talent_mean" = "Merit: Talent (BE) x Perception inequality (BE)"
-  )
-
-texreg::knitreg(list(m8.1,m8.2,m9.1,m9.2),
-               custom.model.names = c("Model 8", "Model 9", "Model 10", "Model 11"),
-               caption = NULL,
-               stars = c(0.05, 0.01, 0.001),
-               custom.coef.map = ccoef,
-               digits = 3,
-               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
-               leading.zero = T,
-               use.packages = F,
-               booktabs = F,
-               scalebox = 0.8,
-               include.loglik = FALSE,
-               include.aic = FALSE,
-               center = T,
-               custom.gof.rows = list("Controls"=c(rep("Yes",4))))
-
-```
-
-```{r}
-#| label: fig-interact
-#| out-width: '100%'
-#| fig-asp: 1
-#| fig-cap: "Within effects of perception of inequality and meritocratic perceptions on market justice preferences"
-#| fig-cap-location: top
-#| echo: false
-#| results: asis
-
-library(ggeffects)
-
-p6 <- ggpredict(m9.1, terms = c("merit_effort_mean", "perc_inequality_mean")) %>% 
-  plot() + 
- # scale_x_continuous(n.breaks = 6) +
-  scale_color_manual(
-    name = "Perception\nInequality (BE)",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  scale_fill_manual(
-    name = "",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")) +
-  labs(x = "Merit: Effort (BE)",
-       title = "(A) People are rewarded for their efforts",
-       color = NULL) +
-  ggdist::theme_ggdist() +
-  theme(legend.position = "bottom")
-
-p7 <- ggpredict(m9.2, terms = c("merit_talent_mean", "perc_inequality_mean")) %>% 
-  plot() +
-#  scale_x_continuous(n.breaks = 6) +
-  scale_color_manual(
-    name = "Perception\nInequality (BE)",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  scale_fill_manual(
-    name = "",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  labs(
-    x = "Merit: Talent (BE)",
-    title = "(B) People are rewarded for their intelligence",
-    caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)"
-  ) +
-  ggdist::theme_ggdist() +
-  theme(legend.position = "bottom")
-
-ggpubr::ggarrange(p6, p7, common.legend=TRUE, ncol=1)
-
-```
-
-@tbl-interactions1 reports whether perceived economic inequality moderates the impact of meritocratic perceptions on market-justice preferences.
-The interaction terms in the within-person specification (Model 8 and Model 9) do not reach statistical significance.
-However, in the between-person specification (Model 10 and Model 11), interaction terms reach statistical significance, indicating that perceptions of economic inequality do meaningfully alter the effect of meritocracy on support for market allocation of social services.
-These effects are positive and mean that when the perception of inequality is greater, the positive effect of the meritocratic variables on market justice preferences is greater.
-These interaction effects are visualized in @fig-interact
-
-```{r echo=FALSE, results='asis'}
-#| label: tbl-interactions2
-#| tbl-cap: "Interactions for meritocracy, perceived economic inequality and market justice preferences"
-#| tbl-cap-location: top
-
-ccoef <- list(
-  "(Intercept)" = "Intercept",
-  "ola_num:perc_inequality_cwc" = "Wave x Perception inequality (WE)",
-  "ola_num:merit_effort_cwc" = "Wave x Merit: effort (WE)",
-  "ola_num:merit_talent_cwc" = "Wave x Merit: talent (WE)",
-  "ola_num:perc_inequality_mean" = "Wave x Perception inequality (BE)",
-  "ola_num:merit_effort_mean" = "Wave x Merit: effort (BE)",
-  "ola_num:merit_talent_mean" = "Wave x Merit: talent (BE)"
-    )
-
-texreg::knitreg(list(m10.1, m11.1, m12.1, m10.2, m11.2, m12.2),
-               custom.model.names = c("Model 12", "Model 13", "Model 14", "Model 15", "Model 16", "Model 17"),
-               caption = NULL,
-               stars = c(0.05, 0.01, 0.001),
-               custom.coef.map = ccoef,
-               digits = 3,
-               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
-               leading.zero = T,
-               use.packages = F,
-               booktabs = F,
-               scalebox = 0.8,
-               include.loglik = FALSE,
-               include.aic = FALSE,
-               center = T,
-               custom.gof.rows = list("Controls"=c(rep("Yes",6))))
-
-```
-
-```{r}
-#| label: fig-growth
-#| out-width: '100%'
-#| fig-asp: 1
-#| fig-cap: "Time varying within effects of meritocratic perceptions on market justice preferences"
-#| fig-cap-location: top
-#| echo: false
-#| results: asis
-
-library(ggeffects)
-
-p4 <- ggpredict(m11.2, terms = c("ola_num [all]", "merit_effort_mean")) %>% 
-  plot() + 
-  scale_x_continuous(n.breaks = 6) +
-  scale_color_manual(
-    name = "Merit:\nEffort",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  scale_fill_manual(
-    name = "Merit:\nEffort",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")) +
-  labs(x = "Wave",
-       title = "(A) People are rewarded for their efforts",
-       color = NULL) +
-  ggdist::theme_ggdist() +
-  theme(legend.position = "top")
-
-p5 <- ggpredict(m12.2, terms = c("ola_num [all]", "merit_talent_mean")) %>% 
-  plot() +
-  scale_x_continuous(n.breaks = 6) +
-  scale_color_manual(
-    name = "Merit:\nTalent",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  scale_fill_manual(
-    name = "Merit:\nTalent",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  labs(
-    x = "Wave",
-    title = "(B) People are rewarded for their intelligence",
-    caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)"
-  ) +
-  ggdist::theme_ggdist() +
-  theme(legend.position = "top")
-
-ggpubr::ggarrange(p4, p5, common.legend=FALSE, ncol=1)
-
-```
-
-With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time only has significant effects on the between effects of the meritocratic variables (Model 16 and Model 17).
-These interaction effects are negative and mean that in the last waves, the impact of meritocratic variables on market justice preferences is attenuated.
-@fig-growth illustrates the interaction between survey wave and the within-person effects of the meritocracy variables on market-justice preferences.
-As time advances, the positive within-person effect of perceiving that effort is rewarded (panel A) or that intelligence is rewarded (panel B) progressively attenuates.
-Put differently, the uplift in market justice preferences associated with a strong meritocratic perception becomes smaller in each successive wave.

### 2025-05-21 05-discussion.qmd
ajustes

@@
 The analysis indicated that the positive association between meritocratic beliefs and support for market-based distribution tended to weaken when perceptions of economic inequality were particularly high.
-This pattern suggested that high awareness of disparities could challenge the moral high ground of meritocracy, as individuals might question whether success was truly based on effort or whether systemic biases distorted outcomes.
-Although the moderation was not consistently statistically significant across all models (i.e.within subjects effects), the observed trend aligned with theoretical expectations that perceptions of inequality undermined the legitimacy of meritocratic narratives and fragmented support for existing economic arrangements.
-This interaction revealed the complex psychological landscape in which deeply held beliefs about effort and reward could be destabilized by heightened awareness of systemic disparities, especially during times of social upheaval or mobilization.
+Although this moderation is not significant for within-subject effects (i.e.
+over time), when taking into account between-subjects differences we found a positive effect.
+This appears as contrary to our initial hypothesis, as it would mean that effect of meritocratic perceptions becomes stronger as perception of inequality increases.
+Nevertheless, we need to be cautious about this interpretation, as the main effect of inequality perception on market justice is negative.
+Therefore, as perceived inequality become less negative (i.e.
+more positive), or closer to zero, then the meritocratic perception has an stronger positive relationship on market justice preferences.
+This could mean that meritocratic perceptions play an stronger role in justification of market justice inequalities for those who are less capable or less prone to perceive inequality.
 

### 2025-05-21 06-conclusion.qmd
ajustes

@@
 
-The findings confirm that higher perceptions of economic inequality are associated with increased support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health.
+The findings confirm that higher perceptions of economic inequality are associated with less support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health.
 Meritocratic beliefs strongly reinforce support for market justice, serving as moral justifications for structural inequalities.
-Additionally, perceptions of inequality moderate the influence of meritocratic beliefs, such that high awareness of disparities can weaken support for existing market-based allocations.
+Additionally, perceptions of inequality moderate the influence of meritocratic beliefs, such that as the awareness of disparities are attenuated (i.e.
+less negative) the role of meritocratic beliefs on existing market-based allocations become stronger.
 Notably, the social mobilizations of 2019 and subsequent periods seem to have contributed to shifting attitudes, fostering greater skepticism of systemic fairness and reducing the legitimizing power of meritocratic narratives.
@@
 Finally, the potential influence of media, political discourse, and peer effects on perceptions was not directly assessed.
-
 Future research could explore the causal mechanisms linking perceptions and attitudes, possibly through experimental or mixed-method designs.

### 2025-05-22 03-methods.qmd
Up correciones N y efectos principales + render

@@
 
-This study draws on data from the Chilean Longitudinal Social Survey (ELSOC), a panel study collected annually from 2016 to 2023.
-The survey evaluates how individuals think, feel, and behave regarding conflict and social cohesion in Chile.
-ELSOC employs a probabilistic, stratified, clustered, multistage sampling design encompassing major urban centers (Santiago, Valparaíso, and Concepción) and smaller cities.
-The target population includes women and men aged 18--75 who are habitual residents of private dwellings.
-The first wave included 2,927 participants, representing northern and southern regions, covering 77% of the country's total population and 93% of its urban population, with a response rate of 62.4% [@elsoc_estudio_2022].
+This study draws on data from the Chilean Longitudinal Social Survey (ELSOC), a panel study conducted annually from 2016 to 2023.
+The survey assesses how individuals think, feel, and behave regarding social conflict and cohesion in Chile.
+ELSOC employs a probabilistic, stratified, clustered, multistage sampling design covering major urban centers (Santiago, Valparaíso, and Concepción) as well as smaller cities.
+The target population includes men and women aged 18 to 75 who are habitual residents of private dwellings.
 
-The survey has been conducted yearly since 2016, except in 2020, when it was suspended due to the pandemic.
-Waves 2016, 2017, 2018, 2019, 2022, and 2023 used computer-assisted personal interviewing (CAPI), while a reduced wave in 2021 employed computer-assisted telephone interviews (CATI).
-Wave 3 included a refreshment sample (N = 1,519) to counter attrition, but these cases are not used here to capture longer response trends.
-Between wave 1 and wave 7, attrition amounted to 40%, achieving a final sample of N = 1,741.
-Longitudinal weights adjust for both the sampling design and potential biases arising from systematic non-participation.
-Further details on sampling, attrition, and weighting can be found at <https://coes.cl/encuesta-panel/>, and the dataset is publicly available at <https://dataverse.harvard.edu/dataverse/elsoc>.
+The survey has been conducted every year since 2016, except in 2020, when it was suspended due to the COVID-19 pandemic.
+The first wave included 2,927 participants from both northern and southern regions, covering 77% of the national population and 93% of the urban population, with a response rate of 62.4% [@elsoc_estudio_2022].
+This study uses six waves: 2016, 2017, 2018, 2019, 2022, and 2023.
+The 2021 wave was excluded because a reduced version of the questionnaire omitted key variables of interest.
+Between waves 1 and 6, panel attrition reached 40%, resulting in a final two-level sample comprising N = 8,643 observations nested within N = 1,687 individuals.
+Longitudinal weights are applied to adjust for both the sampling design and potential biases from systematic non-response.
+Further details on sampling, attrition, and weighting procedures are available at https://coes.cl/encuesta-panel/, and the dataset is publicly accessible at https://dataverse.harvard.edu/dataverse/elsoc.
 

### 2025-05-22 04-analisis.qmd
Up correciones N y efectos principales + render

@@
        title = "c. Education",
-       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)") +
+       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (N obs = 8,643; N groups = 1,687)") +
   theme_ggdist() +
@@
        shape = NULL,
-       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)") +
+       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (N obs = 8,643; N groups = 1,687)") +
   theme_ggdist() +
@@
   "(Intercept)" = "Intercept",
+  perc_inequality_cwc = "Perception inequality (WE)",
+  merit_effort_cwc = "Merit: Effort (WE)",
+  merit_talent_cwc = "Merit: Talent (WE)",
+  perc_inequality_mean = "Perception inequality (BE)",
+  merit_effort_mean = "Merit: Effort (BE)",
+  merit_talent_mean = "Merit: Talent (BE)",
   "perc_inequality_cwc:merit_effort_cwc" = "Merit: Effort (WE) x Perception inequality (WE)",

### 2025-05-22 paper.qmd
Up graficos interaccion, tablas, citas y material suplementario + render
Up correciones N y efectos principales + render

@@
   This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and preferences for market justice in Chile between 2016 and 2023.
-Using six waves of panel data from the Chilean Longitudinal Social Survey (ELSOC; N = 1,519), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions.
+Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions.
 Results indicate that rising perceptions of inequality are associated with greater support for redistributive policies; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities.

### 2025-05-22 paper.tex
Up correciones N y efectos principales + render

@@
 \begin{abstract}
-This study examines the interplay between perceptions of economic
-inequality and meritocracy in shaping preferences for market justice in
-Chile from 2016 to 2023. Economic inequality has intensified globally,
-raising concerns about its social impacts and the legitimacy of
-market-based distributions of resources. By utilizing longitudinal data
-from the Chilean Longitudinal Social Survey (ELSOC), the research
-investigates how subjective assessments of inequality influence
-attitudes towards social justice, focusing on the extent to which
-individuals believe access to essential services like healthcare,
-education, and pensions should depend on personal merit. The findings
-suggest that as perceptions of economic inequality rise, individuals
-tend to favor redistributive policies, although those with strong
-meritocratic beliefs are more likely to justify existing inequalities.
-The analysis also highlights the effect of major social movements in
-Chile during this period, which have shifted public discourse and
-potentially altered perceptions of inequality and meritocracy. This
-paper contributes to the understanding of how individual beliefs about
-fairness and equity evolve within a socio-political context
-characterized by high inequality and extensive free-market principles.
-\newline \textbf{Keywords}: Economic inequality, meritocracy, market
-justice, Chile, public preferences
+This study investigates the relationship between perceptions of economic
+inequality, meritocratic beliefs, and preferences for market justice in
+Chile between 2016 and 2023. Using six waves of panel data from the
+Chilean Longitudinal Social Survey - ELSOC (\(N_{observations}\) =
+8,643; \(N_{individuals}\) = 1,687), the analysis examines how
+subjective assessments of inequality shape attitudes toward the role of
+merit in access to key social services such as healthcare, education,
+and pensions. Results indicate that rising perceptions of inequality are
+associated with greater support for redistributive policies; however,
+individuals with strong meritocratic convictions are more likely to
+legitimize existing disparities. The study also considers the influence
+of major social movements during this period, which appear to have
+reshaped public discourse and perceptions of fairness. These findings
+contribute to a deeper understanding of how beliefs about justice and
+equity evolve in contexts marked by persistent inequality and entrenched
+market-oriented frameworks. \newline \textbf{Keywords}: Economic
+inequality, meritocracy, market justice, Chile, public preferences,
+inequality perception
 \end{abstract}
@@
 as a normative principle that legitimates the distribution of economic
-rewards based on individual merit and effort. It is possible to trace
-the concept to the understanding of Lane
-(\citeproc{ref-lane_market_1986}{1986}), who defines market justice as a
-system of ``earned deserts,'' whereby individuals are seen as deserving
-of a determined distributive outcome due to their effort and skills. In
-contrast, political justice, emphasizes principles of equality and need,
-which are often represented by the welfare state action through social
-policies. An important remark is that the principles of market justice
-rely on the assumption that markets are neutral and self-regulating
-arenas, where individuals are treated fairly because they face the same
-formal rules of engagement (\citeproc{ref-lane_market_1986}{Lane,
-1986}). Consequently, the legitimacy of market justice stems from the
-assumption that inequalities are not only inevitable but fair---so long
-as the rules are transparent and opportunities are open. In this way,
-market justice provides a moral justification for inequality by framing
-it as a necessary outcome of individual responsibility.
+rewards based on individual merit. It is possible to trace the concept
+to the understanding of Lane (\citeproc{ref-lane_market_1986}{1986}),
+who makes a contrast between market justice and political justice. The
+author defines market justice as a system of ``earned deserts,'' whereby
+individuals are seen as deserving of a determined distributive outcome
+due to their effort and skills. In contrast, political justice
+emphasizes principles of equality and need, which are often represented
+by the welfare state action through social policies. An important remark
+is that the principles of market justice rely on the assumption that
+markets are neutral and self-regulating arenas, where individuals are
+treated fairly because they face the same formal rules of engagement
+(\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently, the
+legitimacy of market justice stems from the assumption that inequalities
+are not only inevitable but fair---so long as the rules are transparent
+and opportunities are open. In this way, market justice provides a moral
+justification for inequality by framing it as a necessary outcome of
+individual responsibility.
 
@@
 across different contexts. Similarly, comparative studies have also
-relied on these indicators to study the support for market-based
+considered these indicators to study the support for market-based
 distribution in welfare systems (\citeproc{ref-lindh_public_2015}{Lindh,
@@
 that market justice is also shaped contextually, which is represented by
-the development of welfare institutions.
+the development of welfare institutions, in line with the
+policy-feedbacks literature {[}campbell\_policy\_2012;
+pierson\_when\_1993{]}.
 
@@
 particular, we explore whether the political outburst of 2019 and the
-subsequent constitutional processes in Chile have altered how
-perceptions of inequality and meritocracy shape market justice
-attitudes. We expect that after these critical events, perceptions of
-inequality may have a stronger negative effect on market justice
-preferences, reflecting increased societal questioning of market-based
-allocation mechanisms.
+subsequent constitutional processes in Chile --- where socio-economic
+inequality was widely challenged --- have altered how perceptions of
+inequality and meritocracy shape market justice attitudes. We expect
+that after these critical events, perceptions of inequality may have a
+stronger negative effect on market justice preferences, reflecting
+increased societal questioning of market-based allocation mechanisms.
 
@@
 
-\(H1\): Higher perceived economic inequality is associated with higher
+\(H1\): Higher perceived economic inequality is associated with less
 market justice preferences.
@@
 This study draws on data from the Chilean Longitudinal Social Survey
-(ELSOC), a panel study collected annually from 2016 to 2023. The survey
-evaluates how individuals think, feel, and behave regarding conflict and
-social cohesion in Chile. ELSOC employs a probabilistic, stratified,
-clustered, multistage sampling design encompassing major urban centers
-(Santiago, Valparaíso, and Concepción) and smaller cities. The target
-population includes women and men aged 18--75 who are habitual residents
-of private dwellings. The first wave included 2,927 participants,
-representing northern and southern regions, covering 77\% of the
-country's total population and 93\% of its urban population, with a
-response rate of 62.4\% (\citeproc{ref-elsoc_estudio_2022}{ELSOC,
-2022}).
-
-The survey has been conducted yearly since 2016, except in 2020, when it
-was suspended due to the pandemic. Waves 2016, 2017, 2018, 2019, 2022,
-and 2023 used computer-assisted personal interviewing (CAPI), while a
-reduced wave in 2021 employed computer-assisted telephone interviews
-(CATI). Wave 3 included a refreshment sample (N = 1,519) to counter
-attrition, but these cases are not used here to capture longer response
-trends. Between wave 1 and wave 7, attrition amounted to 40\%, achieving
-a final sample of N = 1,741. Longitudinal weights adjust for both the
-sampling design and potential biases arising from systematic
-non-participation. Further details on sampling, attrition, and weighting
-can be found at \url{https://coes.cl/encuesta-panel/}, and the dataset
-is publicly available at
-\url{https://dataverse.harvard.edu/dataverse/elsoc}.
+(ELSOC), a panel study conducted annually from 2016 to 2023. The survey
+assesses how individuals think, feel, and behave regarding social
+conflict and cohesion in Chile. ELSOC employs a probabilistic,
+stratified, clustered, multistage sampling design covering major urban
+centers (Santiago, Valparaíso, and Concepción) as well as smaller
+cities. The target population includes men and women aged 18 to 75 who
+are habitual residents of private dwellings.
+
+The survey has been conducted every year since 2016, except in 2020,
+when it was suspended due to the COVID-19 pandemic. The first wave
+included 2,927 participants from both northern and southern regions,
+covering 77\% of the national population and 93\% of the urban
+population, with a response rate of 62.4\%
+(\citeproc{ref-elsoc_estudio_2022}{ELSOC, 2022}). This study uses six
+waves: 2016, 2017, 2018, 2019, 2022, and 2023. The 2021 wave was
+excluded because a reduced version of the questionnaire omitted key
+variables of interest. Between waves 1 and 6, panel attrition reached
+40\%, resulting in a final two-level sample comprising N = 8,643
+observations nested within N = 1,687 individuals. Longitudinal weights
+are applied to adjust for both the sampling design and potential biases
+from systematic non-response. Further details on sampling, attrition,
+and weighting procedures are available at
+https://coes.cl/encuesta-panel/, and the dataset is publicly accessible
+at https://dataverse.harvard.edu/dataverse/elsoc.
 
@@
 individuals justify conditioning access to social services---healthcare,
-pensions, and education---on income. Specifically, the justification of
-inequality in healthcare is assessed by the question: ``Is it fair in
-Chile that people with higher incomes can access better healthcare than
-people with lower incomes?'' The same question is posed for pensions and
-education. In all cases, respondents indicate their level of agreement
-on a five-point Likert scale ranging from 1 (``strongly disagree'') to 5
-(``strongly agree''). Additionally, we include a composite measure of
-``market justice preferences'', calculated as the average of these three
-items (\(\alpha\) = 0.84). This index ranges from 1 to 5, with higher
-values indicating stronger preferences for market justice (see
-Table~\ref{tbl-summary1}).
+pensions, and education--- basen on individual income. Specifically, the
+justification of inequality in healthcare is assessed by the question:
+``Is it fair in Chile that people with higher incomes can access better
+healthcare than people with lower incomes?'' The same question is posed
+for pensions and education. In all cases, respondents indicate their
+level of agreement on a five-point Likert scale ranging from 1
+(``strongly disagree'') to 5 (``strongly agree''). Additionally, we
+include a composite measure of ``market justice preferences'',
+calculated as the average of these three items (\(\alpha\) = 0.84). This
+index ranges from 1 to 5, with higher values indicating stronger
+preferences for market justice (see Table~\ref{tbl-summary1}).
 
@@
 variable refers to the perception of economic inequality, measured
-through the perceived wage gap.
+through the perceived wage gap
 (\citeproc{ref-castillo_cual_2009}{Castillo, 2009};
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1038}}@{}}
-\caption{Independent variables for the first wave
-(2016)}\label{tbl-summary2}\tabularnewline
+\caption{Independent variables ELSOC survey (descriptives for first wave
+2016)}\label{tbl-summary2}\tabularnewline
 \toprule\noalign{}
@@
 2=Center, 3=Right, 4=No identification) and subjective social status
-(ranging from 1 to 10) as they may confound the relationship between
-market justice preferences and perceptions of inequality and meritocracy
+(ranging from 1 to 10) as they may affect the relationship between
+market justice preferences, perceptions of inequality, and meritocracy
 (\citeproc{ref-schneider_poverty_2015}{Schneider \& Castillo, 2015}).
@@
 a slight decrease in disagreement and a corresponding rise in support
-for inequality. Specifically, in healthcare and education, although
-disagreement remains substantial, agreement (agree + strongly agree)
-increased from 7.4\% and 7.2\% in 2019 to 13.1\% and 14.2\% in 2023,
-respectively. This shift is most evident in pensions, where the combined
-agree/strongly agree category grew by about 10 percentage points, from
-16.9\% in 2016 to 28\% in 2023.
+for market-justice inequality. Specifically, in healthcare and
+education, although disagreement remains substantial, agreement (agree +
+strongly agree) increased from 7.4\% and 7.2\% in 2019 to 13.1\% and
+14.2\% in 2023, respectively. This shift is most evident in pensions,
+where the combined agree/strongly agree category grew by about 10
+percentage points, from 16.9\% in 2016 to 28\% in 2023.
 
@@
 
-Regarding the main variables of this study, Figure~\ref{fig-meanchange}
-depicts their average changes over the years. We observe an increase in
-the average level of market justice preferences in the most recent
-waves, which begins in 2019. The highest average consistently appears
-for perceived economic inequality, although this variable shows a
-downward trend of roughly one point over time. Interestingly, while
-perceptions of economic inequality declined in the latest measurements
-(2022--2023), market justice preferences increased. The meritocracy
-measures remain stable, though the perception that individuals are
-rewarded for intelligence is slightly higher than the perception that
-they are rewarded for their effort.
+Regarding the main dependent and independent variables of this study,
+Figure~\ref{fig-meanchange} depicts their average changes over the
+years. We observe an increase in the average level of market justice
+preferences in the most recent waves, which begins in 2019. The highest
+average consistently appears for perceived economic inequality, although
+this variable shows a downward trend of roughly one point over time.
+Interestingly, while perceptions of economic inequality declined in the
+latest measurements (2022--2023), market justice preferences increased.
+The meritocracy measures remain stable, though the perception that
+individuals are rewarded for intelligence is slightly higher than the
+perception that they are rewarded for their effort.
 
@@
 
-\begin{center}
-\scalebox{0.79}{
-\begin{tabular}{l c c c c c c c}
-\hline
- & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 \\
-\hline
-Intercept                            & $1.968^{***}$  & $2.015^{***}$  & $2.030^{***}$  & $2.028^{***}$  & $2.169^{***}$  & $1.456^{***}$  & $1.547^{***}$  \\
-                                     & $(0.021)$      & $(0.035)$      & $(0.035)$      & $(0.035)$      & $(0.070)$      & $(0.097)$      & $(0.115)$      \\
-Wave (Ref.= 2016)                    &                &                &                &                &                &                &                \\
-                                     &                &                &                &                &                &                &                \\
-\quad Wave 2017                      & $-0.167^{***}$ &                &                &                &                &                &                \\
-                                     & $(0.027)$      &                &                &                &                &                &                \\
-\quad Wave 2018                      & $-0.025$       &                &                &                &                &                &                \\
-                                     & $(0.027)$      &                &                &                &                &                &                \\
-\quad Wave 2019                      & $-0.046$       &                &                &                &                &                &                \\
-                                     & $(0.027)$      &                &                &                &                &                &                \\
-\quad Wave 2022                      & $0.285^{***}$  &                &                &                &                &                &                \\
-                                     & $(0.028)$      &                &                &                &                &                &                \\
-\quad Wave 2023                      & $0.284^{***}$  &                &                &                &                &                &                \\
-                                     & $(0.027)$      &                &                &                &                &                &                \\
-Wave                                 &                & $-0.121^{***}$ & $-0.125^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.129^{***}$ \\
-                                     &                & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      \\
-Wave$^2$                             &                & $0.028^{***}$  & $0.028^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  \\
-                                     &                & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      \\
-Perception inequality (WE)           &                &                & $-0.039^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ \\
-                                     &                &                & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
-Merit: Effort (WE)                   &                &                &                & $0.070^{***}$  & $0.070^{***}$  & $0.069^{***}$  & $0.069^{***}$  \\
-                                     &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
-Merit: Talent (WE)                   &                &                &                & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    \\
-                                     &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
-Perception inequality (BE)           &                &                &                &                & $-0.042^{*}$   & $-0.006$       & $-0.033$       \\
-                                     &                &                &                &                & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
-Merit: Effort (BE)                   &                &                &                &                &                & $0.186^{***}$  & $0.178^{***}$  \\
-                                     &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
-Merit: Talent (BE)                   &                &                &                &                &                & $0.040$        & $0.019$        \\
-                                     &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
-\hline
-Controls                             & No             & No             & No             & No             & No             & No             & Yes            \\
-BIC                                  & $20780.594$    & $20780.617$    & $20781.113$    & $20726.753$    & $20736.627$    & $20655.172$    & $20749.047$    \\
-Num. obs.                            & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         \\
-Num. groups: idencuesta              & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         \\
-Var: idencuesta (Intercept)          & $0.172$        & $0.223$        & $0.219$        & $0.209$        & $0.207$        & $0.182$        & $0.171$        \\
-Var: Residual                        & $0.530$        & $0.499$        & $0.498$        & $0.495$        & $0.495$        & $0.495$        & $0.495$        \\
-Var: idencuesta ola\_num             & $$             & $0.011$        & $0.010$        & $0.009$        & $0.009$        & $0.009$        & $0.009$        \\
-Cov: idencuesta (Intercept) ola\_num & $$             & $-0.025$       & $-0.024$       & $-0.021$       & $-0.020$       & $-0.019$       & $-0.019$       \\
-\hline
-\multicolumn{8}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
-\end{tabular}
-}
-\caption{}
-\label{table:coefficients}
-\end{center}
+~
+
+Model 1
+
+Model 2
+
+Model 3
+
+Model 4
+
+Model 5
+
+Model 6
+
+Model 7
+
+Intercept
+
+1.968***
+
+2.015***
+
+2.030***
+
+2.028***
+
+2.169***
+
+1.456***
+
+1.547***
+
+~
+
+(0.021)
+
+(0.035)
+
+(0.035)
+
+(0.035)
+
+(0.070)
+
+(0.097)
+
+(0.115)
+
+Wave (Ref.= 2016)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2017
+
+-0.167***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.027)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2018
+
+-0.025
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.027)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2019
+
+-0.046
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.027)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2022
+
+0.285***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.028)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~~~~~Wave 2023
+
+0.284***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.027)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+Wave
+
+~
+
+-0.121***
+
+-0.125***
+
+-0.128***
+
+-0.128***
+
+-0.128***
+
+-0.129***
+
+~
+
+~
+
+(0.022)
+
+(0.022)
+
+(0.022)
+
+(0.022)
+
+(0.022)
+
+(0.022)
+
+Wave\^{}2
+
+~
+
+0.028***
+
+0.028***
+
+0.029***
+
+0.029***
+
+0.029***
+
+0.029***
+
+~
+
+~
+
+(0.003)
+
+(0.003)
+
+(0.003)
+
+(0.003)
+
+(0.003)
+
+(0.003)
+
+Perception inequality (WE)
+
+~
+
+~
+
+-0.039***
+
+-0.034***
+
+-0.034***
+
+-0.034***
+
+-0.034***
+
+~
+
+~
+
+~
+
+(0.010)
+
+(0.010)
+
+(0.010)
+
+(0.010)
+
+(0.010)
+
+Merit: Effort (WE)
+
+~
+
+~
+
+~
+
+0.070***
+
+0.070***
+
+0.069***
+
+0.069***
+
+~
+
+~
+
+~
+
+~
+
+(0.012)
+
+(0.012)
+
+(0.012)
+
+(0.012)
+
+Merit: Talent (WE)
+
+~
+
+~
+
+~
+
+0.027*
+
+0.027*
+
+0.027*
+
+0.027*
+
+~
+
+~
+
+~
+
+~
+
+(0.012)
+
+(0.012)
+
+(0.012)
+
+(0.012)
+
+Perception inequality (BE)
+
+~
+
+~
+
+~
+
+~
+
+-0.042*
+
+-0.006
+
+-0.033
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.018)
+
+(0.018)
+
+(0.018)
+
+Merit: Effort (BE)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+0.186***
+
+0.178***
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.032)
+
+(0.031)
+
+Merit: Talent (BE)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+0.040
+
+0.019
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.032)
+
+(0.031)
+
+Controls
+
+No
+
+No
+
+No
+
+No
+
+No
+
+No
+
+Yes
+
+BIC
+
+20780.594
+
+20780.617
+
+20781.113
+
+20726.753
+
+20736.627
+
+20655.172
+
+20749.047
+
+Numb. obs.
+
+8643
+
+8643
+
+8643
+
+8643
+
+8643
+
+8643
+
+8643
+
+Num. groups: individuals
+
+1687
+
+1687
+
+1687
+
+1687
+
+1687
+
+1687
+
+1687
+
+Var: individuals (Intercept)
+
+0.172
+
+0.223
+
+0.219
+
+0.209
+
+0.207
+
+0.182
+
+0.171
+
+Var: Residual
+
+0.530
+
+0.499
+
+0.498
+
+0.495
+
+0.495
+
+0.495
+
+0.495
+
+Var: individuals, wave
+
+~
+
+0.011
+
+0.010
+
+0.009
+
+0.009
+
+0.009
+
+0.009
+
+Cov: individuals (Intercept), wave
+
+~
+
+-0.025
+
+-0.024
+
+-0.021
+
+-0.020
+
+-0.019
+
+-0.019
+
+Note: Cells contain regression coefficients with standard errors in
+parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
+\textless{} 0.05.
 
@@
 
-\begin{center}
-\scalebox{0.8}{
-\begin{tabular}{l c c}
-\hline
- & Model 8 & Model 9 \\
-\hline
-Intercept                                                & $1.558^{***}$ & $2.361^{***}$ \\
-                                                         & $(0.115)$     & $(0.287)$     \\
-Merit: Effort (WE) x Perception inequality (WE)          & $0.019$       &               \\
-                                                         & $(0.016)$     &               \\
-Merit: Talent (WE) x Perception inequality (WE)          & $-0.002$      &               \\
-                                                         & $(0.016)$     &               \\
-Merit: Effort (BE) x Perception inequality (BE)          &               & $0.016$       \\
-                                                         &               & $(0.045)$     \\
-Merit: Talent (BE) x Perception inequality (BE)          &               & $0.070$       \\
-                                                         &               & $(0.045)$     \\
-\hline
-Controls                                                 & Yes           & Yes           \\
-BIC                                                      & $20879.829$   & $20819.684$   \\
-Num. obs.                                                & $8643$        & $8643$        \\
-Num. groups: idencuesta                                  & $1687$        & $1687$        \\
-Var: idencuesta (Intercept)                              & $0.147$       & $0.141$       \\
-Var: idencuesta merit\_effort\_cwc                       & $0.016$       & $$            \\
-Var: idencuesta merit\_talent\_cwc                       & $0.018$       & $$            \\
-Var: idencuesta perc\_inequality\_cwc                    & $0.000$       & $$            \\
-Cov: idencuesta (Intercept) merit\_effort\_cwc           & $-0.005$      & $$            \\
-Cov: idencuesta (Intercept) merit\_talent\_cwc           & $0.015$       & $$            \\
-Cov: idencuesta (Intercept) perc\_inequality\_cwc        & $-0.006$      & $$            \\
-Cov: idencuesta merit\_effort\_cwc merit\_talent\_cwc    & $-0.003$      & $$            \\
-Cov: idencuesta merit\_effort\_cwc perc\_inequality\_cwc & $0.000$       & $$            \\
-Cov: idencuesta merit\_talent\_cwc perc\_inequality\_cwc & $0.001$       & $$            \\
-Var: Residual                                            & $0.503$       & $0.528$       \\
-\hline
-\multicolumn{3}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
-\end{tabular}
+~
+
+Model 8
+
+Model 9
+
+Model 10
+
+Model 11
+
+Intercept
+
+1.561***
+
+1.555***
+
+2.187***
+
+2.340***
+
+~
+
+(0.115)
+
+(0.115)
+
+(0.265)
+
+(0.281)
+
+Perception inequality (WE)
+
+-0.035***
+
+-0.035***
+
+-0.035***
+
+-0.035***
+
+~
+
+(0.010)
+
+(0.010)
+
+(0.010)
+
+(0.010)
+
+Merit: Effort (WE)
+
+0.072***
+
+0.074***
+
+0.073***
+
+0.073***
+
+~
+
+(0.013)
+
+(0.012)
+
+(0.012)
+
+(0.012)
+
+Merit: Talent (WE)
+
+0.031*
+
+0.031*
+
+0.030*
+
+0.030*
+
+~
+
+(0.012)
+
+(0.013)
+
+(0.012)
+
+(0.012)
+
+Perception inequality (BE)
+
+-0.032
+
+-0.033
+
+-0.213**
+
+-0.255***
+
+~
+
+(0.018)
+
+(0.018)
+
+(0.073)
+
+(0.077)
+
+Merit: Effort (BE)
+
+0.177***
+
+0.173***
+
+-0.062
+
+0.179***
+
+~
+
+(0.032)
+
+(0.031)
+
+(0.097)
+
+(0.032)
+
+Merit: Talent (BE)
+
+0.015
+
+0.021
+
+0.014
+
+-0.266**
+
+~
+
+(0.031)
+
+(0.031)
+
+(0.031)
+
+(0.099)
+
+Merit: Effort (WE) x Perception inequality (WE)
+
+0.017
+
+~
+
+~
+
+~
+
+~
+
+(0.012)
+
+~
+
+~
+
+~
+
+Merit: Talent (WE) x Perception inequality (WE)
+
+~
+
+0.011
+
+~
+
+~
+
+~
+
+~
+
+(0.012)
+
+~
+
+~
+
+Merit: Effort (BE) x Perception inequality (BE)
+
+~
+
+~
+
+0.072**
+
+~
+
+~
+
+~
+
+~
+
+(0.028)
+
+~
+
+Merit: Talent (BE) x Perception inequality (BE)
+
+~
+
+~
+
+~
+
+0.082**
+
+~
+
+~
+
+~
+
+~
+
+(0.027)
+
+Controls
+
+Yes
+
+Yes
+
+Yes
+
+Yes
+
+BIC
+
+20842.694
+
+20835.920
+
+20808.677
+
+20806.384
+
+Numb. obs.
+
+8643
+
+8643
+
+8643
+
+8643
+
+Num. groups: individuals
+
+1687
+
+1687
+
+1687
+
+1687
+
+Var: individuals (Intercept)
+
+0.145
+
+0.145
+
+0.141
+
+0.141
+
+Var: individuals, merit effort cwc
+
+0.020
+
+~
+
+~
+
+~
+
+Var: individuals, perception inequality cwc
+
+0.000
+
+0.000
+
+~
+
+~
+
+Cov: individuals (Intercept), merit effort cwc
+
+0.004
+
+~
+
+~
+
+~
+
+Cov: individuals (Intercept), perception inequality cwc
+
+-0.006
+
+-0.005
+
+~
+
+~
+
+Cov: individuals, merit effort cwc, perception inequality cwc
+
+0.000
+
+~
+
+~
+
+~
+
+Var: Residuals
+
+0.512
+
+0.511
+
+0.528
+
+0.528
+
+Var: individuals, merit talent cwc
+
+~
+
+0.020
+
+~
+
+~
+
+Cov: individuals (Intercept), merit talent cwc
+
+~
+
+0.012
+
+~
+
+~
+
+Cov: individuals, merit talent cwc, perception inequality cwc
+
+~
+
+0.001
+
+~
+
+~
+
+Note: Cells contain regression coefficients with standard errors in
+parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
+\textless{} 0.05. CWC = centered within group.
+
 }
-\caption{}
-\label{table:coefficients}
-\end{center}
+
+\end{table}%
+
+\begin{figure}[H]
+
+\caption{\label{fig-interact}Between effects of meritocratic perceptions
+on market justice preferences by economic inequality perception}
+
+\centering{
+
+\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-interact-1.pdf}
 
@@
 
+\end{figure}%
+
+Table~\ref{tbl-interactions1} examines whether perceived economic
+inequality moderates the effect of meritocratic beliefs on market
+justice preferences. Contrary to our hypothesis, the interaction terms
+in the within-person specification (Models 8 and 9) do not reach
+statistical significance. However, in the between-person specification
+(Models 10 and 11), the interaction terms are statistically significant,
+suggesting that perceptions of economic inequality meaningfully alter
+the effect of meritocracy on support for market-based allocation of
+social services. These positive effects indicate that, as perceptions of
+inequality increase, the positive association between meritocratic
+perceptions and market justice preferences becomes stronger. The
+marginal effects of this interaction are illustrated in
+Figure~\ref{fig-interact}, which shows that the effect of meritocratic
+perceptions on market justice preferences intensifies as perceived
+economic inequality rises.
+
+\begin{table}
+
+\caption{\label{tbl-interactions2}Interactions for meritocracy,
+perceived economic inequality and market justice preferences}
+
+\centering{
+
+~
+
+Model 12
+
+Model 13
+
+Model 14
+
+Model 15
+
+Model 16
+
+Model 17
+
+Intercept
+
+1.543***
+
+1.527***
+
+1.521***
+
+1.680***
+
+1.374***
+
+1.362***
+
+~
+
+(0.115)
+
+(0.114)
+
+(0.114)
+
+(0.139)
+
+(0.133)
+
+(0.134)
+
+Wave x Perception inequality (WE)
+
+-0.004
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.006)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+Wave x Merit: effort (WE)
+
+~
+
+-0.008
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.006)
+
+~
+
+~
+
+~
+
+~
+
+Wave x Merit: talent (WE)
+
+~
+
+~
+
+-0.006
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.006)
+
+~
+
+~
+
+~
+
+Wave x Perception inequality (BE)
+
+~
+
+~
+
+~
+
+0.011
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.007)
+
+~
+
+~
+
+Wave x Merit: effort (BE)
+
+~
+
+~
+
+~
+
+~
+
+-0.021*
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.008)
+
+~
+
+Wave x Merit: talent (BE)
+
+~
+
+~
+
+~
+
+~
+
+~
+
+-0.020*
+
+~
+
+~
+
+~
+
+~
+
+~
+
+~
+
+(0.008)
+
+Controls
+
+Yes
+
+Yes
+
+Yes
+
+Yes
+
+Yes
+
+Yes
+
+BIC
+
+20792.207
+
+20767.371
+
+20754.324
+
+20782.059
+
+20749.596
+
+20754.417
+
+Numb. obs.
+
+8643
+
+8643
+
+8643
+
+8643
+
+8643
+
+8643
+
+Num. groups: individuals
+
+1687
+
+1687
+
+1687
+
+1687
+
+1687
+
+1687
+
+Var: individuals (Intercept)
+
+0.176
+
+0.167
+
+0.168
+
+0.383
+
+0.254
+
+0.010
+
+Var: individuals, perception inequality cwc
+
+0.000
+
+~
+
+~
+
+~
+
+~
+
+~
+
+Var: individuals, wave
+
+0.010
+
+0.009
+
+0.009
+
+0.009
+
+0.009
+
+0.009
+
+Cov: individuals (Intercept), perception inequality cwc
+
+-0.007
+
+~
+
+~
+
+~
+
+~
+
+~
+
+Cov: individuals (Intercept), wave
+
+-0.021
+
+-0.018
+
+-0.018
+
+-0.048
+
+-0.003
+
+0.006
+
+Cov: individuals, perception inequality cwc, wave
+
+0.002
+
+~
+
+~
+
+~
+
+~
+
+~
+
+Var: Residuals
+
+0.495
+
+0.480
+
+0.479
+
+0.494
+
+0.494
+
+0.494
+
+Var: individuals, merit effort cwc
+
+~
+
+0.019
+
+~
+
+~
+
+~
+
+~
+
+Cov: individuals (Intercept), merit effort cwc
+
+~
+
+0.023
+
+~
+
+~
+
+~
+
+~
+
+Cov: individuals, merit effort cwc, wave
+
+~
+
+-0.005
+
+~
+
+~
+
+~
+
+~
+
+Var: individuals, merit talent cwc
+
+~
+
+~
+
+0.019
+
+~
+
+~
+
+~
+
+Cov: individuals (Intercept), merit talent cwc
+
+~
+
+~
+
+0.034
+
+~
+
+~
+
+~
+
+Cov: individuals, merit talent cwc, wave
+
+~
+
+~
+
+-0.006
+
+~
+
+~
+
+~
+
+Var: individuals, perception inequality mean
+
+~
+
+~
+
+~
+
+0.008
+
+~
+
+~
+
+Cov: individuals (Intercept), perception inequality mean
+
+~
+
+~
+
+~
+
+-0.045
+
+~
+
+~
+
+Cov: individuals, perception inequality mean, wave
+
+~
+
+~
+
+~
+
+0.009
+
+~
+
+~
+
+Var: individuals, merit effort mean
+
+~
+
+~
+
+~
+
+~
+
+0.053
+
+~
+
+Cov: individuals (Intercept), merit effort mean
+
+~
+
+~
+
+~
+
+~
+
+-0.089
+
+~
+
+Cov: individuals, merit effort mean, wave
+
+~
+
+~
+
+~
+
+~
+
+-0.006
+
+~
+
+Var: individuals, merit talent mean
+
+~
+
+~
+
+~
+
+~
+
+~
+
+0.025
+
+Cov: individuals (Intercept), merit talent mean
+
+~
+
+~
+
+~
+
+~
+
+~
+
+-0.007
+
+Cov: individuals, merit talent mean, wave
+
+~
+
+~
+
+~
+
+~
+
+~
+
+-0.009
+
+Note: Cells contain regression coefficients with standard errors in
+parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
+\textless{} 0.05. CWC = centered within group.
+
+}
+
 \end{table}%
 
-Table~\ref{tbl-interactions1} reports whether perceived economic
-inequality moderates the impact of meritocratic perceptions on
-market-justice preferences. In neither the within-person specification
-(Model 8) nor the between-person specification (Model 9) do the
-interaction terms reach statistical significance, indicating that higher
-perceptions of economic inequality do not meaningfully alter the effect
-of meritocracy on support for market allocation of social services.
-
-\begin{figure}[H]
-
-\caption{\label{fig-growth}Time varying within effects of meritocratic
-perceptions on market justice preferences}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-growth-1.pdf}
-
-}
-
-\end{figure}%
-
 With respect to the temporal dynamics of the key predictors,
-Figure~\ref{fig-growth} illustrates the interaction between survey wave
-and the within-person effects of the meritocracy variables on
-market-justice preferences (complete models appear in the Supplementary
-Material). As time advances, the positive within-person effect of
-perceiving that effort is rewarded (panel A) or that intelligence is
-rewarded (panel B) progressively attenuates. Put differently, the uplift
-in market justice preferences associated with a strong meritocratic
-perception becomes smaller in each successive wave.
+Table~\ref{tbl-interactions2} shows that time only has significant
+effects on the between effects of the meritocratic variables. These
+interaction effects are negative and mean that in the last waves, the
+impact of meritocratic variables on market justice preferences is
+attenuated. Specifically, as time advances, the positive within-person
+effect of perceiving that effort is rewarded (Model 16) or that
+intelligence is rewarded (Model 17) progressively attenuates. Put
+differently, the uplift in market justice preferences associated with a
+strong meritocratic perception becomes smaller in each successive wave.
 
@@
 
-\section{Conclusion}\label{conclusion}
+This study tests several key hypotheses concerning the relationships
+between perceptions of economic inequality, meritocratic beliefs, and
+attitudes toward market justice in Chile. It posits that higher
+perceived inequality and stronger meritocratic beliefs both
+independently contribute to greater support for market-based
+distributions of social goods. Furthermore, the study examines how
+perceived economic inequality moderates the influence of meritocratic
+beliefs on market justice preferences. Additionally, it considers how
+major social and political mobilizations, such as those occurring around
+2019 and subsequent years, may have weakened these relationships,
+reflecting increased critical perspectives on systemic inequalities and
+market mechanisms. The subsequent discussion interprets the empirical
+findings in light of these hypotheses, considering their implications
+for understanding attitudes toward inequality and social justice in a
+context marked by profound social change.
+
+The first hypothesis suggested that perceptions of economic inequality
+were strongly linked to attitudes toward market justice. The findings
+demonstrated that individuals perceiving higher income disparities
+showed stronger support for the distribution of social services based on
+individual income, aligning with theories emphasizing that awareness of
+inequality could fuel a critical view of such policies
+(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
+\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). This relationship
+reflected a moral economy where perceptions of systemic unfairness
+threatened the legitimacy of existing distributions, prompting calls for
+greater equity. Interestingly, the negative within-effects observed over
+time implied that as perceptions of inequality became more salient, some
+individuals experienced a decline in support for market justice,
+possibly due to increasing distrust in market mechanisms or
+disillusionment with the fairness of the system. This indicated that
+while perception heightened awareness, prolonged exposure or social
+mobilization could also foster skepticism about whether redistribution
+effectively addressed inequality or simply perpetuated systemic flaws.
+
+Regarding the second hypothesis, the results reinforced that
+meritocratic perceptions ---particularly those emphasizing individual
+effort and talent--- were associated with stronger support for
+market-based distribution systems. Individuals who viewed success as
+primarily earned through effort were more inclined to justify existing
+inequalities, seeing them as outcomes of individual merit rather than
+systemic injustice. This aligned with research indicating that
+meritocratic narratives served as moral justifications that reinforced
+societal stratification (\citeproc{ref-Castillo2012a_justice}{Castillo,
+2012}; \citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These
+beliefs functioned symbolically to legitimize structural disparities,
+reducing support for redistributive measures by framing inequality as
+fair and deserved. The findings suggested that these perceptions
+bolstered the social hierarchy by fostering acceptance of inequality as
+a reflection of individual virtue rather than structural failure, which
+was especially pertinent in neoliberal contexts like Chile's, where
+market logic heavily influenced social attitudes
+(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).
+
+The third hypothesis concerned the moderating role of perceptions of
+inequality on the relationship between meritocracy and market justice
+attitudes. The analysis indicated that the positive association between
+meritocratic beliefs and support for market-based distribution tended to
+weaken when perceptions of economic inequality were particularly high.
+Although this moderation is not significant for within-subject effects
+(i.e.~over time), when taking into account between-subjects differences
+we found a positive effect. This appears as contrary to our initial
+hypothesis, as it would mean that effect of meritocratic perceptions
+becomes stronger as perception of inequality increases. Nevertheless, we
+need to be cautious about this interpretation, as the main effect of
+inequality perception on market justice is negative. Therefore, as
+perceived inequality become less negative (i.e.~more positive), or
+closer to zero, then the meritocratic perception has an stronger
+positive relationship on market justice preferences. This could mean
+that meritocratic perceptions play an stronger role in justification of
+market justice inequalities for those who are less capable or less prone
+to perceive inequality.
+
+Finally, the fourth hypothesis addressed the impact of recent social
+movements, particularly from 2019 onwards, on the relationship between
+meritocratic beliefs, perceptions of inequality, and attitudes toward
+market justice. The results suggested that these movements contributed
+to a more critical stance toward the fairness of existing distributional
+mechanisms. Post-mobilization, the positive relationship between
+meritocratic beliefs and support for inequality appeared attenuated,
+indicating that collective action and public discourse during this
+period fostered questioning of the legitimacy of systemic inequalities
+justified solely through meritocratic narratives. These social
+mobilizations challenged dominant narratives, emphasizing structural
+barriers and systemic injustices, which in turn weakened the association
+between individual effort and deservingness. This is aligned with
+literature emphasizing that social protests act as catalysts for
+transforming societal attitudes by disrupting normative beliefs and
+encouraging more critical evaluations of institutional legitimacy
+(\citeproc{ref-tilly_contentious_2015}{Tilly \& Tarrow, 2015}). The
+findings underscored the importance of socio-political context in
+shaping and reshaping attitudes toward inequality and justice,
+highlighting how collective action could temporarily or permanently
+alter the normative foundations that sustained support for market
+justice.
+
+\section{Conclusions}\label{conclusions}
+
+This study examined the complex interplay between perceptions of
+economic inequality, meritocratic beliefs, and attitudes toward market
+justice in Chile from 2016 to 2023, drawing on longitudinal data from
+the ELSOC survey. By exploring how subjective assessments and social
+contexts influence support for redistribution and market-based resource
+allocation, the research offers different elements that contribute to
+the understanding of the normative foundations underpinning social
+justice attitudes in a highly unequal, neoliberal environment.
+
+The findings confirm that higher perceptions of economic inequality are
+associated with less support for market justice attitudes, this is, the
+belief that it is fine that those with higher income have better social
+services such as education, pensions and health. Meritocratic beliefs
+strongly reinforce support for market justice, serving as moral
+justifications for structural inequalities. Additionally, perceptions of
+inequality moderate the influence of meritocratic beliefs, such that as
+the awareness of disparities are attenuated (i.e.~less negative) the
+role of meritocratic beliefs on existing market-based allocations become
+stronger. Notably, the social mobilizations of 2019 and subsequent
+periods seem to have contributed to shifting attitudes, fostering
+greater skepticism of systemic fairness and reducing the legitimizing
+power of meritocratic narratives.
+
+This research advances the extant literature by integrating subjective
+perceptions with social and political contexts to explain attitudes
+toward economic inequality and distributional justice. While previous
+studies primarily focused on objective measures or individual
+characteristics, this work emphasizes the dynamic and interactional
+nature of perceptions and beliefs over time. It highlights the
+importance of socio-political upheavals in reshaping normative
+attitudes, underscoring the role of collective action in challenging
+entrenched narratives of meritocracy and fairness. The longitudinal
+approach provides a deeper temporal perspective on how societal events
+could influence individual beliefs and preferences.
+
+Despite its contributions, the study has several limitations. First, the
+reliance on self-reported perceptions and attitudes may be subject to
+social desirability bias. Second, the focus on Chile, while offering
+valuable insights into a specific context, limits the generalizability
+of results to other countries with different institutional and cultural
+backgrounds. Third, the observational nature of the data prevents causal
+inferences, and unobserved confounding factors may influence the
+observed relationships. Finally, the potential influence of media,
+political discourse, and peer effects on perceptions was not directly
+assessed. Future research could explore the causal mechanisms linking
+perceptions and attitudes, possibly through experimental or mixed-method
+designs. Comparative studies across different national contexts would
+help determine the universality or specificity of these dynamics.
+Additionally, investigating the role of media, political communication,
+and education in shaping perceptions of inequality and meritocracy would
+deepen understanding of the normative foundations of social justice
+attitudes. Finally, examining how these perceptions influence behavioral
+outcomes, such as political participation or support for social
+movements, would provide valuable insights into the pathways from
+beliefs to collective action and policy change.
 
@@
 
+\bibitem[\citeproctext]{ref-canalesceron_sujeto_2021}
+Canales Cerón, M., Orellana Calderón, V. S., \& Guajardo Mañán, F.
+(2021). Sujeto y cotidiano en la era neoliberal: El caso de la
+educaci{ó}n chilena. \emph{Revista Mexicana de Ciencias Pol{í}ticas y
+Sociales}, \emph{67}(244).
+\url{https://doi.org/10.22201/fcpys.2448492xe.2022.244.70386}
+
 \bibitem[\citeproctext]{ref-castillo_cual_2009}
@@
 \url{https://doi.org/10.1007/s11211-011-0144-5}
+
+\bibitem[\citeproctext]{ref-Castillo2012a_justice}
+Castillo, J. C. (2012). Is {Inequality Becoming Just}? {Changes} in
+{Public Opinion} about {Economic Distribution} in {Chile}.
+\emph{Bulletin of Latin American Research}, \emph{31}(1), 1--18.
+\url{https://doi.org/10.1111/j.1470-9856.2011.00605.x}
 
@@
 
+\bibitem[\citeproctext]{ref-tilly_contentious_2015}
+Tilly, C., \& Tarrow, S. (2015). \emph{Contentious {Politics}} (Second
+Edition, New to this Edition:, Second Edition, New to this Edition:).
+Oxford, New York: Oxford University Press.
+
 \bibitem[\citeproctext]{ref-trump_income_2018}

### 2025-05-22 02-antecedents.qmd
ajustes formato

@@
 In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015].
-These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions, in line with the policy-feedbacks literature [campbell_policy_2012; pierson_when_1993].
+These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions, in line with the policy-feedbacks literature [@campbell_policy_2012; @pierson_when_1993].
 

### 2025-05-22 04-analisis.qmd
ajustes formato

@@
 
-#knitreg(list(m13.1, m13.2,m14.1, m14.2))
+#texreg(list(m13.1, m13.2,m14.1, m14.2))
 ```
@@
 
-texreg::knitreg(list(m1, m2, m3, m4, m5, m6, m7),
-               custom.model.names = c(paste0("Model ", seq(1:7))),
+texreg::texreg(list(m1, m2, m3, m4, m5, m6, m7),
+               custom.model.names = c(paste0("Model ", seq(1:7))),caption.above = T,
                caption = NULL,
@@
                booktabs = F,
-               scalebox = 0.79,
+               scalebox = 0.80,
                include.loglik = FALSE,
@@
 
-texreg::knitreg(list(m8.1,m8.2,m9.1,m9.2),
+texreg::texreg(list(m8.1,m8.2,m9.1,m9.2),
                custom.model.names = c("Model 8", "Model 9", "Model 10", "Model 11"),
                caption = NULL,
+               caption.above = T,
                stars = c(0.05, 0.01, 0.001),
@@
        title = "(B) People are rewarded for their intelligence",
-       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)
-       Confidence intervals at 95%") +
+       caption = "Source: own elaboration with data from ELSOC 2016-2023 (n = 8460). Based in Model 10 and Model 11 in Table 4. Confidence intervals at 95%") +
   theme_ggdist()
@@
 
-texreg::knitreg(list(m10.1, m11.1, m12.1, m10.2, m11.2, m12.2),
+texreg::texreg(list(m10.1, m11.1, m12.1, m10.2, m11.2, m12.2),
                custom.model.names = c("Model 12", "Model 13", "Model 14", "Model 15", "Model 16", "Model 17"),
@@
                digits = 3,
+               caption.above = T,
                custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.",
@@
                booktabs = F,
-               scalebox = 0.8,
+               scalebox = 0.7,
                include.loglik = FALSE,

### 2025-05-22 paper.qmd
ajustes formato

@@
 bibliography: "input/bib/merit-edjust.bib"
-csl: "input/bib/apa6.csl"
+# csl: "input/bib/apa6.csl"
+csl: "input/bib/harvardquarto.csl"
 author:

### 2025-05-22 paper.tex
ajustes formato

@@
 \usepackage{amsmath,amssymb}
+\usepackage{lmodern}
 \usepackage{setspace}
@@
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else  
-    % xetex/luatex font selection
   \setmainfont[]{Times New Roman}
@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
   \fi
   % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \renewcommand\Affilfont{\small}
+\makeatletter
+\makeatother
+\makeatletter
+\makeatother
 \makeatletter
@@
 \makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
 \makeatother
 \makeatletter
-\@ifpackageloaded{caption}{}{\usepackage{caption}}
-\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
 \makeatother
@@
 \fi
-\usepackage{bookmark}
-
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \end{abstract}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
 
 \setstretch{1.15}
-\section{Introduction}\label{introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
@@
 concentrated 76\%, and the wealthiest 1\% captured nearly 38\% of total
-assets (\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}). This
-context of economic disparity has sparked renewed interest in studying
-not only the objective aspects of inequality, such as income and access
-to resources, but also its subjective dimensions, including perceptions,
-beliefs, and associated attitudes
-(\citeproc{ref-janmaat_subjective_2013}{Janmaat, 2013}). This has led,
-among other reasons, to widespread concern in the social sciences about
-the political delegitimization that economic inequality can produce
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}).
-Understanding how people perceive inequality is crucial, as these
-perceptions can influence how societies comprehend and justify the
-distribution of goods and services.
+assets (\protect\hyperlink{ref-chancel_world_2022}{Chancel et al.,
+2022}). This context of economic disparity has sparked renewed interest
+in studying not only the objective aspects of inequality, such as income
+and access to resources, but also its subjective dimensions, including
+perceptions, beliefs, and associated attitudes
+(\protect\hyperlink{ref-janmaat_subjective_2013}{Janmaat, 2013}). This
+has led, among other reasons, to widespread concern in the social
+sciences about the political delegitimization that economic inequality
+can produce (\protect\hyperlink{ref-castillo_perception_2022}{Castillo
+et al., 2022}). Understanding how people perceive inequality is crucial,
+as these perceptions can influence how societies comprehend and justify
+the distribution of goods and services.
 
@@
 individual's subjective assessment of how resources are allocated among
-members of a given society (\citeproc{ref-akyelken_urban_2020}{Akyelken,
-2020}). Regardless of their measurement, various studies have shown that
-this perception often underestimates the gap between the rich and the
-poor, which could have implications for attitudes toward the
-distribution of goods and services
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
-\citeproc{ref-schroder_income_2017}{Schröder, 2017}). Moreover, in
-recent years, there has been a discussion about the subjective and
-objective aspects of inequality, demonstrating inconsistencies between
-the two (\citeproc{ref-trump_income_2018}{Trump, 2018}). This tension is
-relevant because advancing the study of perceptions could help
+members of a given society
+(\protect\hyperlink{ref-akyelken_urban_2020}{Akyelken, 2020}).
+Regardless of their measurement, various studies have shown that this
+perception often underestimates the gap between the rich and the poor,
+which could have implications for attitudes toward the distribution of
+goods and services
+(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
+2022}; \protect\hyperlink{ref-schroder_income_2017}{Schröder, 2017}).
+Moreover, in recent years, there has been a discussion about the
+subjective and objective aspects of inequality, demonstrating
+inconsistencies between the two
+(\protect\hyperlink{ref-trump_income_2018}{Trump, 2018}). This tension
+is relevant because advancing the study of perceptions could help
 understand how perceived inequality affects the distribution of
@@
 inequality, and legitimacy of the economic system
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
-\citeproc{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et al.,
-2020}, \citeproc{ref-garcia-sanchez_perceptions_2018}{2018}). Research
+(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
+2022};
+\protect\hyperlink{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et
+al., 2020},
+\protect\hyperlink{ref-garcia-sanchez_perceptions_2018}{2018}). Research
 shows that greater perceived inequality fosters stronger preferences for
 redistribution, regardless of actual inequality levels
-(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
+(\protect\hyperlink{ref-garcia-sanchez_vicious_2019}{García-Sánchez et
+al., 2019};
+\protect\hyperlink{ref-gimpelson_misperceiving_2018}{Gimpelson \&
 Treisman, 2018}). However, other findings suggest that heightened
@@
 lead to greater justification of inequality
-(\citeproc{ref-Castillo2011}{Castillo, 2011}). A less explored area in
-this regard concerns market justice, which denotes the degree to which
-individuals consider just that social goods and services (e.g., health,
-pensions, education) are allocated according to individual contribution,
-competition, and ability to pay
-(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
-\citeproc{ref-lane_market_1986}{Lane, 1986};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Market justice attitudes
-reflect the belief that the market promotes procedural
+(\protect\hyperlink{ref-Castillo2011}{Castillo, 2011}). A less explored
+area in this regard concerns market justice, which denotes the degree to
+which individuals consider just that social goods and services (e.g.,
+health, pensions, education) are allocated according to individual
+contribution, competition, and ability to pay
+(\protect\hyperlink{ref-kluegel_legitimation_1999}{Kluegel et al.,
+1999}; \protect\hyperlink{ref-lane_market_1986}{Lane, 1986};
+\protect\hyperlink{ref-lindh_public_2015}{Lindh, 2015}). Market justice
+attitudes reflect the belief that the market promotes procedural
 fairness---equality of opportunity---so that outcomes result from
-individual achievement (\citeproc{ref-lane_market_1986}{Lane, 1986}).
-Indeed, Lindh (\citeproc{ref-lindh_public_2015}{2015}) shows that market
-justice preferences are stronger in countries with higher private
-spending on services than in those with more comprehensive welfare
-systems. From this perspective, perceptions of economic inequality may
-be key, as lower perceived gaps can reinforce acceptance of free-market
-systems and justify existing inequalities.
+individual achievement (\protect\hyperlink{ref-lane_market_1986}{Lane,
+1986}). Indeed, Lindh (\protect\hyperlink{ref-lindh_public_2015}{2015})
+shows that market justice preferences are stronger in countries with
+higher private spending on services than in those with more
+comprehensive welfare systems. From this perspective, perceptions of
+economic inequality may be key, as lower perceived gaps can reinforce
+acceptance of free-market systems and justify existing inequalities.
 
@@
 distributive criteria such as effort and talent
-(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
-\citeproc{ref-young_rise_1962}{Young, 1962}). Previous studies have
-demonstrated that people with stronger meritocratic beliefs tend to
+(\protect\hyperlink{ref-davis_principles_2001}{Davis \& Moore, 2001};
+\protect\hyperlink{ref-young_rise_1962}{Young, 1962}). Previous studies
+have demonstrated that people with stronger meritocratic beliefs tend to
 perceive less inequality as they attribute economic differences to
-individual achievements (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
-\citeproc{ref-wilson_role_2003}{Wilson, 2003}) and also to justify more
-inequality as these beliefs are associated with attitudes that
-legitimize status differences
-(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In a context
-like Chile, where the distribution of goods and services is
+individual achievements (\protect\hyperlink{ref-mijs_paradox_2021}{Mijs,
+2021}; \protect\hyperlink{ref-wilson_role_2003}{Wilson, 2003}) and also
+to justify more inequality as these beliefs are associated with
+attitudes that legitimize status differences
+(\protect\hyperlink{ref-batruch_belief_2023}{Batruch et al., 2023}). In
+a context like Chile, where the distribution of goods and services is
 predominantly governed by market logics strongly introduced during the
 military dictatorship (1973-1989)
-(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}), these beliefs can
-play a crucial role in the acceptance of social inequalities.
-Incorporating the perception of meritocracy in this literature allows
-for an understanding of how attitudes toward inequality are shaped not
-only by the perception of economic gaps, but also by beliefs about how
-these gaps are justified. Furthermore, these beliefs are often
-consolidated from an early age, reinforced by institutions promoting
-values such as effort and individual skills to climb socially
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).
+(\protect\hyperlink{ref-boccardo_30_2020}{Boccardo, 2020}), these
+beliefs can play a crucial role in the acceptance of social
+inequalities. Incorporating the perception of meritocracy in this
+literature allows for an understanding of how attitudes toward
+inequality are shaped not only by the perception of economic gaps, but
+also by beliefs about how these gaps are justified. Furthermore, these
+beliefs are often consolidated from an early age, reinforced by
+institutions promoting values such as effort and individual skills to
+climb socially
+(\protect\hyperlink{ref-castillo_socialization_2024}{Castillo et al.,
+2024}; \protect\hyperlink{ref-reynolds_perceptions_2014}{Reynolds \&
+Xian, 2014}).
 
@@
 justifying an unequal system
-(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
-\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). On the other hand, the
-perception of inequality could moderate the impact of meritocracy, as a
-higher perception of economic gaps might challenge the idea that these
-are solely based on merit.
+(\protect\hyperlink{ref-castillo_percepcion_2019}{Castillo et al.,
+2012}; \protect\hyperlink{ref-mijs_paradox_2021}{Mijs, 2021}). On the
+other hand, the perception of inequality could moderate the impact of
+meritocracy, as a higher perception of economic gaps might challenge the
+idea that these are solely based on merit.
 
@@
 justice in a country characterized by high inequality and strong
-free-market influence (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
-\citeproc{ref-flores_top_2020}{Flores et al., 2020}). Additionally, this
-analysis seeks to elucidate how political and social
+free-market influence
+(\protect\hyperlink{ref-boccardo_30_2020}{Boccardo, 2020};
+\protect\hyperlink{ref-flores_top_2020}{Flores et al., 2020}).
+Additionally, this analysis seeks to elucidate how political and social
 contingencies---such as the 2019 and 2022 social movements---might have
@@
 
+\hypertarget{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}{%
 \section{Theoretical views on market justice, inequality perception, and
-meritocracy}\label{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}
-
+meritocracy}\label{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}}
+
+\hypertarget{the-justification-of-market-inequality}{%
 \subsection{The justification of market
-inequality}\label{the-justification-of-market-inequality}
+inequality}\label{the-justification-of-market-inequality}}
 
@@
 rewards based on individual merit. It is possible to trace the concept
-to the understanding of Lane (\citeproc{ref-lane_market_1986}{1986}),
-who makes a contrast between market justice and political justice. The
-author defines market justice as a system of ``earned deserts,'' whereby
-individuals are seen as deserving of a determined distributive outcome
-due to their effort and skills. In contrast, political justice
-emphasizes principles of equality and need, which are often represented
-by the welfare state action through social policies. An important remark
-is that the principles of market justice rely on the assumption that
-markets are neutral and self-regulating arenas, where individuals are
-treated fairly because they face the same formal rules of engagement
-(\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently, the
-legitimacy of market justice stems from the assumption that inequalities
-are not only inevitable but fair---so long as the rules are transparent
-and opportunities are open. In this way, market justice provides a moral
-justification for inequality by framing it as a necessary outcome of
-individual responsibility.
+to the understanding of Lane
+(\protect\hyperlink{ref-lane_market_1986}{1986}), who makes a contrast
+between market justice and political justice. The author defines market
+justice as a system of ``earned deserts,'' whereby individuals are seen
+as deserving of a determined distributive outcome due to their effort
+and skills. In contrast, political justice emphasizes principles of
+equality and need, which are often represented by the welfare state
+action through social policies. An important remark is that the
+principles of market justice rely on the assumption that markets are
+neutral and self-regulating arenas, where individuals are treated fairly
+because they face the same formal rules of engagement
+(\protect\hyperlink{ref-lane_market_1986}{Lane, 1986}). Consequently,
+the legitimacy of market justice stems from the assumption that
+inequalities are not only inevitable but fair---so long as the rules are
+transparent and opportunities are open. In this way, market justice
+provides a moral justification for inequality by framing it as a
+necessary outcome of individual responsibility.
 
@@
 seminal work of Kluegel and Smith
-(\citeproc{ref-kluegel_beliefs_1981}{1981}) who assessed the normative
-foundations that explain public support for economic inequality. Over
-time, this approach has been extended beyond income to include other
-market-mediated outcomes, such as education, healthcare, and/or
-pensions. For example, Von Dem Knesebeck et al.
-(\citeproc{ref-vondemknesebeck_are_2016}{2016}) and Immergut and
-Schneider (\citeproc{ref-immergut_it_2020}{2020}) examine whether
-citizens consider it fair that individuals with higher incomes can
-access better healthcare, while Lee and Stacey
-(\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
-context of education in Australia. These studies usually rely on a
-survey item asking respondents to evaluate the fairness of income-based
-access to welfare services, allowing for comparing justice perceptions
-across different contexts. Similarly, comparative studies have also
-considered these indicators to study the support for market-based
-distribution in welfare systems (\citeproc{ref-lindh_public_2015}{Lindh,
-2015}; \citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). More
+(\protect\hyperlink{ref-kluegel_beliefs_1981}{1981}) who assessed the
+normative foundations that explain public support for economic
+inequality. Over time, this approach has been extended beyond income to
+include other market-mediated outcomes, such as education, healthcare,
+and/or pensions. For example, Von Dem Knesebeck et al.
+(\protect\hyperlink{ref-vondemknesebeck_are_2016}{2016}) and Immergut
+and Schneider (\protect\hyperlink{ref-immergut_it_2020}{2020}) examine
+whether citizens consider it fair that individuals with higher incomes
+can access better healthcare, while Lee and Stacey
+(\protect\hyperlink{ref-lee_fairness_2023}{2023}) apply a similar method
+in the context of education in Australia. These studies usually rely on
+a survey item asking respondents to evaluate the fairness of
+income-based access to welfare services, allowing for comparing justice
+perceptions across different contexts. Similarly, comparative studies
+have also considered these indicators to study the support for
+market-based distribution in welfare systems
+(\protect\hyperlink{ref-lindh_public_2015}{Lindh, 2015};
+\protect\hyperlink{ref-svallfors_political_2007}{Svallfors, 2007}). More
 recently, Castillo et al.
-(\citeproc{ref-castillo_socialization_2024}{2024}) introduced a
+(\protect\hyperlink{ref-castillo_socialization_2024}{2024}) introduced a
 single-item composite measure of market justice to assess student
@@
 higher occupational class, income, and education---are more likely to
-support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
-\& Sachweh, 2019}; \citeproc{ref-svallfors_political_2007}{Svallfors,
-2007}). This tendency reflects not only material self-interest but also
-a broader moral economy, in which winners of the market system
-internalize justifications for the status quo. At the same time,
-political ideology also plays a role ---such as economically
-conservative values --- where more right-wing individuals show higher
-support for meritocracy and more skepticism towards redistribution. This
-is particularly salient in countries with more restricted public
-provision of social services. For example, Castillo et al.
-(\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
-Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
-those with right-leaning individuals are more supportive of market-based
-distribution of welfare. Beyond individual characteristics,
-country-level institutions also play a central role. In liberal welfare
-regimes like those of the United States or the United Kingdom, market
-justice preferences are more widespread, while in coordinated or
-social-democratic regimes---such as Sweden or Germany---citizens are
-generally more critical of market-based inequalities
-(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
-that market justice is also shaped contextually, which is represented by
-the development of welfare institutions, in line with the
-policy-feedbacks literature {[}campbell\_policy\_2012;
-pierson\_when\_1993{]}.
-
-\subsection{Perception of inequality}\label{perception-of-inequality}
+support market justice principles
+(\protect\hyperlink{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
+\protect\hyperlink{ref-svallfors_political_2007}{Svallfors, 2007}). This
+tendency reflects not only material self-interest but also a broader
+moral economy, in which winners of the market system internalize
+justifications for the status quo. At the same time, political ideology
+also plays a role ---such as economically conservative values --- where
+more right-wing individuals show higher support for meritocracy and more
+skepticism towards redistribution. This is particularly salient in
+countries with more restricted public provision of social services. For
+example, Castillo et al.
+(\protect\hyperlink{ref-castillo_socialization_2024}{2024}) in Chile
+and, Lee and Stacey (\protect\hyperlink{ref-lee_fairness_2023}{2023}) in
+Australia show that those with right-leaning individuals are more
+supportive of market-based distribution of welfare. Beyond individual
+characteristics, country-level institutions also play a central role. In
+liberal welfare regimes like those of the United States or the United
+Kingdom, market justice preferences are more widespread, while in
+coordinated or social-democratic regimes---such as Sweden or
+Germany---citizens are generally more critical of market-based
+inequalities (\protect\hyperlink{ref-immergut_it_2020}{Immergut \&
+Schneider, 2020}; \protect\hyperlink{ref-lindh_public_2015}{Lindh,
+2015}). These findings suggest that market justice is also shaped
+contextually, which is represented by the development of welfare
+institutions, in line with the policy-feedbacks literature
+(\protect\hyperlink{ref-campbell_policy_2012}{Campbell, 2012};
+\protect\hyperlink{ref-pierson_when_1993}{Pierson, 1993}).
+
+\hypertarget{perception-of-inequality}{%
+\subsection{Perception of inequality}\label{perception-of-inequality}}
 
 Perceptions of inequality have been associated with attitudes about
-market justice (\citeproc{ref-kluegel_social_1995a}{Kluegel et al.,
-1995}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research
-indicates that lower perceived inequality can reinforce support for
-market-based distributive arrangements by suggesting that the system is
-fair and that outcomes reflect effort and ability
-(\citeproc{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when inequality
-is perceived as excessive or structurally determined, individuals could
-question the legitimacy of market justice and become more supportive of
-redistributive policies
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
-\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}).
+market justice (\protect\hyperlink{ref-kluegel_social_1995a}{Kluegel et
+al., 1995}; \protect\hyperlink{ref-lindh_public_2015}{Lindh, 2015}).
+Research indicates that lower perceived inequality can reinforce support
+for market-based distributive arrangements by suggesting that the system
+is fair and that outcomes reflect effort and ability
+(\protect\hyperlink{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when
+inequality is perceived as excessive or structurally determined,
+individuals could question the legitimacy of market justice and become
+more supportive of redistributive policies
+(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
+2022};
+\protect\hyperlink{ref-garcia-sanchez_vicious_2019}{García-Sánchez et
+al., 2019}).
 
@@
 comparisons, and information environments
-(\citeproc{ref-garcia-castro_perceiving_2020}{García-Castro et al.,
-2020}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
-Treisman, 2018}; \citeproc{ref-mijs_stratified_2016}{Mijs, 2016}).
-Scholars have proposed multiple dimensions of perceived inequality,
-including its magnitude (how significant are the gaps), vertical
-structure (between which groups), the trend over time (increasing or
-decreasing), and legitimacy (whether it is just or not)
-(\citeproc{ref-engelhardt_what_2018}{Engelhardt \& Wagener, 2018};
-\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}). These dimensions encompass both cognitive and normative aspects
-of perceptions of inequality and can vary across societies and social
-groups, depending on exposure, ideology, and personal experience
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
-\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
-2018}).
+(\protect\hyperlink{ref-garcia-castro_perceiving_2020}{García-Castro et
+al., 2020};
+\protect\hyperlink{ref-gimpelson_misperceiving_2018}{Gimpelson \&
+Treisman, 2018}; \protect\hyperlink{ref-mijs_stratified_2016}{Mijs,
+2016}). Scholars have proposed multiple dimensions of perceived
+inequality, including its magnitude (how significant are the gaps),
+vertical structure (between which groups), the trend over time
+(increasing or decreasing), and legitimacy (whether it is just or not)
+(\protect\hyperlink{ref-engelhardt_what_2018}{Engelhardt \& Wagener,
+2018};
+\protect\hyperlink{ref-garcia-sanchez_vicious_2019}{García-Sánchez et
+al., 2019}). These dimensions encompass both cognitive and normative
+aspects of perceptions of inequality and can vary across societies and
+social groups, depending on exposure, ideology, and personal experience
+(\protect\hyperlink{ref-castillo_perception_2022}{Castillo et al.,
+2022};
+\protect\hyperlink{ref-garcia-sanchez_perceptions_2018}{García-Sánchez
+et al., 2018}).
 
@@
 than abstract questions about national income distribution
-(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
-\citeproc{ref-easterbrook_social_2021}{Easterbrook, 2021};
-\citeproc{ref-willis_legitimacy_2015}{Willis et al., 2015}). While it
-enables the estimation of inequality using simple heuristics, this
-method is not without challenges. For instance, people often lack
-reliable knowledge about the earnings of those at the top of the income
-ladder, which leads to high variability in responses and the use of
-biased mental shortcuts (\citeproc{ref-knell_perceptions_2020}{Knell \&
-Stix, 2020}). Despite this, perceived wage gaps are strong predictors of
-political attitudes
-(\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
-2018}; \citeproc{ref-pedersen_attitudes_2019}{Pedersen \& Mutz, 2019}),
-making them a valuable tool for understanding public responses to
-economic disparities.
+(\protect\hyperlink{ref-castillo_percepcion_2019}{Castillo et al.,
+2012}; \protect\hyperlink{ref-easterbrook_social_2021}{Easterbrook,
+2021}; \protect\hyperlink{ref-willis_legitimacy_2015}{Willis et al.,
+2015}). While it enables the estimation of inequality using simple
+heuristics, this method is not without challenges. For instance, people
+often lack reliable knowledge about the earnings of those at the top of
+the income ladder, which leads to high variability in responses and the
+use of biased mental shortcuts
+(\protect\hyperlink{ref-knell_perceptions_2020}{Knell \& Stix, 2020}).
+Despite this, perceived wage gaps are strong predictors of political
+attitudes
+(\protect\hyperlink{ref-garcia-sanchez_perceptions_2018}{García-Sánchez
+et al., 2018}; \protect\hyperlink{ref-pedersen_attitudes_2019}{Pedersen
+\& Mutz, 2019}), making them a valuable tool for understanding public
+responses to economic disparities.
 
@@
 which captures general concern or discomfort rather than a specific
-perception (\citeproc{ref-Castillo2011}{Castillo, 2011};
-\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}). These items mix cognitive estimations with affective
+perception (\protect\hyperlink{ref-Castillo2011}{Castillo, 2011};
+\protect\hyperlink{ref-garcia-sanchez_vicious_2019}{García-Sánchez et
+al., 2019}). These items mix cognitive estimations with affective
 evaluations, complicating the interpretation of what respondents
@@
 comparative measures, as well as between ideal and actual estimates of
-economic gaps (\citeproc{ref-auspurg_why_2017}{Auspurg et al., 2017};
-\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De
-Carvalho, 2022}).
-
-\subsection{Perception of meritocracy}\label{perception-of-meritocracy}
+economic gaps (\protect\hyperlink{ref-auspurg_why_2017}{Auspurg et al.,
+2017};
+\protect\hyperlink{ref-garcia-sanchez_creencias_2022}{García-Sánchez \&
+De Carvalho, 2022}).
+
+\hypertarget{perception-of-meritocracy}{%
+\subsection{Perception of meritocracy}\label{perception-of-meritocracy}}
 
@@
 functions. As initially conceptualized by Michael Young
-(\citeproc{ref-young_rise_1962}{1962}), the term was meant to critique a
-system in which merit-based stratification becomes a new form of
-inequality. However, over time, meritocracy has been widely supported in
-many societies as a fair and desirable principle of distribution,
+(\protect\hyperlink{ref-young_rise_1962}{1962}), the term was meant to
+critique a system in which merit-based stratification becomes a new form
+of inequality. However, over time, meritocracy has been widely supported
+in many societies as a fair and desirable principle of distribution,
 particularly within liberal democracies and market-oriented economies
-(\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
-\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
-perspective, the belief in meritocracy is more than a cognitive
-assessment; it reflects a moral lens through which individuals interpret
-inequality. People who believe that success results from hard work and
-talent are more likely to view social and economic disparities as
-legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
-\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}).
-Conversely, if they see outcomes as driven by luck, social origin, or
-systemic barriers, inequality is more likely to be perceived as unjust.
-This distinction becomes crucial in societies with persistent structural
-inequality, where public narratives often emphasize personal
+(\protect\hyperlink{ref-mijs_paradox_2021}{Mijs, 2021};
+\protect\hyperlink{ref-sandel_tyranny_2020}{Sandel, 2020}). From a
+sociological perspective, the belief in meritocracy is more than a
+cognitive assessment; it reflects a moral lens through which individuals
+interpret inequality. People who believe that success results from hard
+work and talent are more likely to view social and economic disparities
+as legitimate (\protect\hyperlink{ref-batruch_belief_2023}{Batruch et
+al., 2023}; \protect\hyperlink{ref-castillo_percepcion_2019}{Castillo et
+al., 2012}). Conversely, if they see outcomes as driven by luck, social
+origin, or systemic barriers, inequality is more likely to be perceived
+as unjust. This distinction becomes crucial in societies with persistent
+structural inequality, where public narratives often emphasize personal
 responsibility and merit while overlooking entrenched disadvantages.
@@
 how inequality is justified in specific domains
-(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}).
-The relevance of this distinction is supported by recent studies, which
-show that individuals respond differently to these dimensions. For
-instance, perceptions that effort is rewarded in society are more
-strongly associated with positive evaluations of fairness and acceptance
-of unequal outcomes (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
-2023}). This may be because effort is seen as a controllable and morally
+(\protect\hyperlink{ref-castillo_multidimensional_2023}{Castillo et al.,
+2023}). The relevance of this distinction is supported by recent
+studies, which show that individuals respond differently to these
+dimensions. For instance, perceptions that effort is rewarded in society
+are more strongly associated with positive evaluations of fairness and
+acceptance of unequal outcomes
+(\protect\hyperlink{ref-batruch_belief_2023}{Batruch et al., 2023}).
+This may be because effort is seen as a controllable and morally
 virtuous trait, whereas talent is often perceived as a natural
@@
 preferences for redistribution and justice
-(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
-2025}) Meritocratic beliefs serve as symbolic justifications for unequal
-outcomes, particularly when access is stratified by income or social
-opportunity. Prior studies have shown that individuals who perceive
-higher levels of meritocracy tend to express stronger support for
-unequal distributions that reflect market outcomes
-(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
-\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
+(\protect\hyperlink{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina
+et al., 2025}) Meritocratic beliefs serve as symbolic justifications for
+unequal outcomes, particularly when access is stratified by income or
+social opportunity. Prior studies have shown that individuals who
+perceive higher levels of meritocracy tend to express stronger support
+for unequal distributions that reflect market outcomes
+(\protect\hyperlink{ref-castillo_percepcion_2019}{Castillo et al.,
+2012}; \protect\hyperlink{ref-castillo_socialization_2024}{Castillo et
+al., 2024}).
 
@@
 conditions, reducing support for redistributive measures and increasing
-the stigmatization of the poor (\citeproc{ref-hoyt_mindsets_2023}{Hoyt
-et al., 2023}). This reinforces negative stereotypes and reduces empathy
-toward individuals from lower socioeconomic backgrounds. Moreover,
-Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argues
-that meritocratic narratives can serve as feedback mechanisms that shape
+the stigmatization of the poor
+(\protect\hyperlink{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). This
+reinforces negative stereotypes and reduces empathy toward individuals
+from lower socioeconomic backgrounds. Moreover, Busemeyer et al.
+(\protect\hyperlink{ref-busemeyer_positive_2021}{2021}) argues that
+meritocratic narratives can serve as feedback mechanisms that shape
 public opinion and well-being by framing individuals' understanding of
@@
 
-\subsection{This study}\label{this-study}
+\hypertarget{this-study}{%
+\subsection{This study}\label{this-study}}
 
@@
 
-\section{Data, Variables and Methods}\label{data-variables-and-methods}
-
-\subsection{Data}\label{data}
+\hypertarget{data-variables-and-methods}{%
+\section{Data, Variables and Methods}\label{data-variables-and-methods}}
+
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
 
@@
 population, with a response rate of 62.4\%
-(\citeproc{ref-elsoc_estudio_2022}{ELSOC, 2022}). This study uses six
-waves: 2016, 2017, 2018, 2019, 2022, and 2023. The 2021 wave was
-excluded because a reduced version of the questionnaire omitted key
+(\protect\hyperlink{ref-elsoc_estudio_2022}{ELSOC, 2022}). This study
+uses six waves: 2016, 2017, 2018, 2019, 2022, and 2023. The 2021 wave
+was excluded because a reduced version of the questionnaire omitted key
 variables of interest. Between waves 1 and 6, panel attrition reached
@@
 
-\subsection{Variables}\label{variables}
+\hypertarget{variables}{%
+\subsection{Variables}\label{variables}}
 
@@
 
+\hypertarget{tbl-summary1}{}
 \begin{longtable}[]{@{}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1122}}@{}}
-\caption{Dependent variables for the first wave
-(2016)}\label{tbl-summary1}\tabularnewline
-\toprule\noalign{}
+\caption{\label{tbl-summary1}Dependent variables for the first wave
+(2016)}\tabularnewline
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endfirsthead
-\toprule\noalign{}
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Health distributive justice &
@@
 \end{minipage} \\
+\bottomrule()
 \end{longtable}
@@
 through the perceived wage gap
-(\citeproc{ref-castillo_cual_2009}{Castillo, 2009};
-\citeproc{ref-gijsberts_thelegitimation_1999}{Gijsberts, 1999};
-\citeproc{ref-hadler_why_2005}{Hadler, 2005}). This measure is derived
-from the salary gap between the perceived salaries of jobs at opposite
-ends of the occupational hierarchy. Specifically, it relies on the
-division between the perceived salary of a large-company president and
-that of an unskilled worker (\citeproc{ref-Castillo2011}{Castillo,
-2011}). Higher values of this term indicate a greater perception of
-economic inequality between occupations located at the extremes of the
-status continuum. This measure includes a logarithmic term in order to
-adjust income magnitudes (usually fewer cases with high income):
+(\protect\hyperlink{ref-castillo_cual_2009}{Castillo, 2009};
+\protect\hyperlink{ref-gijsberts_thelegitimation_1999}{Gijsberts, 1999};
+\protect\hyperlink{ref-hadler_why_2005}{Hadler, 2005}). This measure is
+derived from the salary gap between the perceived salaries of jobs at
+opposite ends of the occupational hierarchy. Specifically, it relies on
+the division between the perceived salary of a large-company president
+and that of an unskilled worker
+(\protect\hyperlink{ref-Castillo2011}{Castillo, 2011}). Higher values of
+this term indicate a greater perception of economic inequality between
+occupations located at the extremes of the status continuum. This
+measure includes a logarithmic term in order to adjust income magnitudes
+(usually fewer cases with high income):
 
@@
 through two components, namely effort and talent
-(\citeproc{ref-young_rise_1962}{Young, 1962}). The item used to gauge
-effort is: ``In Chile, people are rewarded for their efforts,'' while
-the item for talent is: ``In Chile, people are rewarded for their
+(\protect\hyperlink{ref-young_rise_1962}{Young, 1962}). The item used to
+gauge effort is: ``In Chile, people are rewarded for their efforts,''
+while the item for talent is: ``In Chile, people are rewarded for their
 intelligence and skills''. In both cases, respondents indicate their
@@
 
+\hypertarget{tbl-summary2}{}
 \begin{longtable}[]{@{}
@@
   >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1038}}@{}}
-\caption{Independent variables ELSOC survey (descriptives for first wave
-2016)}\label{tbl-summary2}\tabularnewline
-\toprule\noalign{}
+\caption{\label{tbl-summary2}Independent variables ELSOC survey
+(descriptives for first wave 2016)}\tabularnewline
+\toprule()
 \begin{minipage}[b]{\linewidth}\raggedright
@@
 \end{minipage} \\
-\midrule\noalign{}
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
 Inequality gap perception & \begin{minipage}[t]{\linewidth}\raggedright
@@
 been previously shown to influence market justice preferences
-significantly (\citeproc{ref-castillo_socialization_2024}{Castillo et
-al., 2024}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding
-attitudinal variables, we include political identification (1=Left,
-2=Center, 3=Right, 4=No identification) and subjective social status
-(ranging from 1 to 10) as they may affect the relationship between
-market justice preferences, perceptions of inequality, and meritocracy
-(\citeproc{ref-schneider_poverty_2015}{Schneider \& Castillo, 2015}).
-
-\subsection{Methods}\label{methods}
+significantly
+(\protect\hyperlink{ref-castillo_socialization_2024}{Castillo et al.,
+2024}; \protect\hyperlink{ref-lindh_public_2015}{Lindh, 2015}).
+Regarding attitudinal variables, we include political identification
+(1=Left, 2=Center, 3=Right, 4=No identification) and subjective social
+status (ranging from 1 to 10) as they may affect the relationship
+between market justice preferences, perceptions of inequality, and
+meritocracy (\protect\hyperlink{ref-schneider_poverty_2015}{Schneider \&
+Castillo, 2015}).
+
+\hypertarget{methods}{%
+\subsection{Methods}\label{methods}}
 
@@
 nested in survey waves, we employ longitudinal multilevel linear models
-(\citeproc{ref-singer_applied_2009}{Singer \& Willett, 2009}). In a
-panel-data framework, within-person effects capture how shifts in
+(\protect\hyperlink{ref-singer_applied_2009}{Singer \& Willett, 2009}).
+In a panel-data framework, within-person effects capture how shifts in
 individual-level variables across waves are associated with variations
@@
 All the analyses were conducted using R software and the \emph{lme4}
-package (\citeproc{ref-bates_fitting_2015}{Bates et al., 2015}).
-
-\section{Results}\label{results}
-
-\subsection{Descriptive}\label{descriptive}
+package (\protect\hyperlink{ref-bates_fitting_2015}{Bates et al.,
+2015}).
+
+\hypertarget{results}{%
+\section{Results}\label{results}}
+
+\hypertarget{descriptive}{%
+\subsection{Descriptive}\label{descriptive}}
 
@@
 
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-alluvial-1.pdf}
+
+}
+
 \caption{\label{fig-alluvial}Change in the justification of inequality
@@
 
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-alluvial-1.pdf}
-
-}
-
-\end{figure}%
+\end{figure}
 
@@
 
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-meanchange-1.pdf}
+
+}
+
 \caption{\label{fig-meanchange}Change in the mean of market justice
@@
 
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-meanchange-1.pdf}
-
-}
-
-\end{figure}%
+\end{figure}
 
@@
 
+{\centering \includegraphics[width=0.9\textwidth,height=\textheight]{paper_files/figure-pdf/fig-matrix-1.pdf}
+
+}
+
 \caption{\label{fig-matrix}Correlation matrix of the main variables for
@@
 
-\centering{
-
-\includegraphics[width=0.9\textwidth,height=\textheight]{paper_files/figure-pdf/fig-matrix-1.pdf}
-
-}
-
-\end{figure}%
-
-\subsection{Multilevel models}\label{multilevel-models}
+\end{figure}
+
+\hypertarget{multilevel-models}{%
+\subsection{Multilevel models}\label{multilevel-models}}
 
@@
 (within) and group-level (between) effects. The intraclass correlation
-(\citeproc{ref-hoxMultilevelAnalysisTechniques2017a}{Hox et al., 2017})
-from the empty model (see Supplementary Material), which decomposes the
-variance of market justice preferences, is 0.23, indicating that
-approximately 23\% of the variation is attributable to differences
-between individuals. Complementary, 77\% of the variation corresponds to
-within-individual differences over time.
-
+(\protect\hyperlink{ref-hoxMultilevelAnalysisTechniques2017a}{Hox et
+al., 2017}) from the empty model (see Supplementary Material), which
+decomposes the variance of market justice preferences, is 0.23,
+indicating that approximately 23\% of the variation is attributable to
+differences between individuals. Complementary, 77\% of the variation
+corresponds to within-individual differences over time.
+
+\hypertarget{tbl-modelos}{}
 \begin{table}
-
-\caption{\label{tbl-modelos}Longitudinal multilevel models for market
-justice preferences}
-
-\centering{
-
-~
-
-Model 1
-
-Model 2
-
-Model 3
-
-Model 4
-
-Model 5
-
-Model 6
-
-Model 7
-
-Intercept
-
-1.968***
-
-2.015***
-
-2.030***
-
-2.028***
-
-2.169***
-
-1.456***
-
-1.547***
-
-~
-
-(0.021)
-
-(0.035)
-
-(0.035)
-
-(0.035)
-
-(0.070)
-
-(0.097)
-
-(0.115)
-
-Wave (Ref.= 2016)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2017
-
--0.167***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.027)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2018
-
--0.025
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.027)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2019
-
--0.046
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.027)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2022
-
-0.285***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.028)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2023
-
-0.284***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.027)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Wave
-
-~
-
--0.121***
-
--0.125***
-
--0.128***
-
--0.128***
-
--0.128***
-
--0.129***
-
-~
-
-~
-
-(0.022)
-
-(0.022)
-
-(0.022)
-
-(0.022)
-
-(0.022)
-
-(0.022)
-
-Wave\^{}2
-
-~
-
-0.028***
-
-0.028***
-
-0.029***
-
-0.029***
-
-0.029***
-
-0.029***
-
-~
-
-~
-
-(0.003)
-
-(0.003)
-
-(0.003)
-
-(0.003)
-
-(0.003)
-
-(0.003)
-
-Perception inequality (WE)
-
-~
-
-~
-
--0.039***
-
--0.034***
-
--0.034***
-
--0.034***
-
--0.034***
-
-~
-
-~
-
-~
-
-(0.010)
-
-(0.010)
-
-(0.010)
-
-(0.010)
-
-(0.010)
-
-Merit: Effort (WE)
-
-~
-
-~
-
-~
-
-0.070***
-
-0.070***
-
-0.069***
-
-0.069***
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-Merit: Talent (WE)
-
-~
-
-~
-
-~
-
-0.027*
-
-0.027*
-
-0.027*
-
-0.027*
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-Perception inequality (BE)
-
-~
-
-~
-
-~
-
-~
-
--0.042*
-
--0.006
-
--0.033
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.018)
-
-(0.018)
-
-(0.018)
-
-Merit: Effort (BE)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.186***
-
-0.178***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.032)
-
-(0.031)
-
-Merit: Talent (BE)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.040
-
-0.019
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.032)
-
-(0.031)
-
-Controls
-
-No
-
-No
-
-No
-
-No
-
-No
-
-No
-
-Yes
-
-BIC
-
-20780.594
-
-20780.617
-
-20781.113
-
-20726.753
-
-20736.627
-
-20655.172
-
-20749.047
-
-Numb. obs.
-
-8643
-
-8643
-
-8643
-
-8643
-
-8643
-
-8643
-
-8643
-
-Num. groups: individuals
-
-1687
-
-1687
-
-1687
-
-1687
-
-1687
-
-1687
-
-1687
-
-Var: individuals (Intercept)
-
-0.172
-
-0.223
-
-0.219
-
-0.209
-
-0.207
-
-0.182
-
-0.171
-
-Var: Residual
-
-0.530
-
-0.499
-
-0.498
-
-0.495
-
-0.495
-
-0.495
-
-0.495
-
-Var: individuals, wave
-
-~
-
-0.011
-
-0.010
-
-0.009
-
-0.009
-
-0.009
-
-0.009
-
-Cov: individuals (Intercept), wave
-
-~
-
--0.025
-
--0.024
-
--0.021
-
--0.020
-
--0.019
-
--0.019
-
-Note: Cells contain regression coefficients with standard errors in
-parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
-\textless{} 0.05.
-
+\caption{\label{tbl-modelos}Longitudinal multilevel models for market justice preferences }
+\begin{center}
+\scalebox{0.8}{
+\begin{tabular}{l c c c c c c c}
+\hline
+ & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 \\
+\hline
+Intercept                          & $1.968^{***}$  & $2.015^{***}$  & $2.030^{***}$  & $2.028^{***}$  & $2.169^{***}$  & $1.456^{***}$  & $1.547^{***}$  \\
+                                   & $(0.021)$      & $(0.035)$      & $(0.035)$      & $(0.035)$      & $(0.070)$      & $(0.097)$      & $(0.115)$      \\
+Wave (Ref.= 2016)                  &                &                &                &                &                &                &                \\
+                                   &                &                &                &                &                &                &                \\
+\quad Wave 2017                    & $-0.167^{***}$ &                &                &                &                &                &                \\
+                                   & $(0.027)$      &                &                &                &                &                &                \\
+\quad Wave 2018                    & $-0.025$       &                &                &                &                &                &                \\
+                                   & $(0.027)$      &                &                &                &                &                &                \\
+\quad Wave 2019                    & $-0.046$       &                &                &                &                &                &                \\
+                                   & $(0.027)$      &                &                &                &                &                &                \\
+\quad Wave 2022                    & $0.285^{***}$  &                &                &                &                &                &                \\
+                                   & $(0.028)$      &                &                &                &                &                &                \\
+\quad Wave 2023                    & $0.284^{***}$  &                &                &                &                &                &                \\
+                                   & $(0.027)$      &                &                &                &                &                &                \\
+Wave                               &                & $-0.121^{***}$ & $-0.125^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.129^{***}$ \\
+                                   &                & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      \\
+Wave$^2$                           &                & $0.028^{***}$  & $0.028^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  \\
+                                   &                & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      \\
+Perception inequality (WE)         &                &                & $-0.039^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ \\
+                                   &                &                & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
+Merit: Effort (WE)                 &                &                &                & $0.070^{***}$  & $0.070^{***}$  & $0.069^{***}$  & $0.069^{***}$  \\
+                                   &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
+Merit: Talent (WE)                 &                &                &                & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    \\
+                                   &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
+Perception inequality (BE)         &                &                &                &                & $-0.042^{*}$   & $-0.006$       & $-0.033$       \\
+                                   &                &                &                &                & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
+Merit: Effort (BE)                 &                &                &                &                &                & $0.186^{***}$  & $0.178^{***}$  \\
+                                   &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
+Merit: Talent (BE)                 &                &                &                &                &                & $0.040$        & $0.019$        \\
+                                   &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
+\hline
+Controls                           & No             & No             & No             & No             & No             & No             & Yes            \\
+BIC                                & $20780.594$    & $20780.617$    & $20781.113$    & $20726.753$    & $20736.627$    & $20655.172$    & $20749.047$    \\
+Numb. obs.                         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         \\
+Num. groups: individuals           & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         \\
+Var: individuals (Intercept)       & $0.172$        & $0.223$        & $0.219$        & $0.209$        & $0.207$        & $0.182$        & $0.171$        \\
+Var: Residual                      & $0.530$        & $0.499$        & $0.498$        & $0.495$        & $0.495$        & $0.495$        & $0.495$        \\
+Var: individuals, wave             & $$             & $0.011$        & $0.010$        & $0.009$        & $0.009$        & $0.009$        & $0.009$        \\
+Cov: individuals (Intercept), wave & $$             & $-0.025$       & $-0.024$       & $-0.021$       & $-0.020$       & $-0.019$       & $-0.019$       \\
+\hline
+\multicolumn{8}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
+\end{tabular}
 }
-
-\end{table}%
+\label{table:coefficients}
+\end{center}
+\end{table}
 
@@
 
+\hypertarget{tbl-interactions1}{}
 \begin{table}
-
-\caption{\label{tbl-interactions1}Interactions for meritocracy,
-perceived economic inequality and market justice preferences}
-
-\centering{
-
-~
-
-Model 8
-
-Model 9
-
-Model 10
-
-Model 11
-
-Intercept
-
-1.561***
-
-1.555***
-
-2.187***
-
-2.340***
-
-~
-
-(0.115)
-
-(0.115)
-
-(0.265)
-
-(0.281)
-
-Perception inequality (WE)
-
--0.035***
-
--0.035***
-
--0.035***
-
--0.035***
-
-~
-
-(0.010)
-
-(0.010)
-
-(0.010)
-
-(0.010)
-
-Merit: Effort (WE)
-
-0.072***
-
-0.074***
-
-0.073***
-
-0.073***
-
-~
-
-(0.013)
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-Merit: Talent (WE)
-
-0.031*
-
-0.031*
-
-0.030*
-
-0.030*
-
-~
-
-(0.012)
-
-(0.013)
-
-(0.012)
-
-(0.012)
-
-Perception inequality (BE)
-
--0.032
-
--0.033
-
--0.213**
-
--0.255***
-
-~
-
-(0.018)
-
-(0.018)
-
-(0.073)
-
-(0.077)
-
-Merit: Effort (BE)
-
-0.177***
-
-0.173***
-
--0.062
-
-0.179***
-
-~
-
-(0.032)
-
-(0.031)
-
-(0.097)
-
-(0.032)
-
-Merit: Talent (BE)
-
-0.015
-
-0.021
-
-0.014
-
--0.266**
-
-~
-
-(0.031)
-
-(0.031)
-
-(0.031)
-
-(0.099)
-
-Merit: Effort (WE) x Perception inequality (WE)
-
-0.017
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-~
-
-~
-
-~
-
-Merit: Talent (WE) x Perception inequality (WE)
-
-~
-
-0.011
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-~
-
-~
-
-Merit: Effort (BE) x Perception inequality (BE)
-
-~
-
-~
-
-0.072**
-
-~
-
-~
-
-~
-
-~
-
-(0.028)
-
-~
-
-Merit: Talent (BE) x Perception inequality (BE)
-
-~
-
-~
-
-~
-
-0.082**
-
-~
-
-~
-
-~
-
-~
-
-(0.027)
-
-Controls
-
-Yes
-
-Yes
-
-Yes
-
-Yes
-
-BIC
-
-20842.694
-
-20835.920
-
-20808.677
-
-20806.384
-
-Numb. obs.
-
-8643
-
-8643
-
-8643
-
-8643
-
-Num. groups: individuals
-
-1687
-
-1687
-
-1687
-
-1687
-
-Var: individuals (Intercept)
-
-0.145
-
-0.145
-
-0.141
-
-0.141
-
-Var: individuals, merit effort cwc
-
-0.020
-
-~
-
-~
-
-~
-
-Var: individuals, perception inequality cwc
-
-0.000
-
-0.000
-
-~
-
-~
-
-Cov: individuals (Intercept), merit effort cwc
-
-0.004
-
-~
-
-~
-
-~
-
-Cov: individuals (Intercept), perception inequality cwc
-
--0.006
-
--0.005
-
-~
-
-~
-
-Cov: individuals, merit effort cwc, perception inequality cwc
-
-0.000
-
-~
-
-~
-
-~
-
-Var: Residuals
-
-0.512
-
-0.511
-
-0.528
-
-0.528
-
-Var: individuals, merit talent cwc
-
-~
-
-0.020
-
-~
-
-~
-
-Cov: individuals (Intercept), merit talent cwc
-
-~
-
-0.012
-
-~
-
-~
-
-Cov: individuals, merit talent cwc, perception inequality cwc
-
-~
-
-0.001
-
-~
-
-~
-
-Note: Cells contain regression coefficients with standard errors in
-parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
-\textless{} 0.05. CWC = centered within group.
-
+\caption{\label{tbl-interactions1}Interactions for meritocracy, perceived economic inequality and market
+justice preferences }
+\begin{center}
+\scalebox{0.8}{
+\begin{tabular}{l c c c c}
+\hline
+ & Model 8 & Model 9 & Model 10 & Model 11 \\
+\hline
+Intercept                                                     & $1.561^{***}$  & $1.555^{***}$  & $2.187^{***}$  & $2.340^{***}$  \\
+                                                              & $(0.115)$      & $(0.115)$      & $(0.265)$      & $(0.281)$      \\
+Perception inequality (WE)                                    & $-0.035^{***}$ & $-0.035^{***}$ & $-0.035^{***}$ & $-0.035^{***}$ \\
+                                                              & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
+Merit: Effort (WE)                                            & $0.072^{***}$  & $0.074^{***}$  & $0.073^{***}$  & $0.073^{***}$  \\
+                                                              & $(0.013)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
+Merit: Talent (WE)                                            & $0.031^{*}$    & $0.031^{*}$    & $0.030^{*}$    & $0.030^{*}$    \\
+                                                              & $(0.012)$      & $(0.013)$      & $(0.012)$      & $(0.012)$      \\
+Perception inequality (BE)                                    & $-0.032$       & $-0.033$       & $-0.213^{**}$  & $-0.255^{***}$ \\
+                                                              & $(0.018)$      & $(0.018)$      & $(0.073)$      & $(0.077)$      \\
+Merit: Effort (BE)                                            & $0.177^{***}$  & $0.173^{***}$  & $-0.062$       & $0.179^{***}$  \\
+                                                              & $(0.032)$      & $(0.031)$      & $(0.097)$      & $(0.032)$      \\
+Merit: Talent (BE)                                            & $0.015$        & $0.021$        & $0.014$        & $-0.266^{**}$  \\
+                                                              & $(0.031)$      & $(0.031)$      & $(0.031)$      & $(0.099)$      \\
+Merit: Effort (WE) x Perception inequality (WE)               & $0.017$        &                &                &                \\
+                                                              & $(0.012)$      &                &                &                \\
+Merit: Talent (WE) x Perception inequality (WE)               &                & $0.011$        &                &                \\
+                                                              &                & $(0.012)$      &                &                \\
+Merit: Effort (BE) x Perception inequality (BE)               &                &                & $0.072^{**}$   &                \\
+                                                              &                &                & $(0.028)$      &                \\
+Merit: Talent (BE) x Perception inequality (BE)               &                &                &                & $0.082^{**}$   \\
+                                                              &                &                &                & $(0.027)$      \\
+\hline
+Controls                                                      & Yes            & Yes            & Yes            & Yes            \\
+BIC                                                           & $20842.694$    & $20835.920$    & $20808.677$    & $20806.384$    \\
+Numb. obs.                                                    & $8643$         & $8643$         & $8643$         & $8643$         \\
+Num. groups: individuals                                      & $1687$         & $1687$         & $1687$         & $1687$         \\
+Var: individuals (Intercept)                                  & $0.145$        & $0.145$        & $0.141$        & $0.141$        \\
+Var: individuals, merit effort cwc                            & $0.020$        & $$             & $$             & $$             \\
+Var: individuals, perception inequality cwc                   & $0.000$        & $0.000$        & $$             & $$             \\
+Cov: individuals (Intercept), merit effort cwc                & $0.004$        & $$             & $$             & $$             \\
+Cov: individuals (Intercept), perception inequality cwc       & $-0.006$       & $-0.005$       & $$             & $$             \\
+Cov: individuals, merit effort cwc, perception inequality cwc & $0.000$        & $$             & $$             & $$             \\
+Var: Residuals                                                & $0.512$        & $0.511$        & $0.528$        & $0.528$        \\
+Var: individuals, merit talent cwc                            & $$             & $0.020$        & $$             & $$             \\
+Cov: individuals (Intercept), merit talent cwc                & $$             & $0.012$        & $$             & $$             \\
+Cov: individuals, merit talent cwc, perception inequality cwc & $$             & $0.001$        & $$             & $$             \\
+\hline
+\multicolumn{5}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$. CWC = centered within group.}}
+\end{tabular}
 }
-
-\end{table}%
+\label{table:coefficients}
+\end{center}
+\end{table}
 
 \begin{figure}[H]
+
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-interact-1.pdf}
+
+}
 
@@
 
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-interact-1.pdf}
-
-}
-
-\end{figure}%
+\end{figure}
 
@@
 
+\hypertarget{tbl-interactions2}{}
 \begin{table}
-
-\caption{\label{tbl-interactions2}Interactions for meritocracy,
-perceived economic inequality and market justice preferences}
-
-\centering{
-
-~
-
-Model 12
-
-Model 13
-
-Model 14
-
-Model 15
-
-Model 16
-
-Model 17
-
-Intercept
-
-1.543***
-
-1.527***
-
-1.521***
-
-1.680***
-
-1.374***
-
-1.362***
-
-~
-
-(0.115)
-
-(0.114)
-
-(0.114)
-
-(0.139)
-
-(0.133)
-
-(0.134)
-
-Wave x Perception inequality (WE)
-
--0.004
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.006)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Wave x Merit: effort (WE)
-
-~
-
--0.008
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.006)
-
-~
-
-~
-
-~
-
-~
-
-Wave x Merit: talent (WE)
-
-~
-
-~
-
--0.006
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.006)
-
-~
-
-~
-
-~
-
-Wave x Perception inequality (BE)
-
-~
-
-~
-
-~
-
-0.011
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.007)
-
-~
-
-~
-
-Wave x Merit: effort (BE)
-
-~
-
-~
-
-~
-
-~
-
--0.021*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.008)
-
-~
-
-Wave x Merit: talent (BE)
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.020*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.008)
-
-Controls
-
-Yes
-
-Yes
-
-Yes
-
-Yes
-
-Yes
-
-Yes
-
-BIC
-
-20792.207
-
-20767.371
-
-20754.324
-
-20782.059
-
-20749.596
-
-20754.417
-
-Numb. obs.
-
-8643
-
-8643
-
-8643
-
-8643
-
-8643
-
-8643
-
-Num. groups: individuals
-
-1687
-
-1687
-
-1687
-
-1687
-
-1687
-
-1687
-
-Var: individuals (Intercept)
-
-0.176
-
-0.167
-
-0.168
-
-0.383
-
-0.254
-
-0.010
-
-Var: individuals, perception inequality cwc
-
-0.000
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Var: individuals, wave
-
-0.010
-
-0.009
-
-0.009
-
-0.009
-
-0.009
-
-0.009
-
-Cov: individuals (Intercept), perception inequality cwc
-
--0.007
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Cov: individuals (Intercept), wave
-
--0.021
-
--0.018
-
--0.018
-
--0.048
-
--0.003
-
-0.006
-
-Cov: individuals, perception inequality cwc, wave
-
-0.002
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Var: Residuals
-
-0.495
-
-0.480
-
-0.479
-
-0.494
-
-0.494
-
-0.494
-
-Var: individuals, merit effort cwc
-
-~
-
-0.019
-
-~
-
-~
-
-~
-
-~
-
-Cov: individuals (Intercept), merit effort cwc
-
-~
-
-0.023
-
-~
-
-~
-
-~
-
-~
-
-Cov: individuals, merit effort cwc, wave
-
-~
-
--0.005
-
-~
-
-~
-
-~
-
-~
-
-Var: individuals, merit talent cwc
-
-~
-
-~
-
-0.019
-
-~
-
-~
-
-~
-
-Cov: individuals (Intercept), merit talent cwc
-
-~
-
-~
-
-0.034
-
-~
-
-~
-
-~
-
-Cov: individuals, merit talent cwc, wave
-
-~
-
-~
-
--0.006
-
-~
-
-~
-
-~
-
-Var: individuals, perception inequality mean
-
-~
-
-~
-
-~
-
-0.008
-
-~
-
-~
-
-Cov: individuals (Intercept), perception inequality mean
-
-~
-
-~
-
-~
-
--0.045
-
-~
-
-~
-
-Cov: individuals, perception inequality mean, wave
-
-~
-
-~
-
-~
-
-0.009
-
-~
-
-~
-
-Var: individuals, merit effort mean
-
-~
-
-~
-
-~
-
-~
-
-0.053
-
-~
-
-Cov: individuals (Intercept), merit effort mean
-
-~
-
-~
-
-~
-
-~
-
--0.089
-
-~
-
-Cov: individuals, merit effort mean, wave
-
-~
-
-~
-
-~
-
-~
-
--0.006
-
-~
-
-Var: individuals, merit talent mean
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.025
-
-Cov: individuals (Intercept), merit talent mean
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.007
-
-Cov: individuals, merit talent mean, wave
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.009
-
-Note: Cells contain regression coefficients with standard errors in
-parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
-\textless{} 0.05. CWC = centered within group.
-
+\caption{\label{tbl-interactions2}Interactions for meritocracy, perceived economic inequality and market
+justice preferences }
+\begin{center}
+\scalebox{0.7}{
+\begin{tabular}{l c c c c c c}
+\hline
+ & Model 12 & Model 13 & Model 14 & Model 15 & Model 16 & Model 17 \\
+\hline
+Intercept                                                & $1.543^{***}$ & $1.527^{***}$ & $1.521^{***}$ & $1.680^{***}$ & $1.374^{***}$ & $1.362^{***}$ \\
+                                                         & $(0.115)$     & $(0.114)$     & $(0.114)$     & $(0.139)$     & $(0.133)$     & $(0.134)$     \\
+Wave x Perception inequality (WE)                        & $-0.004$      &               &               &               &               &               \\
+                                                         & $(0.006)$     &               &               &               &               &               \\
+Wave x Merit: effort (WE)                                &               & $-0.008$      &               &               &               &               \\
+                                                         &               & $(0.006)$     &               &               &               &               \\
+Wave x Merit: talent (WE)                                &               &               & $-0.006$      &               &               &               \\
+                                                         &               &               & $(0.006)$     &               &               &               \\
+Wave x Perception inequality (BE)                        &               &               &               & $0.011$       &               &               \\
+                                                         &               &               &               & $(0.007)$     &               &               \\
+Wave x Merit: effort (BE)                                &               &               &               &               & $-0.021^{*}$  &               \\
+                                                         &               &               &               &               & $(0.008)$     &               \\
+Wave x Merit: talent (BE)                                &               &               &               &               &               & $-0.020^{*}$  \\
+                                                         &               &               &               &               &               & $(0.008)$     \\
+\hline
+Controls                                                 & Yes           & Yes           & Yes           & Yes           & Yes           & Yes           \\
+BIC                                                      & $20792.203$   & $20767.371$   & $20754.324$   & $20782.059$   & $20749.596$   & $20754.417$   \\
+Numb. obs.                                               & $8643$        & $8643$        & $8643$        & $8643$        & $8643$        & $8643$        \\
+Num. groups: individuals                                 & $1687$        & $1687$        & $1687$        & $1687$        & $1687$        & $1687$        \\
+Var: individuals (Intercept)                             & $0.176$       & $0.166$       & $0.168$       & $0.387$       & $0.254$       & $0.010$       \\
+Var: individuals, perception inequality cwc              & $0.000$       & $$            & $$            & $$            & $$            & $$            \\
+Var: individuals, wave                                   & $0.010$       & $0.009$       & $0.009$       & $0.009$       & $0.009$       & $0.009$       \\
+Cov: individuals (Intercept), perception inequality cwc  & $-0.007$      & $$            & $$            & $$            & $$            & $$            \\
+Cov: individuals (Intercept), wave                       & $-0.021$      & $-0.018$      & $-0.018$      & $-0.049$      & $-0.003$      & $0.006$       \\
+Cov: individuals, perception inequality cwc, wave        & $0.002$       & $$            & $$            & $$            & $$            & $$            \\
+Var: Residuals                                           & $0.495$       & $0.480$       & $0.479$       & $0.494$       & $0.494$       & $0.494$       \\
+Var: individuals, merit effort cwc                       & $$            & $0.019$       & $$            & $$            & $$            & $$            \\
+Cov: individuals (Intercept), merit effort cwc           & $$            & $0.023$       & $$            & $$            & $$            & $$            \\
+Cov: individuals, merit effort cwc, wave                 & $$            & $-0.005$      & $$            & $$            & $$            & $$            \\
+Var: individuals, merit talent cwc                       & $$            & $$            & $0.019$       & $$            & $$            & $$            \\
+Cov: individuals (Intercept), merit talent cwc           & $$            & $$            & $0.034$       & $$            & $$            & $$            \\
+Cov: individuals, merit talent cwc, wave                 & $$            & $$            & $-0.006$      & $$            & $$            & $$            \\
+Var: individuals, perception inequality mean             & $$            & $$            & $$            & $0.008$       & $$            & $$            \\
+Cov: individuals (Intercept), perception inequality mean & $$            & $$            & $$            & $-0.046$      & $$            & $$            \\
+Cov: individuals, perception inequality mean, wave       & $$            & $$            & $$            & $0.009$       & $$            & $$            \\
+Var: individuals, merit effort mean                      & $$            & $$            & $$            & $$            & $0.053$       & $$            \\
+Cov: individuals (Intercept), merit effort mean          & $$            & $$            & $$            & $$            & $-0.089$      & $$            \\
+Cov: individuals, merit effort mean, wave                & $$            & $$            & $$            & $$            & $-0.006$      & $$            \\
+Var: individuals, merit talent mean                      & $$            & $$            & $$            & $$            & $$            & $0.025$       \\
+Cov: individuals (Intercept), merit talent mean          & $$            & $$            & $$            & $$            & $$            & $-0.007$      \\
+Cov: individuals, merit talent mean, wave                & $$            & $$            & $$            & $$            & $$            & $-0.009$      \\
+\hline
+\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$. CWC = centered within group.}}
+\end{tabular}
 }
-
-\end{table}%
+\label{table:coefficients}
+\end{center}
+\end{table}
 
@@
 
-\section{Discussion}\label{discussion}
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
 
@@
 inequality could fuel a critical view of such policies
-(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
-\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). This relationship
-reflected a moral economy where perceptions of systemic unfairness
-threatened the legitimacy of existing distributions, prompting calls for
-greater equity. Interestingly, the negative within-effects observed over
-time implied that as perceptions of inequality became more salient, some
-individuals experienced a decline in support for market justice,
-possibly due to increasing distrust in market mechanisms or
-disillusionment with the fairness of the system. This indicated that
-while perception heightened awareness, prolonged exposure or social
-mobilization could also foster skepticism about whether redistribution
-effectively addressed inequality or simply perpetuated systemic flaws.
+(\protect\hyperlink{ref-Castillo2012a_justice}{Castillo, 2012};
+\protect\hyperlink{ref-mijs_paradox_2021}{Mijs, 2021}). This
+relationship reflected a moral economy where perceptions of systemic
+unfairness threatened the legitimacy of existing distributions,
+prompting calls for greater equity. Interestingly, the negative
+within-effects observed over time implied that as perceptions of
+inequality became more salient, some individuals experienced a decline
+in support for market justice, possibly due to increasing distrust in
+market mechanisms or disillusionment with the fairness of the system.
+This indicated that while perception heightened awareness, prolonged
+exposure or social mobilization could also foster skepticism about
+whether redistribution effectively addressed inequality or simply
+perpetuated systemic flaws.
 
@@
 meritocratic narratives served as moral justifications that reinforced
-societal stratification (\citeproc{ref-Castillo2012a_justice}{Castillo,
-2012}; \citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These
+societal stratification
+(\protect\hyperlink{ref-Castillo2012a_justice}{Castillo, 2012};
+\protect\hyperlink{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These
 beliefs functioned symbolically to legitimize structural disparities,
@@
 market logic heavily influenced social attitudes
-(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).
+(\protect\hyperlink{ref-canalesceron_sujeto_2021}{Canales Cerón et al.,
+2021}).
 
@@
 encouraging more critical evaluations of institutional legitimacy
-(\citeproc{ref-tilly_contentious_2015}{Tilly \& Tarrow, 2015}). The
-findings underscored the importance of socio-political context in
+(\protect\hyperlink{ref-tilly_contentious_2015}{Tilly \& Tarrow, 2015}).
+The findings underscored the importance of socio-political context in
 shaping and reshaping attitudes toward inequality and justice,
@@
 
-\section{Conclusions}\label{conclusions}
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
 
@@
 
-\section{References}\label{references}
-
-\phantomsection\label{refs}
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\hypertarget{refs}{}
 \begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-akyelken_urban_2020}
+\leavevmode\vadjust pre{\hypertarget{ref-akyelken_urban_2020}{}}%
 Akyelken, N. (2020). Urban conceptions of economic inequalities.
@@
 
-\bibitem[\citeproctext]{ref-auspurg_why_2017}
+\leavevmode\vadjust pre{\hypertarget{ref-auspurg_why_2017}{}}%
 Auspurg, K., Hinz, T., \& Sauer, C. (2017). Why {Should Women Get Less}?
@@
 
-\bibitem[\citeproctext]{ref-bates_fitting_2015}
+\leavevmode\vadjust pre{\hypertarget{ref-bates_fitting_2015}{}}%
 Bates, D., Mächler, M., Bolker, B., \& Walker, S. (2015). Fitting linear
@@
 
-\bibitem[\citeproctext]{ref-batruch_belief_2023}
+\leavevmode\vadjust pre{\hypertarget{ref-batruch_belief_2023}{}}%
 Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F.
@@
 
-\bibitem[\citeproctext]{ref-boccardo_30_2020}
+\leavevmode\vadjust pre{\hypertarget{ref-boccardo_30_2020}{}}%
 Boccardo, G. (2020). \emph{30 a{ñ}os de privatizaciones en {Chile}: Lo
@@
 
-\bibitem[\citeproctext]{ref-busemeyer_positive_2021}
+\leavevmode\vadjust pre{\hypertarget{ref-busemeyer_positive_2021}{}}%
 Busemeyer, M. R., Abrassart, A., \& Nezi, R. (2021). Beyond {Positive}
@@
 
-\bibitem[\citeproctext]{ref-canalesceron_sujeto_2021}
+\leavevmode\vadjust pre{\hypertarget{ref-campbell_policy_2012}{}}%
+Campbell, A. L. (2012). Policy {Makes Mass Politics}. \emph{Annual
+Review of Political Science}, \emph{15}(Volume 15, 2012), 333--351.
+\url{https://doi.org/10.1146/annurev-polisci-012610-135202}
+
+\leavevmode\vadjust pre{\hypertarget{ref-canalesceron_sujeto_2021}{}}%
 Canales Cerón, M., Orellana Calderón, V. S., \& Guajardo Mañán, F.
@@
 
-\bibitem[\citeproctext]{ref-castillo_cual_2009}
+\leavevmode\vadjust pre{\hypertarget{ref-castillo_cual_2009}{}}%
 Castillo, J. C. (2009). {{\textquestiondown}Cu{á}l es la brecha salarial
@@
 
-\bibitem[\citeproctext]{ref-Castillo2011}
+\leavevmode\vadjust pre{\hypertarget{ref-Castillo2011}{}}%
 Castillo, J. C. (2011). Legitimacy of {Inequality} in a {Highly Unequal
@@
 
-\bibitem[\citeproctext]{ref-Castillo2012a_justice}
+\leavevmode\vadjust pre{\hypertarget{ref-Castillo2012a_justice}{}}%
 Castillo, J. C. (2012). Is {Inequality Becoming Just}? {Changes} in
@@
 
-\bibitem[\citeproctext]{ref-castillo_perception_2022}
+\leavevmode\vadjust pre{\hypertarget{ref-castillo_perception_2022}{}}%
 Castillo, J. C., García-Castro, J.-D., \& Venegas, M. (2022). Perception
@@
 
-\bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
+\leavevmode\vadjust pre{\hypertarget{ref-castillo_multidimensional_2023}{}}%
 Castillo, J. C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F.
@@
 
-\bibitem[\citeproctext]{ref-castillo_percepcion_2019}
+\leavevmode\vadjust pre{\hypertarget{ref-castillo_percepcion_2019}{}}%
 Castillo, J. C., Miranda, D., \& Carrasco, D. (2012). Percepci{ó}n de
@@
 
-\bibitem[\citeproctext]{ref-castillo_socialization_2024}
+\leavevmode\vadjust pre{\hypertarget{ref-castillo_socialization_2024}{}}%
 Castillo, J. C., Salgado, M., Carrasco, K., \& Laffert, A. (2024). The
@@
 
-\bibitem[\citeproctext]{ref-chancel_world_2022}
+\leavevmode\vadjust pre{\hypertarget{ref-chancel_world_2022}{}}%
 Chancel, L., Piketty, T., Saez, E., \& Zucman, G. (2022). World
@@
 
-\bibitem[\citeproctext]{ref-davis_principles_2001}
+\leavevmode\vadjust pre{\hypertarget{ref-davis_principles_2001}{}}%
 Davis, K., \& Moore, W. E. (2001). Some {Principles} of
@@
 
-\bibitem[\citeproctext]{ref-easterbrook_social_2021}
+\leavevmode\vadjust pre{\hypertarget{ref-easterbrook_social_2021}{}}%
 Easterbrook, M. J. (2021). \emph{The social psychology of economic
@@
 
-\bibitem[\citeproctext]{ref-elsoc_estudio_2022}
+\leavevmode\vadjust pre{\hypertarget{ref-elsoc_estudio_2022}{}}%
 ELSOC, S. T. (2022). Estudio {Longitudinal Social} de {Chile}. Harvard
@@
 
-\bibitem[\citeproctext]{ref-engelhardt_what_2018}
+\leavevmode\vadjust pre{\hypertarget{ref-engelhardt_what_2018}{}}%
 Engelhardt, C., \& Wagener, A. (2018). What do {Germans} think and know
@@
 
-\bibitem[\citeproctext]{ref-flores_top_2020}
+\leavevmode\vadjust pre{\hypertarget{ref-flores_top_2020}{}}%
 Flores, I., Sanhueza, C., Atria, J., \& Mayer, R. (2020). Top {Incomes}
@@
 
-\bibitem[\citeproctext]{ref-garcia-castro_perceiving_2020}
+\leavevmode\vadjust pre{\hypertarget{ref-garcia-castro_perceiving_2020}{}}%
 García-Castro, J. D., Rodríguez-Bailón, R., \& Willis, G. B. (2020).
@@
 
-\bibitem[\citeproctext]{ref-garcia-sanchez_creencias_2022}
+\leavevmode\vadjust pre{\hypertarget{ref-garcia-sanchez_creencias_2022}{}}%
 García-Sánchez, E., \& De Carvalho, S. (2022). Las creencias que
@@
 
-\bibitem[\citeproctext]{ref-garcia-sanchez_attitudes_2020}
+\leavevmode\vadjust pre{\hypertarget{ref-garcia-sanchez_attitudes_2020}{}}%
 García-Sánchez, E., Osborne, D., Willis, G. B., \& Rodríguez-Bailón, R.
@@
 
-\bibitem[\citeproctext]{ref-garcia-sanchez_vicious_2019}
+\leavevmode\vadjust pre{\hypertarget{ref-garcia-sanchez_vicious_2019}{}}%
 García-Sánchez, E., Van Der Toorn, J., Rodríguez-Bailón, R., \& Willis,
@@
 
-\bibitem[\citeproctext]{ref-garcia-sanchez_perceptions_2018}
+\leavevmode\vadjust pre{\hypertarget{ref-garcia-sanchez_perceptions_2018}{}}%
 García-Sánchez, E., Willis, G. B., Rodríguez-Bailón, R., Palacio Sañudo,
@@
 
-\bibitem[\citeproctext]{ref-gijsberts_thelegitimation_1999}
+\leavevmode\vadjust pre{\hypertarget{ref-gijsberts_thelegitimation_1999}{}}%
 Gijsberts, M. (1999). \emph{{Thelegitimation of inequality in state-
@@
 
-\bibitem[\citeproctext]{ref-gimpelson_misperceiving_2018}
+\leavevmode\vadjust pre{\hypertarget{ref-gimpelson_misperceiving_2018}{}}%
 Gimpelson, V., \& Treisman, D. (2018). Misperceiving inequality.
@@
 
-\bibitem[\citeproctext]{ref-hadler_why_2005}
+\leavevmode\vadjust pre{\hypertarget{ref-hadler_why_2005}{}}%
 Hadler, M. (2005). Why {Do People Accept Different Income Ratios}?: {A
@@
 
-\bibitem[\citeproctext]{ref-hoxMultilevelAnalysisTechniques2017a}
+\leavevmode\vadjust pre{\hypertarget{ref-hoxMultilevelAnalysisTechniques2017a}{}}%
 Hox, J. J., Moerbeek, M., \& Van de Schoot, R. (2017). \emph{Multilevel
@@
 
-\bibitem[\citeproctext]{ref-hoyt_mindsets_2023}
+\leavevmode\vadjust pre{\hypertarget{ref-hoyt_mindsets_2023}{}}%
 Hoyt, C. L., Burnette, J. L., Billingsley, J., Becker, W., \& Babij, A.
@@
 
-\bibitem[\citeproctext]{ref-immergut_it_2020}
+\leavevmode\vadjust pre{\hypertarget{ref-immergut_it_2020}{}}%
 Immergut, E. M., \& Schneider, S. M. (2020). Is it unfair for the
@@
 
-\bibitem[\citeproctext]{ref-janmaat_subjective_2013}
+\leavevmode\vadjust pre{\hypertarget{ref-janmaat_subjective_2013}{}}%
 Janmaat, J. G. (2013). Subjective inequality: {A} review of
@@
 
-\bibitem[\citeproctext]{ref-kluegel_social_1995a}
+\leavevmode\vadjust pre{\hypertarget{ref-kluegel_social_1995a}{}}%
 Kluegel, J. R., Mason, D. S., \& Wegener, B. (Eds.). (1995).
@@
 
-\bibitem[\citeproctext]{ref-kluegel_legitimation_1999}
+\leavevmode\vadjust pre{\hypertarget{ref-kluegel_legitimation_1999}{}}%
 Kluegel, J. R., Mason, D. S., \& Wegener, B. (1999). The {Legitimation}
@@
 
-\bibitem[\citeproctext]{ref-kluegel_beliefs_1981}
+\leavevmode\vadjust pre{\hypertarget{ref-kluegel_beliefs_1981}{}}%
 Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}.
@@
 
-\bibitem[\citeproctext]{ref-knell_perceptions_2020}
+\leavevmode\vadjust pre{\hypertarget{ref-knell_perceptions_2020}{}}%
 Knell, M., \& Stix, H. (2020). Perceptions of inequality. \emph{European
@@
 
-\bibitem[\citeproctext]{ref-koos_moral_2019}
+\leavevmode\vadjust pre{\hypertarget{ref-koos_moral_2019}{}}%
 Koos, S., \& Sachweh, P. (2019). The moral economies of market
@@
 
-\bibitem[\citeproctext]{ref-kuhn_eye_2011}
+\leavevmode\vadjust pre{\hypertarget{ref-kuhn_eye_2011}{}}%
 Kuhn, A. (2011). In the eye of the beholder: {Subjective} inequality
@@
 
-\bibitem[\citeproctext]{ref-lane_market_1986}
+\leavevmode\vadjust pre{\hypertarget{ref-lane_market_1986}{}}%
 Lane, R. E. (1986). Market {Justice}, {Political Justice}.
@@
 
-\bibitem[\citeproctext]{ref-lee_fairness_2023}
+\leavevmode\vadjust pre{\hypertarget{ref-lee_fairness_2023}{}}%
 Lee, J.-S., \& Stacey, M. (2023). Fairness perceptions of educational
@@
 
-\bibitem[\citeproctext]{ref-lindh_public_2015}
+\leavevmode\vadjust pre{\hypertarget{ref-lindh_public_2015}{}}%
 Lindh, A. (2015). Public {Opinion} against {Markets}? {Attitudes}
@@
 
-\bibitem[\citeproctext]{ref-mijs_stratified_2016}
+\leavevmode\vadjust pre{\hypertarget{ref-mijs_stratified_2016}{}}%
 Mijs, J. (2016). Stratified {Failure}: {Educational Stratification} and
@@
 
-\bibitem[\citeproctext]{ref-mijs_paradox_2021}
+\leavevmode\vadjust pre{\hypertarget{ref-mijs_paradox_2021}{}}%
 Mijs, J. (2021). The paradox of inequality: Income inequality and belief
@@
 
-\bibitem[\citeproctext]{ref-pedersen_attitudes_2019}
+\leavevmode\vadjust pre{\hypertarget{ref-pedersen_attitudes_2019}{}}%
 Pedersen, R. T., \& Mutz, D. C. (2019). Attitudes {Toward Economic
@@
 
-\bibitem[\citeproctext]{ref-reynolds_perceptions_2014}
+\leavevmode\vadjust pre{\hypertarget{ref-pierson_when_1993}{}}%
+Pierson, P. (1993). When {Effect Becomes Cause}: {Policy Feedback} and
+{Political Change}. \emph{World Politics}, \emph{45}(4), 595--628.
+\url{https://doi.org/10.2307/2950710}
+
+\leavevmode\vadjust pre{\hypertarget{ref-reynolds_perceptions_2014}{}}%
 Reynolds, J., \& Xian, H. (2014). Perceptions of meritocracy in the land
@@
 
-\bibitem[\citeproctext]{ref-sandel_tyranny_2020}
+\leavevmode\vadjust pre{\hypertarget{ref-sandel_tyranny_2020}{}}%
 Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the
@@
 
-\bibitem[\citeproctext]{ref-schneider_poverty_2015}
+\leavevmode\vadjust pre{\hypertarget{ref-schneider_poverty_2015}{}}%
 Schneider, S. M., \& Castillo, J. C. (2015). Poverty {Attributions} and
@@
 
-\bibitem[\citeproctext]{ref-schroder_income_2017}
+\leavevmode\vadjust pre{\hypertarget{ref-schroder_income_2017}{}}%
 Schröder, M. (2017). Is {Income Inequality Related} to {Tolerance} for
@@
 
-\bibitem[\citeproctext]{ref-singer_applied_2009}
+\leavevmode\vadjust pre{\hypertarget{ref-singer_applied_2009}{}}%
 Singer, J. D., \& Willett, J. B. (2009). \emph{Applied longitudinal data
@@
 
-\bibitem[\citeproctext]{ref-svallfors_political_2007}
+\leavevmode\vadjust pre{\hypertarget{ref-svallfors_political_2007}{}}%
 Svallfors, S. (Ed.). (2007). \emph{The {Political Sociology} of the
@@
 
-\bibitem[\citeproctext]{ref-tejero-peregrina_perceived_2025}
+\leavevmode\vadjust pre{\hypertarget{ref-tejero-peregrina_perceived_2025}{}}%
 Tejero-Peregrina, L., Willis, G., Sánchez-Rodríguez, Á., \&
@@
 
-\bibitem[\citeproctext]{ref-tilly_contentious_2015}
+\leavevmode\vadjust pre{\hypertarget{ref-tilly_contentious_2015}{}}%
 Tilly, C., \& Tarrow, S. (2015). \emph{Contentious {Politics}} (Second
@@
 
-\bibitem[\citeproctext]{ref-trump_income_2018}
+\leavevmode\vadjust pre{\hypertarget{ref-trump_income_2018}{}}%
 Trump, K.-S. (2018). Income {Inequality Influences Perceptions} of
@@
 
-\bibitem[\citeproctext]{ref-vondemknesebeck_are_2016}
+\leavevmode\vadjust pre{\hypertarget{ref-vondemknesebeck_are_2016}{}}%
 Von Dem Knesebeck, O., Vonneilich, N., \& Kim, T. J. (2016). Are health
@@
 
-\bibitem[\citeproctext]{ref-willis_legitimacy_2015}
+\leavevmode\vadjust pre{\hypertarget{ref-willis_legitimacy_2015}{}}%
 Willis, G. B., Rodríguez-Bailón, R., López-Rodríguez, L., \&
@@
 
-\bibitem[\citeproctext]{ref-wilson_role_2003}
+\leavevmode\vadjust pre{\hypertarget{ref-wilson_role_2003}{}}%
 Wilson, C. (2003). The {Role} of a {Merit Principle} in {Distributive
@@
 
-\bibitem[\citeproctext]{ref-young_rise_1962}
+\leavevmode\vadjust pre{\hypertarget{ref-young_rise_1962}{}}%
 Young, M. (1962). \emph{The rise of the meritocracy}. Baltimore: Penguin
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-akyelken_urban_2020}{}}%
+Akyelken, N. (2020). Urban conceptions of economic inequalities.
+\emph{Regional Studies}, \emph{54}(6), 863--872.
+\url{https://doi.org/10.1080/00343404.2020.1732902}
+
+\leavevmode\vadjust pre{\hypertarget{ref-auspurg_why_2017}{}}%
+Auspurg, K., Hinz, T., \& Sauer, C. (2017). Why {Should Women Get Less}?
+{Evidence} on the {Gender Pay Gap} from {Multifactorial Survey
+Experiments}. \emph{American Sociological Review}, \emph{82}(1),
+179--210. \url{https://doi.org/10.1177/0003122416683393}
+
+\leavevmode\vadjust pre{\hypertarget{ref-bates_fitting_2015}{}}%
+Bates, D., Mächler, M., Bolker, B., \& Walker, S. (2015). Fitting linear
+mixed-effects models using {lme4}. \emph{Journal of Statistical
+Software}, \emph{67}(1), 1--48.
+\url{https://doi.org/10.18637/jss.v067.i01}
+
+\leavevmode\vadjust pre{\hypertarget{ref-batruch_belief_2023}{}}%
+Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F.
+(2023). Belief in {School Meritocracy} and the {Legitimization} of
+{Social} and {Income Inequality}. \emph{Social Psychological and
+Personality Science}, \emph{14}(5), 621--635.
+\url{https://doi.org/10.1177/19485506221111017}
+
+\leavevmode\vadjust pre{\hypertarget{ref-boccardo_30_2020}{}}%
+Boccardo, G. (2020). \emph{30 a{ñ}os de privatizaciones en {Chile}: Lo
+que la pandemia revel{ó}} (Nodo XXI). Santiago.
+
+\leavevmode\vadjust pre{\hypertarget{ref-busemeyer_positive_2021}{}}%
+Busemeyer, M. R., Abrassart, A., \& Nezi, R. (2021). Beyond {Positive}
+and {Negative}: {New Perspectives} on {Feedback Effects} in {Public
+Opinion} on the {Welfare State}. \emph{British Journal of Political
+Science}, \emph{51}(1), 137--162.
+\url{https://doi.org/10.1017/S0007123418000534}
+
+\leavevmode\vadjust pre{\hypertarget{ref-campbell_policy_2012}{}}%
+Campbell, A. L. (2012). Policy {Makes Mass Politics}. \emph{Annual
+Review of Political Science}, \emph{15}(Volume 15, 2012), 333--351.
+\url{https://doi.org/10.1146/annurev-polisci-012610-135202}
+
+\leavevmode\vadjust pre{\hypertarget{ref-canalesceron_sujeto_2021}{}}%
+Canales Cerón, M., Orellana Calderón, V. S., \& Guajardo Mañán, F.
+(2021). Sujeto y cotidiano en la era neoliberal: El caso de la
+educaci{ó}n chilena. \emph{Revista Mexicana de Ciencias Pol{í}ticas y
+Sociales}, \emph{67}(244).
+\url{https://doi.org/10.22201/fcpys.2448492xe.2022.244.70386}
+
+\leavevmode\vadjust pre{\hypertarget{ref-castillo_cual_2009}{}}%
+Castillo, J. C. (2009). {{\textquestiondown}Cu{á}l es la brecha salarial
+justa? Opini{ó}n p{ú}blica y legitimaci{ó}n de la desigualdad en Chile}.
+\emph{Estudios P{ú}blicos}, (113).
+
+\leavevmode\vadjust pre{\hypertarget{ref-Castillo2011}{}}%
+Castillo, J. C. (2011). Legitimacy of {Inequality} in a {Highly Unequal
+Context}: {Evidence} from the {Chilean Case}. \emph{Social Justice
+Research}, \emph{24}(4), 314--340.
+\url{https://doi.org/10.1007/s11211-011-0144-5}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Castillo2012a_justice}{}}%
+Castillo, J. C. (2012). Is {Inequality Becoming Just}? {Changes} in
+{Public Opinion} about {Economic Distribution} in {Chile}.
+\emph{Bulletin of Latin American Research}, \emph{31}(1), 1--18.
+\url{https://doi.org/10.1111/j.1470-9856.2011.00605.x}
+
+\leavevmode\vadjust pre{\hypertarget{ref-castillo_perception_2022}{}}%
+Castillo, J. C., García-Castro, J.-D., \& Venegas, M. (2022). Perception
+of economic inequality: Concepts, associated factors and prospects of a
+burgeoning research agenda. \emph{International Journal of Social
+Psychology}, \emph{37}(1), 180--207.
+\url{https://doi.org/10.1080/02134748.2021.2009275}
+
+\leavevmode\vadjust pre{\hypertarget{ref-castillo_multidimensional_2023}{}}%
+Castillo, J. C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F.
+(2023). A {Multidimensional Approach} for {Measuring Meritocratic
+Beliefs}: {Advantages}, {Limitations} and {Alternatives} to the {ISSP
+Social Inequality Survey}. \emph{International Journal of Sociology},
+1--25. \url{https://doi.org/10.1080/00207659.2023.2274712}
+
+\leavevmode\vadjust pre{\hypertarget{ref-castillo_percepcion_2019}{}}%
+Castillo, J. C., Miranda, D., \& Carrasco, D. (2012). Percepci{ó}n de
+{Desigualdad Econ{ó}mica} en {Chile}: {Medici{ó}n}, {Diferencias} y
+{Determinantes}. \emph{Psykhe (Santiago)}, \emph{21}(1), 99--114.
+\url{https://doi.org/10.4067/S0718-22282012000100007}
+
+\leavevmode\vadjust pre{\hypertarget{ref-castillo_socialization_2024}{}}%
+Castillo, J. C., Salgado, M., Carrasco, K., \& Laffert, A. (2024). The
+{Socialization} of {Meritocracy} and {Market Justice Preferences} at
+{School}. \emph{Societies}, \emph{14}(11), 214.
+\url{https://doi.org/10.3390/soc14110214}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chancel_world_2022}{}}%
+Chancel, L., Piketty, T., Saez, E., \& Zucman, G. (2022). World
+inequality report 2022.
+https://bibliotecadigital.ccb.org.co/handle/11520/27510.
+
+\leavevmode\vadjust pre{\hypertarget{ref-davis_principles_2001}{}}%
+Davis, K., \& Moore, W. E. (2001). Some {Principles} of
+{Stratification}. In \emph{Social {Stratification}, {Class}, {Race}, and
+{Gender} in {Sociological Perspective}, {Second Edition}} (2nd ed.).
+Routledge.
+
+\leavevmode\vadjust pre{\hypertarget{ref-easterbrook_social_2021}{}}%
+Easterbrook, M. J. (2021). \emph{The social psychology of economic
+inequality} (43rd ed., Vol. 2021). UNU-WIDER.
+\url{https://doi.org/10.35188/UNU-WIDER/2021/981-5}
+
+\leavevmode\vadjust pre{\hypertarget{ref-elsoc_estudio_2022}{}}%
+ELSOC, S. T. (2022). Estudio {Longitudinal Social} de {Chile}. Harvard
+Dataverse. \url{https://doi.org/10.7910/dvn/0kirbj}
+
+\leavevmode\vadjust pre{\hypertarget{ref-engelhardt_what_2018}{}}%
+Engelhardt, C., \& Wagener, A. (2018). What do {Germans} think and know
+about income inequality? {A} survey experiment. \emph{Socio-Economic
+Review}, \emph{16}(4), 743--767.
+\url{https://doi.org/10.1093/ser/mwx036}
+
+\leavevmode\vadjust pre{\hypertarget{ref-flores_top_2020}{}}%
+Flores, I., Sanhueza, C., Atria, J., \& Mayer, R. (2020). Top {Incomes}
+in {Chile}: {A Historical Perspective} on {Income Inequality},
+1964--2017. \emph{Review of Income and Wealth}, \emph{66}(4), 850--874.
+\url{https://doi.org/10.1111/roiw.12441}
+
+\leavevmode\vadjust pre{\hypertarget{ref-garcia-castro_perceiving_2020}{}}%
+García-Castro, J. D., Rodríguez-Bailón, R., \& Willis, G. B. (2020).
+Perceiving economic inequality in everyday life decreases tolerance to
+inequality. \emph{Journal of Experimental Social Psychology}, \emph{90},
+104019. \url{https://doi.org/10.1016/j.jesp.2020.104019}
+
+\leavevmode\vadjust pre{\hypertarget{ref-garcia-sanchez_creencias_2022}{}}%
+García-Sánchez, E., \& De Carvalho, S. (2022). Las creencias que
+justifican la desigualdad moderan la relaci{ó}n entre el estatus
+socioecon{ó}mico y el apoyo a la redistribuci{ó}n. \emph{Revista
+Internacional de Sociolog{í}a}, \emph{80}(3), e210.
+\url{https://doi.org/10.3989/ris.2022.80.3.21.29}
+
+\leavevmode\vadjust pre{\hypertarget{ref-garcia-sanchez_attitudes_2020}{}}%
+García-Sánchez, E., Osborne, D., Willis, G. B., \& Rodríguez-Bailón, R.
+(2020). Attitudes towards redistribution and the interplay between
+perceptions and beliefs about inequality. \emph{British Journal of
+Social Psychology}, \emph{59}(1), 111--136.
+\url{https://doi.org/10.1111/bjso.12326}
+
+\leavevmode\vadjust pre{\hypertarget{ref-garcia-sanchez_vicious_2019}{}}%
+García-Sánchez, E., Van Der Toorn, J., Rodríguez-Bailón, R., \& Willis,
+G. B. (2019). The {Vicious Cycle} of {Economic Inequality}: {The Role}
+of {Ideology} in {Shaping} the {Relationship Between} {``{What Is}''}
+and {``{What Ought} to {Be}''} in 41 {Countries}. \emph{Social
+Psychological and Personality Science}, \emph{10}(8), 991--1001.
+\url{https://doi.org/10.1177/1948550618811500}
+
+\leavevmode\vadjust pre{\hypertarget{ref-garcia-sanchez_perceptions_2018}{}}%
+García-Sánchez, E., Willis, G. B., Rodríguez-Bailón, R., Palacio Sañudo,
+J., David Polo, J., \& Rentería Pérez, E. (2018). Perceptions of
+{Economic Inequality} and {Support} for {Redistribution}: {The} role of
+{Existential} and {Utopian Standards}. \emph{Social Justice Research},
+\emph{31}(4), 335--354. \url{https://doi.org/10.1007/s11211-018-0317-6}
+
+\leavevmode\vadjust pre{\hypertarget{ref-gijsberts_thelegitimation_1999}{}}%
+Gijsberts, M. (1999). \emph{{Thelegitimation of inequality in state-
+socialist and market societies, 1987 - 1996}}. Amsterdam: Thela Thesis.
+
+\leavevmode\vadjust pre{\hypertarget{ref-gimpelson_misperceiving_2018}{}}%
+Gimpelson, V., \& Treisman, D. (2018). Misperceiving inequality.
+\emph{Economics \& Politics}, \emph{30}(1), 27--54.
+\url{https://doi.org/10.1111/ecpo.12103}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hadler_why_2005}{}}%
+Hadler, M. (2005). Why {Do People Accept Different Income Ratios}?: {A
+Multi-level Comparison} of {Thirty Countries}. \emph{Acta Sociologica},
+\emph{48}(2), 131--154. \url{https://doi.org/10.1177/0001699305053768}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hoxMultilevelAnalysisTechniques2017a}{}}%
+Hox, J. J., Moerbeek, M., \& Van de Schoot, R. (2017). \emph{Multilevel
+{Analysis}: {Techniques} and {Applications}}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hoyt_mindsets_2023}{}}%
+Hoyt, C. L., Burnette, J. L., Billingsley, J., Becker, W., \& Babij, A.
+D. (2023). Mindsets of poverty: {Implications} for redistributive policy
+support. \emph{Analyses of Social Issues and Public Policy},
+\emph{23}(3), 668--693. \url{https://doi.org/10.1111/asap.12367}
+
+\leavevmode\vadjust pre{\hypertarget{ref-immergut_it_2020}{}}%
+Immergut, E. M., \& Schneider, S. M. (2020). Is it unfair for the
+affluent to be able to purchase {``better''} healthcare? {Existential}
+standards and institutional norms in healthcare attitudes across 28
+countries. \emph{Social Science \& Medicine}, \emph{267}, 113146.
+\url{https://doi.org/10.1016/j.socscimed.2020.113146}
+
+\leavevmode\vadjust pre{\hypertarget{ref-janmaat_subjective_2013}{}}%
+Janmaat, J. G. (2013). Subjective inequality: {A} review of
+international comparative studies on people's views about inequality.
+\emph{Archives Europeennes de Sociologie}, \emph{54}(3), 357--389.
+\url{https://doi.org/10.1017/S0003975613000209}
+
+\leavevmode\vadjust pre{\hypertarget{ref-kluegel_social_1995a}{}}%
+Kluegel, J. R., Mason, D. S., \& Wegener, B. (Eds.). (1995).
+\emph{Social {Justice} and {Political Change}: {Public Opinion} in
+{Capitalist} and {Post-Communist States}} (1st ed.). Routledge.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kluegel_legitimation_1999}{}}%
+Kluegel, J. R., Mason, D. S., \& Wegener, B. (1999). The {Legitimation}
+of {Capitalism} in the {Postcommunist Transition}: {Public Opinion}
+about {Market Justice}, 1991-1996. \emph{European Sociological Review},
+\emph{15}(3), 251--283. Retrieved from
+\url{https://www.jstor.org/stable/522731}
+
+\leavevmode\vadjust pre{\hypertarget{ref-kluegel_beliefs_1981}{}}%
+Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}.
+\emph{Annual Review of Sociology}, 29--56.
+
+\leavevmode\vadjust pre{\hypertarget{ref-knell_perceptions_2020}{}}%
+Knell, M., \& Stix, H. (2020). Perceptions of inequality. \emph{European
+Journal of Political Economy}, \emph{65}, 101927.
+\url{https://doi.org/10.1016/j.ejpoleco.2020.101927}
+
+\leavevmode\vadjust pre{\hypertarget{ref-koos_moral_2019}{}}%
+Koos, S., \& Sachweh, P. (2019). The moral economies of market
+societies: Popular attitudes towards market competition, redistribution
+and reciprocity in comparative perspective. \emph{Socio-Economic
+Review}, \emph{17}(4), 793--821.
+\url{https://doi.org/10.1093/ser/mwx045}
+
+\leavevmode\vadjust pre{\hypertarget{ref-kuhn_eye_2011}{}}%
+Kuhn, A. (2011). In the eye of the beholder: {Subjective} inequality
+measures and individuals' assessment of market justice. \emph{European
+Journal of Political Economy}, \emph{27}(4), 625--641.
+\url{https://doi.org/10.1016/j.ejpoleco.2011.06.002}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lane_market_1986}{}}%
+Lane, R. E. (1986). Market {Justice}, {Political Justice}.
+\emph{American Political Science Review}, \emph{80}(2), 383--402.
+\url{https://doi.org/10.2307/1958264}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lee_fairness_2023}{}}%
+Lee, J.-S., \& Stacey, M. (2023). Fairness perceptions of educational
+inequality: The effects of self-interest and neoliberal orientations.
+\emph{The Australian Educational Researcher}.
+\url{https://doi.org/10.1007/s13384-023-00636-6}
+
+\leavevmode\vadjust pre{\hypertarget{ref-lindh_public_2015}{}}%
+Lindh, A. (2015). Public {Opinion} against {Markets}? {Attitudes}
+towards {Market Distribution} of {Social Services} -- {A Comparison} of
+17 {Countries}. \emph{Social Policy \& Administration}, \emph{49}(7),
+887--910. \url{https://doi.org/10.1111/spol.12105}
+
+\leavevmode\vadjust pre{\hypertarget{ref-mijs_stratified_2016}{}}%
+Mijs, J. (2016). Stratified {Failure}: {Educational Stratification} and
+{Students}' {Attributions} of {Their Mathematics Performance} in 24
+{Countries}. \emph{Sociology of Education}, \emph{89}(2), 137--153.
+\url{https://doi.org/10.1177/0038040716636434}
+
+\leavevmode\vadjust pre{\hypertarget{ref-mijs_paradox_2021}{}}%
+Mijs, J. (2021). The paradox of inequality: Income inequality and belief
+in meritocracy go hand in hand. \emph{Socio-Economic Review},
+\emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pedersen_attitudes_2019}{}}%
+Pedersen, R. T., \& Mutz, D. C. (2019). Attitudes {Toward Economic
+Inequality}: {The Illusory Agreement}. \emph{Political Science Research
+and Methods}, \emph{7}(04), 835--851.
+\url{https://doi.org/10.1017/psrm.2018.18}
+
+\leavevmode\vadjust pre{\hypertarget{ref-pierson_when_1993}{}}%
+Pierson, P. (1993). When {Effect Becomes Cause}: {Policy Feedback} and
+{Political Change}. \emph{World Politics}, \emph{45}(4), 595--628.
+\url{https://doi.org/10.2307/2950710}
+
+\leavevmode\vadjust pre{\hypertarget{ref-reynolds_perceptions_2014}{}}%
+Reynolds, J., \& Xian, H. (2014). Perceptions of meritocracy in the land
+of opportunity. \emph{Research in Social Stratification and Mobility},
+\emph{36}, 121--137. \url{https://doi.org/10.1016/j.rssm.2014.03.001}
+
+\leavevmode\vadjust pre{\hypertarget{ref-sandel_tyranny_2020}{}}%
+Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the
+common good?} (First edition). New York: {Farrar, Straus and Giroux}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-schneider_poverty_2015}{}}%
+Schneider, S. M., \& Castillo, J. C. (2015). Poverty {Attributions} and
+the {Perceived Justice} of {Income Inequality} : {A Comparison} of
+{East} and {West Germany}.
+\url{https://doi.org/10.1177/0190272515589298}
+
+\leavevmode\vadjust pre{\hypertarget{ref-schroder_income_2017}{}}%
+Schröder, M. (2017). Is {Income Inequality Related} to {Tolerance} for
+{Inequality}? \emph{Social Justice Research}, \emph{30}(1), 23--47.
+\url{https://doi.org/10.1007/s11211-016-0276-8}
+
+\leavevmode\vadjust pre{\hypertarget{ref-singer_applied_2009}{}}%
+Singer, J. D., \& Willett, J. B. (2009). \emph{Applied longitudinal data
+analysis: Modeling change and event occurence}. New York: Oxford
+University Press, Incorporated.
+
+\leavevmode\vadjust pre{\hypertarget{ref-svallfors_political_2007}{}}%
+Svallfors, S. (Ed.). (2007). \emph{The {Political Sociology} of the
+{Welfare State}: {Institutions}, {Social Cleavages}, and {Orientations}}
+(1st ed.). Stanford University Press.
+\url{https://doi.org/10.2307/j.ctvr0qv0q}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tejero-peregrina_perceived_2025}{}}%
+Tejero-Peregrina, L., Willis, G., Sánchez-Rodríguez, Á., \&
+Rodríguez-Bailón, R. (2025). From {Perceived Economic Inequality} to
+{Support} for {Redistribution}: {The Role} of {Meritocracy Perception}.
+\emph{International Review of Social Psychology}, \emph{38}(1), 4.
+\url{https://doi.org/10.5334/irsp.1013}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tilly_contentious_2015}{}}%
+Tilly, C., \& Tarrow, S. (2015). \emph{Contentious {Politics}} (Second
+Edition, New to this Edition:, Second Edition, New to this Edition:).
+Oxford, New York: Oxford University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-trump_income_2018}{}}%
+Trump, K.-S. (2018). Income {Inequality Influences Perceptions} of
+{Legitimate Income Differences}. \emph{British Journal of Political
+Science}, \emph{48}(4), 929--952.
+\url{https://doi.org/10.1017/S0007123416000326}
+
+\leavevmode\vadjust pre{\hypertarget{ref-vondemknesebeck_are_2016}{}}%
+Von Dem Knesebeck, O., Vonneilich, N., \& Kim, T. J. (2016). Are health
+care inequalities unfair? {A} study on public attitudes in 23 countries.
+\emph{International Journal for Equity in Health}, \emph{15}(1), 61.
+\url{https://doi.org/10.1186/s12939-016-0350-8}
+
+\leavevmode\vadjust pre{\hypertarget{ref-willis_legitimacy_2015}{}}%
+Willis, G. B., Rodríguez-Bailón, R., López-Rodríguez, L., \&
+García-Sánchez, E. (2015). Legitimacy {Moderates} the {Relation Between
+Perceived} and {Ideal Economic Inequalities}. \emph{Social Justice
+Research}, \emph{28}(4), 493--508.
+\url{https://doi.org/10.1007/s11211-015-0253-7}
+
+\leavevmode\vadjust pre{\hypertarget{ref-wilson_role_2003}{}}%
+Wilson, C. (2003). The {Role} of a {Merit Principle} in {Distributive
+Justice}. \emph{The Journal of Ethics}, \emph{7}(3), 277--314.
+\url{https://doi.org/10.1023/A:1024667228488}
+
+\leavevmode\vadjust pre{\hypertarget{ref-young_rise_1962}{}}%
+Young, M. (1962). \emph{The rise of the meritocracy}. Baltimore: Penguin
+Books.
+
 \end{CSLReferences}

### 2025-05-22 Title-Page.Rmd
Up title page

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
linestretch: '1.5'
output:
  pdf_document:
    toc: false
    number_sections: false
    keep-tex: true
geometry: margin=0.78in
header-includes:
  - \usepackage{times}           # Times New Roman
  - \usepackage{caption}
  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
  - \usepackage[figurename=Fig.]{caption}
  - \usepackage{graphicx}
  - \usepackage{float}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
fontsize: 12pt

---

## Authors Identification

Juan Carlos Castillo, Department of Sociology, Universidad de Chile, Santiago de Chile[^*].  
ORCID: 0000-0003-1265-7854

Andreas Laffert, Institute of Sociology, Pontificia Universidad Católica de Chile.  
ORCID: 0000-0002-9008-2454

Kevin Carrasco, Center for Social Conflict and Cohesion Studies.  
ORCID: 0000-0002-9348-8361

Julio Iturra, Bremen International Graduate School of Social Sciences, Bremen, Germany.
ORCID: 0000-0003-1993-3479

## Exclusive Submission Statement

The authors confirm that this article has been submitted exclusively to this journal.

## Acknowledgments

The authors recognize the support of the National Research and Development Agency (ANID) through the grant from the National Fund for Scientific and Technological Development (FONDECYT) N° 1210847; from the Center for Conflict and Social Cohesion Studies (COES) grant Financing Fund for Research Centers in Priority Areas (FONDAP) N° 1523A0005; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046.

## Authors' Contributions

Author roles were classified using the Contributor Role Taxonomy [CRediT](https://credit.niso.org/) as follows: Juan Carlos Castillo: Conceptualization, Supervision, Methodology, Writing - review and editing; Andreas Laffert: Conceptualization, Methodology, Formal analysis, Software, Visualization, Writing - review and editing; Kevin Carrasco: Methodology, Formal analysis, Writing - review and editing. Julio Iturra: Conceptualization, Writing - review and editing.

## Conflict of Interest

The authors declare no conflict of interest.

## Funding

This research received funding from ANID/FONDECYT grant 1210847 and the Center for Conflict and Social Cohesion Studies (COES) through ANID/FONDAP grant 1513009.

## Data Avaibility Statement

The original data presented in the study are openly available in <https://dataverse.harvard.edu/api/access/datafile/10797987>

Access to the data, code, and analysis presented in the study are openly available in <https://github.com/justicia-distributiva/market-justice-panel-elsoc>.

[^*]: Corresponding author (juancastillov@uchile.cl - [jc-castillo.com](https://juancarloscastillo.github.io/jc-castillo/) - Ignacio Carrera Pinto 1045 Ñuñoa, Santiago de Chile).

### 2025-05-22 cover_letter.qmd
Up title page

--- 2025-05-22 Title-Page.Rmd
@@
 title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
-linestretch: '1.5'
-output:
-  pdf_document:
-    toc: false
-    number_sections: false
+format:
+  html: 
+    code-fold: true
+    html-math-method: katex
+    number-sections: true
+  pdf: 
+    geometry: margin=2cm
+    template-partials: 
+      - title.tex
     keep-tex: true
-geometry: margin=0.78in
-header-includes:
-  - \usepackage{times}           # Times New Roman
-  - \usepackage{caption}
-  - \captionsetup[figure, table]{labelfont={bf},labelformat={default},labelsep=period}
-  - \usepackage[figurename=Fig.]{caption}
-  - \usepackage{graphicx}
-  - \usepackage{float}
-  - \usepackage{booktabs}
-  - \usepackage{longtable}
-  - \usepackage{array}
-  - \usepackage{multirow}
-  - \usepackage{wrapfig}
-  - \usepackage{float}
-  - \usepackage{colortbl}
-  - \usepackage{pdflscape}
-  - \usepackage{tabu}
-  - \usepackage{threeparttable}
+    number-sections: true
+editor: source
+link-citations: true
+linestretch: 1.15       
+mainfont: Times New Roman
 fontsize: 12pt          
-
+colorlinks: true
+fig-height: 4
+fig-width: 7.5
 ---

### 2025-05-22 cover_letter.tex
Up title page

--- 2025-04-30 paper.tex
@@
 \makeatother
-% definitions for citeproc citations
-\NewDocumentCommand\citeproctext{}{}
-\NewDocumentCommand\citeproc{mm}{%
-  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
-\makeatletter
- % allow citations to break across lines
- \let\@cite@ofmt\@firstofone
- % avoid brackets around text for \cite:
- \def\@biblabel#1{}
- \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
-\makeatother
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
- {\begin{list}{}{%
-  \setlength{\itemindent}{0pt}
-  \setlength{\leftmargin}{0pt}
-  \setlength{\parsep}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-   \setlength{\leftmargin}{\cslhangindent}
-   \setlength{\itemindent}{-1\cslhangindent}
-  \fi
-  % set entry spacing
-  \setlength{\itemsep}{#2\baselineskip}}}
- {\end{list}}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
-\usepackage[noblocks]{authblk}
-\renewcommand*{\Authsep}{, }
-\renewcommand*{\Authand}{, }
-\renewcommand*{\Authands}{, }
-\renewcommand\Affilfont{\small}
 \makeatletter
@@
   pdftitle={Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)},
-  pdfauthor={Equipo EDUMER},
   colorlinks=true,
@@
 
-  \author{Juan Carlos Castillo}
-            \affil{%
-                  Departamento de Sociología, Universidad de Chile
-              }
-          \affil{%
-                  Centro de estudios del conflicto y cohesión social
-                  (COES)
-              }
-          \affil{%
-                  Núcleo milenio de desigualdades y oportunidades
-                  digitales (NUDOS)
-              }
-        \author{Andreas Laffert}
-            \affil{%
-                  Instituto de Sociología, Pontificia Universidad
-                  Católica de Chile
-              }
-        \author{Kevin Carrasco}
-            \affil{%
-                  Centro de estudios del conflicto y cohesión social
-                  (COES)
-              }
-        \author{Julio Iturra}
-            \affil{%
-                  International Graduate School of Social Sciencies
-                  (BIGSSS), University of Bremen, Germany
-              }
-
 \date{}
@@
 \maketitle
-\begin{abstract}
-This study examines the interplay between perceptions of economic
-inequality and meritocracy in shaping preferences for market justice in
-Chile from 2016 to 2023. Economic inequality has intensified globally,
-raising concerns about its social impacts and the legitimacy of
-market-based distributions of resources. By utilizing longitudinal data
-from the Chilean Longitudinal Social Survey (ELSOC), the research
-investigates how subjective assessments of inequality influence
-attitudes towards social justice, focusing on the extent to which
-individuals believe access to essential services like healthcare,
-education, and pensions should depend on personal merit. The findings
-suggest that as perceptions of economic inequality rise, individuals
-tend to favor redistributive policies, although those with strong
-meritocratic beliefs are more likely to justify existing inequalities.
-The analysis also highlights the effect of major social movements in
-Chile during this period, which have shifted public discourse and
-potentially altered perceptions of inequality and meritocracy. This
-paper contributes to the understanding of how individual beliefs about
-fairness and equity evolve within a socio-political context
-characterized by high inequality and extensive free-market principles.
-\newline \textbf{Keywords}: Economic inequality, meritocracy, market
-justice, Chile, public preferences
-\end{abstract}
 
 \setstretch{1.15}
-\section{Introduction}\label{introduction}
+\section{Authors Identification}\label{authors-identification}
 
-Since 1980, economic inequality and wealth concentration have
-dramatically increased worldwide, becoming one of the main challenges
-for the social sciences. Globally, in 2021, less than 50\% of the world
-population owned only 2\% of the wealth, while the richest 10\%
-concentrated 76\%, and the wealthiest 1\% captured nearly 38\% of total
-assets (\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}). This
-context of economic disparity has sparked renewed interest in studying
-not only the objective aspects of inequality, such as income and access
-to resources, but also its subjective dimensions, including perceptions,
-beliefs, and associated attitudes
-(\citeproc{ref-janmaat_subjective_2013}{Janmaat, 2013}). This has led,
-among other reasons, to widespread concern in the social sciences about
-the political delegitimization that economic inequality can produce
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}).
-Understanding how people perceive inequality is crucial, as these
-perceptions can influence how societies comprehend and justify the
-distribution of goods and services.
+Juan Carlos Castillo, Department of Sociology, Universidad de Chile,
+Santiago de Chile\footnote{Corresponding author (juancastillov@uchile.cl
+  -
+  \href{https://juancarloscastillo.github.io/jc-castillo/}{jc-castillo.com}
+  - Ignacio Carrera Pinto 1045 Ñuñoa, Santiago de Chile).}.\\
+ORCID: 0000-0003-1265-7854
 
-The perception of economic inequality can be understood as an
-individual's subjective assessment of how resources are allocated among
-members of a given society (\citeproc{ref-akyelken_urban_2020}{Akyelken,
-2020}). Regardless of their measurement, various studies have shown that
-this perception often underestimates the gap between the rich and the
-poor, which could have implications for attitudes toward the
-distribution of goods and services
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
-\citeproc{ref-schroder_income_2017}{Schröder, 2017}). Moreover, in
-recent years, there has been a discussion about the subjective and
-objective aspects of inequality, demonstrating inconsistencies between
-the two (\citeproc{ref-trump_income_2018}{Trump, 2018}). This tension is
-relevant because advancing the study of perceptions could help
-understand how perceived inequality affects the distribution of
-resources within a society (Becker, 2020).
+Andreas Laffert, Institute of Sociology, Pontificia Universidad Católica
+de Chile.\\
+ORCID: 0000-0002-9008-2454
 
-Perception of economic inequality has been associated with important
-outcomes such as redistributive preferences, justification of
-inequality, and legitimacy of the economic system
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
-\citeproc{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et al.,
-2020}, \citeproc{ref-garcia-sanchez_perceptions_2018}{2018}). Research
-shows that greater perceived inequality fosters stronger preferences for
-redistribution, regardless of actual inequality levels
-(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
-Treisman, 2018}). However, other findings suggest that heightened
-perceptions of income gaps between high- and low-paying occupations may
-lead to greater justification of inequality
-(\citeproc{ref-Castillo2011}{Castillo, 2011}). A less explored area in
-this regard concerns market justice, which denotes the degree to which
-individuals consider just that social goods and services (e.g., health,
-pensions, education) are allocated according to individual contribution,
-competition, and ability to pay
-(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
-\citeproc{ref-lane_market_1986}{Lane, 1986};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Market justice attitudes
-reflect the belief that the market promotes procedural
-fairness---equality of opportunity---so that outcomes result from
-individual achievement (\citeproc{ref-lane_market_1986}{Lane, 1986}).
-Indeed, Lindh (\citeproc{ref-lindh_public_2015}{2015}) shows that market
-justice preferences are stronger in countries with higher private
-spending on services than in those with more comprehensive welfare
-systems. From this perspective, perceptions of economic inequality may
-be key, as lower perceived gaps can reinforce acceptance of free-market
-systems and justify existing inequalities.
+Kevin Carrasco, Center for Social Conflict and Cohesion Studies.\\
+ORCID: 0000-0002-9348-8361
 
-Meritocracy posits that inequality can be legitimized through
-distributive criteria such as effort and talent
-(\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
-\citeproc{ref-young_rise_1962}{Young, 1962}). Previous studies have
-demonstrated that people with stronger meritocratic beliefs tend to
-perceive less inequality as they attribute economic differences to
-individual achievements (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
-\citeproc{ref-wilson_role_2003}{Wilson, 2003}) and also to justify more
-inequality as these beliefs are associated with attitudes that
-legitimize status differences
-(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In a context
-like Chile, where the distribution of goods and services is
-predominantly governed by market logics strongly introduced during the
-military dictatorship (1973-1989)
-(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}), these beliefs can
-play a crucial role in the acceptance of social inequalities.
-Incorporating the perception of meritocracy in this literature allows
-for an understanding of how attitudes toward inequality are shaped not
-only by the perception of economic gaps, but also by beliefs about how
-these gaps are justified. Furthermore, these beliefs are often
-consolidated from an early age, reinforced by institutions promoting
-values such as effort and individual skills to climb socially
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).
+Julio Iturra, Bremen International Graduate School of Social Sciences,
+Bremen, Germany. ORCID: 0000-0003-1993-3479
 
-The perception of meritocracy and the perception of economic inequality
-would interact intricately in shaping preferences for market justice. On
-the one hand, a stronger perception of merit as a criterion for
-distribution could minimize the perception of economic gaps, thus
-justifying an unequal system
-(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
-\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). On the other hand, the
-perception of inequality could moderate the impact of meritocracy, as a
-higher perception of economic gaps might challenge the idea that these
-are solely based on merit.
+\section{Exclusive Submission
+Statement}\label{exclusive-submission-statement}
 
-The primary objective of this study is to analyze the interplay between
-perceptions of inequality and meritocracy and their joint influence on
-preferences for market justice in Chile from 2016 to 2023, using
-longitudinal data from the ELSOC survey. This interaction is expected to
-provide a more comprehensive explanation of attitudes toward market
-justice in a country characterized by high inequality and strong
-free-market influence (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
-\citeproc{ref-flores_top_2020}{Flores et al., 2020}). Additionally, this
-analysis seeks to elucidate how political and social
-contingencies---such as the 2019 and 2022 social movements---might have
-moderated these relationships by prompting more critical reflection on
-the commodification of social services. Recognizing the temporal
-dimension in shaping market justice preferences is essential, given that
-such preferences are not static but are influenced by historical and
-contextual factors that challenge or reaffirm prevailing social norms.
-In this regard, variations in perceived economic inequality and
-meritocracy over time can affect how individuals endorse market-based
-approaches. Consequently, public opinion may shift from supporting
-market justice to embracing redistributive policies aimed at mitigating
-inequalities.
+The authors confirm that this article has been submitted exclusively to
+this journal.
 
-\section{Theoretical views on market justice, inequality perception, and
-meritocracy}\label{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}
+\section{Acknowledgments}\label{acknowledgments}
 
-\subsection{The justification of market
-inequality}\label{the-justification-of-market-inequality}
+The authors recognize the support of the National Research and
+Development Agency (ANID) through the grant from the National Fund for
+Scientific and Technological Development (FONDECYT) N° 1210847; from the
+Center for Conflict and Social Cohesion Studies (COES) grant Financing
+Fund for Research Centers in Priority Areas (FONDAP) N° 1523A0005; and
+Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS)
+NCS2022\_046.
 
-Conceptually, \emph{market justice} has been discussed in the literature
-as a normative principle that legitimates the distribution of economic
-rewards based on individual merit and effort. It is possible to trace
-the concept to the understanding of Lane
-(\citeproc{ref-lane_market_1986}{1986}), who defines market justice as a
-system of ``earned deserts,'' whereby individuals are seen as deserving
-of a determined distributive outcome due to their effort and skills. In
-contrast, political justice, emphasizes principles of equality and need,
-which are often represented by the welfare state action through social
-policies. An important remark is that the principles of market justice
-rely on the assumption that markets are neutral and self-regulating
-arenas, where individuals are treated fairly because they face the same
-formal rules of engagement (\citeproc{ref-lane_market_1986}{Lane,
-1986}). Consequently, the legitimacy of market justice stems from the
-assumption that inequalities are not only inevitable but fair---so long
-as the rules are transparent and opportunities are open. In this way,
-market justice provides a moral justification for inequality by framing
-it as a necessary outcome of individual responsibility.
+\section{Authors' Contributions}\label{authors-contributions}
 
-Empirical studies have shown different strategies for the study of
-market justice preferences. A common approach in the literature is to
-gauge attitudes toward the legitimacy of inequality in specific domains,
-especially when linked to income differences. This can be traced to the
-seminal work of Kluegel and Smith
-(\citeproc{ref-kluegel_beliefs_1981}{1981}) who assessed the normative
-foundations that explain public support for economic inequality. Over
-time, this approach has been extended beyond income to include other
-market-mediated outcomes, such as education, healthcare, and/or
-pensions. For example, Von Dem Knesebeck et al.
-(\citeproc{ref-vondemknesebeck_are_2016}{2016}) and Immergut and
-Schneider (\citeproc{ref-immergut_it_2020}{2020}) examine whether
-citizens consider it fair that individuals with higher incomes can
-access better healthcare, while Lee and Stacey
-(\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
-context of education in Australia. These studies usually rely on a
-survey item asking respondents to evaluate the fairness of income-based
-access to welfare services, allowing for comparing justice perceptions
-across different contexts. Similarly, comparative studies have also
-relied on these indicators to study the support for market-based
-distribution in welfare systems (\citeproc{ref-lindh_public_2015}{Lindh,
-2015}; \citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). More
-recently, Castillo et al.
-(\citeproc{ref-castillo_socialization_2024}{2024}) introduced a
-single-item composite measure of market justice to assess student
-attitudes toward income-based access to education, healthcare, and
-pensions in Chile. These empirical strategies all aim to capture the
-extent to which individuals accept inequality when framed as a
-reflection of market outcomes.
+Author roles were classified using the Contributor Role Taxonomy
+\href{https://credit.niso.org/}{CRediT} as follows: Juan Carlos
+Castillo: Conceptualization, Supervision, Methodology, Writing - review
+and editing; Andreas Laffert: Conceptualization, Methodology, Formal
+analysis, Software, Visualization, Writing - review and editing; Kevin
+Carrasco: Methodology, Formal analysis, Writing - review and editing.
+Julio Iturra: Conceptualization, Writing - review and editing.
 
-The study of market justice preferences has increasingly focused on how
-they are shaped by individuals' socioeconomic position, normative
-beliefs, and the institutional context in which they are embedded.
-Across the literature, there is empirical evidence suggesting that
-individuals in more advantaged socioeconomic positions---those with
-higher occupational class, income, and education---are more likely to
-support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
-\& Sachweh, 2019}; \citeproc{ref-svallfors_political_2007}{Svallfors,
-2007}). This tendency reflects not only material self-interest but also
-a broader moral economy, in which winners of the market system
-internalize justifications for the status quo. At the same time,
-political ideology also plays a role ---such as economically
-conservative values --- where more right-wing individuals show higher
-support for meritocracy and more skepticism towards redistribution. This
-is particularly salient in countries with more restricted public
-provision of social services. For example, Castillo et al.
-(\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
-Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
-those with right-leaning individuals are more supportive of market-based
-distribution of welfare. Beyond individual characteristics,
-country-level institutions also play a central role. In liberal welfare
-regimes like those of the United States or the United Kingdom, market
-justice preferences are more widespread, while in coordinated or
-social-democratic regimes---such as Sweden or Germany---citizens are
-generally more critical of market-based inequalities
-(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
-that market justice is also shaped contextually, which is represented by
-the development of welfare institutions.
+\section{Conflict of Interest}\label{conflict-of-interest}
 
-\subsection{Perception of inequality}\label{perception-of-inequality}
+The authors declare no conflict of interest.
 
-Perceptions of inequality have been associated with attitudes about
-market justice (\citeproc{ref-kluegel_social_1995a}{Kluegel et al.,
-1995}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research
-indicates that lower perceived inequality can reinforce support for
-market-based distributive arrangements by suggesting that the system is
-fair and that outcomes reflect effort and ability
-(\citeproc{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when inequality
-is perceived as excessive or structurally determined, individuals could
-question the legitimacy of market justice and become more supportive of
-redistributive policies
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
-\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}).
+\section{Funding}\label{funding}
 
-Perceptions of inequality refer to individuals' subjective evaluations
-of the extent, causes, and consequences of income and wealth
-disparities. Unlike objective measures such as the Gini index, perceived
-inequality captures how individuals make sense of distributive
-hierarchies in their everyday lives, shaped by reference groups, social
-comparisons, and information environments
-(\citeproc{ref-garcia-castro_perceiving_2020}{García-Castro et al.,
-2020}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
-Treisman, 2018}; \citeproc{ref-mijs_stratified_2016}{Mijs, 2016}).
-Scholars have proposed multiple dimensions of perceived inequality,
-including its magnitude (how significant are the gaps), vertical
-structure (between which groups), the trend over time (increasing or
-decreasing), and legitimacy (whether it is just or not)
-(\citeproc{ref-engelhardt_what_2018}{Engelhardt \& Wagener, 2018};
-\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}). These dimensions encompass both cognitive and normative aspects
-of perceptions of inequality and can vary across societies and social
-groups, depending on exposure, ideology, and personal experience
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
-\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
-2018}).
+This research received funding from ANID/FONDECYT grant 1210847 and the
+Center for Conflict and Social Cohesion Studies (COES) through
+ANID/FONDAP grant 1513009.
 
-Among the different approaches to measuring perceived inequality, one of
-the most widely used is the estimation of wage gaps between occupational
-extremes, such as between a CEO and a manual worker. This type of item
-provides a concrete frame that respondents can relate to more easily
-than abstract questions about national income distribution
-(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
-\citeproc{ref-easterbrook_social_2021}{Easterbrook, 2021};
-\citeproc{ref-willis_legitimacy_2015}{Willis et al., 2015}). While it
-enables the estimation of inequality using simple heuristics, this
-method is not without challenges. For instance, people often lack
-reliable knowledge about the earnings of those at the top of the income
-ladder, which leads to high variability in responses and the use of
-biased mental shortcuts (\citeproc{ref-knell_perceptions_2020}{Knell \&
-Stix, 2020}). Despite this, perceived wage gaps are strong predictors of
-political attitudes
-(\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
-2018}; \citeproc{ref-pedersen_attitudes_2019}{Pedersen \& Mutz, 2019}),
-making them a valuable tool for understanding public responses to
-economic disparities.
+\section{Data Avaibility Statement}\label{data-avaibility-statement}
 
-Another relevant critique of this approach lies in its conflation of
-different psychological constructs. Many surveys assess perceived
-inequality through Likert-type items that ask respondents to agree or
-disagree with statements, such as ``income differences are too large,''
-which captures general concern or discomfort rather than a specific
-perception (\citeproc{ref-Castillo2011}{Castillo, 2011};
-\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}). These items mix cognitive estimations with affective
-evaluations, complicating the interpretation of what respondents
-perceive versus what they morally reject. As a result, the conceptual
-clarity between perceived inequality and inequality aversion remains
-blurred in many empirical studies. To address this limitation, recent
-work has emphasized the need to distinguish between absolute and
-comparative measures, as well as between ideal and actual estimates of
-economic gaps (\citeproc{ref-auspurg_why_2017}{Auspurg et al., 2017};
-\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De
-Carvalho, 2022}).
+The original data presented in the study are openly available in
+\url{https://dataverse.harvard.edu/api/access/datafile/10797987}
 
-\subsection{Perception of meritocracy}\label{perception-of-meritocracy}
-
-Meritocracy constitutes a central ideological framework for legitimizing
-different types of social inequality, for instance through market
-justice beliefs. Rooted in the belief that rewards and positions should
-be allocated based on individual effort and talent, meritocracy operates
-as a normative ideal and a descriptive belief about how society
-functions. As initially conceptualized by Michael Young
-(\citeproc{ref-young_rise_1962}{1962}), the term was meant to critique a
-system in which merit-based stratification becomes a new form of
-inequality. However, over time, meritocracy has been widely supported in
-many societies as a fair and desirable principle of distribution,
-particularly within liberal democracies and market-oriented economies
-(\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
-\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
-perspective, the belief in meritocracy is more than a cognitive
-assessment; it reflects a moral lens through which individuals interpret
-inequality. People who believe that success results from hard work and
-talent are more likely to view social and economic disparities as
-legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
-\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}).
-Conversely, if they see outcomes as driven by luck, social origin, or
-systemic barriers, inequality is more likely to be perceived as unjust.
-This distinction becomes crucial in societies with persistent structural
-inequality, where public narratives often emphasize personal
-responsibility and merit while overlooking entrenched disadvantages.
-
-We adopt a multidimensional perspective on meritocracy, distinguishing
-between two key dimensions: effort-based and talent-based perceptions.
-This distinction is essential, as it captures different pathways through
-which individuals justify inequality. Effort-based meritocracy
-emphasizes hard work and perseverance as the basis for social rewards,
-aligning closely with cultural narratives of personal responsibility. A
-talent-based meritocracy, by contrast, emphasizes intelligence and
-innate abilities, which are often perceived as less malleable and more
-unequally distributed. Both dimensions have been shown to correlate with
-acceptance of inequality, but they may carry distinct implications for
-how inequality is justified in specific domains
-(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}).
-The relevance of this distinction is supported by recent studies, which
-show that individuals respond differently to these dimensions. For
-instance, perceptions that effort is rewarded in society are more
-strongly associated with positive evaluations of fairness and acceptance
-of unequal outcomes (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
-2023}). This may be because effort is seen as a controllable and morally
-virtuous trait, whereas talent is often perceived as a natural
-advantage. Consequently, effort-based meritocracy is likely more potent
-in legitimizing inequality, particularly in neoliberal contexts.
-
-These dimensions of meritocracy reflect how respondents perceive
-society's distributive logic, regardless of whether they endorse
-meritocratic principles. This distinction aligns with recent findings
-indicating that individuals distinguish between how merit is perceived
-in society and how it should ideally operate, which in turn shapes their
-preferences for redistribution and justice
-(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
-2025}) Meritocratic beliefs serve as symbolic justifications for unequal
-outcomes, particularly when access is stratified by income or social
-opportunity. Prior studies have shown that individuals who perceive
-higher levels of meritocracy tend to express stronger support for
-unequal distributions that reflect market outcomes
-(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
-\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
-
-In addition to influencing individual attitudes toward inequality,
-meritocratic beliefs can contribute to social division and the
-stigmatization of disadvantaged groups. Recent research has demonstrated
-that exposure to meritocratic narratives can reinforce the belief that
-poverty results from individual failings rather than systemic
-conditions, reducing support for redistributive measures and increasing
-the stigmatization of the poor (\citeproc{ref-hoyt_mindsets_2023}{Hoyt
-et al., 2023}). This reinforces negative stereotypes and reduces empathy
-toward individuals from lower socioeconomic backgrounds. Moreover,
-Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argues
-that meritocratic narratives can serve as feedback mechanisms that shape
-public opinion and well-being by framing individuals' understanding of
-welfare outcomes as deserved or undeserved within existing institutional
-structures. This psychological mechanism highlights the normative power
-of meritocracy in stabilizing unequal systems by shaping political
-attitudes and personal perceptions of success and failure.
-
-\subsection{This study}\label{this-study}
-
-Building upon the previous literature, this study proposes that
-attitudes toward market justice are shaped by a dynamic interplay
-between individuals' perceptions of economic inequality and their
-beliefs in meritocracy. Specifically, we argue that both perceptions
-independently and interactively influence the extent to which
-individuals endorse market-based distributions of social goods and
-services in Chile.
-
-First, consistent with previous findings, we expect that a lower
-perception of economic inequality will be associated with stronger
-market justice preferences. When individuals perceive smaller income
-gaps, they are more likely to view market mechanisms as fair and
-legitimate, reinforcing the acceptance of outcomes based on competition
-and ability to pay. Conversely, a heightened perception of inequality
-may erode confidence in market fairness, weakening support for
-market-based distribution. This relationship is particularly relevant in
-the context of Chile, where the neoliberal economic model has been a
-dominant force in shaping public attitudes toward inequality and
-justice.
-
-Second, higher perceived meritocracy is expected to be positively
-associated with market justice preferences. Individuals who believe that
-effort and talent primarily determine success are more likely to justify
-unequal outcomes and endorse the notion that markets allocate resources
-fairly according to individual merit. This aligns with the idea that
-meritocratic beliefs serve as a moral framework that legitimizes
-market-based inequalities, as individuals perceive the system as just
-when they believe that rewards are based on individual merit. This is
-particularly relevant in contexts where neoliberal ideologies dominate,
-as they often emphasize individual responsibility and competition as the
-basis for social order.
-
-Third, we propose that perceptions of meritocracy and perceptions of
-economic inequality interact in shaping market justice attitudes.
-Specifically, we argue that the legitimizing effect of perceived
-meritocracy on market justice preferences is moderated by perceived
-economic inequality: when perceived inequality is low, the positive
-association between meritocratic beliefs and market justice preferences
-will be stronger. However, when perceived inequality is high, this
-association will weaken, as greater awareness of large economic gaps may
-challenge the view that outcomes are purely merit-based.
-
-Additionally, this study examines whether these relationships vary over
-time in response to significant social and political events. In
-particular, we explore whether the political outburst of 2019 and the
-subsequent constitutional processes in Chile have altered how
-perceptions of inequality and meritocracy shape market justice
-attitudes. We expect that after these critical events, perceptions of
-inequality may have a stronger negative effect on market justice
-preferences, reflecting increased societal questioning of market-based
-allocation mechanisms.
-
-Based on these arguments, we propose the following hypotheses:
-
-\(H1\): Higher perceived economic inequality is associated with higher
-market justice preferences.
-
-\(H2\): Higher meritocratic beliefs are associated with higher market
-justice preferences.
-
-\(H3\): The positive association between meritocratic beliefs and market
-justice preferences is moderated by perceived economic inequality;
-specifically, this association is weaker when perceived economic
-inequality is high.
-
-\(H4\): The effects described in H1--H3 are attenuated after major
-social mobilizations (2019--2022), reflecting increased critical views
-of the market's role in allocating social goods.
-
-\section{Data, Variables and Methods}\label{data-variables-and-methods}
-
-\subsection{Data}\label{data}
-
-This study draws on data from the Chilean Longitudinal Social Survey
-(ELSOC), a panel study collected annually from 2016 to 2023. The survey
-evaluates how individuals think, feel, and behave regarding conflict and
-social cohesion in Chile. ELSOC employs a probabilistic, stratified,
-clustered, multistage sampling design encompassing major urban centers
-(Santiago, Valparaíso, and Concepción) and smaller cities. The target
-population includes women and men aged 18--75 who are habitual residents
-of private dwellings. The first wave included 2,927 participants,
-representing northern and southern regions, covering 77\% of the
-country's total population and 93\% of its urban population, with a
-response rate of 62.4\% (\citeproc{ref-elsoc_estudio_2022}{ELSOC,
-2022}).
-
-The survey has been conducted yearly since 2016, except in 2020, when it
-was suspended due to the pandemic. Waves 2016, 2017, 2018, 2019, 2022,
-and 2023 used computer-assisted personal interviewing (CAPI), while a
-reduced wave in 2021 employed computer-assisted telephone interviews
-(CATI). Wave 3 included a refreshment sample (N = 1,519) to counter
-attrition, but these cases are not used here to capture longer response
-trends. Between wave 1 and wave 7, attrition amounted to 40\%, achieving
-a final sample of N = 1,741. Longitudinal weights adjust for both the
-sampling design and potential biases arising from systematic
-non-participation. Further details on sampling, attrition, and weighting
-can be found at \url{https://coes.cl/encuesta-panel/}, and the dataset
-is publicly available at
-\url{https://dataverse.harvard.edu/dataverse/elsoc}.
-
-\subsection{Variables}\label{variables}
-
-\textbf{Market justice preferences}: The dependent variable in this
-study is preferences for market justice. This construct is
-operationalized through three items that capture how strongly
-individuals justify conditioning access to social services---healthcare,
-pensions, and education---on income. Specifically, the justification of
-inequality in healthcare is assessed by the question: ``Is it fair in
-Chile that people with higher incomes can access better healthcare than
-people with lower incomes?'' The same question is posed for pensions and
-education. In all cases, respondents indicate their level of agreement
-on a five-point Likert scale ranging from 1 (``strongly disagree'') to 5
-(``strongly agree''). Additionally, we include a composite measure of
-``market justice preferences'', calculated as the average of these three
-items (\(\alpha\) = 0.84). This index ranges from 1 to 5, with higher
-values indicating stronger preferences for market justice (see
-Table~\ref{tbl-summary1}).
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3367}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3367}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2143}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1122}}@{}}
-\caption{Dependent variables for the first wave
-(2016)}\label{tbl-summary1}\tabularnewline
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Label
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Stats / Values
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Freqs (\% of Valid)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Valid
-\end{minipage} \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Label
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Stats / Values
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Freqs (\% of Valid)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Valid
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-Health distributive justice &
-\begin{minipage}[t]{\linewidth}\raggedright
-1. Strongly desagree\\
-2. Desagree\\
-3. Neither agree nor desagre\\
-4. Agree\\
-5. Strongly agree\strut
-\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-558 (37.2\%)\\
-729 (48.6\%)\\
-63 ( 4.2\%)\\
-133 ( 8.9\%)\\
-18 ( 1.2\%)\strut
-\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-1501\\
-(100.0\%)\strut
-\end{minipage} \\
-Pension distributive justice &
-\begin{minipage}[t]{\linewidth}\raggedright
-1. Strongly desagree\\
-2. Desagree\\
-3. Neither agree nor desagre\\
-4. Agree\\
-5. Strongly agree\strut
-\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-426 (28.4\%)\\
-718 (47.8\%)\\
-108 ( 7.2\%)\\
-226 (15.1\%)\\
-23 ( 1.5\%)\strut
-\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-1501\\
-(100.0\%)\strut
-\end{minipage} \\
-Education distributive justice &
-\begin{minipage}[t]{\linewidth}\raggedright
-1. Strongly desagree\\
-2. Desagree\\
-3. Neither agree nor desagre\\
-4. Agree\\
-5. Strongly agree\strut
-\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-521 (34.7\%)\\
-783 (52.2\%)\\
-73 ( 4.9\%)\\
-113 ( 7.5\%)\\
-11 ( 0.7\%)\strut
-\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-1501\\
-(100.0\%)\strut
-\end{minipage} \\
-Market justice preferences & \begin{minipage}[t]{\linewidth}\raggedright
-Mean (sd) : 2 (0.8)\\
-min \textless{} med \textless{} max:\\
-1 \textless{} 2 \textless{} 5\\
-IQR (CV) : 0.7 (0.4)\strut
-\end{minipage} & 12 distinct values &
-\begin{minipage}[t]{\linewidth}\raggedright
-1501\\
-(100.0\%)\strut
-\end{minipage} \\
-\end{longtable}
-
-\textbf{Perception of economic inequality}: The main independent
-variable refers to the perception of economic inequality, measured
-through the perceived wage gap.
-(\citeproc{ref-castillo_cual_2009}{Castillo, 2009};
-\citeproc{ref-gijsberts_thelegitimation_1999}{Gijsberts, 1999};
-\citeproc{ref-hadler_why_2005}{Hadler, 2005}). This measure is derived
-from the salary gap between the perceived salaries of jobs at opposite
-ends of the occupational hierarchy. Specifically, it relies on the
-division between the perceived salary of a large-company president and
-that of an unskilled worker (\citeproc{ref-Castillo2011}{Castillo,
-2011}). Higher values of this term indicate a greater perception of
-economic inequality between occupations located at the extremes of the
-status continuum. This measure includes a logarithmic term in order to
-adjust income magnitudes (usually fewer cases with high income):
-
-\[
-\text{perceived wage gap} = \log_{10}\left(\frac{\text{perceived salary of a large-company president}}{\text{perceived salary of an unskilled worker}}\right)
-\]
-
-\textbf{Perception of Meritocracy}: this variable is operationalized
-through two components, namely effort and talent
-(\citeproc{ref-young_rise_1962}{Young, 1962}). The item used to gauge
-effort is: ``In Chile, people are rewarded for their efforts,'' while
-the item for talent is: ``In Chile, people are rewarded for their
-intelligence and skills''. In both cases, respondents indicate their
-level of agreement on a five-point Likert scale, ranging from 1
-(``strongly disagree'') to 5 (``strongly agree'').
-
-Table 2 shows the independent variables used, their response categories
-and their frequencies.
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3774}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.3113}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2075}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1038}}@{}}
-\caption{Independent variables for the first wave
-(2016)}\label{tbl-summary2}\tabularnewline
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Label
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Stats / Values
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Freqs (\% of Valid)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Valid
-\end{minipage} \\
-\midrule\noalign{}
-\endfirsthead
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Label
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Stats / Values
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Freqs (\% of Valid)
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Valid
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-Inequality gap perception & \begin{minipage}[t]{\linewidth}\raggedright
-Mean (sd) : 3.7 (1.1)\\
-min \textless{} med \textless{} max:\\
-0.4 \textless{} 3.7 \textless{} 6.9\\
-IQR (CV) : 1.6 (0.3)\strut
-\end{minipage} & 296 distinct values &
-\begin{minipage}[t]{\linewidth}\raggedright
-1501\\
-(100.0\%)\strut
-\end{minipage} \\
-People are rewarded for their efforts &
-\begin{minipage}[t]{\linewidth}\raggedright
-1. Strongly desagree\\
-2. Desagree\\
-3. Neither agree nor desagre\\
-4. Agree\\
-5. Strongly agree\strut
-\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-169 (11.3\%)\\
-693 (46.2\%)\\
-263 (17.5\%)\\
-328 (21.9\%)\\
-48 ( 3.2\%)\strut
-\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-1501\\
-(100.0\%)\strut
-\end{minipage} \\
-People are rewarded for their intelligence &
-\begin{minipage}[t]{\linewidth}\raggedright
-1. Strongly desagree\\
-2. Desagree\\
-3. Neither agree nor desagre\\
-4. Agree\\
-5. Strongly agree\strut
-\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-134 ( 8.9\%)\\
-617 (41.1\%)\\
-294 (19.6\%)\\
-401 (26.7\%)\\
-55 ( 3.7\%)\strut
-\end{minipage} & \begin{minipage}[t]{\linewidth}\raggedright
-1501\\
-(100.0\%)\strut
-\end{minipage} \\
-\end{longtable}
-
-\emph{Controls}
-
-Sociodemographic and attitudinal variables are included to control for
-potential composition effects in the population. In terms of
-sociodemographic characteristics, we incorporate per capita household
-income quintile, educational level (1=Less than Universitary,
-2=Universitary), age (in years), and sex (1=Male, 2=Female), which have
-been previously shown to influence market justice preferences
-significantly (\citeproc{ref-castillo_socialization_2024}{Castillo et
-al., 2024}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Regarding
-attitudinal variables, we include political identification (1=Left,
-2=Center, 3=Right, 4=No identification) and subjective social status
-(ranging from 1 to 10) as they may confound the relationship between
-market justice preferences and perceptions of inequality and meritocracy
-(\citeproc{ref-schneider_poverty_2015}{Schneider \& Castillo, 2015}).
-
-\subsection{Methods}\label{methods}
-
-Given the data's hierarchical structure, in which observations are
-nested in survey waves, we employ longitudinal multilevel linear models
-(\citeproc{ref-singer_applied_2009}{Singer \& Willett, 2009}). In a
-panel-data framework, within-person effects capture how shifts in
-individual-level variables across waves are associated with variations
-in market justice preferences. By contrast, between-person effects focus
-on differences among individuals, explaining how long-term (or average)
-values relate to overall levels of market justice preferences.
-
-To estimate within-person effects, we use group-mean centering, where
-each respondent functions as the ``group'' (i.e., observations nested
-within persons). Meanwhile, the between-person effects are derived from
-each individual's average on these variables, calculated across the
-waves of panel data.
-
-All the analyses were conducted using R software and the \emph{lme4}
-package (\citeproc{ref-bates_fitting_2015}{Bates et al., 2015}).
-
-\section{Results}\label{results}
-
-\subsection{Descriptive}\label{descriptive}
-
-Figure~\ref{fig-alluvial} shows the annual frequencies of market justice
-preferences for healthcare, pensions, and education from 2016 to 2023.
-Each year presents stacked percentage frequencies, and the flows between
-them reflect opinion changes among the same individuals from one year to
-the next, given that we are using panel data. For instance, of the
-40.8\% who strongly disagreed with justifying inequality in healthcare
-in 2019, around 24.3\% maintained that position in 2022, while the
-remaining 16.5\% shifted toward other response categories---primarily
-moving into disagreement rather than strong disagreement. Overall, more
-than half of the respondents exhibit a high level of disagreement
-(disagree + strongly disagree) with inequality in these three social
-service areas over time. Despite this general pattern, recent waves show
-a slight decrease in disagreement and a corresponding rise in support
-for inequality. Specifically, in healthcare and education, although
-disagreement remains substantial, agreement (agree + strongly agree)
-increased from 7.4\% and 7.2\% in 2019 to 13.1\% and 14.2\% in 2023,
-respectively. This shift is most evident in pensions, where the combined
-agree/strongly agree category grew by about 10 percentage points, from
-16.9\% in 2016 to 28\% in 2023.
-
-\begin{figure}[H]
-
-\caption{\label{fig-alluvial}Change in the justification of inequality
-in healthcare, pensions and education over time (2016-2023)}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-alluvial-1.pdf}
-
-}
-
-\end{figure}%
-
-Regarding the main variables of this study, Figure~\ref{fig-meanchange}
-depicts their average changes over the years. We observe an increase in
-the average level of market justice preferences in the most recent
-waves, which begins in 2019. The highest average consistently appears
-for perceived economic inequality, although this variable shows a
-downward trend of roughly one point over time. Interestingly, while
-perceptions of economic inequality declined in the latest measurements
-(2022--2023), market justice preferences increased. The meritocracy
-measures remain stable, though the perception that individuals are
-rewarded for intelligence is slightly higher than the perception that
-they are rewarded for their effort.
-
-\begin{figure}[H]
-
-\caption{\label{fig-meanchange}Change in the mean of market justice
-preferences, economic inequality perception, and meritocracy
-(2016-2023)}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-meanchange-1.pdf}
-
-}
-
-\end{figure}%
-
-Figure~\ref{fig-matrix} presents the correlation matrix for the main
-variables in the latest wave (2023). Overall, the coefficients range
-from low to moderate. The association between market justice preferences
-and economic inequality perception is negative but small and
-statistically significant (\(r\) = -0.05, \(p\) \textless{} .05). By
-contrast, market justice preferences positively and significantly
-correlate with the two meritocracy variables (\(r\) = 0.15, \(p\)
-\textless{} .01; \(r\) = 0.13, \(p\) \textless{} .01). Perceived
-economic inequality has negative and nonsignificant correlations with
-both meritocracy perceptions (\(r\) = -0.04, \(p\) \textgreater{} .05;
-\(r\) = -0.04, \(p\) \textgreater{} .05). Finally, the two meritocracy
-variables exhibit a strong positive association with each other (\(r\) =
-0.69, \(p\) \textless{} .01).
-
-\begin{figure}[H]
-
-\caption{\label{fig-matrix}Correlation matrix of the main variables for
-the last wave (2023)}
-
-\centering{
-
-\includegraphics[width=0.9\textwidth,height=\textheight]{paper_files/figure-pdf/fig-matrix-1.pdf}
-
-}
-
-\end{figure}%
-
-\subsection{Multilevel models}\label{multilevel-models}
-
-Table~\ref{tbl-modelos} presents the results of the multilevel models
-estimated for market justice preferences, examining both individuals
-(within) and group-level (between) effects. The intraclass correlation
-(\citeproc{ref-hoxMultilevelAnalysisTechniques2017a}{Hox et al., 2017})
-from the empty model (see Supplementary Material), which decomposes the
-variance of market justice preferences, is 0.23, indicating that
-approximately 23\% of the variation is attributable to differences
-between individuals. Complementary, 77\% of the variation corresponds to
-within-individual differences over time.
-
-\begin{table}
-
-\caption{\label{tbl-modelos}Longitudinal multilevel models for market
-justice preferences}
-
-\centering{
-
-\begin{center}
-\scalebox{0.79}{
-\begin{tabular}{l c c c c c c c}
-\hline
- & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 \\
-\hline
-Intercept                            & $1.968^{***}$  & $2.015^{***}$  & $2.030^{***}$  & $2.028^{***}$  & $2.169^{***}$  & $1.456^{***}$  & $1.547^{***}$  \\
-                                     & $(0.021)$      & $(0.035)$      & $(0.035)$      & $(0.035)$      & $(0.070)$      & $(0.097)$      & $(0.115)$      \\
-Wave (Ref.= 2016)                    &                &                &                &                &                &                &                \\
-                                     &                &                &                &                &                &                &                \\
-\quad Wave 2017                      & $-0.167^{***}$ &                &                &                &                &                &                \\
-                                     & $(0.027)$      &                &                &                &                &                &                \\
-\quad Wave 2018                      & $-0.025$       &                &                &                &                &                &                \\
-                                     & $(0.027)$      &                &                &                &                &                &                \\
-\quad Wave 2019                      & $-0.046$       &                &                &                &                &                &                \\
-                                     & $(0.027)$      &                &                &                &                &                &                \\
-\quad Wave 2022                      & $0.285^{***}$  &                &                &                &                &                &                \\
-                                     & $(0.028)$      &                &                &                &                &                &                \\
-\quad Wave 2023                      & $0.284^{***}$  &                &                &                &                &                &                \\
-                                     & $(0.027)$      &                &                &                &                &                &                \\
-Wave                                 &                & $-0.121^{***}$ & $-0.125^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.129^{***}$ \\
-                                     &                & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      \\
-Wave$^2$                             &                & $0.028^{***}$  & $0.028^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  \\
-                                     &                & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      \\
-Perception inequality (WE)           &                &                & $-0.039^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ \\
-                                     &                &                & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
-Merit: Effort (WE)                   &                &                &                & $0.070^{***}$  & $0.070^{***}$  & $0.069^{***}$  & $0.069^{***}$  \\
-                                     &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
-Merit: Talent (WE)                   &                &                &                & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    \\
-                                     &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
-Perception inequality (BE)           &                &                &                &                & $-0.042^{*}$   & $-0.006$       & $-0.033$       \\
-                                     &                &                &                &                & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
-Merit: Effort (BE)                   &                &                &                &                &                & $0.186^{***}$  & $0.178^{***}$  \\
-                                     &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
-Merit: Talent (BE)                   &                &                &                &                &                & $0.040$        & $0.019$        \\
-                                     &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
-\hline
-Controls                             & No             & No             & No             & No             & No             & No             & Yes            \\
-BIC                                  & $20780.594$    & $20780.617$    & $20781.113$    & $20726.753$    & $20736.627$    & $20655.172$    & $20749.047$    \\
-Num. obs.                            & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         \\
-Num. groups: idencuesta              & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         \\
-Var: idencuesta (Intercept)          & $0.172$        & $0.223$        & $0.219$        & $0.209$        & $0.207$        & $0.182$        & $0.171$        \\
-Var: Residual                        & $0.530$        & $0.499$        & $0.498$        & $0.495$        & $0.495$        & $0.495$        & $0.495$        \\
-Var: idencuesta ola\_num             & $$             & $0.011$        & $0.010$        & $0.009$        & $0.009$        & $0.009$        & $0.009$        \\
-Cov: idencuesta (Intercept) ola\_num & $$             & $-0.025$       & $-0.024$       & $-0.021$       & $-0.020$       & $-0.019$       & $-0.019$       \\
-\hline
-\multicolumn{8}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
-\end{tabular}
-}
-\caption{}
-\label{table:coefficients}
-\end{center}
-
-}
-
-\end{table}%
-
-According to Model 1, which includes the survey waves to capture
-intertemporal variations in the dependent variable, there is a decrease
-in 2017 (\(\beta\) = -0.167, \(p\) \textless{} .001) relative to 2016,
-and similarly in 2018 (\(\beta\) = -0.025, \(p\) \textgreater{} .05) and
-2019 (\(\beta\) = -0.046, \(p\) \textgreater{} .05), although the latter
-effects are not statistically significant. In contrast, in the more
-recent waves of 2022 and 2023, there is a statistically significant
-increase in market justice preferences (\(\beta\) = 0.285, \(p\)
-\textless{} .001; \(\beta\) = 0.284, \(p\) \textless{} .001), suggesting
-a non-linear effect. To model this trajectory over time, Model 2
-incorporates time (survey waves) as a continuous variable, along with
-its quadratic term, representing the non-linear association initially
-observed in Model 1. While the linear term (survey wave) shows a
-negative association, reflecting an overall decline in market
-preferences over time, the positive quadratic term indicates a reversal
-of this pattern in the final measurement points.
-
-Models 3 and 4 incorporate the within-group effects (WE) of the primary
-independent variables, capturing how individual changes in these
-variables over time shape the dependent variable. The results in Model 3
-suggest that the within effect of perceived economic inequality is
-negative and statistically significant (\(p\) \textless{} .001).
-Specifically, each one-point increase in an individual's perception of
-economic inequality between waves is associated with a 0.034 point
-decrease in market justice preferences. Model 4 shows that meritocratic
-beliefs operate in the opposite direction. An upward shift in the
-perception that effort is rewarded exerts a positive within effect
-(\(\beta\) = 0.071, \(p\) \textless{} .001), and a parallel increase in
-the perception that intelligence and ability are rewarded is likewise
-associated with higher market-justice preferences (\(\beta\) = 0.028,
-\(p\) \textless{} .05). Taken together, these results suggest that
-individuals who increasingly perceive meritocracy---whether through
-effort or talent---tend to hold stronger market justice preferences.
-
-When examining the between-group effects (BE) in Model 5 and 6, which
-capture differences between individuals in the average of the main
-variables, a similar pattern emerges. Individuals who perceive higher
-levels of economic inequality tend to prefer less market justice
-(\(\beta\) = -0.042, \(p\) \textless{} .05). In Model 6, the
-meritocratic perception that effort is rewarded is positively associated
-with market justice preferences (\(\beta\) = 0.186, \(p\) \textless{}
-.001), whereas the perception that talent is rewarded shows a positive
-but non-significant coefficient (\(\beta\) = 0.040). Notably, once the
-meritocratic variables are included, the negative coefficient for
-perceived economic inequality remains but is no longer statistically
-significant.
-
-Model 7 adds the control variables. With the exception of the
-between-effect for perceived economic inequality---which becomes
-nonsignificant---the within- and between-effects of the principal
-predictors retain both their direction and statistical significance,
-confirming the robustness of the associations (see Supplementary
-Material for effects of control variables).
-
-\begin{table}
-
-\caption{\label{tbl-interactions1}Interactions for meritocracy,
-perceived economic inequality and market justice preferences}
-
-\centering{
-
-\begin{center}
-\scalebox{0.8}{
-\begin{tabular}{l c c}
-\hline
- & Model 8 & Model 9 \\
-\hline
-Intercept                                                & $1.558^{***}$ & $2.361^{***}$ \\
-                                                         & $(0.115)$     & $(0.287)$     \\
-Merit: Effort (WE) x Perception inequality (WE)          & $0.019$       &               \\
-                                                         & $(0.016)$     &               \\
-Merit: Talent (WE) x Perception inequality (WE)          & $-0.002$      &               \\
-                                                         & $(0.016)$     &               \\
-Merit: Effort (BE) x Perception inequality (BE)          &               & $0.016$       \\
-                                                         &               & $(0.045)$     \\
-Merit: Talent (BE) x Perception inequality (BE)          &               & $0.070$       \\
-                                                         &               & $(0.045)$     \\
-\hline
-Controls                                                 & Yes           & Yes           \\
-BIC                                                      & $20879.829$   & $20819.684$   \\
-Num. obs.                                                & $8643$        & $8643$        \\
-Num. groups: idencuesta                                  & $1687$        & $1687$        \\
-Var: idencuesta (Intercept)                              & $0.147$       & $0.141$       \\
-Var: idencuesta merit\_effort\_cwc                       & $0.016$       & $$            \\
-Var: idencuesta merit\_talent\_cwc                       & $0.018$       & $$            \\
-Var: idencuesta perc\_inequality\_cwc                    & $0.000$       & $$            \\
-Cov: idencuesta (Intercept) merit\_effort\_cwc           & $-0.005$      & $$            \\
-Cov: idencuesta (Intercept) merit\_talent\_cwc           & $0.015$       & $$            \\
-Cov: idencuesta (Intercept) perc\_inequality\_cwc        & $-0.006$      & $$            \\
-Cov: idencuesta merit\_effort\_cwc merit\_talent\_cwc    & $-0.003$      & $$            \\
-Cov: idencuesta merit\_effort\_cwc perc\_inequality\_cwc & $0.000$       & $$            \\
-Cov: idencuesta merit\_talent\_cwc perc\_inequality\_cwc & $0.001$       & $$            \\
-Var: Residual                                            & $0.503$       & $0.528$       \\
-\hline
-\multicolumn{3}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
-\end{tabular}
-}
-\caption{}
-\label{table:coefficients}
-\end{center}
-
-}
-
-\end{table}%
-
-Table~\ref{tbl-interactions1} reports whether perceived economic
-inequality moderates the impact of meritocratic perceptions on
-market-justice preferences. In neither the within-person specification
-(Model 8) nor the between-person specification (Model 9) do the
-interaction terms reach statistical significance, indicating that higher
-perceptions of economic inequality do not meaningfully alter the effect
-of meritocracy on support for market allocation of social services.
-
-\begin{figure}[H]
-
-\caption{\label{fig-growth}Time varying within effects of meritocratic
-perceptions on market justice preferences}
-
-\centering{
-
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-growth-1.pdf}
-
-}
-
-\end{figure}%
-
-With respect to the temporal dynamics of the key predictors,
-Figure~\ref{fig-growth} illustrates the interaction between survey wave
-and the within-person effects of the meritocracy variables on
-market-justice preferences (complete models appear in the Supplementary
-Material). As time advances, the positive within-person effect of
-perceiving that effort is rewarded (panel A) or that intelligence is
-rewarded (panel B) progressively attenuates. Put differently, the uplift
-in market justice preferences associated with a strong meritocratic
-perception becomes smaller in each successive wave.
-
-\section{Discussion}\label{discussion}
-
-\section{Conclusion}\label{conclusion}
-
-\section{References}\label{references}
-
-\phantomsection\label{refs}
-\begin{CSLReferences}{1}{0}
-\bibitem[\citeproctext]{ref-akyelken_urban_2020}
-Akyelken, N. (2020). Urban conceptions of economic inequalities.
-\emph{Regional Studies}, \emph{54}(6), 863--872.
-\url{https://doi.org/10.1080/00343404.2020.1732902}
-
-\bibitem[\citeproctext]{ref-auspurg_why_2017}
-Auspurg, K., Hinz, T., \& Sauer, C. (2017). Why {Should Women Get Less}?
-{Evidence} on the {Gender Pay Gap} from {Multifactorial Survey
-Experiments}. \emph{American Sociological Review}, \emph{82}(1),
-179--210. \url{https://doi.org/10.1177/0003122416683393}
-
-\bibitem[\citeproctext]{ref-bates_fitting_2015}
-Bates, D., Mächler, M., Bolker, B., \& Walker, S. (2015). Fitting linear
-mixed-effects models using {lme4}. \emph{Journal of Statistical
-Software}, \emph{67}(1), 1--48.
-\url{https://doi.org/10.18637/jss.v067.i01}
-
-\bibitem[\citeproctext]{ref-batruch_belief_2023}
-Batruch, A., Jetten, J., Van De Werfhorst, H., Darnon, C., \& Butera, F.
-(2023). Belief in {School Meritocracy} and the {Legitimization} of
-{Social} and {Income Inequality}. \emph{Social Psychological and
-Personality Science}, \emph{14}(5), 621--635.
-\url{https://doi.org/10.1177/19485506221111017}
-
-\bibitem[\citeproctext]{ref-boccardo_30_2020}
-Boccardo, G. (2020). \emph{30 a{ñ}os de privatizaciones en {Chile}: Lo
-que la pandemia revel{ó}} (Nodo XXI). Santiago.
-
-\bibitem[\citeproctext]{ref-busemeyer_positive_2021}
-Busemeyer, M. R., Abrassart, A., \& Nezi, R. (2021). Beyond {Positive}
-and {Negative}: {New Perspectives} on {Feedback Effects} in {Public
-Opinion} on the {Welfare State}. \emph{British Journal of Political
-Science}, \emph{51}(1), 137--162.
-\url{https://doi.org/10.1017/S0007123418000534}
-
-\bibitem[\citeproctext]{ref-castillo_cual_2009}
-Castillo, J. C. (2009). {{\textquestiondown}Cu{á}l es la brecha salarial
-justa? Opini{ó}n p{ú}blica y legitimaci{ó}n de la desigualdad en Chile}.
-\emph{Estudios P{ú}blicos}, (113).
-
-\bibitem[\citeproctext]{ref-Castillo2011}
-Castillo, J. C. (2011). Legitimacy of {Inequality} in a {Highly Unequal
-Context}: {Evidence} from the {Chilean Case}. \emph{Social Justice
-Research}, \emph{24}(4), 314--340.
-\url{https://doi.org/10.1007/s11211-011-0144-5}
-
-\bibitem[\citeproctext]{ref-castillo_perception_2022}
-Castillo, J. C., García-Castro, J.-D., \& Venegas, M. (2022). Perception
-of economic inequality: Concepts, associated factors and prospects of a
-burgeoning research agenda. \emph{International Journal of Social
-Psychology}, \emph{37}(1), 180--207.
-\url{https://doi.org/10.1080/02134748.2021.2009275}
-
-\bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
-Castillo, J. C., Iturra, J., Maldonado, L., Atria, J., \& Meneses, F.
-(2023). A {Multidimensional Approach} for {Measuring Meritocratic
-Beliefs}: {Advantages}, {Limitations} and {Alternatives} to the {ISSP
-Social Inequality Survey}. \emph{International Journal of Sociology},
-1--25. \url{https://doi.org/10.1080/00207659.2023.2274712}
-
-\bibitem[\citeproctext]{ref-castillo_percepcion_2019}
-Castillo, J. C., Miranda, D., \& Carrasco, D. (2012). Percepci{ó}n de
-{Desigualdad Econ{ó}mica} en {Chile}: {Medici{ó}n}, {Diferencias} y
-{Determinantes}. \emph{Psykhe (Santiago)}, \emph{21}(1), 99--114.
-\url{https://doi.org/10.4067/S0718-22282012000100007}
-
-\bibitem[\citeproctext]{ref-castillo_socialization_2024}
-Castillo, J. C., Salgado, M., Carrasco, K., \& Laffert, A. (2024). The
-{Socialization} of {Meritocracy} and {Market Justice Preferences} at
-{School}. \emph{Societies}, \emph{14}(11), 214.
-\url{https://doi.org/10.3390/soc14110214}
-
-\bibitem[\citeproctext]{ref-chancel_world_2022}
-Chancel, L., Piketty, T., Saez, E., \& Zucman, G. (2022). World
-inequality report 2022.
-https://bibliotecadigital.ccb.org.co/handle/11520/27510.
-
-\bibitem[\citeproctext]{ref-davis_principles_2001}
-Davis, K., \& Moore, W. E. (2001). Some {Principles} of
-{Stratification}. In \emph{Social {Stratification}, {Class}, {Race}, and
-{Gender} in {Sociological Perspective}, {Second Edition}} (2nd ed.).
-Routledge.
-
-\bibitem[\citeproctext]{ref-easterbrook_social_2021}
-Easterbrook, M. J. (2021). \emph{The social psychology of economic
-inequality} (43rd ed., Vol. 2021). UNU-WIDER.
-\url{https://doi.org/10.35188/UNU-WIDER/2021/981-5}
-
-\bibitem[\citeproctext]{ref-elsoc_estudio_2022}
-ELSOC, S. T. (2022). Estudio {Longitudinal Social} de {Chile}. Harvard
-Dataverse. \url{https://doi.org/10.7910/dvn/0kirbj}
-
-\bibitem[\citeproctext]{ref-engelhardt_what_2018}
-Engelhardt, C., \& Wagener, A. (2018). What do {Germans} think and know
-about income inequality? {A} survey experiment. \emph{Socio-Economic
-Review}, \emph{16}(4), 743--767.
-\url{https://doi.org/10.1093/ser/mwx036}
-
-\bibitem[\citeproctext]{ref-flores_top_2020}
-Flores, I., Sanhueza, C., Atria, J., \& Mayer, R. (2020). Top {Incomes}
-in {Chile}: {A Historical Perspective} on {Income Inequality},
-1964--2017. \emph{Review of Income and Wealth}, \emph{66}(4), 850--874.
-\url{https://doi.org/10.1111/roiw.12441}
-
-\bibitem[\citeproctext]{ref-garcia-castro_perceiving_2020}
-García-Castro, J. D., Rodríguez-Bailón, R., \& Willis, G. B. (2020).
-Perceiving economic inequality in everyday life decreases tolerance to
-inequality. \emph{Journal of Experimental Social Psychology}, \emph{90},
-104019. \url{https://doi.org/10.1016/j.jesp.2020.104019}
-
-\bibitem[\citeproctext]{ref-garcia-sanchez_creencias_2022}
-García-Sánchez, E., \& De Carvalho, S. (2022). Las creencias que
-justifican la desigualdad moderan la relaci{ó}n entre el estatus
-socioecon{ó}mico y el apoyo a la redistribuci{ó}n. \emph{Revista
-Internacional de Sociolog{í}a}, \emph{80}(3), e210.
-\url{https://doi.org/10.3989/ris.2022.80.3.21.29}
-
-\bibitem[\citeproctext]{ref-garcia-sanchez_attitudes_2020}
-García-Sánchez, E., Osborne, D., Willis, G. B., \& Rodríguez-Bailón, R.
-(2020). Attitudes towards redistribution and the interplay between
-perceptions and beliefs about inequality. \emph{British Journal of
-Social Psychology}, \emph{59}(1), 111--136.
-\url{https://doi.org/10.1111/bjso.12326}
-
-\bibitem[\citeproctext]{ref-garcia-sanchez_vicious_2019}
-García-Sánchez, E., Van Der Toorn, J., Rodríguez-Bailón, R., \& Willis,
-G. B. (2019). The {Vicious Cycle} of {Economic Inequality}: {The Role}
-of {Ideology} in {Shaping} the {Relationship Between} {``{What Is}''}
-and {``{What Ought} to {Be}''} in 41 {Countries}. \emph{Social
-Psychological and Personality Science}, \emph{10}(8), 991--1001.
-\url{https://doi.org/10.1177/1948550618811500}
-
-\bibitem[\citeproctext]{ref-garcia-sanchez_perceptions_2018}
-García-Sánchez, E., Willis, G. B., Rodríguez-Bailón, R., Palacio Sañudo,
-J., David Polo, J., \& Rentería Pérez, E. (2018). Perceptions of
-{Economic Inequality} and {Support} for {Redistribution}: {The} role of
-{Existential} and {Utopian Standards}. \emph{Social Justice Research},
-\emph{31}(4), 335--354. \url{https://doi.org/10.1007/s11211-018-0317-6}
-
-\bibitem[\citeproctext]{ref-gijsberts_thelegitimation_1999}
-Gijsberts, M. (1999). \emph{{Thelegitimation of inequality in state-
-socialist and market societies, 1987 - 1996}}. Amsterdam: Thela Thesis.
-
-\bibitem[\citeproctext]{ref-gimpelson_misperceiving_2018}
-Gimpelson, V., \& Treisman, D. (2018). Misperceiving inequality.
-\emph{Economics \& Politics}, \emph{30}(1), 27--54.
-\url{https://doi.org/10.1111/ecpo.12103}
-
-\bibitem[\citeproctext]{ref-hadler_why_2005}
-Hadler, M. (2005). Why {Do People Accept Different Income Ratios}?: {A
-Multi-level Comparison} of {Thirty Countries}. \emph{Acta Sociologica},
-\emph{48}(2), 131--154. \url{https://doi.org/10.1177/0001699305053768}
-
-\bibitem[\citeproctext]{ref-hoxMultilevelAnalysisTechniques2017a}
-Hox, J. J., Moerbeek, M., \& Van de Schoot, R. (2017). \emph{Multilevel
-{Analysis}: {Techniques} and {Applications}}.
-
-\bibitem[\citeproctext]{ref-hoyt_mindsets_2023}
-Hoyt, C. L., Burnette, J. L., Billingsley, J., Becker, W., \& Babij, A.
-D. (2023). Mindsets of poverty: {Implications} for redistributive policy
-support. \emph{Analyses of Social Issues and Public Policy},
-\emph{23}(3), 668--693. \url{https://doi.org/10.1111/asap.12367}
-
-\bibitem[\citeproctext]{ref-immergut_it_2020}
-Immergut, E. M., \& Schneider, S. M. (2020). Is it unfair for the
-affluent to be able to purchase {``better''} healthcare? {Existential}
-standards and institutional norms in healthcare attitudes across 28
-countries. \emph{Social Science \& Medicine}, \emph{267}, 113146.
-\url{https://doi.org/10.1016/j.socscimed.2020.113146}
-
-\bibitem[\citeproctext]{ref-janmaat_subjective_2013}
-Janmaat, J. G. (2013). Subjective inequality: {A} review of
-international comparative studies on people's views about inequality.
-\emph{Archives Europeennes de Sociologie}, \emph{54}(3), 357--389.
-\url{https://doi.org/10.1017/S0003975613000209}
-
-\bibitem[\citeproctext]{ref-kluegel_social_1995a}
-Kluegel, J. R., Mason, D. S., \& Wegener, B. (Eds.). (1995).
-\emph{Social {Justice} and {Political Change}: {Public Opinion} in
-{Capitalist} and {Post-Communist States}} (1st ed.). Routledge.
-
-\bibitem[\citeproctext]{ref-kluegel_legitimation_1999}
-Kluegel, J. R., Mason, D. S., \& Wegener, B. (1999). The {Legitimation}
-of {Capitalism} in the {Postcommunist Transition}: {Public Opinion}
-about {Market Justice}, 1991-1996. \emph{European Sociological Review},
-\emph{15}(3), 251--283. Retrieved from
-\url{https://www.jstor.org/stable/522731}
-
-\bibitem[\citeproctext]{ref-kluegel_beliefs_1981}
-Kluegel, J. R., \& Smith, E. R. (1981). Beliefs {About Stratification}.
-\emph{Annual Review of Sociology}, 29--56.
-
-\bibitem[\citeproctext]{ref-knell_perceptions_2020}
-Knell, M., \& Stix, H. (2020). Perceptions of inequality. \emph{European
-Journal of Political Economy}, \emph{65}, 101927.
-\url{https://doi.org/10.1016/j.ejpoleco.2020.101927}
-
-\bibitem[\citeproctext]{ref-koos_moral_2019}
-Koos, S., \& Sachweh, P. (2019). The moral economies of market
-societies: Popular attitudes towards market competition, redistribution
-and reciprocity in comparative perspective. \emph{Socio-Economic
-Review}, \emph{17}(4), 793--821.
-\url{https://doi.org/10.1093/ser/mwx045}
-
-\bibitem[\citeproctext]{ref-kuhn_eye_2011}
-Kuhn, A. (2011). In the eye of the beholder: {Subjective} inequality
-measures and individuals' assessment of market justice. \emph{European
-Journal of Political Economy}, \emph{27}(4), 625--641.
-\url{https://doi.org/10.1016/j.ejpoleco.2011.06.002}
-
-\bibitem[\citeproctext]{ref-lane_market_1986}
-Lane, R. E. (1986). Market {Justice}, {Political Justice}.
-\emph{American Political Science Review}, \emph{80}(2), 383--402.
-\url{https://doi.org/10.2307/1958264}
-
-\bibitem[\citeproctext]{ref-lee_fairness_2023}
-Lee, J.-S., \& Stacey, M. (2023). Fairness perceptions of educational
-inequality: The effects of self-interest and neoliberal orientations.
-\emph{The Australian Educational Researcher}.
-\url{https://doi.org/10.1007/s13384-023-00636-6}
-
-\bibitem[\citeproctext]{ref-lindh_public_2015}
-Lindh, A. (2015). Public {Opinion} against {Markets}? {Attitudes}
-towards {Market Distribution} of {Social Services} -- {A Comparison} of
-17 {Countries}. \emph{Social Policy \& Administration}, \emph{49}(7),
-887--910. \url{https://doi.org/10.1111/spol.12105}
-
-\bibitem[\citeproctext]{ref-mijs_stratified_2016}
-Mijs, J. (2016). Stratified {Failure}: {Educational Stratification} and
-{Students}' {Attributions} of {Their Mathematics Performance} in 24
-{Countries}. \emph{Sociology of Education}, \emph{89}(2), 137--153.
-\url{https://doi.org/10.1177/0038040716636434}
-
-\bibitem[\citeproctext]{ref-mijs_paradox_2021}
-Mijs, J. (2021). The paradox of inequality: Income inequality and belief
-in meritocracy go hand in hand. \emph{Socio-Economic Review},
-\emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}
-
-\bibitem[\citeproctext]{ref-pedersen_attitudes_2019}
-Pedersen, R. T., \& Mutz, D. C. (2019). Attitudes {Toward Economic
-Inequality}: {The Illusory Agreement}. \emph{Political Science Research
-and Methods}, \emph{7}(04), 835--851.
-\url{https://doi.org/10.1017/psrm.2018.18}
-
-\bibitem[\citeproctext]{ref-reynolds_perceptions_2014}
-Reynolds, J., \& Xian, H. (2014). Perceptions of meritocracy in the land
-of opportunity. \emph{Research in Social Stratification and Mobility},
-\emph{36}, 121--137. \url{https://doi.org/10.1016/j.rssm.2014.03.001}
-
-\bibitem[\citeproctext]{ref-sandel_tyranny_2020}
-Sandel, M. J. (2020). \emph{The tyranny of merit: {What}'s become of the
-common good?} (First edition). New York: {Farrar, Straus and Giroux}.
-
-\bibitem[\citeproctext]{ref-schneider_poverty_2015}
-Schneider, S. M., \& Castillo, J. C. (2015). Poverty {Attributions} and
-the {Perceived Justice} of {Income Inequality} : {A Comparison} of
-{East} and {West Germany}.
-\url{https://doi.org/10.1177/0190272515589298}
-
-\bibitem[\citeproctext]{ref-schroder_income_2017}
-Schröder, M. (2017). Is {Income Inequality Related} to {Tolerance} for
-{Inequality}? \emph{Social Justice Research}, \emph{30}(1), 23--47.
-\url{https://doi.org/10.1007/s11211-016-0276-8}
-
-\bibitem[\citeproctext]{ref-singer_applied_2009}
-Singer, J. D., \& Willett, J. B. (2009). \emph{Applied longitudinal data
-analysis: Modeling change and event occurence}. New York: Oxford
-University Press, Incorporated.
-
-\bibitem[\citeproctext]{ref-svallfors_political_2007}
-Svallfors, S. (Ed.). (2007). \emph{The {Political Sociology} of the
-{Welfare State}: {Institutions}, {Social Cleavages}, and {Orientations}}
-(1st ed.). Stanford University Press.
-\url{https://doi.org/10.2307/j.ctvr0qv0q}
-
-\bibitem[\citeproctext]{ref-tejero-peregrina_perceived_2025}
-Tejero-Peregrina, L., Willis, G., Sánchez-Rodríguez, Á., \&
-Rodríguez-Bailón, R. (2025). From {Perceived Economic Inequality} to
-{Support} for {Redistribution}: {The Role} of {Meritocracy Perception}.
-\emph{International Review of Social Psychology}, \emph{38}(1), 4.
-\url{https://doi.org/10.5334/irsp.1013}
-
-\bibitem[\citeproctext]{ref-trump_income_2018}
-Trump, K.-S. (2018). Income {Inequality Influences Perceptions} of
-{Legitimate Income Differences}. \emph{British Journal of Political
-Science}, \emph{48}(4), 929--952.
-\url{https://doi.org/10.1017/S0007123416000326}
-
-\bibitem[\citeproctext]{ref-vondemknesebeck_are_2016}
-Von Dem Knesebeck, O., Vonneilich, N., \& Kim, T. J. (2016). Are health
-care inequalities unfair? {A} study on public attitudes in 23 countries.
-\emph{International Journal for Equity in Health}, \emph{15}(1), 61.
-\url{https://doi.org/10.1186/s12939-016-0350-8}
-
-\bibitem[\citeproctext]{ref-willis_legitimacy_2015}
-Willis, G. B., Rodríguez-Bailón, R., López-Rodríguez, L., \&
-García-Sánchez, E. (2015). Legitimacy {Moderates} the {Relation Between
-Perceived} and {Ideal Economic Inequalities}. \emph{Social Justice
-Research}, \emph{28}(4), 493--508.
-\url{https://doi.org/10.1007/s11211-015-0253-7}
-
-\bibitem[\citeproctext]{ref-wilson_role_2003}
-Wilson, C. (2003). The {Role} of a {Merit Principle} in {Distributive
-Justice}. \emph{The Journal of Ethics}, \emph{7}(3), 277--314.
-\url{https://doi.org/10.1023/A:1024667228488}
-
-\bibitem[\citeproctext]{ref-young_rise_1962}
-Young, M. (1962). \emph{The rise of the meritocracy}. Baltimore: Penguin
-Books.
-
-\end{CSLReferences}
+Access to the data, code, and analysis presented in the study are openly
+available in
+\url{https://github.com/justicia-distributiva/market-justice-panel-elsoc}.
 

### 2025-05-22 paper.qmd
Up title page

--- 2025-05-22 paper.qmd
@@
 bibliography: "input/bib/merit-edjust.bib"
-csl: "input/bib/apa6.csl"
+# csl: "input/bib/apa6.csl"
+csl: "input/bib/harvard.csl"
 author:

### 2025-05-22 02-antecedents.qmd
type changes

@@
 It is possible to trace the concept to the understanding of Lane [-@lane_market_1986], who makes a contrast between market justice and political justice.
-The author defines market justice as a system of "earned deserts," whereby individuals are seen as deserving of a determined distributive outcome due to their effort and skills.
+The author defines market justice as a system of "earned deserts", whereby individuals are seen as deserving of a determined distributive outcome due to their effort and skills.
 In contrast, political justice emphasizes principles of equality and need, which are often represented by the welfare state action through social policies.
@@
 These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they endorse meritocratic principles.
-This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025] Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity.
+This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025].
+Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity.
 Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_percepcion_2019; @castillo_socialization_2024].
@@
 
-First, consistent with previous findings, we expect that a lower perception of economic inequality will be associated with stronger market justice preferences.
+First, consistent with previous findings, we expect that a higher perception of economic inequality will be associated with lower market justice preferences.
 When individuals perceive smaller income gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing the acceptance of outcomes based on competition and ability to pay.

### 2025-05-22 04-analisis.qmd
type changes

@@
                summarytools,
-               interactions)
+               interactions,
+               cowplot)
 
@@
        title = "(B) People are rewarded for their intelligence",
-       caption = "Source: own elaboration with data from ELSOC 2016-2023 (n = 8460). Based in Model 10 and Model 11 in Table 4. Confidence intervals at 95%") +
+       caption = "Source: own elaboration with data from ELSOC 2016-2023 (n = 8460). Based in Model 10 and Model 11 in Table 4.
+Confidence intervals at 95%") +
   theme_ggdist()

### 2025-05-22 cover_letter.qmd
Up nombres y cover letter

@@
 
-Julio Iturra, Bremen International Graduate School of Social Sciences, Bremen, Germany.
+Julio Iturra-Sanhueza, Bremen International Graduate School of Social Sciences, Bremen, Germany.
 ORCID: 0000-0003-1993-3479
@@
 
-The authors recognize the support of the National Research and Development Agency (ANID) through the grant from the National Fund for Scientific and Technological Development (FONDECYT) N° 1210847; from the Center for Conflict and Social Cohesion Studies (COES) grant Financing Fund for Research Centers in Priority Areas (FONDAP) N° 1523A0005; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046.
+The authors recognize the support of the National Research and Development Agency (ANID) through the grant from the National Fund for Scientific and Technological Development (FONDECYT) N° 1250518; from the Center for Conflict and Social Cohesion Studies (COES) grant Financing Fund for Research Centers in Priority Areas (FONDAP) N° 1523A0005; and Millenium Nucleus on Digital Inequalities and Opportunities (NUDOS) NCS2022_046.
 
@@
 
-This research received funding from ANID/FONDECYT grant 1210847 and the Center for Conflict and Social Cohesion Studies (COES) through ANID/FONDAP grant 1513009.
+This research received funding from ANID/FONDECYT grant 1250518 and the Center for Conflict and Social Cohesion Studies (COES) through ANID/FONDAP grant 1523A0005.
 

### 2025-05-22 paper.qmd
Up nombres y cover letter

--- 2025-05-22 paper.qmd
@@
       - ref: 2
-  - name: Julio Iturra
+  - name: Julio Iturra-Sanhueza
     orcid: 0000-0003-1993-3479

### 2025-05-22 paper.tex
Up nombres y cover letter

--- 2025-05-22 paper.tex
@@
               }
-        \author{Julio Iturra}
+        \author{Julio Iturra-Sanhueza}
             \affil{%
@@
 who makes a contrast between market justice and political justice. The
-author defines market justice as a system of ``earned deserts,'' whereby
+author defines market justice as a system of ``earned deserts'', whereby
 individuals are seen as deserving of a determined distributive outcome
@@
 the development of welfare institutions, in line with the
-policy-feedbacks literature {[}campbell\_policy\_2012;
-pierson\_when\_1993{]}.
+policy-feedbacks literature
+(\citeproc{ref-campbell_policy_2012}{Campbell, 2012};
+\citeproc{ref-pierson_when_1993}{Pierson, 1993}).
 
@@
 (\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
-2025}) Meritocratic beliefs serve as symbolic justifications for unequal
-outcomes, particularly when access is stratified by income or social
-opportunity. Prior studies have shown that individuals who perceive
-higher levels of meritocracy tend to express stronger support for
-unequal distributions that reflect market outcomes
+2025}). Meritocratic beliefs serve as symbolic justifications for
+unequal outcomes, particularly when access is stratified by income or
+social opportunity. Prior studies have shown that individuals who
+perceive higher levels of meritocracy tend to express stronger support
+for unequal distributions that reflect market outcomes
 (\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
@@
 
-First, consistent with previous findings, we expect that a lower
-perception of economic inequality will be associated with stronger
-market justice preferences. When individuals perceive smaller income
-gaps, they are more likely to view market mechanisms as fair and
-legitimate, reinforcing the acceptance of outcomes based on competition
-and ability to pay. Conversely, a heightened perception of inequality
-may erode confidence in market fairness, weakening support for
-market-based distribution. This relationship is particularly relevant in
-the context of Chile, where the neoliberal economic model has been a
-dominant force in shaping public attitudes toward inequality and
-justice.
+First, consistent with previous findings, we expect that a higher
+perception of economic inequality will be associated with lower market
+justice preferences. When individuals perceive smaller income gaps, they
+are more likely to view market mechanisms as fair and legitimate,
+reinforcing the acceptance of outcomes based on competition and ability
+to pay. Conversely, a heightened perception of inequality may erode
+confidence in market fairness, weakening support for market-based
+distribution. This relationship is particularly relevant in the context
+of Chile, where the neoliberal economic model has been a dominant force
+in shaping public attitudes toward inequality and justice.
 
@@
 
-~
-
-Model 1
-
-Model 2
-
-Model 3
-
-Model 4
-
-Model 5
-
-Model 6
-
-Model 7
-
-Intercept
-
-1.968***
-
-2.015***
-
-2.030***
-
-2.028***
-
-2.169***
-
-1.456***
-
-1.547***
-
-~
-
-(0.021)
-
-(0.035)
-
-(0.035)
-
-(0.035)
-
-(0.070)
-
-(0.097)
-
-(0.115)
-
-Wave (Ref.= 2016)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2017
-
--0.167***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.027)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2018
-
--0.025
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.027)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2019
-
--0.046
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.027)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2022
-
-0.285***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.028)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~~~~~Wave 2023
-
-0.284***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.027)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Wave
-
-~
-
--0.121***
-
--0.125***
-
--0.128***
-
--0.128***
-
--0.128***
-
--0.129***
-
-~
-
-~
-
-(0.022)
-
-(0.022)
-
-(0.022)
-
-(0.022)
-
-(0.022)
-
-(0.022)
-
-Wave\^{}2
-
-~
-
-0.028***
-
-0.028***
-
-0.029***
-
-0.029***
-
-0.029***
-
-0.029***
-
-~
-
-~
-
-(0.003)
-
-(0.003)
-
-(0.003)
-
-(0.003)
-
-(0.003)
-
-(0.003)
-
-Perception inequality (WE)
-
-~
-
-~
-
--0.039***
-
--0.034***
-
--0.034***
-
--0.034***
-
--0.034***
-
-~
-
-~
-
-~
-
-(0.010)
-
-(0.010)
-
-(0.010)
-
-(0.010)
-
-(0.010)
-
-Merit: Effort (WE)
-
-~
-
-~
-
-~
-
-0.070***
-
-0.070***
-
-0.069***
-
-0.069***
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-Merit: Talent (WE)
-
-~
-
-~
-
-~
-
-0.027*
-
-0.027*
-
-0.027*
-
-0.027*
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-Perception inequality (BE)
-
-~
-
-~
-
-~
-
-~
-
--0.042*
-
--0.006
-
--0.033
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.018)
-
-(0.018)
-
-(0.018)
-
-Merit: Effort (BE)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.186***
-
-0.178***
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.032)
-
-(0.031)
-
-Merit: Talent (BE)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.040
-
-0.019
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.032)
-
-(0.031)
-
-Controls
-
-No
-
-No
-
-No
-
-No
-
-No
-
-No
-
-Yes
-
-BIC
-
-20780.594
-
-20780.617
-
-20781.113
-
-20726.753
-
-20736.627
-
-20655.172
-
-20749.047
-
-Numb. obs.
-
-8643
-
-8643
-
-8643
-
-8643
-
-8643
-
-8643
-
-8643
-
-Num. groups: individuals
-
-1687
-
-1687
-
-1687
-
-1687
-
-1687
-
-1687
-
-1687
-
-Var: individuals (Intercept)
-
-0.172
-
-0.223
-
-0.219
-
-0.209
-
-0.207
-
-0.182
-
-0.171
-
-Var: Residual
-
-0.530
-
-0.499
-
-0.498
-
-0.495
-
-0.495
-
-0.495
-
-0.495
-
-Var: individuals, wave
-
-~
-
-0.011
-
-0.010
-
-0.009
-
-0.009
-
-0.009
-
-0.009
-
-Cov: individuals (Intercept), wave
-
-~
-
--0.025
-
--0.024
-
--0.021
-
--0.020
-
--0.019
-
--0.019
-
-Note: Cells contain regression coefficients with standard errors in
-parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
-\textless{} 0.05.
+\caption{}
+\begin{center}
+\scalebox{0.8}{
+\begin{tabular}{l c c c c c c c}
+\hline
+ & Model 1 & Model 2 & Model 3 & Model 4 & Model 5 & Model 6 & Model 7 \\
+\hline
+Intercept                          & $1.968^{***}$  & $2.015^{***}$  & $2.030^{***}$  & $2.028^{***}$  & $2.169^{***}$  & $1.456^{***}$  & $1.547^{***}$  \\
+                                   & $(0.021)$      & $(0.035)$      & $(0.035)$      & $(0.035)$      & $(0.070)$      & $(0.097)$      & $(0.115)$      \\
+Wave (Ref.= 2016)                  &                &                &                &                &                &                &                \\
+                                   &                &                &                &                &                &                &                \\
+\quad Wave 2017                    & $-0.167^{***}$ &                &                &                &                &                &                \\
+                                   & $(0.027)$      &                &                &                &                &                &                \\
+\quad Wave 2018                    & $-0.025$       &                &                &                &                &                &                \\
+                                   & $(0.027)$      &                &                &                &                &                &                \\
+\quad Wave 2019                    & $-0.046$       &                &                &                &                &                &                \\
+                                   & $(0.027)$      &                &                &                &                &                &                \\
+\quad Wave 2022                    & $0.285^{***}$  &                &                &                &                &                &                \\
+                                   & $(0.028)$      &                &                &                &                &                &                \\
+\quad Wave 2023                    & $0.284^{***}$  &                &                &                &                &                &                \\
+                                   & $(0.027)$      &                &                &                &                &                &                \\
+Wave                               &                & $-0.121^{***}$ & $-0.125^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.129^{***}$ \\
+                                   &                & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      \\
+Wave$^2$                           &                & $0.028^{***}$  & $0.028^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  \\
+                                   &                & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      \\
+Perception inequality (WE)         &                &                & $-0.039^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ \\
+                                   &                &                & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
+Merit: Effort (WE)                 &                &                &                & $0.070^{***}$  & $0.070^{***}$  & $0.069^{***}$  & $0.069^{***}$  \\
+                                   &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
+Merit: Talent (WE)                 &                &                &                & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    \\
+                                   &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
+Perception inequality (BE)         &                &                &                &                & $-0.042^{*}$   & $-0.006$       & $-0.033$       \\
+                                   &                &                &                &                & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
+Merit: Effort (BE)                 &                &                &                &                &                & $0.186^{***}$  & $0.178^{***}$  \\
+                                   &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
+Merit: Talent (BE)                 &                &                &                &                &                & $0.040$        & $0.019$        \\
+                                   &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
+\hline
+Controls                           & No             & No             & No             & No             & No             & No             & Yes            \\
+BIC                                & $20780.594$    & $20780.617$    & $20781.113$    & $20726.753$    & $20736.627$    & $20655.172$    & $20749.047$    \\
+Numb. obs.                         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         \\
+Num. groups: individuals           & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         \\
+Var: individuals (Intercept)       & $0.172$        & $0.223$        & $0.219$        & $0.209$        & $0.207$        & $0.182$        & $0.171$        \\
+Var: Residual                      & $0.530$        & $0.499$        & $0.498$        & $0.495$        & $0.495$        & $0.495$        & $0.495$        \\
+Var: individuals, wave             & $$             & $0.011$        & $0.010$        & $0.009$        & $0.009$        & $0.009$        & $0.009$        \\
+Cov: individuals (Intercept), wave & $$             & $-0.025$       & $-0.024$       & $-0.021$       & $-0.020$       & $-0.019$       & $-0.019$       \\
+\hline
+\multicolumn{8}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$.}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
 
@@
 
-~
-
-Model 8
-
-Model 9
-
-Model 10
-
-Model 11
-
-Intercept
-
-1.561***
-
-1.555***
-
-2.187***
-
-2.340***
-
-~
-
-(0.115)
-
-(0.115)
-
-(0.265)
-
-(0.281)
-
-Perception inequality (WE)
-
--0.035***
-
--0.035***
-
--0.035***
-
--0.035***
-
-~
-
-(0.010)
-
-(0.010)
-
-(0.010)
-
-(0.010)
-
-Merit: Effort (WE)
-
-0.072***
-
-0.074***
-
-0.073***
-
-0.073***
-
-~
-
-(0.013)
-
-(0.012)
-
-(0.012)
-
-(0.012)
-
-Merit: Talent (WE)
-
-0.031*
-
-0.031*
-
-0.030*
-
-0.030*
-
-~
-
-(0.012)
-
-(0.013)
-
-(0.012)
-
-(0.012)
-
-Perception inequality (BE)
-
--0.032
-
--0.033
-
--0.213**
-
--0.255***
-
-~
-
-(0.018)
-
-(0.018)
-
-(0.073)
-
-(0.077)
-
-Merit: Effort (BE)
-
-0.177***
-
-0.173***
-
--0.062
-
-0.179***
-
-~
-
-(0.032)
-
-(0.031)
-
-(0.097)
-
-(0.032)
-
-Merit: Talent (BE)
-
-0.015
-
-0.021
-
-0.014
-
--0.266**
-
-~
-
-(0.031)
-
-(0.031)
-
-(0.031)
-
-(0.099)
-
-Merit: Effort (WE) x Perception inequality (WE)
-
-0.017
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-~
-
-~
-
-~
-
-Merit: Talent (WE) x Perception inequality (WE)
-
-~
-
-0.011
-
-~
-
-~
-
-~
-
-~
-
-(0.012)
-
-~
-
-~
-
-Merit: Effort (BE) x Perception inequality (BE)
-
-~
-
-~
-
-0.072**
-
-~
-
-~
-
-~
-
-~
-
-(0.028)
-
-~
-
-Merit: Talent (BE) x Perception inequality (BE)
-
-~
-
-~
-
-~
-
-0.082**
-
-~
-
-~
-
-~
-
-~
-
-(0.027)
-
-Controls
-
-Yes
-
-Yes
-
-Yes
-
-Yes
-
-BIC
-
-20842.694
-
-20835.920
-
-20808.677
-
-20806.384
-
-Numb. obs.
-
-8643
-
-8643
-
-8643
-
-8643
-
-Num. groups: individuals
-
-1687
-
-1687
-
-1687
-
-1687
-
-Var: individuals (Intercept)
-
-0.145
-
-0.145
-
-0.141
-
-0.141
-
-Var: individuals, merit effort cwc
-
-0.020
-
-~
-
-~
-
-~
-
-Var: individuals, perception inequality cwc
-
-0.000
-
-0.000
-
-~
-
-~
-
-Cov: individuals (Intercept), merit effort cwc
-
-0.004
-
-~
-
-~
-
-~
-
-Cov: individuals (Intercept), perception inequality cwc
-
--0.006
-
--0.005
-
-~
-
-~
-
-Cov: individuals, merit effort cwc, perception inequality cwc
-
-0.000
-
-~
-
-~
-
-~
-
-Var: Residuals
-
-0.512
-
-0.511
-
-0.528
-
-0.528
-
-Var: individuals, merit talent cwc
-
-~
-
-0.020
-
-~
-
-~
-
-Cov: individuals (Intercept), merit talent cwc
-
-~
-
-0.012
-
-~
-
-~
-
-Cov: individuals, merit talent cwc, perception inequality cwc
-
-~
-
-0.001
-
-~
-
-~
-
-Note: Cells contain regression coefficients with standard errors in
-parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
-\textless{} 0.05. CWC = centered within group.
+\caption{}
+\begin{center}
+\scalebox{0.8}{
+\begin{tabular}{l c c c c}
+\hline
+ & Model 8 & Model 9 & Model 10 & Model 11 \\
+\hline
+Intercept                                                     & $1.561^{***}$  & $1.555^{***}$  & $2.187^{***}$  & $2.340^{***}$  \\
+                                                              & $(0.115)$      & $(0.115)$      & $(0.265)$      & $(0.281)$      \\
+Perception inequality (WE)                                    & $-0.035^{***}$ & $-0.035^{***}$ & $-0.035^{***}$ & $-0.035^{***}$ \\
+                                                              & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
+Merit: Effort (WE)                                            & $0.072^{***}$  & $0.074^{***}$  & $0.073^{***}$  & $0.073^{***}$  \\
+                                                              & $(0.013)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
+Merit: Talent (WE)                                            & $0.031^{*}$    & $0.031^{*}$    & $0.030^{*}$    & $0.030^{*}$    \\
+                                                              & $(0.012)$      & $(0.013)$      & $(0.012)$      & $(0.012)$      \\
+Perception inequality (BE)                                    & $-0.032$       & $-0.033$       & $-0.213^{**}$  & $-0.255^{***}$ \\
+                                                              & $(0.018)$      & $(0.018)$      & $(0.073)$      & $(0.077)$      \\
+Merit: Effort (BE)                                            & $0.177^{***}$  & $0.173^{***}$  & $-0.062$       & $0.179^{***}$  \\
+                                                              & $(0.032)$      & $(0.031)$      & $(0.097)$      & $(0.032)$      \\
+Merit: Talent (BE)                                            & $0.015$        & $0.021$        & $0.014$        & $-0.266^{**}$  \\
+                                                              & $(0.031)$      & $(0.031)$      & $(0.031)$      & $(0.099)$      \\
+Merit: Effort (WE) x Perception inequality (WE)               & $0.017$        &                &                &                \\
+                                                              & $(0.012)$      &                &                &                \\
+Merit: Talent (WE) x Perception inequality (WE)               &                & $0.011$        &                &                \\
+                                                              &                & $(0.012)$      &                &                \\
+Merit: Effort (BE) x Perception inequality (BE)               &                &                & $0.072^{**}$   &                \\
+                                                              &                &                & $(0.028)$      &                \\
+Merit: Talent (BE) x Perception inequality (BE)               &                &                &                & $0.082^{**}$   \\
+                                                              &                &                &                & $(0.027)$      \\
+\hline
+Controls                                                      & Yes            & Yes            & Yes            & Yes            \\
+BIC                                                           & $20842.694$    & $20835.920$    & $20808.677$    & $20806.384$    \\
+Numb. obs.                                                    & $8643$         & $8643$         & $8643$         & $8643$         \\
+Num. groups: individuals                                      & $1687$         & $1687$         & $1687$         & $1687$         \\
+Var: individuals (Intercept)                                  & $0.145$        & $0.145$        & $0.141$        & $0.141$        \\
+Var: individuals, merit effort cwc                            & $0.020$        & $$             & $$             & $$             \\
+Var: individuals, perception inequality cwc                   & $0.000$        & $0.000$        & $$             & $$             \\
+Cov: individuals (Intercept), merit effort cwc                & $0.004$        & $$             & $$             & $$             \\
+Cov: individuals (Intercept), perception inequality cwc       & $-0.006$       & $-0.005$       & $$             & $$             \\
+Cov: individuals, merit effort cwc, perception inequality cwc & $0.000$        & $$             & $$             & $$             \\
+Var: Residuals                                                & $0.512$        & $0.511$        & $0.528$        & $0.528$        \\
+Var: individuals, merit talent cwc                            & $$             & $0.020$        & $$             & $$             \\
+Cov: individuals (Intercept), merit talent cwc                & $$             & $0.012$        & $$             & $$             \\
+Cov: individuals, merit talent cwc, perception inequality cwc & $$             & $0.001$        & $$             & $$             \\
+\hline
+\multicolumn{5}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$. CWC = centered within group.}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
 
@@
 
-~
-
-Model 12
-
-Model 13
-
-Model 14
-
-Model 15
-
-Model 16
-
-Model 17
-
-Intercept
-
-1.543***
-
-1.527***
-
-1.521***
-
-1.680***
-
-1.374***
-
-1.362***
-
-~
-
-(0.115)
-
-(0.114)
-
-(0.114)
-
-(0.139)
-
-(0.133)
-
-(0.134)
-
-Wave x Perception inequality (WE)
-
--0.004
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.006)
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Wave x Merit: effort (WE)
-
-~
-
--0.008
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.006)
-
-~
-
-~
-
-~
-
-~
-
-Wave x Merit: talent (WE)
-
-~
-
-~
-
--0.006
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.006)
-
-~
-
-~
-
-~
-
-Wave x Perception inequality (BE)
-
-~
-
-~
-
-~
-
-0.011
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.007)
-
-~
-
-~
-
-Wave x Merit: effort (BE)
-
-~
-
-~
-
-~
-
-~
-
--0.021*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.008)
-
-~
-
-Wave x Merit: talent (BE)
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.020*
-
-~
-
-~
-
-~
-
-~
-
-~
-
-~
-
-(0.008)
-
-Controls
-
-Yes
-
-Yes
-
-Yes
-
-Yes
-
-Yes
-
-Yes
-
-BIC
-
-20792.207
-
-20767.371
-
-20754.324
-
-20782.059
-
-20749.596
-
-20754.417
-
-Numb. obs.
-
-8643
-
-8643
-
-8643
-
-8643
-
-8643
-
-8643
-
-Num. groups: individuals
-
-1687
-
-1687
-
-1687
-
-1687
-
-1687
-
-1687
-
-Var: individuals (Intercept)
-
-0.176
-
-0.167
-
-0.168
-
-0.383
-
-0.254
-
-0.010
-
-Var: individuals, perception inequality cwc
-
-0.000
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Var: individuals, wave
-
-0.010
-
-0.009
-
-0.009
-
-0.009
-
-0.009
-
-0.009
-
-Cov: individuals (Intercept), perception inequality cwc
-
--0.007
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Cov: individuals (Intercept), wave
-
--0.021
-
--0.018
-
--0.018
-
--0.048
-
--0.003
-
-0.006
-
-Cov: individuals, perception inequality cwc, wave
-
-0.002
-
-~
-
-~
-
-~
-
-~
-
-~
-
-Var: Residuals
-
-0.495
-
-0.480
-
-0.479
-
-0.494
-
-0.494
-
-0.494
-
-Var: individuals, merit effort cwc
-
-~
-
-0.019
-
-~
-
-~
-
-~
-
-~
-
-Cov: individuals (Intercept), merit effort cwc
-
-~
-
-0.023
-
-~
-
-~
-
-~
-
-~
-
-Cov: individuals, merit effort cwc, wave
-
-~
-
--0.005
-
-~
-
-~
-
-~
-
-~
-
-Var: individuals, merit talent cwc
-
-~
-
-~
-
-0.019
-
-~
-
-~
-
-~
-
-Cov: individuals (Intercept), merit talent cwc
-
-~
-
-~
-
-0.034
-
-~
-
-~
-
-~
-
-Cov: individuals, merit talent cwc, wave
-
-~
-
-~
-
--0.006
-
-~
-
-~
-
-~
-
-Var: individuals, perception inequality mean
-
-~
-
-~
-
-~
-
-0.008
-
-~
-
-~
-
-Cov: individuals (Intercept), perception inequality mean
-
-~
-
-~
-
-~
-
--0.045
-
-~
-
-~
-
-Cov: individuals, perception inequality mean, wave
-
-~
-
-~
-
-~
-
-0.009
-
-~
-
-~
-
-Var: individuals, merit effort mean
-
-~
-
-~
-
-~
-
-~
-
-0.053
-
-~
-
-Cov: individuals (Intercept), merit effort mean
-
-~
-
-~
-
-~
-
-~
-
--0.089
-
-~
-
-Cov: individuals, merit effort mean, wave
-
-~
-
-~
-
-~
-
-~
-
--0.006
-
-~
-
-Var: individuals, merit talent mean
-
-~
-
-~
-
-~
-
-~
-
-~
-
-0.025
-
-Cov: individuals (Intercept), merit talent mean
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.007
-
-Cov: individuals, merit talent mean, wave
-
-~
-
-~
-
-~
-
-~
-
-~
-
--0.009
-
-Note: Cells contain regression coefficients with standard errors in
-parentheses. ***p \textless{} 0.001; **p \textless{} 0.01; *p
-\textless{} 0.05. CWC = centered within group.
+\caption{}
+\begin{center}
+\scalebox{0.7}{
+\begin{tabular}{l c c c c c c}
+\hline
+ & Model 12 & Model 13 & Model 14 & Model 15 & Model 16 & Model 17 \\
+\hline
+Intercept                                                & $1.543^{***}$ & $1.527^{***}$ & $1.521^{***}$ & $1.680^{***}$ & $1.374^{***}$ & $1.362^{***}$ \\
+                                                         & $(0.115)$     & $(0.114)$     & $(0.114)$     & $(0.139)$     & $(0.133)$     & $(0.134)$     \\
+Wave x Perception inequality (WE)                        & $-0.004$      &               &               &               &               &               \\
+                                                         & $(0.006)$     &               &               &               &               &               \\
+Wave x Merit: effort (WE)                                &               & $-0.008$      &               &               &               &               \\
+                                                         &               & $(0.006)$     &               &               &               &               \\
+Wave x Merit: talent (WE)                                &               &               & $-0.006$      &               &               &               \\
+                                                         &               &               & $(0.006)$     &               &               &               \\
+Wave x Perception inequality (BE)                        &               &               &               & $0.011$       &               &               \\
+                                                         &               &               &               & $(0.007)$     &               &               \\
+Wave x Merit: effort (BE)                                &               &               &               &               & $-0.021^{*}$  &               \\
+                                                         &               &               &               &               & $(0.008)$     &               \\
+Wave x Merit: talent (BE)                                &               &               &               &               &               & $-0.020^{*}$  \\
+                                                         &               &               &               &               &               & $(0.008)$     \\
+\hline
+Controls                                                 & Yes           & Yes           & Yes           & Yes           & Yes           & Yes           \\
+BIC                                                      & $20792.207$   & $20767.371$   & $20754.324$   & $20782.059$   & $20749.596$   & $20754.417$   \\
+Numb. obs.                                               & $8643$        & $8643$        & $8643$        & $8643$        & $8643$        & $8643$        \\
+Num. groups: individuals                                 & $1687$        & $1687$        & $1687$        & $1687$        & $1687$        & $1687$        \\
+Var: individuals (Intercept)                             & $0.176$       & $0.167$       & $0.168$       & $0.383$       & $0.254$       & $0.010$       \\
+Var: individuals, perception inequality cwc              & $0.000$       & $$            & $$            & $$            & $$            & $$            \\
+Var: individuals, wave                                   & $0.010$       & $0.009$       & $0.009$       & $0.009$       & $0.009$       & $0.009$       \\
+Cov: individuals (Intercept), perception inequality cwc  & $-0.007$      & $$            & $$            & $$            & $$            & $$            \\
+Cov: individuals (Intercept), wave                       & $-0.021$      & $-0.018$      & $-0.018$      & $-0.048$      & $-0.003$      & $0.006$       \\
+Cov: individuals, perception inequality cwc, wave        & $0.002$       & $$            & $$            & $$            & $$            & $$            \\
+Var: Residuals                                           & $0.495$       & $0.480$       & $0.479$       & $0.494$       & $0.494$       & $0.494$       \\
+Var: individuals, merit effort cwc                       & $$            & $0.019$       & $$            & $$            & $$            & $$            \\
+Cov: individuals (Intercept), merit effort cwc           & $$            & $0.023$       & $$            & $$            & $$            & $$            \\
+Cov: individuals, merit effort cwc, wave                 & $$            & $-0.005$      & $$            & $$            & $$            & $$            \\
+Var: individuals, merit talent cwc                       & $$            & $$            & $0.019$       & $$            & $$            & $$            \\
+Cov: individuals (Intercept), merit talent cwc           & $$            & $$            & $0.034$       & $$            & $$            & $$            \\
+Cov: individuals, merit talent cwc, wave                 & $$            & $$            & $-0.006$      & $$            & $$            & $$            \\
+Var: individuals, perception inequality mean             & $$            & $$            & $$            & $0.008$       & $$            & $$            \\
+Cov: individuals (Intercept), perception inequality mean & $$            & $$            & $$            & $-0.045$      & $$            & $$            \\
+Cov: individuals, perception inequality mean, wave       & $$            & $$            & $$            & $0.009$       & $$            & $$            \\
+Var: individuals, merit effort mean                      & $$            & $$            & $$            & $$            & $0.053$       & $$            \\
+Cov: individuals (Intercept), merit effort mean          & $$            & $$            & $$            & $$            & $-0.089$      & $$            \\
+Cov: individuals, merit effort mean, wave                & $$            & $$            & $$            & $$            & $-0.006$      & $$            \\
+Var: individuals, merit talent mean                      & $$            & $$            & $$            & $$            & $$            & $0.025$       \\
+Cov: individuals (Intercept), merit talent mean          & $$            & $$            & $$            & $$            & $$            & $-0.007$      \\
+Cov: individuals, merit talent mean, wave                & $$            & $$            & $$            & $$            & $$            & $-0.009$      \\
+\hline
+\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$. CWC = centered within group.}}
+\end{tabular}
+}
+\label{table:coefficients}
+\end{center}
 
@@
 
+\bibitem[\citeproctext]{ref-campbell_policy_2012}
+Campbell, A. L. (2012). Policy {Makes Mass Politics}. \emph{Annual
+Review of Political Science}, \emph{15}(Volume 15, 2012), 333--351.
+\url{https://doi.org/10.1146/annurev-polisci-012610-135202}
+
 \bibitem[\citeproctext]{ref-canalesceron_sujeto_2021}
@@
 
+\bibitem[\citeproctext]{ref-pierson_when_1993}
+Pierson, P. (1993). When {Effect Becomes Cause}: {Policy Feedback} and
+{Political Change}. \emph{World Politics}, \emph{45}(4), 595--628.
+\url{https://doi.org/10.2307/2950710}
+
 \bibitem[\citeproctext]{ref-reynolds_perceptions_2014}

### 2025-05-23 07-supplementary-material.qmd
submission materials

@@
 ---
-author: "Equipo EDUMER"
-bibliography: "input/bib/merit-edjust.bib"
-csl: "input/bib/apa6.csl"
+title: "Supplementary material"
+format:
+  html: 
+    code-fold: true
+    html-math-method: katex
+    number-sections: true
+  docx:
+    number-sections: true
+  pdf: 
+    geometry: margin=0.5cm
+    template-partials: 
+      - title.tex
+    keep-tex: true
+    number-sections: true
+editor: source
+link-citations: true
+linestretch: 1.15       
+mainfont: Times New Roman
+fontsize: 12pt          
+colorlinks: true
+fig-height: 4
+fig-width: 7.5
 --- 
@@
 
-# Supplementary material
-
 ```{r echo=FALSE, results='asis'}
-#| label: tbl-suple1
-#| tbl-cap: "Complete longitudinal multilevel models for market justice preferences"
-#| tbl-cap-location: top
-
 ccoef <- list(
@@
 
-texreg::knitreg(list(m0, m1, m2, m3, m4, m5, m6, m7),
+texreg::texreg(list(m0, m1, m2, m3, m4, m5, m6, m7),
                custom.model.names = c("Model 0",paste0("Model ", seq(1:7))),
-               caption = NULL,
+               caption = "Complete longitudinal multilevel models for market justice preferences",
+               caption.above = T,
                stars = c(0.05, 0.01, 0.001),
@@
                              "Income quintile (Ref.= Quintile 1)" = 16:20,
-                             "Political identification (Ref.= Left)" = 21:23,
-                             "Age (Ref.= 18-29)" = 25:27),
+                             "Political identification (Ref.= Left)" = 22:24,
+                             "Age (Ref.= 18-29)" = 26:28),
                custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
@@
                booktabs = F,
-               scalebox = 0.7,
+               scalebox = 0.5,
                include.loglik = FALSE,

### 2025-05-23 cover_letter.tex
submission materials

@@
 \makeatletter
+\makeatother
+\makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \maketitle
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners, breakable, enhanced, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
 \setstretch{1.15}
-\section{Authors Identification}\label{authors-identification}
+\hypertarget{authors-identification}{%
+\section{Authors Identification}\label{authors-identification}}
 
@@
 
-Julio Iturra, Bremen International Graduate School of Social Sciences,
-Bremen, Germany. ORCID: 0000-0003-1993-3479
+Julio Iturra-Sanhueza, Bremen International Graduate School of Social
+Sciences, Bremen, Germany. ORCID: 0000-0003-1993-3479
 
+\hypertarget{exclusive-submission-statement}{%
 \section{Exclusive Submission
-Statement}\label{exclusive-submission-statement}
+Statement}\label{exclusive-submission-statement}}
 
@@
 
-\section{Acknowledgments}\label{acknowledgments}
+\hypertarget{acknowledgments}{%
+\section{Acknowledgments}\label{acknowledgments}}
 
@@
 Development Agency (ANID) through the grant from the National Fund for
-Scientific and Technological Development (FONDECYT) N° 1210847; from the
+Scientific and Technological Development (FONDECYT) N° 1250518; from the
 Center for Conflict and Social Cohesion Studies (COES) grant Financing
@@
 
-\section{Authors' Contributions}\label{authors-contributions}
+\hypertarget{authors-contributions}{%
+\section{Authors' Contributions}\label{authors-contributions}}
 
@@
 
-\section{Conflict of Interest}\label{conflict-of-interest}
+\hypertarget{conflict-of-interest}{%
+\section{Conflict of Interest}\label{conflict-of-interest}}
 
@@
 
-\section{Funding}\label{funding}
+\hypertarget{funding}{%
+\section{Funding}\label{funding}}
 
-This research received funding from ANID/FONDECYT grant 1210847 and the
+This research received funding from ANID/FONDECYT grant 1250518 and the
 Center for Conflict and Social Cohesion Studies (COES) through
-ANID/FONDAP grant 1513009.
+ANID/FONDAP grant 1523A0005.
 
-\section{Data Avaibility Statement}\label{data-avaibility-statement}
+\hypertarget{data-avaibility-statement}{%
+\section{Data Avaibility Statement}\label{data-avaibility-statement}}
 

### 2025-05-23 paper-blinded.qmd
submission materials

--- 2025-05-22 paper.qmd
@@
 csl: "input/bib/harvard.csl"
-author:
-  - name: Juan Carlos Castillo
-    corresponding: true
-    orcid: 0000-0003-1265-7854
-    email: juancastillov@uchile.cl
-    affiliations:
-      - ref: 1
-      - ref: 2
-      - ref: 3
-  - name: Andreas Laffert
-    orcid: 0000-0002-9008-2454
-    email: alaffertt@estudiante.uc.cl 
-    affiliations:
-      - ref: 4
-  - name: Kevin Carrasco
-    orcid: 0000-0002-9348-8361
-    email: kevin.carrasco@ug.uchile.cl
-    affiliations:
-      - ref: 2
-  - name: Julio Iturra
-    orcid: 0000-0003-1993-3479
-    email: julioiturrasanhueza@gmail.com 
-    affiliations:
-      - ref: 5     
-affiliations: 
-  - id: 1
-    name: Departamento de Sociología, Universidad de Chile
-  - id: 2
-    name: Centro de estudios del conflicto y cohesión social (COES)
-  - id: 3
-    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
-  - id: 4
-    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
-  - id: 5
-    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
+# author:
+#   - name: Juan Carlos Castillo
+#     corresponding: true
+#     orcid: 0000-0003-1265-7854
+#     email: juancastillov@uchile.cl
+#     affiliations:
+#       - ref: 1
+#       - ref: 2
+#       - ref: 3
+#   - name: Andreas Laffert
+#     orcid: 0000-0002-9008-2454
+#     email: alaffertt@estudiante.uc.cl 
+#     affiliations:
+#       - ref: 4
+#   - name: Kevin Carrasco
+#     orcid: 0000-0002-9348-8361
+#     email: kevin.carrasco@ug.uchile.cl
+#     affiliations:
+#       - ref: 2
+#   - name: Julio Iturra-Sanhueza
+#     orcid: 0000-0003-1993-3479
+#     email: julioiturrasanhueza@gmail.com 
+#     affiliations:
+#       - ref: 5     
+# affiliations: 
+#   - id: 1
+#     name: Departamento de Sociología, Universidad de Chile
+#   - id: 2
+#     name: Centro de estudios del conflicto y cohesión social (COES)
+#   - id: 3
+#     name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
+#   - id: 4
+#     name: Instituto de Sociología, Pontificia Universidad Católica de Chile
+#   - id: 5
+#     name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
 citeproc: true

### 2025-05-23 submission/Main_blinded.tex
submission materials

--- 2025-05-22 paper.tex
@@
 \usepackage{lmodern}
+\usepackage{times}
 \usepackage{setspace}
 \usepackage{iftex}
+
 \ifPDFTeX
@@
 Shaping Preferences for Market Justice in Chile (2016-2023)}
-
-  \author{Juan Carlos Castillo}
-            \affil{%
-                  Departamento de Sociología, Universidad de Chile
-              }
-          \affil{%
-                  Centro de estudios del conflicto y cohesión social
-                  (COES)
-              }
-          \affil{%
-                  Núcleo milenio de desigualdades y oportunidades
-                  digitales (NUDOS)
-              }
-        \author{Andreas Laffert}
-            \affil{%
-                  Instituto de Sociología, Pontificia Universidad
-                  Católica de Chile
-              }
-        \author{Kevin Carrasco}
-            \affil{%
-                  Centro de estudios del conflicto y cohesión social
-                  (COES)
-              }
-        \author{Julio Iturra}
-            \affil{%
-                  International Graduate School of Social Sciencies
-                  (BIGSSS), University of Bremen, Germany
-              }
+\author{ }
 
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, frame hidden, interior hidden, breakable, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, breakable, sharp corners, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 between market justice and political justice. The author defines market
-justice as a system of ``earned deserts,'' whereby individuals are seen
+justice as a system of ``earned deserts'', whereby individuals are seen
 as deserving of a determined distributive outcome due to their effort
@@
 (\protect\hyperlink{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina
-et al., 2025}) Meritocratic beliefs serve as symbolic justifications for
-unequal outcomes, particularly when access is stratified by income or
-social opportunity. Prior studies have shown that individuals who
+et al., 2025}). Meritocratic beliefs serve as symbolic justifications
+for unequal outcomes, particularly when access is stratified by income
+or social opportunity. Prior studies have shown that individuals who
 perceive higher levels of meritocracy tend to express stronger support
@@
 
-Figure~\ref{fig-alluvial} shows the annual frequencies of market justice
+Figure~\ref{figure1} shows the annual frequencies of market justice
 preferences for healthcare, pensions, and education from 2016 to 2023.
@@
 
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-alluvial-1.pdf}
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{figure1.pdf}
 
@@
 
-\caption{\label{fig-alluvial}Change in the justification of inequality
+\caption{\label{figure1}Change in the justification of inequality
 in healthcare, pensions and education over time (2016-2023)}
@@
 Regarding the main dependent and independent variables of this study,
-Figure~\ref{fig-meanchange} depicts their average changes over the
+Figure~\ref{figure2} depicts their average changes over the
 years. We observe an increase in the average level of market justice
@@
 
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-meanchange-1.pdf}
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{figure2.pdf}
 
@@
 
-\caption{\label{fig-meanchange}Change in the mean of market justice
+\caption{\label{figure2}Change in the mean of market justice
 preferences, economic inequality perception, and meritocracy
@@
 
-Figure~\ref{fig-matrix} presents the correlation matrix for the main
+Figure~\ref{figure3} presents the correlation matrix for the main
 variables in the latest wave (2023). Overall, the coefficients range
@@
 
-{\centering \includegraphics[width=0.9\textwidth,height=\textheight]{paper_files/figure-pdf/fig-matrix-1.pdf}
+{\centering \includegraphics[width=0.9\textwidth,height=\textheight]{figure3.pdf}
 
@@
 
-\caption{\label{fig-matrix}Correlation matrix of the main variables for
+\caption{\label{figure3}Correlation matrix of the main variables for
 the last wave (2023)}
@@
 
-{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-interact-1.pdf}
+{\centering \includegraphics[width=1\textwidth,height=\textheight]{figure4.pdf}
 
@@
 
-\caption{\label{fig-interact}Between effects of meritocratic perceptions
+\caption{\label{figure4}Between effects of meritocratic perceptions
 on market justice preferences by economic inequality perception}
@@
 marginal effects of this interaction are illustrated in
-Figure~\ref{fig-interact}, which shows that the effect of meritocratic
+Figure~\ref{figure4}, which shows that the effect of meritocratic
 perceptions on market justice preferences intensifies as perceived

### 2025-07-11 03-methods.qmd
Update carpeta resubmission y varios (archivos no renderizados)

@@
                performance,
-               influence.ME,
                marginaleffects,
-               MLMusingR,
                texreg, 
                ggdist,
-               misty,
                kableExtra,
@@
 
-This study draws on data from the Chilean Longitudinal Social Survey (ELSOC), a panel study conducted annually from 2016 to 2023.
-The survey assesses how individuals think, feel, and behave regarding social conflict and cohesion in Chile.
-ELSOC employs a probabilistic, stratified, clustered, multistage sampling design covering major urban centers (Santiago, Valparaíso, and Concepción) as well as smaller cities.
+This study draws on data from the Chilean Longitudinal Social Survey (ELSOC), a nationally representative panel study of the urban adult population in Chile, conducted annually between 2016 and 2023.
+Designed to examine individuals’ attitudes, emotions, and behaviors regarding social conflict and cohesion, ELSOC employs a probabilistic, stratified, clustered, and multistage sampling design covering both major urban centers and smaller cities.
+The sampling frame was proportionally stratified into six categories of urban population size (e.g., large and small cities), followed by the random selection of households within 1,067 city blocks.
 The target population includes men and women aged 18 to 75 who are habitual residents of private dwellings.
@@
 The survey has been conducted every year since 2016, except in 2020, when it was suspended due to the COVID-19 pandemic.
-The first wave included 2,927 participants from both northern and southern regions, covering 77% of the national population and 93% of the urban population, with a response rate of 62.4% [@elsoc_estudio_2022].
 This study uses six waves: 2016, 2017, 2018, 2019, 2022, and 2023.
@@
 
-**Market justice preferences**: The dependent variable in this study is preferences for market justice.
-This construct is operationalized through three items that capture how strongly individuals justify conditioning access to social services---healthcare, pensions, and education--- basen on individual income.
+**Market justice preferences**: the dependent variable in this study is preferences for market justice.
+This construct is operationalized through three items that capture how strongly individuals justify conditioning access to core services---healthcare, pensions, and education--- basen on individual income.
 Specifically, the justification of inequality in healthcare is assessed by the question: "Is it fair in Chile that people with higher incomes can access better healthcare than people with lower incomes?" The same question is posed for pensions and education.
@@
 
-**Perception of economic inequality**: The main independent variable refers to the perception of economic inequality, measured through the perceived wage gap [@castillo_cual_2009; @hadler_why_2005; @gijsberts_thelegitimation_1999].
+**Perception of economic inequality**: this variable is measured through the perceived wage gap [@castillo_cual_2009; @hadler_why_2005; @gijsberts_thelegitimation_1999].
 This measure is derived from the salary gap between the perceived salaries of jobs at opposite ends of the occupational hierarchy.
@@
 Sociodemographic and attitudinal variables are included to control for potential composition effects in the population.
-In terms of sociodemographic characteristics, we incorporate per capita household income quintile, educational level (1=Less than Universitary, 2=Universitary), age (in years), and sex (1=Male, 2=Female), which have been previously shown to influence market justice preferences significantly [@lindh_public_2015; @castillo_socialization_2024].
-Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (ranging from 1 to 10) as they may affect the relationship between market justice preferences, perceptions of inequality, and meritocracy [@schneider_poverty_2015].
+In terms of sociodemographic characteristics, we incorporate per capita household income quintile, educational level (1=Less than Universitary, 2=Universitary), age (in ranges), and sex (1=Male, 2=Female), which have been previously shown to influence market justice preferences significantly [@lindh_public_2015; @castillo_socialization_2024].
+Regarding attitudinal variables, we include political identification (1=Left, 2=Center, 3=Right, 4=No identification) and subjective social status (measured through a scale from 1 to 10) as they may affect the relationship between market justice preferences, perceptions of inequality, and meritocracy [@schneider_poverty_2015].
+Descriptive statistics for the control variables can be found in the supplementary material.
 

### 2025-07-11 04-analisis.qmd
Update carpeta resubmission y varios (archivos no renderizados)

--- 2025-05-22 04-analisis.qmd
@@
                performance,
-               influence.ME, 
                marginaleffects,
-               MLMusingR,
                texreg, 
                ggdist,
-               misty,
                kableExtra,
@@
 
-## Descriptive
+## Descriptive statistics
 
@@
 
-Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their average changes over the years.
+Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their standardized average changes over the years.
 We observe an increase in the average level of market justice preferences in the most recent waves, which begins in 2019.
-The highest average consistently appears for perceived economic inequality, although this variable shows a downward trend of roughly one point over time.
+The highest average appears for perceived economic inequality in 2016, although this variable shows a downward trend over time, with the exception of 2019, which can be associated with the “social crisis” in the country.
 Interestingly, while perceptions of economic inequality declined in the latest measurements (2022–2023), market justice preferences increased.
-The meritocracy measures remain stable, though the perception that individuals are rewarded for intelligence is slightly higher than the perception that they are rewarded for their effort.
+The meritocracy measures remain stable, though the perception that individuals are rewarded for effort is slightly higher than the perception that they are rewarded for their intelligence
 
@@
 #| fig-asp: 0.8
-#| fig-cap: "Change in the mean of market justice preferences, economic inequality perception, and meritocracy (2016-2023)"
+#| fig-cap: "Change in the standarized mean of market justice preferences, economic inequality perception, and meritocracy (2016-2023)"
 #| fig-cap-location: top
@@
 
-df_pond %>% 
+df_pond <- df_pond %>% 
   select(ola, mjp, perc_inequality, merit_effort, merit_talent) %>% 
   mutate_at(.vars = 2:5, .funs = ~ as.numeric(.)) %>% 
+  mutate(
+    mjp_z            = as.numeric(scale(mjp)),
+    perc_inequality_z = as.numeric(scale(perc_inequality)),
+    merit_effort_z    = as.numeric(scale(merit_effort)),
+    merit_talent_z    = as.numeric(scale(merit_talent))
+  )
+
+df_pond %>% 
+  select(ola, mjp_z, perc_inequality_z, merit_effort_z, merit_talent_z) %>% 
   group_by(ola) %>% 
@@
   ) %>% 
-  mutate(variable = case_when(variable == "mjp" ~ "Market justice preferences",
-                              variable == "perc_inequality"  ~ "Inequality perception",
-                              variable == "merit_effort"     ~ "Merit: Effort",
-                              variable == "merit_talent"     ~ "Merit: Talent"),
+  mutate(variable = case_when(variable == "mjp_z" ~ "Market justice preferences",
+                              variable == "perc_inequality_z"  ~ "Inequality perception",
+                              variable == "merit_effort_z"     ~ "Merit: Effort",
+                              variable == "merit_talent_z"     ~ "Merit: Talent"),
          variable = factor(variable, levels = c("Market justice preferences",
@@
   scale_shape_manual(values=c(15, 16, 17, 18)) +
-  labs(y = "Mean",
+  labs(y = "Mean (Z-score)",
        x = "Wave",
@@
         text = element_text(size = 12))
-
-```
-
-@fig-matrix presents the correlation matrix for the main variables in the latest wave (2023).
-Overall, the coefficients range from low to moderate.
-The association between market justice preferences and economic inequality perception is negative but small and statistically significant ($r$ = -0.05, $p$ < .05).
-By contrast, market justice preferences positively and significantly correlate with the two meritocracy variables ($r$ = 0.15, $p$ < .01; $r$ = 0.13, $p$ < .01).
-Perceived economic inequality has negative and nonsignificant correlations with both meritocracy perceptions ($r$ = -0.04, $p$ > .05; $r$ = -0.04, $p$ > .05).
-Finally, the two meritocracy variables exhibit a strong positive association with each other ($r$ = 0.69, $p$ < .01).
-
-```{r}
-#| label: fig-matrix
-#| out-width: '90%'
-#| fig-asp: 1
-#| fig-cap: "Correlation matrix of the main variables for the last wave (2023)"
-#| fig-cap-location: top
-#| echo: false
-#| results: asis
-
-M <- df_study1 %>%
-  filter(ola == 6) %>% 
-  select(mjp, perc_inequality, merit_effort, merit_talent) %>% 
-  mutate_all(.funs = ~ as.numeric(.)) %>% 
-  cor(use = "pairwise.complete.obs", method = "pearson")
-
-diag(M) <- NA
-
-rownames(M) <- c("A. Marjet justice preferences",
-                 "B. Inequality gap perception",
-                 "C. People are rewarded for their efforts ",
-                 "D. People are rewarded for their intelligence")
-
-#set Column names of the matrix
-colnames(M) <-c("(A)", "(B)","(C)","(D)")
-
-#Plot the matrix using corrplot
-corrplot::corrplot(M,
-                   method = "color",
-                   addCoef.col = "black",
-                   type = "upper",
-                   tl.col = "black",
-                   col = colorRampPalette(c("#E16462", "white", "#0D0887"))(12),
-                   bg = "white",
-                   na.label = "-")
 
@@
 Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.034 point decrease in market justice preferences.
-Model 4 shows that meritocratic beliefs operate in the opposite direction.
+Model 4 shows that meritocratic perceptions operate in the opposite direction.
 An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.071, $p$ < .001), and a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with higher market-justice preferences ($\beta$ = 0.028, $p$ < .05).
@@
 
+ggeffects::ggpredict(model = m9.1,
+                     terms = c("merit_effort_mean",
+                               "perc_inequality_mean [min, max]"),
+                     ci_level = .95
+                     ) %>% 
+  plot()
+
+predictions(m9.1,
+            condition = c("merit_effort_mean", "perc_inequality_mean"),
+            newdata = datagrid(merit_effort_mean = df_study1$merit_effort_mean,
+                               perc_inequality_mean = seq(min(df_study1$perc_inequality_mean), max(df_study1$perc_inequality_mean), by= 0.1)),
+            re.form = NA) %>% 
+  ggplot(aes(y =estimate,x=merit_effort_mean, fill=perc_inequality_mean, color=perc_inequality_mean)) +
+  geom_ribbon(alpha=0.8,size=1,linetype="solid",color="black") +
+  geom_line(size=1,color="black") +
+  geom_point(size=2,color="black")
+
 p6 <- marginaleffects::plot_slopes(model = m9.1,
@@
 
-@tbl-interactions1 examines whether perceived economic inequality moderates the effect of meritocratic beliefs on market justice preferences.
+@tbl-interactions1 examines whether perceived economic inequality moderates the effect of meritocratic perceptions on market justice preferences.
 Contrary to our hypothesis, the interaction terms in the within-person specification (Models 8 and 9) do not reach statistical significance.

### 2025-07-11 05-discussion.qmd
Update carpeta resubmission y varios (archivos no renderizados)

@@
 
-This study tests several key hypotheses concerning the relationships between perceptions of economic inequality, meritocratic beliefs, and attitudes toward market justice in Chile.
-It posits that higher perceived inequality and stronger meritocratic beliefs both independently contribute to greater support for market-based distributions of social goods.
-Furthermore, the study examines how perceived economic inequality moderates the influence of meritocratic beliefs on market justice preferences.
+This study tests several key hypotheses concerning the relationships between perceptions of economic inequality, meritocratic perceptions, and attitudes toward market justice in Chile.
+It posits that higher perceived inequality and stronger meritocratic perceptions both independently contribute to greater support for market-based distributions of social goods.
+Furthermore, the study examines how perceived economic inequality moderates the influence of meritocratic perceptions on market justice preferences.
 Additionally, it considers how major social and political mobilizations, such as those occurring around 2019 and subsequent years, may have weakened these relationships, reflecting increased critical perspectives on systemic inequalities and market mechanisms.
@@
 This aligned with research indicating that meritocratic narratives served as moral justifications that reinforced societal stratification [@Castillo2012a_justice; @hoyt_mindsets_2023].
-These beliefs functioned symbolically to legitimize structural disparities, reducing support for redistributive measures by framing inequality as fair and deserved.
+These perceptions functioned symbolically to legitimize structural disparities, reducing support for redistributive measures by framing inequality as fair and deserved.
 The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes [@canalesceron_sujeto_2021].
@@
 The third hypothesis concerned the moderating role of perceptions of inequality on the relationship between meritocracy and market justice attitudes.
-The analysis indicated that the positive association between meritocratic beliefs and support for market-based distribution tended to weaken when perceptions of economic inequality were particularly high.
+The analysis indicated that the positive association between meritocratic perceptions and support for market-based distribution tended to weaken when perceptions of economic inequality were particularly high.
 Although this moderation is not significant for within-subject effects (i.e.
@@
 
-Finally, the fourth hypothesis addressed the impact of recent social movements, particularly from 2019 onwards, on the relationship between meritocratic beliefs, perceptions of inequality, and attitudes toward market justice.
+Finally, the fourth hypothesis addressed the impact of recent social movements, particularly from 2019 onwards, on the relationship between meritocratic perceptions, perceptions of inequality, and attitudes toward market justice.
 The results suggested that these movements contributed to a more critical stance toward the fairness of existing distributional mechanisms.
-Post-mobilization, the positive relationship between meritocratic beliefs and support for inequality appeared attenuated, indicating that collective action and public discourse during this period fostered questioning of the legitimacy of systemic inequalities justified solely through meritocratic narratives.
+Post-mobilization, the positive relationship between meritocratic perceptions and support for inequality appeared attenuated, indicating that collective action and public discourse during this period fostered questioning of the legitimacy of systemic inequalities justified solely through meritocratic narratives.
 These social mobilizations challenged dominant narratives, emphasizing structural barriers and systemic injustices, which in turn weakened the association between individual effort and deservingness.

### 2025-07-11 06-conclusion.qmd
Update carpeta resubmission y varios (archivos no renderizados)

@@
 
-This study examined the complex interplay between perceptions of economic inequality, meritocratic beliefs, and attitudes toward market justice in Chile from 2016 to 2023, drawing on longitudinal data from the ELSOC survey.
+This study examined the complex interplay between perceptions of economic inequality, meritocratic perceptions, and attitudes toward market justice in Chile from 2016 to 2023, drawing on longitudinal data from the ELSOC survey.
 By exploring how subjective assessments and social contexts influence support for redistribution and market-based resource allocation, the research offers different elements that contribute to the understanding of the normative foundations underpinning social justice attitudes in a highly unequal, neoliberal environment.
@@
 The findings confirm that higher perceptions of economic inequality are associated with less support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health.
-Meritocratic beliefs strongly reinforce support for market justice, serving as moral justifications for structural inequalities.
-Additionally, perceptions of inequality moderate the influence of meritocratic beliefs, such that as the awareness of disparities are attenuated (i.e.
-less negative) the role of meritocratic beliefs on existing market-based allocations become stronger.
+Meritocratic perceptions strongly reinforce support for market justice, serving as moral justifications for structural inequalities.
+Additionally, perceptions of inequality moderate the influence of meritocratic perceptions, such that as the awareness of disparities are attenuated (i.e.
+less negative) the role of meritocratic perceptions on existing market-based allocations become stronger.
 Notably, the social mobilizations of 2019 and subsequent periods seem to have contributed to shifting attitudes, fostering greater skepticism of systemic fairness and reducing the legitimizing power of meritocratic narratives.
@@
 It highlights the importance of socio-political upheavals in reshaping normative attitudes, underscoring the role of collective action in challenging entrenched narratives of meritocracy and fairness.
-The longitudinal approach provides a deeper temporal perspective on how societal events could influence individual beliefs and preferences.
+The longitudinal approach provides a deeper temporal perspective on how societal events could influence individual perceptions and preferences.
 

### 2025-07-11 07-supplementary-material.qmd
Update carpeta resubmission y varios (archivos no renderizados)

@@
 fig-width: 7.5
+editor_options: 
+  chunk_output_type: console
 --- 
@@
                performance,
-               influence.ME, 
-               marginaleffects,
-               MLMusingR,
                texreg, 
                ggdist,
-               misty,
                kableExtra,
-               ggalluvial, 
                shadowtext,
@@
                sjlabelled,
-               summarytools,
-               interactions)
+               summarytools)
 
@@
   ungroup()
+```
+
+```{r}
+#| label: tbl-summary-controls
+#| tbl-cap: "Descriptive statistics for control variables for the first wave (2016)"
+#| tbl-cap-location: top
+#| results: asis
+#| echo: false
+
+t2 <- df_study1 %>% 
+  filter(ola == "2016") %>% 
+  select(educ,
+         quintil1,
+         sex,
+         age,
+         ess, 
+         ideo)
+
+t2$educ <- sjlabelled::set_label(t2$educ, "Educational level")
+t2$quintil1 <- sjlabelled::set_label(t2$quintil1, "Household income quintile (per capita)")
+t2$sex <- sjlabelled::set_label(t2$sex, "Sex")
+t2$age <- sjlabelled::set_label(t2$age, "Age")
+t2$ess <- sjlabelled::set_label(t2$ess, "Subjective social status")
+t2$ideo <- sjlabelled::set_label(t2$ideo, "Political identification")
+
+df<-dfSummary(t2,
+               plain.ascii = FALSE,
+               style = "multiline",
+               tmp.img.dir = "/tmp",
+               graph.magnif = 0.75,
+               headings = F,  # encabezado
+               varnumbers = F, # num variable
+               labels.col = T, # etiquetas
+               na.col = F,    # missing
+               graph.col = F, # plot
+               valid.col = T, # n valido
+               col.widths = c(30,10,10,10))
+
+df$Variable <- NULL # delete variable column
+
+print(df)
 ```

### 2025-07-11 paper.qmd
Update carpeta resubmission y varios (archivos no renderizados)

--- 2025-05-22 paper.qmd
@@
 abstract: | 
-  This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and preferences for market justice in Chile between 2016 and 2023.
+  This study investigates the relationship between perceptions of economic inequality, meritocratic perceptions, and preferences for market justice in Chile between 2016 and 2023.
 Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions.
-Results indicate that rising perceptions of inequality are associated with greater support for redistributive policies; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities.
+Results indicate that rising perceptions of inequality are associated with lower support for market justice; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities.
 The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness.

### 2025-07-11 resubmission/02-antecedents.qmd
Update carpeta resubmission y varios (archivos no renderizados)

--- 2025-05-22 02-antecedents.qmd
@@
 For example, Castillo et al.
-[-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that those with right-leaning individuals are more supportive of market-based distribution of welfare.
+[-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that right-wing individuals are more supportive of market-based distribution of welfare.
 Beyond individual characteristics, country-level institutions also play a central role.
@@
 Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011].
-In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].
+In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].ƒ
 
@@
 Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit.
-This aligns with the idea that meritocratic beliefs serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit.
+This aligns with the idea that meritocratic perceptions serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit.
 This is particularly relevant in contexts where neoliberal ideologies dominate, as they often emphasize individual responsibility and competition as the basis for social order.
@@
 Third, we propose that perceptions of meritocracy and perceptions of economic inequality interact in shaping market justice attitudes.
-Specifically, we argue that the legitimizing effect of perceived meritocracy on market justice preferences is moderated by perceived economic inequality: when perceived inequality is low, the positive association between meritocratic beliefs and market justice preferences will be stronger.
+Specifically, we argue that the legitimizing effect of perceived meritocracy on market justice preferences is moderated by perceived economic inequality: when perceived inequality is low, the positive association between meritocratic perceptions and market justice preferences will be stronger.
 However, when perceived inequality is high, this association will weaken, as greater awareness of large economic gaps may challenge the view that outcomes are purely merit-based.
@@
 
-$H2$: Higher meritocratic beliefs are associated with higher market justice preferences.
+$H2$: Higher meritocratic perceptions are associated with higher market justice preferences.
 
-$H3$: The positive association between meritocratic beliefs and market justice preferences is moderated by perceived economic inequality; specifically, this association is weaker when perceived economic inequality is high.
+$H3$: The positive association between meritocratic perceptions and market justice preferences is moderated by perceived economic inequality; specifically, this association is weaker when perceived economic inequality is high.
 

### 2025-07-11 resubmission/03-methods.qmd
Update carpeta resubmission y varios (archivos no renderizados)

--- 2025-07-11 03-methods.qmd
@@
 author: "Equipo EDUMER"
-bibliography: "input/bib/merit-edjust.bib"
-csl: "input/bib/apa6.csl"
+bibliography: "../input/bib/merit-edjust.bib"
+csl: "../input/bib/apa6.csl"
 ---
@@
                performance,
+               influence.ME,
                marginaleffects,
+               #MLMusingR,
                texreg, 
                ggdist,
+               misty,
                kableExtra,

### 2025-07-11 resubmission/04-analisis.qmd
Update carpeta resubmission y varios (archivos no renderizados)

--- 2025-07-11 04-analisis.qmd
@@
 author: "Equipo EDUMER"
-bibliography: "input/bib/merit-edjust.bib"
-csl: "input/bib/apa6.csl"
+bibliography: "../input/bib/merit-edjust.bib"
+csl: "../input/bib/apa6.csl"
 editor_options: 
@@
 
-Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their standardized average changes over the years.
-We observe an increase in the average level of market justice preferences in the most recent waves, which begins in 2019.
-The highest average appears for perceived economic inequality in 2016, although this variable shows a downward trend over time, with the exception of 2019, which can be associated with the “social crisis” in the country.
-Interestingly, while perceptions of economic inequality declined in the latest measurements (2022–2023), market justice preferences increased.
-The meritocracy measures remain stable, though the perception that individuals are rewarded for effort is slightly higher than the perception that they are rewarded for their intelligence
+Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their standardized average values across survey waves.
+The results show a notable upward trend in market justice preferences, particularly in the most recent waves.
+Perceived economic inequality exhibited the highest standardized mean in 2016; however, this variable shows a general downward trajectory over time, with a temporary increase in 2019, possibly reflecting the effects of the social uprising that year.
+Interestingly, while perceptions of inequality declined in the latest waves (2022–2023), market justice preferences continued to rise.
+Meritocratic beliefs, in contrast, remain relatively stable overall.
+Nevertheless, perceptions that individuals are rewarded based on talent tend to be slightly higher than those based on effort.
+Both meritocracy-related measures follow a similar temporal pattern: an increase from 2016 to 2018, a decline between 2019 and 2022—potentially associated with the social unrest and the consequences of the COVID-19 pandemic—and a subsequent rise in 2023, returning to levels comparable to those observed in 2016.
 
@@
 @tbl-modelos presents the results of the multilevel models estimated for market justice preferences, examining both individuals (within) and group-level (between) effects.
-The intraclass correlation [@hoxMultilevelAnalysisTechniques2017a] from the empty model (see Supplementary Material), which decomposes the variance of market justice preferences, is 0.23, indicating that approximately 23% of the variation is attributable to differences between individuals.
-Complementary, 77% of the variation corresponds to within-individual differences over time.
+The intraclass correlation [@hoxMultilevelAnalysisTechniques2017a] from the empty model (see Supplementary Material), which decomposes the variance of market justice preferences, is 0.31, indicating that approximately 31% of the variation is attributable to differences between individuals.
+Complementary, 69% of the variation corresponds to within-individual differences over time.
 
@@
 m0 <- lmer(mjp ~ 1 + (1 | idencuesta), 
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
 m1 <- lmer(mjp ~ 1 + ola + (1 | idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
 m1.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + (1 | idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
 m2 <- lmer(mjp ~ 1 + ola_num + ola_2 + (1 + ola_num | idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
 m3 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + (1 + ola_num | idencuesta), 
-           data = df_study1)
+           data = df_study1, weights = ponderador_long_total)
 
 m4 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + (1 + ola_num | idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
 m5 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  (1 + ola_num | idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
 m6 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + (1 + ola_num | idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
 m7 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + ola_num | idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 + merit_effort_cwc + merit_talent_cwc + perc_inequality_cwc | idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 + merit_effort_cwc + perc_inequality_cwc | idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 + merit_talent_cwc + perc_inequality_cwc | idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 | idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
 m10 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality*ola_num + merit_effort + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
 m10.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc*ola_num + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality_cwc + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
 m10.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean*ola_num + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality_mean + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
 m11 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality + merit_effort*ola_num  + merit_talent + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
 m11.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc*ola_num  + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort_cwc + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
 m11.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc  + merit_talent_cwc + perc_inequality_mean + merit_effort_mean*ola_num + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort_mean + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 + merit_talent + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 + merit_talent_cwc + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 + merit_talent_mean + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 + perc_inequality + merit_effort  + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 + perc_inequality_cwc + merit_effort_cwc + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 + perc_inequality_mean + merit_effort_mean  + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 + perc_inequality + merit_talent + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 + perc_inequality_cwc + merit_talent_cwc + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
               (1 + perc_inequality_mean + merit_talent_mean + ola_num| idencuesta),
-                data = df_study1)
+                data = df_study1, weights = ponderador_long_total)
 
@@
 The results in Model 3 suggest that the within effect of perceived economic inequality is negative and statistically significant ($p$ < .001).
-Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.034 point decrease in market justice preferences.
+Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.039 point decrease in market justice preferences.
 Model 4 shows that meritocratic perceptions operate in the opposite direction.
-An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.071, $p$ < .001), and a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with higher market-justice preferences ($\beta$ = 0.028, $p$ < .05).
+An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.070, $p$ < .001), and a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with higher market-justice preferences ($\beta$ = 0.027, $p$ < .05).
 Taken together, these results suggest that individuals who increasingly perceive meritocracy—whether through effort or talent—tend to hold stronger market justice preferences.
@@
 #| fig-asp: 1
-#| fig-cap: "Between effects of meritocratic perceptions on market justice preferences by economic inequality perception"
+#| fig-cap: "Predicted values of market justice preferences by perceptions of meritocraticy and economic inequality"
 #| fig-cap-location: top
@@
 
-ggeffects::ggpredict(model = m9.1,
+min_val <- min(df_study1$perc_inequality_mean, na.rm = TRUE)
+max_val <- max(df_study1$perc_inequality_mean, na.rm = TRUE)
+mean_val <- mean(df_study1$perc_inequality_mean, na.rm = TRUE)
+
+p6 <- ggpredict(
+  m9.1,
   terms = c("merit_effort_mean", 
-                               "perc_inequality_mean [min, max]"),
-                     ci_level = .95
+            sprintf("perc_inequality_mean [%.2f, %.2f, %.2f]", min_val, mean_val, max_val))
 ) %>% 
-  plot()
-
-predictions(m9.1,
-            condition = c("merit_effort_mean", "perc_inequality_mean"),
-            newdata = datagrid(merit_effort_mean = df_study1$merit_effort_mean,
-                               perc_inequality_mean = seq(min(df_study1$perc_inequality_mean), max(df_study1$perc_inequality_mean), by= 0.1)),
-            re.form = NA) %>% 
-  ggplot(aes(y =estimate,x=merit_effort_mean, fill=perc_inequality_mean, color=perc_inequality_mean)) +
-  geom_ribbon(alpha=0.8,size=1,linetype="solid",color="black") +
-  geom_line(size=1,color="black") +
-  geom_point(size=2,color="black")
-
-p6 <- marginaleffects::plot_slopes(model = m9.1,
-                             variables = "merit_effort_mean",
-                             condition = "perc_inequality_mean",
-                             conf_level = .95) +
-  geom_hline(yintercept = 0,
-             color = "red",
-             linetype = "dashed") +
-  labs(x = "Perception inequality (BE)",
-       y = "Marginal effect of Merit: Effort (BE)",
-       fill = NULL,
+  plot() +
+  scale_x_continuous(n.breaks = 5) +
+  scale_color_manual(
+    name = "Perception of inequality (BE)",
+    values = c("#0571B0", "black", "#B2182B"),
+    labels = c("Min", "Mean", "Max")
+  ) +
+  scale_fill_manual(
+    name = "Perception of inequality (BE)",
+    values = c("#0571B0", "black", "#B2182B"),
+    labels = c("Min", "Mean", "Max")) +
+  labs(
+    x = "Merit: Effort (BE)",
+    y = "Market justice preferences",
     title = "(A) People are rewarded by their efforts") +
@@
 
-p7 <- marginaleffects::plot_slopes(model = m9.2,
-                             variables = "merit_talent_mean",
-                             condition = "perc_inequality_mean",
-                             conf_level = .95) +
-  geom_hline(yintercept = 0,
-             color = "red",
-             linetype = "dashed") +
-  labs(x = "Perception inequality (BE)",
-       y = "Marginal effect of Merit: Talent (BE)",
-       fill = NULL,
+p7 <- ggpredict(
+  m9.2,
+  terms = c("merit_talent_mean", 
+            sprintf("perc_inequality_mean [%.2f, %.2f, %.2f]", min_val, mean_val, max_val))
+) %>% 
+  plot() +
+  scale_x_continuous(n.breaks = 5) +
+  scale_color_manual(
+    name = "Perception of inequality (BE)",
+    values = c("#0571B0", "black", "#B2182B"),
+    labels = c("Min", "Mean", "Max")
+  ) +
+  scale_fill_manual(
+    name = "Perception of inequality (BE)",
+    values = c("#0571B0", "black", "#B2182B"),
+    labels = c("Min", "Mean", "Max")) +
+  labs(
+    x = "Merit: Talent (BE)",
+    y = "Market justice preferences",
     title = "(B) People are rewarded for their intelligence",
-       caption = "Source: own elaboration with data from ELSOC 2016-2023 (n = 8460). Based in Model 10 and Model 11 in Table 4.
-Confidence intervals at 95%") +
+  caption = "Source: own elaboration with data from ELSOC 2016-2023 (n = 8460). Based in Model 10 and Model 11 in Table 4") +
   theme_ggdist()
@@
 library(ggpubr)
-ggpubr::ggarrange(p6, p7, common.legend=TRUE, ncol=1)
+ggpubr::ggarrange(p6, p7, common.legend=TRUE, ncol=1, legend = "top")
 
@@
 However, in the between-person specification (Models 10 and 11), the interaction terms are statistically significant, suggesting that perceptions of economic inequality meaningfully alter the effect of meritocracy on support for market-based allocation of social services.
-These positive effects indicate that, as perceptions of inequality increase, the positive association between meritocratic perceptions and market justice preferences becomes stronger.
-The marginal effects of this interaction are illustrated in @fig-interact, which shows that the effect of meritocratic perceptions on market justice preferences intensifies as perceived economic inequality rises.
+Specifically, these positive effects indicate that, as perceptions of inequality increase, the positive association between meritocratic perceptions and market justice preferences becomes stronger.
+As shown in @fig-interact, Panel A, the effect of meritocratic perceptions about effort on market justice preferences strengthens as perceived inequality increases, particularly among individuals who perceive the highest levels of inequality, whereas this effect weakens among those who perceive low levels of inequality.
+Panel B reveals a similar pattern for meritocratic perceptions about talent, although the effect is substantially weaker at low levels of perceived inequality.
+Substantively, these findings suggest that the legitimizing power of meritocratic perceptions is amplified in contexts perceived as highly unequal, reinforcing individualistic understandings of justice and strengthening support for market-based access to essential social services.
 

### 2025-07-11 resubmission/07-supplementary-material.qmd
Update carpeta resubmission y varios (archivos no renderizados)

--- 2025-07-11 resubmission/04-analisis.qmd
@@
 ---
-author: "Equipo EDUMER"
-bibliography: "../input/bib/merit-edjust.bib"
-csl: "../input/bib/apa6.csl"
+title: "Supplementary material"
+format:
+  html: 
+    code-fold: true
+    html-math-method: katex
+    number-sections: true
+  docx:
+    number-sections: true
+  pdf: 
+    geometry: margin=0.5cm
+    template-partials: 
+      - title.tex
+    keep-tex: true
+    number-sections: true
+editor: source
+link-citations: true
+linestretch: 1.15       
+mainfont: Times New Roman
+fontsize: 12pt          
+colorlinks: true
+fig-height: 4
+fig-width: 7.5
 editor_options: 
@@
 --- 
-
 ```{r}
@@
                performance,
-               marginaleffects,
                texreg, 
@@
                kableExtra,
-               ggalluvial, 
                shadowtext,
@@
                sjlabelled,
-               summarytools,
-               interactions,
-               cowplot)
+               summarytools)
 
@@
 ```
-# Results
-
-## Descriptive statistics
-
-@fig-alluvial shows the annual frequencies of market justice preferences for healthcare, pensions, and education from 2016 to 2023.
-Each year presents stacked percentage frequencies, and the flows between them reflect opinion changes among the same individuals from one year to the next, given that we are using panel data.
-For instance, of the 40.8% who strongly disagreed with justifying inequality in healthcare in 2019, around 24.3% maintained that position in 2022, while the remaining 16.5% shifted toward other response categories—primarily moving into disagreement rather than strong disagreement.
-Overall, more than half of the respondents exhibit a high level of disagreement (disagree + strongly disagree) with inequality in these three social service areas over time.
-Despite this general pattern, recent waves show a slight decrease in disagreement and a corresponding rise in support for market-justice inequality.
-Specifically, in healthcare and education, although disagreement remains substantial, agreement (agree + strongly agree) increased from 7.4% and 7.2% in 2019 to 13.1% and 14.2% in 2023, respectively.
-This shift is most evident in pensions, where the combined agree/strongly agree category grew by about 10 percentage points, from 16.9% in 2016 to 28% in 2023.
-
-```{r}
-#| label: fig-alluvial
-#| out-width: '100%'
-#| fig-asp: 1.2
-#| fig-cap: "Change in the justification of inequality in healthcare, pensions and education over time (2016-2023)"
-#| fig-cap-location: top
-#| echo: false
-#| results: asis
-
-datos.health <- df_study1 %>% 
-  mutate(just_health = factor(just_health, 
-                              levels = c("Strongly agree",
-                                         "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"),
-                              ordered = T)) %>% 
-  group_by(idencuesta, ola) %>% 
-  count(just_health) %>% 
-  group_by(ola) %>% 
-  mutate(porcentaje=n/sum(n)) %>% 
-  ungroup() %>% 
-  na.omit() %>% 
-  mutate(wave = case_when(ola == 1 ~ "2016",
-                          ola == 2 ~ "2017",
-                          ola == 3 ~ "2018",
-                          ola == 4 ~ "2019",
-                          ola == 5 ~ "2022",
-                          ola == 6 ~ "2023"),
-         wave = factor(wave, levels = c("2016",
-                                      "2017",
-                                      "2018",
-                                      "2019",
-                                      "2022",
-                                      "2023")))
-
-etiquetas.health <- df_study1 %>%
-  mutate(just_health = factor(just_health, 
-                              levels = c("Strongly agree",
-                                         "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"),
-                              ordered = T)) %>% 
-  group_by(ola, just_health) %>%
-  summarise(count = n(), .groups = "drop") %>%
-  group_by(ola) %>%
-  mutate(porcentaje = count / sum(count)) %>% 
-  na.omit() %>% 
-  mutate(idencuesta = 1,
-         wave = case_when(ola == 1 ~ "2016",
-                          ola == 2 ~ "2017",
-                          ola == 3 ~ "2018",
-                          ola == 4 ~ "2019",
-                          ola == 5 ~ "2022",
-                          ola == 6 ~ "2023"),
-         wave = factor(wave, levels = c("2016",
-                                        "2017",
-                                        "2018",
-                                        "2019",
-                                        "2022",
-                                        "2023")))
-
-p1 <- datos.health %>% 
-  ggplot(aes(x = wave, fill = just_health, stratum = just_health,
-             alluvium = idencuesta, y = porcentaje)) +
-  ggalluvial::geom_flow(alpha = .4) + 
-  ggalluvial::geom_stratum(linetype = 0) +
-  scale_y_continuous(labels = scales::percent) + 
-  scale_fill_manual(values =  c("#0571B0","#92C5DE","#b3b3b3ff","#F4A582","#CA0020")) +
-  geom_shadowtext(data = etiquetas.health,
-                  aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
-                  position = position_stack(vjust = .5),
-                  show.legend = FALSE,
-                  size = 3,
-                  color = rep('white'),
-                  bg.colour='grey30')+
-  labs(y = "%",
-       x = NULL,
-       fill = NULL,
-       title = "a. Healthcare") +
-  theme_ggdist() +
-  theme(legend.position = "none")
-
-datos.pension <- df_study1 %>% 
-   mutate(just_pension = factor(just_pension, 
-                              levels = c("Strongly agree",
-                                         "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"),
-                              ordered = T)) %>% 
-  group_by(idencuesta, ola) %>% 
-  count(just_pension) %>% 
-  group_by(ola) %>% 
-  mutate(porcentaje=n/sum(n)) %>% 
-  ungroup() %>% 
-  na.omit() %>% 
-  mutate(wave = case_when(ola == 1 ~ "2016",
-                          ola == 2 ~ "2017",
-                          ola == 3 ~ "2018",
-                          ola == 4 ~ "2019",
-                          ola == 5 ~ "2022",
-                          ola == 6 ~ "2023"),
-         wave = factor(wave, levels = c("2016",
-                                        "2017",
-                                        "2018",
-                                        "2019",
-                                        "2022",
-                                        "2023")))
-
-etiquetas.pension <- df_study1 %>%
-  mutate(just_pension = factor(just_pension, 
-                              levels = c("Strongly agree",
-                                         "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"),
-                              ordered = T)) %>% 
-  group_by(ola, just_pension) %>%
-  summarise(count = n(), .groups = "drop") %>%
-  group_by(ola) %>%
-  mutate(porcentaje = count / sum(count)) %>% 
-  na.omit() %>% 
-  mutate(idencuesta = 1,
-         wave = case_when(ola == 1 ~ "2016",
-                          ola == 2 ~ "2017",
-                          ola == 3 ~ "2018",
-                          ola == 4 ~ "2019",
-                          ola == 5 ~ "2022",
-                          ola == 6 ~ "2023"),
-         wave = factor(wave, levels = c("2016",
-                                        "2017",
-                                        "2018",
-                                        "2019",
-                                        "2022",
-                                        "2023")))
-
-p2 <- datos.pension %>% 
-  ggplot(aes(x = wave, fill = just_pension, stratum = just_pension,
-             alluvium = idencuesta, y = porcentaje)) +
-  ggalluvial::geom_flow(alpha = .4) + 
-  ggalluvial::geom_stratum(linetype = 0) +
-  scale_y_continuous(labels = scales::percent) + 
-  scale_fill_manual(values =  c("#0571B0","#92C5DE","#b3b3b3ff","#F4A582","#CA0020")) +
-  geom_shadowtext(data = etiquetas.pension,
-                  aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
-                  position = position_stack(vjust = .5),
-                  show.legend = FALSE,
-                  size = 3,
-                  color = rep('white'),
-                  bg.colour='grey30')+
-  labs(y = "%",
-       x = NULL,
-       fill = NULL,
-       title = "b. Pensions") +
-  theme_ggdist() +
-  theme(legend.position = "none")
-
-datos.educ <- df_study1 %>% 
-  mutate(just_educ = factor(just_educ, 
-                            levels = c("Strongly agree",
-                                         "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"),
-                              ordered = T)) %>%
-  group_by(idencuesta, ola) %>% 
-  count(just_educ) %>% 
-  group_by(ola) %>% 
-  mutate(porcentaje=n/sum(n)) %>% 
-  ungroup() %>% 
-  na.omit() %>% 
-  mutate(wave = case_when(ola == 1 ~ "2016",
-                          ola == 2 ~ "2017",
-                          ola == 3 ~ "2018",
-                          ola == 4 ~ "2019",
-                          ola == 5 ~ "2022",
-                          ola == 6 ~ "2023"),
-         wave = factor(wave, levels = c("2016",
-                                        "2017",
-                                        "2018",
-                                        "2019",
-                                        "2022",
-                                        "2023")))
-
-etiquetas.educ <- df_study1 %>%
-    mutate(just_educ = factor(just_educ, 
-                              levels = c("Strongly agree",
-                                         "Agree",
-                                         "Neither agree nor desagree",
-                                         "Desagree",
-                                         "Strongly desagree"),
-                              ordered = T)) %>%
-  group_by(ola, just_educ) %>%
-  summarise(count = n(), .groups = "drop") %>%
-  group_by(ola) %>%
-  mutate(porcentaje = count / sum(count)) %>% 
-  na.omit() %>% 
-  mutate(idencuesta = 1,
-         wave = case_when(ola == 1 ~ "2016",
-                          ola == 2 ~ "2017",
-                          ola == 3 ~ "2018",
-                          ola == 4 ~ "2019",
-                          ola == 5 ~ "2022",
-                          ola == 6 ~ "2023"),
-         wave = factor(wave, levels = c("2016",
-                                        "2017",
-                                        "2018",
-                                        "2019",
-                                        "2022",
-                                        "2023")))
-
-p3 <- datos.educ %>% 
-  ggplot(aes(x = wave, fill = just_educ, stratum = just_educ,
-             alluvium = idencuesta, y = porcentaje)) +
-  ggalluvial::geom_flow(alpha = .4) + 
-  ggalluvial::geom_stratum(linetype = 0) +
-  scale_y_continuous(labels = scales::percent) + 
-  scale_fill_manual(values =  c("#0571B0","#92C5DE","#b3b3b3ff","#F4A582","#CA0020")) +
-  geom_shadowtext(data = etiquetas.educ,
-                  aes(label = ifelse(porcentaje > 0 , scales::percent(porcentaje, accuracy = .1),"")),
-                  position = position_stack(vjust = .5),
-                  show.legend = FALSE,
-                  size = 3,
-                  color = rep('white'),
-                  bg.colour='grey30')+
-  labs(y = "%",
-       x = "Wave",
-       fill = NULL,
-       title = "c. Education",
-       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (N obs = 8,643; N groups = 1,687)") +
-  theme_ggdist() +
-  theme(legend.position = "bottom")
-
-(p1 / p2 / p3)
-
-```
-
-Regarding the main dependent and independent variables of this study, @fig-meanchange depicts their standardized average values across survey waves.
-The results show a notable upward trend in market justice preferences, particularly in the most recent waves.
-Perceived economic inequality exhibited the highest standardized mean in 2016; however, this variable shows a general downward trajectory over time, with a temporary increase in 2019, possibly reflecting the effects of the social uprising that year.
-Interestingly, while perceptions of inequality declined in the latest waves (2022–2023), market justice preferences continued to rise.
-Meritocratic beliefs, in contrast, remain relatively stable overall.
-Nevertheless, perceptions that individuals are rewarded based on talent tend to be slightly higher than those based on effort.
-Both meritocracy-related measures follow a similar temporal pattern: an increase from 2016 to 2018, a decline between 2019 and 2022—potentially associated with the social unrest and the consequences of the COVID-19 pandemic—and a subsequent rise in 2023, returning to levels comparable to those observed in 2016.
-
-```{r}
-#| label: fig-meanchange
-#| out-width: '100%'
-#| fig-asp: 0.8
-#| fig-cap: "Change in the standarized mean of market justice preferences, economic inequality perception, and meritocracy (2016-2023)"
-#| fig-cap-location: top
-#| echo: false
-#| results: asis
-
-library(srvyr)
-
-df_pond <- df_study1 %>% 
-  mutate(ola = case_when(ola == 1 ~ "2016",
-                         ola == 2 ~ "2017",
-                         ola == 3 ~ "2018",
-                         ola == 4 ~ "2019",
-                         ola == 5 ~ "2022",
-                         ola == 6 ~ "2023"),
-         ola = factor(ola, levels = c("2016",
-                                        "2017",
-                                        "2018",
-                                        "2019",
-                                        "2022",
-                                        "2023"))) %>% 
-  as_survey_design(.data = .,
-                   ids = segmento, 
-                   strata = estrato, 
-                   weights = ponderador_long_total)
-
-df_pond <- df_pond %>% 
-  select(ola, mjp, perc_inequality, merit_effort, merit_talent) %>% 
-  mutate_at(.vars = 2:5, .funs = ~ as.numeric(.)) %>% 
-  mutate(
-    mjp_z            = as.numeric(scale(mjp)),
-    perc_inequality_z = as.numeric(scale(perc_inequality)),
-    merit_effort_z    = as.numeric(scale(merit_effort)),
-    merit_talent_z    = as.numeric(scale(merit_talent))
-  )
-
-df_pond %>% 
-  select(ola, mjp_z, perc_inequality_z, merit_effort_z, merit_talent_z) %>% 
-  group_by(ola) %>% 
-  summarise_all(~survey_mean(., vartype = "ci")) %>% 
-  pivot_longer(cols = -ola,
-               names_to = "temp",
-               values_to = "valor") %>%
-  mutate(
-    ci = case_when(
-      str_ends(temp, "_low") ~ "ic_low",
-      str_ends(temp, "_upp") ~ "ic_upp",
-      TRUE                   ~ "mean"
-    ),
-    variable = str_remove(temp, "_low|_upp")
-  ) %>%
-  select(ola, variable, ci, valor) %>%
-  pivot_wider(
-    names_from  = ci,
-    values_from = valor
-  ) %>% 
-  mutate(variable = case_when(variable == "mjp_z" ~ "Market justice preferences",
-                              variable == "perc_inequality_z"  ~ "Inequality perception",
-                              variable == "merit_effort_z"     ~ "Merit: Effort",
-                              variable == "merit_talent_z"     ~ "Merit: Talent"),
-         variable = factor(variable, levels = c("Market justice preferences",
-                                      "Inequality perception",
-                                      "Merit: Effort",
-                                      "Merit: Talent")
-                              )) %>% 
-  ggplot(aes(x = ola, y = mean, group = variable)) +
-  geom_point(aes(shape=variable, color=variable), size = 3.5) +
-  geom_line(aes(color = variable), linewidth = 0.8) +
-  geom_errorbar(aes(ymin = ic_low, ymax = ic_upp, color = variable),
-                width = 0.1) +
-  scale_color_manual(values = c("#023858", # azul oscuro
-                                "#0571B0", # azul intermedio
-                                "#F4A582", # salmón
-                                "#B2182B"  # rojo profundo
-                                         )) +
-  scale_shape_manual(values=c(15, 16, 17, 18)) +
-  labs(y = "Mean (Z-score)",
-       x = "Wave",
-       color = NULL,
-       shape = NULL,
-       caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (N obs = 8,643; N groups = 1,687)") +
-  theme_ggdist() +
-  theme(legend.position = "top",
-        text = element_text(size = 12))
-
-```
-
-## Multilevel models
-
-@tbl-modelos presents the results of the multilevel models estimated for market justice preferences, examining both individuals (within) and group-level (between) effects.
-The intraclass correlation [@hoxMultilevelAnalysisTechniques2017a] from the empty model (see Supplementary Material), which decomposes the variance of market justice preferences, is 0.31, indicating that approximately 31% of the variation is attributable to differences between individuals.
-Complementary, 69% of the variation corresponds to within-individual differences over time.
 
@@
   ungroup()
+```
+
+```{r}
+#| label: tbl-summary-controls
+#| tbl-cap: "Descriptive statistics for control variables for the first wave (2016)"
+#| tbl-cap-location: top
+#| results: asis
+#| echo: false
+
+t2 <- df_study1 %>% 
+  filter(ola == "2016") %>% 
+  select(educ,
+         quintil1,
+         sex,
+         age,
+         ess, 
+         ideo)
+
+t2$educ <- sjlabelled::set_label(t2$educ, "Educational level")
+t2$quintil1 <- sjlabelled::set_label(t2$quintil1, "Household income quintile (per capita)")
+t2$sex <- sjlabelled::set_label(t2$sex, "Sex")
+t2$age <- sjlabelled::set_label(t2$age, "Age")
+t2$ess <- sjlabelled::set_label(t2$ess, "Subjective social status")
+t2$ideo <- sjlabelled::set_label(t2$ideo, "Political identification")
+
+df<-dfSummary(t2,
+               plain.ascii = FALSE,
+               style = "multiline",
+               tmp.img.dir = "/tmp",
+               graph.magnif = 0.75,
+               headings = F,  # encabezado
+               varnumbers = F, # num variable
+               labels.col = T, # etiquetas
+               na.col = F,    # missing
+               graph.col = F, # plot
+               valid.col = T, # n valido
+               col.widths = c(30,10,10,10))
+
+df$Variable <- NULL # delete variable column
+
+print(df)
+```
+
+```{r}
+#| label: fig-matrix
+#| out-width: '90%'
+#| fig-asp: 1
+#| fig-cap: "Correlation matrix of the main variables for the last wave (2023)"
+#| fig-cap-location: top
+#| echo: false
+#| results: asis
+
+M <- df_study1 %>%
+  filter(ola == "2016") %>% 
+  select(mjp, perc_inequality, merit_effort, merit_talent) %>% 
+  mutate_all(.funs = ~ as.numeric(.)) %>% 
+  cor(use = "pairwise.complete.obs", method = "pearson")
+
+diag(M) <- NA
+
+rownames(M) <- c("A. Marjet justice preferences",
+                 "B. Inequality gap perception",
+                 "C. People are rewarded for their efforts ",
+                 "D. People are rewarded for their intelligence")
+
+#set Column names of the matrix
+colnames(M) <-c("(A)", "(B)","(C)","(D)")
+
+#Plot the matrix using corrplot
+corrplot::corrplot(M,
+                   method = "color",
+                   addCoef.col = "black",
+                   type = "upper",
+                   tl.col = "black",
+                   col = colorRampPalette(c("#E16462", "white", "#0D0887"))(12),
+                   bg = "white",
+                   na.label = "-")
+
 ```
@@
 
-#texreg(list(m13.1, m13.2,m14.1, m14.2))
-```
-
-```{r, include=FALSE, eval=FALSE}
-interactions::interact_plot(m13.2, pred = "merit_effort_mean", modx="perc_inequality_mean", mod2="ola_num")
+#knitreg(list(m13.1, m13.2,m14.1, m14.2))
 ```
@@
 ```{r echo=FALSE, results='asis'}
-#| label: tbl-modelos
-#| tbl-cap: "Longitudinal multilevel models for market justice preferences"
-#| tbl-cap-location: top
-
-omit <- "(educ)|(quintil1)|(ess)|(ideo)|(sex)|(age)"
-
 ccoef <- list(
@@
   merit_effort_mean = "Merit: Effort (BE)",
-  merit_talent_mean = "Merit: Talent (BE)"
+  merit_talent_mean = "Merit: Talent (BE)",
+  "educUniversitary" = "Universitary education (Ref.= Less than Universitary)",
+  quintil1Q2 = "Quintile Q2",
+  quintil1Q3 = "Quintile Q3",
+  quintil1Q4 = "Quintile Q4",
+  quintil1Q5 = "Quintile Q5",
+  quintil1QNA = "Quintile no information",
+  ess = "Subjective social status",
+  ideoCenter = "Center",
+  ideoRight = "Right",
+  "ideoDoes not identify" = "Does not identify",
+  sexFemale = "Female (Ref.= Male)",
+  "age30-49" = "Age 30-49",
+  "age50-64" = "Age 50-64",
+  "age65 or more" = "Age 65 or more"
   )
 
-texreg::texreg(list(m1, m2, m3, m4, m5, m6, m7),
-               custom.model.names = c(paste0("Model ", seq(1:7))),caption.above = T,
-               caption = NULL,
+texreg::texreg(list(m0, m1, m2, m3, m4, m5, m6, m7),
+               custom.model.names = c("Model 0",paste0("Model ", seq(1:7))),
+               caption = "Complete longitudinal multilevel models for market justice preferences",
+               caption.above = T,
                stars = c(0.05, 0.01, 0.001),
-               omit.coef = omit,
                custom.coef.map = ccoef,
                digits = 3,
-               groups = list("Wave (Ref.= 2016)" = 2:6),
+               groups = list("Wave (Ref.= 2016)" = 2:6,
+                             "Income quintile (Ref.= Quintile 1)" = 16:20,
+                             "Political identification (Ref.= Left)" = 22:24,
+                             "Age (Ref.= 18-29)" = 26:28),
                custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
@@
                booktabs = F,
-               scalebox = 0.80,
+               scalebox = 0.5,
                include.loglik = FALSE,
@@
                center = T,
-               custom.gof.names = c("BIC", "Numb. obs.", "Num. groups: individuals", "Var: individuals (Intercept)", "Var: Residual", "Var: individuals, wave", "Cov: individuals (Intercept), wave"),
-               custom.gof.rows = list("Controls"=c(rep("No",6), "Yes")))
+               custom.gof.names = c("BIC", "Numb. obs.", "Num. groups: individuals", "Var: individuals (Intercept)", "Var: Residual", "Var: individuals, wave", "Cov: individuals (Intercept), wave"))
 
@@
 
-According to Model 1, which includes the survey waves to capture intertemporal variations in the dependent variable, there is a decrease in 2017 ($\beta$ = -0.167, $p$ < .001) relative to 2016, and similarly in 2018 ($\beta$ = -0.025, $p$ > .05) and 2019 ($\beta$ = -0.046, $p$ > .05), although the latter effects are not statistically significant.
-In contrast, in the more recent waves of 2022 and 2023, there is a statistically significant increase in market justice preferences ($\beta$ = 0.285, $p$ < .001; $\beta$ = 0.284, $p$ < .001), suggesting a non-linear effect.
-To model this trajectory over time, Model 2 incorporates time (survey waves) as a continuous variable, along with its quadratic term, representing the non-linear association initially observed in Model 1.
-While the linear term (survey wave) shows a negative association, reflecting an overall decline in market preferences over time, the positive quadratic term indicates a reversal of this pattern in the final measurement points.
+```{r}
+#| echo: false
+#| include: false
 
-Models 3 and 4 incorporate the within-group effects (WE) of the primary independent variables, capturing how individual changes in these variables over time shape the dependent variable.
-The results in Model 3 suggest that the within effect of perceived economic inequality is negative and statistically significant ($p$ < .001).
-Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.039 point decrease in market justice preferences.
-Model 4 shows that meritocratic perceptions operate in the opposite direction.
-An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.070, $p$ < .001), and a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with higher market-justice preferences ($\beta$ = 0.027, $p$ < .05).
-Taken together, these results suggest that individuals who increasingly perceive meritocracy—whether through effort or talent—tend to hold stronger market justice preferences.
+# growth curves con ola categorica
 
-When examining the between-group effects (BE) in Model 5 and 6, which capture differences between individuals in the average of the main variables, a similar pattern emerges.
-Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.042, $p$ < .05).
-In Model 6, the meritocratic perception that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.186, $p$ < .001), whereas the perception that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.040).
-Notably, once the meritocratic variables are included, the negative coefficient for perceived economic inequality remains but is no longer statistically significant.
+m15.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc*ola + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality_cwc + ola_num| idencuesta),
+                data = df_study1, weights = ponderador_long_total)
 
-Model 7 adds the control variables.
-With the exception of the between-effect for perceived economic inequality—which becomes nonsignificant—the within- and between-effects of the principal predictors retain both their direction and statistical significance, confirming the robustness of the associations (see Supplementary Material for effects of control variables).
+m15.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean*ola + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + perc_inequality_mean + ola_num| idencuesta),
+                data = df_study1, weights = ponderador_long_total)
+
+m16.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc*ola  + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort_cwc + ola_num| idencuesta),
+                data = df_study1, weights = ponderador_long_total)
+
+m16.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc + merit_effort_cwc  + merit_talent_cwc + perc_inequality_mean + merit_effort_mean*ola + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age + (1 + merit_effort_mean + ola_num| idencuesta),
+                data = df_study1, weights = ponderador_long_total)
+
+m17.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc*ola + perc_inequality_mean +  merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+              (1 + merit_talent_cwc + ola_num| idencuesta),
+                data = df_study1, weights = ponderador_long_total)
+
+m17.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean +  merit_effort_mean + merit_talent_mean*ola + educ + quintil1 + ess + ideo + sex + age +
+              (1 + merit_talent_mean + ola_num| idencuesta),
+                data = df_study1, weights = ponderador_long_total)
+
+m18.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc  + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+             merit_effort_cwc*perc_inequality_cwc*ola +
+              (1 + perc_inequality_cwc + merit_effort_cwc + ola_num| idencuesta),
+                data = df_study1, weights = ponderador_long_total)
+
+m18.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc  + merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+             ola*merit_effort_mean*perc_inequality_mean +
+              (1 + perc_inequality_mean + merit_effort_mean  + ola_num| idencuesta),
+                data = df_study1, weights = ponderador_long_total)
+
+m19.1 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+              merit_talent_cwc*perc_inequality_cwc*ola +
+              (1 + perc_inequality_cwc + merit_talent_cwc + ola_num| idencuesta),
+                data = df_study1, weights = ponderador_long_total)
+
+m19.2 <- lmer(mjp ~ 1 + ola_num + ola_2 + perc_inequality_cwc +  merit_effort_cwc + merit_talent_cwc + perc_inequality_mean + merit_effort_mean + merit_talent_mean + educ + quintil1 + ess + ideo + sex + age +
+              merit_talent_mean*perc_inequality_mean*ola +
+              (1 + perc_inequality_mean + merit_talent_mean + ola_num| idencuesta),
+                data = df_study1, weights = ponderador_long_total)
+
+```
 
 ```{r echo=FALSE, results='asis'}
-#| label: tbl-interactions1
+#| label: tbl-interactions_olacateg
 #| tbl-cap: "Interactions for meritocracy, perceived economic inequality and market justice preferences"
@@
   "(Intercept)" = "Intercept",
-  perc_inequality_cwc = "Perception inequality (WE)",
-  merit_effort_cwc = "Merit: Effort (WE)",
-  merit_talent_cwc = "Merit: Talent (WE)",
-  perc_inequality_mean = "Perception inequality (BE)",
-  merit_effort_mean = "Merit: Effort (BE)",
-  merit_talent_mean = "Merit: Talent (BE)",
-  "perc_inequality_cwc:merit_effort_cwc" = "Merit: Effort (WE) x Perception inequality (WE)",
-  "perc_inequality_cwc:merit_talent_cwc" = "Merit: Talent (WE) x Perception inequality (WE)",
-  "perc_inequality_mean:merit_effort_mean" = "Merit: Effort (BE) x Perception inequality (BE)",
-  "perc_inequality_mean:merit_talent_mean" = "Merit: Talent (BE) x Perception inequality (BE)"
+  "perc_inequality_cwc:ola2017" = "Wave 2017 x Perception inequality (WE)",
+  "perc_inequality_cwc:ola2018" = "Wave 2018 x Perception inequality (WE)",
+  "perc_inequality_cwc:ola2019" = "Wave 2019 x Perception inequality (WE)",
+  "perc_inequality_cwc:ola2022" = "Wave 2022 x Perception inequality (WE)",
+  "perc_inequality_cwc:ola2023" = "Wave 2023 x Perception inequality (WE)",
+
+  "merit_effort_cwc:ola2017" = "Wave 2017 x Merit: effort (WE)",
+  "merit_effort_cwc:ola2018" = "Wave 2018 x Merit: effort (WE)",
+  "merit_effort_cwc:ola2019" = "Wave 2019 x Merit: effort (WE)",
+  "merit_effort_cwc:ola2022" = "Wave 2022 x Merit: effort (WE)",
+  "merit_effort_cwc:ola2023" = "Wave 2023 x Merit: effort (WE)",
+
+  "merit_talent_cwc:ola2017" = "Wave 2017 x Merit: talent (WE)",
+  "merit_talent_cwc:ola2018" = "Wave 2018 x Merit: talent (WE)",
+  "merit_talent_cwc:ola2019" = "Wave 2019 x Merit: talent (WE)",
+  "merit_talent_cwc:ola2022" = "Wave 2022 x Merit: talent (WE)",
+  "merit_talent_cwc:ola2023" = "Wave 2023 x Merit: talent (WE)",
+
+  "perc_inequality_mean:ola2017" = "Wave 2017 x Perception inequality (BE)",
+  "perc_inequality_mean:ola2018" = "Wave 2018 x Perception inequality (BE)",
+  "perc_inequality_mean:ola2019" = "Wave 2019 x Perception inequality (BE)",
+  "perc_inequality_mean:ola2022" = "Wave 2022 x Perception inequality (BE)",
+  "perc_inequality_mean:ola2023" = "Wave 2023 x Perception inequality (BE)",
+
+  "merit_effort_mean:ola2017" = "Wave 2017 x Merit: effort (BE)",
+  "merit_effort_mean:ola2018" = "Wave 2018 x Merit: effort (BE)",
+  "merit_effort_mean:ola2019" = "Wave 2019 x Merit: effort (BE)",
+  "merit_effort_mean:ola2022" = "Wave 2022 x Merit: effort (BE)",
+  "merit_effort_mean:ola2023" = "Wave 2023 x Merit: effort (BE)",
+
+   "merit_talent_mean:ola2017" = "Wave 2017 x Merit: talent (BE)",
+  "merit_talent_mean:ola2018" = "Wave 2018 x Merit: talent (BE)",
+  "merit_talent_mean:ola2019" = "Wave 2019 x Merit: talent (BE)",
+  "merit_talent_mean:ola2022" = "Wave 2022 x Merit: talent (BE)",
+  "merit_talent_mean:ola2023" = "Wave 2023 x Merit: talent (BE)"
+
     )
 
-texreg::texreg(list(m8.1,m8.2,m9.1,m9.2),
-               custom.model.names = c("Model 8", "Model 9", "Model 10", "Model 11"),
-               caption = NULL,
-               caption.above = T,
-               stars = c(0.05, 0.01, 0.001),
-               custom.coef.map = ccoef,
-               digits = 3,
-               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.",
-               leading.zero = T,
-               use.packages = F,
-               booktabs = F,
-               scalebox = 0.8,
-               include.loglik = FALSE,
-               include.aic = FALSE,
-               center = T,
-               custom.gof.names = c("BIC", 
-                                    "Numb. obs.", 
-                                    "Num. groups: individuals", 
-                                    "Var: individuals (Intercept)", 
-                                    "Var: individuals, merit effort cwc", 
-                                    "Var: individuals, perception inequality cwc", 
-                                    "Cov: individuals (Intercept), merit effort cwc",
-                                    "Cov: individuals (Intercept), perception inequality cwc",
-                                    "Cov: individuals, merit effort cwc, perception inequality cwc",
-                                    "Var: Residuals",
-                                    "Var: individuals, merit talent cwc", 
-                                    "Cov: individuals (Intercept), merit talent cwc",
-                                    "Cov: individuals, merit talent cwc, perception inequality cwc"),
-               custom.gof.rows = list("Controls"=c(rep("Yes",4))))
-
-```
-
-```{r}
-#| label: fig-interact
-#| out-width: '100%'
-#| fig-asp: 1
-#| fig-cap: "Predicted values of market justice preferences by perceptions of meritocraticy and economic inequality"
-#| fig-cap-location: top
-#| echo: false
-#| results: asis
-
-library(ggeffects)
-
-min_val <- min(df_study1$perc_inequality_mean, na.rm = TRUE)
-max_val <- max(df_study1$perc_inequality_mean, na.rm = TRUE)
-mean_val <- mean(df_study1$perc_inequality_mean, na.rm = TRUE)
-
-p6 <- ggpredict(
-  m9.1,
-  terms = c("merit_effort_mean", 
-            sprintf("perc_inequality_mean [%.2f, %.2f, %.2f]", min_val, mean_val, max_val))
-) %>% 
-  plot() +
-  scale_x_continuous(n.breaks = 5) +
-  scale_color_manual(
-    name = "Perception of inequality (BE)",
-    values = c("#0571B0", "black", "#B2182B"),
-    labels = c("Min", "Mean", "Max")
-  ) +
-  scale_fill_manual(
-    name = "Perception of inequality (BE)",
-    values = c("#0571B0", "black", "#B2182B"),
-    labels = c("Min", "Mean", "Max")) +
-  labs(
-    x = "Merit: Effort (BE)",
-    y = "Market justice preferences",
-    title = "(A) People are rewarded by their efforts") +
-  theme_ggdist()
-
-p7 <- ggpredict(
-  m9.2,
-  terms = c("merit_talent_mean", 
-            sprintf("perc_inequality_mean [%.2f, %.2f, %.2f]", min_val, mean_val, max_val))
-) %>% 
-  plot() +
-  scale_x_continuous(n.breaks = 5) +
-  scale_color_manual(
-    name = "Perception of inequality (BE)",
-    values = c("#0571B0", "black", "#B2182B"),
-    labels = c("Min", "Mean", "Max")
-  ) +
-  scale_fill_manual(
-    name = "Perception of inequality (BE)",
-    values = c("#0571B0", "black", "#B2182B"),
-    labels = c("Min", "Mean", "Max")) +
-  labs(
-    x = "Merit: Talent (BE)",
-    y = "Market justice preferences",
-    title = "(B) People are rewarded for their intelligence",
-  caption = "Source: own elaboration with data from ELSOC 2016-2023 (n = 8460). Based in Model 10 and Model 11 in Table 4") +
-  theme_ggdist()
-
-library(ggpubr)
-ggpubr::ggarrange(p6, p7, common.legend=TRUE, ncol=1, legend = "top")
-
-```
-
-@tbl-interactions1 examines whether perceived economic inequality moderates the effect of meritocratic perceptions on market justice preferences.
-Contrary to our hypothesis, the interaction terms in the within-person specification (Models 8 and 9) do not reach statistical significance.
-However, in the between-person specification (Models 10 and 11), the interaction terms are statistically significant, suggesting that perceptions of economic inequality meaningfully alter the effect of meritocracy on support for market-based allocation of social services.
-Specifically, these positive effects indicate that, as perceptions of inequality increase, the positive association between meritocratic perceptions and market justice preferences becomes stronger.
-As shown in @fig-interact, Panel A, the effect of meritocratic perceptions about effort on market justice preferences strengthens as perceived inequality increases, particularly among individuals who perceive the highest levels of inequality, whereas this effect weakens among those who perceive low levels of inequality.
-Panel B reveals a similar pattern for meritocratic perceptions about talent, although the effect is substantially weaker at low levels of perceived inequality.
-Substantively, these findings suggest that the legitimizing power of meritocratic perceptions is amplified in contexts perceived as highly unequal, reinforcing individualistic understandings of justice and strengthening support for market-based access to essential social services.
-
-```{r echo=FALSE, results='asis'}
-#| label: tbl-interactions2
-#| tbl-cap: "Interactions for meritocracy, perceived economic inequality and market justice preferences"
-#| tbl-cap-location: top
-
-ccoef <- list(
-  "(Intercept)" = "Intercept",
-  "ola_num:perc_inequality_cwc" = "Wave x Perception inequality (WE)",
-  "ola_num:merit_effort_cwc" = "Wave x Merit: effort (WE)",
-  "ola_num:merit_talent_cwc" = "Wave x Merit: talent (WE)",
-  "ola_num:perc_inequality_mean" = "Wave x Perception inequality (BE)",
-  "ola_num:merit_effort_mean" = "Wave x Merit: effort (BE)",
-  "ola_num:merit_talent_mean" = "Wave x Merit: talent (BE)"
-    )
-
-texreg::texreg(list(m10.1, m11.1, m12.1, m10.2, m11.2, m12.2),
-               custom.model.names = c("Model 12", "Model 13", "Model 14", "Model 15", "Model 16", "Model 17"),
+texreg::texreg(list(m15.1, m16.1, m17.1, m15.2, m16.2, m17.2),
+               custom.model.names = c("Model 18", "Model 19", "Model 20", "Model 21", "Model 22", "Model 23"),
                caption = NULL,
@@
                center = T,
-               custom.gof.names = c("BIC", 
-                                    "Numb. obs.", 
-                                    "Num. groups: individuals", 
-                                    "Var: individuals (Intercept)",
-
-                                    "Var: individuals, perception inequality cwc",
-                                    "Var: individuals, wave",
-                                    "Cov: individuals (Intercept), perception inequality cwc",
-                                    "Cov: individuals (Intercept), wave",
-                                    "Cov: individuals, perception inequality cwc, wave",
-                                    "Var: Residuals",
-                                    "Var: individuals, merit effort cwc", 
-                                    "Cov: individuals (Intercept), merit effort cwc",
-                                    "Cov: individuals, merit effort cwc, wave",
-
-                                    "Var: individuals, merit talent cwc", 
-                                    "Cov: individuals (Intercept), merit talent cwc",
-                                    "Cov: individuals, merit talent cwc, wave",
-
-                                    "Var: individuals, perception inequality mean", 
-                                    "Cov: individuals (Intercept), perception inequality mean",
-                                    "Cov: individuals, perception inequality mean, wave",
-
-                                     "Var: individuals, merit effort mean", 
-                                    "Cov: individuals (Intercept), merit effort mean",
-                                    "Cov: individuals, merit effort mean, wave",
-
-                                    "Var: individuals, merit talent mean", 
-                                    "Cov: individuals (Intercept), merit talent mean",
-                                    "Cov: individuals, merit talent mean, wave"
-                                    ),
                custom.gof.rows = list("Controls"=c(rep("Yes",6))))
@@
 ```
-
-```{r}
-#| label: fig-growth
-#| out-width: '100%'
-#| fig-asp: 1
-#| fig-cap: "Time varying within effects of meritocratic perceptions on market justice preferences"
-#| fig-cap-location: top
-#| echo: false
-#| results: asis
-#| include: false
-#| eval: false
-
-library(ggeffects)
-
-p4 <- ggpredict(m11.2, terms = c("ola_num [all]", "merit_effort_mean")) %>% 
-  plot() + 
-  scale_x_continuous(n.breaks = 6) +
-  scale_color_manual(
-    name = "Merit:\nEffort",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  scale_fill_manual(
-    name = "Merit:\nEffort",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")) +
-  labs(x = "Wave",
-       title = "(A) People are rewarded for their efforts",
-       color = NULL) +
-  ggdist::theme_ggdist() +
-  theme(legend.position = "top")
-
-p5 <- ggpredict(m12.2, terms = c("ola_num [all]", "merit_talent_mean")) %>% 
-  plot() +
-  scale_x_continuous(n.breaks = 6) +
-  scale_color_manual(
-    name = "Merit:\nTalent",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  scale_fill_manual(
-    name = "Merit:\nTalent",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  labs(
-    x = "Wave",
-    title = "(B) People are rewarded for their intelligence",
-    caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)"
-  ) +
-  ggdist::theme_ggdist() +
-  theme(legend.position = "top")
-
-ggpubr::ggarrange(p4, p5, common.legend=FALSE, ncol=1)
-
-```
-
-With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time only has significant effects on the between effects of the meritocratic variables.
-These interaction effects are negative and mean that in the last waves, the impact of meritocratic variables on market justice preferences is attenuated.
-Specifically, as time advances, the positive within-person effect of perceiving that effort is rewarded (Model 16) or that intelligence is rewarded (Model 17) progressively attenuates.
-Put differently, the uplift in market justice preferences associated with a strong meritocratic perception becomes smaller in each successive wave.

### 2025-07-11 resubmission/paper-blinded-resubmit.qmd
Update carpeta resubmission y varios (archivos no renderizados)

--- 2025-07-11 paper.qmd
@@
 editor: source
-bibliography: "input/bib/merit-edjust.bib"
+bibliography: "../input/bib/merit-edjust.bib"
 # csl: "input/bib/apa6.csl"
-csl: "input/bib/harvard.csl"
-author:
-  - name: Juan Carlos Castillo
-    corresponding: true
-    orcid: 0000-0003-1265-7854
-    email: juancastillov@uchile.cl
-    affiliations:
-      - ref: 1
-      - ref: 2
-      - ref: 3
-  - name: Andreas Laffert
-    orcid: 0000-0002-9008-2454
-    email: alaffertt@estudiante.uc.cl 
-    affiliations:
-      - ref: 4
-  - name: Kevin Carrasco
-    orcid: 0000-0002-9348-8361
-    email: kevin.carrasco@ug.uchile.cl
-    affiliations:
-      - ref: 2
-  - name: Julio Iturra-Sanhueza
-    orcid: 0000-0003-1993-3479
-    email: julioiturrasanhueza@gmail.com 
-    affiliations:
-      - ref: 5     
-affiliations: 
-  - id: 1
-    name: Departamento de Sociología, Universidad de Chile
-  - id: 2
-    name: Centro de estudios del conflicto y cohesión social (COES)
-  - id: 3
-    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
-  - id: 4
-    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
-  - id: 5
-    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
+csl: "../input/bib/harvard.csl"
+# author:
+#   - name: Juan Carlos Castillo
+#     corresponding: true
+#     orcid: 0000-0003-1265-7854
+#     email: juancastillov@uchile.cl
+#     affiliations:
+#       - ref: 1
+#       - ref: 2
+#       - ref: 3
+#   - name: Andreas Laffert
+#     orcid: 0000-0002-9008-2454
+#     email: alaffertt@estudiante.uc.cl 
+#     affiliations:
+#       - ref: 4
+#   - name: Kevin Carrasco
+#     orcid: 0000-0002-9348-8361
+#     email: kevin.carrasco@ug.uchile.cl
+#     affiliations:
+#       - ref: 2
+#   - name: Julio Iturra-Sanhueza
+#     orcid: 0000-0003-1993-3479
+#     email: julioiturrasanhueza@gmail.com 
+#     affiliations:
+#       - ref: 5     
+# affiliations: 
+#   - id: 1
+#     name: Departamento de Sociología, Universidad de Chile
+#   - id: 2
+#     name: Centro de estudios del conflicto y cohesión social (COES)
+#   - id: 3
+#     name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
+#   - id: 4
+#     name: Instituto de Sociología, Pontificia Universidad Católica de Chile
+#   - id: 5
+#     name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
 citeproc: true

### 2025-07-11 resubmission/paper-resubmit.qmd
Update carpeta resubmission y varios (archivos no renderizados)

--- 2025-07-11 paper.qmd
@@
 editor: source
-bibliography: "input/bib/merit-edjust.bib"
+bibliography: "../input/bib/merit-edjust.bib"
 # csl: "input/bib/apa6.csl"
-csl: "input/bib/harvard.csl"
+csl: "../input/bib/harvard.csl"
 author:

### 2025-07-24 resubmission/02-antecedents.qmd
Seccion caso Chileno

@@
 
+## The case of Chile
+
+Chile offers an interesting case for examining how public attitudes toward the distribution of social services evolve amid declining poverty and persistently high income inequality within a residual welfare framework [@huber_democracy_2012].
+Following the neoliberal reforms of the 1980s, the Chilean welfare state has become increasingly reliant on private provision, with essential services often commodified and largely reserved for those with sufficient purchasing power [@ferre_welfare_2023].
+This model has favored the affluent, while lower-income groups are left dependent on limited and underfunded public alternatives.
+Research on social stratification characterizes Chile as a society where mobility into intermediate classes is feasible, but access to elite strata remains largely restricted [@torche_unequal_2005].
+While some indicators—such as education and occupation—suggest a relatively fluid class structure, income mobility remains constrained [@EspinozaNunez2014].
+Despite steady economic expansion, Chile continues to rank among the most unequal nations in the OECD, marked by a high Gini coefficient and substantial wealth concentration at the top [@flores_top_2019].
+In this context, Chile experienced a wave of mass protests from October 2019 to March 2020.
+Initially led by high school students, the demonstrations quickly gained momentum as broader segments of society mobilized around demands for more equality of opportunities, particularly regarding precarious access to welfare services.
+The political system interpreted this unrest as a call to reform existing political institutions, with a special focus on equalizing opportunities and providing more egalitarian access to welfare by reducing reliance on out-of-pocket mechanisms [@somma_no_2021].
+
+Concerning public opinion, the ELSOC survey shows that Chile has a low level of agreement with the idea that it is fair for those with higher incomes to buy better welfare in the areas of education, healthcare, and pensions, with average support of 8.5% between 2016 and 2023.
+At the same time, around 55% and 47% of the public between 2016 and 2023 disagreed that effort and talent are rewarded accordingly.
+This coincides with 90% of citizens agreeing that income differences in Chile are too large [@castillo_changes_2025].
+In sum, Chile represents an illustrative case of a particular form of moral economy [@koos_moral_2019], in which socioeconomic inequalities, institutional arrangements, and public attitudes coexist.
+This suggests that the country combines persistently high levels of economic and social inequality with a strong institutionalization of market-based mechanisms within the welfare system.
+At the same time, perceptions of inequality are high, and there is widespread public criticism of the commodification of social services.
+For the above reasons, we argue that the Chilean case represents a relevant context to study the role of perceived income inequality and meritocracy in shaping public preferences regarding the commodification of welfare services, particularly from the perspective of individual change.
+
 ## This study

### 2025-07-24 resubmission/02-antecedents.qmd
typos

@@
 Concerning public opinion, the ELSOC survey shows that Chile has a low level of agreement with the idea that it is fair for those with higher incomes to buy better welfare in the areas of education, healthcare, and pensions, with average support of 8.5% between 2016 and 2023.
-At the same time, around 55% and 47% of the public between 2016 and 2023 disagreed that effort and talent are rewarded accordingly.
-This coincides with 90% of citizens agreeing that income differences in Chile are too large [@castillo_changes_2025].
-In sum, Chile represents an illustrative case of a particular form of moral economy [@koos_moral_2019], in which socioeconomic inequalities, institutional arrangements, and public attitudes coexist.
+At the same time, around 55% and 47% of the public between 2016 and 2023 *disagreed* that effort and talent are rewarded accordingly.
+This coincides with 90% of citizens *agreeing* that income differences in Chile are *too large* [@castillo_changes_2025].
+
+In sum, Chile represents an illustrative case of a particular form of *moral economy* [@koos_moral_2019], in which national socioeconomic characteristics, institutional arrangements, and public attitudes coexist.
 This suggests that the country combines persistently high levels of economic and social inequality with a strong institutionalization of market-based mechanisms within the welfare system.

### 2025-07-24 resubmission/02-antecedents.qmd
ajuste hipotesis interaccion con tiempo

@@
 
-Additionally, this study examines whether these relationships vary over time in response to significant social and political events.
-In particular, we explore whether the political outburst of 2019 and the subsequent constitutional processes in Chile --- where socio-economic inequality was widely challenged --- have altered how perceptions of inequality and meritocracy shape market justice attitudes.
-We expect that after these critical events, perceptions of inequality may have a stronger negative effect on market justice preferences, reflecting increased societal questioning of market-based allocation mechanisms.
+Additionally, this study examines whether these relationships vary over time, particularly in light of broader social and political transformations.
+We are especially interested in whether changes in individuals’ perceptions of inequality and meritocracy are associated with shifts in their market justice attitudes across different time points.
+The primary focus of this study is thus to scrutinize how individual-level changes in perceptions relate to changes in preferences over time.
+While we acknowledge the relevance of contextual events — such as the 2019 political outburst or the subsequent constitutional process — for public opinion and political orientation [@disi_pavlic_justification_2025], identifying causal effects of such events (e.g., mass protest exposure) on individual attitudes lies outside the scope of this research.
+Instead, we focus on how intra-individual trajectories in perceived inequality and merit shape evolving attitudes toward the commodification of social services.
 
@@
 
-$H4$: The effects described in H1–H3 are attenuated after major social mobilizations (2019–2022), reflecting increased critical views of the market's role in allocating social goods.
+$H4a$: Changes in perceived income inequality are less strongly associated with changes in market justice attitudes over time.
+
+$H4b$: Changes in meritocracy beliefs are less strongly associated with changes in market justice attitudes over time.

### 2025-07-25 resubmission/01-introduction.qmd
Update carpeta resubmission y varios (archivos no renderizados)
intro reviewed jc- open for comments

--- 2025-04-28 01-introduction.qmd
@@
 Since 1980, economic inequality and wealth concentration have dramatically increased worldwide, becoming one of the main challenges for the social sciences.
-Globally, in 2021, less than 50% of the world population owned only 2% of the wealth, while the richest 10% concentrated 76%, and the wealthiest 1% captured nearly 38% of total assets [@chancel_world_2022].
+Globally, in 2021, less than 50% of the world population owned only 2% of the wealth, while the richest 10% concentrated 76%.
+Futhermore, the wealthiest 1% captured nearly 38% of total assets [@chancel_world_2022].
 This context of economic disparity has sparked renewed interest in studying not only the objective aspects of inequality, such as income and access to resources, but also its subjective dimensions, including perceptions, beliefs, and associated attitudes [@janmaat_subjective_2013].
-This has led, among other reasons, to widespread concern in the social sciences about the political delegitimization that economic inequality can produce [@castillo_perception_2022].
-Understanding how people perceive inequality is crucial, as these perceptions can influence how societies comprehend and justify the distribution of goods and services.
+Such dimensions are deemed as key to understand the political delegitimation that economic inequality can elicit [@castillo_perception_2022].
+Therefore, understanding how people perceive inequality is crucial, as these perceptions can influence how societies comprehend and justify (or not) inequality in the distribution of goods and services, as well as their capacity to funnel political discontent and eventually opposition to the current economic distribution.
+
+The justification of economic inequality has a long tradition in sociology, with seminal works such as the ones from Kluegel [-@kluegel_social_1995a; -@kluegel_beliefs_1981c] and Lane [@lane_political_1962; @lane_market_1986].
+Lane's main aim was to "help explain the tenacious hold of market capitalism on the public mind" (@lane_market_1986, p.384), proposing the distinction between market and political justice.
+Whereas political justice refer to government actions aligned with justice principles such equality or need, market justice involves the allocation linked to earned desserts, particularly through individual contributions based on as effort and ability.
+Sociological survey research in this area has considered questions related to the extent to which it is considered just to get better access to social services (such as pensions, health and education) based on payment capacity.
+Although such services could be considered basic citizens rights, the evidence suggest that a relevant share of the population justify inequality in these domains [@lindh_public_2015].
+In the present paper, we argue that there are two relevant factors to take into account when analyzing differences between individuals and groups to this regard: the perception of economic inequality, and the perception of meritocracy.
 
 The perception of economic inequality can be understood as an individual's subjective assessment of how resources are allocated among members of a given society [@akyelken_urban_2020].
-Regardless of their measurement, various studies have shown that this perception often underestimates the gap between the rich and the poor, which could have implications for attitudes toward the distribution of goods and services [@castillo_perception_2022; @schroder_income_2017].
-Moreover, in recent years, there has been a discussion about the subjective and objective aspects of inequality, demonstrating inconsistencies between the two [@trump_income_2018].
-This tension is relevant because advancing the study of perceptions could help understand how perceived inequality affects the distribution of resources within a society (Becker, 2020).
+Regardless of their measurement, various studies have shown that this perception often underestimates the gap between the rich and the poor, revealing inconsistencies between objective and subjective aspects of inequality [@trump_income_2018].
+Moreover, there is consistent evidence that perceived inequality, even under control of objective inequality, could have implications for attitudes toward the distribution of goods and services [@castillo_perception_2022; @schroder_income_2017].
+redistributive preferences, justification of inequality, and legitimacy of the economic system [@castillo_perception_2022; @garcia-sanchez_perceptions_2018; @garcia-sanchez_attitudes_2020].
+From this perspective, in this paper we argue that perceptions of economic inequality may play a relevant aspect when understanding market justice preferences, as a lower perceived inequality could reinforce the idea that the differences in access to social services could be justified --- a link that becomes difficult to hold when high inequality is perceived.
 
-Perception of economic inequality has been associated with important outcomes such as redistributive preferences, justification of inequality, and legitimacy of the economic system (Castillo et al., 2022 \[acá citar gente de Granada\]).
+Along with perception of inequality, a second focus of this paper is the perception of meritocracy.
+Meritocracy posits that inequality can be legitimized through distributive criteria such as effort and talent [@young_rise_1962; @davis_principles_2001].
+Previous studies have demonstrated that people with stronger meritocratic beliefs tend to perceive less inequality as they attribute economic differences to individual achievements [@mijs_paradox_2021; @wilson_role_2003] and also to justify more inequality as these beliefs are associated with attitudes that legitimize status differences [@batruch_belief_2023].
+In a context as Chile, where the distribution of goods and services is predominantly governed by market logics strongly introduced during the military dictatorship (1973-1989) [@boccardo_30_2020], meritocratic perceptions have the potential to affect the acceptance of social inequalities.
+Therefore, incorporating the perception of meritocracy in this literature allows for an understanding of how attitudes toward inequality are shaped not only by the perception of economic gaps, but also by beliefs about how these gaps are justified.
+Furthermore, these beliefs are often consolidated from an early age, reinforced by institutions promoting values such as effort and individual skills to climb socially [@castillo_socialization_2024; @reynolds_perceptions_2014].
+
+At this point we can propose the following research questions: To what extent the preferences for market justice are associated with perceptions of inequality and meritocracy?
+How stable are these associations (within subjects) along time?
+The primary objective of this study is to analyze the interplay between perceptions of inequality and meritocracy and their joint influence on preferences for market justice along time in Chile using the panel survey ELSOC (Social Chilean Longitudinal Study, 2016-2023).
+The chilean case appears as a particular case for studying market justice preferences given their story of high inequality and strong free-market policies introduced in the dictatorship period [@boccardo_30_2020; @flores_top_2020].
+Additionally, this analysis seeks to explore how political and social contingencies---such as the 2019 and 2022 social movements---might have moderated these relationships by prompting more critical reflection on the commodification of social services.
+Recognizing the temporal dimension in shaping market justice preferences is essential, given that historical and contextual factors that could challenge or reaffirm the way in which individuals persceive inequality and meritocracy, and consequantely --- following our argument --- the preferences for market justice.
+
+------------------------------------------------------------------------
+
+Papelera
+
 Research shows that greater perceived inequality fosters stronger preferences for redistribution, regardless of actual inequality levels [@garcia-sanchez_vicious_2019; @gimpelson_misperceiving_2018].
 However, other findings suggest that heightened perceptions of income gaps between high- and low-paying occupations may lead to greater justification of inequality [@Castillo2011].
+
 A less explored area in this regard concerns market justice, which denotes the degree to which individuals consider just that social goods and services (e.g., health, pensions, education) are allocated according to individual contribution, competition, and ability to pay [@lane_market_1986; @lindh_public_2015; @kluegel_legitimation_1999].
 Market justice attitudes reflect the belief that the market promotes procedural fairness---equality of opportunity---so that outcomes result from individual achievement [@lane_market_1986].
-Indeed, Lindh [-@lindh_public_2015] shows that market justice preferences are stronger in countries with higher private spending on services than in those with more comprehensive welfare systems.
-From this perspective, perceptions of economic inequality may be key, as lower perceived gaps can reinforce acceptance of free-market systems and justify existing inequalities.
+Indeed, Lindh \[-\] shows that market justice preferences are stronger in countries with higher private spending on services than in those with more comprehensive welfare systems.
 
-Meritocracy posits that inequality can be legitimized through distributive criteria such as effort and talent [@young_rise_1962; @davis_principles_2001].
-Previous studies have demonstrated that people with stronger meritocratic beliefs tend to perceive less inequality as they attribute economic differences to individual achievements [@mijs_paradox_2021; @wilson_role_2003] and also to justify more inequality as these beliefs are associated with attitudes that  legitimize status differences [@batruch_belief_2023].
-In a context like Chile, where the distribution of goods and services is predominantly governed by market logics strongly introduced during the military dictatorship (1973-1989) [@boccardo_30_2020], these beliefs can play a crucial role in the acceptance of social inequalities.
-Incorporating the perception of meritocracy in this literature allows for an understanding of how attitudes toward inequality are shaped not only by the perception of economic gaps, but also by beliefs about how these gaps are justified.
-Furthermore, these beliefs are often consolidated from an early age, reinforced by institutions promoting values such as effort and individual skills to climb socially [@castillo_socialization_2024; @reynolds_perceptions_2014].
-
+[lo de la interacción, me parece que es too much para la introducción]
 The perception of meritocracy and the perception of economic inequality would interact intricately in shaping preferences for market justice.
@@
 On the other hand, the perception of inequality could moderate the impact of meritocracy, as a higher perception of economic gaps might challenge the idea that these are solely based on merit.
-
-The primary objective of this study is to analyze the interplay between perceptions of inequality and meritocracy and their joint influence on preferences for market justice in Chile from 2016 to 2023, using longitudinal data from the ELSOC survey.
-This interaction is expected to provide a more comprehensive explanation of attitudes toward market justice in a country characterized by high inequality and strong free-market influence [@boccardo_30_2020; @flores_top_2020].
-Additionally, this analysis seeks to elucidate how political and social contingencies---such as the 2019 and 2022 social movements---might have moderated these relationships by prompting more critical reflection on the commodification of social services.
-Recognizing the temporal dimension in shaping market justice preferences is essential, given that such preferences are not static but are influenced by historical and contextual factors that challenge or reaffirm prevailing social norms.
-In this regard, variations in perceived economic inequality and meritocracy over time can affect how individuals endorse market-based approaches.
-Consequently, public opinion may shift from supporting market justice to embracing redistributive policies aimed at mitigating inequalities.

### 2025-07-28 resubmission/04-analisis.qmd
Update varios ajustes

@@
 ```
+
 # Results
@@
 
-According to Model 1, which includes the survey waves to capture intertemporal variations in the dependent variable, there is a decrease in 2017 ($\beta$ = -0.167, $p$ < .001) relative to 2016, and similarly in 2018 ($\beta$ = -0.025, $p$ > .05) and 2019 ($\beta$ = -0.046, $p$ > .05), although the latter effects are not statistically significant.
-In contrast, in the more recent waves of 2022 and 2023, there is a statistically significant increase in market justice preferences ($\beta$ = 0.285, $p$ < .001; $\beta$ = 0.284, $p$ < .001), suggesting a non-linear effect.
+According to Model 1, which includes the survey waves to capture intertemporal variations in the dependent variable, there is a decrease in 2017 ($\beta$ = -0.183, $p$ < .001) relative to 2016, and similarly in 2018 ($\beta$ = -0.009, $p$ > .05) and 2019 ($\beta$ = -0.009, $p$ > .05), although the latter effects are not statistically significant.
+In contrast, in the more recent waves of 2022 and 2023, there is a statistically significant increase in market justice preferences ($\beta$ = 0.300, $p$ < .001; $\beta$ = 0.320, $p$ < .001), suggesting a non-linear effect.
 To model this trajectory over time, Model 2 incorporates time (survey waves) as a continuous variable, along with its quadratic term, representing the non-linear association initially observed in Model 1.
@@
 The results in Model 3 suggest that the within effect of perceived economic inequality is negative and statistically significant ($p$ < .001).
-Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.039 point decrease in market justice preferences.
-Model 4 shows that meritocratic perceptions operate in the opposite direction.
-An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.070, $p$ < .001), and a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with higher market-justice preferences ($\beta$ = 0.027, $p$ < .05).
-Taken together, these results suggest that individuals who increasingly perceive meritocracy—whether through effort or talent—tend to hold stronger market justice preferences.
+Specifically, each one-point increase in an individual’s perception of economic inequality between waves is associated with a 0.027 point decrease in market justice preferences.
+Model 4 shows that meritocratic perceptions operate in distinct directions.
+An upward shift in the perception that effort is rewarded exerts a positive within effect ($\beta$ = 0.070, $p$ < .001), while a parallel increase in the perception that intelligence and ability are rewarded is likewise associated with lower market-justice preferences ($\beta$ = -0.027, $p$ < .05).
+Taken together, these results suggest that people who increasingly perceive meritocracy based on effort tend to have stronger preferences for market justice, and that the opposite is true for those who increasingly perceive meritocracy based on talent.
 
 When examining the between-group effects (BE) in Model 5 and 6, which capture differences between individuals in the average of the main variables, a similar pattern emerges.
-Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.042, $p$ < .05).
-In Model 6, the meritocratic perception that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.186, $p$ < .001), whereas the perception that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.040).
-Notably, once the meritocratic variables are included, the negative coefficient for perceived economic inequality remains but is no longer statistically significant.
+Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.002, $p$ > .05).
+However, this effects is no longer statistically significant.
+In Model 6, the meritocratic perception that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.206, $p$ < .001), whereas the perception that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.036).
 
 Model 7 adds the control variables.
-With the exception of the between-effect for perceived economic inequality—which becomes nonsignificant—the within- and between-effects of the principal predictors retain both their direction and statistical significance, confirming the robustness of the associations (see Supplementary Material for effects of control variables).
+The within- and between-effects of the principal predictors retain both their direction and statistical significance, confirming the robustness of the associations (see Supplementary Material for effects of control variables).
 
@@
 @tbl-interactions1 examines whether perceived economic inequality moderates the effect of meritocratic perceptions on market justice preferences.
-Contrary to our hypothesis, the interaction terms in the within-person specification (Models 8 and 9) do not reach statistical significance.
-However, in the between-person specification (Models 10 and 11), the interaction terms are statistically significant, suggesting that perceptions of economic inequality meaningfully alter the effect of meritocracy on support for market-based allocation of social services.
-Specifically, these positive effects indicate that, as perceptions of inequality increase, the positive association between meritocratic perceptions and market justice preferences becomes stronger.
-As shown in @fig-interact, Panel A, the effect of meritocratic perceptions about effort on market justice preferences strengthens as perceived inequality increases, particularly among individuals who perceive the highest levels of inequality, whereas this effect weakens among those who perceive low levels of inequality.
-Panel B reveals a similar pattern for meritocratic perceptions about talent, although the effect is substantially weaker at low levels of perceived inequality.
-Substantively, these findings suggest that the legitimizing power of meritocratic perceptions is amplified in contexts perceived as highly unequal, reinforcing individualistic understandings of justice and strengthening support for market-based access to essential social services.
+Contrary to our expectations, the interaction terms in the within-person specification of Model 9 indicate that the negative effect of talent-based meritocratic perceptions becomes stronger as perceived inequality increases ($\beta$ = -0.032, $p$ < .05).
+In contrast, in the between-person specification (Model 11), the interaction term is positive ($\beta$ = 0.099, $p$ < .01), suggesting that perceptions of economic inequality significantly shape the influence of meritocratic perceptions on support for market-based allocation of social services.
+Specifically, these positive effects indicate that, as perceptions of inequality increase, the association between talent-based meritocratic perceptions and market justice preferences shifts in a positive direction.
+
+The effects of effort-based meritocratic perceptions are not statistically significant in either the within- or between-person specifications (Models 8 and 10).
+However, the direction of the coefficients is negative in the former and positive in the latter.
+As shown in @fig-interact, Panel A, the between-person effect of effort-based meritocratic perceptions on market justice preferences strengthens as perceived inequality increases across individuals, whereas this effect weakens among those who perceive lower levels of inequality.
+Panel B reveals a similar pattern for the between-person effect of talent-based meritocratic perceptions, although the effect is substantially weaker at low levels of perceived inequality.
+
+Substantively, these findings suggest that the legitimizing function of meritocratic perceptions is amplified in contexts perceived as highly unequal.
+In such contexts, meritocratic narratives may reinforce individualistic understandings of distributive justice and intensify support for market-based access to essential social services.
 
@@
 
-With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time only has significant effects on the between effects of the meritocratic variables.
-These interaction effects are negative and mean that in the last waves, the impact of meritocratic variables on market justice preferences is attenuated.
-Specifically, as time advances, the positive within-person effect of perceiving that effort is rewarded (Model 16) or that intelligence is rewarded (Model 17) progressively attenuates.
-Put differently, the uplift in market justice preferences associated with a strong meritocratic perception becomes smaller in each successive wave.
+With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time has no statistically significant effects on either the within- or between-person effects of perceived inequality and meritocratic perceptions.
+The within-person interaction effects are positive, suggesting that in later waves, the influence of perceived inequality and meritocratic beliefs on market justice preferences tends to increase.
+Conversely, in the between-person specification, the interaction effects are negative, indicating that over time, the impact of perceived inequality and meritocratic beliefs on support for market-based justice tends to diminish.
+However, these trends are not statistically significant, and thus no definitive conclusions can be drawn.

### 2025-07-28 resubmission/blinded-paper-resubmit.tex
Update varios ajustes

--- 2025-05-22 paper.tex
@@
 
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{threeparttablex}
-\usepackage[normalem]{ulem}
-\usepackage{makecell}
-\usepackage{xcolor}
 \usepackage[noblocks]{authblk}
@@
 
-  \author{Juan Carlos Castillo}
-            \affil{%
-                  Departamento de Sociología, Universidad de Chile
-              }
-          \affil{%
-                  Centro de estudios del conflicto y cohesión social
-                  (COES)
-              }
-          \affil{%
-                  Núcleo milenio de desigualdades y oportunidades
-                  digitales (NUDOS)
-              }
-        \author{Andreas Laffert}
-            \affil{%
-                  Instituto de Sociología, Pontificia Universidad
-                  Católica de Chile
-              }
-        \author{Kevin Carrasco}
-            \affil{%
-                  Centro de estudios del conflicto y cohesión social
-                  (COES)
-              }
-        \author{Julio Iturra-Sanhueza}
-            \affil{%
-                  International Graduate School of Social Sciencies
-                  (BIGSSS), University of Bremen, Germany
-              }
-
 \date{}
@@
 \begin{abstract}
-This study investigates the relationship between perceptions of economic
-inequality, meritocratic beliefs, and preferences for market justice in
-Chile between 2016 and 2023. Using six waves of panel data from the
-Chilean Longitudinal Social Survey - ELSOC (\(N_{observations}\) =
-8,643; \(N_{individuals}\) = 1,687), the analysis examines how
-subjective assessments of inequality shape attitudes toward the role of
-merit in access to key social services such as healthcare, education,
-and pensions. Results indicate that rising perceptions of inequality are
-associated with greater support for redistributive policies; however,
-individuals with strong meritocratic convictions are more likely to
-legitimize existing disparities. The study also considers the influence
-of major social movements during this period, which appear to have
-reshaped public discourse and perceptions of fairness. These findings
-contribute to a deeper understanding of how beliefs about justice and
-equity evolve in contexts marked by persistent inequality and entrenched
-market-oriented frameworks. \newline \textbf{Keywords}: Economic
-inequality, meritocracy, market justice, Chile, public preferences,
-inequality perception
+Several countries have experienced a shift toward the privatization and
+commodification of public goods, welfare policies, and social services.
+In Latin America, Chile stands out as a paradigmatic case where this
+trend has led to the extensive marketization of essential services. From
+a moral economy perspective, the extent to which individuals consider it
+fair for access to such services to depend on market criteria has been
+conceptualized as market justice preferences. This study investigates
+the relationship between perceptions of economic inequality,
+meritocratic beliefs, and market justice preferences in Chile between
+2016 and 2023. Using six waves of panel data from the Chilean
+Longitudinal Social Survey -- ELSOC (\(N_{observations}\) = 8,643;
+\(N_{individuals}\) = 1,687), the analysis examines how subjective
+assessments of inequality shape attitudes toward the role of merit in
+access to key social services such as healthcare, education, and
+pensions. The findings show that greater perceived inequality is
+associated with lower market justice preferences. However, individuals
+who believe that effort is rewarded are more likely to legitimize
+existing disparities. In contrast, the perception that talent is
+rewarded shows a negative effect on market justice preferences; an
+effect that intensifies as perceived inequality increases over time. The
+study also considers the influence of major social movements during this
+period, which appear to have reshaped public discourse on justice and
+fairness. These findings contribute to a deeper understanding of how
+distributive beliefs evolve in contexts marked by persistent inequality
+and entrenched neoliberal frameworks \newline \textbf{Keywords}:
+Economic inequality, meritocracy, market justice, Chile, public
+preferences, inequality perception
 \end{abstract}
@@
 population owned only 2\% of the wealth, while the richest 10\%
-concentrated 76\%, and the wealthiest 1\% captured nearly 38\% of total
-assets (\citeproc{ref-chancel_world_2022}{Chancel et al., 2022}). This
-context of economic disparity has sparked renewed interest in studying
-not only the objective aspects of inequality, such as income and access
-to resources, but also its subjective dimensions, including perceptions,
-beliefs, and associated attitudes
-(\citeproc{ref-janmaat_subjective_2013}{Janmaat, 2013}). This has led,
-among other reasons, to widespread concern in the social sciences about
-the political delegitimization that economic inequality can produce
+concentrated 76\%. Futhermore, the wealthiest 1\% captured nearly 38\%
+of total assets (\citeproc{ref-chancel_world_2022}{Chancel et al.,
+2022}). This context of economic disparity has sparked renewed interest
+in studying not only the objective aspects of inequality, such as income
+and access to resources, but also its subjective dimensions, including
+perceptions, beliefs, and associated attitudes
+(\citeproc{ref-janmaat_subjective_2013}{Janmaat, 2013}). Such dimensions
+are deemed as key to understand the political delegitimation that
+economic inequality can elicit
 (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}).
-Understanding how people perceive inequality is crucial, as these
-perceptions can influence how societies comprehend and justify the
-distribution of goods and services.
+Therefore, understanding how people perceive inequality is crucial, as
+these perceptions can influence how societies comprehend and justify (or
+not) inequality in the distribution of goods and services, as well as
+their capacity to funnel political discontent and eventually opposition
+to the current economic distribution.
+
+The justification of economic inequality has a long tradition in
+sociology, with seminal works such as the ones from Kluegel
+(\citeproc{ref-kluegel_social_1995a}{1995};
+\citeproc{ref-kluegel_beliefs_1986}{1986}) and Lane
+(\citeproc{ref-lane_political_1962}{1962};
+\citeproc{ref-lane_market_1986}{1986}). Lane's main aim was to ``help
+explain the tenacious hold of market capitalism on the public mind''
+(\citeproc{ref-lane_market_1986}{1986, p. 384}), proposing the
+distinction between market and political justice. Whereas political
+justice refer to government actions aligned with justice principles such
+equality or need, market justice involves the allocation linked to
+earned desserts, particularly through individual contributions based on
+as effort and ability. Sociological survey research in this area has
+considered questions related to the extent to which it is considered
+just to get better access to social services (such as pensions, health
+and education) based on payment capacity
+(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
+\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
+\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
+\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015};
+\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). Although
+these services could be considered basic rights of citizens, evidence to
+date suggests that a significant portion of the population justifies
+inequality in these areas (\citeproc{ref-koos_moral_2019}{Koos \&
+Sachweh, 2019}) and that such preferences vary according to objective
+socioeconomic factors and the contextual characteristics of countries
+(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In the present paper, we
+argue that there are two other relevant factors to take into account
+when analyzing differences between individuals and groups to this
+regard: the perception of economic inequality and the perception of
+meritocracy.
 
@@
 this perception often underestimates the gap between the rich and the
-poor, which could have implications for attitudes toward the
-distribution of goods and services
+poor, revealing inconsistencies between objective and subjective aspects
+of inequality (\citeproc{ref-trump_income_2018}{Trump, 2018}). Moreover,
+there is consistent evidence that perceived inequality, even under
+control of objective inequality, could have implications for attitudes
+toward the distribution of goods and services
 (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
-\citeproc{ref-schroder_income_2017}{Schröder, 2017}). Moreover, in
-recent years, there has been a discussion about the subjective and
-objective aspects of inequality, demonstrating inconsistencies between
-the two (\citeproc{ref-trump_income_2018}{Trump, 2018}). This tension is
-relevant because advancing the study of perceptions could help
-understand how perceived inequality affects the distribution of
-resources within a society (Becker, 2020).
-
-Perception of economic inequality has been associated with important
-outcomes such as redistributive preferences, justification of
-inequality, and legitimacy of the economic system
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
+\citeproc{ref-schroder_income_2017}{Schröder, 2017}), redistributive
+preferences, justification of inequality, and legitimacy of the economic
+system (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
 \citeproc{ref-garcia-sanchez_attitudes_2020}{García-Sánchez et al.,
-2020}, \citeproc{ref-garcia-sanchez_perceptions_2018}{2018}). Research
-shows that greater perceived inequality fosters stronger preferences for
-redistribution, regardless of actual inequality levels
-(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
-Treisman, 2018}). However, other findings suggest that heightened
-perceptions of income gaps between high- and low-paying occupations may
-lead to greater justification of inequality
-(\citeproc{ref-Castillo2011}{Castillo, 2011}). A less explored area in
-this regard concerns market justice, which denotes the degree to which
-individuals consider just that social goods and services (e.g., health,
-pensions, education) are allocated according to individual contribution,
-competition, and ability to pay
-(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
-\citeproc{ref-lane_market_1986}{Lane, 1986};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Market justice attitudes
-reflect the belief that the market promotes procedural
-fairness---equality of opportunity---so that outcomes result from
-individual achievement (\citeproc{ref-lane_market_1986}{Lane, 1986}).
-Indeed, Lindh (\citeproc{ref-lindh_public_2015}{2015}) shows that market
-justice preferences are stronger in countries with higher private
-spending on services than in those with more comprehensive welfare
-systems. From this perspective, perceptions of economic inequality may
-be key, as lower perceived gaps can reinforce acceptance of free-market
-systems and justify existing inequalities.
-
-Meritocracy posits that inequality can be legitimized through
-distributive criteria such as effort and talent
+2020}, \citeproc{ref-garcia-sanchez_perceptions_2018}{2018}). From this
+perspective, in this paper we argue that perceptions of economic
+inequality may play a relevant aspect for understanding market justice
+preferences, as a lower perceived inequality could reinforce the idea
+that the differences in access to social services could be justified ---
+a link that becomes difficult to hold when high inequality is perceived.
+
+Along with perception of inequality, a second focus of this paper is the
+perception of meritocracy. Meritocracy posits that inequality can be
+legitimized through distributive criteria such as effort and talent
 (\citeproc{ref-davis_principles_2001}{Davis \& Moore, 2001};
@@
 (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In a context
-like Chile, where the distribution of goods and services is
-predominantly governed by market logics strongly introduced during the
-military dictatorship (1973-1989)
-(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020}), these beliefs can
-play a crucial role in the acceptance of social inequalities.
-Incorporating the perception of meritocracy in this literature allows
-for an understanding of how attitudes toward inequality are shaped not
-only by the perception of economic gaps, but also by beliefs about how
-these gaps are justified. Furthermore, these beliefs are often
-consolidated from an early age, reinforced by institutions promoting
-values such as effort and individual skills to climb socially
+as Chile, where the distribution of goods and services is predominantly
+governed by market logics strongly introduced during the military
+dictatorship (1973-1989) (\citeproc{ref-boccardo_30_2020}{Boccardo,
+2020}; \citeproc{ref-madariaga_three_2020}{Madariaga, 2020}),
+meritocratic perceptions have the potential to affect the acceptance of
+social inequalities. Therefore, incorporating the perception of
+meritocracy in this literature allows for an understanding of how
+attitudes toward inequality are shaped not only by the perception of
+economic gaps, but also by beliefs about how these gaps are justified.
+Furthermore, these beliefs are often consolidated from an early age,
+reinforced by institutions promoting values such as effort and
+individual skills to climb socially
 (\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
@@
 
-The perception of meritocracy and the perception of economic inequality
-would interact intricately in shaping preferences for market justice. On
-the one hand, a stronger perception of merit as a criterion for
-distribution could minimize the perception of economic gaps, thus
-justifying an unequal system
-(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
-\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). On the other hand, the
-perception of inequality could moderate the impact of meritocracy, as a
-higher perception of economic gaps might challenge the idea that these
-are solely based on merit.
-
-The primary objective of this study is to analyze the interplay between
-perceptions of inequality and meritocracy and their joint influence on
-preferences for market justice in Chile from 2016 to 2023, using
-longitudinal data from the ELSOC survey. This interaction is expected to
-provide a more comprehensive explanation of attitudes toward market
-justice in a country characterized by high inequality and strong
-free-market influence (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
-\citeproc{ref-flores_top_2020}{Flores et al., 2020}). Additionally, this
-analysis seeks to elucidate how political and social
+At this point we can propose the following research questions: To what
+extent the preferences for market justice are associated with
+perceptions of inequality and meritocracy? How stable are these
+associations (within subjects) along time? The primary objective of this
+study is to analyze the interplay between perceptions of inequality and
+meritocracy and their joint influence on preferences for market justice
+along time in Chile using the panel survey ELSOC (Social Chilean
+Longitudinal Study, 2016-2023). The chilean case appears as a particular
+case for studying market justice preferences given their story of high
+inequality and strong free-market policies introduced in the
+dictatorship period (\citeproc{ref-flores_top_2020}{Flores et al.,
+2020}; \citeproc{ref-madariaga_three_2020}{Madariaga, 2020}).
+Additionally, this analysis seeks to explore how political and social
 contingencies---such as the 2019 and 2022 social movements---might have
 moderated these relationships by prompting more critical reflection on
-the commodification of social services. Recognizing the temporal
-dimension in shaping market justice preferences is essential, given that
-such preferences are not static but are influenced by historical and
-contextual factors that challenge or reaffirm prevailing social norms.
-In this regard, variations in perceived economic inequality and
-meritocracy over time can affect how individuals endorse market-based
-approaches. Consequently, public opinion may shift from supporting
-market justice to embracing redistributive policies aimed at mitigating
-inequalities.
+the commodification of social services
+(\citeproc{ref-somma_no_2021}{Somma et al., 2021}). Recognizing the
+temporal dimension in shaping market justice preferences is essential,
+given that historical and contextual factors that could challenge or
+reaffirm the way in which individuals persceive inequality and
+meritocracy, and consequantely --- following our argument --- the
+preferences for market justice.
 
@@
 treated fairly because they face the same formal rules of engagement
-(\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently, the
-legitimacy of market justice stems from the assumption that inequalities
-are not only inevitable but fair---so long as the rules are transparent
-and opportunities are open. In this way, market justice provides a moral
-justification for inequality by framing it as a necessary outcome of
-individual responsibility.
+(\citeproc{ref-lane_market_1986}{Robert E. Lane, 1986}). Consequently,
+the legitimacy of market justice stems from the assumption that
+inequalities are not only inevitable but fair---so long as the rules are
+transparent and opportunities are open. In this way, market justice
+provides a moral justification for inequality by framing it as a
+necessary outcome of individual responsibility
+(\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}).
 
@@
 support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
-\& Sachweh, 2019}; \citeproc{ref-svallfors_political_2007}{Svallfors,
-2007}). This tendency reflects not only material self-interest but also
-a broader moral economy, in which winners of the market system
-internalize justifications for the status quo. At the same time,
-political ideology also plays a role ---such as economically
-conservative values --- where more right-wing individuals show higher
-support for meritocracy and more skepticism towards redistribution. This
-is particularly salient in countries with more restricted public
-provision of social services. For example, Castillo et al.
+\& Sachweh, 2019}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
+\citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). This tendency
+reflects not only material self-interest but also a broader moral
+economy, in which winners of the market system internalize
+justifications for the status quo. At the same time, political ideology
+also plays a role ---such as economically conservative values --- where
+more right-wing individuals show higher support for meritocracy and more
+skepticism towards redistribution. This is particularly salient in
+countries with more restricted public provision of social services. For
+example, Castillo et al.
 (\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
 Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
-those with right-leaning individuals are more supportive of market-based
-distribution of welfare. Beyond individual characteristics,
-country-level institutions also play a central role. In liberal welfare
-regimes like those of the United States or the United Kingdom, market
-justice preferences are more widespread, while in coordinated or
+right-wing individuals are more supportive of market-based distribution
+of welfare. Beyond individual characteristics, country-level
+institutions also play a central role. In liberal welfare regimes like
+those of the United States or the United Kingdom, market justice
+preferences are more widespread, while in coordinated or
 social-democratic regimes---such as Sweden or Germany---citizens are
@@
 
-Perceptions of inequality have been associated with attitudes about
-market justice (\citeproc{ref-kluegel_social_1995a}{Kluegel et al.,
-1995}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}). Research
-indicates that lower perceived inequality can reinforce support for
-market-based distributive arrangements by suggesting that the system is
-fair and that outcomes reflect effort and ability
-(\citeproc{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when inequality
-is perceived as excessive or structurally determined, individuals could
-question the legitimacy of market justice and become more supportive of
-redistributive policies
-(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022};
-\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
+Perceptions of inequality have been associated with a range of
+distributive attitudes (\citeproc{ref-castillo_perception_2022}{Castillo
+et al., 2022}; \citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez
+et al., 2019}; \citeproc{ref-mccall_exposure_2017}{McCall et al., 2017};
+\citeproc{ref-trump_income_2018}{Trump, 2018}). Research indicates that
+perceptions of inequality influence redistributive preferences to a
+greater extent than objective or actual economic inequality
+(\citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \& Treisman,
+2018}). Others have shown that the perception of inequality, through the
+perceived wage gap, is related to the justification of greater
+inequality (\citeproc{ref-Castillo2011}{Castillo, 2011}). In that sense,
+lower perceived inequality can reinforce support for market-based
+distributive arrangements by suggesting that the system is fair and that
+outcomes reflect effort and ability (\citeproc{ref-kuhn_eye_2011}{Kuhn,
+2011}). In contrast, when inequality is perceived as excessive or
+structurally determined, individuals could question the legitimacy of
+market justice and become more supportive of redistributive policies
+(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
 2019}).
@@
 Perceptions of inequality refer to individuals' subjective evaluations
-of the extent, causes, and consequences of income and wealth
-disparities. Unlike objective measures such as the Gini index, perceived
-inequality captures how individuals make sense of distributive
-hierarchies in their everyday lives, shaped by reference groups, social
-comparisons, and information environments
-(\citeproc{ref-garcia-castro_perceiving_2020}{García-Castro et al.,
-2020}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
+of the extent, causes, and consequences of income and wealth disparities
+(\citeproc{ref-akyelken_urban_2020}{Akyelken, 2020}). Unlike objective
+measures such as the Gini index, perceived inequality captures how
+individuals make sense of distributive hierarchies in their everyday
+lives, shaped by reference groups, social comparisons, and information
+environments (\citeproc{ref-garcia-castro_perceiving_2020}{García-Castro
+et al., 2020}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
 Treisman, 2018}; \citeproc{ref-mijs_stratified_2016}{Mijs, 2016}).
@@
 In addition to influencing individual attitudes toward inequality,
-meritocratic beliefs can contribute to social division and the
+meritocratic perceptions can contribute to social division and the
 stigmatization of disadvantaged groups. Recent research has demonstrated
@@
 
+\subsection{The Chilean context}\label{the-chilean-context}
+
+Chile offers an interesting case for examining how public attitudes
+toward the distribution of social services evolve amid declining poverty
+and persistently high income inequality within a residual welfare
+framework (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). Following
+the neoliberal reforms of the 1980s, the Chilean welfare state has
+become increasingly reliant on private provision, with essential
+services often commodified and largely reserved for those with
+sufficient purchasing power (\citeproc{ref-boccardo_30_2020}{Boccardo,
+2020}; \citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). This model
+has favored the affluent, while lower-income groups are left dependent
+on limited and underfunded public alternatives. Research on social
+stratification characterizes Chile as a society where mobility into
+intermediate classes is feasible, but access to elite strata remains
+largely restricted
+(\citeproc{ref-lopez-roldan_comparative_2021}{López-Roldán \& Fachelli,
+2021}; \citeproc{ref-torche_intergenerational_2014a}{Torche, 2014}).
+While some indicators---such as education and occupation---suggest a
+relatively fluid class structure, income mobility remains constrained
+(\citeproc{ref-espinoza_movilidad_2014}{Espinoza \& Núñez, 2014}).
+
+Despite steady economic expansion, Chile continues to rank among the
+most unequal nations in the OECD, marked by a high Gini coefficient and
+substantial wealth concentration at the top
+(\citeproc{ref-flores_top_2020}{Flores et al., 2020}). In this context,
+Chile experienced a wave of mass protests from October 2019 to March
+2020. Initially led by high school students, the demonstrations quickly
+gained momentum as broader segments of society mobilized around demands
+for more equality of opportunities, particularly regarding precarious
+access to welfare services. The political system interpreted this unrest
+as a call to reform existing political institutions, with a special
+focus on equalizing opportunities and providing more egalitarian access
+to welfare by reducing reliance on out-of-pocket mechanisms
+(\citeproc{ref-somma_no_2021}{Somma et al., 2021}).
+
+Concerning public opinion, the ELSOC survey shows that Chile has a low
+level of agreement with the idea that it is fair for those with higher
+incomes to buy better welfare in the areas of education, healthcare, and
+pensions, with average support of 8.5\% between 2016 and 2023. At the
+same time, around 55\% and 47\% of the public between 2016 and 2023
+\emph{disagreed} that effort and talent are rewarded accordingly. This
+coincides with 90\% of citizens \emph{agreeing} that income differences
+in Chile are \emph{too large}
+(\citeproc{ref-castillo_changes_2025}{Castillo et al., 2025}).
+
+In sum, Chile represents an illustrative case of a particular form of
+\emph{moral economy} (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh,
+2019}), in which national socioeconomic characteristics, institutional
+arrangements, and public attitudes coexist. This suggests that the
+country combines persistently high levels of economic and social
+inequality with a strong institutionalization of market-based mechanisms
+within the welfare system
+(\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). At the same
+time, perceptions of inequality are high, and there is widespread public
+criticism of the commodification of social services
+(\citeproc{ref-somma_no_2021}{Somma et al., 2021}). For the above
+reasons, we argue that the Chilean case represents a relevant context to
+study the role of perceived income inequality and meritocracy in shaping
+public preferences regarding the commodification of welfare services,
+particularly from the perspective of individual change.
+
 \subsection{This study}\label{this-study}
@@
 between individuals' perceptions of economic inequality and their
-beliefs in meritocracy. Specifically, we argue that both perceptions
+perceptions of meritocracy. Specifically, we argue that both perceptions
 independently and interactively influence the extent to which
@@
 
-First, consistent with previous findings, we expect that a higher
-perception of economic inequality will be associated with lower market
-justice preferences. When individuals perceive smaller income gaps, they
-are more likely to view market mechanisms as fair and legitimate,
-reinforcing the acceptance of outcomes based on competition and ability
-to pay. Conversely, a heightened perception of inequality may erode
-confidence in market fairness, weakening support for market-based
-distribution. This relationship is particularly relevant in the context
-of Chile, where the neoliberal economic model has been a dominant force
-in shaping public attitudes toward inequality and justice.
+First, consistent with previous findings about redistributive
+preferences (\citeproc{ref-castillo_perception_2022}{Castillo et al.,
+2022}), we expect that a higher perception of economic inequality will
+be associated with lower market justice preferences. When individuals
+perceive smaller income gaps, they are more likely to view market
+mechanisms as fair and legitimate, reinforcing the acceptance of
+outcomes based on competition and ability to pay. Conversely, a
+heightened perception of inequality may erode confidence in market
+fairness, weakening support for market-based distribution. This
+relationship is particularly relevant in the context of Chile, where the
+neoliberal economic model has been a dominant force in shaping public
+attitudes toward inequality and justice
+(\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021};
+\citeproc{ref-mac-clure_justicia_2024}{Mac-Clure et al., 2024}).
 
@@
 unequal outcomes and endorse the notion that markets allocate resources
-fairly according to individual merit. This aligns with the idea that
-meritocratic beliefs serve as a moral framework that legitimizes
-market-based inequalities, as individuals perceive the system as just
-when they believe that rewards are based on individual merit. This is
-particularly relevant in contexts where neoliberal ideologies dominate,
-as they often emphasize individual responsibility and competition as the
-basis for social order.
+fairly according to individual merit
+(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
+This aligns with the idea that meritocratic perceptions serve as a moral
+framework that legitimizes market-based inequalities, as individuals
+perceive the system as just when they believe that rewards are based on
+individual merit. This is particularly relevant in contexts where
+neoliberal ideologies dominate, as they often emphasize individual
+responsibility and competition as the basis for social order.
 
@@
 economic inequality: when perceived inequality is low, the positive
-association between meritocratic beliefs and market justice preferences
-will be stronger. However, when perceived inequality is high, this
-association will weaken, as greater awareness of large economic gaps may
-challenge the view that outcomes are purely merit-based.
+association between meritocratic perceptions and market justice
+preferences will be stronger. However, when perceived inequality is
+high, this association will weaken, as greater awareness of large
+economic gaps may challenge the view that outcomes are purely
+merit-based.
 
 Additionally, this study examines whether these relationships vary over
-time in response to significant social and political events. In
-particular, we explore whether the political outburst of 2019 and the
-subsequent constitutional processes in Chile --- where socio-economic
-inequality was widely challenged --- have altered how perceptions of
-inequality and meritocracy shape market justice attitudes. We expect
-that after these critical events, perceptions of inequality may have a
-stronger negative effect on market justice preferences, reflecting
-increased societal questioning of market-based allocation mechanisms.
+time, particularly in light of broader social and political
+transformations. We are especially interested in whether changes in
+individuals' perceptions of inequality and meritocracy are associated
+with shifts in their market justice attitudes across different time
+points. The primary focus of this study is thus to scrutinize how
+individual-level changes in perceptions relate to changes in preferences
+over time. While we acknowledge the relevance of contextual events ---
+such as the 2019 political outburst or the subsequent constitutional
+process --- for public opinion and political orientation
+(\citeproc{ref-disipavlic_justification_2025}{Disi Pavlic et al.,
+2025}), identifying causal effects of such events (e.g., mass protest
+exposure) on individual attitudes lies outside the scope of this
+research. Instead, we focus on how intra-individual trajectories in
+perceived inequality and merit shape evolving attitudes toward the
+commodification of social services.
 
@@
 
-\(H1\): Higher perceived economic inequality is associated with less
-market justice preferences.
-
-\(H2\): Higher meritocratic beliefs are associated with higher market
-justice preferences.
-
-\(H3\): The positive association between meritocratic beliefs and market
-justice preferences is moderated by perceived economic inequality;
-specifically, this association is weaker when perceived economic
-inequality is high.
-
-\(H4\): The effects described in H1--H3 are attenuated after major
-social mobilizations (2019--2022), reflecting increased critical views
-of the market's role in allocating social goods.
+\(H1_{a} \; (\text{Between-person}):\) Individuals who perceive higher
+levels of economic inequality will report lower support for market
+justice preferences, on average, compared to those who perceive lower
+levels of inequality.
+
+\(H1_{b} \; (\text{Within-person}):\) Increases in an individual's
+perception of economic inequality over time will be associated with
+decreases in their support for market justice preferences.
+
+\(H2_{a} \; (\text{Between-person}):\) Individuals who endorse stronger
+meritocratic perceptions will report higher support for market justice
+preferences, on average, compared to those with weaker meritocratic
+perceptions
+
+\(H2_{b} \; (\text{Within-person}):\) Increases in an individual's
+meritocratic perceptions over time will be associated with increases in
+their support for market justice preferences.
+
+\(H3_{a} \; (\text{Between-person}):\) The positive association between
+meritocratic perceptions and market justice preferences will be weaker
+among individuals who perceive higher levels of economic inequality,
+compared to those who perceive lower inequality.
+
+\(H3_{b} \; (\text{Within-person}):\) The positive association between
+meritocratic perceptions and market justice preferences will be weaker
+at times when individuals perceive higher levels of inequality, relative
+to times when they perceive lower inequality.
+
+\(H4_{a} \; (\text{Growth moderation – Inequality}):\) The longitudinal
+association between perceived economic inequality and market justice
+preferences will strengthen over time, both within and between
+individuals.
+
+\(H4_{b} \; (\text{Growth moderation – Meritocracy}):\) The longitudinal
+association between meritocratic perceptions and market justice
+preferences will weaken over time, both within and between individuals.
 
@@
 This study draws on data from the Chilean Longitudinal Social Survey
-(ELSOC), a panel study conducted annually from 2016 to 2023. The survey
-assesses how individuals think, feel, and behave regarding social
-conflict and cohesion in Chile. ELSOC employs a probabilistic,
-stratified, clustered, multistage sampling design covering major urban
-centers (Santiago, Valparaíso, and Concepción) as well as smaller
-cities. The target population includes men and women aged 18 to 75 who
-are habitual residents of private dwellings.
+(ELSOC), a nationally representative panel study of the urban adult
+population in Chile, conducted annually between 2016 and 2023. Designed
+to examine individuals' attitudes, emotions, and behaviors regarding
+social conflict and cohesion, ELSOC employs a probabilistic, stratified,
+clustered, and multistage sampling design covering both major urban
+centers and smaller cities. The sampling frame was proportionally
+stratified into six categories of urban population size (e.g., large and
+small cities), followed by the random selection of households within
+1,067 city blocks. The target population includes men and women aged 18
+to 75 who are habitual residents of private dwellings.
 
 The survey has been conducted every year since 2016, except in 2020,
-when it was suspended due to the COVID-19 pandemic. The first wave
-included 2,927 participants from both northern and southern regions,
-covering 77\% of the national population and 93\% of the urban
-population, with a response rate of 62.4\%
-(\citeproc{ref-elsoc_estudio_2022}{ELSOC, 2022}). This study uses six
+when it was suspended due to the COVID-19 pandemic. This study uses six
 waves: 2016, 2017, 2018, 2019, 2022, and 2023. The 2021 wave was
@@
 
-\textbf{Market justice preferences}: The dependent variable in this
+\textbf{Market justice preferences}: the dependent variable in this
 study is preferences for market justice. This construct is
 operationalized through three items that capture how strongly
-individuals justify conditioning access to social services---healthcare,
+individuals justify conditioning access to core services---healthcare,
 pensions, and education--- basen on individual income. Specifically, the
@@
 
-\textbf{Perception of economic inequality}: The main independent
-variable refers to the perception of economic inequality, measured
+\textbf{Perception of economic inequality}: this variable is measured
 through the perceived wage gap
@@
 income quintile, educational level (1=Less than Universitary,
-2=Universitary), age (in years), and sex (1=Male, 2=Female), which have
+2=Universitary), age (in ranges), and sex (1=Male, 2=Female), which have
 been previously shown to influence market justice preferences
@@
 2=Center, 3=Right, 4=No identification) and subjective social status
-(ranging from 1 to 10) as they may affect the relationship between
-market justice preferences, perceptions of inequality, and meritocracy
+(measured through a scale from 1 to 10) as they may affect the
+relationship between market justice preferences, perceptions of
+inequality, and meritocracy
 (\citeproc{ref-schneider_poverty_2015}{Schneider \& Castillo, 2015}).
+Descriptive statistics for the control variables can be found in the
+supplementary material.
 
@@
 
-\subsection{Descriptive}\label{descriptive}
+\subsection{Descriptive statistics}\label{descriptive-statistics}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-alluvial-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{blinded-paper-resubmit_files/figure-pdf/fig-alluvial-1.pdf}
 
@@
 Regarding the main dependent and independent variables of this study,
-Figure~\ref{fig-meanchange} depicts their average changes over the
-years. We observe an increase in the average level of market justice
-preferences in the most recent waves, which begins in 2019. The highest
-average consistently appears for perceived economic inequality, although
-this variable shows a downward trend of roughly one point over time.
-Interestingly, while perceptions of economic inequality declined in the
-latest measurements (2022--2023), market justice preferences increased.
-The meritocracy measures remain stable, though the perception that
-individuals are rewarded for intelligence is slightly higher than the
-perception that they are rewarded for their effort.
+Figure~\ref{fig-meanchange} depicts their standardized average values
+across survey waves. The results show a notable upward trend in market
+justice preferences, particularly in the most recent waves. Perceived
+economic inequality exhibited the highest standardized mean in 2016;
+however, this variable shows a general downward trajectory over time,
+with a temporary increase in 2019, possibly reflecting the effects of
+the social uprising that year. Interestingly, while perceptions of
+inequality declined in the latest waves (2022--2023), market justice
+preferences continued to rise. Meritocratic beliefs, in contrast, remain
+relatively stable overall. Nevertheless, perceptions that individuals
+are rewarded based on talent tend to be slightly higher than those based
+on effort. Both meritocracy-related measures follow a similar temporal
+pattern: an increase from 2016 to 2018, a decline between 2019 and
+2022---potentially associated with the social unrest and the
+consequences of the COVID-19 pandemic---and a subsequent rise in 2023,
+returning to levels comparable to those observed in 2016.
 
@@
 
-\caption{\label{fig-meanchange}Change in the mean of market justice
-preferences, economic inequality perception, and meritocracy
+\caption{\label{fig-meanchange}Change in the standarized mean of market
+justice preferences, economic inequality perception, and meritocracy
 (2016-2023)}
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-meanchange-1.pdf}
-
-}
-
-\end{figure}%
-
-Figure~\ref{fig-matrix} presents the correlation matrix for the main
-variables in the latest wave (2023). Overall, the coefficients range
-from low to moderate. The association between market justice preferences
-and economic inequality perception is negative but small and
-statistically significant (\(r\) = -0.05, \(p\) \textless{} .05). By
-contrast, market justice preferences positively and significantly
-correlate with the two meritocracy variables (\(r\) = 0.15, \(p\)
-\textless{} .01; \(r\) = 0.13, \(p\) \textless{} .01). Perceived
-economic inequality has negative and nonsignificant correlations with
-both meritocracy perceptions (\(r\) = -0.04, \(p\) \textgreater{} .05;
-\(r\) = -0.04, \(p\) \textgreater{} .05). Finally, the two meritocracy
-variables exhibit a strong positive association with each other (\(r\) =
-0.69, \(p\) \textless{} .01).
-
-\begin{figure}[H]
-
-\caption{\label{fig-matrix}Correlation matrix of the main variables for
-the last wave (2023)}
-
-\centering{
-
-\includegraphics[width=0.9\textwidth,height=\textheight]{paper_files/figure-pdf/fig-matrix-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{blinded-paper-resubmit_files/figure-pdf/fig-meanchange-1.pdf}
 
@@
 from the empty model (see Supplementary Material), which decomposes the
-variance of market justice preferences, is 0.23, indicating that
-approximately 23\% of the variation is attributable to differences
-between individuals. Complementary, 77\% of the variation corresponds to
+variance of market justice preferences, is 0.31, indicating that
+approximately 31\% of the variation is attributable to differences
+between individuals. Complementary, 69\% of the variation corresponds to
 within-individual differences over time.
@@
 \hline
-Intercept                          & $1.968^{***}$  & $2.015^{***}$  & $2.030^{***}$  & $2.028^{***}$  & $2.169^{***}$  & $1.456^{***}$  & $1.547^{***}$  \\
-                                   & $(0.021)$      & $(0.035)$      & $(0.035)$      & $(0.035)$      & $(0.070)$      & $(0.097)$      & $(0.115)$      \\
+Intercept                          & $1.938^{***}$  & $1.948^{***}$  & $1.965^{***}$  & $1.967^{***}$  & $1.974^{***}$  & $1.186^{***}$  & $1.250^{***}$  \\
+                                   & $(0.023)$      & $(0.037)$      & $(0.037)$      & $(0.037)$      & $(0.087)$      & $(0.124)$      & $(0.144)$      \\
 Wave (Ref.= 2016)                  &                &                &                &                &                &                &                \\
                                    &                &                &                &                &                &                &                \\
-\quad Wave 2017                    & $-0.167^{***}$ &                &                &                &                &                &                \\
-                                   & $(0.027)$      &                &                &                &                &                &                \\
-\quad Wave 2018                    & $-0.025$       &                &                &                &                &                &                \\
-                                   & $(0.027)$      &                &                &                &                &                &                \\
-\quad Wave 2019                    & $-0.046$       &                &                &                &                &                &                \\
-                                   & $(0.027)$      &                &                &                &                &                &                \\
-\quad Wave 2022                    & $0.285^{***}$  &                &                &                &                &                &                \\
-                                   & $(0.028)$      &                &                &                &                &                &                \\
-\quad Wave 2023                    & $0.284^{***}$  &                &                &                &                &                &                \\
-                                   & $(0.027)$      &                &                &                &                &                &                \\
-Wave                               &                & $-0.121^{***}$ & $-0.125^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.128^{***}$ & $-0.129^{***}$ \\
-                                   &                & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      & $(0.022)$      \\
-Wave$^2$                           &                & $0.028^{***}$  & $0.028^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  & $0.029^{***}$  \\
+\quad Wave 2017                    & $-0.183^{***}$ &                &                &                &                &                &                \\
+                                   & $(0.025)$      &                &                &                &                &                &                \\
+\quad Wave 2018                    & $-0.009$       &                &                &                &                &                &                \\
+                                   & $(0.025)$      &                &                &                &                &                &                \\
+\quad Wave 2019                    & $-0.009$       &                &                &                &                &                &                \\
+                                   & $(0.025)$      &                &                &                &                &                &                \\
+\quad Wave 2022                    & $0.300^{***}$  &                &                &                &                &                &                \\
+                                   & $(0.025)$      &                &                &                &                &                &                \\
+\quad Wave 2023                    & $0.320^{***}$  &                &                &                &                &                &                \\
+                                   & $(0.025)$      &                &                &                &                &                &                \\
+Wave                               &                & $-0.088^{***}$ & $-0.095^{***}$ & $-0.096^{***}$ & $-0.096^{***}$ & $-0.096^{***}$ & $-0.096^{***}$ \\
+                                   &                & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      & $(0.020)$      \\
+Wave$^2$                           &                & $0.024^{***}$  & $0.024^{***}$  & $0.025^{***}$  & $0.025^{***}$  & $0.025^{***}$  & $0.025^{***}$  \\
                                    &                & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      & $(0.003)$      \\
-Perception inequality (WE)         &                &                & $-0.039^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ & $-0.034^{***}$ \\
-                                   &                &                & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
-Merit: Effort (WE)                 &                &                &                & $0.070^{***}$  & $0.070^{***}$  & $0.069^{***}$  & $0.069^{***}$  \\
-                                   &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
-Merit: Talent (WE)                 &                &                &                & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    & $0.027^{*}$    \\
-                                   &                &                &                & $(0.012)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
-Perception inequality (BE)         &                &                &                &                & $-0.042^{*}$   & $-0.006$       & $-0.033$       \\
-                                   &                &                &                &                & $(0.018)$      & $(0.018)$      & $(0.018)$      \\
-Merit: Effort (BE)                 &                &                &                &                &                & $0.186^{***}$  & $0.178^{***}$  \\
-                                   &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
-Merit: Talent (BE)                 &                &                &                &                &                & $0.040$        & $0.019$        \\
-                                   &                &                &                &                &                & $(0.032)$      & $(0.031)$      \\
+Perception inequality (WE)         &                &                & $-0.027^{**}$  & $-0.025^{**}$  & $-0.025^{**}$  & $-0.025^{**}$  & $-0.025^{**}$  \\
+                                   &                &                & $(0.009)$      & $(0.009)$      & $(0.009)$      & $(0.009)$      & $(0.009)$      \\
+Merit: Effort (WE)                 &                &                &                & $0.070^{***}$  & $0.070^{***}$  & $0.070^{***}$  & $0.070^{***}$  \\
+                                   &                &                &                & $(0.011)$      & $(0.011)$      & $(0.011)$      & $(0.011)$      \\
+Merit: Talent (WE)                 &                &                &                & $-0.027^{*}$   & $-0.027^{*}$   & $-0.027^{*}$   & $-0.027^{*}$   \\
+                                   &                &                &                & $(0.011)$      & $(0.011)$      & $(0.011)$      & $(0.011)$      \\
+Perception inequality (BE)         &                &                &                &                & $-0.002$       & $0.043$        & $0.008$        \\
+                                   &                &                &                &                & $(0.023)$      & $(0.023)$      & $(0.024)$      \\
+Merit: Effort (BE)                 &                &                &                &                &                & $0.206^{***}$  & $0.191^{***}$  \\
+                                   &                &                &                &                &                & $(0.041)$      & $(0.040)$      \\
+Merit: Talent (BE)                 &                &                &                &                &                & $0.036$        & $0.021$        \\
+                                   &                &                &                &                &                & $(0.040)$      & $(0.040)$      \\
 \hline
 Controls                           & No             & No             & No             & No             & No             & No             & Yes            \\
-BIC                                & $20780.594$    & $20780.617$    & $20781.113$    & $20726.753$    & $20736.627$    & $20655.172$    & $20749.047$    \\
+BIC                                & $32146.711$    & $31406.958$    & $31414.699$    & $31404.308$    & $31419.062$    & $31366.239$    & $31473.850$    \\
 Numb. obs.                         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         & $8643$         \\
 Num. groups: individuals           & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         & $1687$         \\
-Var: individuals (Intercept)       & $0.172$        & $0.223$        & $0.219$        & $0.209$        & $0.207$        & $0.182$        & $0.171$        \\
-Var: Residual                      & $0.530$        & $0.499$        & $0.498$        & $0.495$        & $0.495$        & $0.495$        & $0.495$        \\
-Var: individuals, wave             & $$             & $0.011$        & $0.010$        & $0.009$        & $0.009$        & $0.009$        & $0.009$        \\
-Cov: individuals (Intercept), wave & $$             & $-0.025$       & $-0.024$       & $-0.021$       & $-0.020$       & $-0.019$       & $-0.019$       \\
+Var: individuals (Intercept)       & $0.205$        & $0.370$        & $0.366$        & $0.363$        & $0.364$        & $0.336$        & $0.326$        \\
+Var: Residual                      & $0.416$        & $0.345$        & $0.345$        & $0.343$        & $0.343$        & $0.343$        & $0.343$        \\
+Var: individuals, wave             & $$             & $0.022$        & $0.021$        & $0.021$        & $0.021$        & $0.021$        & $0.021$        \\
+Cov: individuals (Intercept), wave & $$             & $-0.061$       & $-0.060$       & $-0.059$       & $-0.059$       & $-0.058$       & $-0.059$       \\
 \hline
@@
 intertemporal variations in the dependent variable, there is a decrease
-in 2017 (\(\beta\) = -0.167, \(p\) \textless{} .001) relative to 2016,
-and similarly in 2018 (\(\beta\) = -0.025, \(p\) \textgreater{} .05) and
-2019 (\(\beta\) = -0.046, \(p\) \textgreater{} .05), although the latter
+in 2017 (\(\beta\) = -0.183, \(p\) \textless{} .001) relative to 2016,
+and similarly in 2018 (\(\beta\) = -0.009, \(p\) \textgreater{} .05) and
+2019 (\(\beta\) = -0.009, \(p\) \textgreater{} .05), although the latter
 effects are not statistically significant. In contrast, in the more
 recent waves of 2022 and 2023, there is a statistically significant
-increase in market justice preferences (\(\beta\) = 0.285, \(p\)
-\textless{} .001; \(\beta\) = 0.284, \(p\) \textless{} .001), suggesting
+increase in market justice preferences (\(\beta\) = 0.300, \(p\)
+\textless{} .001; \(\beta\) = 0.320, \(p\) \textless{} .001), suggesting
 a non-linear effect. To model this trajectory over time, Model 2
@@
 Specifically, each one-point increase in an individual's perception of
-economic inequality between waves is associated with a 0.034 point
+economic inequality between waves is associated with a 0.027 point
 decrease in market justice preferences. Model 4 shows that meritocratic
-beliefs operate in the opposite direction. An upward shift in the
+perceptions operate in distinct directions. An upward shift in the
 perception that effort is rewarded exerts a positive within effect
-(\(\beta\) = 0.071, \(p\) \textless{} .001), and a parallel increase in
-the perception that intelligence and ability are rewarded is likewise
-associated with higher market-justice preferences (\(\beta\) = 0.028,
+(\(\beta\) = 0.070, \(p\) \textless{} .001), while a parallel increase
+in the perception that intelligence and ability are rewarded is likewise
+associated with lower market-justice preferences (\(\beta\) = -0.027,
 \(p\) \textless{} .05). Taken together, these results suggest that
-individuals who increasingly perceive meritocracy---whether through
-effort or talent---tend to hold stronger market justice preferences.
+people who increasingly perceive meritocracy based on effort tend to
+have stronger preferences for market justice, and that the opposite is
+true for those who increasingly perceive meritocracy based on talent.
 
@@
 levels of economic inequality tend to prefer less market justice
-(\(\beta\) = -0.042, \(p\) \textless{} .05). In Model 6, the
-meritocratic perception that effort is rewarded is positively associated
-with market justice preferences (\(\beta\) = 0.186, \(p\) \textless{}
-.001), whereas the perception that talent is rewarded shows a positive
-but non-significant coefficient (\(\beta\) = 0.040). Notably, once the
-meritocratic variables are included, the negative coefficient for
-perceived economic inequality remains but is no longer statistically
-significant.
-
-Model 7 adds the control variables. With the exception of the
-between-effect for perceived economic inequality---which becomes
-nonsignificant---the within- and between-effects of the principal
-predictors retain both their direction and statistical significance,
-confirming the robustness of the associations (see Supplementary
-Material for effects of control variables).
+(\(\beta\) = -0.002, \(p\) \textgreater{} .05). However, this effects is
+no longer statistically significant. In Model 6, the meritocratic
+perception that effort is rewarded is positively associated with market
+justice preferences (\(\beta\) = 0.206, \(p\) \textless{} .001), whereas
+the perception that talent is rewarded shows a positive but
+non-significant coefficient (\(\beta\) = 0.036).
+
+Model 7 adds the control variables. The within- and between-effects of
+the principal predictors retain both their direction and statistical
+significance, confirming the robustness of the associations (see
+Supplementary Material for effects of control variables).
 
@@
 \hline
-Intercept                                                     & $1.561^{***}$  & $1.555^{***}$  & $2.187^{***}$  & $2.340^{***}$  \\
-                                                              & $(0.115)$      & $(0.115)$      & $(0.265)$      & $(0.281)$      \\
-Perception inequality (WE)                                    & $-0.035^{***}$ & $-0.035^{***}$ & $-0.035^{***}$ & $-0.035^{***}$ \\
-                                                              & $(0.010)$      & $(0.010)$      & $(0.010)$      & $(0.010)$      \\
-Merit: Effort (WE)                                            & $0.072^{***}$  & $0.074^{***}$  & $0.073^{***}$  & $0.073^{***}$  \\
-                                                              & $(0.013)$      & $(0.012)$      & $(0.012)$      & $(0.012)$      \\
-Merit: Talent (WE)                                            & $0.031^{*}$    & $0.031^{*}$    & $0.030^{*}$    & $0.030^{*}$    \\
-                                                              & $(0.012)$      & $(0.013)$      & $(0.012)$      & $(0.012)$      \\
-Perception inequality (BE)                                    & $-0.032$       & $-0.033$       & $-0.213^{**}$  & $-0.255^{***}$ \\
-                                                              & $(0.018)$      & $(0.018)$      & $(0.073)$      & $(0.077)$      \\
-Merit: Effort (BE)                                            & $0.177^{***}$  & $0.173^{***}$  & $-0.062$       & $0.179^{***}$  \\
-                                                              & $(0.032)$      & $(0.031)$      & $(0.097)$      & $(0.032)$      \\
-Merit: Talent (BE)                                            & $0.015$        & $0.021$        & $0.014$        & $-0.266^{**}$  \\
-                                                              & $(0.031)$      & $(0.031)$      & $(0.031)$      & $(0.099)$      \\
-Merit: Effort (WE) x Perception inequality (WE)               & $0.017$        &                &                &                \\
-                                                              & $(0.012)$      &                &                &                \\
-Merit: Talent (WE) x Perception inequality (WE)               &                & $0.011$        &                &                \\
-                                                              &                & $(0.012)$      &                &                \\
-Merit: Effort (BE) x Perception inequality (BE)               &                &                & $0.072^{**}$   &                \\
-                                                              &                &                & $(0.028)$      &                \\
-Merit: Talent (BE) x Perception inequality (BE)               &                &                &                & $0.082^{**}$   \\
-                                                              &                &                &                & $(0.027)$      \\
+Intercept                                                     & $1.328^{***}$ & $1.346^{***}$ & $1.956^{***}$  & $2.281^{***}$  \\
+                                                              & $(0.143)$     & $(0.143)$     & $(0.350)$      & $(0.369)$      \\
+Perception inequality (WE)                                    & $-0.037^{*}$  & $-0.036^{*}$  & $-0.040^{***}$ & $-0.040^{***}$ \\
+                                                              & $(0.015)$     & $(0.015)$     & $(0.009)$      & $(0.009)$      \\
+Merit: Effort (WE)                                            & $0.075^{***}$ & $0.085^{***}$ & $0.081^{***}$  & $0.081^{***}$  \\
+                                                              & $(0.018)$     & $(0.012)$     & $(0.011)$      & $(0.011)$      \\
+Merit: Talent (WE)                                            & $-0.021$      & $-0.017$      & $-0.026^{*}$   & $-0.026^{*}$   \\
+                                                              & $(0.011)$     & $(0.016)$     & $(0.011)$      & $(0.011)$      \\
+Perception inequality (BE)                                    & $-0.003$      & $-0.003$      & $-0.176$       & $-0.268^{**}$  \\
+                                                              & $(0.023)$     & $(0.023)$     & $(0.096)$      & $(0.101)$      \\
+Merit: Effort (BE)                                            & $0.180^{***}$ & $0.183^{***}$ & $-0.045$       & $0.191^{***}$  \\
+                                                              & $(0.040)$     & $(0.040)$     & $(0.127)$      & $(0.041)$      \\
+Merit: Talent (BE)                                            & $0.022$       & $0.018$       & $0.017$        & $-0.320^{*}$   \\
+                                                              & $(0.039)$     & $(0.039)$     & $(0.041)$      & $(0.129)$      \\
+Merit: Effort (WE) x Perception inequality (WE)               & $-0.014$      &               &                &                \\
+                                                              & $(0.013)$     &               &                &                \\
+Merit: Talent (WE) x Perception inequality (WE)               &               & $-0.032^{*}$  &                &                \\
+                                                              &               & $(0.013)$     &                &                \\
+Merit: Effort (BE) x Perception inequality (BE)               &               &               & $0.070$        &                \\
+                                                              &               &               & $(0.036)$      &                \\
+Merit: Talent (BE) x Perception inequality (BE)               &               &               &                & $0.099^{**}$   \\
+                                                              &               &               &                & $(0.036)$      \\
 \hline
 Controls                                                      & Yes           & Yes           & Yes            & Yes            \\
-BIC                                                           & $20842.694$    & $20835.920$    & $20808.677$    & $20806.384$    \\
+BIC                                                           & $31201.714$   & $31254.484$   & $32281.698$    & $32277.809$    \\
 Numb. obs.                                                    & $8643$        & $8643$        & $8643$         & $8643$         \\
 Num. groups: individuals                                      & $1687$        & $1687$        & $1687$         & $1687$         \\
-Var: individuals (Intercept)                                  & $0.145$        & $0.145$        & $0.141$        & $0.141$        \\
-Var: individuals, merit effort cwc                            & $0.020$        & $$             & $$             & $$             \\
-Var: individuals, perception inequality cwc                   & $0.000$        & $0.000$        & $$             & $$             \\
-Cov: individuals (Intercept), merit effort cwc                & $0.004$        & $$             & $$             & $$             \\
-Cov: individuals (Intercept), perception inequality cwc       & $-0.006$       & $-0.005$       & $$             & $$             \\
-Cov: individuals, merit effort cwc, perception inequality cwc & $0.000$        & $$             & $$             & $$             \\
-Var: Residuals                                                & $0.512$        & $0.511$        & $0.528$        & $0.528$        \\
-Var: individuals, merit talent cwc                            & $$             & $0.020$        & $$             & $$             \\
-Cov: individuals (Intercept), merit talent cwc                & $$             & $0.012$        & $$             & $$             \\
-Cov: individuals, merit talent cwc, perception inequality cwc & $$             & $0.001$        & $$             & $$             \\
+Var: individuals (Intercept)                                  & $0.178$       & $0.178$       & $0.170$        & $0.169$        \\
+Var: individuals, merit effort cwc                            & $0.127$       & $$            & $$             & $$             \\
+Var: individuals, perception inequality cwc                   & $0.090$       & $0.088$       & $$             & $$             \\
+Cov: individuals (Intercept), merit effort cwc                & $-0.009$      & $$            & $$             & $$             \\
+Cov: individuals (Intercept), perception inequality cwc       & $-0.015$      & $-0.013$      & $$             & $$             \\
+Cov: individuals, merit effort cwc, perception inequality cwc & $0.003$       & $$            & $$             & $$             \\
+Var: Residuals                                                & $0.296$       & $0.303$       & $0.417$        & $0.417$        \\
+Var: individuals, merit talent cwc                            & $$            & $0.097$       & $$             & $$             \\
+Cov: individuals (Intercept), merit talent cwc                & $$            & $-0.011$      & $$             & $$             \\
+Cov: individuals, merit talent cwc, perception inequality cwc & $$            & $-0.004$      & $$             & $$             \\
 \hline
@@
 
-\caption{\label{fig-interact}Between effects of meritocratic perceptions
-on market justice preferences by economic inequality perception}
+\caption{\label{fig-interact}Predicted values of market justice
+preferences by perceptions of meritocraticy and economic inequality}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-interact-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{blinded-paper-resubmit_files/figure-pdf/fig-interact-1.pdf}
 
@@
 Table~\ref{tbl-interactions1} examines whether perceived economic
-inequality moderates the effect of meritocratic beliefs on market
-justice preferences. Contrary to our hypothesis, the interaction terms
-in the within-person specification (Models 8 and 9) do not reach
-statistical significance. However, in the between-person specification
-(Models 10 and 11), the interaction terms are statistically significant,
-suggesting that perceptions of economic inequality meaningfully alter
-the effect of meritocracy on support for market-based allocation of
-social services. These positive effects indicate that, as perceptions of
-inequality increase, the positive association between meritocratic
-perceptions and market justice preferences becomes stronger. The
-marginal effects of this interaction are illustrated in
-Figure~\ref{fig-interact}, which shows that the effect of meritocratic
-perceptions on market justice preferences intensifies as perceived
-economic inequality rises.
+inequality moderates the effect of meritocratic perceptions on market
+justice preferences. Contrary to our expectations, the interaction terms
+in the within-person specification of Model 9 indicate that the negative
+effect of talent-based meritocratic perceptions becomes stronger as
+perceived inequality increases (\(\beta\) = -0.032, \(p\) \textless{}
+.05). In contrast, in the between-person specification (Model 11), the
+interaction term is positive (\(\beta\) = 0.099, \(p\) \textless{} .01),
+suggesting that perceptions of economic inequality significantly shape
+the influence of meritocratic perceptions on support for market-based
+allocation of social services. Specifically, these positive effects
+indicate that, as perceptions of inequality increase, the association
+between talent-based meritocratic perceptions and market justice
+preferences shifts in a positive direction.
+
+The effects of effort-based meritocratic perceptions are not
+statistically significant in either the within- or between-person
+specifications (Models 8 and 10). However, the direction of the
+coefficients is negative in the former and positive in the latter. As
+shown in Figure~\ref{fig-interact}, Panel A, the between-person effect
+of effort-based meritocratic perceptions on market justice preferences
+strengthens as perceived inequality increases across individuals,
+whereas this effect weakens among those who perceive lower levels of
+inequality. Panel B reveals a similar pattern for the between-person
+effect of talent-based meritocratic perceptions, although the effect is
+substantially weaker at low levels of perceived inequality.
+
+Substantively, these findings suggest that the legitimizing function of
+meritocratic perceptions is amplified in contexts perceived as highly
+unequal. In such contexts, meritocratic narratives may reinforce
+individualistic understandings of distributive justice and intensify
+support for market-based access to essential social services.
 
@@
 \hline
-Intercept                                                & $1.543^{***}$ & $1.527^{***}$ & $1.521^{***}$ & $1.680^{***}$ & $1.374^{***}$ & $1.362^{***}$ \\
-                                                         & $(0.115)$     & $(0.114)$     & $(0.114)$     & $(0.139)$     & $(0.133)$     & $(0.134)$     \\
-Wave x Perception inequality (WE)                        & $-0.004$      &               &               &               &               &               \\
+Intercept                                                & $1.263^{***}$ & $1.265^{***}$ & $1.246^{***}$ & $1.128^{***}$ & $1.099^{***}$ & $1.090^{***}$ \\
+                                                         & $(0.143)$     & $(0.141)$     & $(0.142)$     & $(0.176)$     & $(0.169)$     & $(0.170)$     \\
+Wave x Perception inequality (WE)                        & $0.003$       &               &               &               &               &               \\
                                                          & $(0.006)$     &               &               &               &               &               \\
-Wave x Merit: effort (WE)                                &               & $-0.008$      &               &               &               &               \\
+Wave x Merit: effort (WE)                                &               & $0.012$       &               &               &               &               \\
                                                          &               & $(0.006)$     &               &               &               &               \\
-Wave x Merit: talent (WE)                                &               &               & $-0.006$      &               &               &               \\
+Wave x Merit: talent (WE)                                &               &               & $0.007$       &               &               &               \\
                                                          &               &               & $(0.006)$     &               &               &               \\
-Wave x Perception inequality (BE)                        &               &               &               & $0.011$       &               &               \\
-                                                         &               &               &               & $(0.007)$     &               &               \\
-Wave x Merit: effort (BE)                                &               &               &               &               & $-0.021^{*}$  &               \\
-                                                         &               &               &               &               & $(0.008)$     &               \\
-Wave x Merit: talent (BE)                                &               &               &               &               &               & $-0.020^{*}$  \\
-                                                         &               &               &               &               &               & $(0.008)$     \\
+Wave x Perception inequality (BE)                        &               &               &               & $-0.011$      &               &               \\
+                                                         &               &               &               & $(0.009)$     &               &               \\
+Wave x Merit: effort (BE)                                &               &               &               &               & $-0.020$      &               \\
+                                                         &               &               &               &               & $(0.011)$     &               \\
+Wave x Merit: talent (BE)                                &               &               &               &               &               & $-0.016$      \\
+                                                         &               &               &               &               &               & $(0.010)$     \\
 \hline
 Controls                                                 & Yes           & Yes           & Yes           & Yes           & Yes           & Yes           \\
-BIC                                                      & $20792.207$   & $20767.371$   & $20754.324$   & $20782.059$   & $20749.596$   & $20754.417$   \\
+BIC                                                      & $31010.853$   & $30888.601$   & $30785.269$   & $31515.978$   & $31493.536$   & $31501.405$   \\
 Numb. obs.                                               & $8643$        & $8643$        & $8643$        & $8643$        & $8643$        & $8643$        \\
 Num. groups: individuals                                 & $1687$        & $1687$        & $1687$        & $1687$        & $1687$        & $1687$        \\
-Var: individuals (Intercept)                             & $0.176$       & $0.167$       & $0.168$       & $0.383$       & $0.254$       & $0.010$       \\
-Var: individuals, perception inequality cwc              & $0.000$       & $$            & $$            & $$            & $$            & $$            \\
-Var: individuals, wave                                   & $0.010$       & $0.009$       & $0.009$       & $0.009$       & $0.009$       & $0.009$       \\
-Cov: individuals (Intercept), perception inequality cwc  & $-0.007$      & $$            & $$            & $$            & $$            & $$            \\
-Cov: individuals (Intercept), wave                       & $-0.021$      & $-0.018$      & $-0.018$      & $-0.048$      & $-0.003$      & $0.006$       \\
-Cov: individuals, perception inequality cwc, wave        & $0.002$       & $$            & $$            & $$            & $$            & $$            \\
-Var: Residuals                                           & $0.495$       & $0.480$       & $0.479$       & $0.494$       & $0.494$       & $0.494$       \\
-Var: individuals, merit effort cwc                       & $$            & $0.019$       & $$            & $$            & $$            & $$            \\
-Cov: individuals (Intercept), merit effort cwc           & $$            & $0.023$       & $$            & $$            & $$            & $$            \\
-Cov: individuals, merit effort cwc, wave                 & $$            & $-0.005$      & $$            & $$            & $$            & $$            \\
-Var: individuals, merit talent cwc                       & $$            & $$            & $0.019$       & $$            & $$            & $$            \\
-Cov: individuals (Intercept), merit talent cwc           & $$            & $$            & $0.034$       & $$            & $$            & $$            \\
-Cov: individuals, merit talent cwc, wave                 & $$            & $$            & $-0.006$      & $$            & $$            & $$            \\
-Var: individuals, perception inequality mean             & $$            & $$            & $$            & $0.008$       & $$            & $$            \\
-Cov: individuals (Intercept), perception inequality mean & $$            & $$            & $$            & $-0.045$      & $$            & $$            \\
-Cov: individuals, perception inequality mean, wave       & $$            & $$            & $$            & $0.009$       & $$            & $$            \\
-Var: individuals, merit effort mean                      & $$            & $$            & $$            & $$            & $0.053$       & $$            \\
-Cov: individuals (Intercept), merit effort mean          & $$            & $$            & $$            & $$            & $-0.089$      & $$            \\
-Cov: individuals, merit effort mean, wave                & $$            & $$            & $$            & $$            & $-0.006$      & $$            \\
-Var: individuals, merit talent mean                      & $$            & $$            & $$            & $$            & $$            & $0.025$       \\
-Cov: individuals (Intercept), merit talent mean          & $$            & $$            & $$            & $$            & $$            & $-0.007$      \\
-Cov: individuals, merit talent mean, wave                & $$            & $$            & $$            & $$            & $$            & $-0.009$      \\
+Var: individuals (Intercept)                             & $0.388$       & $0.334$       & $0.379$       & $0.274$       & $0.532$       & $0.372$       \\
+Var: individuals, perception inequality cwc              & $0.100$       & $$            & $$            & $$            & $$            & $$            \\
+Var: individuals, wave                                   & $0.025$       & $0.023$       & $0.024$       & $0.021$       & $0.021$       & $0.021$       \\
+Cov: individuals (Intercept), perception inequality cwc  & $-0.081$      & $$            & $$            & $$            & $$            & $$            \\
+Cov: individuals (Intercept), wave                       & $-0.074$      & $-0.062$      & $-0.072$      & $-0.054$      & $-0.056$      & $-0.049$      \\
+Cov: individuals, perception inequality cwc, wave        & $0.021$       & $$            & $$            & $$            & $$            & $$            \\
+Var: Residuals                                           & $0.291$       & $0.280$       & $0.275$       & $0.343$       & $0.342$       & $0.343$       \\
+Var: individuals, merit effort cwc                       & $$            & $0.122$       & $$            & $$            & $$            & $$            \\
+Cov: individuals (Intercept), merit effort cwc           & $$            & $-0.009$      & $$            & $$            & $$            & $$            \\
+Cov: individuals, merit effort cwc, wave                 & $$            & $0.000$       & $$            & $$            & $$            & $$            \\
+Var: individuals, merit talent cwc                       & $$            & $$            & $0.112$       & $$            & $$            & $$            \\
+Cov: individuals (Intercept), merit talent cwc           & $$            & $$            & $0.003$       & $$            & $$            & $$            \\
+Cov: individuals, merit talent cwc, wave                 & $$            & $$            & $-0.003$      & $$            & $$            & $$            \\
+Var: individuals, perception inequality mean             & $$            & $$            & $$            & $0.000$       & $$            & $$            \\
+Cov: individuals (Intercept), perception inequality mean & $$            & $$            & $$            & $0.007$       & $$            & $$            \\
+Cov: individuals, perception inequality mean, wave       & $$            & $$            & $$            & $-0.001$      & $$            & $$            \\
+Var: individuals, merit effort mean                      & $$            & $$            & $$            & $$            & $0.051$       & $$            \\
+Cov: individuals (Intercept), merit effort mean          & $$            & $$            & $$            & $$            & $-0.110$      & $$            \\
+Cov: individuals, merit effort mean, wave                & $$            & $$            & $$            & $$            & $-0.001$      & $$            \\
+Var: individuals, merit talent mean                      & $$            & $$            & $$            & $$            & $$            & $0.031$       \\
+Cov: individuals (Intercept), merit talent mean          & $$            & $$            & $$            & $$            & $$            & $-0.053$      \\
+Cov: individuals, merit talent mean, wave                & $$            & $$            & $$            & $$            & $$            & $-0.003$      \\
 \hline
@@
 With respect to the temporal dynamics of the key predictors,
-Table~\ref{tbl-interactions2} shows that time only has significant
-effects on the between effects of the meritocratic variables. These
-interaction effects are negative and mean that in the last waves, the
-impact of meritocratic variables on market justice preferences is
-attenuated. Specifically, as time advances, the positive within-person
-effect of perceiving that effort is rewarded (Model 16) or that
-intelligence is rewarded (Model 17) progressively attenuates. Put
-differently, the uplift in market justice preferences associated with a
-strong meritocratic perception becomes smaller in each successive wave.
+Table~\ref{tbl-interactions2} shows that time has no statistically
+significant effects on either the within- or between-person effects of
+perceived inequality and meritocratic perceptions. The within-person
+interaction effects are positive, suggesting that in later waves, the
+influence of perceived inequality and meritocratic beliefs on market
+justice preferences tends to increase. Conversely, in the between-person
+specification, the interaction effects are negative, indicating that
+over time, the impact of perceived inequality and meritocratic beliefs
+on support for market-based justice tends to diminish. However, these
+trends are not statistically significant, and thus no definitive
+conclusions can be drawn.
 
@@
 This study tests several key hypotheses concerning the relationships
-between perceptions of economic inequality, meritocratic beliefs, and
-attitudes toward market justice in Chile. It posits that higher
-perceived inequality and stronger meritocratic beliefs both
+between perceptions of economic inequality, meritocratic perceptions,
+and attitudes toward market justice in Chile. It posits that higher
+perceived inequality and stronger meritocratic perceptions both
 independently contribute to greater support for market-based
@@
 perceived economic inequality moderates the influence of meritocratic
-beliefs on market justice preferences. Additionally, it considers how
-major social and political mobilizations, such as those occurring around
-2019 and subsequent years, may have weakened these relationships,
+perceptions on market justice preferences. Additionally, it considers
+how major social and political mobilizations, such as those occurring
+around 2019 and subsequent years, may have weakened these relationships,
 reflecting increased critical perspectives on systemic inequalities and
@@
 2012}; \citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These
-beliefs functioned symbolically to legitimize structural disparities,
-reducing support for redistributive measures by framing inequality as
-fair and deserved. The findings suggested that these perceptions
-bolstered the social hierarchy by fostering acceptance of inequality as
-a reflection of individual virtue rather than structural failure, which
-was especially pertinent in neoliberal contexts like Chile's, where
-market logic heavily influenced social attitudes
+perceptions functioned symbolically to legitimize structural
+disparities, reducing support for redistributive measures by framing
+inequality as fair and deserved. The findings suggested that these
+perceptions bolstered the social hierarchy by fostering acceptance of
+inequality as a reflection of individual virtue rather than structural
+failure, which was especially pertinent in neoliberal contexts like
+Chile's, where market logic heavily influenced social attitudes
 (\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).
@@
 attitudes. The analysis indicated that the positive association between
-meritocratic beliefs and support for market-based distribution tended to
-weaken when perceptions of economic inequality were particularly high.
-Although this moderation is not significant for within-subject effects
-(i.e.~over time), when taking into account between-subjects differences
-we found a positive effect. This appears as contrary to our initial
-hypothesis, as it would mean that effect of meritocratic perceptions
-becomes stronger as perception of inequality increases. Nevertheless, we
-need to be cautious about this interpretation, as the main effect of
-inequality perception on market justice is negative. Therefore, as
-perceived inequality become less negative (i.e.~more positive), or
-closer to zero, then the meritocratic perception has an stronger
-positive relationship on market justice preferences. This could mean
-that meritocratic perceptions play an stronger role in justification of
-market justice inequalities for those who are less capable or less prone
-to perceive inequality.
+meritocratic perceptions and support for market-based distribution
+tended to weaken when perceptions of economic inequality were
+particularly high. Although this moderation is not significant for
+within-subject effects (i.e.~over time), when taking into account
+between-subjects differences we found a positive effect. This appears as
+contrary to our initial hypothesis, as it would mean that effect of
+meritocratic perceptions becomes stronger as perception of inequality
+increases. Nevertheless, we need to be cautious about this
+interpretation, as the main effect of inequality perception on market
+justice is negative. Therefore, as perceived inequality become less
+negative (i.e.~more positive), or closer to zero, then the meritocratic
+perception has an stronger positive relationship on market justice
+preferences. This could mean that meritocratic perceptions play an
+stronger role in justification of market justice inequalities for those
+who are less capable or less prone to perceive inequality.
 
@@
 movements, particularly from 2019 onwards, on the relationship between
-meritocratic beliefs, perceptions of inequality, and attitudes toward
-market justice. The results suggested that these movements contributed
-to a more critical stance toward the fairness of existing distributional
-mechanisms. Post-mobilization, the positive relationship between
-meritocratic beliefs and support for inequality appeared attenuated,
-indicating that collective action and public discourse during this
-period fostered questioning of the legitimacy of systemic inequalities
-justified solely through meritocratic narratives. These social
-mobilizations challenged dominant narratives, emphasizing structural
-barriers and systemic injustices, which in turn weakened the association
-between individual effort and deservingness. This is aligned with
-literature emphasizing that social protests act as catalysts for
+meritocratic perceptions, perceptions of inequality, and attitudes
+toward market justice. The results suggested that these movements
+contributed to a more critical stance toward the fairness of existing
+distributional mechanisms. Post-mobilization, the positive relationship
+between meritocratic perceptions and support for inequality appeared
+attenuated, indicating that collective action and public discourse
+during this period fostered questioning of the legitimacy of systemic
+inequalities justified solely through meritocratic narratives. These
+social mobilizations challenged dominant narratives, emphasizing
+structural barriers and systemic injustices, which in turn weakened the
+association between individual effort and deservingness. This is aligned
+with literature emphasizing that social protests act as catalysts for
 transforming societal attitudes by disrupting normative beliefs and
@@
 This study examined the complex interplay between perceptions of
-economic inequality, meritocratic beliefs, and attitudes toward market
-justice in Chile from 2016 to 2023, drawing on longitudinal data from
-the ELSOC survey. By exploring how subjective assessments and social
-contexts influence support for redistribution and market-based resource
-allocation, the research offers different elements that contribute to
-the understanding of the normative foundations underpinning social
-justice attitudes in a highly unequal, neoliberal environment.
+economic inequality, meritocratic perceptions, and attitudes toward
+market justice in Chile from 2016 to 2023, drawing on longitudinal data
+from the ELSOC survey. By exploring how subjective assessments and
+social contexts influence support for redistribution and market-based
+resource allocation, the research offers different elements that
+contribute to the understanding of the normative foundations
+underpinning social justice attitudes in a highly unequal, neoliberal
+environment.
 
@@
 belief that it is fine that those with higher income have better social
-services such as education, pensions and health. Meritocratic beliefs
-strongly reinforce support for market justice, serving as moral
-justifications for structural inequalities. Additionally, perceptions of
-inequality moderate the influence of meritocratic beliefs, such that as
-the awareness of disparities are attenuated (i.e.~less negative) the
-role of meritocratic beliefs on existing market-based allocations become
-stronger. Notably, the social mobilizations of 2019 and subsequent
-periods seem to have contributed to shifting attitudes, fostering
-greater skepticism of systemic fairness and reducing the legitimizing
-power of meritocratic narratives.
+services such as education, pensions and health. Meritocratic
+perceptions strongly reinforce support for market justice, serving as
+moral justifications for structural inequalities. Additionally,
+perceptions of inequality moderate the influence of meritocratic
+perceptions, such that as the awareness of disparities are attenuated
+(i.e.~less negative) the role of meritocratic perceptions on existing
+market-based allocations become stronger. Notably, the social
+mobilizations of 2019 and subsequent periods seem to have contributed to
+shifting attitudes, fostering greater skepticism of systemic fairness
+and reducing the legitimizing power of meritocratic narratives.
 
@@
 approach provides a deeper temporal perspective on how societal events
-could influence individual beliefs and preferences.
+could influence individual perceptions and preferences.
 
@@
 \bibitem[\citeproctext]{ref-boccardo_30_2020}
-Boccardo, G. (2020). \emph{30 a{ñ}os de privatizaciones en {Chile}: Lo
+Boccardo, G. (2020). \emph{30 a{ñ}os de privatizaciones en {Chile}: {Lo}
 que la pandemia revel{ó}} (Nodo XXI). Santiago.
 
+\bibitem[\citeproctext]{ref-busemeyer_skills_2014}
+Busemeyer, M. (2014). \emph{Skills and {Inequality}: {Partisan Politics}
+and the {Political Economy} of {Education Reforms} in {Western Welfare
+States}}. Cambridge University Press.
+
 \bibitem[\citeproctext]{ref-busemeyer_positive_2021}
-Busemeyer, M. R., Abrassart, A., \& Nezi, R. (2021). Beyond {Positive}
-and {Negative}: {New Perspectives} on {Feedback Effects} in {Public
-Opinion} on the {Welfare State}. \emph{British Journal of Political
-Science}, \emph{51}(1), 137--162.
-\url{https://doi.org/10.1017/S0007123418000534}
+Busemeyer, M., Abrassart, A., \& Nezi, R. (2021). Beyond {Positive} and
+{Negative}: {New Perspectives} on {Feedback Effects} in {Public Opinion}
+on the {Welfare State}. \emph{British Journal of Political Science},
+\emph{51}(1), 137--162. \url{https://doi.org/10.1017/S0007123418000534}
 
@@
 
+\bibitem[\citeproctext]{ref-castillo_changes_2025}
+Castillo, J. C., Iturra, J., \& Carrasco, K. (2025). Changes in the
+{Justification} of {Educational Inequalities}: {The Role} of
+{Perceptions} of {Inequality} and {Meritocracy During} the {COVID
+Pandemic}. \emph{Social Justice Research}.
+\url{https://doi.org/10.1007/s11211-025-00458-0}
+
 \bibitem[\citeproctext]{ref-castillo_multidimensional_2023}
@@
 
+\bibitem[\citeproctext]{ref-disipavlic_justification_2025}
+Disi Pavlic, R., Medel, R. M., Bargsted, M., \& Somma, N. M. (2025).
+Justification of violence, ideological preferences, and exposure to
+protests: Causal evidence from the 2019 {Chilean} social unrest.
+\emph{Social Forces}, soaf102. \url{https://doi.org/10.1093/sf/soaf102}
+
 \bibitem[\citeproctext]{ref-easterbrook_social_2021}
@@
 \url{https://doi.org/10.35188/UNU-WIDER/2021/981-5}
-
-\bibitem[\citeproctext]{ref-elsoc_estudio_2022}
-ELSOC, S. T. (2022). Estudio {Longitudinal Social} de {Chile}. Harvard
-Dataverse. \url{https://doi.org/10.7910/dvn/0kirbj}
 
@@
 \url{https://doi.org/10.1093/ser/mwx036}
+
+\bibitem[\citeproctext]{ref-espinoza_movilidad_2014}
+Espinoza, V., \& Núñez, J. (2014). Movilidad ocupacional en {Chile}
+2001-2009. {\textquestiondown}{Desigualdad} de ingresos con igualdad de
+oportunidades? \emph{Revista Internacional de Sociolog{í}a},
+\emph{72}(1), 57--82. \url{https://doi.org/10.3989/ris.2011.11.08}
+
+\bibitem[\citeproctext]{ref-ferre_welfare_2023}
+Ferre, J. C. (2023). Welfare regimes in twenty-first-century {Latin
+America}. \emph{Journal of International and Comparative Social Policy},
+\emph{39}(2), 101--127. \url{https://doi.org/10.1017/ics.2023.16}
 
@@
 
+\bibitem[\citeproctext]{ref-kluegel_beliefs_1986}
+Kluegel, J. R., \& Smith, E. R. (1986). \emph{Beliefs about
+{Inequality}: {Americans}' {Views} of {What Is} and {What Ought} to
+{Be}} (1st ed.). Routledge.
+
 \bibitem[\citeproctext]{ref-knell_perceptions_2020}
@@
 
+\bibitem[\citeproctext]{ref-lane_political_1962}
+Lane, R. E. (1962). \emph{Political ideology: Why the {American} common
+man believes what he does}. Oxford, England: Free Press of Glencoe.
+
 \bibitem[\citeproctext]{ref-lane_market_1986}
-Lane, R. E. (1986). Market {Justice}, {Political Justice}.
+Lane, Robert E. (1986). Market {Justice}, {Political Justice}.
 \emph{American Political Science Review}, \emph{80}(2), 383--402.
@@
 
+\bibitem[\citeproctext]{ref-lindh_bringing_2023}
+Lindh, A., \& McCall, L. (2023). Bringing the market in: An expanded
+framework for understanding popular responses to economic inequality.
+\emph{Socio-Economic Review}, \emph{21}(2), 1035--1055.
+\url{https://doi.org/10.1093/ser/mwac018}
+
+\bibitem[\citeproctext]{ref-lopez-roldan_comparative_2021}
+López-Roldán, P., \& Fachelli, S. (Eds.). (2021). \emph{Towards a
+{Comparative Analysis} of {Social Inequalities} between {Europe} and
+{Latin America}}. Cham: Springer International Publishing.
+\url{https://doi.org/10.1007/978-3-030-48442-2}
+
+\bibitem[\citeproctext]{ref-mac-clure_justicia_2024}
+Mac-Clure, O., Barozet, E., \& Franetovic, G. (2024). {Justicia
+distributiva y posici{ó}n social subjetiva: {\textquestiondown}la
+meritocracia justifica la desigualdad de ingresos?} \emph{Convergencia
+Revista de Ciencias Sociales}, \emph{31}, 1.
+\url{https://doi.org/10.29101/crcs.v31i0.22258}
+
+\bibitem[\citeproctext]{ref-madariaga_three_2020}
+Madariaga, A. (2020). The three pillars of neoliberalism: {Chile}'s
+economic policy trajectory in comparative perspective.
+\emph{Contemporary Politics}, \emph{26}(3), 308--329.
+\url{https://doi.org/10.1080/13569775.2020.1735021}
+
+\bibitem[\citeproctext]{ref-mccall_exposure_2017}
+McCall, L., Burk, D., Laperrière, M., \& Richeson, J. A. (2017).
+Exposure to rising inequality shapes {Americans}' opportunity beliefs
+and policy support. \emph{Proceedings of the National Academy of
+Sciences}, \emph{114}(36), 9593--9598.
+\url{https://doi.org/10.1073/pnas.1706253114}
+
 \bibitem[\citeproctext]{ref-mijs_stratified_2016}
@@
 University Press, Incorporated.
+
+\bibitem[\citeproctext]{ref-somma_no_2021}
+Somma, N. M., Bargsted, M., Disi Pavlic, R., \& Medel, R. M. (2021). No
+water in the oasis: The {Chilean Spring} of 2019--2020. \emph{Social
+Movement Studies}, \emph{20}(4), 495--502.
+\url{https://doi.org/10.1080/14742837.2020.1727737}
 
@@
 
+\bibitem[\citeproctext]{ref-torche_intergenerational_2014a}
+Torche, F. (2014). Intergenerational {Mobility} and {Inequality}: {The
+Latin American Case}. \emph{Annual Review of Sociology}, \emph{40}(1),
+619--642. \url{https://doi.org/10.1146/annurev-soc-071811-145521}
+
 \bibitem[\citeproctext]{ref-trump_income_2018}

### 2025-07-28 resubmission/paper-resubmit.tex
Update varios ajustes

--- 2025-07-28 resubmission/blinded-paper-resubmit.tex
@@
 Shaping Preferences for Market Justice in Chile (2016-2023)}
+
+  \author{Juan Carlos Castillo}
+            \affil{%
+                  Departamento de Sociología, Universidad de Chile
+              }
+          \affil{%
+                  Centro de estudios del conflicto y cohesión social
+                  (COES)
+              }
+          \affil{%
+                  Núcleo milenio de desigualdades y oportunidades
+                  digitales (NUDOS)
+              }
+        \author{Andreas Laffert}
+            \affil{%
+                  Instituto de Sociología, Pontificia Universidad
+                  Católica de Chile
+              }
+        \author{Kevin Carrasco}
+            \affil{%
+                  Centro de estudios del conflicto y cohesión social
+                  (COES)
+              }
+        \author{Julio Iturra-Sanhueza}
+            \affil{%
+                  International Graduate School of Social Sciencies
+                  (BIGSSS), University of Bremen, Germany
+              }
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{blinded-paper-resubmit_files/figure-pdf/fig-alluvial-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{paper-resubmit_files/figure-pdf/fig-alluvial-1.pdf}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{blinded-paper-resubmit_files/figure-pdf/fig-meanchange-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{paper-resubmit_files/figure-pdf/fig-meanchange-1.pdf}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{blinded-paper-resubmit_files/figure-pdf/fig-interact-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{paper-resubmit_files/figure-pdf/fig-interact-1.pdf}
 

### 2025-07-28 resubmission/04-analisis.qmd
update discussion and conclusion (no render)

@@
 
-```{r echo=FALSE, results='asis'}
+```{r echo=FALSE, results='asis', eval=FALSE, include=FALSE}
 #| label: tbl-interactions2
@@
 
-With respect to the temporal dynamics of the key predictors, @tbl-interactions2 shows that time has no statistically significant effects on either the within- or between-person effects of perceived inequality and meritocratic perceptions.
-The within-person interaction effects are positive, suggesting that in later waves, the influence of perceived inequality and meritocratic beliefs on market justice preferences tends to increase.
-Conversely, in the between-person specification, the interaction effects are negative, indicating that over time, the impact of perceived inequality and meritocratic beliefs on support for market-based justice tends to diminish.
-However, these trends are not statistically significant, and thus no definitive conclusions can be drawn.
+Regarding the temporal dynamics of the key predictors, the analysis shows that time has no statistically significant effects on the intrapersonal or interpersonal effects of perceived inequality and meritocratic perceptions.
+Further details about this analysis can be found in supplementary material.

### 2025-07-28 resubmission/05-discussion.qmd
Update carpeta resubmission y varios (archivos no renderizados)
update discussion and conclusion (no render)

--- 2025-07-11 05-discussion.qmd
@@
 The first hypothesis suggested that perceptions of economic inequality were strongly linked to attitudes toward market justice.
-The findings demonstrated that individuals perceiving higher income disparities showed stronger support for the distribution of social services based on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021].
+The findings demonstrated that individuals perceiving higher income disparities showed less support for the distribution of social services based on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021].
 This relationship reflected a moral economy where perceptions of systemic unfairness threatened the legitimacy of existing distributions, prompting calls for greater equity.
-Interestingly, the negative within-effects observed over time implied that as perceptions of inequality became more salient, some individuals experienced a decline in support for market justice, possibly due to increasing distrust in market mechanisms or disillusionment with the fairness of the system.
-This indicated that while perception heightened awareness, prolonged exposure or social mobilization could also foster skepticism about whether redistribution effectively addressed inequality or simply perpetuated systemic flaws.
+Similarly, the negative within-effects observed over time ($H1_{b}$) implied that as perceptions of inequality became more salient, some individuals experienced a decline in support for market justice, possibly due to increasing distrust in market mechanisms or disillusionment with the fairness of the system.
+This indicated that while perception heightened awareness, prolonged exposure or social mobilization could also further foster skepticism about whether redistribution effectively addressed inequality or simply perpetuated systemic flaws.
 
-Regarding the second hypothesis, the results reinforced that meritocratic perceptions ---particularly those emphasizing individual effort and talent--- were associated with stronger support for market-based distribution systems.
+Regarding the second hypothesis, the results reinforced that meritocratic perceptions ---particularly those emphasizing individual effort--- were associated with stronger support for market-based distribution systems.
 Individuals who viewed success as primarily earned through effort were more inclined to justify existing inequalities, seeing them as outcomes of individual merit rather than systemic injustice.
@@
 The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes [@canalesceron_sujeto_2021].
+Interestingly, intraindividual changes in meritocratic perceptions over time ($H2_{b}$) reveal mixed effects.
+While an increase in the perception that rewards are based on individual effort is associated with stronger preferences for market fairness, an increase in the belief that rewards come from talent is linked to lower support for such principles.
+One possible explanation is that effort is generally perceived as a controllable and malleable trait, which aligns with the normative foundations of market fairness.
+Conversely, talent may be viewed as an innate and less alterable attribute, making talent-based reward allocation seem less fair or legitimate.
+Moreover, greater exposure to real-life situations where outcomes are strongly determined by inherent characteristics, rather than effort or hard work, may lead people to question the fairness of market-based reward systems.
 
 The third hypothesis concerned the moderating role of perceptions of inequality on the relationship between meritocracy and market justice attitudes.
-The analysis indicated that the positive association between meritocratic perceptions and support for market-based distribution tended to weaken when perceptions of economic inequality were particularly high.
-Although this moderation is not significant for within-subject effects (i.e.
-over time), when taking into account between-subjects differences we found a positive effect.
+The analysis indicated that the positive association between meritocratic perceptions and support for market-based distribution tended to strengthen when perceptions of economic inequality were particularly high.
+While this moderation is not significant for effort-related effects, it is significant for talent in the within- and between-individual effects.
 This appears as contrary to our initial hypothesis, as it would mean that effect of meritocratic perceptions becomes stronger as perception of inequality increases.
@@
 
-Finally, the fourth hypothesis addressed the impact of recent social movements, particularly from 2019 onwards, on the relationship between meritocratic perceptions, perceptions of inequality, and attitudes toward market justice.
-The results suggested that these movements contributed to a more critical stance toward the fairness of existing distributional mechanisms.
-Post-mobilization, the positive relationship between meritocratic perceptions and support for inequality appeared attenuated, indicating that collective action and public discourse during this period fostered questioning of the legitimacy of systemic inequalities justified solely through meritocratic narratives.
-These social mobilizations challenged dominant narratives, emphasizing structural barriers and systemic injustices, which in turn weakened the association between individual effort and deservingness.
-This is aligned with literature emphasizing that social protests act as catalysts for transforming societal attitudes by disrupting normative beliefs and encouraging more critical evaluations of institutional legitimacy [@tilly_contentious_2015].
-The findings underscored the importance of socio-political context in shaping and reshaping attitudes toward inequality and justice, highlighting how collective action could temporarily or permanently alter the normative foundations that sustained support for market justice.
+Finally, the fourth hypothesis addressed the impact of recent social movements—particularly those that emerged in 2019—on the relationship between meritocratic beliefs, perceptions of inequality, and attitudes toward market justice.
+The main effect of time suggests that, on average, support for market justice increased after 2019.
+However, the results do not reveal any significant interactions between distributive mechanisms and market justice attitudes, indicating that the 2019 social mobilizations did not substantially alter the underlying associations between these variables.
+This may suggest a certain stability or resilience in the normative frameworks that shape individuals’ evaluations of justice, even in the context of major collective political events.
+Alternatively, it is possible that while the protests challenged institutional legitimacy, they did not fundamentally shift individuals’ views regarding how rewards should be allocated in society.

### 2025-07-28 resubmission/06-conclusion.qmd
Update carpeta resubmission y varios (archivos no renderizados)
update discussion and conclusion (no render)

--- 2025-07-11 06-conclusion.qmd
@@
 The findings confirm that higher perceptions of economic inequality are associated with less support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health.
-Meritocratic perceptions strongly reinforce support for market justice, serving as moral justifications for structural inequalities.
-Additionally, perceptions of inequality moderate the influence of meritocratic perceptions, such that as the awareness of disparities are attenuated (i.e.
-less negative) the role of meritocratic perceptions on existing market-based allocations become stronger.
-Notably, the social mobilizations of 2019 and subsequent periods seem to have contributed to shifting attitudes, fostering greater skepticism of systemic fairness and reducing the legitimizing power of meritocratic narratives.
+At the same time, meritocratic beliefs—particularly those emphasizing individual effort—are strongly associated with support for market-based distributions, suggesting that these beliefs serve as moral justifications for structural inequalities.
+However, changes in meritocratic perceptions over time reveal a more nuanced picture: while increased emphasis on effort reinforces support for market justice, increased emphasis on talent tends to reduce it.
+This distinction likely reflects broader beliefs about the controllability and fairness of different meritocratic traits.
+Moreover, the interaction between inequality perceptions and meritocratic beliefs indicates that the legitimizing power of meritocracy becomes stronger as perceptions of inequality become less negative—that is, when individuals are less aware or concerned about economic disparities.
+This finding highlights a potential feedback mechanism by which lower sensitivity to inequality may enable stronger endorsement of merit-based explanations for unequal outcomes.
 

### 2025-07-28 resubmission/04-analisis.qmd
Up ajustes (renderizado)

@@
 Interestingly, while perceptions of inequality declined in the latest waves (2022–2023), market justice preferences continued to rise.
-Meritocratic beliefs, in contrast, remain relatively stable overall.
+Meritocratic perceptions, in contrast, remain relatively stable overall.
 Nevertheless, perceptions that individuals are rewarded based on talent tend to be slightly higher than those based on effort.
@@
 Individuals who perceive higher levels of economic inequality tend to prefer less market justice ($\beta$ = -0.002, $p$ > .05).
-However, this effects is no longer statistically significant.
-In Model 6, the meritocratic perception that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.206, $p$ < .001), whereas the perception that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.036).
+However, this effect is no longer statistically significant.
+In Model 6, the meritocratic perception that effort is rewarded is positively associated with market justice preferences ($\beta$ = 0.206, $p$ < .001), whereas the perception that talent is rewarded shows a positive but non-significant coefficient ($\beta$ = 0.036, $p$ > .05).
 

### 2025-07-28 resubmission/05-discussion.qmd
Up ajustes (renderizado)

@@
 
-This study tests several key hypotheses concerning the relationships between perceptions of economic inequality, meritocratic perceptions, and attitudes toward market justice in Chile.
+This study tests several key hypotheses concerning the relationships between perceptions of economic inequality, meritocratic perceptions, and preferences toward market justice in Chile.
 It posits that higher perceived inequality and stronger meritocratic perceptions both independently contribute to greater support for market-based distributions of social goods.
@@
 
-The first hypothesis suggested that perceptions of economic inequality were strongly linked to attitudes toward market justice.
-The findings demonstrated that individuals perceiving higher income disparities showed less support for the distribution of social services based on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021].
+The first hypothesis suggested that perceptions of economic inequality were strongly linked to preferences for market justice.
+The findings demonstrated that individuals perceiving higher income disparities showed less support for the idea that the access to core social services depends on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021].
 This relationship reflected a moral economy where perceptions of systemic unfairness threatened the legitimacy of existing distributions, prompting calls for greater equity.
 Similarly, the negative within-effects observed over time ($H1_{b}$) implied that as perceptions of inequality became more salient, some individuals experienced a decline in support for market justice, possibly due to increasing distrust in market mechanisms or disillusionment with the fairness of the system.
-This indicated that while perception heightened awareness, prolonged exposure or social mobilization could also further foster skepticism about whether redistribution effectively addressed inequality or simply perpetuated systemic flaws.
 

### 2025-07-28 resubmission/06-conclusion.qmd
Up ajustes (renderizado)

@@
 
-This study examined the complex interplay between perceptions of economic inequality, meritocratic perceptions, and attitudes toward market justice in Chile from 2016 to 2023, drawing on longitudinal data from the ELSOC survey.
-By exploring how subjective assessments and social contexts influence support for redistribution and market-based resource allocation, the research offers different elements that contribute to the understanding of the normative foundations underpinning social justice attitudes in a highly unequal, neoliberal environment.
+This study examined the complex interplay between perceptions of economic inequality, meritocratic perceptions, and preferences toward market justice in Chile from 2016 to 2023, drawing on longitudinal data from the ELSOC survey.
+By exploring how subjective assessments and social contexts influence support for redistribution and market-based resource allocation, the research offers different elements that contribute to the understanding of the normative foundations underpinning social justice attitudes in a highly unequal and commoditized environment.
 
-The findings confirm that higher perceptions of economic inequality are associated with less support for market justice attitudes, this is, the belief that it is fine that those with higher income have better social services such as education, pensions and health.
-At the same time, meritocratic beliefs—particularly those emphasizing individual effort—are strongly associated with support for market-based distributions, suggesting that these beliefs serve as moral justifications for structural inequalities.
+The findings confirm that higher perceptions of economic inequality are associated with less support for market justice preferences, this is, the belief that it is fair that those with higher income have better social services such as education, pensions and health.
+At the same time, meritocratic perceptions—particularly those emphasizing individual effort—are strongly associated with support for market-based distributions, suggesting that meritocracy serve as a moral justification for structural inequalities.
 However, changes in meritocratic perceptions over time reveal a more nuanced picture: while increased emphasis on effort reinforces support for market justice, increased emphasis on talent tends to reduce it.
 This distinction likely reflects broader beliefs about the controllability and fairness of different meritocratic traits.
-Moreover, the interaction between inequality perceptions and meritocratic beliefs indicates that the legitimizing power of meritocracy becomes stronger as perceptions of inequality become less negative—that is, when individuals are less aware or concerned about economic disparities.
+Moreover, the interaction between inequality perceptions and meritocratic perceptions indicates that the legitimizing power of meritocracy becomes stronger as perceptions of inequality become less negative—that is, when individuals are less aware or concerned about economic disparities.
 This finding highlights a potential feedback mechanism by which lower sensitivity to inequality may enable stronger endorsement of merit-based explanations for unequal outcomes.
@@
 This research advances the extant literature by integrating subjective perceptions with social and political contexts to explain attitudes toward economic inequality and distributional justice.
-While previous studies primarily focused on objective measures or individual characteristics, this work emphasizes the dynamic and interactional nature of perceptions and beliefs over time.
+While previous studies primarily focused on objective measures or individual characteristics [@busemeyer_skills_2014; @immergut_it_2020; @lindh_public_2015], this work emphasizes the dynamic and interactional nature of perceptions and beliefs over time.
 It highlights the importance of socio-political upheavals in reshaping normative attitudes, underscoring the role of collective action in challenging entrenched narratives of meritocracy and fairness.
@@
 Finally, the potential influence of media, political discourse, and peer effects on perceptions was not directly assessed.
+
 Future research could explore the causal mechanisms linking perceptions and attitudes, possibly through experimental or mixed-method designs.

### 2025-07-28 resubmission/blinded-paper-resubmit.tex
Up ajustes (renderizado)

@@
 \citeproc{ref-kluegel_beliefs_1986}{1986}) and Lane
-(\citeproc{ref-lane_political_1962}{1962};
+(\citeproc{ref-lane_political_1962}{1962},
 \citeproc{ref-lane_market_1986}{1986}). Lane's main aim was to ``help
@@
 treated fairly because they face the same formal rules of engagement
-(\citeproc{ref-lane_market_1986}{Robert E. Lane, 1986}). Consequently,
-the legitimacy of market justice stems from the assumption that
-inequalities are not only inevitable but fair---so long as the rules are
-transparent and opportunities are open. In this way, market justice
-provides a moral justification for inequality by framing it as a
-necessary outcome of individual responsibility
+(\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently, the
+legitimacy of market justice stems from the assumption that inequalities
+are not only inevitable but fair---so long as the rules are transparent
+and opportunities are open. In this way, market justice provides a moral
+justification for inequality by framing it as a necessary outcome of
+individual responsibility
 (\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
@@
 inequality declined in the latest waves (2022--2023), market justice
-preferences continued to rise. Meritocratic beliefs, in contrast, remain
-relatively stable overall. Nevertheless, perceptions that individuals
-are rewarded based on talent tend to be slightly higher than those based
-on effort. Both meritocracy-related measures follow a similar temporal
-pattern: an increase from 2016 to 2018, a decline between 2019 and
-2022---potentially associated with the social unrest and the
-consequences of the COVID-19 pandemic---and a subsequent rise in 2023,
-returning to levels comparable to those observed in 2016.
+preferences continued to rise. Meritocratic perceptions, in contrast,
+remain relatively stable overall. Nevertheless, perceptions that
+individuals are rewarded based on talent tend to be slightly higher than
+those based on effort. Both meritocracy-related measures follow a
+similar temporal pattern: an increase from 2016 to 2018, a decline
+between 2019 and 2022---potentially associated with the social unrest
+and the consequences of the COVID-19 pandemic---and a subsequent rise in
+2023, returning to levels comparable to those observed in 2016.
 
@@
 levels of economic inequality tend to prefer less market justice
-(\(\beta\) = -0.002, \(p\) \textgreater{} .05). However, this effects is
+(\(\beta\) = -0.002, \(p\) \textgreater{} .05). However, this effect is
 no longer statistically significant. In Model 6, the meritocratic
@@
 the perception that talent is rewarded shows a positive but
-non-significant coefficient (\(\beta\) = 0.036).
+non-significant coefficient (\(\beta\) = 0.036, \(p\) \textgreater{}
+.05).
 
@@
 
-\begin{table}
-
-\caption{\label{tbl-interactions2}Interactions for meritocracy,
-perceived economic inequality and market justice preferences}
-
-\centering{
-
-\caption{}
-\begin{center}
-\scalebox{0.7}{
-\begin{tabular}{l c c c c c c}
-\hline
- & Model 12 & Model 13 & Model 14 & Model 15 & Model 16 & Model 17 \\
-\hline
-Intercept                                                & $1.263^{***}$ & $1.265^{***}$ & $1.246^{***}$ & $1.128^{***}$ & $1.099^{***}$ & $1.090^{***}$ \\
-                                                         & $(0.143)$     & $(0.141)$     & $(0.142)$     & $(0.176)$     & $(0.169)$     & $(0.170)$     \\
-Wave x Perception inequality (WE)                        & $0.003$       &               &               &               &               &               \\
-                                                         & $(0.006)$     &               &               &               &               &               \\
-Wave x Merit: effort (WE)                                &               & $0.012$       &               &               &               &               \\
-                                                         &               & $(0.006)$     &               &               &               &               \\
-Wave x Merit: talent (WE)                                &               &               & $0.007$       &               &               &               \\
-                                                         &               &               & $(0.006)$     &               &               &               \\
-Wave x Perception inequality (BE)                        &               &               &               & $-0.011$      &               &               \\
-                                                         &               &               &               & $(0.009)$     &               &               \\
-Wave x Merit: effort (BE)                                &               &               &               &               & $-0.020$      &               \\
-                                                         &               &               &               &               & $(0.011)$     &               \\
-Wave x Merit: talent (BE)                                &               &               &               &               &               & $-0.016$      \\
-                                                         &               &               &               &               &               & $(0.010)$     \\
-\hline
-Controls                                                 & Yes           & Yes           & Yes           & Yes           & Yes           & Yes           \\
-BIC                                                      & $31010.853$   & $30888.601$   & $30785.269$   & $31515.978$   & $31493.536$   & $31501.405$   \\
-Numb. obs.                                               & $8643$        & $8643$        & $8643$        & $8643$        & $8643$        & $8643$        \\
-Num. groups: individuals                                 & $1687$        & $1687$        & $1687$        & $1687$        & $1687$        & $1687$        \\
-Var: individuals (Intercept)                             & $0.388$       & $0.334$       & $0.379$       & $0.274$       & $0.532$       & $0.372$       \\
-Var: individuals, perception inequality cwc              & $0.100$       & $$            & $$            & $$            & $$            & $$            \\
-Var: individuals, wave                                   & $0.025$       & $0.023$       & $0.024$       & $0.021$       & $0.021$       & $0.021$       \\
-Cov: individuals (Intercept), perception inequality cwc  & $-0.081$      & $$            & $$            & $$            & $$            & $$            \\
-Cov: individuals (Intercept), wave                       & $-0.074$      & $-0.062$      & $-0.072$      & $-0.054$      & $-0.056$      & $-0.049$      \\
-Cov: individuals, perception inequality cwc, wave        & $0.021$       & $$            & $$            & $$            & $$            & $$            \\
-Var: Residuals                                           & $0.291$       & $0.280$       & $0.275$       & $0.343$       & $0.342$       & $0.343$       \\
-Var: individuals, merit effort cwc                       & $$            & $0.122$       & $$            & $$            & $$            & $$            \\
-Cov: individuals (Intercept), merit effort cwc           & $$            & $-0.009$      & $$            & $$            & $$            & $$            \\
-Cov: individuals, merit effort cwc, wave                 & $$            & $0.000$       & $$            & $$            & $$            & $$            \\
-Var: individuals, merit talent cwc                       & $$            & $$            & $0.112$       & $$            & $$            & $$            \\
-Cov: individuals (Intercept), merit talent cwc           & $$            & $$            & $0.003$       & $$            & $$            & $$            \\
-Cov: individuals, merit talent cwc, wave                 & $$            & $$            & $-0.003$      & $$            & $$            & $$            \\
-Var: individuals, perception inequality mean             & $$            & $$            & $$            & $0.000$       & $$            & $$            \\
-Cov: individuals (Intercept), perception inequality mean & $$            & $$            & $$            & $0.007$       & $$            & $$            \\
-Cov: individuals, perception inequality mean, wave       & $$            & $$            & $$            & $-0.001$      & $$            & $$            \\
-Var: individuals, merit effort mean                      & $$            & $$            & $$            & $$            & $0.051$       & $$            \\
-Cov: individuals (Intercept), merit effort mean          & $$            & $$            & $$            & $$            & $-0.110$      & $$            \\
-Cov: individuals, merit effort mean, wave                & $$            & $$            & $$            & $$            & $-0.001$      & $$            \\
-Var: individuals, merit talent mean                      & $$            & $$            & $$            & $$            & $$            & $0.031$       \\
-Cov: individuals (Intercept), merit talent mean          & $$            & $$            & $$            & $$            & $$            & $-0.053$      \\
-Cov: individuals, merit talent mean, wave                & $$            & $$            & $$            & $$            & $$            & $-0.003$      \\
-\hline
-\multicolumn{7}{l}{\scriptsize{Note: Cells contain regression coefficients with standard errors in parentheses. $^{***}p<0.001$; $^{**}p<0.01$; $^{*}p<0.05$. CWC = centered within group.}}
-\end{tabular}
-}
-\label{table:coefficients}
-\end{center}
-
-}
-
-\end{table}%
-
-With respect to the temporal dynamics of the key predictors,
-Table~\ref{tbl-interactions2} shows that time has no statistically
-significant effects on either the within- or between-person effects of
-perceived inequality and meritocratic perceptions. The within-person
-interaction effects are positive, suggesting that in later waves, the
-influence of perceived inequality and meritocratic beliefs on market
-justice preferences tends to increase. Conversely, in the between-person
-specification, the interaction effects are negative, indicating that
-over time, the impact of perceived inequality and meritocratic beliefs
-on support for market-based justice tends to diminish. However, these
-trends are not statistically significant, and thus no definitive
-conclusions can be drawn.
+Regarding the temporal dynamics of the key predictors, the analysis
+shows that time has no statistically significant effects on the
+intrapersonal or interpersonal effects of perceived inequality and
+meritocratic perceptions. Further details about this analysis can be
+found in supplementary material.
 
@@
 between perceptions of economic inequality, meritocratic perceptions,
-and attitudes toward market justice in Chile. It posits that higher
+and preferences toward market justice in Chile. It posits that higher
 perceived inequality and stronger meritocratic perceptions both
@@
 The first hypothesis suggested that perceptions of economic inequality
-were strongly linked to attitudes toward market justice. The findings
+were strongly linked to preferences for market justice. The findings
 demonstrated that individuals perceiving higher income disparities
-showed stronger support for the distribution of social services based on
-individual income, aligning with theories emphasizing that awareness of
-inequality could fuel a critical view of such policies
+showed less support for the idea that the access to core social services
+depends on individual income, aligning with theories emphasizing that
+awareness of inequality could fuel a critical view of such policies
 (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
@@
 threatened the legitimacy of existing distributions, prompting calls for
-greater equity. Interestingly, the negative within-effects observed over
-time implied that as perceptions of inequality became more salient, some
-individuals experienced a decline in support for market justice,
-possibly due to increasing distrust in market mechanisms or
-disillusionment with the fairness of the system. This indicated that
-while perception heightened awareness, prolonged exposure or social
-mobilization could also foster skepticism about whether redistribution
-effectively addressed inequality or simply perpetuated systemic flaws.
+greater equity. Similarly, the negative within-effects observed over
+time (\(H1_{b}\)) implied that as perceptions of inequality became more
+salient, some individuals experienced a decline in support for market
+justice, possibly due to increasing distrust in market mechanisms or
+disillusionment with the fairness of the system.
 
@@
 meritocratic perceptions ---particularly those emphasizing individual
-effort and talent--- were associated with stronger support for
-market-based distribution systems. Individuals who viewed success as
-primarily earned through effort were more inclined to justify existing
-inequalities, seeing them as outcomes of individual merit rather than
-systemic injustice. This aligned with research indicating that
-meritocratic narratives served as moral justifications that reinforced
-societal stratification (\citeproc{ref-Castillo2012a_justice}{Castillo,
-2012}; \citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These
-perceptions functioned symbolically to legitimize structural
-disparities, reducing support for redistributive measures by framing
-inequality as fair and deserved. The findings suggested that these
-perceptions bolstered the social hierarchy by fostering acceptance of
-inequality as a reflection of individual virtue rather than structural
-failure, which was especially pertinent in neoliberal contexts like
-Chile's, where market logic heavily influenced social attitudes
+effort--- were associated with stronger support for market-based
+distribution systems. Individuals who viewed success as primarily earned
+through effort were more inclined to justify existing inequalities,
+seeing them as outcomes of individual merit rather than systemic
+injustice. This aligned with research indicating that meritocratic
+narratives served as moral justifications that reinforced societal
+stratification (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
+\citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These perceptions
+functioned symbolically to legitimize structural disparities, reducing
+support for redistributive measures by framing inequality as fair and
+deserved. The findings suggested that these perceptions bolstered the
+social hierarchy by fostering acceptance of inequality as a reflection
+of individual virtue rather than structural failure, which was
+especially pertinent in neoliberal contexts like Chile's, where market
+logic heavily influenced social attitudes
 (\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).
+Interestingly, intraindividual changes in meritocratic perceptions over
+time (\(H2_{b}\)) reveal mixed effects. While an increase in the
+perception that rewards are based on individual effort is associated
+with stronger preferences for market fairness, an increase in the belief
+that rewards come from talent is linked to lower support for such
+principles. One possible explanation is that effort is generally
+perceived as a controllable and malleable trait, which aligns with the
+normative foundations of market fairness. Conversely, talent may be
+viewed as an innate and less alterable attribute, making talent-based
+reward allocation seem less fair or legitimate. Moreover, greater
+exposure to real-life situations where outcomes are strongly determined
+by inherent characteristics, rather than effort or hard work, may lead
+people to question the fairness of market-based reward systems.
 
@@
 meritocratic perceptions and support for market-based distribution
-tended to weaken when perceptions of economic inequality were
-particularly high. Although this moderation is not significant for
-within-subject effects (i.e.~over time), when taking into account
-between-subjects differences we found a positive effect. This appears as
-contrary to our initial hypothesis, as it would mean that effect of
-meritocratic perceptions becomes stronger as perception of inequality
-increases. Nevertheless, we need to be cautious about this
-interpretation, as the main effect of inequality perception on market
-justice is negative. Therefore, as perceived inequality become less
-negative (i.e.~more positive), or closer to zero, then the meritocratic
-perception has an stronger positive relationship on market justice
-preferences. This could mean that meritocratic perceptions play an
-stronger role in justification of market justice inequalities for those
-who are less capable or less prone to perceive inequality.
+tended to strengthen when perceptions of economic inequality were
+particularly high. While this moderation is not significant for
+effort-related effects, it is significant for talent in the within- and
+between-individual effects. This appears as contrary to our initial
+hypothesis, as it would mean that effect of meritocratic perceptions
+becomes stronger as perception of inequality increases. Nevertheless, we
+need to be cautious about this interpretation, as the main effect of
+inequality perception on market justice is negative. Therefore, as
+perceived inequality become less negative (i.e.~more positive), or
+closer to zero, then the meritocratic perception has an stronger
+positive relationship on market justice preferences. This could mean
+that meritocratic perceptions play an stronger role in justification of
+market justice inequalities for those who are less capable or less prone
+to perceive inequality.
 
 Finally, the fourth hypothesis addressed the impact of recent social
-movements, particularly from 2019 onwards, on the relationship between
-meritocratic perceptions, perceptions of inequality, and attitudes
-toward market justice. The results suggested that these movements
-contributed to a more critical stance toward the fairness of existing
-distributional mechanisms. Post-mobilization, the positive relationship
-between meritocratic perceptions and support for inequality appeared
-attenuated, indicating that collective action and public discourse
-during this period fostered questioning of the legitimacy of systemic
-inequalities justified solely through meritocratic narratives. These
-social mobilizations challenged dominant narratives, emphasizing
-structural barriers and systemic injustices, which in turn weakened the
-association between individual effort and deservingness. This is aligned
-with literature emphasizing that social protests act as catalysts for
-transforming societal attitudes by disrupting normative beliefs and
-encouraging more critical evaluations of institutional legitimacy
-(\citeproc{ref-tilly_contentious_2015}{Tilly \& Tarrow, 2015}). The
-findings underscored the importance of socio-political context in
-shaping and reshaping attitudes toward inequality and justice,
-highlighting how collective action could temporarily or permanently
-alter the normative foundations that sustained support for market
-justice.
+movements---particularly those that emerged in 2019---on the
+relationship between meritocratic beliefs, perceptions of inequality,
+and attitudes toward market justice. The main effect of time suggests
+that, on average, support for market justice increased after 2019.
+However, the results do not reveal any significant interactions between
+distributive mechanisms and market justice attitudes, indicating that
+the 2019 social mobilizations did not substantially alter the underlying
+associations between these variables. This may suggest a certain
+stability or resilience in the normative frameworks that shape
+individuals' evaluations of justice, even in the context of major
+collective political events. Alternatively, it is possible that while
+the protests challenged institutional legitimacy, they did not
+fundamentally shift individuals' views regarding how rewards should be
+allocated in society.
 
@@
 This study examined the complex interplay between perceptions of
-economic inequality, meritocratic perceptions, and attitudes toward
+economic inequality, meritocratic perceptions, and preferences toward
 market justice in Chile from 2016 to 2023, drawing on longitudinal data
@@
 contribute to the understanding of the normative foundations
-underpinning social justice attitudes in a highly unequal, neoliberal
-environment.
+underpinning social justice attitudes in a highly unequal and
+commoditized environment.
 
 The findings confirm that higher perceptions of economic inequality are
-associated with less support for market justice attitudes, this is, the
-belief that it is fine that those with higher income have better social
-services such as education, pensions and health. Meritocratic
-perceptions strongly reinforce support for market justice, serving as
-moral justifications for structural inequalities. Additionally,
-perceptions of inequality moderate the influence of meritocratic
-perceptions, such that as the awareness of disparities are attenuated
-(i.e.~less negative) the role of meritocratic perceptions on existing
-market-based allocations become stronger. Notably, the social
-mobilizations of 2019 and subsequent periods seem to have contributed to
-shifting attitudes, fostering greater skepticism of systemic fairness
-and reducing the legitimizing power of meritocratic narratives.
+associated with less support for market justice preferences, this is,
+the belief that it is fair that those with higher income have better
+social services such as education, pensions and health. At the same
+time, meritocratic perceptions---particularly those emphasizing
+individual effort---are strongly associated with support for
+market-based distributions, suggesting that meritocracy serve as a moral
+justification for structural inequalities. However, changes in
+meritocratic perceptions over time reveal a more nuanced picture: while
+increased emphasis on effort reinforces support for market justice,
+increased emphasis on talent tends to reduce it. This distinction likely
+reflects broader beliefs about the controllability and fairness of
+different meritocratic traits. Moreover, the interaction between
+inequality perceptions and meritocratic perceptions indicates that the
+legitimizing power of meritocracy becomes stronger as perceptions of
+inequality become less negative---that is, when individuals are less
+aware or concerned about economic disparities. This finding highlights a
+potential feedback mechanism by which lower sensitivity to inequality
+may enable stronger endorsement of merit-based explanations for unequal
+outcomes.
 
@@
 studies primarily focused on objective measures or individual
-characteristics, this work emphasizes the dynamic and interactional
-nature of perceptions and beliefs over time. It highlights the
-importance of socio-political upheavals in reshaping normative
-attitudes, underscoring the role of collective action in challenging
-entrenched narratives of meritocracy and fairness. The longitudinal
-approach provides a deeper temporal perspective on how societal events
-could influence individual perceptions and preferences.
+characteristics (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
+\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}), this work emphasizes the
+dynamic and interactional nature of perceptions and beliefs over time.
+It highlights the importance of socio-political upheavals in reshaping
+normative attitudes, underscoring the role of collective action in
+challenging entrenched narratives of meritocracy and fairness. The
+longitudinal approach provides a deeper temporal perspective on how
+societal events could influence individual perceptions and preferences.
 
@@
 political discourse, and peer effects on perceptions was not directly
-assessed. Future research could explore the causal mechanisms linking
-perceptions and attitudes, possibly through experimental or mixed-method
-designs. Comparative studies across different national contexts would
-help determine the universality or specificity of these dynamics.
+assessed.
+
+Future research could explore the causal mechanisms linking perceptions
+and attitudes, possibly through experimental or mixed-method designs.
+Comparative studies across different national contexts would help
+determine the universality or specificity of these dynamics.
 Additionally, investigating the role of media, political communication,
@@
 \bibitem[\citeproctext]{ref-lane_political_1962}
-Lane, R. E. (1962). \emph{Political ideology: Why the {American} common
-man believes what he does}. Oxford, England: Free Press of Glencoe.
+Lane, R. (1962). \emph{Political ideology: Why the {American} common man
+believes what he does}. Oxford, England: Free Press of Glencoe.
 
 \bibitem[\citeproctext]{ref-lane_market_1986}
-Lane, Robert E. (1986). Market {Justice}, {Political Justice}.
-\emph{American Political Science Review}, \emph{80}(2), 383--402.
+Lane, R. (1986). Market {Justice}, {Political Justice}. \emph{American
+Political Science Review}, \emph{80}(2), 383--402.
 \url{https://doi.org/10.2307/1958264}
@@
 
-\bibitem[\citeproctext]{ref-tilly_contentious_2015}
-Tilly, C., \& Tarrow, S. (2015). \emph{Contentious {Politics}} (Second
-Edition, New to this Edition:, Second Edition, New to this Edition:).
-Oxford, New York: Oxford University Press.
-
 \bibitem[\citeproctext]{ref-torche_intergenerational_2014a}

### 2025-08-04 resubmission/01-introduction.qmd
update resubmit

@@
 
-The justification of economic inequality has a long tradition in sociology, with seminal works such as the ones from Kluegel [-@kluegel_social_1995a; -@kluegel_beliefs_1981c] and Lane [@lane_political_1962; @lane_market_1986].
-Lane's main aim was to "help explain the tenacious hold of market capitalism on the public mind" (@lane_market_1986, p.384), proposing the distinction between market and political justice.
+The justification of economic inequality has a long tradition in sociology, with seminal works such as the ones from Kluegel [-@kluegel_social_1995a; -@kluegel_beliefs_1986] and Lane [-@lane_political_1962; -@lane_market_1986].
+Lane's main aim was to "help explain the tenacious hold of market capitalism on the public mind" [-@lane_market_1986, p.384], proposing the distinction between market and political justice.
 Whereas political justice refer to government actions aligned with justice principles such equality or need, market justice involves the allocation linked to earned desserts, particularly through individual contributions based on as effort and ability.
-Sociological survey research in this area has considered questions related to the extent to which it is considered just to get better access to social services (such as pensions, health and education) based on payment capacity.
-Although such services could be considered basic citizens rights, the evidence suggest that a relevant share of the population justify inequality in these domains [@lindh_public_2015].
-In the present paper, we argue that there are two relevant factors to take into account when analyzing differences between individuals and groups to this regard: the perception of economic inequality, and the perception of meritocracy.
+Sociological survey research in this area has considered questions related to the extent to which it is considered just to get better access to social services (such as pensions, health and education) based on payment capacity [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015].
+Although these services can be considered fundamental rights of citizenship, existing evidence indicates that a substantial share of the population legitimizes inequality in these domains, and that such preferences are shaped by individuals' socioeconomic characteristics as well as the broader contextual features of their countries [@koos_moral_2019; @lindh_public_2015; @busemeyer_skills_2014].
+In the present paper, we argue that there are two other relevant factors to take into account when analyzing differences between individuals and groups to this regard: the perception of economic inequality and the perception of meritocracy.
 
@@
 Regardless of their measurement, various studies have shown that this perception often underestimates the gap between the rich and the poor, revealing inconsistencies between objective and subjective aspects of inequality [@trump_income_2018].
-Moreover, there is consistent evidence that perceived inequality, even under control of objective inequality, could have implications for attitudes toward the distribution of goods and services [@castillo_perception_2022; @schroder_income_2017].
-redistributive preferences, justification of inequality, and legitimacy of the economic system [@castillo_perception_2022; @garcia-sanchez_perceptions_2018; @garcia-sanchez_attitudes_2020].
-From this perspective, in this paper we argue that perceptions of economic inequality may play a relevant aspect when understanding market justice preferences, as a lower perceived inequality could reinforce the idea that the differences in access to social services could be justified --- a link that becomes difficult to hold when high inequality is perceived.
+Moreover, there is consistent evidence that perceived inequality, even under control of objective inequality, could have implications for attitudes toward the distribution of goods and services [@castillo_perception_2022; @schroder_income_2017], redistributive preferences, justification of inequality, and legitimacy of the economic system [@castillo_perception_2022; @garcia-sanchez_perceptions_2018; @garcia-sanchez_attitudes_2020].
+From this perspective, in this paper we argue that perceptions of economic inequality may play a relevant aspect for understanding market justice preferences, as a lower perceived inequality could reinforce the idea that the differences in access to social services could be justified --- a link that becomes difficult to hold when high inequality is perceived.
 
-Along with perception of inequality, a second focus of this paper is the perception of meritocracy.
+Along with perception of inequality, a second focus of this study is the perception of meritocracy.
 Meritocracy posits that inequality can be legitimized through distributive criteria such as effort and talent [@young_rise_1962; @davis_principles_2001].
 Previous studies have demonstrated that people with stronger meritocratic beliefs tend to perceive less inequality as they attribute economic differences to individual achievements [@mijs_paradox_2021; @wilson_role_2003] and also to justify more inequality as these beliefs are associated with attitudes that legitimize status differences [@batruch_belief_2023].
-In a context as Chile, where the distribution of goods and services is predominantly governed by market logics strongly introduced during the military dictatorship (1973-1989) [@boccardo_30_2020], meritocratic perceptions have the potential to affect the acceptance of social inequalities.
+In a context as Chile, where the distribution of goods and services is predominantly governed by market logics strongly introduced during the military dictatorship (1973-1989) [@madariaga_three_2020; @boccardo_30_2020], meritocratic perceptions have the potential to affect the acceptance of social inequalities.
 Therefore, incorporating the perception of meritocracy in this literature allows for an understanding of how attitudes toward inequality are shaped not only by the perception of economic gaps, but also by beliefs about how these gaps are justified.
@@
 How stable are these associations (within subjects) along time?
-The primary objective of this study is to analyze the interplay between perceptions of inequality and meritocracy and their joint influence on preferences for market justice along time in Chile using the panel survey ELSOC (Social Chilean Longitudinal Study, 2016-2023).
-The chilean case appears as a particular case for studying market justice preferences given their story of high inequality and strong free-market policies introduced in the dictatorship period [@boccardo_30_2020; @flores_top_2020].
-Additionally, this analysis seeks to explore how political and social contingencies---such as the 2019 and 2022 social movements---might have moderated these relationships by prompting more critical reflection on the commodification of social services.
+The primary objective of this study is to examine the interplay between perceptions of inequality and beliefs in meritocracy, and how their interaction shapes preferences for market justice over time in Chile, using data from the ELSOC panel survey (Chilean Longitudinal Social Study, 2016–2023).
+The chilean context appears as a particular case for studying market justice preferences given their story of high inequality and strong free-market policies introduced in the dictatorship period and expanded under democratic governments [@madariaga_three_2020; @flores_top_2020].
+Additionally, this analysis seeks to explore how political and social contingencies---such as the 2019 and 2022 social movements---might have moderated these relationships by prompting more critical reflection on the commodification of social services [@somma_no_2021].
 Recognizing the temporal dimension in shaping market justice preferences is essential, given that historical and contextual factors that could challenge or reaffirm the way in which individuals persceive inequality and meritocracy, and consequantely --- following our argument --- the preferences for market justice.
-
-------------------------------------------------------------------------
-
-Papelera
-
-Research shows that greater perceived inequality fosters stronger preferences for redistribution, regardless of actual inequality levels [@garcia-sanchez_vicious_2019; @gimpelson_misperceiving_2018].
-However, other findings suggest that heightened perceptions of income gaps between high- and low-paying occupations may lead to greater justification of inequality [@Castillo2011].
-
-A less explored area in this regard concerns market justice, which denotes the degree to which individuals consider just that social goods and services (e.g., health, pensions, education) are allocated according to individual contribution, competition, and ability to pay [@lane_market_1986; @lindh_public_2015; @kluegel_legitimation_1999].
-Market justice attitudes reflect the belief that the market promotes procedural fairness---equality of opportunity---so that outcomes result from individual achievement [@lane_market_1986].
-Indeed, Lindh \[-\] shows that market justice preferences are stronger in countries with higher private spending on services than in those with more comprehensive welfare systems.
-
-[lo de la interacción, me parece que es too much para la introducción]
-The perception of meritocracy and the perception of economic inequality would interact intricately in shaping preferences for market justice.
-On the one hand, a stronger perception of merit as a criterion for distribution could minimize the perception of economic gaps, thus justifying an unequal system [@castillo_percepcion_2019; @mijs_paradox_2021].
-On the other hand, the perception of inequality could moderate the impact of meritocracy, as a higher perception of economic gaps might challenge the idea that these are solely based on merit.

### 2025-08-04 resubmission/02-antecedents.qmd
update resubmit

@@
 In contrast, political justice emphasizes principles of equality and need, which are often represented by the welfare state action through social policies.
-An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement [@lane_market_1986].
+An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement and procedures [@lane_market_1986].
 Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open.
-In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of individual responsibility.
+In this way, market justice provides a moral justification for inequality by framing it as a necessary outcome of individual responsibility [@kluegel_legitimation_1999; @lindh_public_2015].
 
@@
 The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded.
-Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019].
+Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019; @lindh_public_2015].
 This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo.
@@
 
-Perceptions of inequality have been associated with attitudes about market justice [@lindh_public_2015; @kluegel_social_1995a].
-Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011].
-In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].ƒ
+Perceptions of inequality have been associated with a range of distributive attitudes [@castillo_perception_2022; @garcia-sanchez_vicious_2019; @trump_income_2018; @mccall_exposure_2017].
+Research indicates that perceptions of inequality influence redistributive preferences to a greater extent than objective or actual economic inequality [@gimpelson_misperceiving_2018].
+Others have shown that the perception of inequality, through the perceived wage gap, is related to the justification of greater inequality [@Castillo2011].
+In that sense, lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011].
+In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@garcia-sanchez_vicious_2019].
 
-Perceptions of inequality refer to individuals' subjective evaluations of the extent, causes, and consequences of income and wealth disparities.
+Perceptions of inequality refer to individuals' subjective evaluations of the extent, causes, and consequences of income and wealth disparities [@akyelken_urban_2020].
 Unlike objective measures such as the Gini index, perceived inequality captures how individuals make sense of distributive hierarchies in their everyday lives, shaped by reference groups, social comparisons, and information environments [@mijs_stratified_2016; @gimpelson_misperceiving_2018; @garcia-castro_perceiving_2020].
@@
 
-In addition to influencing individual attitudes toward inequality, meritocratic beliefs can contribute to social division and the stigmatization of disadvantaged groups.
+In addition to influencing individual attitudes toward inequality, meritocratic perceptions can contribute to social division and the stigmatization of disadvantaged groups.
 Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023].
@@
 
-## The case of Chile
+## The Chilean context
 
-Chile offers an interesting case for examining how public attitudes toward the distribution of social services evolve amid declining poverty and persistently high income inequality within a residual welfare framework [@huber_democracy_2012].
-Following the neoliberal reforms of the 1980s, the Chilean welfare state has become increasingly reliant on private provision, with essential services often commodified and largely reserved for those with sufficient purchasing power [@ferre_welfare_2023].
+Chile offers an interesting case for examining how public attitudes toward the distribution of social services evolve amid declining poverty and persistently high income inequality within a residual welfare framework [@ferre_welfare_2023].
+Following the neoliberal reforms of the 1980s introduced during the civil-military dictatorship (1973-1989), the Chilean welfare state has become increasingly reliant on private provision, with essential services often commodified and largely reserved for those with sufficient purchasing power [@madariaga_three_2020; @boccardo_30_2020].
 This model has favored the affluent, while lower-income groups are left dependent on limited and underfunded public alternatives.
-Research on social stratification characterizes Chile as a society where mobility into intermediate classes is feasible, but access to elite strata remains largely restricted [@torche_unequal_2005].
-While some indicators—such as education and occupation—suggest a relatively fluid class structure, income mobility remains constrained [@EspinozaNunez2014].
-Despite steady economic expansion, Chile continues to rank among the most unequal nations in the OECD, marked by a high Gini coefficient and substantial wealth concentration at the top [@flores_top_2019].
+Research on social stratification characterizes Chile as a society where mobility into intermediate classes is feasible, but access to elite strata remains largely restricted [@lopez-roldan_comparative_2021; @torche_intergenerational_2014a].
+While some indicators—such as education and occupation—suggest a relatively fluid class structure, income mobility remains constrained [@espinoza_movilidad_2014].
+
+Despite steady economic expansion, Chile continues to rank among the most unequal nations in the OECD, marked by a high Gini coefficient and substantial wealth concentration at the top [@flores_top_2020].
 In this context, Chile experienced a wave of mass protests from October 2019 to March 2020.
@@
 In sum, Chile represents an illustrative case of a particular form of *moral economy* [@koos_moral_2019], in which national socioeconomic characteristics, institutional arrangements, and public attitudes coexist.
-This suggests that the country combines persistently high levels of economic and social inequality with a strong institutionalization of market-based mechanisms within the welfare system.
-At the same time, perceptions of inequality are high, and there is widespread public criticism of the commodification of social services.
+This suggests that the country combines persistently high levels of economic and social inequality with a strong institutionalization of market-based mechanisms within the welfare system [@madariaga_three_2020].
+At the same time, perceptions of inequality are high, and there is widespread public criticism of the commodification of social services [@somma_no_2021].
 For the above reasons, we argue that the Chilean case represents a relevant context to study the role of perceived income inequality and meritocracy in shaping public preferences regarding the commodification of welfare services, particularly from the perspective of individual change.
@@
 
-Building upon the previous literature, this study proposes that attitudes toward market justice are shaped by a dynamic interplay between individuals' perceptions of economic inequality and their beliefs in meritocracy.
+Building upon the previous literature, this study proposes that attitudes toward market justice are shaped by a dynamic interplay between individuals' perceptions of economic inequality and their perceptions of meritocracy.
 Specifically, we argue that both perceptions independently and interactively influence the extent to which individuals endorse market-based distributions of social goods and services in Chile.
 
-First, consistent with previous findings, we expect that a higher perception of economic inequality will be associated with lower market justice preferences.
+First, consistent with previous findings about redistributive preferences [@castillo_perception_2022], we expect that a higher perception of economic inequality will be associated with lower market justice preferences.
 When individuals perceive smaller income gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing the acceptance of outcomes based on competition and ability to pay.
 Conversely, a heightened perception of inequality may erode confidence in market fairness, weakening support for market-based distribution.
-This relationship is particularly relevant in the context of Chile, where the neoliberal economic model has been a dominant force in shaping public attitudes toward inequality and justice.
+This relationship is particularly relevant in the context of Chile, where the neoliberal economic model has been a dominant force in shaping public attitudes toward inequality and justice [@canalesceron_sujeto_2021; @mac-clure_justicia_2024].
 
 Second, higher perceived meritocracy is expected to be positively associated with market justice preferences.
-Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit.
+Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit [@castillo_socialization_2024].
 This aligns with the idea that meritocratic perceptions serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit.
@@
 The primary focus of this study is thus to scrutinize how individual-level changes in perceptions relate to changes in preferences over time.
-While we acknowledge the relevance of contextual events — such as the 2019 political outburst or the subsequent constitutional process — for public opinion and political orientation [@disi_pavlic_justification_2025], identifying causal effects of such events (e.g., mass protest exposure) on individual attitudes lies outside the scope of this research.
+While we acknowledge the relevance of contextual events — such as the 2019 political outburst or the subsequent constitutional process — for public opinion and political orientation [@disipavlic_justification_2025], identifying causal effects of such events (e.g., mass protest exposure) on individual attitudes lies outside the scope of this research.
 Instead, we focus on how intra-individual trajectories in perceived inequality and merit shape evolving attitudes toward the commodification of social services.
@@
 
-$H1$: Higher perceived economic inequality is associated with less market justice preferences.
+$H1_{a} \; (\text{Between-person}):$ Individuals who perceive higher levels of economic inequality will report lower support for market justice preferences, on average, compared to those who perceive lower levels of inequality.
 
-$H2$: Higher meritocratic perceptions are associated with higher market justice preferences.
+$H1_{b} \; (\text{Within-person}):$ Increases in an individual's perception of economic inequality over time will be associated with decreases in their support for market justice preferences.
 
-$H3$: The positive association between meritocratic perceptions and market justice preferences is moderated by perceived economic inequality; specifically, this association is weaker when perceived economic inequality is high.
+$H2_{a} \; (\text{Between-person}):$ Individuals who endorse stronger meritocratic perceptions will report higher support for market justice preferences, on average, compared to those with weaker meritocratic perceptions
 
-$H4a$: Changes in perceived income inequality are less strongly associated with changes in market justice attitudes over time.
+$H2_{b} \; (\text{Within-person}):$ Increases in an individual's meritocratic perceptions over time will be associated with increases in their support for market justice preferences.
 
-$H4b$: Changes in meritocracy beliefs are less strongly associated with changes in market justice attitudes over time.
+$H3_{a} \; (\text{Between-person}):$ The positive association between meritocratic perceptions and market justice preferences will be weaker among individuals who perceive higher levels of economic inequality, compared to those who perceive lower inequality.
+
+$H3_{b} \; (\text{Within-person}):$ The positive association between meritocratic perceptions and market justice preferences will be weaker at times when individuals perceive higher levels of inequality, relative to times when they perceive lower inequality.
+
+$H4_{a} \; (\text{Growth moderation – Inequality}):$ The longitudinal association between perceived economic inequality and market justice preferences will strengthen over time, both within and between individuals.
+
+$H4_{b} \; (\text{Growth moderation – Meritocracy}):$ The longitudinal association between meritocratic perceptions and market justice preferences will weaken over time, both within and between individuals.

### 2025-08-04 resubmission/blinded-paper-resubmit.qmd
update resubmit

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "../input/bib/merit-edjust.bib"
# csl: "input/bib/apa6.csl"
csl: "../input/bib/harvard.csl"
# author:
#   - name: Juan Carlos Castillo
#     corresponding: true
#     orcid: 0000-0003-1265-7854
#     email: juancastillov@uchile.cl
#     affiliations:
#       - ref: 1
#       - ref: 2
#       - ref: 3
#   - name: Andreas Laffert
#     orcid: 0000-0002-9008-2454
#     email: alaffertt@estudiante.uc.cl 
#     affiliations:
#       - ref: 4
#   - name: Kevin Carrasco
#     orcid: 0000-0002-9348-8361
#     email: kevin.carrasco@ug.uchile.cl
#     affiliations:
#       - ref: 2
#   - name: Julio Iturra-Sanhueza
#     orcid: 0000-0003-1993-3479
#     email: julioiturrasanhueza@gmail.com 
#     affiliations:
#       - ref: 5     
# affiliations: 
#   - id: 1
#     name: Departamento de Sociología, Universidad de Chile
#   - id: 2
#     name: Centro de estudios del conflicto y cohesión social (COES)
#   - id: 3
#     name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
#   - id: 4
#     name: Instituto de Sociología, Pontificia Universidad Católica de Chile
#   - id: 5
#     name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
 Several countries have experienced a shift toward the privatization and commodification of public goods, welfare policies, and social services. In Latin America, Chile stands out as a paradigmatic case where this trend has led to the extensive marketization of essential services. From a moral economy perspective, the extent to which individuals consider it fair for access to such services to depend on market criteria has been conceptualized as market justice preferences. This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and market justice preferences in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey – ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. The findings show that greater perceived inequality is associated with lower market justice preferences. However, individuals who believe that effort is rewarded are more likely to legitimize existing disparities. In contrast, the perception that talent is rewarded shows a negative effect on market justice preferences; an effect that intensifies as perceived inequality increases over time. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse on justice and fairness. These findings contribute to a deeper understanding of how distributive beliefs evolve in contexts marked by persistent inequality and entrenched neoliberal frameworks.
  \newline
  **Keywords**: Inequality perception, meritocracy, commodification, Chile, public preferences 
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---

```{r setup, include=FALSE}
 knitr::opts_chunk$set(echo=FALSE, warning = FALSE,message = FALSE, cache = TRUE,out.width = '85%',fig.pos= "H"
                       # , fig.align = 'center'
                       )
 # knitr::opts_knit$set(base.url = "../") #relative path for .html output file
 # knitr::opts_knit$set(root.dir = "../") #relative path for chunks within .rmd files
 options(scipen=999)
 options(kableExtra.auto_format = FALSE)
 rm(list=ls())
 options(knitr.kable.NA = '')
 options(knitr.graphics.error = FALSE)
 Sys.setlocale("LC_ALL", "ES_ES.UTF-8")
```

```{r}
# This document was last modified at {{< modified >}}
# 
# and it was last rendered at {{< now >}}

```

{{< include 01-introduction.qmd >}}

{{< include 02-antecedents.qmd >}}

{{< include 03-methods.qmd >}}

{{< include 04-analisis.qmd >}}

{{< include 05-discussion.qmd >}}

{{< include 06-conclusion.qmd >}}

# References

::: {#refs}
:::

### 2025-08-04 resubmission/paper-resubmit.qmd
update resubmit

--- 2025-07-11 paper.qmd
@@
 abstract: | 
-  This study investigates the relationship between perceptions of economic inequality, meritocratic perceptions, and preferences for market justice in Chile between 2016 and 2023.
-Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions.
-Results indicate that rising perceptions of inequality are associated with lower support for market justice; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities.
-The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness.
-These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
+ Several countries have experienced a shift toward the privatization and commodification of public goods, welfare policies, and social services.
+In Latin America, Chile stands out as a paradigmatic case where this trend has led to the extensive marketization of essential services.
+From a moral economy perspective, the extent to which individuals consider it fair for access to such services to depend on market criteria has been conceptualized as market justice preferences.
+This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and market justice preferences in Chile between 2016 and 2023.
+Using six waves of panel data from the Chilean Longitudinal Social Survey – ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions.
+The findings show that greater perceived inequality is associated with lower market justice preferences.
+However, individuals who believe that effort is rewarded are more likely to legitimize existing disparities.
+In contrast, the perception that talent is rewarded shows a negative effect on market justice preferences; an effect that intensifies as perceived inequality increases over time.
+The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse on justice and fairness.
+These findings contribute to a deeper understanding of how distributive beliefs evolve in contexts marked by persistent inequality and entrenched neoliberal frameworks
   \newline
-  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
+ **Keywords**: Inequality perception, meritocracy, commodification, Chile, public preferences 
 link-citations: true

### 2025-08-05 resubmission/01-introduction.qmd
Revert "update resubmit"
This reverts commit 27ec0a816645e6bef02aa45f0b148543bc942541.

@@
 Sociological survey research in this area has considered questions related to the extent to which it is considered just to get better access to social services (such as pensions, health and education) based on payment capacity [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015].
-Although these services can be considered fundamental rights of citizenship, existing evidence indicates that a substantial share of the population legitimizes inequality in these domains, and that such preferences are shaped by individuals' socioeconomic characteristics as well as the broader contextual features of their countries [@koos_moral_2019; @lindh_public_2015; @busemeyer_skills_2014].
+Although these services could be considered basic rights of citizens, evidence to date suggests that a significant portion of the population justifies inequality in these areas [@koos_moral_2019] and that such preferences vary according to objective socioeconomic factors and the contextual characteristics of countries [@lindh_public_2015; @busemeyer_skills_2014].
 In the present paper, we argue that there are two other relevant factors to take into account when analyzing differences between individuals and groups to this regard: the perception of economic inequality and the perception of meritocracy.
@@
 
-Along with perception of inequality, a second focus of this study is the perception of meritocracy.
+Along with perception of inequality, a second focus of this paper is the perception of meritocracy.
 Meritocracy posits that inequality can be legitimized through distributive criteria such as effort and talent [@young_rise_1962; @davis_principles_2001].
@@
 How stable are these associations (within subjects) along time?
-The primary objective of this study is to examine the interplay between perceptions of inequality and beliefs in meritocracy, and how their interaction shapes preferences for market justice over time in Chile, using data from the ELSOC panel survey (Chilean Longitudinal Social Study, 2016–2023).
-The chilean context appears as a particular case for studying market justice preferences given their story of high inequality and strong free-market policies introduced in the dictatorship period and expanded under democratic governments [@madariaga_three_2020; @flores_top_2020].
+The primary objective of this study is to analyze the interplay between perceptions of inequality and meritocracy and their joint influence on preferences for market justice along time in Chile using the panel survey ELSOC (Social Chilean Longitudinal Study, 2016-2023).
+The chilean case appears as a particular case for studying market justice preferences given their story of high inequality and strong free-market policies introduced in the dictatorship period [@madariaga_three_2020; @flores_top_2020].
 Additionally, this analysis seeks to explore how political and social contingencies---such as the 2019 and 2022 social movements---might have moderated these relationships by prompting more critical reflection on the commodification of social services [@somma_no_2021].

### 2025-08-05 resubmission/02-antecedents.qmd
Revert "update resubmit"
This reverts commit 27ec0a816645e6bef02aa45f0b148543bc942541.

@@
 In contrast, political justice emphasizes principles of equality and need, which are often represented by the welfare state action through social policies.
-An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement and procedures [@lane_market_1986].
+An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement [@lane_market_1986].
 Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open.
@@
 Chile offers an interesting case for examining how public attitudes toward the distribution of social services evolve amid declining poverty and persistently high income inequality within a residual welfare framework [@ferre_welfare_2023].
-Following the neoliberal reforms of the 1980s introduced during the civil-military dictatorship (1973-1989), the Chilean welfare state has become increasingly reliant on private provision, with essential services often commodified and largely reserved for those with sufficient purchasing power [@madariaga_three_2020; @boccardo_30_2020].
+Following the neoliberal reforms of the 1980s, the Chilean welfare state has become increasingly reliant on private provision, with essential services often commodified and largely reserved for those with sufficient purchasing power [@madariaga_three_2020; @boccardo_30_2020].
 This model has favored the affluent, while lower-income groups are left dependent on limited and underfunded public alternatives.

### 2025-08-05 resubmission/blinded-paper-resubmit.qmd
Revert "update resubmit"
This reverts commit 27ec0a816645e6bef02aa45f0b148543bc942541.

@@
 editor: source
-bibliography: "../input/bib/merit-edjust.bib"
+bibliography: "input/bib/merit-edjust.bib"
 # csl: "input/bib/apa6.csl"
-csl: "../input/bib/harvard.csl"
+csl: "input/bib/harvard.csl"
 # author:
@@
 The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse on justice and fairness.
-These findings contribute to a deeper understanding of how distributive beliefs evolve in contexts marked by persistent inequality and entrenched neoliberal frameworks.
+These findings contribute to a deeper understanding of how distributive beliefs evolve in contexts marked by persistent inequality and entrenched neoliberal frameworks
   \newline
-  **Keywords**: Inequality perception, meritocracy, commodification, Chile, public preferences 
+  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
 link-citations: true

### 2025-08-05 resubmission/paper-resubmit.qmd
Revert "update resubmit"
This reverts commit 27ec0a816645e6bef02aa45f0b148543bc942541.

@@
   \newline
- **Keywords**: Inequality perception, meritocracy, commodification, Chile, public preferences 
+  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
 link-citations: true

### 2025-08-05 resubmission/01-introduction.qmd
review introduction & antecedentes

@@
 This context of economic disparity has sparked renewed interest in studying not only the objective aspects of inequality, such as income and access to resources, but also its subjective dimensions, including perceptions, beliefs, and associated attitudes [@janmaat_subjective_2013].
-Such dimensions are deemed as key to understand the political delegitimation that economic inequality can elicit [@castillo_perception_2022].
+Such dimensions are deemed as key to understand the  delegitimation of political institutions that economic inequality can elicit [@castillo_perception_2022].
 Therefore, understanding how people perceive inequality is crucial, as these perceptions can influence how societies comprehend and justify (or not) inequality in the distribution of goods and services, as well as their capacity to funnel political discontent and eventually opposition to the current economic distribution.
 
-The justification of economic inequality has a long tradition in sociology, with seminal works such as the ones from Kluegel [-@kluegel_social_1995a; -@kluegel_beliefs_1986] and Lane [-@lane_political_1962; -@lane_market_1986].
+The justification of economic inequality is a topic with a long tradition in sociology, with seminal works such as the ones from Kluegel [-@kluegel_social_1995a; -@kluegel_beliefs_1986] and Lane [-@lane_political_1962; -@lane_market_1986].
 Lane's main aim was to "help explain the tenacious hold of market capitalism on the public mind" [-@lane_market_1986, p.384], proposing the distinction between market and political justice.
-Whereas political justice refer to government actions aligned with justice principles such equality or need, market justice involves the allocation linked to earned desserts, particularly through individual contributions based on as effort and ability.
-Sociological survey research in this area has considered questions related to the extent to which it is considered just to get better access to social services (such as pensions, health and education) based on payment capacity [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015].
-Although these services could be considered basic rights of citizens, evidence to date suggests that a significant portion of the population justifies inequality in these areas [@koos_moral_2019] and that such preferences vary according to objective socioeconomic factors and the contextual characteristics of countries [@lindh_public_2015; @busemeyer_skills_2014].
-In the present paper, we argue that there are two other relevant factors to take into account when analyzing differences between individuals and groups to this regard: the perception of economic inequality and the perception of meritocracy.
+Whereas political justice refer to government actions aligned with justice principles such equality or need, market justice involves the allocation linked to earned desserts, particularly through individual contributions based on individual merit.
+Part of this agenda has focused on analyzing the extent to which individuals consider as just to get better access to social services (such as pensions, health and education) based on payment capacity [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015], usually under the label of market justice preferences.
+The empirical evidence to date suggests that a significant portion of the population justifies inequality in these areas [@koos_moral_2019], and that such preferences vary according to objective socioeconomic factors and countries contextual characteristics [@lindh_public_2015; @busemeyer_skills_2014].
+Within this framework, we argue that there are two relevant factors that are usually sidelined in this literature and that are key to understand market justice preferences: the perception of economic inequality and the perception of meritocracy.
 
@@
 Moreover, there is consistent evidence that perceived inequality, even under control of objective inequality, could have implications for attitudes toward the distribution of goods and services [@castillo_perception_2022; @schroder_income_2017], redistributive preferences, justification of inequality, and legitimacy of the economic system [@castillo_perception_2022; @garcia-sanchez_perceptions_2018; @garcia-sanchez_attitudes_2020].
-From this perspective, in this paper we argue that perceptions of economic inequality may play a relevant aspect for understanding market justice preferences, as a lower perceived inequality could reinforce the idea that the differences in access to social services could be justified --- a link that becomes difficult to hold when high inequality is perceived.
+From this perspective, in this paper we argue that perceptions of economic inequality would play a relevant role for understanding market justice preferences, as a lower perceived inequality could reinforce the idea that the differences in access to social services could be justified --- a link that becomes harder to hold when high inequality is perceived.
 
@@
 Meritocracy posits that inequality can be legitimized through distributive criteria such as effort and talent [@young_rise_1962; @davis_principles_2001].
-Previous studies have demonstrated that people with stronger meritocratic beliefs tend to perceive less inequality as they attribute economic differences to individual achievements [@mijs_paradox_2021; @wilson_role_2003] and also to justify more inequality as these beliefs are associated with attitudes that legitimize status differences [@batruch_belief_2023].
-In a context as Chile, where the distribution of goods and services is predominantly governed by market logics strongly introduced during the military dictatorship (1973-1989) [@madariaga_three_2020; @boccardo_30_2020], meritocratic perceptions have the potential to affect the acceptance of social inequalities.
-Therefore, incorporating the perception of meritocracy in this literature allows for an understanding of how attitudes toward inequality are shaped not only by the perception of economic gaps, but also by beliefs about how these gaps are justified.
-Furthermore, these beliefs are often consolidated from an early age, reinforced by institutions promoting values such as effort and individual skills to climb socially [@castillo_socialization_2024; @reynolds_perceptions_2014].
+Previous studies have shown that those with stronger meritocratic perceptions tend to justify larger inequality, as economic differences are legitimized by being attributed to individual achievements [@mijs_paradox_2021; @wilson_role_2003; @batruch_belief_2023].
+We argue that in contexts where the distribution of goods and services is predominantly governed by market logics, as in the chilean case [@madariaga_three_2020; @boccardo_30_2020], meritocratic perceptions could play a role in the preferences of market justice.
 
-At this point we can propose the following research questions: To what extent the preferences for market justice are associated with perceptions of inequality and meritocracy?
-How stable are these associations (within subjects) along time?
 The primary objective of this study is to analyze the interplay between perceptions of inequality and meritocracy and their joint influence on preferences for market justice along time in Chile using the panel survey ELSOC (Social Chilean Longitudinal Study, 2016-2023).
-The chilean case appears as a particular case for studying market justice preferences given their story of high inequality and strong free-market policies introduced in the dictatorship period [@madariaga_three_2020; @flores_top_2020].
-Additionally, this analysis seeks to explore how political and social contingencies---such as the 2019 and 2022 social movements---might have moderated these relationships by prompting more critical reflection on the commodification of social services [@somma_no_2021].
-Recognizing the temporal dimension in shaping market justice preferences is essential, given that historical and contextual factors that could challenge or reaffirm the way in which individuals persceive inequality and meritocracy, and consequantely --- following our argument --- the preferences for market justice.
+Furthermore, we explore how political and social contingencies---such as the 2019 and 2022 social movements---might have moderated these relationships by prompting more critical reflection on the commodification of social services [@somma_no_2021].
+Recognizing the temporal dimension in shaping preferences is essential, given that historical and contextual factors that could challenge or reaffirm the way in which individuals perceive inequality and meritocracy, resulting in changes in market justice preferences.

### 2025-08-05 resubmission/02-antecedents.qmd
review introduction & antecedentes

@@
-# Theoretical views on market justice, inequality perception, and meritocracy
-
-## The justification of market inequality
+# Market justice preferences
 
@@
 For example, @vondemknesebeck_are_2016 and Immergut and Schneider [-@immergut_it_2020] examine whether citizens consider it fair that individuals with higher incomes can access better healthcare, while Lee and Stacey [-@lee_fairness_2023] apply a similar method in the context of education in Australia.
-These studies usually rely on a survey item asking respondents to evaluate the fairness of income-based access to welfare services, allowing for comparing justice perceptions across different contexts.
-Similarly, comparative studies have also considered these indicators to study the support for market-based distribution in welfare systems [@svallfors_political_2007;@lindh_public_2015].
+These studies usually rely on survey items asking respondents to evaluate the fairness of income-based access to welfare and social services[@svallfors_political_2007;@lindh_public_2015].
 More recently, Castillo et al.
 [-@castillo_socialization_2024] introduced a single-item composite measure of market justice to assess student attitudes toward income-based access to education, healthcare, and pensions in Chile.
-These empirical strategies all aim to capture the extent to which individuals accept inequality when framed as a reflection of market outcomes.
+Such empirical strategies aim to capture the extent to which individuals accept inequality when framed as a reflection of market outcomes.
 
-The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded.
+The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded [@pierson_when_1993].
 Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019; @lindh_public_2015].
 This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo.
-At the same time, political ideology also plays a role ---such as economically conservative values ---  where more right-wing individuals show higher support for meritocracy and more skepticism towards redistribution.
+At the same time, political ideology also plays a role ---such as economically conservative values ---  where  right-wing individuals show higher support for meritocracy and more skepticism towards redistribution.
 This is particularly salient in countries with more restricted public provision of social services.
-For example, Castillo et al.
-[-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that right-wing individuals are more supportive of market-based distribution of welfare.
-Beyond individual characteristics, country-level institutions also play a central role.
-In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015].
-These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions, in line with the policy-feedbacks literature [@campbell_policy_2012; @pierson_when_1993].
+For example, in cases such as Chile [@castillo_socialization_2024] and Australia[@lee_fairness_2023] the evidence suggests that right-wing individuals are more supportive of market-based distribution of welfare.
+Regarding contextual factors, in liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015].
 
-## Perception of inequality
+# Perceptions of inequality and meritocracy
 
-Perceptions of inequality have been associated with a range of distributive attitudes [@castillo_perception_2022; @garcia-sanchez_vicious_2019; @trump_income_2018; @mccall_exposure_2017].
-Research indicates that perceptions of inequality influence redistributive preferences to a greater extent than objective or actual economic inequality [@gimpelson_misperceiving_2018].
-Others have shown that the perception of inequality, through the perceived wage gap, is related to the justification of greater inequality [@Castillo2011].
-In that sense, lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011].
-In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@garcia-sanchez_vicious_2019].
+_The perception of economic inequality_
 
@@
 
-Among the different approaches to measuring perceived inequality, one of the most widely used is the estimation of wage gaps between occupational extremes, such as between a CEO and a manual worker.
-This type of item provides a concrete frame that respondents can relate to more easily than abstract questions about national income distribution [@castillo_percepcion_2019; @easterbrook_social_2021; @willis_legitimacy_2015].
-While it enables the estimation of inequality using simple heuristics, this method is not without challenges.
-For instance, people often lack reliable knowledge about the earnings of those at the top of the income ladder, which leads to high variability in responses and the use of biased mental shortcuts [@knell_perceptions_2020].
-Despite this, perceived wage gaps are strong predictors of political attitudes [@pedersen_attitudes_2019; @garcia-sanchez_perceptions_2018], making them a valuable tool for understanding public responses to economic disparities.
+Perceptions of inequality have been associated with a range of distributive attitudes [@castillo_perception_2022; @garcia-sanchez_vicious_2019; @trump_income_2018; @mccall_exposure_2017].
+Research indicates that perceptions of inequality influence redistributive preferences to a greater extent than objective or actual economic inequality [@gimpelson_misperceiving_2018; @Castillo2011].
+Lower perceived inequality can reinforce support for market-based distributive arrangements [@kuhn_eye_2011].
+In contrast, when inequality is perceived as excessive, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@garcia-sanchez_vicious_2019].
 
-Another relevant critique of this approach lies in its conflation of different psychological constructs.
-Many surveys assess perceived inequality through Likert-type items that ask respondents to agree or disagree with statements, such as "income differences are too large," which captures general concern or discomfort rather than a specific perception [@Castillo2011; @garcia-sanchez_vicious_2019].
-These items mix cognitive estimations with affective evaluations, complicating the interpretation of what respondents perceive versus what they morally reject.
+Assessing the perception of inequality in empirical research is an ongoing challenge.
+Specifically in survey research, the assessment of perceived inequality through Likert-type items such as "income differences are too large" have been questioned as they captures general concern or discomfort rather than a specific perception [@Castillo2011; @garcia-sanchez_vicious_2019].
 As a result, the conceptual clarity between perceived inequality and inequality aversion remains blurred in many empirical studies.
-To address this limitation, recent work has emphasized the need to distinguish between absolute and comparative measures, as well as between ideal and actual estimates of economic gaps [@garcia-sanchez_creencias_2022; @auspurg_why_2017].
+To address this limitation, recent work has emphasized the need to distinguish between absolute and comparative measures, as well as between ideal and perceived estimates of economic gaps [@garcia-sanchez_creencias_2022; @auspurg_why_2017].
+Through asking perceived salaries for occupations in the extremes of the status continuum (such as a CEO and a manual worker), an indirect measure of perceived inequality is built by the ratio of the high and the low status occupation.
+This type of measure provides a concrete frame that respondents can relate to more easily than abstract questions about national income distribution [@castillo_percepcion_2019; @easterbrook_social_2021; @willis_legitimacy_2015].
+Evidence to dat shows that perceived wage gaps are strong predictors of political attitudes [@pedersen_attitudes_2019; @garcia-sanchez_perceptions_2018], making them a valuable tool for understanding public responses to economic disparities.
 
-## Perception of meritocracy
+_Perception of meritocracy_
 

### 2025-08-05 resubmission/paper-resubmit.qmd
review introduction & antecedentes

--- 2025-07-11 resubmission/paper-resubmit.qmd
@@
 abstract: | 
-  This study investigates the relationship between perceptions of economic inequality, meritocratic perceptions, and preferences for market justice in Chile between 2016 and 2023.
-Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions.
-Results indicate that rising perceptions of inequality are associated with lower support for market justice; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities.
-The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness.
-These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
+ Several countries have experienced a shift towards an increasing commodification and private provision of public policies, in areas such as health, education, and pensions.
+In Latin America, Chile stands out as a paradigmatic case of privatizations, which were strongly introduced during the military dictatorship (1973-1989), and maintained after the return of democracy.
+From a moral economy perspective, such institutional design has had deeply cultural consequences in terms of attitudes, ideologies and preferences towards market mechanisms.
+Within this framework, the present study delves into the relationship between perceptions of economic inequality, meritocratic beliefs, and market justice preferences in Chile between 2016 and 2023.
+Using six waves of panel data from the Chilean Longitudinal Social Survey – ELSOC (N=1,687; 8,643 observations along time), we assess whether a lower inequality perception and a higher meritocratic perception increases the preferences for private provision in health, education and pensions.
+The findings support the role of inequality perception, whereas the role of merit perceptions are not as straightforward.
+These findings contribute to a deeper understanding of how distributive beliefs evolve in contexts marked by persistent inequality and entrenched neoliberal frameworks.
   \newline

### 2025-08-05 resubmission/01-introduction.qmd
Update + render
Creo que solo falta mirar en detalle la conclusión @juancarloscastillo y luego lo cerramos con un vistazo final

@@
 Part of this agenda has focused on analyzing the extent to which individuals consider as just to get better access to social services (such as pensions, health and education) based on payment capacity [@busemeyer_skills_2014; @immergut_it_2020; @castillo_socialization_2024; @koos_moral_2019; @lindh_bringing_2023; @lindh_public_2015], usually under the label of market justice preferences.
-The empirical evidence to date suggests that a significant portion of the population justifies inequality in these areas [@koos_moral_2019], and that such preferences vary according to objective socioeconomic factors and countries contextual characteristics [@lindh_public_2015; @busemeyer_skills_2014].
+Although these services can be considered fundamental rights of citizenship, existing evidence indicates that a substantial share of the population legitimizes inequality in these domains, and that such preferences are shaped by individuals' socioeconomic characteristics as well as the broader contextual features of their countries [@koos_moral_2019; @lindh_public_2015; @busemeyer_skills_2014].
 Within this framework, we argue that there are two relevant factors that are usually sidelined in this literature and that are key to understand market justice preferences: the perception of economic inequality and the perception of meritocracy.

### 2025-08-05 resubmission/02-antecedents.qmd
Update + render
Creo que solo falta mirar en detalle la conclusión @juancarloscastillo y luego lo cerramos con un vistazo final

@@
 In contrast, political justice emphasizes principles of equality and need, which are often represented by the welfare state action through social policies.
-An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement [@lane_market_1986].
+An important remark is that the principles of market justice rely on the assumption that markets are neutral and self-regulating arenas, where individuals are treated fairly because they face the same formal rules of engagement and procedures [@lane_market_1986].
 Consequently, the legitimacy of market justice stems from the assumption that inequalities are not only inevitable but fair---so long as the rules are transparent and opportunities are open.
@@
 This type of measure provides a concrete frame that respondents can relate to more easily than abstract questions about national income distribution [@castillo_percepcion_2019; @easterbrook_social_2021; @willis_legitimacy_2015].
-Evidence to dat shows that perceived wage gaps are strong predictors of political attitudes [@pedersen_attitudes_2019; @garcia-sanchez_perceptions_2018], making them a valuable tool for understanding public responses to economic disparities.
+Evidence to date shows that perceived wage gaps are strong predictors of political attitudes [@pedersen_attitudes_2019; @garcia-sanchez_perceptions_2018], making them a valuable tool for understanding public responses to economic disparities.
 
@@
 Chile offers an interesting case for examining how public attitudes toward the distribution of social services evolve amid declining poverty and persistently high income inequality within a residual welfare framework [@ferre_welfare_2023].
-Following the neoliberal reforms of the 1980s, the Chilean welfare state has become increasingly reliant on private provision, with essential services often commodified and largely reserved for those with sufficient purchasing power [@madariaga_three_2020; @boccardo_30_2020].
+Following the neoliberal reforms of the 1980s introduced during the civil-military dictatorship (1973-1989), the Chilean welfare state has become increasingly reliant on private provision, with essential services often commodified and largely reserved for those with sufficient purchasing power [@madariaga_three_2020; @boccardo_30_2020].
 This model has favored the affluent, while lower-income groups are left dependent on limited and underfunded public alternatives.

### 2025-08-05 resubmission/03-methods.qmd
Update + render
Creo que solo falta mirar en detalle la conclusión @juancarloscastillo y luego lo cerramos con un vistazo final

@@
 Longitudinal weights are applied to adjust for both the sampling design and potential biases from systematic non-response.
-Further details on sampling, attrition, and weighting procedures are available at https://coes.cl/encuesta-panel/, and the dataset is publicly accessible at https://dataverse.harvard.edu/dataverse/elsoc.
+Further details on sampling, attrition, and weighting procedures are available at [https://coes.cl/encuesta-panel/](https://coes.cl/encuesta-panel/), and the dataset is publicly accessible at [https://dataverse.harvard.edu/dataverse/elsoc](https://dataverse.harvard.edu/dataverse/elsoc).
 

### 2025-08-05 resubmission/05-discussion.qmd
Update + render
Creo que solo falta mirar en detalle la conclusión @juancarloscastillo y luego lo cerramos con un vistazo final

@@
 
-The first hypothesis suggested that perceptions of economic inequality were strongly linked to preferences for market justice.
-The findings demonstrated that individuals perceiving higher income disparities showed less support for the idea that the access to core social services depends on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021].
-This relationship reflected a moral economy where perceptions of systemic unfairness threatened the legitimacy of existing distributions, prompting calls for greater equity.
-Similarly, the negative within-effects observed over time ($H1_{b}$) implied that as perceptions of inequality became more salient, some individuals experienced a decline in support for market justice, possibly due to increasing distrust in market mechanisms or disillusionment with the fairness of the system.
+The first set of hypothesis suggested that perceptions of economic inequality were strongly linked to preferences for market justice.
+The findings demonstrated that individuals perceiving higher income disparities showed less support for the idea that the access to core social services depends on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021; @garcia-sanchez_vicious_2019].
+Such a relationship reflects a moral economy in which perceptions of systemic unfairness undermine the legitimacy of existing distributions and fuel demands for greater equity.
+In particular, the negative within-individual effect observed over time ($H1_{b}$) suggests that as inequality becomes more salient for some individuals, their support for market justice declines—potentially due to rising distrust in market mechanisms or growing disillusionment with the fairness of the system.
+However, although the effect of perceived inequality at the between-individual level is also negative, it is not statistically significant.
+Therefore, we cannot conclude that individuals who, on average, perceive higher inequality are systematically less supportive of market justice.
+As a result, hypothesis $H1_{a}$ does not receive empirical support.
 
-Regarding the second hypothesis, the results reinforced that meritocratic perceptions ---particularly those emphasizing individual effort--- were associated with stronger support for market-based distribution systems.
-Individuals who viewed success as primarily earned through effort were more inclined to justify existing inequalities, seeing them as outcomes of individual merit rather than systemic injustice.
-This aligned with research indicating that meritocratic narratives served as moral justifications that reinforced societal stratification [@Castillo2012a_justice; @hoyt_mindsets_2023].
-These perceptions functioned symbolically to legitimize structural disparities, reducing support for redistributive measures by framing inequality as fair and deserved.
-The findings suggested that these perceptions bolstered the social hierarchy by fostering acceptance of inequality as a reflection of individual virtue rather than structural failure, which was especially pertinent in neoliberal contexts like Chile's, where market logic heavily influenced social attitudes [@canalesceron_sujeto_2021].
-Interestingly, intraindividual changes in meritocratic perceptions over time ($H2_{b}$) reveal mixed effects.
-While an increase in the perception that rewards are based on individual effort is associated with stronger preferences for market fairness, an increase in the belief that rewards come from talent is linked to lower support for such principles.
-One possible explanation is that effort is generally perceived as a controllable and malleable trait, which aligns with the normative foundations of market fairness.
-Conversely, talent may be viewed as an innate and less alterable attribute, making talent-based reward allocation seem less fair or legitimate.
-Moreover, greater exposure to real-life situations where outcomes are strongly determined by inherent characteristics, rather than effort or hard work, may lead people to question the fairness of market-based reward systems.
+Regarding the second set of hypotheses, the results confirmed that meritocratic perceptions—particularly those emphasizing individual effort—were associated with stronger support for market-based distribution systems.
+Individuals who believed that success is primarily achieved through personal effort were more likely to justify unequal access to core social services based on income, interpreting such disparities as outcomes of individual merit rather than systemic injustice ($H2_{a}$).
+This finding aligns with previous research showing that meritocratic narratives serve as moral justifications that legitimize social stratification [@Castillo2012a_justice; @hoyt_mindsets_2023].
+These perceptions operate symbolically to reinforce structural inequalities by reducing support for redistributive policies, framing inequality as both fair and deserved.
+Consistent with prior work by @castillo_socialization_2024 on Chilean students, the findings suggest that such meritocratic perceptions uphold existing hierarchies by promoting the acceptance of inequality as a reflection of individual virtue rather than structural failure.
+This mechanism is particularly salient in neoliberal contexts like Chile, where market logics heavily shape social attitudes [@canalesceron_sujeto_2021].
 
-The third hypothesis concerned the moderating role of perceptions of inequality on the relationship between meritocracy and market justice attitudes.
-The analysis indicated that the positive association between meritocratic perceptions and support for market-based distribution tended to strengthen when perceptions of economic inequality were particularly high.
-While this moderation is not significant for effort-related effects, it is significant for talent in the within- and between-individual effects.
-This appears as contrary to our initial hypothesis, as it would mean that effect of meritocratic perceptions becomes stronger as perception of inequality increases.
-Nevertheless, we need to be cautious about this interpretation, as the main effect of inequality perception on market justice is negative.
-Therefore, as perceived inequality become less negative (i.e.
-more positive), or closer to zero, then the meritocratic perception has an stronger positive relationship on market justice preferences.
-This could mean that meritocratic perceptions play an stronger role in justification of market justice inequalities for those who are less capable or less prone to perceive inequality.
+Interestingly, intra-individual changes in meritocratic perceptions over time ($H2_{b}$) reveal mixed effects.
+While increases in the belief that rewards are based on individual effort are associated with stronger preferences for market justice, increases in the belief that rewards derive from talent are linked to weaker support for such principles.
+One possible explanation is that effort is generally viewed as a controllable and malleable trait, aligning with the normative foundations of market fairness.
+In contrast, talent tends to be perceived as innate and less subject to personal control, rendering talent-based inequality less legitimate.
+Furthermore, increased exposure to real-world scenarios in which outcomes are clearly shaped by inherent traits rather than hard work may lead individuals to question the fairness and legitimacy of market-based reward systems.
 
-Finally, the fourth hypothesis addressed the impact of recent social movements—particularly those that emerged in 2019—on the relationship between meritocratic beliefs, perceptions of inequality, and attitudes toward market justice.
-The main effect of time suggests that, on average, support for market justice increased after 2019.
-However, the results do not reveal any significant interactions between distributive mechanisms and market justice attitudes, indicating that the 2019 social mobilizations did not substantially alter the underlying associations between these variables.
-This may suggest a certain stability or resilience in the normative frameworks that shape individuals’ evaluations of justice, even in the context of major collective political events.
-Alternatively, it is possible that while the protests challenged institutional legitimacy, they did not fundamentally shift individuals’ views regarding how rewards should be allocated in society.
+The third set of hypotheses addressed the moderating role of perceptions of inequality in the relationship between meritocratic perceptions and preferences for market justice.
+The analysis revealed that the positive association between meritocratic perceptions and support for market-based distribution systems tends to be stronger when perceived economic inequality is high—contradicting our initial expectations ($H3_{a}$ and $H3_{b}$).
+Although this moderating effect is not statistically significant for effort-related meritocratic perceptions, it is significant for talent-related perceptions at both the within- and between-individual levels.
+This finding runs counter to our hypotheses, as it suggests that the effect of meritocratic perceptions—specifically, talent-based meritocracy—intensifies as perceptions of inequality increase.
+However, this interpretation requires caution.
+The main effect of perceived inequality on support for market justice is negative.
+Thus, as perceived inequality becomes less negative (i.e., closer to zero), the positive relationship between talent-based meritocratic perceptions and market justice preferences becomes stronger.
+This suggests a more nuanced dynamic: meritocratic perceptions, particularly those emphasizing talent, may serve as a stronger justificatory mechanism for market-based inequalities among individuals who are less inclined to perceive inequality as a problem.
+In other words, among those who do not strongly perceive systemic disparities, meritocratic narratives may play a more influential role in legitimizing unequal outcomes.
+
+Finally, the fourth hypothesis examined the potential impact of recent social mobilizations—particularly those that erupted in 2019—on the relationship between meritocratic perceptions, perceptions of inequality, and preferences for market justice.
+The main effect of time indicates that, on average, support for market justice increased following the 2019 protests.
+However, the analysis did not detect any significant interactions between distributive beliefs and support for market justice, suggesting that the social mobilizations did not substantially modify the underlying associations between these variables.
+As a result, hypotheses $H4_{a}$ and $H4_{b}$ are not supported by the data.
+This may point to a certain stability—or even resilience—in the normative frameworks that guide individuals’ evaluations of distributive justice, despite the occurrence of major collective political events.
+Alternatively, while the protests may have eroded trust in institutional arrangements or governance, they may not have fundamentally altered individuals’ beliefs about how rewards and resources should be allocated in society.

### 2025-08-05 resubmission/blinded-paper-resubmit.qmd
Update + render
Creo que solo falta mirar en detalle la conclusión @juancarloscastillo y luego lo cerramos con un vistazo final

--- 2025-08-04 resubmission/blinded-paper-resubmit.qmd
@@
 The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse on justice and fairness.
-These findings contribute to a deeper understanding of how distributive beliefs evolve in contexts marked by persistent inequality and entrenched neoliberal frameworks.
+These findings contribute to a deeper understanding of how distributive beliefs evolve in contexts marked by persistent inequality and entrenched neoliberal frameworks
   \newline

### 2025-08-05 resubmission/blinded-paper-resubmit.tex
Update + render
Creo que solo falta mirar en detalle la conclusión @juancarloscastillo y luego lo cerramos con un vistazo final

@@
 and entrenched neoliberal frameworks \newline \textbf{Keywords}:
-Economic inequality, meritocracy, market justice, Chile, public
-preferences, inequality perception
+Inequality perception, meritocracy, commodification, Chile, public
+preferences
 \end{abstract}
@@
 (\citeproc{ref-janmaat_subjective_2013}{Janmaat, 2013}). Such dimensions
-are deemed as key to understand the political delegitimation that
-economic inequality can elicit
+are deemed as key to understand the delegitimation of political
+institutions that economic inequality can elicit
 (\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}).
@@
 
-The justification of economic inequality has a long tradition in
-sociology, with seminal works such as the ones from Kluegel
+The justification of economic inequality is a topic with a long
+tradition in sociology, with seminal works such as the ones from Kluegel
 (\citeproc{ref-kluegel_social_1995a}{1995};
@@
 earned desserts, particularly through individual contributions based on
-as effort and ability. Sociological survey research in this area has
-considered questions related to the extent to which it is considered
-just to get better access to social services (such as pensions, health
-and education) based on payment capacity
-(\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
+individual merit. Part of this agenda has focused on analyzing the
+extent to which individuals consider as just to get better access to
+social services (such as pensions, health and education) based on
+payment capacity (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
 \citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
@@
 \citeproc{ref-lindh_public_2015}{Lindh, 2015};
-\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}). Although
-these services could be considered basic rights of citizens, evidence to
-date suggests that a significant portion of the population justifies
-inequality in these areas (\citeproc{ref-koos_moral_2019}{Koos \&
-Sachweh, 2019}) and that such preferences vary according to objective
-socioeconomic factors and the contextual characteristics of countries
+\citeproc{ref-lindh_bringing_2023}{Lindh \& McCall, 2023}), usually
+under the label of market justice preferences. Although these services
+can be considered fundamental rights of citizenship, existing evidence
+indicates that a substantial share of the population legitimizes
+inequality in these domains, and that such preferences are shaped by
+individuals' socioeconomic characteristics as well as the broader
+contextual features of their countries
 (\citeproc{ref-busemeyer_skills_2014}{Busemeyer, 2014};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). In the present paper, we
-argue that there are two other relevant factors to take into account
-when analyzing differences between individuals and groups to this
-regard: the perception of economic inequality and the perception of
+\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015}). Within this framework,
+we argue that there are two relevant factors that are usually sidelined
+in this literature and that are key to understand market justice
+preferences: the perception of economic inequality and the perception of
 meritocracy.
@@
 perspective, in this paper we argue that perceptions of economic
-inequality may play a relevant aspect for understanding market justice
+inequality would play a relevant role for understanding market justice
 preferences, as a lower perceived inequality could reinforce the idea
 that the differences in access to social services could be justified ---
-a link that becomes difficult to hold when high inequality is perceived.
+a link that becomes harder to hold when high inequality is perceived.
 
@@
 \citeproc{ref-young_rise_1962}{Young, 1962}). Previous studies have
-demonstrated that people with stronger meritocratic beliefs tend to
-perceive less inequality as they attribute economic differences to
-individual achievements (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
-\citeproc{ref-wilson_role_2003}{Wilson, 2003}) and also to justify more
-inequality as these beliefs are associated with attitudes that
-legitimize status differences
-(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}). In a context
-as Chile, where the distribution of goods and services is predominantly
-governed by market logics strongly introduced during the military
-dictatorship (1973-1989) (\citeproc{ref-boccardo_30_2020}{Boccardo,
-2020}; \citeproc{ref-madariaga_three_2020}{Madariaga, 2020}),
-meritocratic perceptions have the potential to affect the acceptance of
-social inequalities. Therefore, incorporating the perception of
-meritocracy in this literature allows for an understanding of how
-attitudes toward inequality are shaped not only by the perception of
-economic gaps, but also by beliefs about how these gaps are justified.
-Furthermore, these beliefs are often consolidated from an early age,
-reinforced by institutions promoting values such as effort and
-individual skills to climb socially
-(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024};
-\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian, 2014}).
-
-At this point we can propose the following research questions: To what
-extent the preferences for market justice are associated with
-perceptions of inequality and meritocracy? How stable are these
-associations (within subjects) along time? The primary objective of this
-study is to analyze the interplay between perceptions of inequality and
-meritocracy and their joint influence on preferences for market justice
-along time in Chile using the panel survey ELSOC (Social Chilean
-Longitudinal Study, 2016-2023). The chilean case appears as a particular
-case for studying market justice preferences given their story of high
-inequality and strong free-market policies introduced in the
-dictatorship period (\citeproc{ref-flores_top_2020}{Flores et al.,
-2020}; \citeproc{ref-madariaga_three_2020}{Madariaga, 2020}).
-Additionally, this analysis seeks to explore how political and social
-contingencies---such as the 2019 and 2022 social movements---might have
-moderated these relationships by prompting more critical reflection on
-the commodification of social services
-(\citeproc{ref-somma_no_2021}{Somma et al., 2021}). Recognizing the
-temporal dimension in shaping market justice preferences is essential,
-given that historical and contextual factors that could challenge or
-reaffirm the way in which individuals persceive inequality and
-meritocracy, and consequantely --- following our argument --- the
-preferences for market justice.
-
-\section{Theoretical views on market justice, inequality perception, and
-meritocracy}\label{theoretical-views-on-market-justice-inequality-perception-and-meritocracy}
-
-\subsection{The justification of market
-inequality}\label{the-justification-of-market-inequality}
+shown that those with stronger meritocratic perceptions tend to justify
+larger inequality, as economic differences are legitimized by being
+attributed to individual achievements
+(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
+\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
+\citeproc{ref-wilson_role_2003}{Wilson, 2003}). We argue that in
+contexts where the distribution of goods and services is predominantly
+governed by market logics, as in the chilean case
+(\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
+\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}), meritocratic
+perceptions could play a role in the preferences of market justice.
+
+The primary objective of this study is to analyze the interplay between
+perceptions of inequality and meritocracy and their joint influence on
+preferences for market justice along time in Chile using the panel
+survey ELSOC (Social Chilean Longitudinal Study, 2016-2023).
+Furthermore, we explore how political and social contingencies---such as
+the 2019 and 2022 social movements---might have moderated these
+relationships by prompting more critical reflection on the
+commodification of social services (\citeproc{ref-somma_no_2021}{Somma
+et al., 2021}). Recognizing the temporal dimension in shaping
+preferences is essential, given that historical and contextual factors
+that could challenge or reaffirm the way in which individuals perceive
+inequality and meritocracy, resulting in changes in market justice
+preferences.
+
+\section{Market justice preferences}\label{market-justice-preferences}
 
@@
 markets are neutral and self-regulating arenas, where individuals are
-treated fairly because they face the same formal rules of engagement
-(\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently, the
-legitimacy of market justice stems from the assumption that inequalities
-are not only inevitable but fair---so long as the rules are transparent
-and opportunities are open. In this way, market justice provides a moral
-justification for inequality by framing it as a necessary outcome of
-individual responsibility
+treated fairly because they face the same formal rules of engagement and
+procedures (\citeproc{ref-lane_market_1986}{Lane, 1986}). Consequently,
+the legitimacy of market justice stems from the assumption that
+inequalities are not only inevitable but fair---so long as the rules are
+transparent and opportunities are open. In this way, market justice
+provides a moral justification for inequality by framing it as a
+necessary outcome of individual responsibility
 (\citeproc{ref-kluegel_legitimation_1999}{Kluegel et al., 1999};
@@
 (\citeproc{ref-lee_fairness_2023}{2023}) apply a similar method in the
-context of education in Australia. These studies usually rely on a
-survey item asking respondents to evaluate the fairness of income-based
-access to welfare services, allowing for comparing justice perceptions
-across different contexts. Similarly, comparative studies have also
-considered these indicators to study the support for market-based
-distribution in welfare systems (\citeproc{ref-lindh_public_2015}{Lindh,
+context of education in Australia. These studies usually rely on survey
+items asking respondents to evaluate the fairness of income-based access
+to welfare and social services(\citeproc{ref-lindh_public_2015}{Lindh,
 2015}; \citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). More
@@
 attitudes toward income-based access to education, healthcare, and
-pensions in Chile. These empirical strategies all aim to capture the
-extent to which individuals accept inequality when framed as a
-reflection of market outcomes.
+pensions in Chile. Such empirical strategies aim to capture the extent
+to which individuals accept inequality when framed as a reflection of
+market outcomes.
 
@@
 they are shaped by individuals' socioeconomic position, normative
-beliefs, and the institutional context in which they are embedded.
-Across the literature, there is empirical evidence suggesting that
-individuals in more advantaged socioeconomic positions---those with
-higher occupational class, income, and education---are more likely to
-support market justice principles (\citeproc{ref-koos_moral_2019}{Koos
-\& Sachweh, 2019}; \citeproc{ref-lindh_public_2015}{Lindh, 2015};
+beliefs, and the institutional context in which they are embedded
+(\citeproc{ref-pierson_when_1993}{Pierson, 1993}). Across the
+literature, there is empirical evidence suggesting that individuals in
+more advantaged socioeconomic positions---those with higher occupational
+class, income, and education---are more likely to support market justice
+principles (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh, 2019};
+\citeproc{ref-lindh_public_2015}{Lindh, 2015};
 \citeproc{ref-svallfors_political_2007}{Svallfors, 2007}). This tendency
@@
 also plays a role ---such as economically conservative values --- where
-more right-wing individuals show higher support for meritocracy and more
+right-wing individuals show higher support for meritocracy and more
 skepticism towards redistribution. This is particularly salient in
 countries with more restricted public provision of social services. For
-example, Castillo et al.
-(\citeproc{ref-castillo_socialization_2024}{2024}) in Chile and, Lee and
-Stacey (\citeproc{ref-lee_fairness_2023}{2023}) in Australia show that
-right-wing individuals are more supportive of market-based distribution
-of welfare. Beyond individual characteristics, country-level
-institutions also play a central role. In liberal welfare regimes like
-those of the United States or the United Kingdom, market justice
-preferences are more widespread, while in coordinated or
-social-democratic regimes---such as Sweden or Germany---citizens are
-generally more critical of market-based inequalities
-(\citeproc{ref-immergut_it_2020}{Immergut \& Schneider, 2020};
-\citeproc{ref-lindh_public_2015}{Lindh, 2015}). These findings suggest
-that market justice is also shaped contextually, which is represented by
-the development of welfare institutions, in line with the
-policy-feedbacks literature
-(\citeproc{ref-campbell_policy_2012}{Campbell, 2012};
-\citeproc{ref-pierson_when_1993}{Pierson, 1993}).
-
-\subsection{Perception of inequality}\label{perception-of-inequality}
-
-Perceptions of inequality have been associated with a range of
-distributive attitudes (\citeproc{ref-castillo_perception_2022}{Castillo
-et al., 2022}; \citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez
-et al., 2019}; \citeproc{ref-mccall_exposure_2017}{McCall et al., 2017};
-\citeproc{ref-trump_income_2018}{Trump, 2018}). Research indicates that
-perceptions of inequality influence redistributive preferences to a
-greater extent than objective or actual economic inequality
-(\citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \& Treisman,
-2018}). Others have shown that the perception of inequality, through the
-perceived wage gap, is related to the justification of greater
-inequality (\citeproc{ref-Castillo2011}{Castillo, 2011}). In that sense,
-lower perceived inequality can reinforce support for market-based
-distributive arrangements by suggesting that the system is fair and that
-outcomes reflect effort and ability (\citeproc{ref-kuhn_eye_2011}{Kuhn,
-2011}). In contrast, when inequality is perceived as excessive or
-structurally determined, individuals could question the legitimacy of
-market justice and become more supportive of redistributive policies
-(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}).
+example, in cases such as Chile
+(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}) and
+Australia(\citeproc{ref-lee_fairness_2023}{Lee \& Stacey, 2023}) the
+evidence suggests that right-wing individuals are more supportive of
+market-based distribution of welfare. Regarding contextual factors, in
+liberal welfare regimes like those of the United States or the United
+Kingdom, market justice preferences are more widespread, while in
+coordinated or social-democratic regimes---such as Sweden or
+Germany---citizens are generally more critical of market-based
+inequalities (\citeproc{ref-immergut_it_2020}{Immergut \& Schneider,
+2020}; \citeproc{ref-lindh_public_2015}{Lindh, 2015}).
+
+\section{Perceptions of inequality and
+meritocracy}\label{perceptions-of-inequality-and-meritocracy}
+
+\emph{The perception of economic inequality}
 
@@
 
-Among the different approaches to measuring perceived inequality, one of
-the most widely used is the estimation of wage gaps between occupational
-extremes, such as between a CEO and a manual worker. This type of item
+Perceptions of inequality have been associated with a range of
+distributive attitudes (\citeproc{ref-castillo_perception_2022}{Castillo
+et al., 2022}; \citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez
+et al., 2019}; \citeproc{ref-mccall_exposure_2017}{McCall et al., 2017};
+\citeproc{ref-trump_income_2018}{Trump, 2018}). Research indicates that
+perceptions of inequality influence redistributive preferences to a
+greater extent than objective or actual economic inequality
+(\citeproc{ref-Castillo2011}{Castillo, 2011};
+\citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \& Treisman,
+2018}). Lower perceived inequality can reinforce support for
+market-based distributive arrangements
+(\citeproc{ref-kuhn_eye_2011}{Kuhn, 2011}). In contrast, when inequality
+is perceived as excessive, individuals could question the legitimacy of
+market justice and become more supportive of redistributive policies
+(\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
+2019}).
+
+Assessing the perception of inequality in empirical research is an
+ongoing challenge. Specifically in survey research, the assessment of
+perceived inequality through Likert-type items such as ``income
+differences are too large'' have been questioned as they captures
+general concern or discomfort rather than a specific perception
+(\citeproc{ref-Castillo2011}{Castillo, 2011};
+\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
+2019}). As a result, the conceptual clarity between perceived inequality
+and inequality aversion remains blurred in many empirical studies. To
+address this limitation, recent work has emphasized the need to
+distinguish between absolute and comparative measures, as well as
+between ideal and perceived estimates of economic gaps
+(\citeproc{ref-auspurg_why_2017}{Auspurg et al., 2017};
+\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De
+Carvalho, 2022}). Through asking perceived salaries for occupations in
+the extremes of the status continuum (such as a CEO and a manual
+worker), an indirect measure of perceived inequality is built by the
+ratio of the high and the low status occupation. This type of measure
 provides a concrete frame that respondents can relate to more easily
@@
 \citeproc{ref-easterbrook_social_2021}{Easterbrook, 2021};
-\citeproc{ref-willis_legitimacy_2015}{Willis et al., 2015}). While it
-enables the estimation of inequality using simple heuristics, this
-method is not without challenges. For instance, people often lack
-reliable knowledge about the earnings of those at the top of the income
-ladder, which leads to high variability in responses and the use of
-biased mental shortcuts (\citeproc{ref-knell_perceptions_2020}{Knell \&
-Stix, 2020}). Despite this, perceived wage gaps are strong predictors of
-political attitudes
-(\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez et al.,
-2018}; \citeproc{ref-pedersen_attitudes_2019}{Pedersen \& Mutz, 2019}),
-making them a valuable tool for understanding public responses to
-economic disparities.
-
-Another relevant critique of this approach lies in its conflation of
-different psychological constructs. Many surveys assess perceived
-inequality through Likert-type items that ask respondents to agree or
-disagree with statements, such as ``income differences are too large,''
-which captures general concern or discomfort rather than a specific
-perception (\citeproc{ref-Castillo2011}{Castillo, 2011};
-\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al.,
-2019}). These items mix cognitive estimations with affective
-evaluations, complicating the interpretation of what respondents
-perceive versus what they morally reject. As a result, the conceptual
-clarity between perceived inequality and inequality aversion remains
-blurred in many empirical studies. To address this limitation, recent
-work has emphasized the need to distinguish between absolute and
-comparative measures, as well as between ideal and actual estimates of
-economic gaps (\citeproc{ref-auspurg_why_2017}{Auspurg et al., 2017};
-\citeproc{ref-garcia-sanchez_creencias_2022}{García-Sánchez \& De
-Carvalho, 2022}).
-
-\subsection{Perception of meritocracy}\label{perception-of-meritocracy}
+\citeproc{ref-willis_legitimacy_2015}{Willis et al., 2015}). Evidence to
+date shows that perceived wage gaps are strong predictors of political
+attitudes (\citeproc{ref-garcia-sanchez_perceptions_2018}{García-Sánchez
+et al., 2018}; \citeproc{ref-pedersen_attitudes_2019}{Pedersen \& Mutz,
+2019}), making them a valuable tool for understanding public responses
+to economic disparities.
+
+\emph{Perception of meritocracy}
 
@@
 framework (\citeproc{ref-ferre_welfare_2023}{Ferre, 2023}). Following
-the neoliberal reforms of the 1980s, the Chilean welfare state has
-become increasingly reliant on private provision, with essential
-services often commodified and largely reserved for those with
-sufficient purchasing power (\citeproc{ref-boccardo_30_2020}{Boccardo,
-2020}; \citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). This model
-has favored the affluent, while lower-income groups are left dependent
-on limited and underfunded public alternatives. Research on social
+the neoliberal reforms of the 1980s introduced during the civil-military
+dictatorship (1973-1989), the Chilean welfare state has become
+increasingly reliant on private provision, with essential services often
+commodified and largely reserved for those with sufficient purchasing
+power (\citeproc{ref-boccardo_30_2020}{Boccardo, 2020};
+\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). This model has
+favored the affluent, while lower-income groups are left dependent on
+limited and underfunded public alternatives. Research on social
 stratification characterizes Chile as a society where mobility into
@@
 and weighting procedures are available at
-https://coes.cl/encuesta-panel/, and the dataset is publicly accessible
-at https://dataverse.harvard.edu/dataverse/elsoc.
+\url{https://coes.cl/encuesta-panel/}, and the dataset is publicly
+accessible at \url{https://dataverse.harvard.edu/dataverse/elsoc}.
 
@@
 
-The first hypothesis suggested that perceptions of economic inequality
-were strongly linked to preferences for market justice. The findings
-demonstrated that individuals perceiving higher income disparities
-showed less support for the idea that the access to core social services
-depends on individual income, aligning with theories emphasizing that
-awareness of inequality could fuel a critical view of such policies
+The first set of hypothesis suggested that perceptions of economic
+inequality were strongly linked to preferences for market justice. The
+findings demonstrated that individuals perceiving higher income
+disparities showed less support for the idea that the access to core
+social services depends on individual income, aligning with theories
+emphasizing that awareness of inequality could fuel a critical view of
+such policies (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
+\citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al., 2019};
+\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). Such a relationship
+reflects a moral economy in which perceptions of systemic unfairness
+undermine the legitimacy of existing distributions and fuel demands for
+greater equity. In particular, the negative within-individual effect
+observed over time (\(H1_{b}\)) suggests that as inequality becomes more
+salient for some individuals, their support for market justice
+declines---potentially due to rising distrust in market mechanisms or
+growing disillusionment with the fairness of the system. However,
+although the effect of perceived inequality at the between-individual
+level is also negative, it is not statistically significant. Therefore,
+we cannot conclude that individuals who, on average, perceive higher
+inequality are systematically less supportive of market justice. As a
+result, hypothesis \(H1_{a}\) does not receive empirical support.
+
+Regarding the second set of hypotheses, the results confirmed that
+meritocratic perceptions---particularly those emphasizing individual
+effort---were associated with stronger support for market-based
+distribution systems. Individuals who believed that success is primarily
+achieved through personal effort were more likely to justify unequal
+access to core social services based on income, interpreting such
+disparities as outcomes of individual merit rather than systemic
+injustice (\(H2_{a}\)). This finding aligns with previous research
+showing that meritocratic narratives serve as moral justifications that
+legitimize social stratification
 (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
-\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). This relationship
-reflected a moral economy where perceptions of systemic unfairness
-threatened the legitimacy of existing distributions, prompting calls for
-greater equity. Similarly, the negative within-effects observed over
-time (\(H1_{b}\)) implied that as perceptions of inequality became more
-salient, some individuals experienced a decline in support for market
-justice, possibly due to increasing distrust in market mechanisms or
-disillusionment with the fairness of the system.
-
-Regarding the second hypothesis, the results reinforced that
-meritocratic perceptions ---particularly those emphasizing individual
-effort--- were associated with stronger support for market-based
-distribution systems. Individuals who viewed success as primarily earned
-through effort were more inclined to justify existing inequalities,
-seeing them as outcomes of individual merit rather than systemic
-injustice. This aligned with research indicating that meritocratic
-narratives served as moral justifications that reinforced societal
-stratification (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
 \citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023}). These perceptions
-functioned symbolically to legitimize structural disparities, reducing
-support for redistributive measures by framing inequality as fair and
-deserved. The findings suggested that these perceptions bolstered the
-social hierarchy by fostering acceptance of inequality as a reflection
-of individual virtue rather than structural failure, which was
-especially pertinent in neoliberal contexts like Chile's, where market
-logic heavily influenced social attitudes
+operate symbolically to reinforce structural inequalities by reducing
+support for redistributive policies, framing inequality as both fair and
+deserved. Consistent with prior work by Castillo et al.
+(\citeproc{ref-castillo_socialization_2024}{2024}) on Chilean students,
+the findings suggest that such meritocratic perceptions uphold existing
+hierarchies by promoting the acceptance of inequality as a reflection of
+individual virtue rather than structural failure. This mechanism is
+particularly salient in neoliberal contexts like Chile, where market
+logics heavily shape social attitudes
 (\citeproc{ref-canalesceron_sujeto_2021}{Canales Cerón et al., 2021}).
-Interestingly, intraindividual changes in meritocratic perceptions over
-time (\(H2_{b}\)) reveal mixed effects. While an increase in the
-perception that rewards are based on individual effort is associated
-with stronger preferences for market fairness, an increase in the belief
-that rewards come from talent is linked to lower support for such
-principles. One possible explanation is that effort is generally
-perceived as a controllable and malleable trait, which aligns with the
-normative foundations of market fairness. Conversely, talent may be
-viewed as an innate and less alterable attribute, making talent-based
-reward allocation seem less fair or legitimate. Moreover, greater
-exposure to real-life situations where outcomes are strongly determined
-by inherent characteristics, rather than effort or hard work, may lead
-people to question the fairness of market-based reward systems.
-
-The third hypothesis concerned the moderating role of perceptions of
-inequality on the relationship between meritocracy and market justice
-attitudes. The analysis indicated that the positive association between
-meritocratic perceptions and support for market-based distribution
-tended to strengthen when perceptions of economic inequality were
-particularly high. While this moderation is not significant for
-effort-related effects, it is significant for talent in the within- and
-between-individual effects. This appears as contrary to our initial
-hypothesis, as it would mean that effect of meritocratic perceptions
-becomes stronger as perception of inequality increases. Nevertheless, we
-need to be cautious about this interpretation, as the main effect of
-inequality perception on market justice is negative. Therefore, as
-perceived inequality become less negative (i.e.~more positive), or
-closer to zero, then the meritocratic perception has an stronger
-positive relationship on market justice preferences. This could mean
-that meritocratic perceptions play an stronger role in justification of
-market justice inequalities for those who are less capable or less prone
-to perceive inequality.
-
-Finally, the fourth hypothesis addressed the impact of recent social
-movements---particularly those that emerged in 2019---on the
-relationship between meritocratic beliefs, perceptions of inequality,
-and attitudes toward market justice. The main effect of time suggests
-that, on average, support for market justice increased after 2019.
-However, the results do not reveal any significant interactions between
-distributive mechanisms and market justice attitudes, indicating that
-the 2019 social mobilizations did not substantially alter the underlying
-associations between these variables. This may suggest a certain
-stability or resilience in the normative frameworks that shape
-individuals' evaluations of justice, even in the context of major
-collective political events. Alternatively, it is possible that while
-the protests challenged institutional legitimacy, they did not
-fundamentally shift individuals' views regarding how rewards should be
-allocated in society.
+
+Interestingly, intra-individual changes in meritocratic perceptions over
+time (\(H2_{b}\)) reveal mixed effects. While increases in the belief
+that rewards are based on individual effort are associated with stronger
+preferences for market justice, increases in the belief that rewards
+derive from talent are linked to weaker support for such principles. One
+possible explanation is that effort is generally viewed as a
+controllable and malleable trait, aligning with the normative
+foundations of market fairness. In contrast, talent tends to be
+perceived as innate and less subject to personal control, rendering
+talent-based inequality less legitimate. Furthermore, increased exposure
+to real-world scenarios in which outcomes are clearly shaped by inherent
+traits rather than hard work may lead individuals to question the
+fairness and legitimacy of market-based reward systems.
+
+The third set of hypotheses addressed the moderating role of perceptions
+of inequality in the relationship between meritocratic perceptions and
+preferences for market justice. The analysis revealed that the positive
+association between meritocratic perceptions and support for
+market-based distribution systems tends to be stronger when perceived
+economic inequality is high---contradicting our initial expectations
+(\(H3_{a}\) and \(H3_{b}\)). Although this moderating effect is not
+statistically significant for effort-related meritocratic perceptions,
+it is significant for talent-related perceptions at both the within- and
+between-individual levels. This finding runs counter to our hypotheses,
+as it suggests that the effect of meritocratic
+perceptions---specifically, talent-based meritocracy---intensifies as
+perceptions of inequality increase. However, this interpretation
+requires caution. The main effect of perceived inequality on support for
+market justice is negative. Thus, as perceived inequality becomes less
+negative (i.e., closer to zero), the positive relationship between
+talent-based meritocratic perceptions and market justice preferences
+becomes stronger. This suggests a more nuanced dynamic: meritocratic
+perceptions, particularly those emphasizing talent, may serve as a
+stronger justificatory mechanism for market-based inequalities among
+individuals who are less inclined to perceive inequality as a problem.
+In other words, among those who do not strongly perceive systemic
+disparities, meritocratic narratives may play a more influential role in
+legitimizing unequal outcomes.
+
+Finally, the fourth hypothesis examined the potential impact of recent
+social mobilizations---particularly those that erupted in 2019---on the
+relationship between meritocratic perceptions, perceptions of
+inequality, and preferences for market justice. The main effect of time
+indicates that, on average, support for market justice increased
+following the 2019 protests. However, the analysis did not detect any
+significant interactions between distributive beliefs and support for
+market justice, suggesting that the social mobilizations did not
+substantially modify the underlying associations between these
+variables. As a result, hypotheses \(H4_{a}\) and \(H4_{b}\) are not
+supported by the data. This may point to a certain stability---or even
+resilience---in the normative frameworks that guide individuals'
+evaluations of distributive justice, despite the occurrence of major
+collective political events. Alternatively, while the protests may have
+eroded trust in institutional arrangements or governance, they may not
+have fundamentally altered individuals' beliefs about how rewards and
+resources should be allocated in society.
 
@@
 
-\bibitem[\citeproctext]{ref-campbell_policy_2012}
-Campbell, A. L. (2012). Policy {Makes Mass Politics}. \emph{Annual
-Review of Political Science}, \emph{15}(Volume 15, 2012), 333--351.
-\url{https://doi.org/10.1146/annurev-polisci-012610-135202}
-
 \bibitem[\citeproctext]{ref-canalesceron_sujeto_2021}
@@
 
-\bibitem[\citeproctext]{ref-knell_perceptions_2020}
-Knell, M., \& Stix, H. (2020). Perceptions of inequality. \emph{European
-Journal of Political Economy}, \emph{65}, 101927.
-\url{https://doi.org/10.1016/j.ejpoleco.2020.101927}
-
 \bibitem[\citeproctext]{ref-koos_moral_2019}
@@
 
-\bibitem[\citeproctext]{ref-reynolds_perceptions_2014}
-Reynolds, J., \& Xian, H. (2014). Perceptions of meritocracy in the land
-of opportunity. \emph{Research in Social Stratification and Mobility},
-\emph{36}, 121--137. \url{https://doi.org/10.1016/j.rssm.2014.03.001}
-
 \bibitem[\citeproctext]{ref-sandel_tyranny_2020}

### 2025-08-05 resubmission/02-antecedents.qmd
review antecedentes jc (falta merit)

@@
 For instance, perceptions that effort is rewarded in society are more strongly associated with positive evaluations of fairness and acceptance of unequal outcomes [@batruch_belief_2023].
-This may be because effort is seen as a controllable and morally virtuous trait, whereas talent is often perceived as a natural advantage.
-Consequently, effort-based meritocracy is likely more potent in legitimizing inequality, particularly in neoliberal contexts.
 
@@
 Second, higher perceived meritocracy is expected to be positively associated with market justice preferences.
-Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit [@castillo_socialization_2024].
+Individuals who believe that effort and talent primarily determine success would be more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit [@castillo_socialization_2024].
 This aligns with the idea that meritocratic perceptions serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit.
@@
 We are especially interested in whether changes in individuals’ perceptions of inequality and meritocracy are associated with shifts in their market justice attitudes across different time points.
-The primary focus of this study is thus to scrutinize how individual-level changes in perceptions relate to changes in preferences over time.
 While we acknowledge the relevance of contextual events — such as the 2019 political outburst or the subsequent constitutional process — for public opinion and political orientation [@disipavlic_justification_2025], identifying causal effects of such events (e.g., mass protest exposure) on individual attitudes lies outside the scope of this research.

### 2025-08-05 resubmission/paper-resubmit.tex
Up render html y pdf

@@
 of unequal outcomes (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
-2023}). This may be because effort is seen as a controllable and morally
-virtuous trait, whereas talent is often perceived as a natural
-advantage. Consequently, effort-based meritocracy is likely more potent
-in legitimizing inequality, particularly in neoliberal contexts.
+2023}).
 
@@
 associated with market justice preferences. Individuals who believe that
-effort and talent primarily determine success are more likely to justify
-unequal outcomes and endorse the notion that markets allocate resources
-fairly according to individual merit
+effort and talent primarily determine success would be more likely to
+justify unequal outcomes and endorse the notion that markets allocate
+resources fairly according to individual merit
 (\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
@@
 with shifts in their market justice attitudes across different time
-points. The primary focus of this study is thus to scrutinize how
-individual-level changes in perceptions relate to changes in preferences
-over time. While we acknowledge the relevance of contextual events ---
-such as the 2019 political outburst or the subsequent constitutional
-process --- for public opinion and political orientation
+points. While we acknowledge the relevance of contextual events --- such
+as the 2019 political outburst or the subsequent constitutional process
+--- for public opinion and political orientation
 (\citeproc{ref-disipavlic_justification_2025}{Disi Pavlic et al.,

### 2025-08-06 resubmission/02-antecedents.qmd
Up antecedentes meritocracia + render
@juancarloscastillo todo tuyo

@@
 
-Meritocracy constitutes a central ideological framework for legitimizing different types of social inequality, for instance through market justice beliefs.
-Rooted in the belief that rewards and positions should be allocated based on individual effort and talent, meritocracy operates as a normative ideal and a descriptive belief about how society functions.
-As initially conceptualized by Michael Young [-@young_rise_1962], the term was meant to critique a system in which merit-based stratification becomes a new form of inequality.
-However, over time, meritocracy has been widely supported in many societies as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented economies [@mijs_paradox_2021; @sandel_tyranny_2020].
-From a sociological perspective, the belief in meritocracy is more than a cognitive assessment; it reflects a moral lens through which individuals interpret inequality.
-People who believe that success results from hard work and talent are more likely to view social and economic disparities as legitimate [@castillo_percepcion_2019; @batruch_belief_2023].
-Conversely, if they see outcomes as driven by luck, social origin, or systemic barriers, inequality is more likely to be perceived as unjust.
-This distinction becomes crucial in societies with persistent structural inequality, where public narratives often emphasize personal responsibility and merit while overlooking entrenched disadvantages.
+Meritocracy constitutes a central ideological framework for legitimizing various forms of social inequality [@mijs_unfulfillable_2016; @wilson_role_2003].
+Rooted in the belief that rewards and social positions should be allocated based on individual effort and talent, meritocracy operates both as a descriptive account of how societies function and as a normative ideal of how they ought to function.
+As originally conceptualized by Michael Young [-@young_rise_1962], the term was intended as a critique of a system in which merit-based stratification gives rise to a new form of inequality.
+However, over time, meritocracy has come to be widely endorsed as a fair and desirable principle of distribution, particularly within liberal democracies and market-oriented societies [@mijs_paradox_2021; @sandel_tyranny_2020].
 
-We adopt a multidimensional perspective on meritocracy, distinguishing between two key dimensions: effort-based and talent-based perceptions.
-This distinction is essential, as it captures different pathways through which individuals justify inequality.
-Effort-based meritocracy emphasizes hard work and perseverance as the basis for social rewards, aligning closely with cultural narratives of personal responsibility.
-A talent-based meritocracy, by contrast, emphasizes intelligence and innate abilities, which are often perceived as less malleable and more unequally distributed.
-Both dimensions have been shown to correlate with acceptance of inequality, but they may carry distinct implications for how inequality is justified in specific domains [@castillo_multidimensional_2023].
-The relevance of this distinction is supported by recent studies, which show that individuals respond differently to these dimensions.
-For instance, perceptions that effort is rewarded in society are more strongly associated with positive evaluations of fairness and acceptance of unequal outcomes [@batruch_belief_2023].
+From a sociological perspective, belief in meritocracy extends beyond a cognitive judgment; it constitutes a moral lens through which individuals interpret social and economic disparities [@mijs_paradox_2021].
+Individuals who believe that success results from hard work and talent are more likely to regard inequality as legitimate [@tejero-peregrina_perceived_2025; @castillo_percepcion_2019; @batruch_belief_2023].
+In contrast, when outcomes are attributed to luck, social origin, or structural barriers, inequality is more often perceived as unjust [@reynolds_perceptions_2014; @hauser_misperceptions_2017].
+This distinction becomes especially salient in societies marked by persistent structural inequality, where dominant narratives emphasize personal responsibility while neglecting the role of entrenched disadvantage.
 
-These dimensions of meritocracy reflect how respondents perceive society's distributive logic, regardless of whether they endorse meritocratic principles.
-This distinction aligns with recent findings indicating that individuals distinguish between how merit is perceived in society and how it should ideally operate, which in turn shapes their preferences for redistribution and justice [@tejero-peregrina_perceived_2025].
-Meritocratic beliefs serve as symbolic justifications for unequal outcomes, particularly when access is stratified by income or social opportunity.
-Prior studies have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes [@castillo_percepcion_2019; @castillo_socialization_2024].
+Recent research has highlighted the importance of disaggregating the concept of "meritocratic beliefs," particularly by distinguishing between meritocratic preferences and meritocratic perceptions [@castillo_multidimensional_2023].
+While preferences refer to normative ideals about how rewards should be allocated (i.e., based on merit), perceptions refer to subjective evaluations of how meritocracy actually operates in practice (i.e., how rewards are allocated) [@janmaat_subjective_2013].
+This distinction allows for a more precise understanding of how individuals make sense of inequality: people may endorse meritocracy normatively while simultaneously recognizing its failure in practice.
+In stratified societies, the gap between preferences and perceptions is particularly evident, as individuals may continue to uphold the ideal of meritocracy despite growing awareness of unequal opportunities [@mijs_belief_2022].
 
-In addition to influencing individual attitudes toward inequality, meritocratic perceptions can contribute to social division and the stigmatization of disadvantaged groups.
-Recent research has demonstrated that exposure to meritocratic narratives can reinforce the belief that poverty results from individual failings rather than systemic conditions, reducing support for redistributive measures and increasing the stigmatization of the poor [@hoyt_mindsets_2023].
-This reinforces negative stereotypes and reduces empathy toward individuals from lower socioeconomic backgrounds.
-Moreover, @busemeyer_positive_2021 argues that meritocratic narratives can serve as feedback mechanisms that shape public opinion and well-being by framing individuals' understanding of welfare outcomes as deserved or undeserved within existing institutional structures.
-This psychological mechanism highlights the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal perceptions of success and failure.
+Within the domain of meritocratic perceptions [@castillo_multidimensional_2023], we further distinguish between two core dimensions: effort-based and talent-based perceptions.
+This distinction is analytically important, as it captures different mechanisms through which individuals may justify inequality [@young_rise_1962; @reynolds_perceptions_2014].
+Effort-based perceptions emphasize hard work and perseverance as the key drivers of success, aligning closely with cultural narratives of individual responsibility and moral deservingness [@dubet_repensar_2011].
+In contrast, talent-based perceptions emphasize innate ability and intelligence—traits often regarded as less malleable and more unequally distributed [@young_rise_1962].
+We argue that this latter dimension may shape attitudes differently, as talent tends to be seen as a given attribute rather than the result of personal effort.
+
+Meritocratic perceptions function as a normative framework that legitimizes unequal outcomes, particularly when access to social goods is stratified by income or socioeconomic background.
+Prior studies have consistently shown that individuals who perceive society as meritocratic express lower support for redistribution [@tejero-peregrina_perceived_2025; @hoyt_mindsets_2023], greater legitimization of class inequality [@batruch_belief_2023], increased tolerance for inequality [@day_movin_2017], and lower perceptions of income inequality itself [@castillo_percepcion_2019].
+In the regard of market justice, prior studies in Chile have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes in healthcare, education and pensions [@castillo_socialization_2024].
+Taken together, and as @busemeyer_positive_2021 argues, meritocratic narratives can operate as feedback mechanisms that shape public opinion and individual well-being by framing welfare outcomes as either deserved or undeserved.
+This psychological mechanism underscores the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal understandings of success and failure.
 

### 2025-08-06 resubmission/paper-blinded-resubmit.tex
Up antecedentes meritocracia + render
@juancarloscastillo todo tuyo

--- 2025-08-05 resubmission/paper-resubmit.tex
@@
 \begin{abstract}
-Several countries have experienced a shift toward the privatization and
-commodification of public goods, welfare policies, and social services.
-In Latin America, Chile stands out as a paradigmatic case where this
-trend has led to the extensive marketization of essential services. From
-a moral economy perspective, the extent to which individuals consider it
-fair for access to such services to depend on market criteria has been
-conceptualized as market justice preferences. This study investigates
-the relationship between perceptions of economic inequality,
-meritocratic beliefs, and market justice preferences in Chile between
-2016 and 2023. Using six waves of panel data from the Chilean
-Longitudinal Social Survey -- ELSOC (\(N_{observations}\) = 8,643;
-\(N_{individuals}\) = 1,687), the analysis examines how subjective
-assessments of inequality shape attitudes toward the role of merit in
-access to key social services such as healthcare, education, and
-pensions. The findings show that greater perceived inequality is
-associated with lower market justice preferences. However, individuals
-who believe that effort is rewarded are more likely to legitimize
-existing disparities. In contrast, the perception that talent is
-rewarded shows a negative effect on market justice preferences; an
-effect that intensifies as perceived inequality increases over time. The
-study also considers the influence of major social movements during this
-period, which appear to have reshaped public discourse on justice and
-fairness. These findings contribute to a deeper understanding of how
-distributive beliefs evolve in contexts marked by persistent inequality
-and entrenched neoliberal frameworks \newline \textbf{Keywords}:
-Inequality perception, meritocracy, commodification, Chile, public
-preferences
+This study investigates the relationship between perceptions of economic
+inequality, meritocratic perceptions, and preferences for market justice
+in Chile between 2016 and 2023. Using six waves of panel data from the
+Chilean Longitudinal Social Survey - ELSOC (\(N_{observations}\) =
+8,643; \(N_{individuals}\) = 1,687), the analysis examines how
+subjective assessments of inequality shape attitudes toward the role of
+merit in access to key social services such as healthcare, education,
+and pensions. Results indicate that rising perceptions of inequality are
+associated with lower support for market justice; however, individuals
+with strong meritocratic convictions are more likely to legitimize
+existing disparities. The study also considers the influence of major
+social movements during this period, which appear to have reshaped
+public discourse and perceptions of fairness. These findings contribute
+to a deeper understanding of how beliefs about justice and equity evolve
+in contexts marked by persistent inequality and entrenched
+market-oriented frameworks. \newline \textbf{Keywords}: Economic
+inequality, meritocracy, market justice, Chile, public preferences,
+inequality perception
 \end{abstract}
@@
 et al., 2020}; \citeproc{ref-gimpelson_misperceiving_2018}{Gimpelson \&
-Treisman, 2018}; \citeproc{ref-mijs_stratified_2016}{Mijs, 2016}).
+Treisman, 2018}; \citeproc{ref-mijs_stratified_2016}{Mijs, 2016a}).
 Scholars have proposed multiple dimensions of perceived inequality,
@@
 Meritocracy constitutes a central ideological framework for legitimizing
-different types of social inequality, for instance through market
-justice beliefs. Rooted in the belief that rewards and positions should
-be allocated based on individual effort and talent, meritocracy operates
-as a normative ideal and a descriptive belief about how society
-functions. As initially conceptualized by Michael Young
-(\citeproc{ref-young_rise_1962}{1962}), the term was meant to critique a
-system in which merit-based stratification becomes a new form of
-inequality. However, over time, meritocracy has been widely supported in
-many societies as a fair and desirable principle of distribution,
-particularly within liberal democracies and market-oriented economies
+various forms of social inequality
+(\citeproc{ref-mijs_unfulfillable_2016}{Mijs, 2016b};
+\citeproc{ref-wilson_role_2003}{Wilson, 2003}). Rooted in the belief
+that rewards and social positions should be allocated based on
+individual effort and talent, meritocracy operates both as a descriptive
+account of how societies function and as a normative ideal of how they
+ought to function. As originally conceptualized by Michael Young
+(\citeproc{ref-young_rise_1962}{1962}), the term was intended as a
+critique of a system in which merit-based stratification gives rise to a
+new form of inequality. However, over time, meritocracy has come to be
+widely endorsed as a fair and desirable principle of distribution,
+particularly within liberal democracies and market-oriented societies
 (\citeproc{ref-mijs_paradox_2021}{Mijs, 2021};
-\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}). From a sociological
-perspective, the belief in meritocracy is more than a cognitive
-assessment; it reflects a moral lens through which individuals interpret
-inequality. People who believe that success results from hard work and
-talent are more likely to view social and economic disparities as
-legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023};
-\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}).
-Conversely, if they see outcomes as driven by luck, social origin, or
-systemic barriers, inequality is more likely to be perceived as unjust.
-This distinction becomes crucial in societies with persistent structural
-inequality, where public narratives often emphasize personal
-responsibility and merit while overlooking entrenched disadvantages.
-
-We adopt a multidimensional perspective on meritocracy, distinguishing
-between two key dimensions: effort-based and talent-based perceptions.
-This distinction is essential, as it captures different pathways through
-which individuals justify inequality. Effort-based meritocracy
-emphasizes hard work and perseverance as the basis for social rewards,
-aligning closely with cultural narratives of personal responsibility. A
-talent-based meritocracy, by contrast, emphasizes intelligence and
-innate abilities, which are often perceived as less malleable and more
-unequally distributed. Both dimensions have been shown to correlate with
-acceptance of inequality, but they may carry distinct implications for
-how inequality is justified in specific domains
+\citeproc{ref-sandel_tyranny_2020}{Sandel, 2020}).
+
+From a sociological perspective, belief in meritocracy extends beyond a
+cognitive judgment; it constitutes a moral lens through which
+individuals interpret social and economic disparities
+(\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). Individuals who believe
+that success results from hard work and talent are more likely to regard
+inequality as legitimate (\citeproc{ref-batruch_belief_2023}{Batruch et
+al., 2023}; \citeproc{ref-castillo_percepcion_2019}{Castillo et al.,
+2012}; \citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina
+et al., 2025}). In contrast, when outcomes are attributed to luck,
+social origin, or structural barriers, inequality is more often
+perceived as unjust (\citeproc{ref-hauser_misperceptions_2017}{Hauser \&
+Norton, 2017}; \citeproc{ref-reynolds_perceptions_2014}{Reynolds \&
+Xian, 2014}). This distinction becomes especially salient in societies
+marked by persistent structural inequality, where dominant narratives
+emphasize personal responsibility while neglecting the role of
+entrenched disadvantage.
+
+Recent research has highlighted the importance of disaggregating the
+concept of ``meritocratic beliefs,'' particularly by distinguishing
+between meritocratic preferences and meritocratic perceptions
 (\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}).
-The relevance of this distinction is supported by recent studies, which
-show that individuals respond differently to these dimensions. For
-instance, perceptions that effort is rewarded in society are more
-strongly associated with positive evaluations of fairness and acceptance
-of unequal outcomes (\citeproc{ref-batruch_belief_2023}{Batruch et al.,
-2023}).
-
-These dimensions of meritocracy reflect how respondents perceive
-society's distributive logic, regardless of whether they endorse
-meritocratic principles. This distinction aligns with recent findings
-indicating that individuals distinguish between how merit is perceived
-in society and how it should ideally operate, which in turn shapes their
-preferences for redistribution and justice
-(\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
-2025}). Meritocratic beliefs serve as symbolic justifications for
-unequal outcomes, particularly when access is stratified by income or
-social opportunity. Prior studies have shown that individuals who
-perceive higher levels of meritocracy tend to express stronger support
-for unequal distributions that reflect market outcomes
-(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012};
-\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
-
-In addition to influencing individual attitudes toward inequality,
-meritocratic perceptions can contribute to social division and the
-stigmatization of disadvantaged groups. Recent research has demonstrated
-that exposure to meritocratic narratives can reinforce the belief that
-poverty results from individual failings rather than systemic
-conditions, reducing support for redistributive measures and increasing
-the stigmatization of the poor (\citeproc{ref-hoyt_mindsets_2023}{Hoyt
-et al., 2023}). This reinforces negative stereotypes and reduces empathy
-toward individuals from lower socioeconomic backgrounds. Moreover,
-Busemeyer et al. (\citeproc{ref-busemeyer_positive_2021}{2021}) argues
-that meritocratic narratives can serve as feedback mechanisms that shape
-public opinion and well-being by framing individuals' understanding of
-welfare outcomes as deserved or undeserved within existing institutional
-structures. This psychological mechanism highlights the normative power
-of meritocracy in stabilizing unequal systems by shaping political
-attitudes and personal perceptions of success and failure.
+While preferences refer to normative ideals about how rewards should be
+allocated (i.e., based on merit), perceptions refer to subjective
+evaluations of how meritocracy actually operates in practice (i.e., how
+rewards are allocated) (\citeproc{ref-janmaat_subjective_2013}{Janmaat,
+2013}). This distinction allows for a more precise understanding of how
+individuals make sense of inequality: people may endorse meritocracy
+normatively while simultaneously recognizing its failure in practice. In
+stratified societies, the gap between preferences and perceptions is
+particularly evident, as individuals may continue to uphold the ideal of
+meritocracy despite growing awareness of unequal opportunities
+(\citeproc{ref-mijs_belief_2022}{Mijs et al., 2022}).
+
+Within the domain of meritocratic perceptions
+(\citeproc{ref-castillo_multidimensional_2023}{Castillo et al., 2023}),
+we further distinguish between two core dimensions: effort-based and
+talent-based perceptions. This distinction is analytically important, as
+it captures different mechanisms through which individuals may justify
+inequality (\citeproc{ref-reynolds_perceptions_2014}{Reynolds \& Xian,
+2014}; \citeproc{ref-young_rise_1962}{Young, 1962}). Effort-based
+perceptions emphasize hard work and perseverance as the key drivers of
+success, aligning closely with cultural narratives of individual
+responsibility and moral deservingness
+(\citeproc{ref-dubet_repensar_2011}{Dubet, 2011}). In contrast,
+talent-based perceptions emphasize innate ability and
+intelligence---traits often regarded as less malleable and more
+unequally distributed (\citeproc{ref-young_rise_1962}{Young, 1962}). We
+argue that this latter dimension may shape attitudes differently, as
+talent tends to be seen as a given attribute rather than the result of
+personal effort.
+
+Meritocratic perceptions function as a normative framework that
+legitimizes unequal outcomes, particularly when access to social goods
+is stratified by income or socioeconomic background. Prior studies have
+consistently shown that individuals who perceive society as meritocratic
+express lower support for redistribution
+(\citeproc{ref-hoyt_mindsets_2023}{Hoyt et al., 2023};
+\citeproc{ref-tejero-peregrina_perceived_2025}{Tejero-Peregrina et al.,
+2025}), greater legitimization of class inequality
+(\citeproc{ref-batruch_belief_2023}{Batruch et al., 2023}), increased
+tolerance for inequality (\citeproc{ref-day_movin_2017}{Day \& Fiske,
+2017}), and lower perceptions of income inequality itself
+(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}). In the
+regard of market justice, prior studies in Chile have shown that
+individuals who perceive higher levels of meritocracy tend to express
+stronger support for unequal distributions that reflect market outcomes
+in healthcare, education and pensions
+(\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
+Taken together, and as Busemeyer et al.
+(\citeproc{ref-busemeyer_positive_2021}{2021}) argues, meritocratic
+narratives can operate as feedback mechanisms that shape public opinion
+and individual well-being by framing welfare outcomes as either deserved
+or undeserved. This psychological mechanism underscores the normative
+power of meritocracy in stabilizing unequal systems by shaping political
+attitudes and personal understandings of success and failure.
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper-resubmit_files/figure-pdf/fig-alluvial-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{paper-blinded-resubmit_files/figure-pdf/fig-alluvial-1.pdf}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper-resubmit_files/figure-pdf/fig-meanchange-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{paper-blinded-resubmit_files/figure-pdf/fig-meanchange-1.pdf}
 
@@
 
-\includegraphics[width=1\textwidth,height=\textheight]{paper-resubmit_files/figure-pdf/fig-interact-1.pdf}
+\includegraphics[width=1\textwidth,height=\textheight]{paper-blinded-resubmit_files/figure-pdf/fig-interact-1.pdf}
 
@@
 
+\bibitem[\citeproctext]{ref-day_movin_2017}
+Day, M. V., \& Fiske, S. T. (2017). Movin' on {Up}? {How Perceptions} of
+{Social Mobility Affect Our Willingness} to {Defend} the {System}.
+\emph{Social Psychological and Personality Science}, \emph{8}(3),
+267--274. \url{https://doi.org/10.1177/1948550616678454}
+
 \bibitem[\citeproctext]{ref-disipavlic_justification_2025}
@@
 \emph{Social Forces}, soaf102. \url{https://doi.org/10.1093/sf/soaf102}
+
+\bibitem[\citeproctext]{ref-dubet_repensar_2011}
+Dubet, F. (2011). \emph{{Repensar la justicia social}} (Sexta
+Edici{ó}n). Siglo XXI.
 
@@
 
+\bibitem[\citeproctext]{ref-hauser_misperceptions_2017}
+Hauser, O. P., \& Norton, M. I. (2017). ({Mis})perceptions of
+inequality. \emph{Current Opinion in Psychology}, \emph{18}, 21--25.
+\url{https://doi.org/10.1016/j.copsyc.2017.07.024}
+
 \bibitem[\citeproctext]{ref-hoxMultilevelAnalysisTechniques2017a}
@@
 \bibitem[\citeproctext]{ref-mijs_stratified_2016}
-Mijs, J. (2016). Stratified {Failure}: {Educational Stratification} and
+Mijs, J. (2016a). Stratified {Failure}: {Educational Stratification} and
 {Students}' {Attributions} of {Their Mathematics Performance} in 24
@@
 \url{https://doi.org/10.1177/0038040716636434}
+
+\bibitem[\citeproctext]{ref-mijs_unfulfillable_2016}
+Mijs, J. (2016b). The {Unfulfillable Promise} of {Meritocracy}: {Three
+Lessons} and {Their Implications} for {Justice} in {Education}.
+\emph{Social Justice Research}, \emph{29}(1), 14--34.
+\url{https://doi.org/10.1007/s11211-014-0228-0}
 
@@
 \emph{19}(1), 7--35. \url{https://doi.org/10.1093/ser/mwy051}
+
+\bibitem[\citeproctext]{ref-mijs_belief_2022}
+Mijs, J., Daenekindt, S., de Koster, W., \& van der Waal, J. (2022).
+Belief in {Meritocracy Reexamined}: {Scrutinizing} the {Role} of
+{Subjective Social Mobility}. \emph{Social Psychology Quarterly},
+\emph{85}(2), 131--141. \url{https://doi.org/10.1177/01902725211063818}
 
@@
 \url{https://doi.org/10.2307/2950710}
+
+\bibitem[\citeproctext]{ref-reynolds_perceptions_2014}
+Reynolds, J., \& Xian, H. (2014). Perceptions of meritocracy in the land
+of opportunity. \emph{Research in Social Stratification and Mobility},
+\emph{36}, 121--137. \url{https://doi.org/10.1016/j.rssm.2014.03.001}
 

### 2025-08-08 resubmission/02-antecedents.qmd
jc review

@@
 
-Recent research has highlighted the importance of disaggregating the concept of "meritocratic beliefs," particularly by distinguishing between meritocratic preferences and meritocratic perceptions [@castillo_multidimensional_2023].
+Recent research has highlighted the importance of disentangling the concept of "meritocratic beliefs," particularly by distinguishing between meritocratic preferences and meritocratic perceptions [@castillo_multidimensional_2023].
 While preferences refer to normative ideals about how rewards should be allocated (i.e., based on merit), perceptions refer to subjective evaluations of how meritocracy actually operates in practice (i.e., how rewards are allocated) [@janmaat_subjective_2013].
@@
 Prior studies have consistently shown that individuals who perceive society as meritocratic express lower support for redistribution [@tejero-peregrina_perceived_2025; @hoyt_mindsets_2023], greater legitimization of class inequality [@batruch_belief_2023], increased tolerance for inequality [@day_movin_2017], and lower perceptions of income inequality itself [@castillo_percepcion_2019].
-In the regard of market justice, prior studies in Chile have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes in healthcare, education and pensions [@castillo_socialization_2024].
+Regarding market justice preferences, prior studies in Chile have shown that individuals who perceive higher levels of meritocracy tend to express stronger support for unequal distributions that reflect market outcomes in healthcare, education and pensions [@castillo_socialization_2024].
 Taken together, and as @busemeyer_positive_2021 argues, meritocratic narratives can operate as feedback mechanisms that shape public opinion and individual well-being by framing welfare outcomes as either deserved or undeserved.
-This psychological mechanism underscores the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal understandings of success and failure.
+Such mechanism underscores the normative power of meritocracy in stabilizing unequal systems by shaping political attitudes and personal understandings of success and failure.
 
@@
 
-Concerning public opinion, the ELSOC survey shows that Chile has a low level of agreement with the idea that it is fair for those with higher incomes to buy better welfare in the areas of education, healthcare, and pensions, with average support of 8.5% between 2016 and 2023.
-At the same time, around 55% and 47% of the public between 2016 and 2023 *disagreed* that effort and talent are rewarded accordingly.
-This coincides with 90% of citizens *agreeing* that income differences in Chile are *too large* [@castillo_changes_2025].
-
-In sum, Chile represents an illustrative case of a particular form of *moral economy* [@koos_moral_2019], in which national socioeconomic characteristics, institutional arrangements, and public attitudes coexist.
-This suggests that the country combines persistently high levels of economic and social inequality with a strong institutionalization of market-based mechanisms within the welfare system [@madariaga_three_2020].
-At the same time, perceptions of inequality are high, and there is widespread public criticism of the commodification of social services [@somma_no_2021].
-For the above reasons, we argue that the Chilean case represents a relevant context to study the role of perceived income inequality and meritocracy in shaping public preferences regarding the commodification of welfare services, particularly from the perspective of individual change.
-
 ## This study
@@
 Specifically, we argue that both perceptions independently and interactively influence the extent to which individuals endorse market-based distributions of social goods and services in Chile.
+To our knowledge, there are not studies so fay analyzing such associations, let alone from a longitudinal perspective.
 
-First, consistent with previous findings about redistributive preferences [@castillo_perception_2022], we expect that a higher perception of economic inequality will be associated with lower market justice preferences.
+Firstly, and consistent with previous findings about research on redistributive preferences [@castillo_perception_2022], we expect that a higher perception of economic inequality will be associated with lower market justice preferences.
 When individuals perceive smaller income gaps, they are more likely to view market mechanisms as fair and legitimate, reinforcing the acceptance of outcomes based on competition and ability to pay.
@@
 We are especially interested in whether changes in individuals’ perceptions of inequality and meritocracy are associated with shifts in their market justice attitudes across different time points.
-While we acknowledge the relevance of contextual events — such as the 2019 political outburst or the subsequent constitutional process — for public opinion and political orientation [@disipavlic_justification_2025], identifying causal effects of such events (e.g., mass protest exposure) on individual attitudes lies outside the scope of this research.
-Instead, we focus on how intra-individual trajectories in perceived inequality and merit shape evolving attitudes toward the commodification of social services.
+While we acknowledge the relevance of contextual events for public opinion and political orientation [@disipavlic_justification_2025], such as the 2019 political outburst or the subsequent constitutional process in CHile, we are aware that identifying causal associations lies outside the scope of this research.
+Instead, and from an exploratory perspective, we focus on how intra-individual trajectories in perceived inequality and merit shape evolving attitudes toward the commodification of social services.
 

### 2025-08-08 resubmission/05-discussion.qmd
jc review

@@
 
-This study tests several key hypotheses concerning the relationships between perceptions of economic inequality, meritocratic perceptions, and preferences toward market justice in Chile.
-It posits that higher perceived inequality and stronger meritocratic perceptions both independently contribute to greater support for market-based distributions of social goods.
-Furthermore, the study examines how perceived economic inequality moderates the influence of meritocratic perceptions on market justice preferences.
-Additionally, it considers how major social and political mobilizations, such as those occurring around 2019 and subsequent years, may have weakened these relationships, reflecting increased critical perspectives on systemic inequalities and market mechanisms.
-The subsequent discussion interprets the empirical findings in light of these hypotheses, considering their implications for understanding attitudes toward inequality and social justice in a context marked by profound social change.
-
-The first set of hypothesis suggested that perceptions of economic inequality were strongly linked to preferences for market justice.
+The first set of hypotheses suggested that perceptions of economic inequality were strongly linked to preferences for market justice.
 The findings demonstrated that individuals perceiving higher income disparities showed less support for the idea that the access to core social services depends on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021; @garcia-sanchez_vicious_2019].
-Such a relationship reflects a moral economy in which perceptions of systemic unfairness undermine the legitimacy of existing distributions and fuel demands for greater equity.
-In particular, the negative within-individual effect observed over time ($H1_{b}$) suggests that as inequality becomes more salient for some individuals, their support for market justice declines—potentially due to rising distrust in market mechanisms or growing disillusionment with the fairness of the system.
+In particular, the negative within-individual effect observed over time ($H1_{b}$) suggests that as inequality becomes more salient for some individuals, their support for market justice declines, potentially due to rising distrust in market mechanisms or growing disillusionment with the fairness of the system.
 However, although the effect of perceived inequality at the between-individual level is also negative, it is not statistically significant.
@@
 These perceptions operate symbolically to reinforce structural inequalities by reducing support for redistributive policies, framing inequality as both fair and deserved.
-Consistent with prior work by @castillo_socialization_2024 on Chilean students, the findings suggest that such meritocratic perceptions uphold existing hierarchies by promoting the acceptance of inequality as a reflection of individual virtue rather than structural failure.
+Consistent with prior work by @castillo_socialization_2024 on Chilean students, the results suggest that such meritocratic perceptions uphold existing hierarchies by promoting the acceptance of inequality as a reflection of individual virtue rather than structural failure.
 This mechanism is particularly salient in neoliberal contexts like Chile, where market logics heavily shape social attitudes [@canalesceron_sujeto_2021].
@@
 While increases in the belief that rewards are based on individual effort are associated with stronger preferences for market justice, increases in the belief that rewards derive from talent are linked to weaker support for such principles.
-One possible explanation is that effort is generally viewed as a controllable and malleable trait, aligning with the normative foundations of market fairness.
-In contrast, talent tends to be perceived as innate and less subject to personal control, rendering talent-based inequality less legitimate.
+One possible explanation is that effort is generally viewed as a controllable and malleable trait, whereas talent tends to be perceived as innate and less subject to personal control, rendering talent-based inequality less legitimate.
 Furthermore, increased exposure to real-world scenarios in which outcomes are clearly shaped by inherent traits rather than hard work may lead individuals to question the fairness and legitimacy of market-based reward systems.
@@
 The third set of hypotheses addressed the moderating role of perceptions of inequality in the relationship between meritocratic perceptions and preferences for market justice.
-The analysis revealed that the positive association between meritocratic perceptions and support for market-based distribution systems tends to be stronger when perceived economic inequality is high—contradicting our initial expectations ($H3_{a}$ and $H3_{b}$).
+The analysis revealed that the positive association between meritocratic perceptions and support for market-based distribution systems tends to be stronger when perceived economic inequality is high, contradicting our initial expectations ($H3_{a}$ and $H3_{b}$).
 Although this moderating effect is not statistically significant for effort-related meritocratic perceptions, it is significant for talent-related perceptions at both the within- and between-individual levels.
-This finding runs counter to our hypotheses, as it suggests that the effect of meritocratic perceptions—specifically, talent-based meritocracy—intensifies as perceptions of inequality increase.
-However, this interpretation requires caution.
-The main effect of perceived inequality on support for market justice is negative.
-Thus, as perceived inequality becomes less negative (i.e., closer to zero), the positive relationship between talent-based meritocratic perceptions and market justice preferences becomes stronger.
+However, this interpretation requires caution: the main effect of perceived inequality on support for market justice is negative, meaning that as perceived inequality becomes less negative (i.e., closer to zero), the positive relationship between talent-based meritocratic perceptions and market justice preferences becomes stronger.
 This suggests a more nuanced dynamic: meritocratic perceptions, particularly those emphasizing talent, may serve as a stronger justificatory mechanism for market-based inequalities among individuals who are less inclined to perceive inequality as a problem.
@@
 
-Finally, the fourth hypothesis examined the potential impact of recent social mobilizations—particularly those that erupted in 2019—on the relationship between meritocratic perceptions, perceptions of inequality, and preferences for market justice.
-The main effect of time indicates that, on average, support for market justice increased following the 2019 protests.
-However, the analysis did not detect any significant interactions between distributive beliefs and support for market justice, suggesting that the social mobilizations did not substantially modify the underlying associations between these variables.
-As a result, hypotheses $H4_{a}$ and $H4_{b}$ are not supported by the data.
-This may point to a certain stability—or even resilience—in the normative frameworks that guide individuals’ evaluations of distributive justice, despite the occurrence of major collective political events.
+Finally, the fourth exploratory hypothesis examined the potential impact of social events — particularly those that erupted in 2019 — on the relationship between meritocratic perceptions, perceptions of inequality, and preferences for market justice.
+The main effect of time indicates that, on average, support for market justice increased after 2019 protests.
+However, the analysis did not detect any significant interactions between distributive beliefs and support for market justice, suggesting that events in this period (as the mobilizations, the pandemic and/or the constitucional processes) did not substantially modify the underlying associations between these variables.
+As a result, hypotheses $H4_{a}$ and $H4_{b}$ are not supported by the analysis.
+This may point to a certain stability or even resilience in the normative frameworks that guide individuals’ evaluations of distributive justice, despite the occurrence of major collective political events.
 Alternatively, while the protests may have eroded trust in institutional arrangements or governance, they may not have fundamentally altered individuals’ beliefs about how rewards and resources should be allocated in society.

### 2025-08-08 resubmission/06-conclusion.qmd
jc review

@@
 This study examined the complex interplay between perceptions of economic inequality, meritocratic perceptions, and preferences toward market justice in Chile from 2016 to 2023, drawing on longitudinal data from the ELSOC survey.
-By exploring how subjective assessments and social contexts influence support for redistribution and market-based resource allocation, the research offers different elements that contribute to the understanding of the normative foundations underpinning social justice attitudes in a highly unequal and commoditized environment.
+By exploring how subjective assessments and social contexts influence support for redistribution and market-based resource allocation, the research offers different elements that contribute to the understanding of the normative foundations underpinning social justice attitudes in a highly unequal and commodified environment.
 
-The findings confirm that higher perceptions of economic inequality are associated with less support for market justice preferences, this is, the belief that it is fair that those with higher income have better social services such as education, pensions and health.
-At the same time, meritocratic perceptions—particularly those emphasizing individual effort—are strongly associated with support for market-based distributions, suggesting that meritocracy serve as a moral justification for structural inequalities.
+The findings support that higher perceptions of economic inequality are associated with less support for market justice preferences, this is, the belief that it is fair that those with higher income have better social services such as education, pensions and health.
+At the same time, meritocratic perceptions — particularly those emphasizing individual effort — are strongly associated with support for market-based distributions, suggesting that meritocracy serve as a moral justification for structural inequalities.
 However, changes in meritocratic perceptions over time reveal a more nuanced picture: while increased emphasis on effort reinforces support for market justice, increased emphasis on talent tends to reduce it.
 This distinction likely reflects broader beliefs about the controllability and fairness of different meritocratic traits.
-Moreover, the interaction between inequality perceptions and meritocratic perceptions indicates that the legitimizing power of meritocracy becomes stronger as perceptions of inequality become less negative—that is, when individuals are less aware or concerned about economic disparities.
-This finding highlights a potential feedback mechanism by which lower sensitivity to inequality may enable stronger endorsement of merit-based explanations for unequal outcomes.
+Moreover, the interaction between inequality perceptions and meritocratic perceptions indicates that the legitimizing power of meritocracy becomes stronger as perceptions of inequality become less negative — that is, when individuals are less aware or concerned about economic disparities.
+Such finding highlights a potential feedback mechanism by which lower sensitivity to inequality may enable stronger endorsement of merit-based explanations for unequal outcomes.
 
@@
 While previous studies primarily focused on objective measures or individual characteristics [@busemeyer_skills_2014; @immergut_it_2020; @lindh_public_2015], this work emphasizes the dynamic and interactional nature of perceptions and beliefs over time.
-It highlights the importance of socio-political upheavals in reshaping normative attitudes, underscoring the role of collective action in challenging entrenched narratives of meritocracy and fairness.
+Furthermore, it highlights the importance of socio-political upheavals in reshaping normative attitudes, underscoring the role of collective action in challenging entrenched narratives of meritocracy and fairness.
 The longitudinal approach provides a deeper temporal perspective on how societal events could influence individual perceptions and preferences.
 
-Despite its contributions, the study has several limitations.
-First, the reliance on self-reported perceptions and attitudes may be subject to social desirability bias.
-Second, the focus on Chile, while offering valuable insights into a specific context, limits the generalizability of results to other countries with different institutional and cultural backgrounds.
-Third, the observational nature of the data prevents causal inferences, and unobserved confounding factors may influence the observed relationships.
-Finally, the potential influence of media, political discourse, and peer effects on perceptions was not directly assessed.
-
-Future research could explore the causal mechanisms linking perceptions and attitudes, possibly through experimental or mixed-method designs.
-Comparative studies across different national contexts would help determine the universality or specificity of these dynamics.
-Additionally, investigating the role of media, political communication, and education in shaping perceptions of inequality and meritocracy would deepen understanding of the normative foundations of social justice attitudes.
+Regarding avenues for future research, experimental evidence could help expand the understanding of the association between perceptual variables and market justice preferences.
+International comparative studies are also necessary in order to assess the role of national contexts and the universality or specificity of these dynamics.
+Additionally, investigating the role of media, political communication, and education in shaping perceptions of inequality and meritocracy would deepen understanding of the normative foundations of social justice attitudes and their variation over time.
 Finally, examining how these perceptions influence behavioral outcomes, such as political participation or support for social movements, would provide valuable insights into the pathways from beliefs to collective action and policy change.

### 2025-08-08 resubmission/04-analisis.qmd
Up carpeta resubmit

@@
   terms = c("merit_effort_mean", 
-            sprintf("perc_inequality_mean [%.2f, %.2f, %.2f]", min_val, mean_val, max_val))
+            sprintf("perc_inequality_mean [%.2f, %.2f]", min_val, max_val))
 ) %>% 
@@
     name = "Perception of inequality (BE)",
-    values = c("#0571B0", "black", "#B2182B"),
-    labels = c("Min", "Mean", "Max")
+    values = c("#0571B0",  "#B2182B"),
+    labels = c("Min",  "Max")
   ) +
@@
     name = "Perception of inequality (BE)",
-    values = c("#0571B0", "black", "#B2182B"),
-    labels = c("Min", "Mean", "Max")) +
+    values = c("#0571B0",  "#B2182B"),
+    labels = c("Min",  "Max")) +
   labs(
@@
   terms = c("merit_talent_mean", 
-            sprintf("perc_inequality_mean [%.2f, %.2f, %.2f]", min_val, mean_val, max_val))
+            sprintf("perc_inequality_mean [%.2f, %.2f]", min_val, max_val))
 ) %>% 
@@
 
-```{r echo=FALSE, results='asis', eval=FALSE, include=FALSE}
-#| label: tbl-interactions2
-#| tbl-cap: "Interactions for meritocracy, perceived economic inequality and market justice preferences"
-#| tbl-cap-location: top
-
-ccoef <- list(
-  "(Intercept)" = "Intercept",
-  "ola_num:perc_inequality_cwc" = "Wave x Perception inequality (WE)",
-  "ola_num:merit_effort_cwc" = "Wave x Merit: effort (WE)",
-  "ola_num:merit_talent_cwc" = "Wave x Merit: talent (WE)",
-  "ola_num:perc_inequality_mean" = "Wave x Perception inequality (BE)",
-  "ola_num:merit_effort_mean" = "Wave x Merit: effort (BE)",
-  "ola_num:merit_talent_mean" = "Wave x Merit: talent (BE)"
-    )
-
-texreg::texreg(list(m10.1, m11.1, m12.1, m10.2, m11.2, m12.2),
-               custom.model.names = c("Model 12", "Model 13", "Model 14", "Model 15", "Model 16", "Model 17"),
-               caption = NULL,
-               stars = c(0.05, 0.01, 0.001),
-               custom.coef.map = ccoef,
-               digits = 3,
-               caption.above = T,
-               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.",
-               leading.zero = T,
-               use.packages = F,
-               booktabs = F,
-               scalebox = 0.7,
-               include.loglik = FALSE,
-               include.aic = FALSE,
-               center = T,
-               custom.gof.names = c("BIC", 
-                                    "Numb. obs.", 
-                                    "Num. groups: individuals", 
-                                    "Var: individuals (Intercept)",
-
-                                    "Var: individuals, perception inequality cwc",
-                                    "Var: individuals, wave",
-                                    "Cov: individuals (Intercept), perception inequality cwc",
-                                    "Cov: individuals (Intercept), wave",
-                                    "Cov: individuals, perception inequality cwc, wave",
-                                    "Var: Residuals",
-                                    "Var: individuals, merit effort cwc", 
-                                    "Cov: individuals (Intercept), merit effort cwc",
-                                    "Cov: individuals, merit effort cwc, wave",
-
-                                    "Var: individuals, merit talent cwc", 
-                                    "Cov: individuals (Intercept), merit talent cwc",
-                                    "Cov: individuals, merit talent cwc, wave",
-
-                                    "Var: individuals, perception inequality mean", 
-                                    "Cov: individuals (Intercept), perception inequality mean",
-                                    "Cov: individuals, perception inequality mean, wave",
-
-                                     "Var: individuals, merit effort mean", 
-                                    "Cov: individuals (Intercept), merit effort mean",
-                                    "Cov: individuals, merit effort mean, wave",
-
-                                    "Var: individuals, merit talent mean", 
-                                    "Cov: individuals (Intercept), merit talent mean",
-                                    "Cov: individuals, merit talent mean, wave"
-                                    ),
-               custom.gof.rows = list("Controls"=c(rep("Yes",6))))
-
-```
-
-```{r}
-#| label: fig-growth
-#| out-width: '100%'
-#| fig-asp: 1
-#| fig-cap: "Time varying within effects of meritocratic perceptions on market justice preferences"
-#| fig-cap-location: top
-#| echo: false
-#| results: asis
-#| include: false
-#| eval: false
-
-library(ggeffects)
-
-p4 <- ggpredict(m11.2, terms = c("ola_num [all]", "merit_effort_mean")) %>% 
-  plot() + 
-  scale_x_continuous(n.breaks = 6) +
-  scale_color_manual(
-    name = "Merit:\nEffort",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  scale_fill_manual(
-    name = "Merit:\nEffort",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")) +
-  labs(x = "Wave",
-       title = "(A) People are rewarded for their efforts",
-       color = NULL) +
-  ggdist::theme_ggdist() +
-  theme(legend.position = "top")
-
-p5 <- ggpredict(m12.2, terms = c("ola_num [all]", "merit_talent_mean")) %>% 
-  plot() +
-  scale_x_continuous(n.breaks = 6) +
-  scale_color_manual(
-    name = "Merit:\nTalent",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  scale_fill_manual(
-    name = "Merit:\nTalent",
-    values = c("#D55E00", "#0072B2", "#009E73"),
-    labels = c("-1 SD", "Mean", "+1 SD")
-  ) +
-  labs(
-    x = "Wave",
-    title = "(B) People are rewarded for their intelligence",
-    caption = "Source: own elaboration with pooled data from ELSOC 2016-2023 (n = 8460)"
-  ) +
-  ggdist::theme_ggdist() +
-  theme(legend.position = "top")
-
-ggpubr::ggarrange(p4, p5, common.legend=FALSE, ncol=1)
-
-```
-
 Regarding the temporal dynamics of the key predictors, the analysis shows that time has no statistically significant effects on the intrapersonal or interpersonal effects of perceived inequality and meritocratic perceptions.

### 2025-08-08 resubmission/05-discussion.qmd
Up carpeta resubmit

@@
 
-The first set of hypotheses suggested that perceptions of economic inequality were strongly linked to preferences for market justice.
-The findings demonstrated that individuals perceiving higher income disparities showed less support for the idea that the access to core social services depends on individual income, aligning with theories emphasizing that awareness of inequality could fuel a critical view of such policies [@Castillo2012a_justice; @mijs_paradox_2021; @garcia-sanchez_vicious_2019].
-In particular, the negative within-individual effect observed over time ($H1_{b}$) suggests that as inequality becomes more salient for some individuals, their support for market justice declines, potentially due to rising distrust in market mechanisms or growing disillusionment with the fairness of the system.
-However, although the effect of perceived inequality at the between-individual level is also negative, it is not statistically significant.
-Therefore, we cannot conclude that individuals who, on average, perceive higher inequality are systematically less supportive of market justice.
-As a result, hypothesis $H1_{a}$ does not receive empirical support.
+The first set of hypotheses proposed that perceptions of economic inequality are strongly associated with preferences for market justice.
+In the between-individual specification, the effect of perceived inequality is not statistically significant.
+Therefore, we cannot conclude that individuals who, on average, perceive higher levels of inequality are systematically less supportive of market justice.
+As a result, hypothesis $H1_{a}$ is not supported by the data.
+However, in the within-individual specification, the findings reveal a significant negative association: individuals who, at a given point in time, perceive higher income disparities are less supportive of the idea that access to core social services should depend on individual income.
+This result aligns with theories suggesting that increased awareness of inequality fosters a more critical stance toward market-based distributive arrangements [@Castillo2012a_justice; @mijs_paradox_2021; @garcia-sanchez_vicious_2019].
+Such a pattern reflects a moral economy logic in which perceptions of systemic unfairness undermine the legitimacy of existing distributions and strengthen demands for greater equity.
+Specifically, the negative within-individual effect over time ($H1_{b}$) suggests that as inequality becomes more salient for some individuals, their support for market justice declines—possibly due to growing distrust in market mechanisms or increasing disillusionment with the perceived fairness of the system.
 

### 2025-08-08 resubmission/07-supplementary-material.qmd
Up carpeta resubmit

@@
 ```{r echo=FALSE, results='asis'}
+#| label: tbl-interactions2
+#| tbl-cap: "Growth curves for meritocracy, perceived economic inequality and market justice preferences"
+#| tbl-cap-location: top
+
+ccoef <- list(
+  "(Intercept)" = "Intercept",
+  "ola_num:perc_inequality_cwc" = "Wave x Perception inequality (WE)",
+  "ola_num:merit_effort_cwc" = "Wave x Merit: effort (WE)",
+  "ola_num:merit_talent_cwc" = "Wave x Merit: talent (WE)",
+  "ola_num:perc_inequality_mean" = "Wave x Perception inequality (BE)",
+  "ola_num:merit_effort_mean" = "Wave x Merit: effort (BE)",
+  "ola_num:merit_talent_mean" = "Wave x Merit: talent (BE)"
+    )
+
+texreg::texreg(list(m10.1, m11.1, m12.1, m10.2, m11.2, m12.2),
+               custom.model.names = c("Model 12", "Model 13", "Model 14", "Model 15", "Model 16", "Model 17"),
+               caption = NULL,
+               stars = c(0.05, 0.01, 0.001),
+               custom.coef.map = ccoef,
+               digits = 3,
+               caption.above = T,
+               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.",
+               leading.zero = T,
+               use.packages = F,
+               booktabs = F,
+               scalebox = 0.7,
+               include.loglik = FALSE,
+               include.aic = FALSE,
+               center = T,
+               custom.gof.names = c("BIC", 
+                                    "Numb. obs.", 
+                                    "Num. groups: individuals", 
+                                    "Var: individuals (Intercept)",
+
+                                    "Var: individuals, perception inequality cwc",
+                                    "Var: individuals, wave",
+                                    "Cov: individuals (Intercept), perception inequality cwc",
+                                    "Cov: individuals (Intercept), wave",
+                                    "Cov: individuals, perception inequality cwc, wave",
+                                    "Var: Residuals",
+                                    "Var: individuals, merit effort cwc", 
+                                    "Cov: individuals (Intercept), merit effort cwc",
+                                    "Cov: individuals, merit effort cwc, wave",
+
+                                    "Var: individuals, merit talent cwc", 
+                                    "Cov: individuals (Intercept), merit talent cwc",
+                                    "Cov: individuals, merit talent cwc, wave",
+
+                                    "Var: individuals, perception inequality mean", 
+                                    "Cov: individuals (Intercept), perception inequality mean",
+                                    "Cov: individuals, perception inequality mean, wave",
+
+                                     "Var: individuals, merit effort mean", 
+                                    "Cov: individuals (Intercept), merit effort mean",
+                                    "Cov: individuals, merit effort mean, wave",
+
+                                    "Var: individuals, merit talent mean", 
+                                    "Cov: individuals (Intercept), merit talent mean",
+                                    "Cov: individuals, merit talent mean, wave"
+                                    ),
+               custom.gof.rows = list("Controls"=c(rep("Yes",6))))
+
+```
+
+```{r echo=FALSE, results='asis'}
 #| label: tbl-interactions_olacateg

### 2025-08-08 resubmission/documents for resubmission/Main_blinded.tex
Up carpeta resubmit

--- 2025-08-06 resubmission/paper-blinded-resubmit.tex
@@
 Shaping Preferences for Market Justice in Chile (2016-2023)}
+\author{ }
 
@@
 \begin{abstract}
-This study investigates the relationship between perceptions of economic
-inequality, meritocratic perceptions, and preferences for market justice
-in Chile between 2016 and 2023. Using six waves of panel data from the
-Chilean Longitudinal Social Survey - ELSOC (\(N_{observations}\) =
-8,643; \(N_{individuals}\) = 1,687), the analysis examines how
-subjective assessments of inequality shape attitudes toward the role of
-merit in access to key social services such as healthcare, education,
-and pensions. Results indicate that rising perceptions of inequality are
-associated with lower support for market justice; however, individuals
-with strong meritocratic convictions are more likely to legitimize
-existing disparities. The study also considers the influence of major
-social movements during this period, which appear to have reshaped
-public discourse and perceptions of fairness. These findings contribute
-to a deeper understanding of how beliefs about justice and equity evolve
-in contexts marked by persistent inequality and entrenched
-market-oriented frameworks. \newline \textbf{Keywords}: Economic
-inequality, meritocracy, market justice, Chile, public preferences,
-inequality perception
+Several countries have experienced a shift toward the privatization and
+commodification of public goods, welfare policies, and social services.
+In Latin America, Chile stands out as a paradigmatic case where this
+trend has led to the extensive marketization of essential services. From
+a moral economy perspective, the extent to which individuals consider it
+fair for access to such services to depend on market criteria has been
+conceptualized as market justice preferences. This study investigates
+the relationship between perceptions of economic inequality,
+meritocratic beliefs, and market justice preferences in Chile between
+2016 and 2023. Using six waves of panel data from the Chilean
+Longitudinal Social Survey -- ELSOC (\(N_{observations}\) = 8,643;
+\(N_{individuals}\) = 1,687), the analysis examines how subjective
+assessments of inequality shape attitudes toward the role of merit in
+access to key social services such as healthcare, education, and
+pensions. The findings show that greater perceived inequality is
+associated with lower market justice preferences. However, individuals
+who believe that effort is rewarded are more likely to legitimize
+existing disparities. In contrast, the perception that talent is
+rewarded shows a negative effect on market justice preferences; an
+effect that intensifies as perceived inequality increases over time. The
+study also considers the influence of major social movements during this
+period, which appear to have reshaped public discourse on justice and
+fairness. These findings contribute to a deeper understanding of how
+distributive beliefs evolve in contexts marked by persistent inequality
+and entrenched neoliberal frameworks \newline \textbf{Keywords}:
+Inequality perception, meritocracy, commodification, Chile, public
+preferences
 \end{abstract}
@@
 
-Recent research has highlighted the importance of disaggregating the
+Recent research has highlighted the importance of disentangling the
 concept of ``meritocratic beliefs,'' particularly by distinguishing
@@
 2017}), and lower perceptions of income inequality itself
-(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}). In the
-regard of market justice, prior studies in Chile have shown that
-individuals who perceive higher levels of meritocracy tend to express
-stronger support for unequal distributions that reflect market outcomes
-in healthcare, education and pensions
+(\citeproc{ref-castillo_percepcion_2019}{Castillo et al., 2012}).
+Regarding market justice preferences, prior studies in Chile have shown
+that individuals who perceive higher levels of meritocracy tend to
+express stronger support for unequal distributions that reflect market
+outcomes in healthcare, education and pensions
 (\citeproc{ref-castillo_socialization_2024}{Castillo et al., 2024}).
@@
 and individual well-being by framing welfare outcomes as either deserved
-or undeserved. This psychological mechanism underscores the normative
-power of meritocracy in stabilizing unequal systems by shaping political
+or undeserved. Such mechanism underscores the normative power of
+meritocracy in stabilizing unequal systems by shaping political
 attitudes and personal understandings of success and failure.
@@
 
-Concerning public opinion, the ELSOC survey shows that Chile has a low
-level of agreement with the idea that it is fair for those with higher
-incomes to buy better welfare in the areas of education, healthcare, and
-pensions, with average support of 8.5\% between 2016 and 2023. At the
-same time, around 55\% and 47\% of the public between 2016 and 2023
-\emph{disagreed} that effort and talent are rewarded accordingly. This
-coincides with 90\% of citizens \emph{agreeing} that income differences
-in Chile are \emph{too large}
-(\citeproc{ref-castillo_changes_2025}{Castillo et al., 2025}).
-
-In sum, Chile represents an illustrative case of a particular form of
-\emph{moral economy} (\citeproc{ref-koos_moral_2019}{Koos \& Sachweh,
-2019}), in which national socioeconomic characteristics, institutional
-arrangements, and public attitudes coexist. This suggests that the
-country combines persistently high levels of economic and social
-inequality with a strong institutionalization of market-based mechanisms
-within the welfare system
-(\citeproc{ref-madariaga_three_2020}{Madariaga, 2020}). At the same
-time, perceptions of inequality are high, and there is widespread public
-criticism of the commodification of social services
-(\citeproc{ref-somma_no_2021}{Somma et al., 2021}). For the above
-reasons, we argue that the Chilean case represents a relevant context to
-study the role of perceived income inequality and meritocracy in shaping
-public preferences regarding the commodification of welfare services,
-particularly from the perspective of individual change.
-
 \subsection{This study}\label{this-study}
@@
 individuals endorse market-based distributions of social goods and
-services in Chile.
-
-First, consistent with previous findings about redistributive
-preferences (\citeproc{ref-castillo_perception_2022}{Castillo et al.,
-2022}), we expect that a higher perception of economic inequality will
-be associated with lower market justice preferences. When individuals
+services in Chile. To our knowledge, there are not studies so fay
+analyzing such associations, let alone from a longitudinal perspective.
+
+Firstly, and consistent with previous findings about research on
+redistributive preferences
+(\citeproc{ref-castillo_perception_2022}{Castillo et al., 2022}), we
+expect that a higher perception of economic inequality will be
+associated with lower market justice preferences. When individuals
 perceive smaller income gaps, they are more likely to view market
@@
 with shifts in their market justice attitudes across different time
-points. While we acknowledge the relevance of contextual events --- such
-as the 2019 political outburst or the subsequent constitutional process
---- for public opinion and political orientation
+points. While we acknowledge the relevance of contextual events for
+public opinion and political orientation
 (\citeproc{ref-disipavlic_justification_2025}{Disi Pavlic et al.,
-2025}), identifying causal effects of such events (e.g., mass protest
-exposure) on individual attitudes lies outside the scope of this
-research. Instead, we focus on how intra-individual trajectories in
-perceived inequality and merit shape evolving attitudes toward the
-commodification of social services.
+2025}), such as the 2019 political outburst or the subsequent
+constitutional process in CHile, we are aware that identifying causal
+associations lies outside the scope of this research. Instead, and from
+an exploratory perspective, we focus on how intra-individual
+trajectories in perceived inequality and merit shape evolving attitudes
+toward the commodification of social services.
 
@@
 
-This study tests several key hypotheses concerning the relationships
-between perceptions of economic inequality, meritocratic perceptions,
-and preferences toward market justice in Chile. It posits that higher
-perceived inequality and stronger meritocratic perceptions both
-independently contribute to greater support for market-based
-distributions of social goods. Furthermore, the study examines how
-perceived economic inequality moderates the influence of meritocratic
-perceptions on market justice preferences. Additionally, it considers
-how major social and political mobilizations, such as those occurring
-around 2019 and subsequent years, may have weakened these relationships,
-reflecting increased critical perspectives on systemic inequalities and
-market mechanisms. The subsequent discussion interprets the empirical
-findings in light of these hypotheses, considering their implications
-for understanding attitudes toward inequality and social justice in a
-context marked by profound social change.
-
-The first set of hypothesis suggested that perceptions of economic
-inequality were strongly linked to preferences for market justice. The
-findings demonstrated that individuals perceiving higher income
-disparities showed less support for the idea that the access to core
-social services depends on individual income, aligning with theories
-emphasizing that awareness of inequality could fuel a critical view of
-such policies (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
+The first set of hypotheses proposed that perceptions of economic
+inequality are strongly associated with preferences for market justice.
+In the between-individual specification, the effect of perceived
+inequality is negative but not statistically significant. Therefore, we
+cannot conclude that individuals who, on average, perceive higher levels
+of inequality are systematically less supportive of market justice. As a
+result, hypothesis \(H1_{a}\) is not supported by the data. However, in
+the within-individual specification, the findings reveal a significant
+negative association: individuals who, at a given point in time,
+perceive higher income disparities are less supportive of the idea that
+access to core social services should depend on individual income. This
+result aligns with theories suggesting that increased awareness of
+inequality fosters a more critical stance toward market-based
+distributive arrangements
+(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
 \citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al., 2019};
-\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). Such a relationship
-reflects a moral economy in which perceptions of systemic unfairness
-undermine the legitimacy of existing distributions and fuel demands for
-greater equity. In particular, the negative within-individual effect
-observed over time (\(H1_{b}\)) suggests that as inequality becomes more
+\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). Such a pattern reflects a
+moral economy logic in which perceptions of systemic unfairness
+undermine the legitimacy of existing distributions and strengthen
+demands for greater equity. Specifically, the negative within-individual
+effect over time (\(H1_{b}\)) suggests that as inequality becomes more
 salient for some individuals, their support for market justice
-declines---potentially due to rising distrust in market mechanisms or
-growing disillusionment with the fairness of the system. However,
-although the effect of perceived inequality at the between-individual
-level is also negative, it is not statistically significant. Therefore,
-we cannot conclude that individuals who, on average, perceive higher
-inequality are systematically less supportive of market justice. As a
-result, hypothesis \(H1_{a}\) does not receive empirical support.
+declines---possibly due to growing distrust in market mechanisms or
+increasing disillusionment with the perceived fairness of the system.
 
@@
 (\citeproc{ref-castillo_socialization_2024}{2024}) on Chilean students,
-the findings suggest that such meritocratic perceptions uphold existing
+the results suggest that such meritocratic perceptions uphold existing
 hierarchies by promoting the acceptance of inequality as a reflection of
@@
 possible explanation is that effort is generally viewed as a
-controllable and malleable trait, aligning with the normative
-foundations of market fairness. In contrast, talent tends to be
-perceived as innate and less subject to personal control, rendering
-talent-based inequality less legitimate. Furthermore, increased exposure
-to real-world scenarios in which outcomes are clearly shaped by inherent
+controllable and malleable trait, whereas talent tends to be perceived
+as innate and less subject to personal control, rendering talent-based
+inequality less legitimate. Furthermore, increased exposure to
+real-world scenarios in which outcomes are clearly shaped by inherent
 traits rather than hard work may lead individuals to question the
@@
 market-based distribution systems tends to be stronger when perceived
-economic inequality is high---contradicting our initial expectations
+economic inequality is high, contradicting our initial expectations
 (\(H3_{a}\) and \(H3_{b}\)). Although this moderating effect is not
@@
 it is significant for talent-related perceptions at both the within- and
-between-individual levels. This finding runs counter to our hypotheses,
-as it suggests that the effect of meritocratic
-perceptions---specifically, talent-based meritocracy---intensifies as
-perceptions of inequality increase. However, this interpretation
-requires caution. The main effect of perceived inequality on support for
-market justice is negative. Thus, as perceived inequality becomes less
+between-individual levels. However, this interpretation requires
+caution: the main effect of perceived inequality on support for market
+justice is negative, meaning that as perceived inequality becomes less
 negative (i.e., closer to zero), the positive relationship between
@@
 
-Finally, the fourth hypothesis examined the potential impact of recent
-social mobilizations---particularly those that erupted in 2019---on the
+Finally, the fourth exploratory hypothesis examined the potential impact
+of social events --- particularly those that erupted in 2019 --- on the
 relationship between meritocratic perceptions, perceptions of
 inequality, and preferences for market justice. The main effect of time
-indicates that, on average, support for market justice increased
-following the 2019 protests. However, the analysis did not detect any
-significant interactions between distributive beliefs and support for
-market justice, suggesting that the social mobilizations did not
-substantially modify the underlying associations between these
-variables. As a result, hypotheses \(H4_{a}\) and \(H4_{b}\) are not
-supported by the data. This may point to a certain stability---or even
-resilience---in the normative frameworks that guide individuals'
-evaluations of distributive justice, despite the occurrence of major
-collective political events. Alternatively, while the protests may have
-eroded trust in institutional arrangements or governance, they may not
-have fundamentally altered individuals' beliefs about how rewards and
-resources should be allocated in society.
+indicates that, on average, support for market justice increased after
+2019 protests. However, the analysis did not detect any significant
+interactions between distributive beliefs and support for market
+justice, suggesting that events in this period (as the mobilizations,
+the pandemic and/or the constitucional processes) did not substantially
+modify the underlying associations between these variables. As a result,
+hypotheses \(H4_{a}\) and \(H4_{b}\) are not supported by the analysis.
+This may point to a certain stability or even resilience in the
+normative frameworks that guide individuals' evaluations of distributive
+justice, despite the occurrence of major collective political events.
+Alternatively, while the protests may have eroded trust in institutional
+arrangements or governance, they may not have fundamentally altered
+individuals' beliefs about how rewards and resources should be allocated
+in society.
 
@@
 underpinning social justice attitudes in a highly unequal and
-commoditized environment.
-
-The findings confirm that higher perceptions of economic inequality are
+commodified environment.
+
+The findings support that higher perceptions of economic inequality are
 associated with less support for market justice preferences, this is,
@@
 social services such as education, pensions and health. At the same
-time, meritocratic perceptions---particularly those emphasizing
-individual effort---are strongly associated with support for
+time, meritocratic perceptions --- particularly those emphasizing
+individual effort --- are strongly associated with support for
 market-based distributions, suggesting that meritocracy serve as a moral
@@
 legitimizing power of meritocracy becomes stronger as perceptions of
-inequality become less negative---that is, when individuals are less
-aware or concerned about economic disparities. This finding highlights a
+inequality become less negative --- that is, when individuals are less
+aware or concerned about economic disparities. Such finding highlights a
 potential feedback mechanism by which lower sensitivity to inequality
@@
 dynamic and interactional nature of perceptions and beliefs over time.
-It highlights the importance of socio-political upheavals in reshaping
-normative attitudes, underscoring the role of collective action in
-challenging entrenched narratives of meritocracy and fairness. The
-longitudinal approach provides a deeper temporal perspective on how
+Furthermore, it highlights the importance of socio-political upheavals
+in reshaping normative attitudes, underscoring the role of collective
+action in challenging entrenched narratives of meritocracy and fairness.
+The longitudinal approach provides a deeper temporal perspective on how
 societal events could influence individual perceptions and preferences.
 
-Despite its contributions, the study has several limitations. First, the
-reliance on self-reported perceptions and attitudes may be subject to
-social desirability bias. Second, the focus on Chile, while offering
-valuable insights into a specific context, limits the generalizability
-of results to other countries with different institutional and cultural
-backgrounds. Third, the observational nature of the data prevents causal
-inferences, and unobserved confounding factors may influence the
-observed relationships. Finally, the potential influence of media,
-political discourse, and peer effects on perceptions was not directly
-assessed.
-
-Future research could explore the causal mechanisms linking perceptions
-and attitudes, possibly through experimental or mixed-method designs.
-Comparative studies across different national contexts would help
-determine the universality or specificity of these dynamics.
-Additionally, investigating the role of media, political communication,
-and education in shaping perceptions of inequality and meritocracy would
-deepen understanding of the normative foundations of social justice
-attitudes. Finally, examining how these perceptions influence behavioral
-outcomes, such as political participation or support for social
-movements, would provide valuable insights into the pathways from
-beliefs to collective action and policy change.
+Regarding avenues for future research, experimental evidence could help
+expand the understanding of the association between perceptual variables
+and market justice preferences. International comparative studies are
+also necessary in order to assess the role of national contexts and the
+universality or specificity of these dynamics. Additionally,
+investigating the role of media, political communication, and education
+in shaping perceptions of inequality and meritocracy would deepen
+understanding of the normative foundations of social justice attitudes
+and their variation over time. Finally, examining how these perceptions
+influence behavioral outcomes, such as political participation or
+support for social movements, would provide valuable insights into the
+pathways from beliefs to collective action and policy change.
 
@@
 \url{https://doi.org/10.1080/02134748.2021.2009275}
-
-\bibitem[\citeproctext]{ref-castillo_changes_2025}
-Castillo, J. C., Iturra, J., \& Carrasco, K. (2025). Changes in the
-{Justification} of {Educational Inequalities}: {The Role} of
-{Perceptions} of {Inequality} and {Meritocracy During} the {COVID
-Pandemic}. \emph{Social Justice Research}.
-\url{https://doi.org/10.1007/s11211-025-00458-0}
 

### 2025-08-08 resubmission/paper-blinded-resubmit.qmd
Up carpeta resubmit

--- 2025-07-11 paper.qmd
@@
 editor: source
-bibliography: "input/bib/merit-edjust.bib"
+bibliography: "../input/bib/merit-edjust.bib"
 # csl: "input/bib/apa6.csl"
-csl: "input/bib/harvard.csl"
-author:
-  - name: Juan Carlos Castillo
-    corresponding: true
-    orcid: 0000-0003-1265-7854
-    email: juancastillov@uchile.cl
-    affiliations:
-      - ref: 1
-      - ref: 2
-      - ref: 3
-  - name: Andreas Laffert
-    orcid: 0000-0002-9008-2454
-    email: alaffertt@estudiante.uc.cl 
-    affiliations:
-      - ref: 4
-  - name: Kevin Carrasco
-    orcid: 0000-0002-9348-8361
-    email: kevin.carrasco@ug.uchile.cl
-    affiliations:
-      - ref: 2
-  - name: Julio Iturra-Sanhueza
-    orcid: 0000-0003-1993-3479
-    email: julioiturrasanhueza@gmail.com 
-    affiliations:
-      - ref: 5     
-affiliations: 
-  - id: 1
-    name: Departamento de Sociología, Universidad de Chile
-  - id: 2
-    name: Centro de estudios del conflicto y cohesión social (COES)
-  - id: 3
-    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
-  - id: 4
-    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
-  - id: 5
-    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
+csl: "../input/bib/harvard.csl"
 citeproc: true
 abstract: | 
-  This study investigates the relationship between perceptions of economic inequality, meritocratic perceptions, and preferences for market justice in Chile between 2016 and 2023.
-Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions.
-Results indicate that rising perceptions of inequality are associated with lower support for market justice; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities.
-The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness.
-These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
+ Several countries have experienced a shift toward the privatization and commodification of public goods, welfare policies, and social services.
+In Latin America, Chile stands out as a paradigmatic case where this trend has led to the extensive marketization of essential services.
+From a moral economy perspective, the extent to which individuals consider it fair for access to such services to depend on market criteria has been conceptualized as market justice preferences.
+This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and market justice preferences in Chile between 2016 and 2023.
+Using six waves of panel data from the Chilean Longitudinal Social Survey – ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions.
+The findings show that greater perceived inequality is associated with lower market justice preferences.
+However, individuals who believe that effort is rewarded are more likely to legitimize existing disparities.
+In contrast, the perception that talent is rewarded shows a negative effect on market justice preferences; an effect that intensifies as perceived inequality increases over time.
+The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse on justice and fairness.
+These findings contribute to a deeper understanding of how distributive beliefs evolve in contexts marked by persistent inequality and entrenched neoliberal frameworks
   \newline
-  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
+ **Keywords**: Inequality perception, meritocracy, commodification, Chile, public preferences
 link-citations: true

### 2025-08-08 resubmission/paper-blinded-resubmit.tex
Up carpeta resubmit

@@
 
-This study tests several key hypotheses concerning the relationships
-between perceptions of economic inequality, meritocratic perceptions,
-and preferences toward market justice in Chile. It posits that higher
-perceived inequality and stronger meritocratic perceptions both
-independently contribute to greater support for market-based
-distributions of social goods. Furthermore, the study examines how
-perceived economic inequality moderates the influence of meritocratic
-perceptions on market justice preferences. Additionally, it considers
-how major social and political mobilizations, such as those occurring
-around 2019 and subsequent years, may have weakened these relationships,
-reflecting increased critical perspectives on systemic inequalities and
-market mechanisms. The subsequent discussion interprets the empirical
-findings in light of these hypotheses, considering their implications
-for understanding attitudes toward inequality and social justice in a
-context marked by profound social change.
-
-The first set of hypothesis suggested that perceptions of economic
-inequality were strongly linked to preferences for market justice. The
-findings demonstrated that individuals perceiving higher income
-disparities showed less support for the idea that the access to core
-social services depends on individual income, aligning with theories
-emphasizing that awareness of inequality could fuel a critical view of
-such policies (\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
+The first set of hypotheses proposed that perceptions of economic
+inequality are strongly associated with preferences for market justice.
+In the between-individual specification, the effect of perceived
+inequality is not statistically significant. Therefore, we cannot
+conclude that individuals who, on average, perceive higher levels of
+inequality are systematically less supportive of market justice. As a
+result, hypothesis \(H1_{a}\) is not supported by the data. However, in
+the within-individual specification, the findings reveal a significant
+negative association: individuals who, at a given point in time,
+perceive higher income disparities are less supportive of the idea that
+access to core social services should depend on individual income. This
+result aligns with theories suggesting that increased awareness of
+inequality fosters a more critical stance toward market-based
+distributive arrangements
+(\citeproc{ref-Castillo2012a_justice}{Castillo, 2012};
 \citeproc{ref-garcia-sanchez_vicious_2019}{García-Sánchez et al., 2019};
-\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). Such a relationship
-reflects a moral economy in which perceptions of systemic unfairness
-undermine the legitimacy of existing distributions and fuel demands for
-greater equity. In particular, the negative within-individual effect
-observed over time (\(H1_{b}\)) suggests that as inequality becomes more
+\citeproc{ref-mijs_paradox_2021}{Mijs, 2021}). Such a pattern reflects a
+moral economy logic in which perceptions of systemic unfairness
+undermine the legitimacy of existing distributions and strengthen
+demands for greater equity. Specifically, the negative within-individual
+effect over time (\(H1_{b}\)) suggests that as inequality becomes more
 salient for some individuals, their support for market justice
-declines---potentially due to rising distrust in market mechanisms or
-growing disillusionment with the fairness of the system. However,
-although the effect of perceived inequality at the between-individual
-level is also negative, it is not statistically significant. Therefore,
-we cannot conclude that individuals who, on average, perceive higher
-inequality are systematically less supportive of market justice. As a
-result, hypothesis \(H1_{a}\) does not receive empirical support.
+declines---possibly due to growing distrust in market mechanisms or
+increasing disillusionment with the perceived fairness of the system.
 

### 2025-09-10 resubmission/07-supplementary-material.qmd
changes supplementary material

@@
 options(knitr.kable.NA = "")
+
 ```
@@
 rm(list = ls())
+
 ```
@@
 #| fig-asp: 1
-#| fig-cap: "Correlation matrix of the main variables for the last wave (2023)"
+#| fig-cap: "Correlation matrix of the main variables for the first wave (2016)"
 #| fig-cap-location: top
@@
                    na.label = "-") 
+grid::grid.text("Source: own elaboration with data from ELSOC 2016 (n = 1687)", x = 0.99, y = 0.1, just = "right", gp = grid::gpar(fontsize = 12, col = "black"))
 
@@
                              "Age (Ref.= 18-29)" = 26:28),
-               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.",
+               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars.
+Source: own elaboration with pooled data from ELSOC 2016-2023 (N obs = 8643; N groups = 1687)",
                leading.zero = T,
@@
                caption.above = T,
-               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.",
+               custom.note = "Note: Cells contain regression coefficients with standard errors in parentheses. %stars. CWC = centered within group.
+Source: own elaboration with pooled data from ELSOC 2016-2023 (N obs = 8643; N groups = 1687)",
                leading.zero = T,
